### 2019-08-14 content/_index.Rmd
Customize course visuals

---
title: "Econometrics"
author: "Ryan Safner"
---

^[{-} `r blogdown::shortcode("courseinfo")`] <span class="newthought">This site contains</span> the syllabus, schedule, and assignments for ECON 480: Econometrics, held during Fall 2019 at Hood College.

By the end of this course, you will:

1. understand how to evaluate statistical and empirical claims;
2. use the fundamental models of causal inference and research design;
2. gather, analyze, and communicate with real data in R.

<figure>
[![](/images/metrics_flowchart_blankbg.png)](/images/metrics_flowchart_blankbg.png)
</figure>

### 2019-08-14 content/syllabus.Rmd
Customize course visuals

---
title: Syllabus
slug: syllabus
citeproc: false
bibliography: ../static/bib/references.bib
csl: ../static/bib/chicago-syllabus-no-bib.csl
---

```{r load-libraries, message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
library(tibble)
library(pander)
```

- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/files/syllabus.pdf)

<span class="newthought">Econometrics</span> [^courseinfo] is the application of statistical tools to quantify and measure economic relationships in the real world. It uses real data to test economic hypotheses, quantitatively estimate causal relationships between economic variables, and to make forecasts of future events. The primary tool that economists use for empirical analysis is ordinary least squares (OLS) linear regression, so the majority of this course will focus on understanding, applying, and extending OLS regressions.

I assume you have *some* working knowledge of  economics at the intermediate level and some basic statistical tools.^[The formal prerequisites for this course are **ECON 205** and **ECON 206**; **ECMG 212** or **MATH 112**; and **ECON 305** or **ECON 306**] We will do some basic review of some necessary statistics and probability at the beginning until everyone is comfortable, before jumping right into regressions.

[^courseinfo]: {-}
  `r blogdown::shortcode("courseinfo")`

# Course objectives

<span class="newthought">By the end of this course,</span> you will:

1. understand how to evaluate statistical and empirical claims;
2. use the fundamental models of causal inference and research design;
2. gather, analyze, and communicate with real data in R.

I am less concerned with forcing you to memorize and recite proofs of statistical estimator properties, and more concerned with the development of your intuitions and the ability to think critically as an empirical social scientist---although this will require you to demonstrate proficiency with some intermediate statistical and mathematical tools.

Given these objectives, this course fulfills three of the four learning outcomes for BYU's [Master of Public Administration (MPA)](https://marriottschool.byu.edu/mpa/) program:

- **Quantitative Analysis**: BYU MPA graduates are skilled at evaluating programs and policies. They know how to gather data, correctly analyze it, and employ the analysis to recommend policy and programmatic action in public service organizations.
- **Public Service Values**: BYU MPA graduates demonstrate an understanding of, passion for, and commitment to public service values, including reverence for the dignity and worth of all people and dedication to ethical governance.
- **Communication**: BYU MPA graduates effectively convey verbal and written information with the polish and professionalism appropriate for the public service context. They listen to and promote understanding among people with diverse viewpoints.

<span class="newthought">\textbf{Fair warning:}</span> **Econometrics is hard.** *It will be one of the hardest economics courses that you will take, primarily due to the mathematical content.* I will do my best to make this class intuitive and helpful, if not interesting. If at any point you find yourself struggling in this course for any reason, please come see me. Do not suffer in silence! Coming to see me for help does not diminish my view of you, in fact I will hold you in *higher* regard for understanding your own needs and taking charge of your own learning. There are also a some fantastic resources on campus, such as the [Center for Academic Achievement and Retention (CAAR)](http://www.hood.edu/campus-services/academic-services/index.html) and the [Beneficial-Hodson Library](http://www.hood.edu/library/}{Beneficial-Hodson Library).

# Course materials

<span class="newthought">Most of the readings in this class are free.</span>

## Books

The following book is **required**^[You are not *obligated* to buy it, I just **strongly recommend** it in the sense that you will still have access to all data and assignments without possessing the book. But this is a course where you really will want to understand the derivations or get additional context beyond just my slides...] and will be available from the campus bookstore.

- [@Bailey], 2^nd^ ed. ISBN: 978-0190857462

You are welcome to purchase the book by other means (e.g. [Amazon](https://www.amazon.com/Real-Econometrics-Answer-Important-Questions/dp/0190857463/ref=dp_ob_title_bk), half.com, etc). I have no financial stake in requiring you to purchase this book. The (cheaper) 1st edition is sufficient, but makes significantly less use of `R` (in favor of `STATA`).

## Software

You are **strongly recommended** to download copies of [`R`](https://www.r-project.org/) and [`R Studio`](http://www.rstudio.com) on your own computers. These software packages are available on all computers in the trading room, and you will have access to them during the week to work on assignments.

We will also have a shared class workspace in [RStudio.cloud](http://rstudio.cloud/) that runs a full instance of R Studio in your web browser (so no need to install anything!) will let you access files and assignments.

## Articles

Throughout the course, I will post both required and supplemental (non-required) readings that enrich your understanding for each topic on Blackboard. Check Blackboard *frequently* for announcements and updates to assignments, readings, and grades.

## Online help and Slack

Economics can be difficult. Computer programming can be difficult. Computers are stupid and little errors in your code can cause hours of headache (even if you've been doing this stuff for years!).

Fortunately there are tons of online resources to help you with this. Two of the most important are [StackOverflow](https://stackoverflow.com/) (a Q&A site with hundreds of thousands of answers to all sorts of programming questions) and [RStudio Community](https://community.rstudio.com/) (a forum specifically designed for people using RStudio and the tidyverse (i.e. you)).

Searching for help with R on Google can be tricky because the program is, um, a single letter. Try searching for "rstats" instead. If you use Twitter, post R-related questions and content with [#rstats](https://twitter.com/search?q=%23rstats).

Additionally, we have a class chatroom at [Slack](https://hoodeconometrics.slack.com/) where anyone in the class can ask questions and anyone can answer. Ask questions about the readings or problem sets [in the class Slack workspace](https://hoodeconometrics.slack.com/). I will monitor Slack regularly, and you should also all do so as well. You'll likely have similar questions as your peers, and you'll likely be able to answer other peoples' questions too.

# Course policies

<span class="newthought">Be nice. Be honest. Don't cheat.</span>

We will also follow the [full list of Marriott School and BYU classroom policies](https://marriottschool.byu.edu/policies/classroom/).

## Counseling and Psychological Services (CAPS)

Life at BYU can be complicated and challenging. You might feel overwhelmed, experience anxiety or depression, or struggle with relationships or family responsibilities. [Counseling and Psychological Services (CAPS)](https://caps.byu.edu/) provides free, *confidential* support for students who are struggling with mental health and emotional challenges. The CAPS office is staffed by professional psychologists who are attuned to the needs of all types of college and professional students. Please do not hesitate to contact CAPS for assistance—getting help is a smart and courageous thing to do.

# Assignments and grades

<span class="newthought">You can find descriptions</span> for all the assignments on the [assignments page](/assignments/).

```{r assignments-grades, include=FALSE}
assignments <- tribble(
  ~"",  ~Assignment, ~Percent,
  1     , "Research Project", "30%",
  "n"  , "Problem sets (Average)", "25%",
  1      , "Midterm", "20%",
  1     , "Final", "25%"
)

grading <- tribble(
  ~Grade, ~Range, ~Grade1, ~Range1,
  "A",  "93–100%", "C",  "73–76%",
  "A−", "90–92%", "C−", "70–72%",
  "B+", "87–89%", "D+", "67–69%",
  "B",  "83–86%", "D",  "63–66%",
  "B−", "80–82%", "D−", "60–62%",
  "C+", "77–79%", "F",  "< 60%"
)
```

```{r}
knitr::kable(assignments)
```

```{r show-assignments-table, eval=F, echo=FALSE, results="asis"}
assignments_total <- tibble(Assignment = "Total",
                            Points = sum(assignments$Points),
                            Percent = NA)

bind_rows(assignments, assignments_total) %>% 
  select(Assignment, Points, Percent) %>% 
  pandoc.table(justify = "lcc", missing = "—")
```

\

```{r show-grades-table, echo=FALSE, results="asis"}
grading %>%
  pandoc.table(justify = "lclc", 
               col.names = c("Grade", "Range", "Grade", "Range"))
```

# Red pandas

[^panda] <span class="newthought">Once you have read</span> this entire syllabus and [the assignments page](/assignments/), please [click here](mailto:andrew_heiss@byu.edu) and e-mail me a picture of a red panda. For real. Brownie points if it's animated.

[^panda]: {-}
  ![Hungry red panda](https://media.giphy.com/media/r7Nbr97FsXxXG/giphy.gif)

### 2019-08-14 content/_index.Rmd
Updated learning goals

@@
 2. use the fundamental models of causal inference and research design;
-2. gather, analyze, and communicate with real data in R.
+3. gather, analyze, and communicate with real data in R.
+
+Given these objectives, this course fulfills all three of the learning outcomes for [the George B.
+Delaplaine, Jr.
+School of Business](https://www.hood.edu/academics/departments/george-b-delaplaine-jr-school-business/student-learning-outcomes) Economics B.A.
+program:
+
+- Use quantitative tools and techniques in the preparation, interpretation, analysis and presentation of data and information for problem solving and decision making [...]
+- Apply economic reasoning and models to understand and analyze problems of public policy [...]
+- Demonstrate effective oral and written communications skills for personal and professional success[...]
 

### 2019-08-14 content/syllabus.Rmd
Updated learning goals

@@
 2. use the fundamental models of causal inference and research design;
-2. gather, analyze, and communicate with real data in R.
+3. gather, analyze, and communicate with real data in R.
 
@@
 
-Given these objectives, this course fulfills three of the four learning outcomes for BYU's [Master of Public Administration (MPA)](https://marriottschool.byu.edu/mpa/) program:
+Given these objectives, this course fulfills all three of the learning outcomes for [the George B.
+Delaplaine, Jr.
+School of Business](https://www.hood.edu/academics/departments/george-b-delaplaine-jr-school-business/student-learning-outcomes) Economics B.A.
+program:
 
-- **Quantitative Analysis**: BYU MPA graduates are skilled at evaluating programs and policies.
-They know how to gather data, correctly analyze it, and employ the analysis to recommend policy and programmatic action in public service organizations.
-- **Public Service Values**: BYU MPA graduates demonstrate an understanding of, passion for, and commitment to public service values, including reverence for the dignity and worth of all people and dedication to ethical governance.
-- **Communication**: BYU MPA graduates effectively convey verbal and written information with the polish and professionalism appropriate for the public service context.
-They listen to and promote understanding among people with diverse viewpoints.
+- Use quantitative tools and techniques in the preparation, interpretation, analysis and presentation of data and information for problem solving and decision making [...]
+- Apply economic reasoning and models to understand and analyze problems of public policy [...]
+- Demonstrate effective oral and written communications skills for personal and professional success[...]
 

### 2019-08-22 content/reference.Rmd
Cleared out old content, updated syllabus, reference (tips for success), and schedule based on devF19

---
title: Reference
slug: reference
output:
  blogdown::html_page:
    template: ../pandoc/toc-title_html.template
    toc: true
---

```{r include=FALSE, eval=F}
library(tidyverse)
library(emo)
library(here)

source(here("lib", "graphics.R"))
```

# Tips for Success {#tips}

1. **Take notes. On paper. Yes.** [Studies show](https://journals.sagepub.com/doi/abs/10.1177/0956797614524581) that using pen and paper trump taking notes on a device. The main reason is because *you cannot possibly write down everything I am saying by hand*. This forces you to selectively filter my lecture to its most essential and important components -- an integral part of the learning and retention process. If you are taking notes on a laptop, you become a court stenographer, thoughtlessly transcribing everything I am saying. At the other extreme, do not assume *everything* is written in my lecture slides. My lecture slides are visuals and cues to organize the content both for me and for you. I try hard to make sure that I have few words on slides, and even fewer answers to problems. 
    - My suggestion: print out my lecture slides in advance (or follow along on your device) and take additional notes by hand.
2. **At least skim all of the readings.** I give out readings for a reason, and that reason is not to bore you or waste your time. The truth is, as a beginner, you can't rely on heuristics or memory to "fill in the blanks." You need the readings to provide context to what I am saying in class. I cannot help you if you are not going to try. Any professor also will lose patience at short notice when it's discovered you haven't done the reading.
3. **Do the homeworks** (if there are any). You might be surprised that I need to say this, but I do. Students that do not do the homeworks do poorly on exams, which often are similar to homework questions. Homeworks are *the best* practice for exams, they give you a sense of the relevant content areas that might come up, the type and style of questions that I ask, and you often get answer keys to help you "get inside my head" and study from. There is no good reason why you should have a low homework grade.
4. **Work on assignments together and study together**. Recognize that you are not in this alone, and other students are just as anxious or uncertain as you are. More surprisingly, other students probably have some answers you are looking for, and you may have answers for some of their problems! You will learn better when you collaborate with others similar to you. More importantly, you don't truly understand something unless you can explain it to others.^[Yes, that means I am doing a *ton* of learning every time I teach!] As a wise woman once told me "whomever is doing the talking is doing the learning."
5. **Learn how to learn**. The most important skill you learn in college is *how to learn*. Taking a course on a subject will *not* make you an expert on that subject. It a) helps you recognize that you do not know everything on that subject, and this prevents you from [actively saying stupid things](http://www.smbc-comics.com/?id=2475); and b) gives you enough context and skills to figure out how to *actually* fill those gaps. This is the actual skill that's relevant in the real world.^[Yes, Google is your best friend. But you do not yet know how to ask the right questions, or understand what constitutes good answers.]

# Installing R and R Studio

1. [Install R](http://www.r-project.org/) from CRAN^[The Comprehensive R Archive Network] by clicking "Download R" (or the CRAN link under Downloads on the left menu). This will take you to a mirrors page, where you can select a location in the U.S. and download a copy of R
2. [Install R Studio (Desktop Version)](https://www.rstudio.com/products/rstudio/download/), choose the "Free" option

# R Packages

Packages come from multiple sources.

The polished, publicly released versions are found on [CRAN](https://cran.r-project.org/). When installing a package available on CRAN, it is sufficient simply to tell R the following:^[Note the plural `s` on `packages`, and the quotes around the "package name"]

```{r, eval=F}
install.packages("packagename") 
```

Other packages, which may be in various developmental states (including perfectly functional!) are often hosted on [GitHub](http://github.com) before they make their way onto CRAN. Simply telling R `install.packages("packagename")` will fail to find it (as R only looks in CRAN for packages), so you must use another package called `devtools`^[Which you will need to install first if you (probably) don't already have it!] to install packages directly from Github:^[Note the `::` allows you to use the function `install_github()` from the `devtools` package without having to first load the devtools package with `library(devtools)`.]

```{r, eval=F}
devtools::install_github("username/packagename") 
```

For example, to install Hadley Wickham's package `r4ds` from its Github page [https://github.com/hadley/r4ds](https://github.com/hadley/r4ds), we would type:

```{r, eval=F}
devtools::install_github("hadley/r4ds")
```

# Getting Help for R

For specific functions or commands, you can simply type:

```{r, eval=F}
?functionname()

# example
?mean()
```

This will display a help page specific to that function in the Viewer pane. R functions and packages are extremely well-documented; help pages normally include a short description of the function, arguments and options (as well as their default values), and several examples or vignettes to demonstrate usage of the function.

Additionally, you can turn to the community.

# R Markdown

## Math and $\LaTeX{}$

Math in R Markdown uses the $\LaTeX{}$ language to typeset beautiful formulas and equations. To include mathematical symbols or expressions inline^[That means, within the text, and not in a separate line, location, or environment within the document.], insert it `$`between two dollar signs`$`. Within a sentence, `$2^2+\frac{\pi}{\pi}=5$` becomes $2^2+\frac{\pi}{\pi}=5$.

If you prefer it to be centered in its own line, put it on its own line, `$$`between two dollar signs`$$`.

`$``$``2^2+\frac{\pi}{\pi}=5``$``$`` becomes:

$$2^2+\frac{\pi}{\pi}=5$$

There is also a shortcut with the R package `equatiomatic`.

## Producing documents

### Producing a PDF Output

PDFs are archaic document formats [with many flaws](https://www.urban.org/urban-wire/your-data-deserve-better-pdf), yet the format remains the industry standard in most cases. PDFs are made with $\LaTeX{}$, a typography language dating to the 1980s. $\LaTeX{}$ [has its own problems](https://www.authorea.com/users/5713/articles/19359-latex-is-dead-long-live-latex-typesetting-in-the-digital-age/_show_article)

One alternative is [tinytex](https://yihui.name/tinytex/)

```{r, eval=F}
install.packages("tinytex")
tinytex::install_tinytex()  # install TinyTeX
```

# Resources

## Cheat Sheets

# R Packages

The following is a compendium of all R packages used, and their main uses

- `tidyverse`
- `DT`
- `knitr`
- `haven`
- `readxl`
- `dplyr`
- `ggplot2`
- `car`
- `sandwich`
- `lmtest`
- `equatiomatic`
- `gganimate`
- `stargazer`
- `broom`
- `modelr`
- `infer`
- `aer`

# Finding Data

- A [near-comprehensive list](https://vincentarelbundock.github.io/Rdatasets/datasets.html) of all existing data sets built-in to R or R packages^[Note: You should use these more for playing around with in R to boost your data wrangling skills. These should not be used for your projects in most circumstances.]

# Citations and bibliography

[^bibtex] You can [download a BibTeX file](/bib/references.bib) of all the non-web-based readings in the course.

[^bibtex]: {-} You can open the file in [BibDesk on macOS](http://bibdesk.sourceforge.net/), [JabRef on Windows](http://www.jabref.org/), or [Zotero](https://www.zotero.org/) or [Mendeley](https://www.mendeley.com/) online.

### 2019-08-22 content/schedule.Rmd
Cleared out old content, updated syllabus, reference (tips for success), and schedule based on devF19

---
title: "Schedule"
slug: schedule
---

This page contains all of the following resources for each class meeting:

- <i class="fas fa-book-reader"></i> **Readings** include textbook chapters and occasional journal articles
- <i class="fas fa-laptop-code"></i> **Assignments** are due by the beginning of class unless otherwise stated
- <i class="fas fa-info-circle"></i> **Class** materials contain extra explanations, math/statistical properties & proofs, and other helpful resources^[These "online appendices" keep the slides nice and de-cluttered!]
- <i class="fas fa-chalkboard-teacher"></i> **Slides** are "[Xaringan](https://github.com/yihui/xaringan)" presentations in html that can be opened in any browser^[`File > Print > Save as PDF` if you wish to download them]
- <i class="fas fa-registered"></i> **R** materials contain extra tutorials, videos, practice exercises for using `R`

Relevant materials (if applicable, icons will become links) will be posted before class meets.

**Last Update**: [`r Sys.time()`](https://github.com/ryansafner/metricsF19/commits/master).

---

`r blogdown::shortcode("schedule")`

### 2019-08-22 content/syllabus.Rmd
Cleared out old content, updated syllabus, reference (tips for success), and schedule based on devF19

@@
 - [<i class="fas fa-file-pdf"></i> `Download as PDF`](/files/syllabus.pdf)
+- **Last Updated**: 2019-08-22
 
@@
 
-<span class="newthought">\textbf{Fair warning:}</span> **Econometrics is hard.** *It will be one of the hardest economics courses that you will take, primarily due to the mathematical content.* I will do my best to make this class intuitive and helpful, if not interesting.
+<span class="newthought">Fair warning:</span> **Econometrics is hard.** *It will be one of the hardest economics courses that you will take, primarily due to the mathematical content.* I will do my best to make this class intuitive and helpful, if not interesting.
 If at any point you find yourself struggling in this course for any reason, please come see me.
@@
 
+See my [tips for success in this course](/reference#tips).
+
 # Course materials
-
-<span class="newthought">Most of the readings in this class are free.</span>
 
@@
 
-- [@Bailey], 2^nd^ ed. ISBN: 978-0190857462
+- Bailey, Michael A, 2019, *Real Econometrics*, New York: Oxford University Press, 2^nd^ ed.
 
 You are welcome to purchase the book by other means (e.g.
-[Amazon](https://www.amazon.com/Real-Econometrics-Answer-Important-Questions/dp/0190857463/ref=dp_ob_title_bk), half.com, etc).
+Amazon, half.com, etc).
 I have no financial stake in requiring you to purchase this book.
 The (cheaper) 1st edition is sufficient, but makes significantly less use of `R` (in favor of `STATA`).
+
+The following two books are **recommended**, and are free online^[You can purchase a hard copy of the first one if you really want.]:
+
+- Grolemund, Garrett and Hadley Wickham, [*R For Data Science*](https://r4ds.had.co.nz/)
+- Ismay, Chester and Albert Y Kim, [*Modern Dive: Statistical Inference Via Data Science*](https://moderndive.com/)
+
+The first book is the number one resource for using `R` and `tidyverse`, and is written for beginners.
+I still look at it *frequently.* The second is another great reference for using `tidyverse` in the context of basic statistics.
 
@@
 
-Throughout the course, I will post both required and supplemental (non-required) readings that enrich your understanding for each topic on Blackboard.
+Throughout the course, I will post both required and supplemental (non-required) readings that enrich your understanding for each topic.
 Check Blackboard *frequently* for announcements and updates to assignments, readings, and grades.
-
-## Online help and Slack
-
-Economics can be difficult. Computer programming can be difficult. Computers are stupid and little errors in your code can cause hours of headache (even if you've been doing this stuff for years!).
-
-Fortunately there are tons of online resources to help you with this.
-Two of the most important are [StackOverflow](https://stackoverflow.com/) (a Q&A site with hundreds of thousands of answers to all sorts of programming questions) and [RStudio Community](https://community.rstudio.com/) (a forum specifically designed for people using RStudio and the tidyverse (i.e.
-you)).
-
-Searching for help with R on Google can be tricky because the program is, um, a single letter.
-Try searching for "rstats" instead.
-If you use Twitter, post R-related questions and content with [#rstats](https://twitter.com/search?q=%23rstats).
-
-Additionally, we have a class chatroom at [Slack](https://hoodeconometrics.slack.com/) where anyone in the class can ask questions and anyone can answer.
-Ask questions about the readings or problem sets [in the class Slack workspace](https://hoodeconometrics.slack.com/).
-I will monitor Slack regularly, and you should also all do so as well.
-You'll likely have similar questions as your peers, and you'll likely be able to answer other peoples' questions too.
-
-# Course policies
-
-<span class="newthought">Be nice. Be honest. Don't cheat.</span>
-
-We will also follow the [full list of Marriott School and BYU classroom policies](https://marriottschool.byu.edu/policies/classroom/).
-
-## Counseling and Psychological Services (CAPS)
-
-Life at BYU can be complicated and challenging.
-You might feel overwhelmed, experience anxiety or depression, or struggle with relationships or family responsibilities.
-[Counseling and Psychological Services (CAPS)](https://caps.byu.edu/) provides free, *confidential* support for students who are struggling with mental health and emotional challenges.
-The CAPS office is staffed by professional psychologists who are attuned to the needs of all types of college and professional students.
-Please do not hesitate to contact CAPS for assistance—getting help is a smart and courageous thing to do.
 
@@
 
-```{r}
+```{r, echo=F}
 knitr::kable(assignments)
@@
 
-\
+All grades are based on the following traditional scale:
 
@@
 
-# Red pandas
+These grades are firm cutoffs, but I do of course round upwards ($\geq 0.5$) for final grades.
+A necessary reminder, as an academic, I am not in the business of *giving* out grades, I merely report the grade that you *earn*.
+I will not alter your grade unless you provide a reasonable argument that I am in error (which does happen from time to time).
 
-[^panda] <span class="newthought">Once you have read</span> this entire syllabus and [the assignments page](/assignments/), please [click here](mailto:andrew_heiss@byu.edu) and e-mail me a picture of a red panda.
-For real.
-Brownie points if it's animated.
+# Policies and Expectations
 
-[^panda]: {-}
-  ![Hungry red panda](https://media.giphy.com/media/r7Nbr97FsXxXG/giphy.gif)
+This syllabus is a contract between you, the student, and me, your instructor.
+It has been carefully and deliberately thought out^[A syllabus can and will be used as a legal document for disputes tried at a court of law.
+Ask me how I know.], and I will uphold my end of the agreement and expect you to uphold yours.
+
+In the language of game theory, this syllabus is my commitment device.
+I am a very understanding person, and I know that exceptions to rules often need to be made for students.
+However, to be *fair* to *all* students the syllabus artificially constrains my ability to make exceptions at a whim for anyone.
+This prevents clever students from exploiting my congenial personality at everyone else's expense.
+Please read and familiarize yourself with the course policies and expectations of you.
+Chances are, if you have a question, it is answered herein.
+
+### Attendance and Participation
+
+I expect you to attend class and to come having already done the reading assigned for that day.
+I will remind you in class and possibly through Blackboard or email which readings I want you to read for the next class.
+You are all adults and I will treat you as such.
+I do not take attendance, nor do I grade formally for participation but I strongly recommended you attend class and participate for your sake and the sake of your classmates.
+If you are too distracted or are not prepared to learn, I suggest you stay home, where you can check Facebook more efficiently.
+I reserve the right to boost the final grades of students that I believe have made consistent, quality contributions above and beyond their peers in class conversations by up to 2.5 points.
+
+### Devices
+
+You are allowed to have and use laptops and tablets in the classroom.
+I will not stop you, but I strongly discourage you from using these to take notes (see Tips for Success).
+As a courtesy to myself and to others, do not use your phone in class.
+I reserve the right to embarrass you in front of everyone if you do so.
+
+### Absences and Make-Ups
+
+You generally do *not* need to let me know if you are unable to make class, *unless* it is on the day of an exam.
+It will however, be your responsibility to acquire the notes from a classmate for any missed classes.
+If you are unable to attend an exam for a legitimate reason (e.g.
+sports/club events, traveling, illness, family issues), please notify me at least *one week* in advance, and we will schedule a make-up exam date.
+If you are ill or otherwise unable to attend on the day of the exam, contact me ASAP to make arrangements.
+Failure to do so, including desperate attempts to make arrangements only *after* the absence will result in a grade of 0 and little sympathy.
+I reserve the right to re-weight other assignments for students who I believe are legitimately unable to complete a particular assignment.
+
+### Late Assignments
+
+I will accept late assignments, but will subtract a specified amount of points as a penalty.
+See individual assignment descriptions for the amount of points taken off (as it varies by assignment).
+If an answer key is posted before you turn in your assignment, the maximum grade you can earn is an 80.
+Even if it is the last week of the semester, I encourage you to turn in late work: some points are better than no points!
+
+### Grading
+
+I will try my best to post grades on Blackboard's Grading Center and return graded assignments to you within about one week of you turning them in.
+There will be exceptions.
+Where applicable, I will post answer keys once I know most homeworks are turned in (see Late Assignments above for penalties).
+Blackboard's Grading Center is the place to look for your most up-to-date grades.
+You will also be given an Excel spreadsheet template where you can calculate your overall grade and forecast "what if" scenarios.
+
+### Email Accounts
+
+Students must regularly monitor their Hood email accounts to receive important college information, including messages related to this class.
+Email through the Blackboard system is my main method of communicating announcements and deadlines regarding your assignments.
+**Please do not reply to any automated Blackboard emails - I may not recieve it!**.
+My Hood email (safner@hood.edu) is the best means of contacting me.
+I will do my best to respond within 24 hours.
+If I do not reply within 48 hours, do not take it personally, and *feel free to send a follow up email* in the very likely event that I genuinely did not see your original message.
+
+### Office Hours
+
+I am generally in my office Monday-Thursday during "normal business hours." You are always welcome to walk-in and chat about class, college, careers, or anything at all.
+Please do try to use the official office hours stated at the head of the syllabus if possible.
+If you need to meet at a different time, I request that you send me an email or let me know after class so I know when to expect you.
+If you want to go over material from class, please have *specific* questions you want help with.
+I am not in the business of giving private lectures (particularly if you missed class without a valid excuse).
+
+Watch this excellent and accurate video explaining office hours:
+
+<iframe src="https://player.vimeo.com/video/270014784" width="640" height="360" frameborder="0" allow="autoplay; fullscreen" allowfullscreen></iframe>
+
+### Van Halen and M&Ms
+
+When you have completed reading the syllabus, email me a picture of the band Van Halen and a picture of a bowl of M&Ms.
+If you do this *before* the date of the first exam, you will get 3 bonus points on the exam.
+Yes, this is real.
+
+### Enrollment
+
+Students are responsible for verifying their enrollment in this class.
+The last day to add or drop this class with no penalty is **Tuesday, September 4**.
+Be aware of [important dates](https://www.hood.edu/offices-services/registrars-office/academic-calendar).
+
+### Honor Code
+
+Hood College has an Academic Honor Code which requires all members of this community to maintain the highest standards of academic honesty and integrity.
+Cheating, plagiarism, lying, and stealing are all prohibited.
+All violations of the Honor Code are taken seriously, will be reported to appropriate authority, and may result in severe penalties, including expulsion from the college.
+See [here](http://hood.smartcatalogiq.com/en/2016-2017/Catalog/The-Spirit-of-Hood/The-Academic-Honor-Code-and-Code-of-Conduct) for more detailed information.
+
+### Accessibility, Equity, and Accommodations
+
+College courses can, and should, be challenging and bring you out of your comfort zone in a safe and equitable environment.
+If, however, you feel at any point in the semester that certain assignments or aspects of the course will be disproportionately uncomfortable or burdensome for you due to any factor beyond your control, please come see me or email me.
+I am a very understanding person and am happy to work out a solution together.
+I reserve the right to modify and reweight assignments at my sole discretion for students that I belive would legitimately be at a disadvantage, through no fault of their own, to complete them as described.
+
+If you are unable to afford required textbooks or other resources for any reason, come see me and we can find a solution that works for you.
+
+This course is intended to be accessible for all students, including those with mental, physical, or cognitive disabilities, illness, injuries, impairments, or any other condition that tends to negatively affect one's equal access to education.
+If at any point in the term, you find yourself not able to fully access the space, content, and experience of this course, you are welcome to contact me to discuss your specific needs.
+I also encourage you to contact the [Office of Accessibility Services](https://www.hood.edu/academics/josephine-steiner-center-academic-achievement-retention/accessibility-services) (301-696-3421).
+If you have a diagnosis or history of accommodations in high school or previous postsecondary institutions, Accessibility Services can help you document your needs and create an accommodation plan.
+By making a plan through Accessibility Services, you can ensure appropriate accommodations without disclosing your condition or diagnosis to course instructors.
+
+# Schedule
+
+<span class="newthought">You can find a full schedule</span> with resources for each class meeting on the [schedule page](/schedule/).

### 2019-08-22 content/assignments.Rmd
Update assignments

---
title: "Assignments"
slug: assignments
---

# Problem Sets

There will be several problem sets (one at the end of each lesson). Problem sets will be a combination of math/statistical theory & application problems and problems that require use of R with real data. You may collaborate with other students to work on problem sets, but *each person must turn in an individual problem set*. Problem sets are due *one week* from the class period where we finish a lesson, and must *emailed* to me by the start of class (so please type or, if you must, hand write and scan them).

# Midterm

After we have finished linear regression, there will be a midterm constituting a combination of multiple choice, problem, and short answer questions. This will cover the content we discuss in class, my lectures, and the readings. The midterm provides feedback both to you and to me that ensures everyone is progressing on schedule and comprehending the material. This is *critical*, as the rest of the course will build off of this foundation.

# Final

On the college-determined date, we will have a comprehensive, closed-book, in-class final exam.

# Empirical Research Paper

You are required to write a short research paper any topic in political economy of your choosing, so long as it has an econometric component (at minimum, features a linear regression). I expect your topic and your findings to be very rudimentary (your results do not need to be significant). The main purpose of this paper is to walk you through the process of conducting research and using econometrics to gain insight on a problem. The most difficult part of this assignment is finding appropriate and sufficient data. I will discuss more about this paper at length in class, and briefly guide you through the craft of writing a good paper. I view this assignment as the primary demonstration of your mastery of this class, and writing is a marketable skill you will need for any gainful employment. I strongly recommend starting early and discussing your topics with me, but this is not required. On the final week of class, we will hold a workshop for each of you to give a short presentation of your paper, and for everyone to provide comments and constructive criticism to help you improve and complete your paper.

### 2019-08-27 content/class/01-class.Rmd
Update for first day!

---
title: "Class 1: Introduction to Econometrics"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

# Overview

An introduction to the course.

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/01-slides.html)

# Slack

- [Click here to join our class Slack channel](https://join.slack.com/t/hoodeconometrics/shared_invite/enQtNzQxMzI1MTU0ODM4LTVhOWI5Zjc5YzdiOGU5YzY0MmZiM2QxOGQwMWQwMjc4Y2FmYTk5Yjg5ZDZhZTY4ZDZiYWU2YzAzMTU1MDIwYTE)^[This link will expire in 7 days]

### 2019-08-27 content/reference.Rmd
Update for first day!

@@
 ```
+
+Look here over the course of the semester for resources, links, and tips on how to succeed in the course, how to write well, and other things of interest related to econometrics, data analysis, managing your worfklow, and using R.
 
@@
 These should not be used for your projects in most circumstances.]
-
-# Citations and bibliography
-
-[^bibtex] You can [download a BibTeX file](/bib/references.bib) of all the non-web-based readings in the course.
-
-[^bibtex]: {-} You can open the file in [BibDesk on macOS](http://bibdesk.sourceforge.net/), [JabRef on Windows](http://www.jabref.org/), or [Zotero](https://www.zotero.org/) or [Mendeley](https://www.mendeley.com/) online.

### 2019-08-27 static/slides/01-slides.Rmd
Add lecture 1 content

---
title: "1.1: Introduction to Econometrics"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: ocean #rainbow #monokai rainbow solarized-light #tomorrow-night -blue -eighties zenburn
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
               warning=F,
               message=F)
```

class: inverse, center, middle

# What is Econometrics?

---

# Why Everyone, Yes *Everyone*, Should Learn Statistics

.pull-left[
.center[
![:scale 75%](https://www.dropbox.com/s/xd47zns2riqtwzz/smbcmathteacher.png?raw=1)

[SMBC](https://www.smbc-comics.com/comic/why-i-couldn39t-be-a-math-teacher)
]
]

--

.pull-right[
.center[
![:scale 60%](https://www.dropbox.com/s/114jbjtdakritv0/smbc2080stats.png?raw=1)

[SMBC](https://www.smbc-comics.com/comic/2010-12-01)
]
]
---

# We're Not so Good at Statistics: Votes I

- Votes in the U.S. House of Representatives in favor of passing the Civil Rights Act of 1964:

.left-column[
| Democrat | Republican |
|----------|------------|
| 61%      | 80%        | 
]

--

.pull-right[
- Simple enough: "on average, Republicans tended to vote for passage more than Democrats"
]

---

# We're Not so Good at Statistics: Votes II

- Broken down further by Northern vs. Southern states:

.left-column[

|       |Democrat  | Republican |
|-------|----------|------------|
| **North** | **94%**      | 85%       | 
|           | (145/154)| (138/162) |
| **South** | **7%**       | 0%        |
|           | (7/94)   | (0/10)    |
| **Overall** | 61%    | **80%**       |
|             | (152/248) | (138/172) |
]

--
.right-column[
- Larger proportion of Democrats $(\frac{94}{248}$, 38%) than Republicans $(\frac{10}{172}$, 6%) were from South

- The 7% of southern Democrats voting *for* the Act dragged down the Democrats' *overall* percentage more than the 0% of southern Republicans
]
---

# We're Not So Good at Statistics: Kidney Stones I

.pull-left[
- Suppose you suffer from kidney stones, your doctor offers you .shout[treatment A] or .whisper[treatment B]

- In clinical trials, .shout[treatment A] was effective for a larger percentage of patients than .whisper[treatment B]

- .whisper[Treatment B] was effective for a higher percentage of patients with *large* stones and a higher percentage of patients with *small* stones

- Wait, what?
]

.pull-right[

.center[
![](https://www.dropbox.com/s/2sos7lpbwp2t6ry/pills2.jpg?raw=1)
]
]
---

# We're Not So Good at Statistics: Kidney Stones II

.pull-left[
From a real [medical study](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1339981/):

|       | Treatment A  | Treatment B |
|-------|----------|------------|
| **Small Stones** | **93%**      | 87%       | 
|                  | (81/87)| (234/270) |
| **Large Stones** | **73%**       | 69%        |
|                  | (192/263)   | (55/80)    |
| **Overall**      | 78%    | **83%**       |
|                  | (273/350) | (289/350) |
]

.footnote[

.source[C R Charig, D R Webb, S R Payne, and J E Wickham, 1986, "Comparison of treatment of renal calculi by open surgery, percutaneous nephrolithotomy, and extracorporeal shockwave lithotripsy," *Br Med J* (Clin Res Ed) 292(6524): 879–882. ]
]

--

.pull-right[
- The *sizes* of the two groups (i.e. who gets A vs B) are *very* different
- A .whisper[lurking variable] in the study is the severity of the case: doctors tended to give treatment B for less severe cases
]

---

# Simpson's Paradox

.pull-left[
.content-box-red[

.red[**Simpson's Paradox**:] The correlation between two variables can change (even reverse!) when additional variables are considered]

]
.pull-right[
.center[![](https://www.dropbox.com/s/0dxltnoenxmljlr/homerdoh.jpg?raw=1)]
]

---

# We're Not so Good at Statistics: Smoking I

.pull-left[
- 1964: U.S. Surgeon General issued a [report](http://profiles.nlm.nih.gov/ps/access/NNBBMQ.pdf) claiming that cigarette smoking causes lung cancer

- Evidence based primarily on *correlations* between cigarette smoking and lung cancer
]
.pull-right[

![](https://www.dropbox.com/s/fkb8z1sipjyknoc/smokingwarning.jpg?raw=1)
]

---

# We're Not so Good at Statistics: Smoking II

.pull-left[
- Tobacco companies attacked the report, naturally
]
.pull-right[

![](https://www.dropbox.com/s/7aqwchwq9vbttnu/bigtobaccotestify.jpg?raw=1)
]

---

# We're Not so Good at Statistics: Smoking III

.left-column[
.center[
![](https://www.dropbox.com/s/zh2vtayx27qtr9q/fishersmoke.png?raw=1)

Ronald A. Fisher

1890--1924
]
]
.right-column[

- But [so did R. A. Fisher](https://priceonomics.com/why-the-father-of-modern-statistics-didnt-believe/), the "father of modern statistics"
]

---

# We're Not so Good at Statistics: Smoking IV

.pull-left[
- There could be a confounding variable ("smoking gene") that causes *both* lung cancer *and* the urge to smoke

- Would imply: decision to smoke or not would have *no impact* on lung cancer!

- Correlation between smoking and cancer is spurious!

]

.pull-right[
.center[
```{r, echo=F, fig.retina=3}
library("ggdag")
dagify(Canc~Gene+Smok,
       Smok~Gene,
       coords=list(
         x=c(Smok=1,Canc=2, Gene=1.5),
         y=c(Smok=1,Canc=1, Gene=2)
         )) %>% 
  #tidy_dagitty() %>%
  ggdag_parents("Canc", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")

```
]
]

---

# Correlation Does Not Imply Causation I

- The goal of every intro statistics class ever

--

.center[
![](https://www.dropbox.com/s/gvd4sii1usda27k/xkcdcorrelation.png?raw=1)

[XKCD: Correlation](https://xkcd.com/552/)
]

---

# Correlation Does Not Imply Causation II

.center[
![](https://www.dropbox.com/s/x01lvngkw9owsli/spuriouscorr1.png?raw=1)

[Spurious Correlations](http://www.tylervigen.com/spurious-correlations)
]

---

# Correlation *Can* Imply Causation...

.pull-left[

- It's always good to be skeptical of causal claims

- But this is actually where .shout[econometrics] shines

]

.pull-right[

.center[
![](https://www.dropbox.com/s/c91a06o91rf3e5h/causation.jpg?raw=1)
]
]

---

# ...With the Right Tools

.pull-left[

- .shout[Econometrics] is the application of statistical tools to *quantify* economic relationships in the real world

- Uses real data to
    - test economic hypotheses
    - quantitatively estimate the magnitude of relationships between economic variables
    - forecast future events

]

.pull-right[

.center[
![](https://www.dropbox.com/s/l2s6c0s55lsws6y/statsgraphs.jpg?raw=1)
]
]

---

# Causal Inference I

.left-column[
.center[
![](https://www.dropbox.com/s/v5vwsadw5vs448t/causality.jpg?raw=1)
]
]

.right-column[

- What sets econometrics apart from mere statistics (or uses of statistics in other disciplines) is its role in .shout[causal inference]

- We can, with proper tools and interprations, make *quantitative* *causal* claims
    - about the effects of individual choices
    - about the effects of policy interventions
    - about the impact of political institutions
    - about economic history and economic development
    - etc...
]

---

# Causal Inference II

.left-column[
.center[
![](https://www.dropbox.com/s/v5vwsadw5vs448t/causality.jpg?raw=1)
]
]

.right-column[

> A 50% increase in police presence in a metropolitan area lowers crime rates by 15%, on average<sup>1</sup>

> Being an incumbent in office raises the probability of re-election by 40-45 percentage points<sup>2</sup>

> European cities with at least one printing press in 1500 were at least 29% more likely to become Protestant by 1600<sup>3</sup>

]

.footnote[
.source[<sup>1</sup> Klick, Jonathan and Alexander Tabarrok, 2005, "Using Terror Alert Levels to Estimate the Effect of Police on Crime," *Journal of Law and Economics* 48(1): 267-279

<sup>2</sup> Lee, David S, 2001, "The Electoral Advantage to Incumbency and Voters' Valuation of Politicians' Experience: A Regression Discontinuity Analysis of Elections to the U.S," *NBER Working Paper* 8441

<sup>3</sup> Rubin, Jared, 2014, "Printing and Protestants: An Empirical Test of the Role of Printing in the Reformation," *Review of Economics and Statistics* 96(2): 270-286
]
]
---

# Example 1: Education

.pull-left[
.content-box-green[
.green[Does reducing class sizes improve student performance?]
]
]

.pull-right[
![](https://www.dropbox.com/s/80sb1duv6e2tazd/smallclass.jpg?raw=1)
]

---

# Example 1: Education

.pull-left[
.content-box-green[
.green[Does reducing class sizes improve student performance?]

- A policy-relevant tradeoff with a budget constraint
- What is the *precise* effect of class size on performance?
- Is it worth hiring new teachers and building more schools over?
]
]

.pull-right[
![](https://www.dropbox.com/s/80sb1duv6e2tazd/smallclass.jpg?raw=1)
]

---

# Example 2: Discrimination in Lending

.pull-left[
.content-box-green[
.green[Is there racial discrimination in home mortgage lending?]
]
]

.pull-right[
![](https://www.dropbox.com/s/pdwqjp7li11m77o/mortgageapp.jpg?raw=1)
]

---

# Example 2: Discrimination in Lending

.pull-left[
.content-box-green[
.green[Is there racial discrimination in home mortgage lending?]

- Boston Fed: 28% of African-Americans are denied mortgages compared to only 9% of White Americans
- Is this due to factors such as credit history, income, or discrimination *purely* because of race? 
]
]

.pull-right[
![](https://www.dropbox.com/s/pdwqjp7li11m77o/mortgageapp.jpg?raw=1)
]

---

# Example 3: Public Health and Public Finance

.pull-left[
.content-box-green[
.green[How much do state cigarette taxes reduce smoking rates?]

]
]

.pull-right[
![](https://www.dropbox.com/s/fkb8z1sipjyknoc/smokingwarning.jpg?raw=1)
]

---

# Example 3: Public Health and Public Finance

.pull-left[
.content-box-green[
.green[How much do state cigarette taxes reduce smoking rates?]

- Econ 101: raise price $\implies$ lower quantity consumed

- What is the *price elasticity of demand* for smoking?

- How much tax revenue will this generate?

- Probably:

$$Taxes \rightarrow Smokers$$

- Maybe?:

$$Taxes \leftarrow Smokers$$
]
]

.pull-right[
![](https://www.dropbox.com/s/fkb8z1sipjyknoc/smokingwarning.jpg?raw=1)
]

---

class: inverse, center, middle

# About this Class

---

# Real Talk I

.center[
![:scale 60%](https://www.dropbox.com/s/k3hhulc5tjy6eis/opengatemath1.jpg?raw=1)
]

---

# Real Talk I

.center[
![:scale 60%](https://www.dropbox.com/s/60kmhih3bpe8l1p/opengatemath2.jpg?raw=1)
]

---

# Real Talk I

.center[
![:scale 60%](https://www.dropbox.com/s/8apd0v6rf5rt165/opengatemath3.jpg?raw=1)
]

---

# Real Talk II

.pull-left[
- This will be one of the hardest courses you take at Hood 
- There will be moments where you have no idea WTF is going on (*this is normal*)
- Yes, you can still get an **A**
]

.pull-right[
.center[
![](https://www.dropbox.com/s/da7m0qx9c2s325g/difficultclass.png?raw=1)
]
]

---

# This Class Is

.left-code[

- **Economics**: take your *preexisting* intuition and models for causal inference
- **Statistics**: add regression and statistical inference
- **Computer Programming**: using `R` and `R Studio` for analyzing and presenting data
]

.right-plot[
.center[
```{r, echo=F}
library("ggplot2")
class<-data.frame(Percentage=c(50,40,10),
                  Subject=c("Economics","Statistics","Computer Programming"))

#ggplot(class, aes(x="", y=Percentage, fill=Subject))+
#  geom_col()+theme_void()+scale_fill_viridis_d()

library(treemapify)
ggplot(class, aes(area=Percentage, fill=Subject))+
  geom_treemap()+
  guides(fill=F)+
  geom_treemap_text(aes(label = Subject),
                  colour = "white", place = "center", grow = TRUE)+
  #scale_fill_brewer(palette="viridis")+
  scale_fill_viridis_d()
```
]
]

---

# This Class Is

.pull-left[

### Old School Statistics Courses

- $\bar{x} = \frac{1}{n} \displaystyle\sum^n_{i=1} x_i$

- $\sigma_x = \displaystyle \sqrt{\frac{1}{n} \sum^n_{i=1} (x_i-\bar{x})^2}$

- $r_{xy}= \displaystyle \frac{\displaystyle\sum^n_{i=1}(x_i-\bar{x})(y_i-\bar{y})}{\sqrt{\displaystyle\sum^n_{i=1}(x_i-\bar{x})^2\sum^n_{i=1}(y_i-\bar{y})^2}}$

- Use pre-cleaned "toy" data, if any
]

--

.pull-right[

### Hip New Data Science Courses

- `mean(x)`

- `sd(x)`

- `cor(x, y)`

- Manipulate raw data from scratch (real life)
]

---

# Prerequisites

- **Courses**: 
  - ECON 205
  - ECON 206
  - ECON 305 or ECON 306
  - MATH 112 or ECMG 212

--

- **Math Skills**:
  - Basic algebra
  - Probability-ish
  - Statistics-ish

--

- **Computer Science Skills**:
  - None

---

# What You'll Get Out of This Class

.pull-left[

By the end of this semester, you will:

1. understand how to evaluate statistical and empirical claims;
2. use the fundamental models of causal inference and research design;
3. gather, analyze, and communicate with real data in R.
]

.pull-right[
.center[
![:scale 80%](https://www.dropbox.com/s/k1xtcebidkonen9/r.png?raw=1)
]
]

---

# This Class Opens Doors

.center[
![](https://www.dropbox.com/s/v33ar033qvc2c0s/metricshelpful.png?raw=1)
]

---

# This Class Gives You a Hybrid of Skills

.pull-left[

- **"Data Science"**: ???
- **Causal Inference**: economists' comparative advantage!

]

.pull-right[
.center[
```{r, echo=F}
library("ggplot2")
class2<-data.frame(Percentage=c(50,50),
                  Subject=c("Data Science","Causal Inference"))

#ggplot(class, aes(x="", y=Percentage, fill=Subject))+
#  geom_col()+theme_void()+scale_fill_viridis_d()

library(treemapify)
ggplot(class2, aes(area=Percentage, fill=Subject))+
  geom_treemap()+
  guides(fill=F)+
  geom_treemap_text(aes(label = Subject),
                  colour = "white", place = "center", grow = TRUE)+
  #scale_fill_brewer(palette="viridis")+
  scale_fill_viridis_d()
```
]
]
---

# Data Science I

.pull-left[
.center[
<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Data Scientist (n.): Person who is better at statistics than any software engineer and better at software engineering than any statistician.</p>&mdash; Josh Wills (@josh_wills) <a href="https://twitter.com/josh_wills/status/198093512149958656?ref_src=twsrc%5Etfw">May 3, 2012</a></blockquote>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
]
]
.pull-right[

.center[
![](https://www.dropbox.com/s/mnjz53pg56gz2s3/datascience1.png?raw=1)
]
]

---

# Data Science II

.pull-left[
.center[
[Harvard Business Review](https://hbr.org/2012/10/data-scientist-the-sexiest-job-of-the-21st-century)

![:scale 95%](https://www.dropbox.com/s/d1z8hw9nhq474ix/datascientist.png?raw=1)

]
]

.pull-right[

.center[
[LinkedIn 2018 Emerging Jobs Report](https://economicgraph.linkedin.com/research/linkedin-2018-emerging-jobs-report)

![:scale 95%](https://www.dropbox.com/s/o2chnosebxiuxac/datasciencelinkedin.png?raw=1)

]
]

---

# R Skills are In Demand

.pull-left[
```{r, echo=F, warning = F, message = F, fig.retina = 3}
library(tidyverse)

kaggle <- read_csv("~/Downloads/kaggle-survey-2018/multipleChoiceResponses.csv")

clean <- kaggle %>%
  select(Q1, Q2, Q3, Q4, Q5, Q6, Q7, Q8, Q9, Q17, Q18) %>%
  rename(gender = Q1,
         age = Q2,
         country = Q3,
         education = Q4,
         major = Q5,
         job_title = Q6,
         industry = Q7,
         experience = Q8,
         salary = Q9,
         top_lang = Q17,
         rec_lang = Q18) %>%
  slice(-1)

clean %>%
  filter(!is.na(top_lang)) %>%
  group_by(top_lang) %>% 
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  mutate(perc = round((n/sum(n) * 100),2)) %>%
  slice(1:10) %>%
  ggplot(., aes(x = reorder(top_lang, n), 
                y = perc,
                fill = top_lang))+
  geom_col()+
  scale_y_continuous(breaks=seq(0,60,10))+
  theme_minimal()+
  labs(x = "Programming Language",
       y = "Percent",
       title = "Programming Languages Used Most by Data Scientists",
       caption = "Source: 2018 Kaggle Data Science Survey")+
  coord_flip()+
  guides(fill = F)+
  scale_fill_viridis_d()
```
]

.pull-right[
.source[[Kaggle Data Science Survey 2018](https://www.kaggle.com/kaggle/kaggle-survey-2018)]
]

---

# Yada Yada Machine Learning

.pull-left[

.center[
![:scale 80%](https://www.dropbox.com/s/z5u9qmm0mskwl2p/machinelearning.jpg?raw=1)
]
]

.pull-right[

> "When you’re fundraising, it’s AI. When you’re hiring, it’s ML. When you’re implementing, it’s logistic regression."

> - everyone on Twitter ever ([Source](https://towardsdatascience.com/no-machine-learning-is-not-just-glorified-statistics-26d3952234e3))
]

---

# Causal Inference I

.pull-left[
- Machine learning and artificial intelligence are "dumb"<sup>1</sup>
- With the right models and research designs, we *can* say "X causes Y" and quantify it!
- Economists are in a unique position to make *causal* claims that mere statistics cannot
]

.pull-right[
![](https://www.dropbox.com/s/yw8t5xsa8dgei71/coding.jpeg?raw=1)
]

.footnote[<sup>1</sup> For more, see [my blog post](https://ryansafner.com/post/econometrics-data-science-and-causal-inference/), and Pearl & MacKenzie (2018), *The Book of Why*]
---

# Causal Inference II

.pull-left[
.center[
![](https://www.dropbox.com/s/jrp67z1umll1urp/econtechcompanies.png?raw=1)
[Harvard Business Review](https://hbr.org/2019/02/why-tech-companies-hire-so-many-economists)
]
]
.pull-right[

.font80[
> "First, the field of economics has spent decades developing a toolkit aimed at investigating empirical relationships, focusing on techniques to help understand which correlations speak to a causal relationship and which do not. This comes up all the time — does Uber Express Pool grow the full Uber user base, or simply draw in users from other Uber products? Should eBay advertise on Google, or does this simply syphon off people who would have come through organic search anyway? Are African-American Airbnb users rejected on the basis of their race? These are just a few of the countless questions that tech companies are grappling with, investing heavily in understanding the extent of a causal relationship."
]
]

---

# Building Good Workflow Habits

.pull-left[

- I will show you the tools to make your workflow:
    - Reproducible
    - Computer- and Human-Readable (!)
    - Automated
    - All in one program
]

.pull-right[

![](https://www.dropbox.com/s/546o01icserlone/rmdout.png?raw=1)
]

---

# A Quick Example

.left-code[

```{r gapplot, eval=F, echo=T}
library("gapminder")

ggplot(data = gapminder, 
       aes(x = gdpPercap,
           y = lifeExp,
           color = continent))+
  geom_point(alpha=0.3)+
  geom_smooth(method = "lm")+
	scale_x_log10(breaks=c(1000,10000, 100000),
	              label=scales::dollar)+
	labs(x = "GDP/Capita",
	     y = "Life Expectancy (Years)")+
  facet_wrap(~continent)+
  guides(color = F)+
  theme_light()
```
]

.right-plot[
.center[
```{r, ref.label="gapplot", echo=F, fig.retina=3, out.width="72.5%"}
```
]
]
---

# Assignments

.pull-left[

- Research project:
    - Come up with a testable research question
    - Find data
    - Analyze data
    - Present your results (in writing and verbally)
- HWs
- Midterm, Final exam (in-class, closed notes)
]

.pull-right[
.center[
```{r, echo = F}
library("tidyverse")
assignments <- tribble(
  ~"",  ~Assignment, ~Percent,
  1, "Research Project", "30%",
  "n"     , "Homeworks (Average)", "25%",
  1, "Midterm", "20%",
  1     , "Final", "25%"
)
knitr::kable(assignments, "html")
```
]
]

---

# Your "Textbooks"

.pull-left[
.center[
![:scale 60%](https://www.dropbox.com/s/me6vp7e536aj75p/baileyeconometrics.jpg?raw=1)
]
]
.pull-right[
.center[
![:scale 60%](https://www.dropbox.com/s/h62ffjo3kswu70m/r4ds.png?raw=1)
]
]

---

# Tips for Success In This Course

.pull-left[
- *Take notes. On paper. Really.*

- **Work together** on assignments and study together.

- Ask questions, come to office hours. Don't struggle in silence, you are not alone!

- .shout[You are learning how to learn]<sup>1</sup>

- See the [reference page](http://metricsf19.classes.ryansafner.com/reference) for more
]

.pull-right[
.center[
![](https://www.dropbox.com/s/rfdf5g27mdtq5wx/education.png?raw=1)
]
]

.footnote[<sup>1</sup> A properly worded Google search will become your secret weapon. Believe me. It's still mine.]

---

# Course Website

.center[
![:scale 30%](https://www.dropbox.com/s/um51cx2bgojrdym/metrics_hex.png?raw=1)

[metricsf19.classes.ryansafner.com](http://metricsf19.classes.ryansafner.com)
]

---

# Let's Try This Out

.center[
![](https://www.dropbox.com/s/r8yzey55g17hi9n/slack.jpg?raw=1)

[hoodeconometrics.slack.com](https://hoodeconometrics.slack.com/)
]

---

# Roadmap for the Semester

.center[
![:scale 75%](https://www.dropbox.com/s/zq71sobk1m6asx0/metrics_flowchart_blankbg.png?raw=1)
]

### 2019-08-27 content/assignment/00-preliminary-survey.Rmd
Fix github link for updates, assignments shortcode to remove due dates; add preliminary survey PDF

---
title: "Preliminary Survey on Statistics and Software"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

*Due by Tuesday, September 3, 2019*

- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/00-preliminary-survey.pdf)

This is an *ungraded* and *anonymous* survey for me to evaluate the distribution of your math and statistics backgrounds. Please write your answers the the following questions on a piece of paper, or download and print a PDF copy (link above) to write on. You do not need to write your name on it. Please complete all problems to the best of your ability. Your responses will help me craft the course to see which material we need to focus on at greater length, especially review material.

# 1

Using the following sample data:

$$8, 12, 9, 10, 11, 5, 15$$

a. Find the median. 
b. Calculate the sample mean, $\bar{x}$
c. Calculate the sample standard deviation, $s$

# 2

For a fair, six-sided die:

a. What is the probability of rolling a 5?
b. What is the probability of rolling an even number?
c. What is the probability of rolling an even number or a 3?
d. If you have two fair dice, what is the probability of rolling a 6 on both dice?

# 3

Hedge fund A earns an average rate of return of 2.5% per year with a standard deviation of 0.5%, while hedge fund B earns an average rate of return of 3.0% per year with a standard deviation of 2.0\%. Which is more unusual, Hedge fund A earning a 4.0% return or hedge fund B earning a return -1.0% return? Why?^[Hint: Standardize the two hedge funds.]

# 4

A discrete random variable $X$ has the following pdf:

```{r, echo=FALSE, results="asis"}
tribble(
  ~x, ~`p(x)`,
  10, 0.10,
  20, 0.20,
  30, 0.30,
  40, 0.40
  ) %>%
  pandoc.table(justify = "ll")
```

Calculate the sample standard deviation, $s$ of $X$.

# 5

The random variable $Y$ is normally distributed with a mean of 50 and standard deviation of 12

$$Y \sim N (50,12)$$

a. What is the $Z$-score for $Y=74$?
b. In your own words, what does this $Z$-score mean?
c. What is the probability that $Y$ takes on a value *greater* than 74?

# 6

On a scale of 1 (least) to 10 (most), how anxious are you about this class? Feel free to share any specific anxieties (they have a better chance to be specifically addressed if you do!).

# 7

On a scale of 1 (least) to 10 (most), how familiar would you say you are with computer programming and/or statistical software?

# 8

List any statistical software packages (e.g. R, Microsoft Excel, Stata, SAS, SPSS, Minitab, etc.) and any programming languages (e.g. html, php, C/++, Python, LaTeX, etc.) you have had any experience with, and rate your proficiency between 1 (least) and 5 (most), if applicable.

### 2019-08-27 static/homeworks/00-preliminary-survey.Rmd
Fix github link for updates, assignments shortcode to remove due dates; add preliminary survey PDF

--- 2019-08-27 content/assignment/00-preliminary-survey.Rmd
@@
 title: "Preliminary Survey on Statistics and Software"
-editor_options: 
-  chunk_output_type: console
+author: "Ryan Safner"
+date: "ECON 480 Fall 2019"
+output: pdf_document
 ---
 
-```{r message=FALSE, warning=FALSE, include=FALSE}
-library(tidyverse)
-library(pander)
-source(file.path(here::here(), "lib", "graphics.R"))
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE,
+                      message = FALSE,
+                      warning = FALSE)
+library("tidyverse")
 ```
@@
 *Due by Tuesday, September 3, 2019*
-
-- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/00-preliminary-survey.pdf)
 
@@
 
-# 1
-
-Using the following sample data:
+### 1. Using the following sample data:
 
@@
 
-a. Find the median. 
-b. Calculate the sample mean, $\bar{x}$
-c. Calculate the sample standard deviation, $s$
+### a. Find the median. 
+### b. Calculate the sample mean, $\bar{x}$
+### c. Calculate the sample standard deviation, $s$
 
-# 2
+\clearpage
 
-For a fair, six-sided die:
+### 2. For a fair, six-sided die:
 
-a. What is the probability of rolling a 5?
-b. What is the probability of rolling an even number?
-c. What is the probability of rolling an even number or a 3?
-d. If you have two fair dice, what is the probability of rolling a 6 on both dice?
+### a. What is the probability of rolling a 5?
+### b. What is the probability of rolling an even number?
+### c. What is the probability of rolling an even number or a 3?
+### d. If you have two fair dice, what is the probability of rolling a 6 on both dice?
 
-# 3
+\clearpage
 
+### 3.
 Hedge fund A earns an average rate of return of 2.5% per year with a standard deviation of 0.5%, while hedge fund B earns an average rate of return of 3.0% per year with a standard deviation of 2.0\%.
@@
 
-# 4
+\clearpage
 
-A discrete random variable $X$ has the following pdf:
+### 4. A discrete random variable $X$ has the following pdf:
 
@@
   ) %>%
-  pandoc.table(justify = "ll")
+  knitr::kable(., "latex")
 ```
@@
 
-# 5
+\clearpage
 
-The random variable $Y$ is normally distributed with a mean of 50 and standard deviation of 12
+### 5. The random variable $Y$ is normally distributed with a mean of 50 and standard deviation of 12
 
@@
 
-a. What is the $Z$-score for $Y=74$?
-b. In your own words, what does this $Z$-score mean?
-c. What is the probability that $Y$ takes on a value *greater* than 74?
+### a. What is the $Z$-score for $Y=74$?
+### b. In your own words, what does this $Z$-score mean?
+### c. What is the probability that $Y$ takes on a value *greater* than 74?
 
-# 6
+\clearpage
 
-On a scale of 1 (least) to 10 (most), how anxious are you about this class? Feel free to share any specific anxieties (they have a better chance to be specifically addressed if you do!).
+### 6 On a scale of 1 (least) to 10 (most), how anxious are you about this class? Feel free to share any specific anxieties (they have a better chance to be specifically addressed if you do!).
 
-# 7
+\vspace{2in}
 
-On a scale of 1 (least) to 10 (most), how familiar would you say you are with computer programming and/or statistical software?
+### 7. On a scale of 1 (least) to 10 (most), how familiar would you say you are with computer programming and/or statistical software?
 
-# 8
+\vspace{2in}
 
+### 8.
 List any statistical software packages (e.g.

### 2019-08-27 static/slides/01-slides.Rmd
Fix treatments in slides

@@
 
-- In clinical trials, .shout[treatment A] was effective for a larger percentage of patients than .whisper[treatment B]
+- In clinical trials, .shout[treatment B] was effective for a larger percentage of patients than .whisper[treatment A]
 
-- .whisper[Treatment B] was effective for a higher percentage of patients with *large* stones and a higher percentage of patients with *small* stones
+- .whisper[Treatment A] was effective for a higher percentage of patients with *large* stones and a higher percentage of patients with *small* stones
 

### 2019-08-27 static/slides/01-slides.Rmd
Fix treatment colors in slides

@@
 
-- In clinical trials, .shout[treatment B] was effective for a larger percentage of patients than .whisper[treatment A]
+- In clinical trials, .whisper[treatment B] was effective for a larger percentage of patients than .shout[treatment A]
 
-- .whisper[Treatment A] was effective for a higher percentage of patients with *large* stones and a higher percentage of patients with *small* stones
+- .shout[Treatment A] was effective for a higher percentage of patients with *large* stones and a higher percentage of patients with *small* stones
 

### 2019-08-29 content/class/02-class.Rmd
Add class 2 content

---
title: "Class 1: Introduction to Econometrics"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

# Overview

Today we will cover the fundamental problem of causal inference. This will help provide us a framework to understand causality and the challenges we will have to grapple with all semester. There is nothing too technical here, and we will not return to this for several weeks (focusing first on building your `R` skills), but these are the most important concepts we will return to countless times.

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/02-slides.html)

# Slack

- [Don't forget to join our class Slack channel](https://join.slack.com/t/hoodeconometrics/shared_invite/enQtNzQxMzI1MTU0ODM4LTVhOWI5Zjc5YzdiOGU5YzY0MmZiM2QxOGQwMWQwMjc4Y2FmYTk5Yjg5ZDZhZTY4ZDZiYWU2YzAzMTU1MDIwYTE)^[This link will expire in 5 days]

Please use your hood email address and real name. Starting next week it may prove helpful you start finding yourself asking a lot of R questions -- and then later, a lot of theory and statistics questions.

### 2019-08-29 static/slides/02-slides.Rmd
Add class 2 content

---
title: "1.2: The Quest for Causality"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: ocean #rainbow #monokai rainbow solarized-light #tomorrow-night -blue -eighties zenburn
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
               warning=F,
               message=F)
set.seed(256)
```

class: inverse, center, middle

# The 2 Big Problems with Data

---

# Two Big Problems with Data

.pull-left[

- We want to use econometrics to **identify** causal relationships and make **inferences** about them

1. Problem for **identification**: .shout[endogeneity]

2. Problem for **inference**: .shout[randomness]

]

.pull-right[
.center[
![:scale 90%](https://www.dropbox.com/s/phhet82tcnuegxp/randomimage.jpg?raw=1)]
]

---

# Identification Problem: Endogeneity I

.pull-left[

- An independent variable $(X)$ is .shout[exogenous] if its variation is *unrelated* to other factors that affect the dependent variable $(Y)$

- An independent variable $(X)$ is .shout[endogenous] if its variation is *related* to other factors that affect the dependent variable $(Y)$
]

.pull-right[

.center[
![](https://www.dropbox.com/s/v5vwsadw5vs448t/causality.jpg?raw=1)
]
]

---

# Identification Problem: Endogeneity II

.pull-left[

- An independent variable $(X)$ is .shout[exogenous] if its variation is *unrelated* to other factors that affect the dependent variable $(Y)$

]

.pull-right[

```{r, fig.retina=3, outwidth="90%", fig.align="center"}
library(ggdag)
#theme_set(theme_dag_blank())
library(gridExtra)

h1 <- dagify(Y ~ X,
             coords=list(
               x=c(X=1,Y=2),
               y=c(X=1,Y=1)
               )) %>% ggdag_parents("Y", stylized = FALSE) + 
  labs(title = "X causes Y") + theme_dag_blank()+theme(legend.position = "none")

h2 <- dagify(Y ~ X+Z,
             coords=list(
               x=c(Z=1,X=1,Y=2),
               y=c(Z=2,X=1,Y=1.5)
             )) %>% ggdag_parents("Y", stylized = FALSE) +
  labs(title = "X and Z (independently) cause Y") + theme_void()+theme(legend.position = "none")

grid.arrange(h1, h2, ncol=1)

```

]

---

# Identification Problem: Endogeneity III

.pull-left[

- An independent variable $(X)$ is .shout[endogenous] if its variation is *related* to other factors that affect the dependent variable $(Y)$

]

.pull-right[

```{r, fig.retina=3, outwidth="90%", fig.align="center"}
h3 <- dagify(Y ~ X+Z, X ~Z,
             coords=list(
               x=c(X=1, Z=1.5, Y=2),
               y=c(X=1, Z=2, Y=1)
             )) %>% ggdag_parents("Y", stylized = FALSE) +
  labs(title = "Z causes X and Y") + theme_void()+theme(legend.position = "none")

h4 <- dagify(Y ~ Z, Z ~ X,
             coords=list(
               x=c(X=1, Z=1.5, Y=2),
               y=c(X=1, Z=1, Y=1)
               )) %>% ggdag_parents("Y", stylized = FALSE) + 
  labs(title = "X Causes Y Indirectly Through Z") + theme_void()+theme(legend.position = "none")

h5 <- dagify(Y ~ X+Z, Z ~X,
             coords=list(
               x=c(X=1, Z=1.5, Y=2),
               y=c(X=1, Z=0.5, Y=1)
             )) %>% ggdag_parents("Y", stylized = FALSE) +
  labs(title = "X Causes Y (Directly and Via Z)") + theme_void()+theme(legend.position = "none")

grid.arrange(h3, h4, h5, nrow=3)

```

]

---

# Identification Problem: Endogeneity IV

.pull-left[

- Confusingly, these terms mean something different in econometrics than in the theoretical economics models where you have heard them before

- In *Theoretical* models:
    - "Exogenous": a parameter determined outside of the model and taken as given
    - "Endogenous": a variable whose value is determined by the model

]

.pull-right[

.center[
![:scale 90%](https://www.dropbox.com/s/qvr240t5j6t3arm/optimize.jpeg?raw=1)
]
]

---

# Identification Problem: Endogeneity V

.pull-left[

.content-box-green[
.green[**Example 1**:] In a classic supply and demand model:

- Exogenous parameters: income, prices of other goods, cost, technology
- Endogenous variables: equilibrium price, equilibrium quantity
]
]

.pull-right[

.center[
![:scale 90%](https://www.dropbox.com/s/qvr240t5j6t3arm/optimize.jpeg?raw=1)
]
]

---

# Identification Problem: Endogeneity VI

.pull-left[

.content-box-green[

.green[**Example 2**:] In a consumer optimization model:

- Exogenous parameters: market prices, income, utility function 
- Endogenous variables: utility-maximizing bundle of goods
]
]

.pull-right[

.center[
![:scale 90%](https://www.dropbox.com/s/qvr240t5j6t3arm/optimize.jpeg?raw=1)
]
]

---

# Inference Problem: Randomness

.pull-left[
- Data is random due to **natural sampling variation**

    - Taking one sample of a population will yield slightly different information than another sample of the same population

- Common in statistics, *easy to fix*

- **Inferential Statistics**: making claims about a wider population using sample data

    - We use common tools and techniques to deal with randomness
]

.pull-right[
.center[
![:scale 90%](https://www.dropbox.com/s/bsdtuddzjouwzr1/sampling.jpg?raw=1)
]
]

---

class: inverse, center, middle

# *Identifying* Causal Effects: Random Controlled Trials

---

# Random Controlled Trials (RCTs) I

.pull-left[

- The *ideal* way to demonstrate causation is through a .hl[randomized controlled trial (RCT)] or "random experiment"
    - *Randomly* assign experimental units (e.g. people, firms, etc.) into groups
    - **Treatment group(s)** get a (kind of) treatment
    - **Control group** gets no treatment 
    - Compare results of treatment and control groups to observe the **average treatment effect (ATE)**
- .hl[We will understand "causality" to mean the ATE from an ideal RCT]
]

.pull-right[

.center[
![:scale 90%](https://www.dropbox.com/s/ninee93he1zgcon/groupsplit.jpeg?raw=1)
]
]

---

# Random Controlled Trials (RCTs) II

.center[
![:scale 75%](https://www.dropbox.com/s/wqjisnpxaoaa2yy/rct.png?raw=1)

Classic (simplified) procedure of a randomized control trial (RCT) from medicine
]

---

# Random Controlled Trials (RCTs) III

.pull-left[

- .hl[Random assignment] to groups ensures that the *only* differences between members of the treatment(s) and control groups is *receiving treatment or not*

]

.pull-right[

.pull-left[
.center[
![:scale 80%](https://www.dropbox.com/s/vcqojr6k058e8hl/3apples.jpg?raw=1)
Treatment Group
]
]

.pull-right[
.center[
![:scale 100%](https://www.dropbox.com/s/xvjphzlwechgip6/3oranges.jpg?raw=1)
Control Group
]
]
]

---

# Random Controlled Trials (RCTs) IV

.pull-left[

- .hl[Random assignment] to groups ensures that the *only* differences between members of the treatment(s) and control groups is *receiving treatment or not*

- .hl[Selection bias:] (pre-existing) differences between members of treatment and control groups *other* than treatment, that affect the outcome

]

.pull-right[

.pull-left[
.center[
![:scale 80%](https://www.dropbox.com/s/vcqojr6k058e8hl/3apples.jpg?raw=1)
Treatment Group
]
]

.pull-right[
.center[
![:scale 100%](https://www.dropbox.com/s/xvjphzlwechgip6/3oranges.jpg?raw=1)
Control Group

(Selection Bias)
]
]
]

---

class: inverse, center, middle

# Some Theory By Example

---

background-image: url(https://www.dropbox.com/s/zsx3pa4m51p82dj/twopaths.jpg?raw=1)
background-size: cover

---

# Example: The Effect of Having Health Insurance I

.pull-left[

.content-box-green[
.green[**Example:**] What is the effect of having health insurance on health outcomes?

- National Health Interview Survey (NHIS) asks "Would you say your health in general is excellent, very good, good, fair, or poor?"

- **Outcome variable $(Y)$**: Index of health (1-poor to 5-excellent) in a sample of married NHIS respondents in 2009 who may or may not have health insurance

- **Treatment $(X)$**: Having health insurance  (vs. not)

]
]

.pull-right[

![](https://www.dropbox.com/s/h9c24dls3cisip1/healthcare.jpg?raw=1)
]

---

# Example: The Effect of Having Health Insurance II

.center[
![:scale 42%](https://www.dropbox.com/s/5lpltvdfim6elr6/econometricshealthnocaption.png?raw=1)]

.source[Angrist, Joshua & Jorn-Steffen Pischke, 2015, *Mostly Harmless Econometrics*]
---

# Example: The Effect of Having Health Insurance III

- $Y$: outcome variable (health index score, 1-5)

- $Y_i$: health score of an individual $i$

- Individual $i$ has a choice, leading to one of two outcomes:

    - $Y^0_i$: individual $i$ has *not* purchased health insurance

    - $Y^1_i$: individual $i$ has purchased health insurance

- $Y^1_i-Y^0_i$: causal effect for individual $i$ of purchasing health insurance

---

# Example: A Hypothetical Comparison

.left-column[

| John      | Maria     |
|-----------|-----------|
| ![](https://www.dropbox.com/s/5vh181yeg0p91t7/coldman2.jpg?raw=1) | ![](https://www.dropbox.com/s/7qa0gtuy29us4p2/femalejogger2.jpg?raw=1) |
| $Y_J^0=3$ | $Y_M^0=5$ | 
| $Y_J^1=4$ | $Y_M^1=5$ |
]

--

.right-column[
- John will choose to buy health insurance

- Maria will choose to not buy health insurance
]

---

# Example: A Hypothetical Comparison

.left-column[

| John      | Maria     |
|-----------|-----------|
| ![](https://www.dropbox.com/s/5vh181yeg0p91t7/coldman2.jpg?raw=1) | ![](https://www.dropbox.com/s/7qa0gtuy29us4p2/femalejogger2.jpg?raw=1) |
| $Y_J^0=3$ | $Y_M^0=5$ | 
| $Y_J^1=4$ | $Y_M^1=5$ |
| $Y_J^1-Y_J^0=1$ | $Y_M^1-Y_M^0=0$ |
]

.right-column[
- John will choose to buy health insurance

- Maria will choose to not buy health insurance

- Health insurance improves John's score by 1, has no effect on Maria's score
]

---

# Example: A Hypothetical Comparison

.left-column[

| John      | Maria     |
|-----------|-----------|
| ![](https://www.dropbox.com/s/5vh181yeg0p91t7/coldman2.jpg?raw=1) | ![](https://www.dropbox.com/s/7qa0gtuy29us4p2/femalejogger2.jpg?raw=1) |
| $Y_J^0=3$ | $Y_M^0=5$ | 
| $Y_J^1=4$ | $Y_M^1=5$ |
| $Y_J^1-Y_J^0=1$ | $Y_M^1-Y_M^0=0$ |
| $Y_J=(Y_J^1)=4$ | $Y_M=(Y_M^0)=5$ |

]

.right-column[
- John will choose to buy health insurance

- Maria will choose to not buy health insurance

- Health insurance improves John's score by 1, has no effect on Maria's score

- Note, all we can observe in the data are their health outcomes *after* they have chosen (not) to buy health insurance: $Y_J=4$ and $Y_M=5$
]

---

# Example: A Hypothetical Comparison

.left-column[

| John      | Maria     |
|-----------|-----------|
| ![](https://www.dropbox.com/s/5vh181yeg0p91t7/coldman2.jpg?raw=1) | ![](https://www.dropbox.com/s/7qa0gtuy29us4p2/femalejogger2.jpg?raw=1) |
| $Y_J^0=3$ | $Y_M^0=5$ | 
| $Y_J^1=4$ | $Y_M^1=5$ |
| $Y_J^1-Y_J^0=1$ | $Y_M^1-Y_M^0=0$ |
| $Y_J=(Y_J^1)=4$ | $Y_M=(Y_M^0)=5$ |

]

.right-column[
- John will choose to buy health insurance

- Maria will choose to not buy health insurance

- Health insurance improves John's score by 1, has no effect on Maria's score

- Note, all we can observe in the data are their health outcomes *after* they have chosen (not) to buy health insurance: $Y_J=4$ and $Y_M=5$

- .whisper[*Observed* difference] between John and Maria: $$Y_J-Y_M=-1$$
]

---

# Counterfactuals

.left-column[

| John      | Maria     |
|-----------|-----------|
| ![](https://www.dropbox.com/s/5vh181yeg0p91t7/coldman2.jpg?raw=1) | ![](https://www.dropbox.com/s/7qa0gtuy29us4p2/femalejogger2.jpg?raw=1) |
| $Y_J=4$ | $Y_M=5$ |

This is all the data we *actually* observe
]

.right-column[
- Observed difference between John and Maria: 
$$Y_J-Y_M=\underbrace{Y^1_J-Y^0_M}_{=-1}$$

- Recall:
    - John has bought health insurance $(Y^1_J)$
    - Maria has not bought insurance $(Y^0_M)$

- We don't see the .shout[counterfactuals]: 
    - John's score *without* insurance
    - Maria score *with* insurance

]

---

# Counterfactuals

.left-column[

| John      | Maria     |
|-----------|-----------|
| ![](https://www.dropbox.com/s/5vh181yeg0p91t7/coldman2.jpg?raw=1) | ![](https://www.dropbox.com/s/7qa0gtuy29us4p2/femalejogger2.jpg?raw=1) |
| $Y_J=4$ | $Y_M=5$ |

This is all the data we *actually* observe

]

.right-column[

- Observed difference between John and Maria: 
$$Y_J-Y_M=\underbrace{Y^1_J-Y^0_M}_{=-1}$$

- Algebra trick: add and subtract $Y^0_J$ to equation

$$\begin{align*}
Y_j-Y_M=\underbrace{Y^1_J-Y^0_J}_{=1}+\underbrace{Y^0_J-Y^0_M}_{=-2}
\end{align*}$$

- $Y^1_J-Y^0_J=1$: Causal effect for John of buying insurance
- $Y^0_J-Y^0_M=-2$: Difference between John & Maria pre-treatment, .hl["selection bias"]

]

---

# Example II

.pull-left[

$$Y^0_J-Y^0_M \neq 0$$

- .hl[Selection bias:] (pre-existing) differences between members of treatment and control groups *other* than treatment, that affect the outcome
    - i.e. John and Maria *start out* with very *different* health scores before either decides to buy insurance or not ("recieve treatment" or not)

]

.pull-right[

.pull-left[
.center[
![:scale 80%](https://www.dropbox.com/s/vcqojr6k058e8hl/3apples.jpg?raw=1)

John (Treated)
]
]

.pull-right[
.center[
![:scale 100%](https://www.dropbox.com/s/xvjphzlwechgip6/3oranges.jpg?raw=1)

Maria (Control)
]
]
]

---

# Example: Thinking about the Data

.pull-left[
```{r, eval = F, results="asis"}
library("tibble")
tribble(
  ~i, ~Insured, ~Not_Insured, ~Diff,
  "John", 4, 3, 1,
  "Maria", 5, 5, 0,
  "AVERAGE", 4.5, 4.0, 0.5
) %>%
  knitr::kable(., format="html")
```

.center[**Ideal** (but impossible) Data]

| Individual | Insured | Not Insured | Diff |
|---|---------|-------------|------|
| John | 4.0 | 3.0 | 1.0 |
| Maria | 5.0 | 5.0 | 0.0 | 
| **Average** | 4.5 | 4.0 | **0.5** |

- .hl[Individual treatment effect] (for individual $i$):
$$TE_i=Y_i^1-Y_i^0$$
- .hl[*Average* treatment effect]:
$$ATE=\frac{1}{n}\sum^n_{i=1}Y_i^1-Y_i^0$$

]

--

.pull-right[

.center[**Actual** (observed) Data]

| Individual | Insured | Not Insured | Diff |
|---|---------|-------------|------|
| John | 4.0 | ? | ? |
| Maria | ? | 5.0 | ? | 
| **Average** | ? | ? | ? |

- We never get to see each person's **counterfactual** state to compare and calculate ITEs or ATE

]

---

# Example: Thinking about the Data

.pull-left[

- Basic comparisons tell us *something* about outcomes, but not **ATE** (the causal effect we're looking for)

$$\text{Difference in Group Outcomes} = ATE + \text{Selection Bias}$$

- Selection bias: difference in average $Y^0_i$ between groups pre-treatment

- $Y^0_i$ includes *everything* about person $i$ relevant to health *except* treatment (insurance) status
    - Age, sex, height, weight, climate, smoker, exercise, diet, etc. 
    - Imagine a world where *nobody* has insurance (treatment), who would have highest health scores?
]

.pull-right[

.center[**Actual** (observed) Data]

| Individual | Insured | Not Insured | Diff |
|---|---------|-------------|------|
| John | 4.0 | ? | ? |
| Maria | ? | 5.0 | ? | 
| **Average** | ? | ? | ? |

]

---

# Random Assignment: The Silver Bullet

.pull-left[

- If treatment is .hl[randomly assigned] for a large sample, it eliminates selection bias!

- Treatment and control groups differ *on average* by nothing *except* treatment status

- Creates *ceterus paribus* conditions in economics: groups are identical *on average* (holding constant age, sex, height, etc.)

]

.pull-right[

.pull-left[
.center[
![](https://www.dropbox.com/s/vcqojr6k058e8hl/3apples.jpg?raw=1)
Treatment Group
]
]

.pull-right[
.center[
![](https://www.dropbox.com/s/vcqojr6k058e8hl/3apples.jpg?raw=1)
Control Group
]
]
]

---

class: inverse, center, middle
# Quasi-Experiments

---

# The Quest for Causal Effects I

.pull-left[
- RCTs are considered the "gold standard" for causal claims

- But society is not our laboratory (probably a good thing!)

- We can rarely conduct experiments to get data
]

.pull-right[
.center[
![:scale 100%](https://www.dropbox.com/s/j0kwzqb2ptusckt/madlaboratory.jpg?raw=1)
]
]

---

# The Quest for Causal Effects II

.pull-left[
- Instead, we often rely on .whisper[observational data]

- This data is *not random*!

- Must take extra care in forming an .shout[identification strategy]

- To make good claims about causation in society, we must get clever! 
]

.pull-right[
.center[
![:scale 100%](https://www.dropbox.com/s/3j2yflalbv2ylo1/fieldobserver.jpg?raw=1)
]
]

---

# Natural Experiments

.pull-left[

- Economists often resort to searching for .shout[natural experiments]

- Some events beyond our control occur that separate *otherwise similar* entities into a "treatment" group and a "control" group that we can compare

- e.g. natural disasters, U.S. State laws, military draft 
]

.pull-right[
.center[
![](https://www.dropbox.com/s/2yiww124mnvhj8l/nskorea2.png?raw=1)
]
]

---

# The First Natural Experiment

.left-column[
.center[
![:scale 100%](https://www.dropbox.com/s/kc1yx6q6iao26mo/jonsnow.jpg?raw=1)

Jon Snow
]
]

.right-column[

]

---

# The First Natural Experiment

.left-column[
.center[
![:scale 80%](https://www.dropbox.com/s/w5jy5an586l7ep7/johnsnow.jpg?raw=1)

1813-1858
]
]

.right-column[

- John Snow utilized the [first famous natural experiment](http://blogs.plos.org/publichealth/2013/03/11/john-snow-the-first-epidemiologist/) to establish the foundations of epidemiology and the germ theory of disease

- Water pumps with sources *downstream* of a sewage dump in the Thames river spread cholera while water pumps with sources *upstream* did not

.center[
![:scale 60%](https://www.dropbox.com/s/s8ar35t7sjklf4k/choleramap.jpg?raw=1)
]
]

---

# Famous Natural Experiments

- **Oregon Health Insurance Experiment**: Oregon used lottery to grant Medicare access to 10,000 people, showing access to Medicaid increased use of health services, lowered debt, etc. relative to those not on Medicaid
- **Angrist (1990)** finds that lifetime earnings of (random) drafted Vietnam veterans is 15% lower than non-veterans
- **Card & Kreuger (1994)** find that minimum wage hike in fast-food restaurants on NJ side of border had no disemployment effects relative to restaurants on PA side of border during the same period
- **Acemoglu, Johnson, and Robinson (2001)** find that inclusive institutions lead to higher economic development than extractive institutions, determined by a colony's disease environment in 1500
- We will look at some of these in greater detail throughout the course
- [A great list, with explanations is here](http://economicspsychologypolicy.blogspot.com/2015/06/list-of-19-natural-experiments.html)

---

# For the Next Few Classes

.center[
![:scale 45%](https://www.dropbox.com/s/k1xtcebidkonen9/r.png?raw=1)
]

### 2019-08-30 content/reading/02-reading.Rmd
Fixed reading pages for classes 1 & 2, added numbers to schedule; added R4DS to bib

---
title: "1.2 The Quest for Causality - Readings"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-fullnote-bibliography-no-bib.csl
---

# Strongly Recommended

- <i class="fas fa-book"></i> @Bailey[Ch. 1]

Bailey dives right in with the basic linear regression model

$$Y=\beta_0+\beta_1 X$$

and also begins with some demonstrations of the model. Don't worry about this just yet. What is more important right now is the **identification problem** (*endogeneity*) and the **inference problem** (*randomness*) and how these challenges motivate a lot of what we are going to do.

### 2019-09-03 content/class/01-class.Rmd
Fix treatments in slides
Add class 3 content

@@
 ---
-title: "Class 1: Introduction to Econometrics"
+title: "1.1 Introduction to Econometrics - Class Notes"
 citeproc: false
@@
 ---
+
+*Tuesday, August 27, 2019*
 
@@
 - [Click here to join our class Slack channel](https://join.slack.com/t/hoodeconometrics/shared_invite/enQtNzQxMzI1MTU0ODM4LTVhOWI5Zjc5YzdiOGU5YzY0MmZiM2QxOGQwMWQwMjc4Y2FmYTk5Yjg5ZDZhZTY4ZDZiYWU2YzAzMTU1MDIwYTE)^[This link will expire in 7 days]
+
+Please use your hood email address and real name.

### 2019-09-03 content/class/02-class.Rmd
Add reminder to do survey to class 2
Add class 3 content

@@
 ---
-title: "Class 1: Introduction to Econometrics"
+title: "Class 2: The Quest for Causality"
 citeproc: false
@@
 
+*Thursday, August 29, 2019*
+
 # Overview
@@
 
+# Statistics Survey
+
+Don't forget to do the [preliminary survey](/assignment/00-preliminary-survey.html) for next class!
+
 # Slack

### 2019-09-03 content/class/03-class.Rmd
Add class 3 content

---
title: "1.3 Meet R - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

*Tuesday, September 3, 2019*

# Overview

Today we begin the long slog to your mastery of `R`.

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/02-slides.html)

# Appendix

## Installing R and R Studio

1. [Install R](http://www.r-project.org/) from CRAN^[The Comprehensive R Archive Network] by clicking "Download R" (or the CRAN link under Downloads on the left menu). This will take you to a mirrors page, where you can select a location in the U.S. and download a copy of R
2. [Install R Studio (Desktop Version)](https://www.rstudio.com/products/rstudio/download/), choose the "Free" option

(This will also be posted on the [Reference](/reference) page.)

## R Packages

Packages come from multiple sources.

The polished, publicly released versions are found on [CRAN](https://cran.r-project.org/). When installing a package available on CRAN, it is sufficient simply to tell R the following:^[Note the plural `s` on `packages`, and the quotes around the "package name"]

```{r, eval=F}
install.packages("packagename") 
```

Other packages, which may be in various developmental states (including perfectly functional!) are often hosted on [GitHub](http://github.com) before they make their way onto CRAN. Simply telling R `install.packages("packagename")` will fail to find it (as R only looks in CRAN for packages), so you must use another package called `devtools`^[Which you will need to install first if you (probably) don't already have it!] to install packages directly from Github:^[Note the `::` allows you to use the function `install_github()` from the `devtools` package without having to first load the devtools package with `library(devtools)`.]

```{r, eval=F}
devtools::install_github("username/packagename") 
```

For example, to install Hadley Wickham's package `r4ds` from its Github page [https://github.com/hadley/r4ds](https://github.com/hadley/r4ds), we would type:

```{r, eval=F}
devtools::install_github("hadley/r4ds")
```

To **use** a package, you need to ensure it is loaded to your workspace (you only need to do this once)^[When we learn how to write `R Markdown` documents, . with `library("package_name")`.^[Quotes are not necessary this time.]

## Getting Help for R

For specific functions or commands, you can simply type:

```{r, eval=F}
?functionname()

# example
?mean()
```

This will display a help page specific to that function in the Viewer pane. R functions and packages are extremely well-documented; help pages [normally ](http://socviz.co/assets/ch-09-read-a-help-page.png)^[This useful guide comes from Kieran Healy's excellent (free online!) [book on Data Visualization](http://socviz.co/).] a short description of the function, arguments and options (as well as their default values), and several examples or vignettes to demonstrate usage of the function.

Additionally, you can turn to the community.

## Other Useful Commands to Know

One of the best/worst things about `R` is that it is a language, meaning there are multiple ways that you can accomplish the same task.

### Creating Vectors

We know vectors can be created with the `c()` command (and stored with `<-`), but there are other shortcuts to combine objects into a vector, particularly numeric data:

- `:` creates a series of integers

```{r}
1:5 # create a vector of 1 through 5

12:17 # create a vector of 12 through 17
```

- `seq(from = , to = , by = )` creates a numeric sequence, and is not restricted to integers

```{r}
seq(from = 1, to = 10, by = 2) # sequence from 1 to 10, by 2s

# note you do not need to fully write out the name of each argument, just the input!

seq(32.5,40,1.5) # sequence from 32.5 to 40, by 1.5
```

- `rep(., times =)`^[`.` is a placeholder here.] repeats an element a specified number of times

```{r}
rep(2, times = 4) # repeat "2" four times

rep(2, 4) # does the same thing

# the thing repeated could itself be a vector
rep(c(1,4,7), 3) # repeat the vector "1, 4, 7" three times
```

- We can combine these:

```{r}

# combine (the sequence of 4 to 8 by 2's repeated three times) and 1 and 5

c(rep(seq(4,8,2),3),1,5)
```

## Style Guide for Coding

We want to maximize *human*-readability of code, not just machine-readability. I try to follow [Hadley Wickham's style guide](http://adv-r.had.co.nz/Style.html) for all of my code, including that in this class.

*You will not be graded on the style of your code.* But now's the best time to learn best practices (while you don't know any alternatives!) to save yourself and your potential colleagues (including your future self) from unnecessary frustration.

### 2019-09-03 content/r/03-r-practice.Rmd
Add class 3 content

---
title: "1.3 Meet R - R Practice"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

# Getting Set Up

Before we begin, start a new file with `File` $\rightarrow$ `New File` $\rightarrow$ `R Script`. As you work through this sheet in the console in `R`, also add (copy/paste) your commands that work into this new file. At the end, save it, and run to execute all of your commands at once.

# Creating Objects

## 1.

a. Create a vector called `me` with two objects, your first name, and your last name. 
b. Call the vector to inspect it.
c. Confirm it is a character class vector.

## 2. Use R's help functions to determine what the `paste()` function does. Then paste together your name from `me`.

## 3. Create a vector called `my_vector` with all the even integers from 2 to 10.

## 4. Find the mean of x with `mean()`.

## 5. Add up all the numbers from 18 to 763, then get the mean

## Playing with Data

For the following questions, we will the `diamonds` dataset, included as part of `ggplot2`.

### 6. Install `ggplot2`.

### 7. Load `ggplot2` with the `library()` command.

### 8. Get the `str`ucture of the diamonds data frame. What are the different variables and what kind of data does each contain?

### 9. Get summary statistics separately for `carat`, `depth`, `table`, and `price`.

### 10. `color`, `cut`, and `clarity` are categorical variables (factors). Use the `table()` command to generate frequency tables for each.

### 11. Now rerun the `summary()` command on the entire data frame.

### 12. Now look only at (subset) the first 4 diamonds in the dataset.

### 13. Now look only at (subset) the third and seventh diamond in the dataset.

### 14. Now look only at (subset) the second column of the dataset.

### 15. Do this again, but look using the `$` to pull up the second column by name.

### 16. Now look only at diamonds that have a `carat` greater than or equal to 1.

### 17. Now look only at diamonds that have a VVS1 `clarity`.

### 17. Now look only at dimaonds that have a color of E, F, I, and J.

### 18. Now look only at diamonds that have a `carat` greater than or equal to 1 and a VVS1 `clarity`.

### 19. Get the average price of diamonds in question 18.^[Hints: use your subset command as an argument to the mean function. You will not need a comma here because you are looking for a single row.]

### 20. What is the highest price for a diamond with a 1.0 carat, D color, and VVS1 clarity?

## Execute your R Script

Save the R Script you created at the beginning and (hopefully) have been pasting all of your valid commands to. This creates a *.R* file wherever you choose to save it to. Now looking at the file in the upper left pane of *R Studio* look for the button in the upper right corner that says **Run**. Sit back and watch `R` redo everything you've carefully worked on, all at once.

### 2019-09-03 content/reading/03-reading.Rmd
Add class 3 content

---
title: "1.3 Meet R - Readings"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-fullnote-bibliography-no-bib.csl
---

# Strongly Recommended

- <i class="fas fa-external-link-square-alt"></i>@R4DS[Ch. 1] 1^[[https://r4ds.had.co.nz/introduction.html](https://r4ds.had.co.nz/introduction.html)]

Now that we start working with R, we will

### 2019-09-03 static/r/03-r-practice.Rmd
Add class 3 content

---
title: "1.3 Meet R - R Practice"
author: "Ryan Safner"
date: "ECON 480 - Fall 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting Set Up

Before we begin, start a new file with \texttt{File} $\rightarrow$ \texttt{New File} $\rightarrow$ \texttt{R Script}. As you work through this sheet in the console in `R`, also add (copy/paste) your commands that work into this new file. At the end, save it, and run to execute all of your commands at once.

## Creating Objects

### 1. Create a vector called "me" with two objects, your first name, and your last name. Then call the vector to inspect it. Confirm it is a character class vector.

### 2. Create a vector called "x" with all the even integers from 2 to 10.

### 3. Find the mean of x with `mean()`

### 4. Now take the following pdf of random variable $Y$:

\begin{table}[h!]
\centering
  \begin{tabular}{ll}
  $y_i$ & $p_i$\\ \hline
  2 & 0.50\\
  4 & 0.25\\
  6 & 0.25\\ \hline
  \end{tabular}
\end{table}

Calculate the standard deviation "manually" using our table method. You can look at the source code of Lecture 4 for my example.

### a. Creating two vectors, one called $y.i$ and one called $p.i$, with the data above.

### b. Merge them into a data frame called $rv$ with `data.frame(y.i,p.i)`. Call `rv` to inspect it.

### c. Find the expected value of $Y$ by taking the sum of each value of `y.i` multiplied by `p.i` with the `sum()` command.

### d. Creating a new column in `rv` called `deviations`, where you subtract the mean from each `y.i` value. Call `rv` again to make sure it's now there.

### e. Create another column in `rv` called `devsq`, where you square the deviations from part d. Call `rv` again to make sure it's now there.

### f. Now add another column in `rv` called `weighteddevsq`, where you multiply the squared deviations in part e. by the associated probability `p.i`. Call `rv` again to make sure it's now there.

### g. Finally, take the sum of `weighteddevsq` to get variance. Square root this to get standard deviation.

### 5. The mean height of adults is 65 inches, with a standard deviation of 4 inches. Use the normal distribution to find the probabilities of the following scenarios:

### a. Find the probability of someone being *at least* 60 inches tall using `pnorm()`.

### b. Find the probability of someone being *at most* 60 inches tall.

### c. Find the probability of someone being between 61 and 69 inches tall. Why is this number familiar?

### d. Find the probability of someone being between 57 and 73 inches tall. Why is this number familiar?

## Playing with a Data Set

For the following questions, use the \texttt{diamonds} dataset, included as part of `ggplot2`.

### 6. Install `ggplot2`

### 7. Load `ggplot2` with the `library()` command

### 8. Get the structure of the diamonds data.frame. What are the different variables and what kind of data does each contain?

### 9. Get summary statistics for `carat`, `depth`, `table`, and `price`

### 10. `color`, `cut`, and `clarity` are categorical variables (factors). Use the `table()` command to generate frequency tables for each.

### 11. Now rerun the `summary()` command on the entire data frame

### 12. Plot a histogram of price.

### 13. Plot a boxplot of price by diamond color.

## Execute your R Script

Save the R Script you created at the beginning and (hopefully) have been pasting all of your valid commands to. This creates a *.R* file wherever you choose to save it to. Now looking at the file in the upper left pane of *R Studio* look for the button in the upper right corner that says **Run**. Sit back and watch `R` redo everything you've carefully worked on, all at once.

### 2019-09-03 static/slides/03-slides.Rmd
Add class 3 content

---
title: "1.3: Meet R"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github #ocean rainbow #monokai rainbow solarized-light #tomorrow-night -blue -eighties zenburn
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(message=F,
                      warning=F)
library(tidyverse)

```

# Data Science

- .font120[.onfire[You go into data analysis with the tools you know, not the tools you need]]

- The next 2-3 weeks are all about giving you the tools you need 
    - Admittedly, a bit before you know what you need them *for*

- We will extend them as we learn specific models

---

# Why Not Excel? I

.center[
![:scale 65%](https://www.dropbox.com/s/t38twgzia49lexn/excelsucks.jpg?raw=1)
]

---

# Why Not Excel? II

.center[
![:scale 65%](https://www.dropbox.com/s/yoiuuifbm3n58bd/excelsucks2.png?raw=1)
]

---

# Why Use R?

.pull-left[

- **Free** and **open source**

- A very large community
    - Written by statisticians for statistics
    - Most packages are written for `R` first

- Can handle virtually any data format

- Makes replication easy

- Can integrate into documents (with `R markdown`)

- R is a *language* so it can do *everything*
    - A good stepping stone to learning other languages like *Python*

]

.pull-right[

.center[
![:scale 90%](https://www.dropbox.com/s/k1xtcebidkonen9/r.png?raw=1)
]

]

---

# Excel and Stata Can't Do This (Execute Inside the Slides)

.left-code[

```{r gapplot, eval=F, echo=T}
library("gapminder")

ggplot(data = gapminder, 
       aes(x = gdpPercap,
           y = lifeExp,
           color = continent))+
  geom_point(alpha=0.3)+
  geom_smooth(method = "lm")+
	scale_x_log10(breaks=c(1000,10000, 100000),
	              label=scales::dollar)+
	labs(x = "GDP/Capita",
	     y = "Life Expectancy (Years)")+
  facet_wrap(~continent)+
  guides(color = F)+
  theme_light()
```
]

.right-plot[
.center[
```{r, ref.label="gapplot", echo=F, fig.retina=3, out.width="72.5%"}
```
]
]

---

# Or This: Execute R Code Inside Your Documents

.pull-left[
# Code

  ```{r, eval=F, echo=T}
  library(gapminder)
  ```

The average GDP per capita is $`` `r
round(mean(gapminder$gdpPercap),2)` `` with a standard deviation of $`` `r
round(sd(gapminder$gdpPercap),2)` ``.  
]

.pull-right[
# Output

```{r, echo=F}
library(gapminder)
```

The average GDP per capita is $`r round(mean(gapminder$gdpPercap),2)` with a standard deviation of $`r round(sd(gapminder$gdpPercap),2)`.

]

---

class: inverse, center, middle
# Meet R and R Studio

---

# *R* and *R Studio* I

.pull-left[

- *R* is the programming language that executes commands

- *R Studio* is an integrated development environment (IDE) that makes your coding life a lot easier
    - Write code in scripts
    - Execture individual commands or entire scripts
    - Auto-complete, highlight syntax
    - View data, objects, and plots
    - Get help and documentation on commands and functions
    - Integration code into documents with *R Markdown*
]

.pull-right[

.center[
![R Studio](https://www.dropbox.com/s/foczpb1z0fixw73/rstudio.png?raw=1)

R Studio
]
]

---

# *R* and *R Studio* II

.pull-left[

- Download and install [R](https://cran.cnr.berkeley.edu/) & [R Studio](https://www.rstudio.com/products/rstudio/download/)<sup>.red[1]</sup> (in that order)

- R is like your car's engine, R Studio is the dashboard

- You will do everything in R Studio and never open the R program itself

- R itself is just a command language 
    - The `R` *app* is basically just a command line
    - You can even just use your computer's command line!<sup>.red[2]</sup>
]

.pull-right[

.center[
![R Studio](https://www.dropbox.com/s/foczpb1z0fixw73/rstudio.png?raw=1)

R Studio
]
]

.footnote[<sup>.red[1]</sup>The (free) Desktop version.

<sup>.red[2]</sup> "Command Prompt" on Windows, "Terminal" on Unix (Mac and Linux). Type `r` and hit enter, and you can now execute `R` commands.]

---

# *R* and *R Studio* III

.pull-left[

- R Studio has 4 window panes:
    1. **Source**<sup>.red[1]</sup>: a text editor for documents, R scripts, etc.
    2. **Console**: type in commands to run
    3. **Browser**: view files, plots, help, etc
    4. **Environment**: view created objects, command history, version control

- Customize the sizes of these panes (I often hide the console and use the text editor)
    - `CTRL+SHIFT+[number]` will maximize a pane. Type again to see all four.

]

.pull-right[

.center[
![R Studio](https://www.dropbox.com/s/foczpb1z0fixw73/rstudio.png?raw=1)

R Studio
]
]

.footnote[<sup>.red[1]</sup>May not be immediately visible until you create new files.]

---

# Learning...

.pull-left[

- .font130[.onfire[You don't "learn R, you learn how to do things in R]]

- In order to do learn this, you need to learn *how to search for what you want to do*

]

--

.pull-right[

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">My <a href="https://twitter.com/hashtag/rstats?src=hash&amp;ref_src=twsrc%5Etfw">#rstats</a> learning path:<br><br>1. Install R<br>2. Install RStudio<br>3. Google &quot;How do I [THING I WANT TO DO] in R?&quot;<br><br>Repeat step 3 ad infinitum.</p>&mdash; Jesse Mostipak (@kierisi) <a href="https://twitter.com/kierisi/status/898534740051062785?ref_src=twsrc%5Etfw">August 18, 2017</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">A surprisingly large part of having expertise in a topic is not so much knowing everything about it but learning the language and sources well enough to be extremely efficient in google searches.</p>&mdash; Katie Mack (@AstroKatie) <a href="https://twitter.com/AstroKatie/status/1071442842873159681?ref_src=twsrc%5Etfw">December 8, 2018</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

]

---

# ...and Sucking

.center[
![:scale 80%](https://www.dropbox.com/s/se0x11ft9xr5ea7/suckingatsomething.gif?raw=1)
]

---

class: inverse, center, middle
# Ways to Use R

---

# 1. Using the Console

.pull-left[

- Type individual commands into the console window

- Great for testing individual commands to see what happens

- Not saved! Not reproducible! Not recommended! 
]

.pull-right[

```{r, echo=T}
2+2
summary(mpg$hwy)
```
]

---

# 1. Using the Console

.pull-left[

- Type individual commands into the console window

- Great for testing individual commands to see what happens

- Not saved! Not reproducible! Not recommended! 
]

.pull-right[

```{r, echo=T, fig.width = 6, fig.asp = 0.618, fig.retina=2, fig.align="center"}
ggplot(mpg, aes(x=displ, y=hwy))+
geom_point(aes(color=class))+
geom_smooth()
```

]

---

# 2. Writing an R Script

.pull-left[

- Source pane is a text-editor

- Make `.R` files: all input commands in a single script

- Comment with `#`

- Can run any or all of script at once

- Can save, reproduce, and send to others!

]

.pull-right[

```{r, eval=F, echo=T}
2+2 # just testing!
head(mpg) # look at mpg data

# create a plot

ggplot(mpg, aes(x=displ, y=hwy))+
geom_point(aes(color=class))+
geom_smooth()
```
]

---

# 3. Using Markdown

- For a later lecture: `R Markdown`, a simple markup language to write documents in
    - **Optional**, but many students have enjoyed it and use it well beyond this class!

- Can integrate text, `R` code, figures, citations and bibliographies into a *single* plain-text file<sup>.red[1]</sup> and then output into a variety of formats: PDF, webpage, slides, Word doc, etc.

.center[
![](https://www.dropbox.com/s/cj48ufcnvszzgbn/rmd3.png?raw=1)
]

.footnote[<sup>.red[1]</sup>OK, to be fair, citations require one additional file!]

---

# For Today

- Practicing typing at the Command line/Console

- Learning different commands and objects relevant for analysis

- Saving and running `.R` scripts

- Later: `R markdown`, literate programming, workflow management

- Today may seem a bit overwhelming
    - You don't need to know or internalize all of this today
    - Use this as a reference to come back to over the semester
    - Last year I started making [a partial book](http://ryansafner.com/tutorial/r4eh), I may revive it

---

class: inverse, center, middle
# Coding Basics

---

# Getting to Know Your Computer

- R assumes a default (often inconvenient) .shout["working directory"] on your computer
    - The first place it looks to `open` or `save` files

- Find out where R this is with `getwd()`

- Change it with `setwd(path/to/folder)`<sup>.red[1]</sup>

- Soon I'll show you better ways where you won't ever have to worry about this

.footnote[<sup>.red[1]</sup> Note the path will be OS-specific. For Windows it might be `C:/Documents/`. For Mac it is often your username folder.]

---

# Coding

.left-column[

.center[
![:scale 80%](https://www.dropbox.com/s/u34dakjyq4hkyov/hadley.jpg?raw=1)

Hadley Wickham

Chief Scientist, R Studio
]
]

.right-column[

> "There’s an implied contract between you and R: it will do the tedious computation for you, but in return, you must be completely precise in your instructions. Typos matter. Case matters." - [R for Data Science, Ch. 4](r4ds.had.co.nz/workflow-basics.html)
]

---

# Say Hello to My Little Friend

.center[
![](https://www.dropbox.com/s/ojoax1g7zqytyqr/googlehelp.png?raw=1)
]

---

# Say Hello to My Better Friend

.center[
![](https://www.dropbox.com/s/wcv8rtxikmei9yn/stackoverflow.png?raw=1)
]

---

# R Is Helpful Too!

- type `help(function_name)` or `?(function_name)` to get documentation on a function
.center[
![:scale 50%](http://socviz.co/assets/ch-09-read-a-help-page.png)

From Kieran Healy's excellent (free online!) [book on Data Visualization](http://socviz.co/).]
]
---

# Tips for Writing Code

- Comment, comment, comment!
- The hashtag `#` starts a comment, R will ignore everything on the rest of that line

```{r, eval=F}
# Run regression of y on x, save as reg1 
reg1<-lm(y~x, data=data) #runs regression 
summary(reg1$coefficients) #prints coefficients
```

- Save often!
  - Write scripts that save the commands that did what you wanted (and comment them!)
  - Better yet, use a version control system like Git (I hope to cover this later)

---

# Style and Naming

- Not relevant yet, but once we start writing longer blocks of code, it helps to have a consistent (and human-readable!) style
    - I follow [this style guide](http://adv-r.had.co.nz/Style.html) (you are not required to)<sup>.red[1]</sup>

- Naming objects and files will become important<sup>.red[2]</sup>
    - DO NOT USE SPACES! You've seen seen webpages intended to be called `my webpage in html` turned into `http://my%20webpage%20in%20html.html`

```{r,eval=F}
i_use_underscores
some.people.use.snake.case
othersUseCamelCase
```

.footnote[<sup>.red[1]</sup> Also described in [today's course notes page](http://metricsf19.classes.ryansafner.com/class/03-class/) and the course [reference page](http://metricsf19.classes.ryansafner.com/reference.

<sup>.red[2]</sup> Consider your folders on your computer as well...]

---

# Coding Basics

- You'll have to get used to the fact that you are coding in commands to execute

- Start with the easiest: simple math operators and calculations:

--

```{r, eval=F}
> 2+2
```

```{r, echo=F}
2+2
```

--

- Note that R will ask for **input** with `>` and give you **output** starting with `## [1]`

---

# Coding Basics II

- We can start using more fancy commands

```{r}
2^3
```

--

```{r}
sqrt(25)
```

--

```{r}
log(6)
```

--

```{r}
pi/2
```

---

# Packages

.pull-left[

- Since R is open source, users contribute **packages**
    - Really it's just users writing custom functions and saving them for others to use
- Load packages with `library()`
    - e.g. `library("package_name")`
- If you don't have a package, you must first `install.packages()`<sup>.red[1]</sup> 
    - e.g. `install.packages("package_name")`
]

.pull-right[

.center[
![:scale 90%](https://www.dropbox.com/s/wxibs3jfsdyuvd4/rpackages.png?raw=1)
]

]
.footnote[<sup>.red[1]</sup> Yes, note the plural, even if it's just for one package!]

---

# Common Packages

.pull-left[

- In this class, we will make extensive use of the following packages:
    - `tidyverse`: really a *meta*-package combining the following packages (among others)
        - `dplyr`: used for better data-wrangling
        - `ggplot2`: used for fancy plotting
    - `huxtable`: used for automatically producing regression tables
]

.pull-right[

.center[
![:scale 75%](https://www.dropbox.com/s/70vjwlyctd55hji/tidyversevertical.png?raw=1)
]
]
---

# R: Objects and Functions

.pull-left[
- R is an "**object-oriented**" programming language
- 99% of the time, you will be:

1. creating `objects`
    - assign values to an object with `<-`

2. running `functions` on `objects`
    - syntax: `function_name(object_name)`
]

.pull-right[

```{r}
# make an object
my_object<-c(1,2,3,4,5)

# look at it 
my_object

# find the sum
sum(my_object)

# find the mean 
mean(my_object)
```
]
---

# R: Objects and Functions II

.pull-left[
- Functions have ".onfire[arguments]," the input(s)

- Some functions may have multiple inputs

- The argument of a function can be another function!

]

.pull-right[

```{r}
# find the sd
sd(my_object)

# round everything in my object to two decimals
round(my_object,2)

# round the sd to two decimals
round(sd(my_object),2)
```
]

---

class: inverse, center, middle

# Types of R Objects

---

# Numeric

.pull-left[
- .shout[Numeric] objects are just numbers<sup>.red[1]</sup>

- Can be mathematically manipulated
]

.pull-right[
```{r}
x <- 2 
y <- 3
x+y
x*y
```
]
.footnote[.red[<sup>1</sup>] If you want to get technical, R may call these `integer` or `double` if there are decimal values.]

---

# Character

.pull-left[
- .shout[Character] objects are **"strings"** of text held inside quote marks

- Can contain spaces, so long as contained within quote marks
]

.pull-right[
```{r}
name <- "Ryan Safner"
address <- "401 Rosemont Ave."

name
address
```
]

---

# Logical

.pull-left[
- .shout[Logical] objects are **binary** `TRUE` or `FALSE` indicators
- Used a lot to evaluate *conditionals*:
  - `>`, `<`: greater than, less than
  - `>=`, `<=`: greater than or equal to, less than or equal to
  - `==`, `!=`: is equal to, is not equal to<sup>.red[1]</sup>
  - `&in&` : Is a member of the set of ($\in$)
  - `&`: "AND"
  - `|`: "OR"

]

.pull-right[

```{r}

z <- 10 # set z equal to 10

z==10 # is z equal to 10?

"red"=="blue" # is red equal to blue?

z > 1 & z < 12 # is z > 1 AND < 12?

z <= 1 | z==10 # is z >= 1 OR equal to 10?

```

]

.footnote[<sup>.red[1]</sup> One `=` *assigns* a value (like `<-`). Two `==` *evaluate* a conditional statement.]
---

# Factor

- .shout[Factor] objects contain **categorical** data - membership in mutually exclusive groups

- Look like strings, behave more like logicals, but with more than two options

```{r, echo=F}
class <- sample(c("freshman", "sophomore", "junior", "senior"),10, replace=T)
```

```{r, echo=F}
class <- as.factor(class)
levels(class)<-c('freshman','sophomore','junior', 'senior')
class
```

- We'll make much more extensive use of them later

```{r, echo=F}
class <- ordered(class, levels=c('freshman','sophomore','junior', 'senior'))
class
```

---

class: inverse, center, middle

# Data Structures

---

# Vectors

.pull-left[

- `Vector`: the simplest type of object, just a collection of objects

- Make a vector using the combine `c()` function
]

.pull-right[

```{r}
# create a vector called vec
vec<-c(1,"orange", 83.5, pi)

# look at vec
vec
```
]

---

# Data Frames I

.pull-left[

- `Data frame`: what we'll be using almost always

- Think like a "spreadsheet"

- Each *column* is a vector (variable)

- Each *row* is an observation (pair of values for all variables)
]

.pull-right[

```{r}
library("ggplot2")

diamonds

```

]
---

# Data Frames II

.pull-left[

- Dataframes are really just combinations of (column) vectors

- You can make data frames by combinining named vectors with `data.frame()` or creating each column/vector in each argument 
]

```{r}
# make two vectors
fruits<-c("apple","orange","pear","kiwi","pineapple")
numbers<-c(3.3,2.0,6.1,7.5,4.2)

# combine into dataframe
df<-data.frame(fruits,numbers)

# do it all in one step (note the = instead of <-)
df<-data.frame(fruits=c("apple","orange","pear","kiwi","pineapple"),
               numbers=c(3.3,2.0,6.1,7.5,4.2))

# look at it
df
```

]

---

class: inverse, center, middle
# Working with Objects

---

# Objects: Storing, Viewing, and Overwriting

- We want to store things in objects to run functions on them later
- Recall, any object is created with the assignment operator `<-`

```{r}
my_vector <- c(1,2,3,4,5)
```

- R will not give any output after an assignment

--

- *View* an object (and list its contents) by typing its name

```{r}
my_vector
```

--

- objects maintain their values until they are assigned different values that will *overwrite* the object

```{r}
my_vector <- c(2,7,9,1,5)
my_vector
```

---

# Objects: Checking and Changing Classes

- Check what type of object something is with `class()`

```{r}
class("six")
class(6)
```

--

- Can also use logical tests of `is.()`

```{r}
is.numeric("six")
is.character("six")
```

- You can convert objects from one class to another with `as.object_class()`
    - Pay attention: you can't convert non-numbers to `numeric`, etc!

```{r}
as.character(6)
as.numeric("six")
```

---

# Objects: Different Classes and Coercion I

- Different types of objects have different rules about mixing classes
- Vectors can *not* contain different types of data
    - Different types of data will be "**coerced**" into the lowest-common denominator type of object

```{r}
mixed_vector <- c(pi, 12, "apple", 6.32)
class(mixed_vector)
mixed_vector
```

---

# Objects: Different Classes and Coercion II

- Data frames can have columns with different types of data, so long as all the elements in each column are the same class<sup>.red[1]</sup>

.pull-left[
```{r}
df
```
]

.pull-right[

```{r}
class(df$fruits)
class(df$numbers)
```
]

.footnote[<sup>.red[1]</sup>remember each column in a data frame is a vector!]

---

# More on Data Frames I

- Learn more about a data frame with the `str()` command to view its structure

```{r}
class(df) 
str(df)
```

---

# More on Data Frames II

- Take a look at the first 5 (or `n`) rows with `head()`

```{r}
head(df)
head(df, n=2)
```

---

# More on Data Frames III

- Get summary statistics<sup>.red[1]</sup> by column (variable) with `summary()`

```{r}
summary(df)
```

.footnote[<sup>.red[1]</sup> for `numeric` data only; a frequency table is displayed for `character` or `factor` data]
---

# More on Data Frames IV

- Note, once you save an object, it shows up in the **Environment Pane** in the upper right window 
- Click the blue arrow button in front of the object for some more information

.center[
![:scale 75%](https://www.dropbox.com/s/ibuv8tqnecnz3u9/renvironmentdf2.png?raw=1)
]
---

# More on Data Frames V

- `data.frame` objects can be viewed in their own panel by clicking on the name of the object
- Note you cannot edit anything in this pane, it is for viewing only

.center[
![:scale 75%](https://www.dropbox.com/s/90kgn2ttkpwqqcz/rdfview.png?raw=1)
]
---

# Functions Again I

- Functions in R are vectorized, meaning running a function on a vector applies it to *each* element

```{r}
my_vector<-c(2,4,5,10)
my_vector+4 # add 4 to all elements
my_vector^2 # square all elements
```

---

# Functions Again II

- But often we want to run functions on vectors that *aggregate* to a result (e.g. a statistic):

```{r}
length(my_vector) # how many elements
sum(my_vector) # add all elements
max(my_vector) # find largest element
min(my_vector) # find smallest element
mean(my_vector) # mean of all elements
median(my_vector) # median of all elements
sd(my_vector) # standard deviation
```

---

# Common Errors

- If you make a coding error (e.g. forget to close a parenthesis), R might show a `+` sign waiting for you to finish the command

```{r, echo=T, eval=F}
> 2+(2*3
+
```

- Either finish the command-- e.g. add `)`--or hit `Esc` to cancel

---

class: inverse, center, middle

# Working with Data

---

# Indexing and Subsetting I

.pull-left[
```{r, echo=F}
mtcars <- mtcars %>%
  select(1:7) %>%
  slice(1:12)
```

```{r}
mtcars
```

]

.pull-right[

- Each element in a data frame is .h1[indexed] by referring to its row and column: `df[r,c]`
- To select elements by row and column (".h1[subset]"), type in the row(s) and/or column(s) to select
    - Leaving `r` or `c` blank selects *all* rows or columns
    - Select multiple values with `c()`<sup>1</sup>
    - Select a range of values with `:`
    - Don't forget the comma between `r` and `c`! 
]

.footnote[<sup>1</sup> You can also "negate" values, selecting everything *except* for values with a `-` in front of them.]

---

# Indexing and Subsetting II

.pull-left[
```{r}
mtcars
```
]

.pull-right[

### Subset by Row (Observations)

```{r}
mtcars[1,] # first row
mtcars[c(1,3,4),] # first, third, and fourth rows
mtcars[1:3,] # first three rows
```

]

---

# Indexing and Subsetting III

.pull-left[
```{r}
mtcars
```
]

.pull-right[

### Subset by Column (Variable)

```{r}
mtcars[,6] # select column 6

mtcars[,2:4] # select columns 2 through 4
```

]
---

# Indexing and Subsetting IV

.pull-left[
```{r}
mtcars
```
]

.pull-right[

### Subset by Column (Variable)

- Alternatively, double brackets `[[]]` selects a column by position

```{r}
mtcars[[6]] # same thing
```

- Data frames can select columns by *name* with `$`

```{r}
mtcars$wt
```

]

---

# Indexing and Subsetting V

.pull-left[
```{r}
mtcars
```

]

.pull-right[

- Select observations (rows) that meet logical criteria

### Subset by Condition

```{r}
mtcars[mtcars$wt>4,] # select obs with wt>4
mtcars[mtcars$cyl==6,] # select obs with exactly 6 cyl
mtcars[mtcars$wt<4 & mtcars$wt>2,] # select obs where 2<wt<4
mtcars[mtcars$cyl==4 | mtcars$cyl==6,] # select obs with 4 OR 6 cyl
```

]

---

class: inverse, center, middle
# Resources

---

# What's To Come

- Next class: data visualization with `ggplot2`

- And then: data wrangling with `tidyverse`

- And then: literate programming and workflow management with `R Markdown`

- Finally: back to econometric theory!

### 2019-09-03 content/class/03-class.Rmd
Update class 3 materials

@@
 
-- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/02-slides.html)
+- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/03-slides.html)
 
-# Appendix
+# Practice Problems
+
+Today you will be working on [R practice problems](/r/03-r-practice.html). Answers will be posted later.
+
+# Statistics and Software Survey
+
+Please turn in your math surveys. I will post "answers" after both sections have turned in their surveys.
+
+# Additional Useful Information
 
 ## Installing R and R Studio
+
+We will be using the trading room computers all semester, and I will ensure you have access to the computers outside of class so that you can work on homeworks and your projects on them.
+However, since R is free, I **strongly recommend** you download and install it on your own computers.
+You may decide also you prefer to use your own computers in class when we work with R.
 
@@
 
+## Other Resources for R
+
+My R4EH guide.
+
 ## Other Useful Commands to Know
@@
 One of the best/worst things about `R` is that it is a language, meaning there are multiple ways that you can accomplish the same task.
+Here are a few alternative methods relevant to what we have learned so far that might prove useful.
 
@@
 
-- `:` creates a series of integers
+1. `:` creates a series of integers
 
@@
 
-- `rep(., times =)`^[`.` is a placeholder here.] repeats an element a specified number of times
+2. `rep(., times =)`^[The `.` is a placeholder here.] repeats an element a specified number of times
 
@@
 
-- We can combine these:
+3. We can combine these:
 
@@
 
+### Other Object Types
+
 ## Style Guide for Coding

### 2019-09-03 content/r/03-r-practice.Rmd
Update class 3 materials

@@
 ```
+- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/r/03-r-practice.pdf)
 

### 2019-09-03 static/r/03-r-practice.Rmd
Update class 3 materials

@@
 
-## Getting Set Up
+# Getting Set Up
 
-Before we begin, start a new file with \texttt{File} $\rightarrow$ \texttt{New File} $\rightarrow$ \texttt{R Script}.
+Before we begin, start a new file with `File` $\rightarrow$ `New File` $\rightarrow$ `R Script`.
 As you work through this sheet in the console in `R`, also add (copy/paste) your commands that work into this new file.
@@
 
-## Creating Objects
+# Creating Objects
 
-### 1. Create a vector called "me" with two objects, your first name, and your last name. Then call the vector to inspect it. Confirm it is a character class vector.
+## 1.
 
-### 2. Create a vector called "x" with all the even integers from 2 to 10.
+a. Create a vector called `me` with two objects, your first name, and your last name.
+b. Call the vector to inspect it.
+c. Confirm it is a character class vector.
 
-### 3. Find the mean of x with `mean()`
+## 2. Use R's help functions to determine what the `paste()` function does. Then paste together your name from `me`.
 
-### 4. Now take the following pdf of random variable $Y$:
+## 3. Create a vector called `my_vector` with all the even integers from 2 to 10.
 
-\begin{table}[h!]
-\centering
-  \begin{tabular}{ll}
-  $y_i$ & $p_i$\\ \hline
-  2 & 0.50\\
-  4 & 0.25\\
-  6 & 0.25\\ \hline
-  \end{tabular}
-\end{table}
+## 4. Find the mean of x with `mean()`.
 
-Calculate the standard deviation "manually" using our table method. You can look at the source code of Lecture 4 for my example.
+## 5. Add up all the numbers from 18 to 763, then get the mean
 
-### a. Creating two vectors, one called $y.i$ and one called $p.i$, with the data above.
+\clearpage
 
-### b. Merge them into a data frame called $rv$ with `data.frame(y.i,p.i)`. Call `rv` to inspect it.
+# Playing with Data
 
-### c. Find the expected value of $Y$ by taking the sum of each value of `y.i` multiplied by `p.i` with the `sum()` command.
+For the following questions, we will the `diamonds` dataset, included as part of `ggplot2`.
 
-### d. Creating a new column in `rv` called `deviations`, where you subtract the mean from each `y.i` value. Call `rv` again to make sure it's now there.
+### 6. Install `ggplot2`.
 
-### e. Create another column in `rv` called `devsq`, where you square the deviations from part d. Call `rv` again to make sure it's now there.
+### 7. Load `ggplot2` with the `library()` command.
 
-### f. Now add another column in `rv` called `weighteddevsq`, where you multiply the squared deviations in part e. by the associated probability `p.i`. Call `rv` again to make sure it's now there.
+### 8. Get the `str`ucture of the diamonds data frame. What are the different variables and what kind of data does each contain?
 
-### g. Finally, take the sum of `weighteddevsq` to get variance. Square root this to get standard deviation.
-
-### 5. The mean height of adults is 65 inches, with a standard deviation of 4 inches. Use the normal distribution to find the probabilities of the following scenarios:
-
-### a. Find the probability of someone being *at least* 60 inches tall using `pnorm()`.
-
-### b. Find the probability of someone being *at most* 60 inches tall.
-
-### c. Find the probability of someone being between 61 and 69 inches tall. Why is this number familiar?
-
-### d. Find the probability of someone being between 57 and 73 inches tall. Why is this number familiar?
-
-## Playing with a Data Set
-
-For the following questions, use the \texttt{diamonds} dataset, included as part of `ggplot2`.
-
-### 6. Install `ggplot2`
-
-### 7. Load `ggplot2` with the `library()` command
-
-### 8. Get the structure of the diamonds data.frame. What are the different variables and what kind of data does each contain?
-
-### 9. Get summary statistics for `carat`, `depth`, `table`, and `price`
+### 9. Get summary statistics separately for `carat`, `depth`, `table`, and `price`.
 
@@
 
-### 11. Now rerun the `summary()` command on the entire data frame
+### 11. Now rerun the `summary()` command on the entire data frame.
 
-### 12. Plot a histogram of price.
+### 12. Now look only at (subset) the first 4 diamonds in the dataset.
 
-### 13. Plot a boxplot of price by diamond color.
+### 13. Now look only at (subset) the third and seventh diamond in the dataset.
 
-## Execute your R Script
+### 14. Now look only at (subset) the second column of the dataset.
+
+### 15. Do this again, but look using the `$` to pull up the second column by name.
+
+### 16. Now look only at diamonds that have a `carat` greater than or equal to 1.
+
+### 17. Now look only at diamonds that have a VVS1 `clarity`.
+
+### 17. Now look only at dimaonds that have a color of E, F, I, and J.
+
+### 18. Now look only at diamonds that have a `carat` greater than or equal to 1 and a VVS1 `clarity`.
+
+### 19. Get the average price of diamonds in question 18.^[Hints: use your subset command as an argument to the mean function. You will not need a comma here because you are looking for a single row.]
+
+### 20. What is the highest price for a diamond with a 1.0 carat, D color, and VVS1 clarity?
+
+# Execute your R Script
 

### 2019-09-03 content/reading/03-reading.Rmd
Fix reading and tweak slides for class 3

@@
 
-- <i class="fas fa-external-link-square-alt"></i>@R4DS[Ch. 1] 1^[[https://r4ds.had.co.nz/introduction.html](https://r4ds.had.co.nz/introduction.html)]
+- <i class="fas fa-external-link-square-alt"></i>Ch.1^[http://r4ds.had.co.nz/introduction.html] in @R4DS[Ch. 1]
 
-Now that we start working with R, we will
+Now that we start working with R, you should consider this book to be your primary reference for R-related questions.
+We will broadly cover the first few chapters in order over the next 2-3 class periods.

### 2019-09-03 static/slides/03-slides.Rmd
Fix reading and tweak slides for class 3

@@
       beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
-      highlightStyle: github #ocean rainbow #monokai rainbow solarized-light #tomorrow-night -blue -eighties zenburn
+      highlightStyle: solarized-dark #ocean rainbow #monokai rainbow solarized-light #tomorrow-night -blue -eighties zenburn
       highlightLines: true
@@
 - R itself is just a command language 
-    - The `R` *app* is basically just a command line
-    - You can even just use your computer's command line!<sup>.red[2]</sup>
+- The `R` *app* is basically just a command line, you can even just use your computer's command line!<sup>.red[2]</sup>
 ]
@@
 
-- R Studio has 4 window panes:
+R Studio has 4 window panes:
+
 1. **Source**<sup>.red[1]</sup>: a text editor for documents, R scripts, etc.
@@
 
-- .font130[.onfire[You don't "learn R, you learn how to do things in R]]
+- .font130[.onfire[You don't "*learn R*", you learn *how to do things in R*]]
 
@@
 
-- Today may seem a bit overwhelming
+- .whisper[Today may seem a bit overwhelming]
     - You don't need to know or internalize all of this today
@@
 .center[
+
+From Kieran Healy's excellent (free online!) [book on Data Visualization](http://socviz.co/).]
+
 ![:scale 50%](http://socviz.co/assets/ch-09-read-a-help-page.png)
 
-From Kieran Healy's excellent (free online!) [book on Data Visualization](http://socviz.co/).]
 ]
@@
 
-- Not relevant yet, but once we start writing longer blocks of code, it helps to have a consistent (and human-readable!) style
+- Once we start writing longer blocks of code, it helps to have a consistent (and human-readable!) style
 - I follow [this style guide](http://adv-r.had.co.nz/Style.html) (you are not required to)<sup>.red[1]</sup>
@@
 
-.footnote[<sup>.red[1]</sup> Also described in [today's course notes page](http://metricsf19.classes.ryansafner.com/class/03-class/) and the course [reference page](http://metricsf19.classes.ryansafner.com/reference.
+.footnote[<sup>.red[1]</sup> Also described in [today's course notes page](http://metricsf19.classes.ryansafner.com/class/03-class/) and the course [reference page](http://metricsf19.classes.ryansafner.com/reference).
 
@@
 .pull-left[
-- R is an "**object-oriented**" programming language
+- R is an .shout[**object-oriented**] programming language
 - 99% of the time, you will be:
@@
 
+.pull-right[
 ```{r}
@@
 
-class: inverse, center, middle
-# Resources
-
----
-
 # What's To Come

### 2019-09-03 content/r/03-r-practice.Rmd
Fix typos on r practice
Fixed typos on R Practice 3, update class 3 notes

@@
 
-## 2. Use R's help functions to determine what the `paste()` function does. Then paste together your name from `me`.
+## 2. Use R's help functions to determine what the `paste()` function does. Then paste together your first name and last name.
 
@@
 
-## 4. Find the mean of x with `mean()`.
+## 4. Find the mean of `my_vector` with `mean()`.
 
-## 5. Add up all the numbers from 18 to 763, then get the mean
+## 5. Take all the integers from 18 to 763,^[Hint: use the `:` operator to create a sequence from a starting number to an ending number] then get the mean.
 
@@
 
-For the following questions, we will the `diamonds` dataset, included as part of `ggplot2`.
+For the following questions, we will use the `diamonds` dataset, included as part of `ggplot2`.
 

### 2019-09-03 static/r/03-r-practice.Rmd
Fix typos on r practice
Fixed typos on R Practice 3, update class 3 notes

@@
 
-## 1.
+## 1. Work on the following parts:
 
 a. Create a vector called `me` with two objects, your first name, and your last name.
+
 b. Call the vector to inspect it.
+
 c. Confirm it is a character class vector.
 
-## 2. Use R's help functions to determine what the `paste()` function does. Then paste together your name from `me`.
+## 2. Use R's help functions to determine what the `paste()` function does. Then paste together your first name and last name.
 

### 2019-09-05 content/class/04-class.Rmd
Add all content for class 1.4

--- 2019-09-03 content/class/03-class.Rmd
@@
 
-Today we begin the long slog to your mastery of `R`.
+We may spend at maximum 10 minutes at the beginning of class continuing to work on last class' R practice questions.
+I will put up the answer key after class.
+Today we start the fun stuff - data visualization.
+We will cover how to build plots with the package `ggplot2` (part of the `tidyverse`): I will lecture for the first half (again, it will be a lot of information that you can refer back to as needed over the semester) and you will practice making plots.
 
@@
 
-- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/02-slides.html)
+- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/04-slides.html)
 
-# Appendix
+# Practice Problems
 
-## Installing R and R Studio
+Today you will be working on [R practice problems](/r/04-r-practice.html).
 
-1.
-[Install R](http://www.r-project.org/) from CRAN^[The Comprehensive R Archive Network] by clicking "Download R" (or the CRAN link under Downloads on the left menu).
-This will take you to a mirrors page, where you can select a location in the U.S.
-and download a copy of R
-2. [Install R Studio (Desktop Version)](https://www.rstudio.com/products/rstudio/download/), choose the "Free" option
+# Statistics and Software Survey
 
-(This will also be posted on the [Reference](/reference) page.)
+I will post "answers" after both sections have turned in their surveys.
 
-## R Packages
-
-Packages come from multiple sources.
-
-The polished, publicly released versions are found on [CRAN](https://cran.r-project.org/).
-When installing a package available on CRAN, it is sufficient simply to tell R the following:^[Note the plural `s` on `packages`, and the quotes around the "package name"]
-
-```{r, eval=F}
-install.packages("packagename") 
-```
-
-Other packages, which may be in various developmental states (including perfectly functional!) are often hosted on [GitHub](http://github.com) before they make their way onto CRAN.
-Simply telling R `install.packages("packagename")` will fail to find it (as R only looks in CRAN for packages), so you must use another package called `devtools`^[Which you will need to install first if you (probably) don't already have it!] to install packages directly from Github:^[Note the `::` allows you to use the function `install_github()` from the `devtools` package without having to first load the devtools package with `library(devtools)`.]
-
-```{r, eval=F}
-devtools::install_github("username/packagename") 
-```
-
-For example, to install Hadley Wickham's package `r4ds` from its Github page [https://github.com/hadley/r4ds](https://github.com/hadley/r4ds), we would type:
-
-```{r, eval=F}
-devtools::install_github("hadley/r4ds")
-```
-
-To **use** a package, you need to ensure it is loaded to your workspace (you only need to do this once)^[When we learn how to write `R Markdown` documents, .
-with `library("package_name")`.^[Quotes are not necessary this time.]
-
-## Getting Help for R
-
-For specific functions or commands, you can simply type:
-
-```{r, eval=F}
-?functionname()
-
-# example
-?mean()
-```
-
-This will display a help page specific to that function in the Viewer pane.
-R functions and packages are extremely well-documented; help pages [normally ](http://socviz.co/assets/ch-09-read-a-help-page.png)^[This useful guide comes from Kieran Healy's excellent (free online!) [book on Data Visualization](http://socviz.co/).] a short description of the function, arguments and options (as well as their default values), and several examples or vignettes to demonstrate usage of the function.
-
-Additionally, you can turn to the community.
-
-## Other Useful Commands to Know
-
-One of the best/worst things about `R` is that it is a language, meaning there are multiple ways that you can accomplish the same task.
-
-### Creating Vectors
-
-We know vectors can be created with the `c()` command (and stored with `<-`), but there are other shortcuts to combine objects into a vector, particularly numeric data:
-
-- `:` creates a series of integers
-
-```{r}
-1:5 # create a vector of 1 through 5
-
-12:17 # create a vector of 12 through 17
-```
-
-- `seq(from = , to = , by = )` creates a numeric sequence, and is not restricted to integers
-
-```{r}
-seq(from = 1, to = 10, by = 2) # sequence from 1 to 10, by 2s
-
-# note you do not need to fully write out the name of each argument, just the input!
-
-seq(32.5,40,1.5) # sequence from 32.5 to 40, by 1.5
-```
-
-- `rep(., times =)`^[`.` is a placeholder here.] repeats an element a specified number of times
-
-```{r}
-rep(2, times = 4) # repeat "2" four times
-
-rep(2, 4) # does the same thing
-
-# the thing repeated could itself be a vector
-rep(c(1,4,7), 3) # repeat the vector "1, 4, 7" three times
-```
-
-- We can combine these:
-
-```{r}
-
-# combine (the sequence of 4 to 8 by 2's repeated three times) and 1 and 5
-
-c(rep(seq(4,8,2),3),1,5)
-```
-
-## Style Guide for Coding
-
-We want to maximize *human*-readability of code, not just machine-readability.
-I try to follow [Hadley Wickham's style guide](http://adv-r.had.co.nz/Style.html) for all of my code, including that in this class.
-
-*You will not be graded on the style of your code.* But now's the best time to learn best practices (while you don't know any alternatives!) to save yourself and your potential colleagues (including your future self) from unnecessary frustration.
+# Additional Useful Information

### 2019-09-05 content/r/04-r-practice.Rmd
Add all content for class 1.4

---
title: "1.4 Data Visualization with ggplot2 - R Practice"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```
- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/r/04-r-practice.pdf)

# Getting Set Up

Before we begin, start a new file with `File` $\rightarrow$ `New File` $\rightarrow$ `R Script`. As you work through this sheet in the console in `R`, also add (copy/paste) your commands that work into this new file. At the end, save it, and run to execute all of your commands at once.

# Exploring the Data

## 1. We will look at GDP per Capita and Life Expectancy using some data from the [gapminder project](http://gapminder.org). There is a [handy package](https://github.com/jennybc/gapminder) called `gapminder` that uses a small snippet of this data for exploratory analysis. Install and load the package `gapminder`. Type `?gapminder` and hit enter to see a description of the data.

## 2. Let's get a quick look at `gapminder` to see what we're dealing with.

a. Get the `str`ucture of the `gapminder` data. 
b. What variables are there?
c. Look at the `head` of the dataset to get an idea of what the data looks like. 
d. Get `summary` statistics of all variables.

# Simple Plots in Base R

## 3. Let's make sure you can do some basic plots before we get into the `gg`. Use base `R`'s `hist()` function to plot a *histogram* of `gdpPercap`.

## 4. Use base `R`'s `boxplot()` function to plot a *boxplot* of `gdpPercap`.

## 5. Now make it a *boxplot* by `continent`.^[Hint: use formula notation with`~`.]

## 6. Now make a *scatterplot* of `gdpPercap` on the $x$-axis and `LifeExp` on the $y$-axis.

# Plots with `ggplot2`

## 7. Load the package `ggplot2` (you should have installed it previously. If not, install first with `install.packages("ggplot2")`).

## 8. Let's first make a `bar` graph to see how many countries are in each continent. The only `aes`thetic you need is to map `continent` to `x`. Bar graphs are great for representing categories, but not quantitative data.

## 9. For quantitative data, we want a `histogram` to visualize the distribution of a variable. Make a `histogram` of `gdpPercap`. Your only `aes`thetic here is to map `gdpPercap` to `x`.

## 10. Now let's try adding some color, specifically, add an `aes`thetic that maps `continent` to `color.`

## 11. Instead of a `histogram`, change the `geom` to make it a `density` graph. To avoid overplotting, add `alpha=0.4` to the `geom` argument (alpha changes the *transparency* of a `fill`).

## 12. Redo your plot from 10 for `lifeExp` instead of `gdpPercap`.

## 13. Now let's try a scatterplot for `lifeExp` (as `y`) on `gdpPercap` (as `x`). You'll need both for `aes`thetics. The `geom` here is `geom_point()`.

## 14. Add some color by mapping `continent` to `color` in your `aes`thetics.

## 15. Now let's try adding a regression line with `geom_smooth()`. Add this layer on top of your `geom_point()` layer.

## 16. Did you notice that you got multiple regression lines (colored by continent)? That's because we set a `global` `aes`thetic of mapping `continent` to `color`. If we want just *one* regression line, we need to instead move the `color = continent` inside the `aes` of `geom_point`. This will only map `continent` to `color` for points, not for anything else.

## 17. Now add an `aes`thetic to your `point`s to map `pop` to `size`.

## 18. Change the color of the regression line to `"black"`. Try first by putting this inside an `aes()` in your `geom_smooth`, and try a second time by just putting it inside `geom_smooth` without an `aes()`. What's the difference, and why?

## 19. Another way to separate out continents is with `facet`ing. Add `+facet_wrap(~continent)` to create subplots by `continent`.

## 20. Remove the `facet` layer. The `scale` is quite annoying for the `x`-axis, a lot of points are clustered on the lower level. Let's try changing the scale by adding a layer: `+scale_x_log10()`.

## 21. Now let's fix the labels by adding `+labs()`. Inside `labs`, make proper axes titles for `x`, `y`, and a `title` to the plot. If you want to change the name of the legends (continent color), add one for `color` and `size`.

## 22. Now let's try subsetting by looking only at North America. Take the `gapminder` dataframe and subset it to only look at `continent=="Americas"`). Assign this to a new dataframe object (call it something like `america`.) Now, use *this* as your `data`, and redo the graph from question 16. (You might want to take a look at your new dataframe to make sure it worked first!)

## 23. Try this again for the *whole* world, but just for observations in the year 2000.

### 2019-09-05 static/r/03-r-practice-answers.Rmd
Add all content for class 1.4

---
title: "1.3 Meet R - R Practice (Answers)"
author: "Ryan Safner"
date: "ECON 480 - Fall 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting Set Up

Before we begin, start a new file with `File` $\rightarrow$ `New File` $\rightarrow$ `R Script`. As you work through this sheet in the console in `R`, also add (copy/paste) your commands that work into this new file. At the end, save it, and run to execute all of your commands at once.

# Creating Objects

## 1. Work on the following parts:

a. Create a vector called `me` with two objects, your first name, and your last name.

```{r}
me <- c("Ryan", "Safner")
```

b. Call the vector to inspect it.

```{r}
me
```

c. Confirm it is a character class vector.

```{r}
class(me)
```

## 2. Use R's help functions to determine what the `paste()` function does. Then paste together your first name and last name.

```{r}
?paste() # or help(paste)
 # paste is a function that combines (concatenates) multiple string objects into a single object
paste("Ryan", "Safner")

# note you can choose how to separate string objects with the "sep" argument
# for example
paste("Ryan", "Safner", sep="") # no separation
paste("Ryan", "Safner", sep=" ") # separate with a space " " (the default)
paste("Ryan", "Safner", sep="_") # separate with underscore
```
## 3. Create a vector called `my_vector` with all the even integers from 2 to 10.

```{r}
my_vector <- c(2,4,6,8,10)

# verify it worked
my_vector
# alternatively, you can use the sequence function, seq()
# see the Class page for more about this function
my_vector <- seq(from = 2, # starting integer
                 to = 10, # ending integer
                 by = 2) # by 2's

# you can shorten it by not including the names of the arguments:
my_vector <- seq(2,10,2)

# verify it worked
my_vector
```

## 4. Find the mean of `my_vector` with `mean()`.

```{r}
mean(my_vector)
```

## 5. Take all the integers from 18 to 763,^[Hint: use the `:` operator to create a sequence from a starting number to an ending number] then get the mean.

```{r}

# create a sequence of integers by 1 with starting_number:ending_number
# see Class 3 page for more

# you can do this all at once without making an object
mean(18:763)

# alternatively you can save this as a vector and run the mean on it
vec1 <- 18:763

mean(vec1)
```

\clearpage

# Playing with Data

For the following questions, we will use the `diamonds` dataset, included as part of `ggplot2`.

### 6. Install `ggplot2`.

```{r, eval=F}
install.packages("ggplot2") # note the s and the quotes
```

### 7. Load `ggplot2` with the `library()` command.

```{r}
library("ggplot2") # quotes not necessary, but can be used
```

### 8. Get the `str`ucture of the diamonds data frame. What are the different variables and what kind of data does each contain?

```{r}
str(diamonds)

# We have
# - carat: a number
# - cut: an ordered factor
# - color: an ordered factor
# - clarity: an ordered factor
# - depth: a number
# - table: a number
# - price: an integer
# - x: a number
# - y: a number
# - z: a number
```

### 9. Get summary statistics separately for `carat`, `depth`, `table`, and `price`.

```{r}
summary(diamonds$carat)
summary(diamonds$depth)
summary(diamonds$table)
summary(diamonds$price)
```

### 10. `color`, `cut`, and `clarity` are categorical variables (factors). Use the `table()` command to generate frequency tables for each.

```{r}
table(diamonds$color)
table(diamonds$cut)
table(diamonds$clarity)
```

### 11. Now rerun the `summary()` command on the entire data frame.

```{r}
summary(diamonds)
```

### 12. Now look only at (subset) the first 4 diamonds in the dataset.

```{r}
# remember, dataframes are indexed by: df[row#s, column#s]
diamonds[1:4,] # select first through fourth rows, all columns

# alternatively
diamonds[c(1,2,3,4),] # using a vector-approach
```

### 13. Now look only at (subset) the third and seventh diamond in the dataset.

```{r}
diamonds[c(3,7),] # select 3rd and 7th row, all columns
```

### 14. Now look only at (subset) the second column of the dataset.

```{r}
diamonds[,2] # select all rows, 2nd column
```
### 15. Do this again, but look using the `$` to pull up the second column by name.

```{r}
# second column is called "cut"
diamonds$cut
```

### 16. Now look only at diamonds that have a `carat` greater than or equal to 1.

```{r}
# use the [square brackets] to subset, 
# first argument (rows) are chosen by conditional: 
# - choose diamonds based on their carat, and only carats >= 1
diamonds[diamonds$carat >= 1,] # select rows on condition, and all columns
```

### 17. Now look only at diamonds that have a VVS1 `clarity`.

```{r}
# we are testing for equality, so we need two ==
# we are selecting based on clarity, a character/factor, so we need quotes
diamonds[diamonds$clarity=="VVS1",] # select rows on condition, and all columns
```

### 18. Now look only at dimaonds that have a color of E, F, I, and J.

```{r}
# same idea as last problem, except now we want one of any of these 4 colors

# first (tedious) way, a series of checking equality and using "OR"s (|) 
diamonds[diamonds$color=="E" | diamonds$color=="F" | diamonds$color=="I" | diamonds$color=="J",] # select rows on condition, and all columns

# second (better) way, using group membership operator (%in%) and list the elements as a vector
diamonds[diamonds$color %in% c("E","F","I","J"),] # select rows on condition, and all columns 
```

### 19. Now look only at diamonds that have a `carat` greater than or equal to 1 and a VVS1 `clarity`.

```{r}
# testing for two conditions (AND)
diamonds[diamonds$carat>=1 & diamonds$clarity=="VVS1",] # select rows on condition, and all columns 
```

### 20. Get the average price of diamonds in question 18.^[Hints: use your subset command as an argument to the mean function. You will not need a comma here because you are looking for a single row.]

```{r}
# use command from last question as the argument to the mean function, 
## but be sure that you look at the price, specifically

mean(diamonds$price[diamonds$carat>=1 & diamonds$color=="D" & diamonds$clarity=="VVS1"])
```

### 21. What is the highest price for a diamond with a 1.0 carat, D color, and VVS1 clarity?

```{r}
max(diamonds$price[diamonds$carat>=1 & diamonds$color=="D" & diamonds$clarity=="VVS1"])
```

# Execute your R Script

Save the R Script you created at the beginning and (hopefully) have been pasting all of your valid commands to. This creates a *.R* file wherever you choose to save it to. Now looking at the file in the upper left pane of *R Studio* look for the button in the upper right corner that says **Run**. Sit back and watch `R` redo everything you've carefully worked on, all at once.

Your `.R` file should look something like this:

```{r, eval=F}

# 1 -------------------

## a 
me <- c("Ryan", "Safner")

## b
me

## c 
class(me)

# 2 ----------------

?paste() # or help(paste)
 # paste is a function that combines (concatenates) multiple string objects into a single object
paste("Ryan", "Safner")

# note you can choose how to separate string objects with the "sep" argument
# for example
paste("Ryan", "Safner", sep="") # no separation
paste("Ryan", "Safner", sep=" ") # separate with a space " " (the default)
paste("Ryan", "Safner", sep="_") # separate with underscore

# 3 ---------------

my_vector <- c(2,4,6,8,10)

# verify it worked
my_vector
# alternatively, you can use the sequence function, seq()
# see the Class page for more about this function
my_vector <- seq(from = 2, # starting integer
                 to = 10, # ending integer
                 by = 2) # by 2's

# you can shorten it by not including the names of the arguments:
my_vector <- seq(2,10,2)

# verify it worked
my_vector

# 4 -------------------

mean(my_vector)

# 5 -------------------

# create a sequence of integers by 1 with starting_number:ending_number
# see Class 3 page for more

# you can do this all at once without making an object
mean(18:763)

# alternatively you can save this as a vector and run the mean on it
vec1 <- 18:763

mean(vec1)

# 6 ------------------

install.packages("ggplot2") # note the s and the quotes

# 7 ------------------

library("ggplot2") # quotes not necessary, but can be used

# 8 ------------------

str(diamonds)

# We have
# - carat: a number
# - cut: an ordered factor
# - color: an ordered factor
# - clarity: an ordered factor
# - depth: a number
# - table: a number
# - price: an integer
# - x: a number
# - y: a number
# - z: a number

# 9 ------------------

summary(diamonds$carat)
summary(diamonds$depth)
summary(diamonds$table)
summary(diamonds$price)

# 10 ------------------

table(diamonds$color)
table(diamonds$cut)
table(diamonds$clarity)

# 11 ------------------

summary(diamonds)

# 12 ------------------

# remember, dataframes are indexed by: df[row#s, column#s]
diamonds[1:4,] # select first through fourth rows, all columns

# alternatively
diamonds[c(1,2,3,4),] # using a vector-approach

# 13 ------------------

diamonds[c(3,7),] # select 3rd and 7th row, all columns

# 14 ------------------
diamonds[,2] # select all rows, 2nd column

# 15 ------------------

# second column is called "cut"
diamonds$cut

# 16 -------------------

# use the [square brackets] to subset, 
# first argument (rows) are chosen by conditional: 
# - choose diamonds based on their carat, and only carats >= 1
diamonds[diamonds$carat >= 1,] # select rows on condition, and all columns

# 17 -------------------

# we are testing for equality, so we need two ==
# we are selecting based on clarity, a character/factor, so we need quotes
diamonds[diamonds$clarity=="VVS1",] # select rows on condition, and all columns

# 18 -------------------

# same idea as last problem, except now we want one of any of these 4 colors

# first (tedious) way, a series of checking equality and using "OR"s (|) 
diamonds[diamonds$color=="E" | diamonds$color=="F" | diamonds$color=="I" | diamonds$color=="J",] # select rows on condition, and all columns

# second (better) way, using group membership operator (%in%) and list the elements as a vector
diamonds[diamonds$color %in% c("E","F","I","J"),] # select rows on condition, and all columns

# 19 -------------------

# testing for two conditions (AND)
diamonds[diamonds$carat>=1 & diamonds$clarity=="VVS1",] # select rows on condition, and all columns

# 20 -------------------

# use command from last question as the argument to the mean function
## but be sure that you look at the price, specifically
mean(diamonds$price[diamonds$carat>=1 & diamonds$color=="D" & diamonds$clarity=="VVS1"])

# 21 -------------------

max(diamonds$price[diamonds$carat>=1 & diamonds$color=="D" & diamonds$clarity=="VVS1"])
```

### 2019-09-05 static/r/04-r-practice.Rmd
Add all content for class 1.4

--- 2019-09-05 content/r/04-r-practice.Rmd
@@
 title: "1.4 Data Visualization with ggplot2 - R Practice"
-editor_options: 
-  chunk_output_type: console
+author: "Ryan Safner"
+date: "ECON 480 - Fall 2019"
+output: pdf_document
 ---
 
-```{r message=FALSE, warning=FALSE, include=FALSE}
-library(tidyverse)
-library(pander)
-source(file.path(here::here(), "lib", "graphics.R"))
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
 ```
-- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/r/04-r-practice.pdf)
 

### 2019-09-05 static/slides/04-slides.Rmd
Add all content for class 1.4

---
title: "1.4: Data Visualization with ggplot2"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: solarized-dark #ocean rainbow #monokai rainbow solarized-light #tomorrow-night -blue -eighties zenburn
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(message=F,
                      warning=F)
library(tidyverse)

```

# Graphics and Statistics

- Admittedly, we still need to cover basic descriptive statistics and data fundamentals
    - continuous, discrete, cross-sectional, time series, panel data
    - mean, median, variance, standard deviation
    - random variables, distributions, PDFs, Z-scores
    - bargraphs, boxplots, histograms, scatterplots

- All of this is coming in 2 weeks as we return to statistics and econometric theory

- But let's start with the fun stuff right away, even if you don't fully know the *reasons*: .shout[data visualiation]

---

class: inverse, center, middle

# Plotting in Base R

---

# Plotting in Base R

- Base `R` is very powerful and intuitive to plot, but not very sexy

- Basic syntax for most types of plots:

```{r, eval=F}
plot_type(my_df$variable)
```

- If using multiple variables, you can avoid typing `$` by just typing the variable names and then in another argument to the plotting function, specify `data = my_df`

```{r, eval=F}
plot_type(my_df$variable1, my_df$variable2, data = my_df)
```

---

# Plotting in Base R: Histogram

.left-code[
- Using the built-in `mtcars` data, plotting a histogram of `mpg`

```{r base-hist, eval=F}
hist(mtcars$mpg)
```

]

.right-plot[
.center[
```{r, ref.label="base-hist", echo=F, fig.retina=3}
```
]
]

---

# Plotting in Base R: Boxplot

.left-code[
- Using the built-in `mtcars` data, plotting a boxplot of `mpg`

```{r base-boxplot, eval=F}
boxplot(mtcars$mpg)
```

]

.right-plot[
.center[
```{r, ref.label="base-boxplot", echo=F, fig.retina=3}
```
]
]

---

# Plotting in Base R: Boxplot by Category

.left-code[
- Using the built-in `mtcars` data, plotting a boxplot of `mpg` by `cyl`

```{r base-boxplot2, eval=F}
boxplot(mtcars$mpg~mtcars$cyl)
```

```{r, eval=F}
# second method
boxplot(mpg~cyl, data = mtcars)
```

- The `~` is part of `R`'s **"formula" notation**: 
    - Dependent variable goes to left
    - Independent variable(s) to right, separated with `+`'s
    - Think `y~x+z` means "`y` is explained by `x` and `z`"

]

.right-plot[
.center[
```{r, ref.label="base-boxplot2", echo=F, fig.retina=3}
```
]
]

---

# Plotting in Base R: Scatterplot

.left-code[
- Using the built-in `mtcars` data, plotting a scatterplot of `mpg` against `disp`

```{r base-scatter, eval=F}
plot(mtcars$mpg~mtcars$disp)
```

```{r, eval=F}
# second method
plot(mpg~disp, data = mtcars)
```

]

.right-plot[
.center[
```{r, ref.label="base-scatter", echo=F, fig.retina=3}
```
]
]

---

class: inverse, center, middle

# ggplot2 and the tidyverse

---

class: blank

background-image: url(https://www.dropbox.com/s/np8v4ihjotkckc1/tidyverse1.png?raw=1)
background-size: cover

---

# The tidyverse

> "The tidyverse is an opinionated collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures.

- Largely (but not only) created by Hadley Wickham

- We will look at this much more extensively next week!

- This "flavor" of `R` will make your coding life *so much easier!*

![](https://www.dropbox.com/s/81ijxswbwqor0up/tidyverse.png?raw=1)

---

# ggplot

.pull-left[

- `ggplot2` is perhaps the most popular package in `R` and a core element of the `tidyverse`

- `gg` stands for a .whisper[grammar of graphics]

- Very powerful and beautiful graphics, very customizable and reproducible, but requires a bit of a learning curve

- All those "cool graphics" you've seen in the New York Times, fivethirtyeight, the Economist, Vox, etc use the grammar of graphics
]

.pull-right[
.center[
![:scale 65%](https://www.dropbox.com/s/4bxo0ac3gwwgd9y/hadleyplot2.png?raw=1)
]
]

---

# ggplot: All Your Figure are Belong to Us

.pull-left[
.center[
![:scale 80%](https://www.dropbox.com/s/4ij96m7h66i9o4p/538gg1.png?raw=1)

Source: [fivethirtyeight](https://fivethirtyeight.com/features/fastest-airlines-fastest-airports/)
]
]
.pull-left[
.center[
![:scale 80%](https://www.dropbox.com/s/2bkklsrsl0ep902/538gg.png?raw=1)

Source: [fivethirtyeight](https://fivethirtyeight.com/features/how-to-tell-someones-age-when-all-you-know-is-her-name/)
]
]

---

# Why Go gg?

.left-column[

.center[
![:scale 80%](https://www.dropbox.com/s/u34dakjyq4hkyov/hadley.jpg?raw=1)

Hadley Wickham

Chief Scientist, R Studio
]
]

.right-column[

> "The transferrable skills from ggplot2 are not the idiosyncracies of plotting syntax, but a powerful way of thinking about visualisation, as a way of **mapping between variables and the visual properties of geometric objects** that you can perceive."

]

.source[[http://disq.us/p/sv640d](http://disq.us/p/sv640d)]

---

# The Grammar of Graphics (gg) I

.pull-left[

- Any graphic can be built from the same components:
    1. **Data** to be drawn from
    2. **Aesthetic mappings** from data to some visual marking
    3. **Geometric objects** that appear on the plot
    4. **Coordinates** to organize location of objects
    5. **Scales** define the range of values
    6. **Facets** group into subplots

- Not every plot needs *every* component, but all plots *must* have the first 3!
]

.pull-right[
.center[
![](https://www.dropbox.com/s/flw1dbvhssn1q23/ggplotlayers.png?raw=1)
]
]

---

# The Grammar of Graphics (gg) II

.pull-left[

- Any graphic can be built from the same components:
    1. `data` to be drawn from
    2. `aes`**thetic mappings** from data to some visual marking
    3. `geom`**etric objects** that appear on the plot
    4. `coord`**inates** to scale axes
    5. `scale`**s** define the range of values
    6. `facet`**s** group into subplots

- Not every plot needs *every* component, but all plots *must* have the first 3!
]

.pull-right[
.center[
![](https://www.dropbox.com/s/flw1dbvhssn1q23/ggplotlayers.png?raw=1)
]
]

---

# Example Data

- We'll use the `mpg` example dataset that is built-in with `ggplot2`

```{r}
head(mpg, n=10)
```

---

# The Grammar of Graphics (gg): Data

.left-code[
### Data
```{r, eval=F}
ggplot(data = mpg)
```

]

.right-plot[

**Data** is the source of our data. As part of the `tidyverse`, `ggplot2` requires data to be .whisper["tidy"]<sup>.red[1]</sup>:

1. Each variable forms a .hl[column]

2. Each observation forms a .hl[row]

3. Each observational unit forms a table

]

.footnote[<sup>.red[1]</sup> Data "tidyness" is the core element of all `tidyverse` packages. Much more on all of this next class.]

---

# The Grammar of Graphics (gg): Adding Layers

.left-code[
### Data
]

.right-plot[
- Add a layer with `+` at the end of a line (never at the beginning!)

- Style recommendation: start a new line after each `+` to improve legibility!

- We will build a plot layer-by-layer
]

---

# The Grammar of Graphics (gg): Aesthetics I

.left-code[
### Data
### Aesthetics
```{r, eval=F}
+ aes()
```
]

.right-plot[
**Aesthetics** map data to visual elements or parameters

.center[
![](https://www.dropbox.com/s/d226uyjrru1sn36/ggplotaes.png?raw=1)
]
]

---

# The Grammar of Graphics (gg): Aesthetics II

.left-code[
### Data
### Aesthetics
```{r, eval=F}
+ aes()
```
]

.right-plot[
**Aesthetics** map data to visual elements or parameters

- `displ`

- `hwy`

- `class`

]

---

# The Grammar of Graphics (gg): Aesthetics II

.left-code[
### Data
### Aesthetics
```{r, eval=F}
+ aes()
```
]

.right-plot[
**Aesthetics** map data to visual elements or parameters

- `displ` → **x**

- `hwy` → **y**

- `class` → *shape*, *size*, **color**, etc.

]

---

# The Grammar of Graphics (gg): Aesthetics II

.left-code[
### Data
### Aesthetics
```{r, eval=F}
+ aes()
```
]

.right-plot[
**Aesthetics** map data to visual elements or parameters

.center[
![](https://www.dropbox.com/s/e1nvb7x330c7ruk/aes2.png?raw=1)
]
]
---

# The Grammar of Graphics (gg): Aesthetics III

.left-code[
### Data
### Aesthetics
```{r, eval=F}
+ aes()
```
]

.right-plot[
**Aesthetics** map data to visual elements or parameters

.center[
![](https://www.dropbox.com/s/f5msffqgvbs4nkb/aes.png?raw=1)
]

---

# The Grammar of Graphics (gg): Aesthetics IV

.left-code[
### Data
### Aesthetics
```{r, eval=F}
+ aes()
```
]

.right-plot[
**Aesthetics** map data to visual elements or parameters

```{r, eval=F}
aes(x = displ,
    y = hwy,
    color = class)
```
]

---

# The Grammar of Graphics (gg): Geoms I

.left-code[
### Data
### Aesthetics
### Geoms
```{r, eval=F}
+ geom_*()
```
]

.right-plot[
**Geometric objects** displayed on the plot

.center[
```{r geom_demo, echo=FALSE, fig.retina=3, fig.height=6}
minimal_theme <- theme_bw() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.title = element_blank(),
    plot.title = element_text(hjust = 0.5),
    text = element_text(family = "Fira Mono"),
    plot.background = element_rect(fill = "#FAFAFA", color = NA),
    panel.background = element_rect(fill = "#FAFAFA", color = NA)
  )

set.seed(20)
df_geom <- data_frame(y = rnorm(10), x = 1:10)

g_geom <- list()
g_geom$point <- ggplot(df_geom, aes(x, y)) + geom_point() + ggtitle("geom_point()")
g_geom$line <- ggplot(df_geom, aes(x, y)) + geom_line() + ggtitle("geom_line()")
g_geom$bar <- ggplot(df_geom, aes(x, y)) + geom_col() + ggtitle("geom_col()")
g_geom$boxplot <- ggplot(df_geom, aes(y = y)) + geom_boxplot() + ggtitle("geom_boxplot()")
g_geom$histogram <- ggplot(df_geom, aes(y)) + geom_histogram(binwidth = 1) + ggtitle("geom_histogram()")
g_geom$density <- ggplot(df_geom, aes(y)) + geom_density(fill = "grey40", alpha = 0.25) + ggtitle("geom_density()") + xlim(-4, 4)

g_geom <- map(g_geom, ~ . + minimal_theme)

cowplot::plot_grid(plotlist = g_geom)
```
]
]

---

# The Grammar of Graphics (gg): Geoms II

.left-code[
### Data
### Aesthetics
### Geoms
```{r, eval=F}
+ geom_*()
```
]

.right-plot[
**Geometric objects** displayed on the plot

- What `geom`s you should use depends on what you want to show:

| Type | `geom` |
|------|--------|
| Point | `geom_point()` |
| Line  | `geom_line()`, `geom_path()`  |
| Bar   | `geom_bar()`, `geom_col()` |
| Histogram | `geom_histogram()` |
| Regression | `geom_smooth()` |
| Boxplot | `geom_boxplot()` |
| Text | `geom_text()` |
| Density | `geom_density()` |
]

---

# The Grammar of Graphics (gg): Geoms III

.left-code[
### Data
### Aesthetics
### Geoms
```{r, eval=F}
+ geom_*()
```
]

.right-plot[
**Geometric objects** displayed on the plot

```{r, eval=F}
##  [1] "geom_abline"     "geom_area"       "geom_bar"        "geom_bin2d"     
##  [5] "geom_blank"      "geom_boxplot"    "geom_col"        "geom_contour"   
##  [9] "geom_count"      "geom_crossbar"   "geom_curve"      "geom_density"   
## [13] "geom_density_2d" "geom_density2d"  "geom_dotplot"    "geom_errorbar"  
## [17] "geom_errorbarh"  "geom_freqpoly"   "geom_hex"        "geom_histogram" 
## [21] "geom_hline"      "geom_jitter"     "geom_label"      "geom_line"      
## [25] "geom_linerange"  "geom_map"        "geom_path"       "geom_point"     
## [29] "geom_pointrange" "geom_polygon"    "geom_qq"         "geom_qq_line"   
## [33] "geom_quantile"   "geom_raster"     "geom_rect"       "geom_ribbon"    
## [37] "geom_rug"        "geom_segment"    "geom_sf"         "geom_sf_label"  
## [41] "geom_sf_text"    "geom_smooth"     "geom_spoke"      "geom_step"      
## [45] "geom_text"       "geom_tile"       "geom_violin"     "geom_vline"
```

See [http://ggplot2.tidyverse.org/reference](http://ggplot2.tidyverse.org/reference) for many more options

]

---

# The Grammar of Graphics (gg): Geoms IV

.left-code[
### Data
### Aesthetics
### Geoms
```{r, eval=F}
+ geom_*()
```
]

.right-plot[
**Geometric objects** displayed on the plot

Or just start typing `geom_` in R Studio!

.center[
![](https://www.dropbox.com/s/xui9bxccztlqmyz/geoms.gif?raw=1)
]
]

---

# Let's Make a Plot!

.left-code[
```{r plot-1, eval=F, echo=T}
ggplot(data = mpg)
```
]

.right-plot[
```{r, ref.label="plot-1", echo=F, fig.retina=3, fig.align="center"}
```

]

---

# Let's Make a Plot!

.left-code[
```{r plot-2, eval=F, echo=T}
ggplot(data = mpg)+
  aes(x = displ, #<<
        y = hwy) #<<
```
]

.right-plot[
```{r, ref.label="plot-2", echo=F, fig.retina=3, fig.align="center"}
```

]

---

# Let's Make a Plot!

.left-code[
```{r plot-3, eval=F, echo=T}
ggplot(data = mpg)+
  aes(x = displ,
        y = hwy)+
  geom_point() #<<
```
]

.right-plot[
```{r, ref.label="plot-3", echo=F, fig.retina=3, fig.align="center"}
```

]

---

# Let's Make a Plot!

.left-code[
```{r plot-4, eval=F, echo=T}
ggplot(data = mpg)+
  aes(x = displ,
        y = hwy)+
  geom_point(aes(color = class)) #<<
```
]

.right-plot[
```{r, ref.label="plot-4", echo=F, fig.retina=3, fig.align="center"}
```

]

---

# Let's Make a Plot!

.left-code[
```{r plot-5, eval=F, echo=T}
ggplot(data = mpg)+
  aes(x = displ,
        y = hwy)+
  geom_point()+
  geom_smooth() #<<
```
]

.right-plot[
```{r, ref.label="plot-5", echo=F, fig.retina=3, fig.align="center"}
```

]

---

# More Geoms

.left-code[
### Data
### Aesthetics
### Geoms
```{r, eval=F}
+ geom_*()
```
]

.right-plot[
`geom_*(aes, data, stat, position)`

- `data`: geoms can have their own data
    - has to map onto global coordinates

- `aes`: geoms can have their own aesthetics
    - inherits global aesthetics by default
    - different geoms have different available aesthetics
]

---

# Change Our Plot

.left-code[
```{r plot-5a, eval=F, echo=T}
ggplot(data = mpg)+
  aes(x = displ,
        y = hwy)+
  geom_point(aes(color = class))+ #<<
  geom_smooth()
```
]

.right-plot[
```{r, ref.label="plot-5a", echo=F, fig.retina=3, fig.align="center"}
```

]

---

# More Geoms II

.left-code[
### Data
### Aesthetics
### Geoms
```{r, eval=F}
+ geom_*()
```
]

.right-plot[
`geom_*(aes, data, stat, position)`

- `stat`: some geoms statistically transform data
    - `geom_histogram()` uses `stat_bin()` to group observations into bins

- `position`: some adjust location of objects
    - `dodge`, `stack`, `jitter`
]

---

# Let's Change Our Plot

.left-code[
```{r plot-box, eval=F, echo=T}
ggplot(data = mpg)+
  aes(x = class, #<<
      y = hwy)+ #<<
  geom_boxplot() #<<
```
]

.right-plot[
```{r, ref.label="plot-box", echo=F, fig.retina=3, fig.align="center"}
```

]

---

# Let's Change Our Plot

.left-code[
```{r plot-6, eval=F, echo=T}
ggplot(data = mpg)+
  aes(x = class)+ #<<
  geom_bar() #<<
```
]

.right-plot[
```{r, ref.label="plot-6", echo=F, fig.retina=3, fig.align="center"}
```

]

---

# Let's Change Our Plot

.left-code[
```{r plot-7, eval=F, echo=T}
ggplot(data = mpg)+
  aes(x = class,
      fill = drv)+ #<<
  geom_bar() 
```
]

.right-plot[
```{r, ref.label="plot-7", echo=F, fig.retina=3, fig.align="center"}
```

]

---

# Let's Change Our Plot

.left-code[
```{r plot-8, eval=F, echo=T}
ggplot(data = mpg)+
  aes(x = class,
      fill = drv)+ 
  geom_bar(position = "dodge") #<<
```
]

.right-plot[
```{r, ref.label="plot-8", echo=F, fig.retina=3, fig.align="center"}
```

]

---

# Back to the Original

.left-code[
```{r plot-8a, eval=F, echo=T}
ggplot(data = mpg)+
  aes(x = displ,
        y = hwy)+
  geom_point(aes(color = class))+
  geom_smooth() 
```
]

.right-plot[
```{r, ref.label="plot-8a", echo=F, fig.retina=3, fig.align="center"}
```

]

---

# The Grammar of Graphics (gg): Facets I

.left-code[
### Data
### Aesthetics
### Geoms
### Facets
```{r, eval=F}
+ facet_wrap()

+ facet_grid()
```
]

.right-plot[

```{r, fig.retina=3, fig.height=3.5, fig.align="center"}
ggplot(data = mpg)+
  aes(x = displ,
        y = hwy)+
  geom_point(aes(color = class))+
  geom_smooth()+
  facet_wrap(~year) #<<
```
]

---

# The Grammar of Graphics (gg): Facets II

.left-code[
### Data
### Aesthetics
### Geoms
### Facets
```{r, eval=F}
+ facet_wrap()

+ facet_grid()
```
]

.right-plot[

```{r, fig.retina=3, fig.height=3.5, fig.align="center"}
ggplot(data = mpg)+
  aes(x = displ,
        y = hwy)+
  geom_point(aes(color = class))+
  geom_smooth()+
  facet_grid(cyl~year) #<<
```
]

---

# The Grammar of Graphics (gg): Labels

.left-code[
### Data
### Aesthetics
### Geoms
### Facets
### Labels
```{r, eval=F}
+ labs()
```
]

.right-plot[

```{r, fig.retina=3, fig.align="center"}
ggplot(data = mpg)+
  aes(x = displ,
        y = hwy)+
  geom_point(aes(color = class))+
  geom_smooth()+
  facet_wrap(~year)+
  labs(x = "Engine Displacement (Liters)",
       y = "Highway MPG",
       title = "Car Mileage and Displacement",
       subtitle = "More Displacement Lowers Highway MPG",
       caption = "Source: EPA",
       color = "Vehicle Class")#<<
```
]

---

# The Grammar of Graphics (gg): Scales

.left-code[
### Data
### Aesthetics
### Geoms
### Facets
### Labels
### Scales
```{r, eval=F}
+ scale_*_*()
```
]

.right-plot[

`scale`+`_`+`<aes>`+`_`+`<type>`+`()`

- `<aes>`: parameter you want to adjust
- `<type`: type of parameter

- I want to change my discrete x-axis: `scale_x_discrete()`
- I want to change my continuous y-axis: `scale_y_continuous()`
- I want to rescale x-axis to log: `scale_x_log10()`
- I want to use a different color palette: `scale_fill_discrete()`, `scale_color_manual()`
]
---

# The Grammar of Graphics (gg): Scales

.left-code[

```{r scales, eval=F}
ggplot(data = mpg)+
  aes(x = displ,
        y = hwy)+
  geom_point(aes(color = class))+
  geom_smooth()+
  facet_wrap(~year)+
  labs(x = "Engine Displacement (Liters)",
       y = "Highway MPG",
       title = "Car Mileage and Displacement",
       subtitle = "More Displacement Lowers Highway MPG",
       caption = "Source: EPA",
       color = "Vehicle Class")+
  scale_color_viridis_d()#<<
```
]

.right-plot[

```{r, echo=F, ref.label="scales", fig.retina=3, fig.align="center"}

```
]

---

# The Grammar of Graphics (gg): Themes

.left-code[
### Data
### Aesthetics
### Geoms
### Facets
### Labels
### Scales
### Theme
```{r, eval=F}
+ theme_*()
```
]

.right-plot[

**Theme** changes appearance of plot decorations (things not mapped to data)

- Some themes that come with `ggplot2`:

- `+ theme_bw()`
- `+ theme_dark()`
- `+ theme_gray()`
- `+ theme_minimal()`
- `+ theme_light()`
- `+ theme_classic()`
]
---

# The Grammar of Graphics (gg): Themes

.left-code[
### Data
### Aesthetics
### Geoms
### Facets
### Labels
### Scales
### Theme
```{r, eval=F}
+ theme_*()
```
]

.right-plot[

**Theme** changes appearance of plot decorations (things not mapped to data)

- Many parameters we could change

- Global options: `line`, `rect`, `text`, `title`
- `axis`: x-, y-, or other axis title, ticks, lines
- `legend`: plot legends for fill or color
- `panel`: actual plot area
- `plot`: whole image
- `strip`: facet labels
]

---

# The Grammar of Graphics (gg): Themes

.left-code[

```{r theme1, eval=F}
ggplot(data = mpg)+
  aes(x = displ,
        y = hwy)+
  geom_point(aes(color = class))+
  geom_smooth()+
  facet_wrap(~year)+
  labs(x = "Engine Displacement (Liters)",
       y = "Highway MPG",
       title = "Car Mileage and Displacement",
       subtitle = "More Displacement Lowers Highway MPG",
       caption = "Source: EPA",
       color = "Vehicle Class")+
  scale_color_viridis_d()+
  theme_bw()#<<
```
]

.right-plot[

```{r, echo=F, ref.label="theme1", fig.retina=3, fig.align="center"}

```
]

---

# The Grammar of Graphics (gg): Themes II

.left-code[

```{r theme2, eval=F}
ggplot(data = mpg)+
  aes(x = displ,
        y = hwy)+
  geom_point(aes(color = class))+
  geom_smooth()+
  facet_wrap(~year)+
  labs(x = "Engine Displacement (Liters)",
       y = "Highway MPG",
       title = "Car Mileage and Displacement",
       subtitle = "More Displacement Lowers Highway MPG",
       caption = "Source: EPA",
       color = "Vehicle Class")+
  scale_color_viridis_d()+
  theme_minimal()#<<
```
]

.right-plot[

```{r, echo=F, ref.label="theme2", fig.retina=3, fig.align="center"}

```
]

---

# The Grammar of Graphics (gg): Themes III

.left-code[

```{r theme3, eval=F}
ggplot(data = mpg)+
  aes(x = displ,
        y = hwy)+
  geom_point(aes(color = class))+
  geom_smooth()+
  facet_wrap(~year)+
  labs(x = "Engine Displacement (Liters)",
       y = "Highway MPG",
       title = "Car Mileage and Displacement",
       subtitle = "More Displacement Lowers Highway MPG",
       caption = "Source: EPA",
       color = "Vehicle Class")+
  scale_color_viridis_d()+
  theme_minimal()+
  theme(text = element_text(family = "Fira Sans"))#<<
```
]

.right-plot[

```{r, echo=F, ref.label="theme3", fig.retina=3, fig.align="center"}

```
]

---

# The Grammar of Graphics (gg): Themes III

.left-code[

```{r theme4, eval=F}
ggplot(data = mpg)+
  aes(x = displ,
        y = hwy)+
  geom_point(aes(color = class))+
  geom_smooth()+
  facet_wrap(~year)+
  labs(x = "Engine Displacement (Liters)",
       y = "Highway MPG",
       title = "Car Mileage and Displacement",
       subtitle = "More Displacement Lowers Highway MPG",
       caption = "Source: EPA",
       color = "Vehicle Class")+
  scale_color_viridis_d()+
  theme_minimal()+
  theme(text = element_text(family = "Fira Sans"),
        legend.position="bottom") #<<
```
]

.right-plot[

```{r, echo=F, ref.label="theme4", fig.retina=3, fig.align="center"}

```
]

---

# The Grammar of Graphics (gg): Themes (ggthemes)

.left-code[
### Data
### Aesthetics
### Geoms
### Facets
### Labels
### Scales
### Theme
```{r, eval=F}
+ theme_*()
```
]

.right-plot[

- `ggthemes` package adds some other nice themes

```{r}
# install if you don't have it
# install.packages("ggthemes")
library("ggthemes") # load package
```

]

---

# The Grammar of Graphics (gg): Themes IV

.left-code[

```{r theme5, eval=F}
library("ggthemes") #<<
ggplot(data = mpg)+
  aes(x = displ,
        y = hwy)+
  geom_point(aes(color = class))+
  geom_smooth()+
  facet_wrap(~year)+
  labs(x = "Engine Displacement (Liters)",
       y = "Highway MPG",
       title = "Car Mileage and Displacement",
       subtitle = "More Displacement Lowers Highway MPG",
       caption = "Source: EPA",
       color = "Vehicle Class")+
  scale_color_viridis_d()+
  theme_economist()+ #<<
  theme(text = element_text(family = "Fira Sans"),
        legend.position="bottom") 
```
]

.right-plot[

```{r, echo=F, ref.label="theme5", fig.retina=3, fig.align="center"}

```
]

---

# The Grammar of Graphics (gg): Themes V

.left-code[

```{r theme6, eval=F}
library("ggthemes") #<<
ggplot(data = mpg)+
  aes(x = displ,
        y = hwy)+
  geom_point(aes(color = class))+
  geom_smooth()+
  facet_wrap(~year)+
  labs(x = "Engine Displacement (Liters)",
       y = "Highway MPG",
       title = "Car Mileage and Displacement",
       subtitle = "More Displacement Lowers Highway MPG",
       caption = "Source: EPA",
       color = "Vehicle Class")+
  scale_color_viridis_d()+
  theme_fivethirtyeight()+ #<<
  theme(text = element_text(family = "Fira Sans"),
        legend.position="bottom") 
```
]

.right-plot[

```{r, echo=F, ref.label="theme6", fig.retina=3, fig.align="center"}

```
]

---

class: inverse, center, middle

# Some Troubleshooting

---

# Global vs. Local Aesthetics

- `aes()` can go in base (`data`) layer and/or in individual `geom()` layers
- All `geoms` will inherit global `aes` from `data` layer unless overridden

.pull-left[

```{r fig.retina=3, fig.height=3.5, fig.align="center"}
# ALL GEOMS will map data to colors
ggplot(data = mpg, aes(x = displ,
                       y = hwy,
                       color = class))+
  geom_point()+
  geom_smooth()
```

]

.pull-right[
```{r fig.retina=3, fig.height=3.5, fig.align="center"}
# ONLY points will map data to colors
ggplot(data = mpg, aes(x = displ,
                       y = hwy))+
  geom_point(aes(color = class))+
  geom_smooth()
```
]

---

# Mapped vs. Set Aesthetics

- `aes`thetics such as `size` and `color` can be mapped from data or set to a single value
- Map *inside* of `aes()`, set *outside* of `aes()`

.pull-left[

```{r fig.retina=3, fig.height=3.5, fig.align="center"}
# Point colors are mapped from class data
ggplot(data = mpg, aes(x = displ,
                       y = hwy))+
  geom_point(aes(color = class))+
  geom_smooth()
```
]

.pull-right[
```{r fig.retina=3, fig.height=3.5, fig.align="center"}
# Point colors are all set to blue
ggplot(data = mpg, aes(x = displ,
                       y = hwy))+
  geom_point(aes(), color = "red")+
  geom_smooth(aes(), color = "blue")
```
]

---

class: inverse, center, middle
# Saving Plots

---

# Saving Plots

---

# Go Crazy

.left-code[
```{r gap-anim, eval=F}
library("gapminder")
library("gganimate")
ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp, size = pop, color = country) +
  geom_point() +
  guides(color = FALSE, size = FALSE) +
  scale_x_log10(
    breaks = c(10^3, 10^4, 10^5), 
    labels = c("$1k", "$10k", "$100k")) +
  scale_color_manual(values = gapminder::country_colors) +
  scale_size(range = c(0.5, 12)) +
  labs(
    x = "GDP per capita",
    y = "Life Expectancy",
    caption = "Source: Hans Rosling's gapminder.org") +
  theme_minimal(14, base_family = "Fira Sans") +
  theme(
    strip.text = element_text(size = 16, face = "bold"),
    panel.border = element_rect(fill = NA, color = "grey40"),
    panel.grid.minor = element_blank())+
  transition_states(year, 1, 0)+
  ggtitle("Income and Life Expectancy - {closest_state}")
```
]

.right-plot[
```{r ref.label="gap-anim", fig.retina=3, fig.height=5.5, fig.align="center", echo=F}
```
]

---

# Data Visualization and Graphic Design Principles

.pull-left[

- We will return to various graphics as we cover descriptive statistics and regression

- I hope to cover some basic principles of good graphic design for figures and plots
    - If not in class, I will make a page on the website, and/or a video
]

--

.pull-right[

Remember:

.center[
![](https://www.dropbox.com/s/noj9n3ohqpgq2pc/powerresponsibility.jpg?raw=1)
]
]

---

# Less is More

.pull-left[

.center[

"Shoot me"
![](https://www.dropbox.com/s/azh5zxzqkmyuxoh/badfigure.png?raw=1)
]

]

--

.pull-right[

.center[
Less is More:

![](http://stat545.com/img/less-is-more-darkhorse-analytics.gif)
]
]
---

# Try to Show One Trend Really Clearly

.center[
![](https://www.dropbox.com/s/gd9eisleu21lgga/goodfigure.png?raw=1)
]

.source[New York Times: ["How Stable Are Democracies? ‘Warning Signs Are Flashing Red’"](https://www.nytimes.com/2016/11/29/world/americas/western-liberal-democracy.html), Nov 29, 2016]

---

# Reference: R Studio Makes Great "Cheat Sheet"s!

.center[
![](https://www.dropbox.com/s/k7fe3fpe6sc5oq7/ggplot2cheat.png?raw=1)
]

.source[[RStudio: ggplot2 Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)]
---

# Reference

On `ggplot2`

- **R Studio's [ggplot2 Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)**
- **`ggplot2`'s website [reference section](https://ggplot2.tidyverse.org/reference/index.html)**
- STHDA's [be awesome in ggplot2](http://www.sthda.com/english/wiki/be-awesome-in-ggplot2-a-practical-guide-to-be-highly-effective-r-software-and-data-visualization)
- r-statistic's [top 50 ggplot2 visualizations](http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html)
- Hadley Wickham's \href{http://r4ds.had.co.nz/data-visualisation.html}{R for Data Science book chapter on ggplot2}

On data visualization

- **Kieran Healy's [Data Visualization: A Practical Guide](http://socviz.co/)**
- **Claus Wilke's [Fundamentals of Data Visualization](https://serialmentor.com/dataviz)**
- PolicyViz [Better Presentations](https://policyviz.com/) 
- Karl Broman's [How to Display Data Badly](https://www.biostat.wisc.edu/~kbroman/presentations/IowaState2013/graphs_combined.pdf)
- [I Want Hue](http://tools.medialab.sciences-po.fr/iwanthue/)

### 2019-09-05 static/slides/04-slides.Rmd
Fix slides 4 final tweak

@@
 
+# The Grammar of Graphics (gg): All at Once
+
+.pull-left[
+
+All in one command
+
+- Produces plot output in viewer
+
+- Does not save plot
+    - Save with `Export` menu in viewer
+
+- Adding layers requires whole code for new plot
+
+]
+
+.pull-right[
+```{r, eval = F}
+ggplot(data = mpg)+
+  aes(x = displ,
+        y = hwy)+
+  geom_point()+
+  geom_smooth()
+```
+]
+
+---
+
+# The Grammar of Graphics (gg): As R Objects
+
+.pull-left[
+- Saving as an Object
+
+- Does not show in viewer
+    - Rune the name of the plot as command to see it
+
+- Saves your plot as an `R` object
+
+- Can add layers by calling the original plot name
+]
+
+.pull-right[
+
+```{r, eval=F}
+# make and save plot
+p <- ggplot(data = mpg)+
+  aes(x = displ,
+        y = hwy)+
+  geom_point()
+
+p # view plot
+
+# add a layer
+
+p + geom_smooth() # shows new plot
+
+p <- p + geom_smooth() # saves and overwrites p
+p2 <- p + geom_smooth() # saves as different object
+
+```
+]
+
+---
+
 # Example Data
@@
 ]
-
+]
 ---
@@
 
-# Back to the Original
+# Back to the Original (and saving it)
 
@@
 ```{r plot-8a, eval=F, echo=T}
-ggplot(data = mpg)+
+p <- ggplot(data = mpg)+ #<<
   aes(x = displ,
@@
   geom_smooth()
+
+p # show plot
 ```
@@
 ```{r, fig.retina=3, fig.height=3.5, fig.align="center"}
-ggplot(data = mpg)+
-  aes(x = displ,
-        y = hwy)+
-  geom_point(aes(color = class))+
-  geom_smooth()+
-  facet_wrap(~year) #<<
+p + facet_wrap(~year) #<<
 ```
@@
 ```{r, fig.retina=3, fig.height=3.5, fig.align="center"}
-ggplot(data = mpg)+
-  aes(x = displ,
-        y = hwy)+
-  geom_point(aes(color = class))+
-  geom_smooth()+
-  facet_grid(cyl~year) #<<
+p + facet_grid(cyl~year) #<<
 ```
@@
 
-```{r, fig.retina=3, fig.align="center"}
-ggplot(data = mpg)+
-  aes(x = displ,
-        y = hwy)+
-  geom_point(aes(color = class))+
-  geom_smooth()+
-  facet_wrap(~year)+
-  labs(x = "Engine Displacement (Liters)",
-       y = "Highway MPG",
-       title = "Car Mileage and Displacement",
-       subtitle = "More Displacement Lowers Highway MPG",
-       caption = "Source: EPA",
+```{r, fig.retina=3, fig.height=4, fig.align="center"}
+p + facet_wrap(~year)+
+  labs(x = "Engine Displacement (Liters)", #<<
+       y = "Highway MPG", #<<
+       title = "Car Mileage and Displacement", #<<
+       subtitle = "More Displacement Lowers Highway MPG", #<<
+       caption = "Source: EPA", #<<
        color = "Vehicle Class")#<<
@@
 
-class: inverse, center, middle
-# Saving Plots
-
----
-
-# Saving Plots
-
----
-
-# Go Crazy
+# Go Crazy I
+
+.left-code[
+```{r gap-map, echo=F}
+library("socviz")
+library(mapproj)
+
+us_states <- map_data("state")
+
+election$region <- tolower(election$state)
+us_states_elec <- left_join(us_states, election)
+
+theme_map <- function(base_size=9, base_family="") {
+    require(grid)
+    theme_bw(base_size=base_size, base_family=base_family) %+replace%
+        theme(axis.line=element_blank(),
+              axis.text=element_blank(),
+              axis.ticks=element_blank(),
+              axis.title=element_blank(),
+              panel.background=element_blank(),
+              panel.border=element_blank(),
+              panel.grid=element_blank(),
+              panel.spacing=unit(0, "lines"),
+              plot.background=element_blank(),
+              legend.justification = c(0,0),
+              legend.position = c(0,0)
+              )
+}
+
+county_full <- left_join(county_map, county_data, by = "id")
+```
+
+```{r gap-map2, eval=F}
+
+# I did some (hidden) data work before this! 
+ggplot(data = county_full,
+            mapping = aes(x = long, y = lat,
+                          fill = pop_dens, 
+                          group = group))+ 
+  geom_polygon(color = "gray90", size = 0.05)+
+  coord_equal()+
+  scale_fill_brewer(palette="Blues",
+                             labels = c("0-10", "10-50", "50-100", "100-500",
+                                        "500-1,000", "1,000-5,000", ">5,000"))+
+  labs(fill = "Population per\nsquare mile") +
+    theme_map() +
+    guides(fill = guide_legend(nrow = 1)) + 
+    theme(legend.position = "bottom")
+```
+]
+
+.right-plot[
+```{r ref.label="gap-map2", fig.retina=3, fig.height=5.5, fig.align="center", echo=F}
+```
+]
+
+---
+
+# Go Crazy II
 
@@
 - r-statistic's [top 50 ggplot2 visualizations](http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html)
-- Hadley Wickham's \href{http://r4ds.had.co.nz/data-visualisation.html}{R for Data Science book chapter on ggplot2}
+- Hadley Wickham's[ R for Data Science book chapter on ggplot2](http://r4ds.had.co.nz/data-visualisation.html)
 

### 2019-09-06 static/slides/04-slides.Rmd
Rescale slide images

@@
 
+.center[
+![:scale 90%](https://www.dropbox.com/s/e1nvb7x330c7ruk/aes2.png?raw=1)
+]
+]
+
+---
+
+# The Grammar of Graphics (gg): Aesthetics III
+
+.left-code[
+### Data
+### Aesthetics
+```{r, eval=F}
++ aes()
+```
+]
+
+.right-plot[
+**Aesthetics** map data to visual elements or parameters
+
 - `displ`
@@
 
-# The Grammar of Graphics (gg): Aesthetics II
+# The Grammar of Graphics (gg): Aesthetics III
 
@@
 
-# The Grammar of Graphics (gg): Aesthetics II
-
-.left-code[
-### Data
-### Aesthetics
-```{r, eval=F}
-+ aes()
-```
-]
-
-.right-plot[
-**Aesthetics** map data to visual elements or parameters
-
-.center[
-![](https://www.dropbox.com/s/e1nvb7x330c7ruk/aes2.png?raw=1)
-]
-]
----
-
-# The Grammar of Graphics (gg): Aesthetics III
+# The Grammar of Graphics (gg): Aesthetics IV
 
@@
         y = hwy)+
-  geom_point()+
+  geom_point(aes(color = class))+
   geom_smooth() #<<
@@
 .center[
-![](https://www.dropbox.com/s/gd9eisleu21lgga/goodfigure.png?raw=1)
+![:scale 70%](https://www.dropbox.com/s/gd9eisleu21lgga/goodfigure.png?raw=1)
 ]
@@
 .center[
-![](https://www.dropbox.com/s/k7fe3fpe6sc5oq7/ggplot2cheat.png?raw=1)
+![:scale 55%](https://www.dropbox.com/s/k7fe3fpe6sc5oq7/ggplot2cheat.png?raw=1)
 ]
@@
 - **`ggplot2`'s website [reference section](https://ggplot2.tidyverse.org/reference/index.html)**
+- Hadley Wickham's[ R for Data Science book chapter on ggplot2](http://r4ds.had.co.nz/data-visualisation.html)
 - STHDA's [be awesome in ggplot2](http://www.sthda.com/english/wiki/be-awesome-in-ggplot2-a-practical-guide-to-be-highly-effective-r-software-and-data-visualization)
 - r-statistic's [top 50 ggplot2 visualizations](http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html)
-- Hadley Wickham's[ R for Data Science book chapter on ggplot2](http://r4ds.had.co.nz/data-visualisation.html)
 

### 2019-09-06 content/r/03-r-practice.Rmd
Cosmetic tweaks to r practice 3, two #17s
Apply cosmetic change to r practice 3 page
Fix formatting for R practice 3 and 4, add answers for practice 3

@@
 - [<i class="fas fa-file-pdf"></i> `Download as PDF`](/r/03-r-practice.pdf)
+- [<i class="fas fa-file-pdf"></i> `Answer Key`](/r/03-r-practice-answers.pdf)
+- [<i class="fas fa-registered"></i> `R Script`](/r/03-r-practice-answers.R)
 
@@
 
-## 1.
+## 1
 
-a. Create a vector called `me` with two objects, your first name, and your last name. 
-b. Call the vector to inspect it.
-c. Confirm it is a character class vector.
+Work on the following parts:
 
-## 2. Use R's help functions to determine what the `paste()` function does. Then paste together your first name and last name.
+- a. Create a vector called `me` with two objects, your first name, and your last name.
+- b. Call the vector to inspect it.
+- c. Confirm it is a character class vector.
 
-## 3. Create a vector called `my_vector` with all the even integers from 2 to 10.
+## 2
 
-## 4. Find the mean of `my_vector` with `mean()`.
+Use R's help functions to determine what the `paste()` function does. Then paste together your first name and last name.
 
-## 5. Take all the integers from 18 to 763,^[Hint: use the `:` operator to create a sequence from a starting number to an ending number] then get the mean.
+## 3
 
-## Playing with Data
+Create a vector called `my_vector` with all the even integers from 2 to 10.
+
+## 4
+
+Find the mean of `my_vector` with `mean()`.
+
+## 5
+
+Take all the integers from 18 to 763,^[Hint: use the `:` operator to create a sequence from a starting number to an ending number] then get the mean.
+
+# Playing with Data
 
@@
 
-### 6. Install `ggplot2`.
+## 6
 
-### 7. Load `ggplot2` with the `library()` command.
+Install `ggplot2`.
 
-### 8. Get the `str`ucture of the diamonds data frame. What are the different variables and what kind of data does each contain?
+## 7
 
-### 9. Get summary statistics separately for `carat`, `depth`, `table`, and `price`.
+Load `ggplot2` with the `library()` command.
 
-### 10. `color`, `cut`, and `clarity` are categorical variables (factors). Use the `table()` command to generate frequency tables for each.
+## 8
 
-### 11. Now rerun the `summary()` command on the entire data frame.
+Get the `str`ucture of the diamonds data frame. What are the different variables and what kind of data does each contain?
 
-### 12. Now look only at (subset) the first 4 diamonds in the dataset.
+## 9
 
-### 13. Now look only at (subset) the third and seventh diamond in the dataset.
+Get summary statistics separately for `carat`, `depth`, `table`, and `price`.
 
-### 14. Now look only at (subset) the second column of the dataset.
+## 10
 
-### 15. Do this again, but look using the `$` to pull up the second column by name.
+`color`, `cut`, and `clarity` are categorical variables (factors). Use the `table()` command to generate frequency tables for each.
 
-### 16. Now look only at diamonds that have a `carat` greater than or equal to 1.
+## 11
 
-### 17. Now look only at diamonds that have a VVS1 `clarity`.
+Now rerun the `summary()` command on the entire data frame.
 
-### 17. Now look only at dimaonds that have a color of E, F, I, and J.
+## 12
 
-### 18. Now look only at diamonds that have a `carat` greater than or equal to 1 and a VVS1 `clarity`.
+Now look only at (subset) the first 4 diamonds in the dataset.
 
-### 19. Get the average price of diamonds in question 18.^[Hints: use your subset command as an argument to the mean function. You will not need a comma here because you are looking for a single row.]
+## 13
 
-### 20. What is the highest price for a diamond with a 1.0 carat, D color, and VVS1 clarity?
+Now look only at (subset) the third and seventh diamond in the dataset.
 
-## Execute your R Script
+## 14
+
+Now look only at (subset) the second column of the dataset.
+
+## 15
+
+Do this again, but look using the `$` to pull up the second column by name.
+
+## 16
+
+Now look only at diamonds that have a `carat` greater than or equal to 1.
+
+## 17
+
+Now look only at diamonds that have a VVS1 `clarity`.
+
+## 18
+
+Now look only at dimaonds that have a color of E, F, I, and J.
+
+## 19
+
+Now look only at diamonds that have a `carat` greater than or equal to 1 and a VVS1 `clarity`.
+
+## 20
+
+Get the average price of diamonds in question 18.^[Hints: use your subset command as an argument to the mean function. You will not need a comma here because you are looking for a single row.]
+
+## 21
+
+What is the highest price for a diamond with a 1.0 carat, D color, and VVS1 clarity?
+
+# Execute your R Script
 

### 2019-09-06 content/r/04-r-practice.Rmd
Cosmetic changes to r practice webpages for classes 3 and 4
Fix formatting for R practice 3 and 4, add answers for practice 3

@@
 
-## 1.
+## 1
+
 We will look at GDP per Capita and Life Expectancy using some data from the [gapminder project](http://gapminder.org).
@@
 
-## 2. Let's get a quick look at `gapminder` to see what we're dealing with.
+## 2
+
+Let's get a quick look at `gapminder` to see what we're dealing with.
 
@@
 
-## 3. Let's make sure you can do some basic plots before we get into the `gg`. Use base `R`'s `hist()` function to plot a *histogram* of `gdpPercap`.
+## 3
 
-## 4. Use base `R`'s `boxplot()` function to plot a *boxplot* of `gdpPercap`.
+Let's make sure you can do some basic plots before we get into the `gg`. Use base `R`'s `hist()` function to plot a *histogram* of `gdpPercap`.
 
-## 5. Now make it a *boxplot* by `continent`.^[Hint: use formula notation with`~`.]
+## 4
 
-## 6. Now make a *scatterplot* of `gdpPercap` on the $x$-axis and `LifeExp` on the $y$-axis.
+Use base `R`'s `boxplot()` function to plot a *boxplot* of `gdpPercap`.
+
+## 5
+
+Now make it a *boxplot* by `continent`.^[Hint: use formula notation with `~`.]
+
+## 6
+
+Now make a *scatterplot* of `gdpPercap` on the $x$-axis and `LifeExp` on the $y$-axis.
 
@@
 
-## 7. Load the package `ggplot2` (you should have installed it previously. If not, install first with `install.packages("ggplot2")`).
+## 7
 
-## 8.
+Load the package `ggplot2` (you should have installed it previously. If not, install first with `install.packages("ggplot2")`).
+
+## 8
+
 Let's first make a `bar` graph to see how many countries are in each continent.
@@
 
-## 9. For quantitative data, we want a `histogram` to visualize the distribution of a variable. Make a `histogram` of `gdpPercap`. Your only `aes`thetic here is to map `gdpPercap` to `x`.
+## 9
 
-## 10. Now let's try adding some color, specifically, add an `aes`thetic that maps `continent` to `color.`
+For quantitative data, we want a `histogram` to visualize the distribution of a variable. Make a `histogram` of `gdpPercap`. Your only `aes`thetic here is to map `gdpPercap` to `x`.
 
-## 11. Instead of a `histogram`, change the `geom` to make it a `density` graph. To avoid overplotting, add `alpha=0.4` to the `geom` argument (alpha changes the *transparency* of a `fill`).
+## 10
 
-## 12. Redo your plot from 10 for `lifeExp` instead of `gdpPercap`.
+Now let's try adding some color, specifically, add an `aes`thetic that maps `continent` to `fill.`^[In general, `color` refers to the outside borders of a `geom` (except points), `fill` is the interior of an object.]
 
-## 13. Now let's try a scatterplot for `lifeExp` (as `y`) on `gdpPercap` (as `x`). You'll need both for `aes`thetics. The `geom` here is `geom_point()`.
+## 11
 
-## 14. Add some color by mapping `continent` to `color` in your `aes`thetics.
+Instead of a `histogram`, change the `geom` to make it a `density` graph. To avoid overplotting, add `alpha=0.4` to the `geom` argument (alpha changes the *transparency* of a `fill`).
 
-## 15. Now let's try adding a regression line with `geom_smooth()`. Add this layer on top of your `geom_point()` layer.
+## 12
 
-## 16.
+Redo your plot from 11 for `lifeExp` instead of `gdpPercap`.
+
+## 13
+
+Now let's try a scatterplot for `lifeExp` (as `y`) on `gdpPercap` (as `x`). You'll need both for `aes`thetics. The `geom` here is `geom_point()`.
+
+## 14
+
+Add some color by mapping `continent` to `color` in your `aes`thetics.
+
+## 15
+
+Now let's try adding a regression line with `geom_smooth()`. Add this layer on top of your `geom_point()` layer.
+
+## 16
+
 Did you notice that you got multiple regression lines (colored by continent)?
@@
 
-## 17. Now add an `aes`thetic to your `point`s to map `pop` to `size`.
+## 17
 
-## 18.
+Now add an `aes`thetic to your `point`s to map `pop` to `size`.
+
+## 18
+
 Change the color of the regression line to `"black"`.
@@
 
-## 19. Another way to separate out continents is with `facet`ing. Add `+facet_wrap(~continent)` to create subplots by `continent`.
+## 19
 
-## 20. Remove the `facet` layer. The `scale` is quite annoying for the `x`-axis, a lot of points are clustered on the lower level. Let's try changing the scale by adding a layer: `+scale_x_log10()`.
+Another way to separate out continents is with `facet`ing. Add `+facet_wrap(~continent)` to create subplots by `continent`.
 
-## 21.
+## 20
+
+Remove the `facet` layer. The `scale` is quite annoying for the `x`-axis, a lot of points are clustered on the lower level. Let's try changing the scale by adding a layer: `+scale_x_log10()`.
+
+## 21
+
 Now let's fix the labels by adding `+labs()`.
@@
 
-## 22.
+## 22
+
 Now let's try subsetting by looking only at North America.
 Take the `gapminder` dataframe and subset it to only look at `continent=="Americas"`).
-Assign this to a new dataframe object (call it something like `america`.) Now, use *this* as your `data`, and redo the graph from question 16.
+Assign this to a new dataframe object (call it something like `america`.) Now, use *this* as your `data`, and redo the graph from question 17.
 (You might want to take a look at your new dataframe to make sure it worked first!)
 
-## 23. Try this again for the *whole* world, but just for observations in the year 2000.
+## 23
+
+Try this again for the *whole* world, but just for observations in the year 2002.

### 2019-09-06 static/r/04-r-practice.Rmd
Fix formatting for R practice 3 and 4, add answers for practice 3

@@
 
-## 1.
+### 1.
 We will look at GDP per Capita and Life Expectancy using some data from the [gapminder project](http://gapminder.org).
@@
 
-## 2. Let's get a quick look at `gapminder` to see what we're dealing with.
+### 2. Let's get a quick look at `gapminder` to see what we're dealing with.
 
 a. Get the `str`ucture of the `gapminder` data.
+
 b. What variables are there?
+
 c. Look at the `head` of the dataset to get an idea of what the data looks like.
+
 d. Get `summary` statistics of all variables.
@@
 
-## 3. Let's make sure you can do some basic plots before we get into the `gg`. Use base `R`'s `hist()` function to plot a *histogram* of `gdpPercap`.
+### 3. Let's make sure you can do some basic plots before we get into the `gg`. Use base `R`'s `hist()` function to plot a *histogram* of `gdpPercap`.
 
-## 4. Use base `R`'s `boxplot()` function to plot a *boxplot* of `gdpPercap`.
+### 4. Use base `R`'s `boxplot()` function to plot a *boxplot* of `gdpPercap`.
 
-## 5. Now make it a *boxplot* by `continent`.^[Hint: use formula notation with`~`.]
+### 5. Now make it a *boxplot* by `continent`.^[Hint: use formula notation with `~`.]
 
-## 6. Now make a *scatterplot* of `gdpPercap` on the $x$-axis and `LifeExp` on the $y$-axis.
+### 6. Now make a *scatterplot* of `gdpPercap` on the $x$-axis and `LifeExp` on the $y$-axis.
 
@@
 
-## 7. Load the package `ggplot2` (you should have installed it previously. If not, install first with `install.packages("ggplot2")`).
+### 7. Load the package `ggplot2` (you should have installed it previously. If not, install first with `install.packages("ggplot2")`).
 
-## 8.
+### 8.
 Let's first make a `bar` graph to see how many countries are in each continent.
@@
 
-## 9. For quantitative data, we want a `histogram` to visualize the distribution of a variable. Make a `histogram` of `gdpPercap`. Your only `aes`thetic here is to map `gdpPercap` to `x`.
+### 9. For quantitative data, we want a `histogram` to visualize the distribution of a variable. Make a `histogram` of `gdpPercap`. Your only `aes`thetic here is to map `gdpPercap` to `x`.
 
-## 10. Now let's try adding some color, specifically, add an `aes`thetic that maps `continent` to `color.`
+### 10.
+Now let's try adding some color, specifically, add an `aes`thetic that maps `continent` to `fill`.^[In general, `color` refers to the outside borders of a `geom` (except points), `fill` is the interior of an object.]
 
-## 11. Instead of a `histogram`, change the `geom` to make it a `density` graph. To avoid overplotting, add `alpha=0.4` to the `geom` argument (alpha changes the *transparency* of a `fill`).
+### 11. Instead of a `histogram`, change the `geom` to make it a `density` graph. To avoid overplotting, add `alpha=0.4` to the `geom` argument (alpha changes the *transparency* of a `fill`).
 
-## 12. Redo your plot from 10 for `lifeExp` instead of `gdpPercap`.
+### 12. Redo your plot from 11 for `lifeExp` instead of `gdpPercap`.
 
-## 13. Now let's try a scatterplot for `lifeExp` (as `y`) on `gdpPercap` (as `x`). You'll need both for `aes`thetics. The `geom` here is `geom_point()`.
+### 13. Now let's try a scatterplot for `lifeExp` (as `y`) on `gdpPercap` (as `x`). You'll need both for `aes`thetics. The `geom` here is `geom_point()`.
 
-## 14. Add some color by mapping `continent` to `color` in your `aes`thetics.
+### 14. Add some color by mapping `continent` to `color` in your `aes`thetics.
 
-## 15. Now let's try adding a regression line with `geom_smooth()`. Add this layer on top of your `geom_point()` layer.
+### 15. Now let's try adding a regression line with `geom_smooth()`. Add this layer on top of your `geom_point()` layer.
 
-## 16.
+### 16.
 Did you notice that you got multiple regression lines (colored by continent)?
@@
 
-## 17. Now add an `aes`thetic to your `point`s to map `pop` to `size`.
+### 17. Now add an `aes`thetic to your `point`s to map `pop` to `size`.
 
-## 18.
+### 18.
 Change the color of the regression line to `"black"`.
@@
 
-## 19. Another way to separate out continents is with `facet`ing. Add `+facet_wrap(~continent)` to create subplots by `continent`.
+### 19. Another way to separate out continents is with `facet`ing. Add `+facet_wrap(~continent)` to create subplots by `continent`.
 
-## 20. Remove the `facet` layer. The `scale` is quite annoying for the `x`-axis, a lot of points are clustered on the lower level. Let's try changing the scale by adding a layer: `+scale_x_log10()`.
+### 20. Remove the `facet` layer. The `scale` is quite annoying for the `x`-axis, a lot of points are clustered on the lower level. Let's try changing the scale by adding a layer: `+scale_x_log10()`.
 
-## 21.
+### 21.
 Now let's fix the labels by adding `+labs()`.
@@
 
-## 22.
+### 22.
 Now let's try subsetting by looking only at North America.
 Take the `gapminder` dataframe and subset it to only look at `continent=="Americas"`).
-Assign this to a new dataframe object (call it something like `america`.) Now, use *this* as your `data`, and redo the graph from question 16.
+Assign this to a new dataframe object (call it something like `america`.) Now, use *this* as your `data`, and redo the graph from question 17.
 (You might want to take a look at your new dataframe to make sure it worked first!)
 
-## 23. Try this again for the *whole* world, but just for observations in the year 2000.
+### 23. Try this again for the *whole* world, but just for observations in the year 2002.

### 2019-09-06 static/homeworks/00-preliminary-survey-answers-pdf.Rmd
Post preliminary survey answers

---
title: "Preliminary Survey on Statistics and Software (Answers)"
author: "Ryan Safner"
date: "ECON 480 - Fall 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
library("tidyverse")
```

This is an *ungraded* and *anonymous* survey for me to evaluate the distribution of your math and statistics backgrounds. Please write your answers the the following questions on a piece of paper, or download and print a PDF copy (link above) to write on. You do not need to write your name on it. Please complete all problems to the best of your ability. Your responses will help me craft the course to see which material we need to focus on at greater length, especially review material.

### 1. Using the following sample data:

$$8, 12, 9, 10, 11, 5, 15$$

### a. Find the median.

First, arrange these in numerical order. 
$$5, 8, 9, \mathbf{10}, 11, 12, 15$$
The middle number is 10, there are 3 on either side.

### b. Calculate the sample mean, $\bar{x}$

$$\begin{aligned}
		\bar{x}&=\frac{1}{n}\displaystyle\sum^n_{i=1} x_i\\
		\bar{x}&=\frac{5+8+9+10+11+12+15}{6}	\\
		&=\frac{70}{7}\\
		&=10\\
		\end{aligned}$$

### c. Calculate the sample standard deviation, $s$

Sample standard deviation is the average deviation of an observation from the mean, calculated as the square root of variance (which is the average squared deviation from the mean), (with a slight correction for samples: $n-1$ rather than $n$). First we find the variance:

$$s^2=\frac{1}{n-1}\displaystyle \sum_{i=1}^n (x_i-\bar{x})^2$$

You can calculate this all at once: 
$$\begin{aligned}
		s^2&=\frac{(5-10)^2+(8-10)^2+(9-10)^2+(10-10)^2+(11-10)^2+(12-10)^2+(15-10)^2}{7-1}\\
		&=\frac{60}{6}\\
		&=10\\
		\end{aligned}$$

But it is often easier to first find the deviations from the mean for each data point (2^nd^ column), square them (3^rd^ column), add them all up (bottom row) and then divide by $n-1$:

| $x_i$ | $(x_i-\bar{x})$ | $(x_i-\bar{x})^2$ |
|------|------|------|
| 5    | $-5$ |  25  |
| 8    | $-2$ |  4   | 
| 9    | $-1$ |  1   |
| 10   | 0    |  0   |
| 11   | 1    |  1   |
| 12   | 2    |  4   |
| 15   | 5    | 25   |
| $\sum^n_{i=1}(x_i-\bar{x})^2$ |     |   $\mathbf{60}$ |

$$\begin{aligned}
		s^2&=\frac{1}{n-1} \displaystyle \sum_{i=1}^n (x_i-\bar{x})^2\\
		&=\frac{1}{7-1}[60] \\
		&=\frac{60}{6}\\
		&=10\\
		\end{aligned}$$

Standard deviation is the square root of variance: 
$$s=\sqrt{s^2}=\sqrt{10}\approx 3.16$$

### 2. For a fair, six-sided die:

### a. What is the probability of rolling a 5?

All outcomes $\{1, 2, 3, 4, 5, 6\}$ are equally likely, so the chance of any one outcome is the same, $\frac{1}{6}$.

### b. What is the probability of rolling an even number?

There are three possible disjoint outcomes, rolling a 2, 4, or 6, so we add up the individual probabilities: 
$$\begin{aligned}
 	P(x=even)&=P(x=2)+P(x=4)+P(x=6)\\
 	&=\bigg( \frac{1}{6} \bigg)+ \bigg( \frac{1}{6} \bigg)+ \bigg(\frac{1}{6} \bigg)\\
 	& = \frac{3}{6}\\
 	& = 0.5\\
 \end{aligned}$$

### c. What is the probability of rolling an even number or a 3?

We know the probability of an even roll is 0.5, and rolling a 3 is disjoint from rolling an even number (a roll cannot be simultaneously 3 and even). So this is the union $\cup$ of two events. 
$$\begin{aligned}
			P(x= \text{even OR } x=3)&= P(x=even)+P(x=3)\\
			&=\bigg( \frac{3}{6} \bigg) + \bigg( \frac{1}{6} \bigg)\\
			&=\frac{4}{6}\\ 
			\end{aligned}$$

### d. If you have two fair dice, what is the probability of rolling a 6 on both dice?

This is the intersection $\cap$ of two *independent* possibilities: that each die will role a six, each with the probability of $\frac{1}{6}$, and that the outcome of the first die has no bearing on the outcome of the second die. Thus, we can simply multiply the two events' probabilities:

$$\begin{aligned}
		P(x_1=6, x_2=6) &= P(x_1=6) \times P(x_2=6)\\
		&= \bigg(\frac{1}{6}\bigg) \times 	\bigg(\frac{1}{6}\bigg)\\
		&=\frac{1}{36}\\
		\end{aligned}$$

\clearpage

### 3. Hedge fund A earns an average rate of return of 2.5% per year with a standard deviation of 0.5%, while hedge fund B earns an average rate of return of 3.0% per year with a standard deviation of 2.0\%. Which is more unusual, Hedge fund A earning a 4.0% return or hedge fund B earning a return -1.0% return? Why?^[Hint: Standardize the two hedge funds.]

We must standardize these variables and compare their $Z$-scores. We calculate an observation's $Z$-score relative to its distribution by taking the observation, subtracting the mean of that variable, and dividing by its standard deviation:

$$Z=\frac{x_i-\bar{X}}{s_X}	$$

For hedge fund A:

$$Z=\frac{4.0-2.5}{0.5}=	3$$

A 4.0% return for Hedge Fund $A$ is 3 standard deviations above its mean return.

For hedge fund B:

$$Z=\frac{-1.0-3.0}{2.0}=-2$$

A $-1.0$% return for Hedge Fund $B$ is 2 standard deviations below its mean return.

The event in hedge fund A is more extreme.

\clearpage

### 4. A discrete random variable $X$ has the following pdf:

```{r, echo=FALSE, results="asis"}
tribble(
  ~x, ~`p(x)`,
  10, 0.10,
  20, 0.20,
  30, 0.30,
  40, 0.40
  ) %>%
  knitr::kable(., "latex")
```

Calculate the sample standard deviation, $s$ of $X$.

First we must calculate the mean, $\mu$ or $E(X)$:

$$\begin{aligned}E(X) &= \sum^n_{i=1} x_ip_i \\
&= 0.1(10)+0.2(20)+0.3(30)+0.4(40)\\
& = 1+4+9+16\\
&=30\end{aligned}$$

Now we must calculate the variance, the average squared distance from the mean (30) weighted by the probability of each value. We can do it all at once:

$$\begin{aligned}var(X) &= E[(X_i-\mu)^2]\\
& = \sum p_i(X_i-\mu)^2\\
&= 0.1(10-30)^2	+0.2(20-30)^2+0.3(30-30)^2+0.4(40-30)^2\\
& = 0.1(400)+0.2(100)+0.3(0)+0.4(100)\\
&= 40+20+0+40=100\\
\end{aligned}$$

Alternatively, we could use the table method to make sure we keep our numbers straight:

| $x_i$ | $p_i$ | $(x_i-\mu)$ | $(x_i-\mu)^2$ | $p_i(x_i-\mu)^2$ | 
|-------|-------|-------|-------|-------|
| 10    | 0.10  | $-20$ | 400   | 40    |
| 20    | 0.20  | $-10$ | 100   | 20    |
| 30    | 0.30  | 0     | 0     | 0     |
| 40    | 0.40  | 10    | 100   | 40    |
| $\displaystyle\sum^n_{i=1} p_i(x_i-\mu)^2$ |      |     |     | $\mathbf{100}$ |

The first column is each individual value. The second is the probability of each value. For the third row, we take the deviation of each value from the mean. The fourth column, we square it. The fifth column, we multiply the squared deviation by the probability of the value. Then, we add up the fifth column to get the variance, 100.

Standard deviation is just the square root of the variance: 
$$sd(X) = \sqrt{var(X)}=\sqrt{100}=10$$

\clearpage

### 5. The random variable $Y$ is normally distributed with a mean of 50 and standard deviation of 12

$$Y \sim N (50,12)$$

### a. What is the $Z$-score for $Y=74$?

The z-score formula subtracts the mean from a value and divides by the standard deviation to convert any normal distribution to the standard normal distribution $(Y \sim N(0,1)$: 
$$\begin{aligned}z&=\frac{y-\mu}{\sigma}\\
z&=\frac{74-50}{12}\\
z&= 2\\ \end{aligned}$$

### b. In your own words, what does this $Z$-score mean?

The z-score or the standard score measures how many standard deviations a value is away from the distribution's mean. Thus, this value, 74, is 2 standard deviations away from the distribution's mean of 50.

### c. What is the probability that $Y$ takes on a value *greater* than 74?

This requires remembering an important property of standard normal distributions, the 68-95-99.7 rule: 68% of a normal distribution falls between 1 standard deviation from the mean, 95% between 2 standard deviations, 	and 99.7 between 3 standard deviations. Our z-score told us that we are 2 standard deviations from the mean, so there is a 95% chance that $Y$ will fall within 2 standard deviations from the mean $(50 \pm 12$, so between 26 and 74).

However, we need to be careful, because the remaining 5% can fall on either side of two deviations from the mean (). What we care about is two deviations *above* the mean, that is, above 74. So divide this 5% in half, to account for the **2.5%** chance that Y takes on a value greater than 74.

\begin{figure}
\centering
  \includegraphics{00-preliminary-survey-5c.png}
\end{figure}

```{r, echo=F, eval=F}
# used to make figure
update_geom_defaults("label", list(family = "Fira Sans Condensed"))

normal.pdf <- ggplot(data = data.frame(x = 2:122), # our "data" is just a sequence of x from -4 to 4
                     aes(x = x))+
  stat_function(fun = dnorm, # dnorm is the normal distributions
                args = list(mean = 50, sd = 12), # set mean = 0 and sd = 1
                color="blue")+
  scale_x_continuous(breaks = seq(2,122,12))+
    xlab("Y")+ylab("p(Y)")+
  theme_classic(base_family = "Fira Sans")

shade95 <- function(x) {
    y <- dnorm(x, mean = 50, sd = 12)
    y[x < 74 | x > 122] <- NA
    return(y)
}

normal.pdf + 
  stat_function(fun = shade95, geom="area", fill="green", alpha=0.5)+
  # add 68% 1 SD 
    annotate("text", x = 50, y = 0.021, label = expression(bar(y) %+-% sigma), color="#000000")+ 
    annotate("text", x = 50, y = 0.019, label = "68%", color="#000000")+ 
  annotate("segment", x = 38, xend = 62, y = 0.02, yend = 0.02, colour = "#000000", size=0.5, linetype = 1, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))+
  annotate("segment", x = 38, xend = 38, y = 0, yend = 0.02, colour = "red", size=0.25, linetype = 2)+
  annotate("segment", x = 62, xend = 62, y = 0, yend = 0.02, colour = "red", size=0.25, linetype = 2)+
  # add 95% 2 SDs 
    annotate("text", x = 50, y = 0.006, label = expression(bar(y) %+-% 2*sigma), color="#000000")+ 
    annotate("text", x = 50, y = 0.004, label = "95%", color="#000000")+ 
  annotate("segment", x = 26, xend = 74, y = 0.005, yend = 0.005, colour = "#000000", size=0.5, linetype = 1, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))+
  annotate("segment", x = 26, xend = 26, y = 0, yend = 0.005, colour = "red", size=0.25, linetype = 2)+
  annotate("segment", x = 74, xend = 74, y = 0, yend = 0.005, colour = "red", size=0.25, linetype = 2)+
  # add 2.5 % 
  annotate("text", x = 98, y = 0.005, label = "2.5%", color="green")+ 
  annotate("segment", x = 74, xend = 122, y = 0.004, yend = 0.004, colour = "green", size=0.5, linetype = 2, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))
```

### 2019-09-06 static/homeworks/00-preliminary-survey-answers.Rmd
Post preliminary survey answers

--- 2019-09-06 static/homeworks/00-preliminary-survey-answers-pdf.Rmd
@@
 date: "ECON 480 - Fall 2019"
-output: pdf_document
+output:
+  html_document:
+    df_print: kable
+    theme: simplex
+    toc: true 
+    toc_depth: 3
+    toc_float: true
+    code_folding: show
+    highlight: tango
+
 ---
@@
 library("tidyverse")
+library("kableExtra")
 ```
@@
 
-### 1. Using the following sample data:
+## 1.
+#### Using the following sample data:
 
@@
 
-### a. Find the median.
+### a.
+
+#### Find the median.
 
@@
 
-### b. Calculate the sample mean, $\bar{x}$
+### b.
+
+#### Calculate the sample mean, $\bar{x}$
 
@@
 
-### c. Calculate the sample standard deviation, $s$
+### c. 
+#### Calculate the sample standard deviation, $s$
 
@@
 
-### 2. For a fair, six-sided die:
+## 2.
 
-### a. What is the probability of rolling a 5?
+For a fair, six-sided die:
+
+### a.
+#### What is the probability of rolling a 5?
 
@@
 
-### b. What is the probability of rolling an even number?
+### b. 
+#### What is the probability of rolling an even number?
 
@@
 
-### c. What is the probability of rolling an even number or a 3?
+### c.
+#### What is the probability of rolling an even number or a 3?
 
@@
 
-### d. If you have two fair dice, what is the probability of rolling a 6 on both dice?
+### d.
+#### If you have two fair dice, what is the probability of rolling a 6 on both dice?
 
@@
 
-\clearpage
+## 3.
 
-### 3.
-Hedge fund A earns an average rate of return of 2.5% per year with a standard deviation of 0.5%, while hedge fund B earns an average rate of return of 3.0% per year with a standard deviation of 2.0\%.
+#### Hedge fund A earns an average rate of return of 2.5% per year with a standard deviation of 0.5%, while hedge fund B earns an average rate of return of 3.0% per year with a standard deviation of 2.0\%.
 Which is more unusual, Hedge fund A earning a 4.0% return or hedge fund B earning a return -1.0% return?
@@
 
-\clearpage
-
-### 4. A discrete random variable $X$ has the following pdf:
+## 4.
+#### A discrete random variable $X$ has the following pdf:
 
@@
   ) %>%
-  knitr::kable(., "latex")
+  knitr::kable()%>%
+  kable_styling()
 ```
 
-Calculate the sample standard deviation, $s$ of $X$.
+#### Calculate the sample standard deviation, $s$ of $X$.
 
@@
 
-\clearpage
-
-### 5. The random variable $Y$ is normally distributed with a mean of 50 and standard deviation of 12
+## 5. 
+#### The random variable $Y$ is normally distributed with a mean of 50 and standard deviation of 12
 
@@
 
-### a. What is the $Z$-score for $Y=74$?
+### a. 
+#### What is the $Z$-score for $Y=74$?
 
@@
 
-### b. In your own words, what does this $Z$-score mean?
+### b.
+#### In your own words, what does this $Z$-score mean?
 
@@
 
-### c. What is the probability that $Y$ takes on a value *greater* than 74?
+### c.
+#### What is the probability that $Y$ takes on a value *greater* than 74?
 
@@
 
-\begin{figure}
-\centering
-  \includegraphics{00-preliminary-survey-5c.png}
-\end{figure}
-
-```{r, echo=F, eval=F}
+```{r, echo=F, fig.retina=3}
 # used to make figure

### 2019-09-06 static/slides/05-slides.Rmd
Post preliminary survey answers

---
title: "1.5: Data Wrangling in the tidyverse"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: solarized-dark #ocean rainbow #monokai rainbow solarized-light #tomorrow-night -blue -eighties zenburn
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(message=F,
                      warning=F)
library(tidyverse)

```

# Data Wrangling

- Most real world data analysis transforming chaos into order
    - Data strewn from multiple sources
    - Missing data
    - Data not in readable form

---

# Workflow of a Data Scientist

.pull-left[
.center[.polaroid[![](https://www.dropbox.com/s/zke6m65wp3upouj/workflow.png?raw=1)]]
]

.pull-right[
- Consider Hadley Wickham's famous diagram from [R for Data Science](http://r4ds.had.co.nz)
1. .shoutcaps[Import] raw data from out there in the world
2. .shoutcaps[Tidy] it into a form that you can use to 
3. .shoutcaps[Explore] the data
4. .shoutcaps[Communicate] results to target audience
]

---

# Workflow of a Data Scientist

.pull-left[
.center[.polaroid[![](https://www.dropbox.com/s/zke6m65wp3upouj/workflow.png?raw=1)]]
]

.pull-right[

3. .shoutcaps[Explore] the data:
- .whisper[Transform] existing variables into more useful ones
- Make plots to .whisper[visualize] .onfire[relationships]
- Create .whisper[models] to analyze, predict, and interpret causal relationships

- 95% of the day-to-day work of data scientists, researchers, econometricians: ".hl[data wrangling]"
]

---

# Workflow of a Data Scientist

.pull-left[
.center[
![](https://www.dropbox.com/s/jd68ky17gb7pmu0/datawranglingnyt.png?raw=1)
]
]

.pull-right[
.center[
![](https://www.dropbox.com/s/8kzy823yubdv8gh/datawranglingnyt2.png?raw=1)
]
]

.center[
[New York Times](https://www.nytimes.com/2014/08/18/technology/for-big-data-scientists-hurdle-to-insights-is-janitor-work.html)
]

---

class: blank

background-image: url(https://www.dropbox.com/s/np8v4ihjotkckc1/tidyverse1.png?raw=1)
background-size: cover

---

# The tidyverse

> "The tidyverse is an opinionated  collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures.

- Core packages include ones we've discussed before: `ggplot2`, `dplyr`, `magrittr` among several others (`tidy` `readr`, `purr`, `forcats`, `stringr`)
- Loading any `tidyverse` package loads `magrittr` (so you can use `%>%`)
- Learn more at [tidyverse.org](tidyverse.org)

![](https://www.dropbox.com/s/81ijxswbwqor0up/tidyverse.png?raw=1)

---

# The tidyverse II

- Easiest to just load the core tidyverse all at once
    - Note loading the tidyverse is "noisy", it will spew a lot of messages
    - Hide them with `suppressPackageStartupMessages()` and insert the `library()` command inside the `()`

```{r, echo=FALSE}
detach("package:tidyverse", TRUE) # just to show start up message of tidyverse, let's unload it first
```

```{r, message=TRUE, warning=TRUE}
library("tidyverse")
```

---

# The tidyverse III

![](https://www.dropbox.com/s/9h1trto0iapnklx/tidyverse2.png?raw=1)

---

# Tidyverse Packages

- We will make **extensive** use of:

1. `dplyr` for data wrangling
2. `tidyr` for tidying data
3. `ggplot2` for plotting data

- Loading `tidyverse` makes **automatically** uses

4. `tibble` for better data frames
5. `magrittr` for "pipeable" code

- We might later look at

6. `readr` for importing data
7. `forcats` for working with factors
8. `stringr` for working with strings

---

# Tibble

.left-column[

.center[![](https://www.dropbox.com/s/u6ubqkng1ja5jwf/tibble.png?raw=1)]
]

.right-column[

- `tibble` converts all `data.frames` into a .whisper[friendlier] version called `tibbles`

]

---

# Tibble

.pull-left[

```{r, eval=F}
diamonds
```

<!--Can't fit in columns so artificially selecting to print only first 7 columns -->

```{r, echo=F} 
diamonds %>% 
  select(1:7)
```

]

.pull-right[

- Prints much nicer output in R
- Shows a bit of the `str`ucture of a dataframe automatically:
    - `nrow() x ncol()`
    - `<dbl>` is numeric ("double")
    - `<ord>` is an ordered factor
    - `<int>` is an integer
- Fundamental grammar of tidyverse: 
    1. start with a tibble 
    2. run a function on it
    3. output a new tibble
]

---

# Magrittr I

.right-column[

- The .hl[magrittr] package allows us to use the **"pipe" operator** (`%>%`)<sup>.red[1]</sup>
- `%>%` "pipes" the *output* of the *left* of the pipe *into* the *argument* of the *right* 
    - Running a function `f` on object `x` as `f(x)` becomes `x %>% f` in pipeable form
    - i.e. "take `x` and then run function `f` on it"
]

.left-column[

.center[![](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)]
]

.footnote[<sup>.red[1]</sup> Keyboard shortcuts in R Studio: `CTRL+Shift+M` (Windows) or `Cmd+Shift+M` (Mac)]

---

# Magrittr II

.right-column[

- With ordinary math functions, read from outside $\leftarrow$ (inside): 
$$g(f(x))$$

i.e. take `x` and perform function `f()` on `x` and then take that result and perform function `g()` on it
- With pipes, read operations from left $\rightarrow$ right:

```{r, eval=FALSE}
x %>% f %>% g
```

take `x` and then perform function `f` on it, then perform function `g` on that result

- So read `%>%` mentally as "and then"

]

.left-column[

.center[![](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)]
]

---

# Magrittr III

.right-column[

.content-box-green[
.green[**Example**]

$$ln(exp(x))$$

- First, exponentiate $x$, then take the natural log of that (resulting in just x)
- In pipes:

```{r, eval=F}
x %>% exp() %>% ln()
```
]

]

.left-column[

.center[![](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)]
]

---

# Magrittr III

.right-column[
.content-box-green[
.green[**Example**]

- Sequence: find keys, unlock car, drive to school, park 
- Using nested functions in pseudo-"code":

```{r, eval=F}
park(drive(start_car(find("keys")), to = "campus"))
```

- Using pipes:

```{r, eval=F}
find("keys") %>%
  start_car() %>%
  drive(to = "campus") %>%
  park()
```

]

]

.left-column[

.center[![](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)]
]

---

# Magrittr: Simple Example

```{r, echo=FALSE}
library("gapminder")
```

.pull-left[

```{r}
# look at top 6 rows
head(gapminder) 
```
]

.pull-right[
```{r}
# use the pipe instead 
gapminder %>% head()
```
]

---

# Magrittr: More Involved Example

- These two methods produce the same output (avg hwy mpg of Audi cars)

- Without the pipe

```{r, eval=F}
summarise(group_by(filter(mpg, manufacturer=="audi"), model), hwy_mean = mean(hwy))
```

- Using the pipe

```{r}
mpg %>%
  filter(manufacturer=="audi") %>%
  group_by(model) %>%
  summarise(hwy_mean = mean(hwy))

```

---

# Readr

.left-column[

.center[![](https://www.dropbox.com/s/cqv60z3cznsmr9a/readr.png?raw=1)]
]

.right-column[

- `Readr` lets you load common spreadsheet files (`.csv`, `.tsv`) with simple commands:
- `read_csv(path/to/my_data.csv)`<sup>.red[1]</sup>
    - `read_tsv(path/to/my_data.tsv)`
- Often this is enough, but there are many more customizations as necessary
- You can also export your data from R into a common spreadsheet file with:
- `write_csv(my_dataframe, path=path/to/file_name.csv)`<sup>.red[2]</sup> 
- Read more on the [tidyverse website](https://readr.tidyverse.org/) and the [Readr Cheatsheet](https://rawgit.com/rstudio/cheatsheets/master/data-import.pdf)
]

.footnote[<sup>.red[1]</sup> If you are using an R Project, this file path is in your main project folder, or any subfolders you decided to create, such as `data/my_data.csv`

<sup>.red[2]</sup> where `my_dataframe` is the name of the dataframe you want to export, and `file_name` is the name of the file you want to create.]

---

# Readxl and Haven: When Readr isn't Enough

.left-column[

.center[![](https://www.dropbox.com/s/rl1sj67y2m8fqy5/readxl.png?raw=1)]
.center[![](https://www.dropbox.com/s/ua6t5wz1se34vmo/haven.png?raw=1)]
]

.right-column[

- Some data types, particularly for other software programs like Excel, STATA, SAS, and SPSS are not `.csv` or `.txt` spreadsheets
- `readxl` has commands equivalent to `readr`'s for Excel data types:
    - `read_xls("path/to/my/data.xls")` for older Excel `.xls` files
    - `read_xlsx("path/to/my/data.xlsx")` for modern Excel `.xls` files
    - `write_xls(my_dataframe, path=path/to/file_name.xls)` or ``write_xlsx(my_dataframe, path=path/to/file_name.xlsx)` to export as `xls` or `xlsx` files
- `haven` has commands equivalent to `readr`'s for these data types:
    - `read_dta("path/to/my_data.dta")` for STATA `.dta` files
    - `read_sav("path/to/my_data.sav")` for SPSS `.sav` files
    - `read_sas("path/to/my_data.sas7bdat")` for SAS files 
]

---

# dplyr I

.left-column[

.center[![](https://www.dropbox.com/s/6ao0u4xc3w492ak/dplyr.png?raw=1)]
]

.right-column[

- `dplyr` uses much more efficient and intuitive commands<sup>.magenta[1]</sup> to manipulate dataframes
- Grammatically:
    - `Base R` is all about passively running functions on nouns: `function(object)`
    - `dplyr` is all about actively using verbs: `dplyr_verb(df, conditions)`<sup>.red[2]</sup>
- Three great features:

1. Allows use of `%>%` pipe operator
2. Input and output is always a dataframe<sup>.magenta[3]</sup>
3. Shows the output from a manipulation, but does not save/overwrite as an object unless explicitly assigned to an object

]

.footnote[<sup>.magenta[1]</sup> Particularly when combined with `magrittr`'s pipe (`%>%`) command!

<sup>.red[2]</sup> With the pipe, even simpler: `df %>% dplyr_verb(conditions)`

<sup>.red[3]</sup> Technically a `tibble` with `tidyverse`!]

---

# dplyr II

.left-column[

.center[![](https://www.dropbox.com/s/6ao0u4xc3w492ak/dplyr.png?raw=1)]
]

.right-column[

- Common `dplyr` functions:

| Verb | Does | 
|----------|------|
| `filter` | Keep only selected *observations* | 
| `select` | Keep only selected *variables* |
| `arrange` | Reorder rows (e.g. in numerical order) |
| `mutate` | Create new variables | 
| `recode` | Change a variable's values or categories/`factor` levels |
| `summarize` | Collapse data into summary statistics| 
| `group_by` | Perform any of the above functions by groups/categories |

]

---

# dplyr::filter

- `filter` keeps only selected **observations** (rows)

--

.left-code[

```{r, eval=F}
# look only at African observations
# syntax without the pipe
filter(gapminder, continent=="Africa")

# using the pipe

gapminder %>%
  filter(continent == "Africa")
```

]

--

.right-plot[

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  filter(continent=="Africa")
```

]

---

# dplyr: saving and storing outputs I

- `dplyr` functions never modify their inputs (i.e. never overwrite the original `tibble`)
- If you want to save a result, use `<-` to assign it to a new `tibble`
- If assigned, you will not see the output until you call up the new `tibble` by name

--

.left-code[

```{r, eval = F}
# base syntax
africa <- filter(gapminder, continent=="Africa")

# using the pipe
africa <- gapminder %>%
  filter(continent == "Africa")

# look at new tibble
africa
```
]

.right-plot[
```{r, echo = F}
africa <- gapminder %>%
  filter(continent == "Africa")

# look at new tibble
africa
```
]

---

# dplyr: saving and storing outputs II

- If you want to *both* store and view the output at the same time, wrap the command in parentheses!

```{r}
(africa <- gapminder %>%
  filter(continent == "Africa"))
```

---

# dplyr: saving and storing outputs III

- If you were to assign the output to the original `tibble`, it would *overwrite* the original!

```{r, eval = F}
# base syntax
gapminder <- filter(gapminder, continent=="Africa")

# using the pipe
gapminder <- gapminder %>%
  filter(continent == "Africa")

# this overwrites gapminder!
```

---

# dplyr conditionals

- In many data wrangling contexts, you will want to select data .whisper[conditionally]
    - In computer code, you are selecting only observations for which a set of logical conditions are `TRUE`
- For example:
    - `>`, `<`: greater than, less than
    - `>=`, `<=`: greater than or equal to, less than or equal to
    - `==`, `!=`: is equal to, is not equal to
    - `%in%`: is a member of some defined set $(\in)$
    - `&`: AND<sup>1</sup>
    - `|`: OR

.footnote[<sup>1</sup> Commas also work in `dplyr`.]

---

# dplyr::filter with conditionals

--

.left-code[

```{r, eval=F}
# look only at African observations in 1997
gapminder %>%
  filter(continent == "Africa", year == 1997)
```
]

--

.right-plot[

```{r, echo=F}
gapminder %>%
filter(continent == "Africa", year == 1997)
```
]

---

# dplyr::filter with conditionals II

--

.left-code[

```{r, eval=F}
# look only at African observations or observations in 1997
gapminder %>%
  filter(continent == "Africa" | year == 1997)
```
]

--

.right-plot[

```{r, echo=F}
gapminder %>%
filter(continent == "Africa" | year == 1997)
```
]

---

# dplyr::filter with conditionals III

--

.left-code[

```{r, eval=F}
# look only at U.S. and U.K. observations in 2002
gapminder %>%
  filter(country %in% c("United States", "United Kingdom"), year == 2002)
```
]

--

.right-plot[

```{r, echo=F}
gapminder %>%
  filter(country %in% c("United States", "United Kingdom"), year == 2002)
```
]

---

# dplyr::arrange I

- `arrange` reorders **observations** (rows) in a logical order
    - e.g. alphabetical, numeric, small to large

--

.left-code[

```{r, eval=F}
# order by smallest to largest population
# syntax without the pipe
arrange(gapminder, pop)

# using the pipe

gapminder %>%
  arrange(pop)
```

]

--

.right-plot[

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  arrange(pop)
```

]

---

# dplyr::arrange II

- Break ties in the value of one variable with the values of additional variables

--

.left-code[

```{r, eval=F}
# order by year, with the smallest to largest population in each year
# syntax without the pipe
arrange(gapminder, year, pop)

# using the pipe

gapminder %>%
  arrange(year, pop)
```

]

--

.right-plot[

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  arrange(year, pop)
```

]

---

# dplyr::arrange III

- Use `desc()` to re-order in the opposite direction

--

.left-code[

```{r, eval=F}
# order by largest to smallest population
# syntax without the pipe
arrange(gapminder, desc(pop))

# using the pipe

gapminder %>%
  arrange(desc(pop))
```

]

--

.right-plot[

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  arrange(desc(pop))
```

]

---

# dplyr::select I

- `select` keeps only selected **variables** (columns)
    - Don't need quotes around column names
--

.left-code[

```{r, eval=F}
# keep only country, year, and population variables
# syntax without the pipe
select(gapminder, country, year, pop)

# using the pipe

gapminder %>%
  select(country, year, pop)
```

]

--

.right-plot[

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  select(country, year, pop)
```

]

---

# dplyr::select II

- `select` "all except" by negating a variable with `-`

--

.left-code[

```{r, eval=F}
# keep all except gdpPercap
# syntax without the pipe
select(gapminder, -gdpPercap)

# using the pipe

gapminder %>%
  select(-gdpPercap)
```

]

--

.right-plot[

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  select(-gdpPercap)
```

]

---

# dplyr::select III

- `select` reorders the columns in the order you provide
    - sometimes useful to keep all variables, and drag one or a few  to the front, add `everything()` at the end

--

.left-code[

```{r, eval=F}
# keep all and move pop first
# syntax without the pipe
select(gapminder, pop, everything())

# using the pipe

gapminder %>%
  select(pop, everything())
```

]

--

.right-plot[

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  select(pop, everything())
```

]

---

# dplyr::select IV

- `select` has a lot of helper functions, useful for when you have hundreds of variables
    - see `?select()` for a list

--

.pull-left[

```{r}
# keep all variables starting with "co"

gapminder %>%
  select(starts_with("co"))
```

]

--

.pull-right[

```{r}
# keep country and all variables containing "per"

gapminder %>%
  select(country, contains("per"))
```

]

---

# dplyr::rename

- `rename` changes the name of a variable (column)
    - Format: `new_name=old_name`
--

.left-code[

```{r, eval=F}
# rename gdpPercap to GDP
# syntax without the pipe
rename(gapminder, GDP = gdpPercap)

# using the pipe

gapminder %>%
  rename(GDP = gdpPercap)
```

]

--

.right-plot[

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  rename(GDP = gdpPercap)
```

]

---

# dplyr::mutate

- `mutate` creates a new variable (column)
    - always adds a new column at the end
    - general formula: `new_variable_name = operation`

---

# dplyr::mutate II

- Three major types of mutates:

1. Create a variable that is a specific value (often categorical)

--

.left-code[

```{r, eval=F}
# create variable "europe" if country is in Europe
# syntax without the pipe
mutate(gapminder, europe = ifelse(continent == "Europe", 
                                  yes = "In Europe", 
                                  no = "Not in Europe"))

# using the pipe

gapminder %>%
  mutate(europe = ifelse(continent == "Europe", 
                                  yes = "In Europe", 
                                  no = "Not in Europe"))
```

]

--

.right-plot[

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  mutate(europe = ifelse(continent == "Europe", 
                                  yes = "In Europe", 
                                  no = "Not in Europe"))
```
]

---

# dplyr::mutate III

- Three major types of mutates:

1. Create a variable that is a specific value (often categorical)
2. Change an existing variable (often rescaling)

--

.left-code[

```{r, eval=F}
# create population in millions 
# syntax without the pipe
mutate(gapminder, pop_millions = pop / 1000000)

# using the pipe

gapminder %>%
  rename(pop_millions = pop / 1000000)
```

]

--

.right-plot[

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  mutate(pop_millions = pop / 1000000)
```
]

---

# dplyr::mutate IV

- Three major types of mutates:

1. Create a variable that is a specific value (often categorical)
2. Change an existing variable (often rescaling)
3. Create a variable based on other variables

--

.left-code[

```{r, eval=F}
# create GDP variable from gdpPercap and pop
# syntax without the pipe
mutate(gapminder, GDP = gdpPercap * pop)

# using the pipe

gapminder %>%
  rename(GDP = gdpPercap * pop)
```

]

--

.right-plot[

```{r, echo=F}
# syntax with the pipe 
gapminder %>%
  mutate(GDP = gdpPercap * pop)
```
]

---

# dplyr::mutate V

- Can create multiple new variables with commas:

--

```{r}
gapminder %>%
  mutate(GDP = gdpPercap * pop,
         pop_millions = pop / 1000000)
```

--

- `transmute` keeps *only* newly created variables (`select`s only the new `mutate`d variables)

```{r}
gapminder %>%
  transmute(GDP = gdpPercap * pop,
         pop_millions = pop / 1000000)
```

--

- Don't forget to assign the output to  new `tibble` if you want to "save" the new variables!

---

# Exporting Tables with knitr::kable

---

# dplyr::tally with group_by

- `tally` provides counts, best used with `group_by` for categorical data

```{r}
gapminder %>%
  tally
```

```{r}
gapminder %>%
  group_by(continent) %>%
  tally
```

---

# Advanced: tidyr and joining data

---

# tidyr: reshaping and tidying data

.left-column[

.center[![](https://www.dropbox.com/s/p6gmb4tds18ueyp/tidyr.png?raw=1)]
]

.right-column[

- `tidyr` helps clean data into more usable format for other `tidyverse` packages
- .hl["tidy" data].footnote[This is the core and eponym of the `tidyverse`: all associated packages and functions use or require this data format!] are (an opinionated view of) data where

1. Each .shout[variable] is in a .shout[column]
2. Each .whisper[observation] is a .whisper[row]
3. Each .onfire[value] is a .onfire[cell]

- Un-tidy data needs to be tidied into this format
    - You will spend less time fighting tools and more time on analysis!

]

---

# Tidying Data: `gather()` wide to long I

.pull-left[

```{r, echo=F}
ex_wide<-tribble(
  ~"Country", ~"2000", ~"2010",
  "United States", 140, 180,
  "Canada", 102, 98,
  "China", 111, 123
)
```

```{r}
ex_wide
```

]

.pull-right[

- Common source of "un-tidy" data: .l[Column headers are values, not variable names]
    - Column names are *values* of a `year` variable!
    - Each row represents *two* observations (one in 2000 and one in 2010)!
- We need to `gather()` these columns into a new pair of variables
    - set of columns that represent values, not variables (`2000` and `2010`)
    - `key`: name of variable whose values form the column names (we'll call it the `year`)
    - `value`: name of the variable whose values are spread over the cells (we'll call it number of `cases`)
]

---

# Tidying Data: `gather()` wide to long II

- `gather()` a wide data frame into a long data frame

.pull-left[

```{r}
ex_wide
```
]

.pull-right[

```{r}
ex_wide %>% gather("2000","2010",
              key = "year",
              value = "cases")
```

]

---

# Tidying Data: `spread()` long to wide I

.pull-left[

```{r, echo=F}
ex_long<-tribble(
  ~"Country", ~"Year", ~"Type", ~"Count",
  "United States", 2000, "Cases", 140,
  "United States", 2000, "Population", 300,
  "United States", 2010, "Cases", 180,
  "United States", 2010, "Population", 310,
  "Canada", 2000, "Cases", 102,
  "Canada", 2000, "Population", 110,
  "Canada", 2010, "Cases", 98,
  "Canada", 2010, "Population", 121,
  "China", 2000, "Cases", 111,
  "China", 2000, "Population", 1201,
  "China", 2010, "Cases", 123,
  "China", 2010, "Population", 1241
)

```

```{r}
ex_long
```

]

.pull-right[

- Common source of "un-tidy" data: .l[observations are scattered across multiple rows]
    - Each country has two rows per observation, one for `Cases` and one for `Population` (categorized by `type` of variable)
- We need to `spread()` these columns into a new pair of variables
    - `key`: column that contains variable names (here, the `type`)
    - `value`: column that contains values from multiple variables (here, the `count`)
]

---

# Tidying Data: `spread()` long to wide II

- `spread()` a long data frame into a wide data frame

.pull-left[

```{r}
ex_long
```
]

.pull-right[

```{r}
ex_long %>% spread(key = "Type",
                   value = "Count")
```

]

---

# Combining Datasets

- Often, data doesn't come from just one source, but several sources
- We can combine datasets into a single dataframe (tibble) using `dplyr` commands in several ways:
    1. `bind` dataframes together by row or by column
        - `bind_rows()` adds observations (rows) to existing dataset<sup>.magenta[1]</sup>
        - `bind_cols()` adds variables (columns) to existing dataset<sup>.magenta[2]</sup>
    2. `join` two existing dataframes by designating a variable(s) as a `key` to match rows by identical values of that `key`

.footnote[<sup>.magenta[1]</sup> Note the columns must be identical between the original dataset and the new observations

<sup>.magenta[1]</sup> Note the rows must be identical between original dataset and new variable]

---

# Two *Similar* Datasets

- Sometimes you want to add rows (observations) or columns (variables) that happen to match up perfectly
    - New observations contain all the same variables as existing data
    - OR
    - New variables contain all the same observations as existing data
- In this case, simply using `bind_*(old, new)` will work
    - `bind_columns(old, new)` adds columns from `new` to `old` dataframe
    - `bind_rows(old, new)` adds rows from `new` to `old` dataframe

---

# Two *Different* Datasets

.pull-left[

- For the following examples, consider the following two dataframes, `x` and `y`<sup>*</sup>
    - each has one unique variable, `x$x` and `y$y`
    - both have values for observations `1` and `2`
    - `x` has observation `3` which `y` does not have
    - `y` has observation `4` which `x` does not have
- We next consider the ways we can merge dataframes `x` and `y` into a single dataframe

]

.pull-right[

![](https://raw.githubusercontent.com/gadenbuie/tidyexplain/master/images/static/png/original-dfs.png)

]

.footnote[<sup>*</sup> Images on all following slides come from Gaden Buie's excellent [tidyexplain](https://github.com/gadenbuie/tidyexplain)]

---

# Inner-Join

.pull-left[

- Merge columns from `x` and `y` for which there are matching rows
    - Rows in `x` with no match in `y` (3) will be dropped
    - Rows in `y` with no match in `x` (4) will be dropped

]

.pull-right[

![](https://raw.githubusercontent.com/gadenbuie/tidyexplain/master/images/inner-join.gif)

]
---

# Left-Join

.pull-left[

- Start with all rows from `x` and add all columns from `y`
    - Rows in `x` with no match in `y` (3) will have `NA`s
    - Rows in `y` with no match in `x` (4) will be dropped

]

.pull-right[

![](https://raw.githubusercontent.com/gadenbuie/tidyexplain/master/images/left-join.gif)

]

---

# Right-Join

.pull-left[

- Start with all rows from `y` and add all columns from `x`
    - Rows in `y` with no match in `x` (4) will have `NA`s
    - Rows in `x` with no match in `y` (3) will be dropped

]

.pull-right[

![](https://raw.githubusercontent.com/gadenbuie/tidyexplain/master/images/right-join.gif)

]

---

# Full-Join

.pull-left[

- All rows and all columns from `x` and `y`
    - Rows that do not match (3 and 4) will have `NA`s

]

.pull-right[

![](https://raw.githubusercontent.com/gadenbuie/tidyexplain/master/images/full-join.gif)

]

### 2019-09-06 content/assignment/00-preliminary-survey.Rmd
Add links to survey answers (testing icons)

@@
 - [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/00-preliminary-survey.pdf)
+- [<i class="far fa-life-ring"></i><i class="fas fa-file-pdf"></i> `Answer Key`](/homeworks/00-preliminary-survey-answers.html)
+- [<i class="far fa-life-ring"></i><i class="fas fa-external-link-square-alt"></i> `Answer Key`](/homeworks/00-preliminary-survey-answers.pdf)
 

### 2019-09-06 content/assignment/00-preliminary-survey.Rmd
Fixed icons for survey answer key

--- 2019-08-27 content/assignment/00-preliminary-survey.Rmd
@@
 - [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/00-preliminary-survey.pdf)
+- [<i class="fas fa-external-link-square-alt"></i> `Answer Key`](/homeworks/00-preliminary-survey-answers-pdf.pdf)
+- [<i class="fas fa-file-pdf"></i> `Answer Key`](/homeworks/00-preliminary-survey-answers.html)
 

### 2019-09-06 content/assignment/00-preliminary-survey.Rmd
Add answer keys (web, pdf) for R practice 1.3

--- 2019-08-27 content/assignment/00-preliminary-survey.Rmd
@@
 - [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/00-preliminary-survey.pdf)
+- [<i class="fas fa-file-pdf"></i> `Answer Key (PDF)`](/homeworks/00-preliminary-survey-answers.html)
+- [<i class="fas fa-external-link-square-alt"></i> `Answer Key (html)`](/homeworks/00-preliminary-survey-answers-pdf.pdf)
 

### 2019-09-06 content/r/03-r-practice.Rmd
Add answer keys (web, pdf) for R practice 1.3

@@
 - [<i class="fas fa-file-pdf"></i> `Download as PDF`](/r/03-r-practice.pdf)
-- [<i class="fas fa-file-pdf"></i> `Answer Key`](/r/03-r-practice-answers.pdf)
+- [<i class="fas fa-external-link-square-alt"></i> `Answer Key (html)`](r/03-r-practice-answers.html)
+- [<i class="fas fa-file-pdf"></i> `Answer Key (PDF)`](/r/03-r-practice-answers-pdf.pdf)
 - [<i class="fas fa-registered"></i> `R Script`](/r/03-r-practice-answers.R)

### 2019-09-06 static/r/03-r-practice-answers.Rmd
Add answer keys (web, pdf) for R practice 1.3

@@
 date: "ECON 480 - Fall 2019"
-output: pdf_document
+output:
+  html_document:
+    df_print: paged
+    theme: simplex
+    toc: true 
+    toc_depth: 3
+    toc_float: true
+    code_folding: show
+    highlight: tango
 ---
@@
 
-## 1. Work on the following parts:
+## 1.
 
-a. Create a vector called `me` with two objects, your first name, and your last name.
+#### Work on the following parts:
+
+### a.
+#### Create a vector called `me` with two objects, your first name, and your last name.
 
@@
 
-b. Call the vector to inspect it.
+### b.
+#### Call the vector to inspect it.
 
@@
 
-c. Confirm it is a character class vector.
+### c.
+#### Confirm it is a character class vector.
 
@@
 
-## 2. Use R's help functions to determine what the `paste()` function does. Then paste together your first name and last name.
+## 2.
+#### Use R's help functions to determine what the `paste()` function does. Then paste together your first name and last name.
 
@@
 ```
-## 3. Create a vector called `my_vector` with all the even integers from 2 to 10.
+## 3.
+#### Create a vector called `my_vector` with all the even integers from 2 to 10.
 
@@
 
-## 4. Find the mean of `my_vector` with `mean()`.
+## 4.
+#### Find the mean of `my_vector` with `mean()`.
 
@@
 
-## 5. Take all the integers from 18 to 763,^[Hint: use the `:` operator to create a sequence from a starting number to an ending number] then get the mean.
+## 5.
+#### Take all the integers from 18 to 763,^[Hint: use the `:` operator to create a sequence from a starting number to an ending number] then get the mean.
 
@@
 
-### 6. Install `ggplot2`.
+### 6. 
+#### Install `ggplot2`.
 
@@
 
-### 7. Load `ggplot2` with the `library()` command.
+### 7. 
+#### Load `ggplot2` with the `library()` command.
 
@@
 
-### 8. Get the `str`ucture of the diamonds data frame. What are the different variables and what kind of data does each contain?
+### 8.
+#### Get the `str`ucture of the diamonds data frame. What are the different variables and what kind of data does each contain?
 
@@
 
-### 9. Get summary statistics separately for `carat`, `depth`, `table`, and `price`.
+### 9. 
+#### Get summary statistics separately for `carat`, `depth`, `table`, and `price`.
 
@@
 
-### 10. `color`, `cut`, and `clarity` are categorical variables (factors). Use the `table()` command to generate frequency tables for each.
+### 10. 
+#### `color`, `cut`, and `clarity` are categorical variables (factors). Use the `table()` command to generate frequency tables for each.
 
@@
 
-### 11. Now rerun the `summary()` command on the entire data frame.
+### 11. 
+#### Now rerun the `summary()` command on the entire data frame.
 
@@
 
-### 12. Now look only at (subset) the first 4 diamonds in the dataset.
+### 12. 
+#### Now look only at (subset) the first 4 diamonds in the dataset.
 
@@
 
-### 13. Now look only at (subset) the third and seventh diamond in the dataset.
+### 13. 
+#### Now look only at (subset) the third and seventh diamond in the dataset.
 
@@
 
-### 14. Now look only at (subset) the second column of the dataset.
+### 14. 
+#### Now look only at (subset) the second column of the dataset.
 
@@
 ```
-### 15. Do this again, but look using the `$` to pull up the second column by name.
+### 15. 
+#### Do this again, but look using the `$` to pull up the second column by name.
 
-```{r}
+```{r, eval=F}
 # second column is called "cut"
-diamonds$cut
+diamonds$cut # dont' run this, it'll print 53,000 rows!
 ```
 
-### 16. Now look only at diamonds that have a `carat` greater than or equal to 1.
+### 16. 
+#### Now look only at diamonds that have a `carat` greater than or equal to 1.
 
@@
 
-### 17. Now look only at diamonds that have a VVS1 `clarity`.
+### 17. 
+#### Now look only at diamonds that have a VVS1 `clarity`.
 
@@
 
-### 18. Now look only at dimaonds that have a color of E, F, I, and J.
+### 18. 
+#### Now look only at dimaonds that have a color of E, F, I, and J.
 
@@
 
-### 19. Now look only at diamonds that have a `carat` greater than or equal to 1 and a VVS1 `clarity`.
+### 19.
+#### Now look only at diamonds that have a `carat` greater than or equal to 1 and a VVS1 `clarity`.
 
@@
 
-### 20. Get the average price of diamonds in question 18.^[Hints: use your subset command as an argument to the mean function. You will not need a comma here because you are looking for a single row.]
+### 20.
+#### Get the average price of diamonds in question 18.^[Hints: use your subset command as an argument to the mean function. You will not need a comma here because you are looking for a single row.]
 
@@
 
-### 21. What is the highest price for a diamond with a 1.0 carat, D color, and VVS1 clarity?
+### 21.
+#### What is the highest price for a diamond with a 1.0 carat, D color, and VVS1 clarity?
 
@@
 # second column is called "cut"
-diamonds$cut
+# diamonds$cut dont' run this, it'll print 53,000 rows!
 

### 2019-09-06 content/r/03-r-practice.Rmd
Fixed link in R practice 1.3 html answers

--- 2019-09-06 content/r/03-r-practice.Rmd
@@
 - [<i class="fas fa-file-pdf"></i> `Download as PDF`](/r/03-r-practice.pdf)
-- [<i class="fas fa-file-pdf"></i> `Answer Key`](/r/03-r-practice-answers.pdf)
+- [<i class="fas fa-external-link-square-alt"></i> `Answer Key (html)`](/r/03-r-practice-answers.html)
+- [<i class="fas fa-file-pdf"></i> `Answer Key (PDF)`](/r/03-r-practice-answers-pdf.pdf)
 - [<i class="fas fa-registered"></i> `R Script`](/r/03-r-practice-answers.R)

### 2019-09-06 static/files/syllabus-pdf.Rmd
Add syllabus PDF

--- 2019-08-22 content/syllabus.Rmd
@@
 ---
-title: Syllabus
-slug: syllabus
-citeproc: false
-bibliography: ../static/bib/references.bib
-csl: ../static/bib/chicago-syllabus-no-bib.csl
+output: 
+  pdf_document:
+    number_sections: false
+    keep_tex: true
+    fig_caption: yes
+    df_print: kable
+    latex_engine: xelatex
+    template: template-syllabus.tex
+    includes:
+      in_header: 
+geometry: margin=1in
+
+course: "Econometrics"
+coursecode: "ECON 480"
+section: ""
+author: Dr. Ryan Safner
+semester: "[metricsF19.classes.ryansafner.com](http://metricsf19.classes.ryansafner.com)"
+
+email: "safner@hood.edu"
+web: "ryansafner.com"
+officehours: "TuTh 10:30--11:30 A.M."
+office: "Rosenstock Hall 118"
+classroom: "Rosenstock Trading Room"
+classhours: "TuTh 2:00 P.M.--3:15 P.M."
+
+fontfamily: Fira Sans
+mainfont: Fira Sans
+fontsize: 12pt
+sansfont: "Yanone Kaffeesatz"
+monofont: "Fira Code"
+mathfont: "Fira Sans"
+
+# bibliography: 306syllabus.bib
 ---
@@
 
-- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/files/syllabus.pdf)
-- **Last Updated**: 2019-08-22
+> "There are three kinds of lies: lies, damned lies, and statistics." -- Benjamin Disraeli
 
-<span class="newthought">Econometrics</span> [^courseinfo] is the application of statistical tools to quantify and measure economic relationships in the real world.
+**Econometrics** is the application of statistical tools to quantify and measure economic relationships in the real world.
 It uses real data to test economic hypotheses, quantitatively estimate causal relationships between economic variables, and to make forecasts of future events.
@@
 
-[^courseinfo]: {-}
-  `r blogdown::shortcode("courseinfo")`
-
 # Course objectives
 
-<span class="newthought">By the end of this course,</span> you will:
+**By the end of this course,** you will:
+
+1. apply the models of microeconomics (constrained optimization and equilibrium) towards explaining real world behavior of individuals, firms, and governments
+2.
+explore the effects of economic and political processes on market performance (competition, market prices, profits and losses, property rights, entrepreneurship, market power, market failures, public policy, government failures)
+3. apply the economic way of thinking to real world issues in writing
 
@@
 
-See my [tips for success in this course](/reference#tips).
+See my [tips for success in this course](http://metricsf19.classes.ryansafner.com/reference#tips).
 
 # Course materials
+
+You can find all course materials at my academic website dedicated for (all sections of) this course [microf19.classes.ryansafner.com](http://microf19.classes.ryansafner.com).
+
+My lecture slides will be shared with you, and serve as your primary resource, but our main "textbook" below is **recommended** as the next best resource and will be available from the campus bookstore.
 
@@
 
-<span class="newthought">You can find descriptions</span> for all the assignments on the [assignments page](/assignments/).
+**You can find descriptions** for all the assignments on the [assignments page](http://metricsf19.classes.ryansafner.com/assignments/).
 
@@
 ```
-
-```{r, echo=F}
-knitr::kable(assignments)
+\begin{center}
+```{r, echo = F, results="asis"}
+knitr::kable(assignments, format="latex", booktabs=TRUE)
 ```
-
+\end{center}
 ```{r show-assignments-table, eval=F, echo=FALSE, results="asis"}
@@
 
+\begin{center}
 ```{r show-grades-table, echo=FALSE, results="asis"}
 grading %>%
-  pandoc.table(justify = "lclc", 
-               col.names = c("Grade", "Range", "Grade", "Range"))
+  knitr::kable(., format="latex", booktabs=TRUE)
 ```
+\end{center}
 
@@
 
-Watch this excellent and accurate video explaining office hours:
-
-<iframe src="https://player.vimeo.com/video/270014784" width="640" height="360" frameborder="0" allow="autoplay; fullscreen" allowfullscreen></iframe>
-
-### Van Halen and M&Ms
-
-When you have completed reading the syllabus, email me a picture of the band Van Halen and a picture of a bowl of M&Ms.
-If you do this *before* the date of the first exam, you will get 3 bonus points on the exam.
-Yes, this is real.
+Watch the excellent and accurate video explaining office hours (on website syllabus page):
 
@@
 See [here](http://hood.smartcatalogiq.com/en/2016-2017/Catalog/The-Spirit-of-Hood/The-Academic-Honor-Code-and-Code-of-Conduct) for more detailed information.
+
+### Van Halen and M&Ms
+
+When you have completed reading the syllabus, email me a picture of the band Van Halen and a picture of a bowl of M&Ms.
+If you do this *before* the date of the first exam, you will get 3 bonus points on the exam.
+Yes, this is real.
 
@@
 
-<span class="newthought">You can find a full schedule</span> with resources for each class meeting on the [schedule page](/schedule/).
+**You can find a full schedule** with resources for each class meeting on the [course website schedule page](http://metricsF19.classes.ryansafner.com/schedule/).

### 2019-09-06 static/files/syllabus-pdf.tex
Add syllabus PDF

\documentclass{article}
\usepackage{booktabs, graphicx, hyperref, fontspec}
\usepackage{sectsty}
\allsectionsfont{\sffamily}
\usepackage[margin=1in]{geometry}
\hypersetup{
  colorlinks = true,
  urlcolor = cyan,
 }
 \providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\newcommand*{\authorfont}{\fontfamily{phv}\selectfont}
\usepackage[]{Fira Sans}

\begin{document}

\sffamily

\centerline{\Huge Econometrics}

\vspace{3 mm}

\centerline{\large Dr.~Ryan Safner}
\vspace{2 mm}
\centerline{\large \href{http://metricsf19.classes.ryansafner.com}{metricsF19.classes.ryansafner.com}}

\vspace{5 mm}

\begin{tabular}{@{}p{3.5in}p{3.5in}}           
\textbf{Course}: ECON 480  & \textbf{Email:}  \href{mailto:safner@hood.edu}{\nolinkurl{safner@hood.edu}}\\
\textbf{Room}: Rosenstock Trading Room & \textbf{Office:}  Rosenstock Hall 118\\
\textbf{Meets}: TuTh 2:00 P.M.--3:15 P.M. & \textbf{Hours:} TuTh 10:30--11:30 A.M.\\ 
\end{tabular}

\vspace{5 mm}

\hrule

\begin{quote}
``There are three kinds of lies: lies, damned lies, and statistics.'' --
Benjamin Disraeli
\end{quote}

\textbf{Econometrics} is the application of statistical tools to
quantify and measure economic relationships in the real world. It uses
real data to test economic hypotheses, quantitatively estimate causal
relationships between economic variables, and to make forecasts of
future events. The primary tool that economists use for empirical
analysis is ordinary least squares (OLS) linear regression, so the
majority of this course will focus on understanding, applying, and
extending OLS regressions.

I assume you have \emph{some} working knowledge of economics at the
intermediate level and some basic statistical tools.\footnote{The formal
  prerequisites for this course are \textbf{ECON 205} and \textbf{ECON
  206}; \textbf{ECMG 212} or \textbf{MATH 112}; and \textbf{ECON 305} or
  \textbf{ECON 306}} We will do some basic review of some necessary
statistics and probability at the beginning until everyone is
comfortable, before jumping right into regressions.

\hypertarget{course-objectives}{%
\section{Course objectives}\label{course-objectives}}

\textbf{By the end of this course,} you will:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  apply the models of microeconomics (constrained optimization and
  equilibrium) towards explaining real world behavior of individuals,
  firms, and governments
\item
  explore the effects of economic and political processes on market
  performance (competition, market prices, profits and losses, property
  rights, entrepreneurship, market power, market failures, public
  policy, government failures)
\item
  apply the economic way of thinking to real world issues in writing
\item
  understand how to evaluate statistical and empirical claims;
\item
  use the fundamental models of causal inference and research design;
\item
  gather, analyze, and communicate with real data in R.
\end{enumerate}

I am less concerned with forcing you to memorize and recite proofs of
statistical estimator properties, and more concerned with the
development of your intuitions and the ability to think critically as an
empirical social scientist---although this will require you to
demonstrate proficiency with some intermediate statistical and
mathematical tools.

Given these objectives, this course fulfills all three of the learning
outcomes for
\href{https://www.hood.edu/academics/departments/george-b-delaplaine-jr-school-business/student-learning-outcomes}{the
George B. Delaplaine, Jr.~School of Business} Economics B.A. program:

\begin{itemize}
\tightlist
\item
  Use quantitative tools and techniques in the preparation,
  interpretation, analysis and presentation of data and information for
  problem solving and decision making {[}\ldots{]}
\item
  Apply economic reasoning and models to understand and analyze problems
  of public policy {[}\ldots{]}
\item
  Demonstrate effective oral and written communications skills for
  personal and professional success{[}\ldots{]}
\end{itemize}

{Fair warning:} \textbf{Econometrics is hard.} \emph{It will be one of
the hardest economics courses that you will take, primarily due to the
mathematical content.} I will do my best to make this class intuitive
and helpful, if not interesting. If at any point you find yourself
struggling in this course for any reason, please come see me. Do not
suffer in silence! Coming to see me for help does not diminish my view
of you, in fact I will hold you in \emph{higher} regard for
understanding your own needs and taking charge of your own learning.
There are also a some fantastic resources on campus, such as the
\href{http://www.hood.edu/campus-services/academic-services/index.html}{Center
for Academic Achievement and Retention (CAAR)} and the
\href{http://www.hood.edu/library/\%7D\%7BBeneficial-Hodson\%20Library}{Beneficial-Hodson
Library}.

See my
\href{http://metricsf19.classes.ryansafner.com/reference\#tips}{tips for
success in this course}.

\hypertarget{course-materials}{%
\section{Course materials}\label{course-materials}}

You can find all course materials at my academic website dedicated for
(all sections of) this course
\href{http://microf19.classes.ryansafner.com}{microf19.classes.ryansafner.com}.

My lecture slides will be shared with you, and serve as your primary
resource, but our main ``textbook'' below is \textbf{recommended} as the
next best resource and will be available from the campus bookstore.

\hypertarget{books}{%
\subsection{Books}\label{books}}

The following book is \textbf{required}\footnote{You are not
  \emph{obligated} to buy it, I just \textbf{strongly recommend} it in
  the sense that you will still have access to all data and assignments
  without possessing the book. But this is a course where you really
  will want to understand the derivations or get additional context
  beyond just my slides\ldots{}} and will be available from the campus
bookstore.

\begin{itemize}
\tightlist
\item
  Bailey, Michael A, 2019, \emph{Real Econometrics}, New York: Oxford
  University Press, 2\textsuperscript{nd} ed.
\end{itemize}

You are welcome to purchase the book by other means (e.g.~Amazon,
half.com, etc). I have no financial stake in requiring you to purchase
this book. The (cheaper) 1st edition is sufficient, but makes
significantly less use of \texttt{R} (in favor of \texttt{STATA}).

The following two books are \textbf{recommended}, and are free
online\footnote{You can purchase a hard copy of the first one if you
  really want.}:

\begin{itemize}
\tightlist
\item
  Grolemund, Garrett and Hadley Wickham,
  \href{https://r4ds.had.co.nz/}{\emph{R For Data Science}}
\item
  Ismay, Chester and Albert Y Kim,
  \href{https://moderndive.com/}{\emph{Modern Dive: Statistical
  Inference Via Data Science}}
\end{itemize}

The first book is the number one resource for using \texttt{R} and
\texttt{tidyverse}, and is written for beginners. I still look at it
\emph{frequently.} The second is another great reference for using
\texttt{tidyverse} in the context of basic statistics.

\hypertarget{software}{%
\subsection{Software}\label{software}}

You are \textbf{strongly recommended} to download copies of
\href{https://www.r-project.org/}{\texttt{R}} and
\href{http://www.rstudio.com}{\texttt{R\ Studio}} on your own computers.
These software packages are available on all computers in the trading
room, and you will have access to them during the week to work on
assignments.

We will also have a shared class workspace in
\href{http://rstudio.cloud/}{RStudio.cloud} that runs a full instance of
R Studio in your web browser (so no need to install anything!) will let
you access files and assignments.

\hypertarget{articles}{%
\subsection{Articles}\label{articles}}

Throughout the course, I will post both required and supplemental
(non-required) readings that enrich your understanding for each topic.
Check Blackboard \emph{frequently} for announcements and updates to
assignments, readings, and grades.

\hypertarget{assignments-and-grades}{%
\section{Assignments and grades}\label{assignments-and-grades}}

\textbf{You can find descriptions} for all the assignments on the
\href{http://metricsf19.classes.ryansafner.com/assignments/}{assignments
page}.

\begin{center}

\begin{tabular}{lll}
\toprule
 & Assignment & Percent\\
\midrule
1 & Research Project & 30\%\\
n & Problem sets (Average) & 25\%\\
1 & Midterm & 20\%\\
1 & Final & 25\%\\
\bottomrule
\end{tabular}
\end{center}

All grades are based on the following traditional scale:

\begin{center}

\begin{tabular}{llll}
\toprule
Grade & Range & Grade1 & Range1\\
\midrule
A & 93–100\% & C & 73–76\%\\
A− & 90–92\% & C− & 70–72\%\\
B+ & 87–89\% & D+ & 67–69\%\\
B & 83–86\% & D & 63–66\%\\
B− & 80–82\% & D− & 60–62\%\\
\addlinespace
C+ & 77–79\% & F & < 60\%\\
\bottomrule
\end{tabular}
\end{center}

These grades are firm cutoffs, but I do of course round upwards
(\(\geq 0.5\)) for final grades. A necessary reminder, as an academic, I
am not in the business of \emph{giving} out grades, I merely report the
grade that you \emph{earn}. I will not alter your grade unless you
provide a reasonable argument that I am in error (which does happen from
time to time).

\hypertarget{policies-and-expectations}{%
\section{Policies and Expectations}\label{policies-and-expectations}}

This syllabus is a contract between you, the student, and me, your
instructor. It has been carefully and deliberately thought out\footnote{A
  syllabus can and will be used as a legal document for disputes tried
  at a court of law. Ask me how I know.}, and I will uphold my end of
the agreement and expect you to uphold yours.

In the language of game theory, this syllabus is my commitment device. I
am a very understanding person, and I know that exceptions to rules
often need to be made for students. However, to be \emph{fair} to
\emph{all} students the syllabus artificially constrains my ability to
make exceptions at a whim for anyone. This prevents clever students from
exploiting my congenial personality at everyone else's expense. Please
read and familiarize yourself with the course policies and expectations
of you. Chances are, if you have a question, it is answered herein.

\hypertarget{attendance-and-participation}{%
\subsubsection{Attendance and
Participation}\label{attendance-and-participation}}

I expect you to attend class and to come having already done the reading
assigned for that day. I will remind you in class and possibly through
Blackboard or email which readings I want you to read for the next
class. You are all adults and I will treat you as such. I do not take
attendance, nor do I grade formally for participation but I strongly
recommended you attend class and participate for your sake and the sake
of your classmates. If you are too distracted or are not prepared to
learn, I suggest you stay home, where you can check Facebook more
efficiently. I reserve the right to boost the final grades of students
that I believe have made consistent, quality contributions above and
beyond their peers in class conversations by up to 2.5 points.

\hypertarget{devices}{%
\subsubsection{Devices}\label{devices}}

You are allowed to have and use laptops and tablets in the classroom. I
will not stop you, but I strongly discourage you from using these to
take notes (see Tips for Success). As a courtesy to myself and to
others, do not use your phone in class. I reserve the right to embarrass
you in front of everyone if you do so.

\hypertarget{absences-and-make-ups}{%
\subsubsection{Absences and Make-Ups}\label{absences-and-make-ups}}

You generally do \emph{not} need to let me know if you are unable to
make class, \emph{unless} it is on the day of an exam. It will however,
be your responsibility to acquire the notes from a classmate for any
missed classes. If you are unable to attend an exam for a legitimate
reason (e.g.~sports/club events, traveling, illness, family issues),
please notify me at least \emph{one week} in advance, and we will
schedule a make-up exam date. If you are ill or otherwise unable to
attend on the day of the exam, contact me ASAP to make arrangements.
Failure to do so, including desperate attempts to make arrangements only
\emph{after} the absence will result in a grade of 0 and little
sympathy. I reserve the right to re-weight other assignments for
students who I believe are legitimately unable to complete a particular
assignment.

\hypertarget{late-assignments}{%
\subsubsection{Late Assignments}\label{late-assignments}}

I will accept late assignments, but will subtract a specified amount of
points as a penalty. See individual assignment descriptions for the
amount of points taken off (as it varies by assignment). If an answer
key is posted before you turn in your assignment, the maximum grade you
can earn is an 80. Even if it is the last week of the semester, I
encourage you to turn in late work: some points are better than no
points!

\hypertarget{grading}{%
\subsubsection{Grading}\label{grading}}

I will try my best to post grades on Blackboard's Grading Center and
return graded assignments to you within about one week of you turning
them in. There will be exceptions. Where applicable, I will post answer
keys once I know most homeworks are turned in (see Late Assignments
above for penalties). Blackboard's Grading Center is the place to look
for your most up-to-date grades. You will also be given an Excel
spreadsheet template where you can calculate your overall grade and
forecast ``what if'' scenarios.

\hypertarget{email-accounts}{%
\subsubsection{Email Accounts}\label{email-accounts}}

Students must regularly monitor their Hood email accounts to receive
important college information, including messages related to this class.
Email through the Blackboard system is my main method of communicating
announcements and deadlines regarding your assignments. \textbf{Please
do not reply to any automated Blackboard emails - I may not recieve
it!}. My Hood email
(\href{mailto:safner@hood.edu}{\nolinkurl{safner@hood.edu}}) is the best
means of contacting me. I will do my best to respond within 24 hours. If
I do not reply within 48 hours, do not take it personally, and
\emph{feel free to send a follow up email} in the very likely event that
I genuinely did not see your original message.

\hypertarget{office-hours}{%
\subsubsection{Office Hours}\label{office-hours}}

I am generally in my office Monday-Thursday during ``normal business
hours.'' You are always welcome to walk-in and chat about class,
college, careers, or anything at all. Please do try to use the official
office hours stated at the head of the syllabus if possible. If you need
to meet at a different time, I request that you send me an email or let
me know after class so I know when to expect you. If you want to go over
material from class, please have \emph{specific} questions you want help
with. I am not in the business of giving private lectures (particularly
if you missed class without a valid excuse).

Watch the excellent and accurate video explaining office hours (on
website syllabus page):

\hypertarget{enrollment}{%
\subsubsection{Enrollment}\label{enrollment}}

Students are responsible for verifying their enrollment in this class.
The last day to add or drop this class with no penalty is
\textbf{Tuesday, September 4}. Be aware of
\href{https://www.hood.edu/offices-services/registrars-office/academic-calendar}{important
dates}.

\hypertarget{honor-code}{%
\subsubsection{Honor Code}\label{honor-code}}

Hood College has an Academic Honor Code which requires all members of
this community to maintain the highest standards of academic honesty and
integrity. Cheating, plagiarism, lying, and stealing are all prohibited.
All violations of the Honor Code are taken seriously, will be reported
to appropriate authority, and may result in severe penalties, including
expulsion from the college. See
\href{http://hood.smartcatalogiq.com/en/2016-2017/Catalog/The-Spirit-of-Hood/The-Academic-Honor-Code-and-Code-of-Conduct}{here}
for more detailed information.

\hypertarget{van-halen-and-mms}{%
\subsubsection{Van Halen and M\&Ms}\label{van-halen-and-mms}}

When you have completed reading the syllabus, email me a picture of the
band Van Halen and a picture of a bowl of M\&Ms.~If you do this
\emph{before} the date of the first exam, you will get 3 bonus points on
the exam. Yes, this is real.

\hypertarget{accessibility-equity-and-accommodations}{%
\subsubsection{Accessibility, Equity, and
Accommodations}\label{accessibility-equity-and-accommodations}}

College courses can, and should, be challenging and bring you out of
your comfort zone in a safe and equitable environment. If, however, you
feel at any point in the semester that certain assignments or aspects of
the course will be disproportionately uncomfortable or burdensome for
you due to any factor beyond your control, please come see me or email
me. I am a very understanding person and am happy to work out a solution
together. I reserve the right to modify and reweight assignments at my
sole discretion for students that I belive would legitimately be at a
disadvantage, through no fault of their own, to complete them as
described.

If you are unable to afford required textbooks or other resources for
any reason, come see me and we can find a solution that works for you.

This course is intended to be accessible for all students, including
those with mental, physical, or cognitive disabilities, illness,
injuries, impairments, or any other condition that tends to negatively
affect one's equal access to education. If at any point in the term, you
find yourself not able to fully access the space, content, and
experience of this course, you are welcome to contact me to discuss your
specific needs. I also encourage you to contact the
\href{https://www.hood.edu/academics/josephine-steiner-center-academic-achievement-retention/accessibility-services}{Office
of Accessibility Services} (301-696-3421). If you have a diagnosis or
history of accommodations in high school or previous postsecondary
institutions, Accessibility Services can help you document your needs
and create an accommodation plan. By making a plan through Accessibility
Services, you can ensure appropriate accommodations without disclosing
your condition or diagnosis to course instructors.

\hypertarget{schedule}{%
\section{Schedule}\label{schedule}}

\textbf{You can find a full schedule} with resources for each class
meeting on the
\href{http://metricsF19.classes.ryansafner.com/schedule/}{course website
schedule page}.

\end{document}

### 2019-09-10 content/r/05-r-practice.Rmd
Finalize 5 slides and 5 practice

---
title: "1.5 Data Wrangling with tidyverse - R Practice"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

# Getting Set Up

Before we begin, start a new file with `File` $\rightarrow$ `New File` $\rightarrow$ `R Script`. As you work through this sheet in the console in `R`, also add (copy/paste) your commands that work into this new file. At the end, save it, and run to execute all of your commands at once.

First things first, load `tidyverse`.

```{r}
library("tidyverse")
```

# Warm Up to `dplyr` With `gapminder`

## 
#### 1. Load `gapminder`. `select()` the variables `year`, `lifeExp`, `country`.

## 
#### 2. `select()` all variables *except* `pop`.

## 
#### 3. `rename()` `continent` to `cont`.

## 4. 
#### `arrange()` by year.

## 5. 
#### `arrange()` by year, but in descending order.

## 6. 
#### `arrange()` by year, then by life expectancy.

## 7. 
#### `filter()` observations with `pop` greater than 1 billion.

## 8. 
#### Of those, look only at `India`.

## 9.
#### Try out the pipe (`%`) if you haven't already, by chaining commands: `select()` your data to look only at `year`, `gdpPercap`, and `country` in the year `1997`, for countries that have a `gdpPercap` greater than 20,000, and `arrange()` them alphabetically.

## 10.
#### `mutate()` a new variable called `GDP` that is equal to `gdpPercap * pop`.

## 11.
#### `mutate()` a new population variable that is the `pop` in millions.

## 12.
#### `summarize()` to get the average GDP per capita.

## 13.
#### Get the number of observations, average, minimum, maximum, and standard deviation for GDP per capita.

## 14.
#### Get the average GDP per capita over time. Hint, first `group_by()` year.

## 15.
#### Get the average GDP per capita by continent.

## 16.
#### Get the average GDP per capita by year and by continent.^[Hint: do `year` first, if you do `continent` first, there are no years to group by!] Then save this as another `tibble` called `gdp`. Create a `ggplot` of a `line` graph of average continent GDP over time using the `gdp` data.

## 17.
#### Try it again all in one command with the pipe `%>%`. Instead of saving the data as `gdp`, pipe it right into `ggplot`!^[Hint: You can use `.` as a placeholder.]

# Example I: Majors

Now let's step it up to work with some data "in the wild" to answer some research questions. This will have you combine your `dplyr` skills and add some new things such as importing with `readr`.

Let's look at fivethirtyeight's article ["
The Economic Guide To Picking A College Major
"](https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/). fivethirtyeight is great about making the data behind their articles public, we can [download all of their data here](https://data.fivethirtyeight.com/). Search for `college majors` and click download (the blue arrow button).^[This will download a `.zip` file that contains many spreadsheets. Unzip it with a program that unzips files (such as WinZip, 7-zip, the Unarchiver, etc).] We will look at the `recent-grads.csv` file.

The description in the `readme` file for the data is as follows:

Header | Description
---|---------
`Rank` | Rank by median earnings
`Major_code` | Major code, FO1DP in ACS PUMS
`Major` | Major description
`Major_category` | Category of major from Carnevale et al
`Total` | Total number of people with major
`Sample_size` | Sample size (unweighted) of full-time, year-round ONLY (used for earnings)
`Men` | Male graduates
`Women` | Female graduates
`ShareWomen` | Women as share of total
`Employed` | Number employed (ESR == 1 or 2)
`Full_time` | Employed 35 hours or more
`Part_time` | Employed less than 35 hours
`Full_time_year_round` | Employed at least 50 weeks (WKW == 1) and at least 35 hours (WKHP >= 35)
`Unemployed` | Number unemployed (ESR == 3)
`Unemployment_rate` | Unemployed / (Unemployed + Employed)
`Median` | Median earnings of full-time, year-round workers
`P25th` | 25th percentile of earnigns
`P75th` | 75th percentile of earnings
`College_jobs` | Number with job requiring a college degree
`Non_college_jobs` | Number with job not requiring a college degree
`Low_wage_jobs` | Number in low-wage service jobs

## 18.

#### Import the data with `read_csv()` and assign it to `tibble` called `majors`.^[One way to avoid error messages is to move it to the same file as R's working directory, which again you can determine with `getwd()`.] The first argument of this command the name of the original file, in quotes.^[If the file is in a different folder, the argument is the full path in quotes.]

## 19. Look at the data with `glimpse()`. This is a suped-up version of `str()` in `tidyverse`.

## 20.
What are all of the unique values of `Major`? How many are there?

## 21.
### Which major has the lowest unemployment rate?

## 22. 
#### What are the top 3 majors that have the highest percentage of women?

### 23. 
#### Make a boxplot of `Median` wage by `Major_Category`.^[You won't be able to read the labels easily, so add `theme(axis.text.x=element_text(angle=45, hjust=1)` to angle `x`-axis labels (and move them down by 1)]

### 24.
#### Which major category is the least popular in this sample?^[Hint: use `group_by` first.]

### 25. 
#### Is there a systematic difference between STEM majors and non-STEM majors? First define:

```{r}
stem_categories <- c("Biology & Life Science",
                     "Computers & Mathematics",
                     "Engineering",
                     "Physical Sciences")
```
Next, make a variable `stem`, for whether or not a `Major_category` is `"stem"` or `"not_stem"`.^[Hint: try out the `ifelse()` function which has three inputs: condition(s) for a variable(s), what to do if `TRUE` (the if), and what to if `FALSE` (the else).]

Then `summarize()` `median` for stem and not stem groups.

### 2019-09-10 static/slides/05-slides.Rmd
Finalize 5 slides and 5 practice

@@
       beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
-      highlightStyle: solarized-dark #ocean rainbow #monokai rainbow solarized-light #tomorrow-night -blue -eighties zenburn
+      highlightStyle: github
       highlightLines: true
@@
 
+class: inverse, middle, center
+
+.pull-left[
+
+### [tibble: friendlier dataframes](#12)
+
+### [magrittr: piping code](#17)
+
+### [readr: importing data](#26)
+
+### [dplyr: wrangling data](#32)
+
+### [dplyr::filter(): select observations](#35)
+
+### [dplyr::arrange(): reorder observations](#50)
+]
+
+.pull-right[
+
+### [dplyr::select(): select variables](#60)
+
+### [dplyr::rename(): rename variables](#73)
+
+### [dplyr::mutate(): create new variables](#77)
+
+### [dplyr::summarize(): create statistics](#98)
+
+### [tidyr: reshaping data](#133)
+
+### [dplyr: combining datasets](#134)
+
+]
+
+---
+
 # Data Wrangling
 
-- Most real world data analysis transforming chaos into order
+.pull-left[
+
+- Most data analysis is taming chaos into order
     - Data strewn from multiple sources
-    - Missing data
-    - Data not in readable form
-
----
-
-# Workflow of a Data Scientist
-
-.pull-left[
-.center[.polaroid[![](https://www.dropbox.com/s/zke6m65wp3upouj/workflow.png?raw=1)]]
-]
-
-.pull-right[
-- Consider Hadley Wickham's famous diagram from [R for Data Science](http://r4ds.had.co.nz)
-1. .shoutcaps[Import] raw data from out there in the world
-2. .shoutcaps[Tidy] it into a form that you can use to 
-3. .shoutcaps[Explore] the data
-4. .shoutcaps[Communicate] results to target audience
-]
-
----
-
-# Workflow of a Data Scientist
-
-.pull-left[
-.center[.polaroid[![](https://www.dropbox.com/s/zke6m65wp3upouj/workflow.png?raw=1)]]
-]
-
-.pull-right[
-
-3. .shoutcaps[Explore] the data:
-- .whisper[Transform] existing variables into more useful ones
-- Make plots to .whisper[visualize] .onfire[relationships]
-- Create .whisper[models] to analyze, predict, and interpret causal relationships
-
-- 95% of the day-to-day work of data scientists, researchers, econometricians: ".hl[data wrangling]"
-]
-
----
-
-# Workflow of a Data Scientist
+    - Missing data ("`NA`")
+    - Data not in a readable form
+
+.center[
+![:scale 90%](https://www.dropbox.com/s/n9fm2lkii5renzf/messyexcel.png?raw=1)
+]
+]
+
+.pull-right[
+.center[
+![](https://www.dropbox.com/s/gp1lxciqaxq141q/datawrangling0.jpg?raw=1)
+]
+]
+---
+
+# Workflow of a Data Scientist I
+
+.pull-left[
+.center[
+.polaroid[![](https://www.dropbox.com/s/zke6m65wp3upouj/workflow.png?raw=1)]
+
+[R for Data Science](http://r4ds.had.co.nz)
+]
+]
+
+.pull-right[
+- Consider Hadley Wickham's famous diagram:
+1. .whisper[Import] raw data from out there in the world
+2. .whisper[Tidy] it into a form that you can use 
+3. .whisper[Explore] the data (do these 3 repetitively!)
+    - **Transform**
+    - **Visualize**
+    - **Model**
+4. .whisper[Communicate] results to target audience
+
+Ideally, you'd want to be able to do all of this in one program
+]
+
+---
+
+# Workflow of a Data Scientist II
 
@@
 ![](https://www.dropbox.com/s/jd68ky17gb7pmu0/datawranglingnyt.png?raw=1)
-]
-]
-
-.pull-right[
-.center[
-![](https://www.dropbox.com/s/8kzy823yubdv8gh/datawranglingnyt2.png?raw=1)
-]
-]
-
-.center[
 [New York Times](https://www.nytimes.com/2014/08/18/technology/for-big-data-scientists-hurdle-to-insights-is-janitor-work.html)
+]
+
+]
+
+.pull-right[
+
+> "Yet far too much handcrafted work - what data scientists call "**data wrangling**," "**data munging**," and "**data janitor work**" - is still required.
+Data scientists, according to interviews and expert estimates, spend from **50 to 80 percent of their time** mired in this more mundane labor of collecting and preparing unruly digital data, before it can be explored for useful nuggets."
+
 ]
@@
 
-# The tidyverse
+# The tidyverse I
 
@@
 
-- Core packages include ones we've discussed before: `ggplot2`, `dplyr`, `magrittr` among several others (`tidy` `readr`, `purr`, `forcats`, `stringr`)
-- Loading any `tidyverse` package loads `magrittr` (so you can use `%>%`)
+- Allows you to do all of those things with one (set of) package(s)!
 - Learn more at [tidyverse.org](tidyverse.org)
 
+.center[
 ![](https://www.dropbox.com/s/81ijxswbwqor0up/tidyverse.png?raw=1)
+]
 
@@
 - Easiest to just load the core tidyverse all at once
+    - First install may take a few minutes - installs a lot of packages!
     - Note loading the tidyverse is "noisy", it will spew a lot of messages
-    - Hide them with `suppressPackageStartupMessages()` and insert the `library()` command inside the `()`
+    - Hide them with `suppressPackageStartupMessages()` and insert `library()` command inside
 
@@
 ```{r, message=TRUE, warning=TRUE}
-library("tidyverse")
+# install for first time
+# install.packages("tidyverse") # this takes a few minutes and may give several prompts
+
+# load tidyverse 
+suppressPackageStartupMessages(library("tidyverse"))
 ```
@@
 
-![](https://www.dropbox.com/s/9h1trto0iapnklx/tidyverse2.png?raw=1)
-
+- `tidyverse` contains a lot of packages, not all are loaded automatically
+
+```{r}
+tidyverse_packages()
+```
+
+---
+
+# Your Workflow in the tidyverse:
+
+.center[
+![:scale 80%](https://www.dropbox.com/s/9h1trto0iapnklx/tidyverse2.png?raw=1)
+]
 ---
@@
 
-- We will make **extensive** use of:
-
-1. `dplyr` for data wrangling
-2. `tidyr` for tidying data
-3. `ggplot2` for plotting data
-
-- Loading `tidyverse` makes **automatically** uses
-
-4. `tibble` for better data frames
-5. `magrittr` for "pipeable" code
-
-- We might later look at
-
-6. `readr` for importing data
-7. `forcats` for working with factors
-8. `stringr` for working with strings
-
----
-
-# Tibble
+- We will make **extensive** use of (and talk today about):
+
+1. `tibble` for friendlier dataframes
+2. `magrittr` for "pipeable" code
+3. `readr` for importing data
+4. `dplyr` for data wrangling
+5. `tidyr` for tidying data
+6. `ggplot2` for plotting data (we've already covered)
+
+- We will (or might) later look at:
+7. `broom` for tidy regression (not part of core tidyverse)
+8. `forcats` for working with factors
+9. `stringr` for working with strings
+10. `lubridate` for working with dates and times
+11. `purrr` for iteration
+
+---
+
+class: inverse, center, middle
+# tibble: friendlier dataframes
+
+---
+
+# tibble I
 
@@
 
-- `tibble` converts all `data.frames` into a .whisper[friendlier] version called `tibbles`
-
-]
-
----
-
-# Tibble
+- `tibble` converts all `data.frames` into a *friendlier* version called `tibbles` (or `tbl_df`)
+
+]
+
+---
+
+# tibble II
 
@@
 
-- Prints much nicer output in R
-- Shows a bit of the `str`ucture of a dataframe automatically:
+- Prints much nicer output
+
+- Shows a bit of the `str`ucture:
     - `nrow() x ncol()`
@@
     - `<int>` is an integer
+
 - Fundamental grammar of tidyverse: 
@@
 
-# Magrittr I
+# tibble III
+
+.left-column[
+
+.center[![](https://www.dropbox.com/s/u6ubqkng1ja5jwf/tibble.png?raw=1)]
+]
 
@@
 
-- The .hl[magrittr] package allows us to use the **"pipe" operator** (`%>%`)<sup>.red[1]</sup>
-- `%>%` "pipes" the *output* of the *left* of the pipe *into* the *argument* of the *right* 
+- Create a `tibble` from a `data.frame` with `as_tibble()`
+
+```{r, eval=F}
+as_tibble(mpg) # take built-in dataframe mpg
+```
+
+- Create a `tibble` from scratch with `tibble()`, works like `data.frame()`
+
+```{r}
+example<-tibble(x = seq(2,6,2), # sequence from 2 to 6 by 2's
+                y = rnorm(3,0,1), # 3 random draws with mean 0, sd 1
+                colors = c("orange", "green", "blue"))
+
+example
+```
+]
+
+---
+
+# tibble IV
+
+.left-column[
+
+.center[![](https://www.dropbox.com/s/u6ubqkng1ja5jwf/tibble.png?raw=1)]
+]
+
+.right-column[
+
+- Create a `tibble` row-by-row with `tribble()`
+
+```{r}
+example_2<-tribble(
+  ~x, ~y, ~color, # each variable name starts with ~
+  2, 1.5, "orange",
+  4, 0.2, "green",
+  6, 0.8, "blue") # last element has no comma
+
+example_2
+```
+]
+
+---
+
+class: inverse, center, middle
+# magrittr: piping code
+
+---
+
+# magrittr I
+
+.left-column[
+
+.center[
+![:scale 100%](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)
+]
+]
+
+.right-column[
+
+- The `magrittr` package allows us to use the **"pipe" operator** (`%>%`)<sup>.red[1]</sup>
+
+- `%>%` "pipes" the *output* of the *left* of the pipe *into* the *(1<sup>st</sup>) argument* of the *right*
+
 - Running a function `f` on object `x` as `f(x)` becomes `x %>% f` in pipeable form
@@
 
+.footnote[<sup>.red[1]</sup> Keyboard shortcuts in R Studio: `CTRL+Shift+M` (Windows) or `Cmd+Shift+M` (Mac)]
+
+---
+
+# magrittr II
+
 .left-column[
 
-.center[![](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)]
-]
-
-.footnote[<sup>.red[1]</sup> Keyboard shortcuts in R Studio: `CTRL+Shift+M` (Windows) or `Cmd+Shift+M` (Mac)]
-
----
-
-# Magrittr II
+.center[
+![:scale 100%](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)
+]
+]
 
@@
 i.e. take `x` and perform function `f()` on `x` and then take that result and perform function `g()` on it
+
 - With pipes, read operations from left $\rightarrow$ right:
@@
 
+---
+
+# magrittr III
+
 .left-column[
 
-.center[![](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)]
-]
-
----
-
-# Magrittr III
+.center[
+![:scale 100%](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)
+]
+]
 
@@
 
+---
+
+# magrittr IV
+
 .left-column[
 
-.center[![](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)]
-]
-
----
-
-# Magrittr III
+.center[
+![:scale 100%](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)
+]
+]
 
@@
 
-.left-column[
-
-.center[![](https://www.dropbox.com/s/74srdjrw13mgx4x/magrittr.png?raw=1)]
-]
-
----
-
-# Magrittr: Simple Example
+---
+
+# magrittr: Simple Example
 
@@
 .pull-left[
-
-```{r}
+```{r, eval=F}
 # look at top 6 rows
 head(gapminder)
-```
-]
-
-.pull-right[
-```{r}
-# use the pipe instead 
+
+# use pipe instead
 gapminder %>% head()
@@
 
----
-
-# Magrittr: More Involved Example
-
-- These two methods produce the same output (avg hwy mpg of Audi cars)
+.pull-right[
+
+```{r, echo=F}
+gapminder %>% head()
+```
+]
+
+---
+
+# magrittr: More Involved Example
+
+- These two methods produce the same output (average hightway mpg of Audi cars)
+
+--
 
@@
 ```
+
+--
 
@@
 
-# Readr
+class: inverse, center, middle
+# readr: importing data
+
+---
+
+# readr
 
@@
 
-- `Readr` lets you load common spreadsheet files (`.csv`, `.tsv`) with simple commands:
-- `read_csv(path/to/my_data.csv)`<sup>.red[1]</sup>
-    - `read_tsv(path/to/my_data.tsv)`
-- Often this is enough, but there are many more customizations as necessary
-- You can also export your data from R into a common spreadsheet file with:
-- `write_csv(my_dataframe, path=path/to/file_name.csv)`<sup>.red[2]</sup> 
+- `readr` helps load common spreadsheet files (`.csv`, `.tsv`) with simple commands:
+- `read_*(path/to/my_data.*)`
+    - where `*` can be `.csv` or `.tsv`
+
+- Often this is enough, but many more customizations possible
+
+- You can also *export* your data from R into a common spreadsheet file with:
+- `write_*(my_df, path = path/to/file_name.*)`
+    - where `my_df` is the name of your `tibble`, and `file_name` is the name of the file you want to save as
+
 - Read more on the [tidyverse website](https://readr.tidyverse.org/) and the [Readr Cheatsheet](https://rawgit.com/rstudio/cheatsheets/master/data-import.pdf)
@@
 
-.footnote[<sup>.red[1]</sup> If you are using an R Project, this file path is in your main project folder, or any subfolders you decided to create, such as `data/my_data.csv`
-
-<sup>.red[2]</sup> where `my_dataframe` is the name of the dataframe you want to export, and `file_name` is the name of the file you want to create.]
-
 ---
@@
 
-.center[![](https://www.dropbox.com/s/rl1sj67y2m8fqy5/readxl.png?raw=1)]
-.center[![](https://www.dropbox.com/s/ua6t5wz1se34vmo/haven.png?raw=1)]
+.center[
+![:scale 60%](https://www.dropbox.com/s/rl1sj67y2m8fqy5/readxl.png?raw=1)
+
+![:scale 60%](https://www.dropbox.com/s/ua6t5wz1se34vmo/haven.png?raw=1)]
 ]
@@
 
-- Some data types, particularly for other software programs like Excel, STATA, SAS, and SPSS are not `.csv` or `.txt` spreadsheets
-- `readxl` has commands equivalent to `readr`'s for Excel data types:
-    - `read_xls("path/to/my/data.xls")` for older Excel `.xls` files
-    - `read_xlsx("path/to/my/data.xlsx")` for modern Excel `.xls` files
-    - `write_xls(my_dataframe, path=path/to/file_name.xls)` or ``write_xlsx(my_dataframe, path=path/to/file_name.xlsx)` to export as `xls` or `xlsx` files
-- `haven` has commands equivalent to `readr`'s for these data types:
-    - `read_dta("path/to/my_data.dta")` for STATA `.dta` files
-    - `read_sav("path/to/my_data.sav")` for SPSS `.sav` files
-    - `read_sas("path/to/my_data.sas7bdat")` for SAS files 
-]
+- For other data types from software programs like Excel, STATA, SAS, and SPSS:
+- `readxl` has equivalent commands for Excel data types:
+    - `read_*("path/to/my/data.*")`
+    - `write_*(my_dataframe, path=path/to/file_name.*)`
+    - where `*` can be `.xls` or `.xlsx`
+
+- `haven` has equivalent commands for other data types:
+    - `read_*("path/to/my_data.dta")` for STATA `.dta` files
+    - `write_*(my_dataframe, path=path/to/file_name.*)`
+    - where `*` can be `.dta` (STATA), `.sav` (SPSS), `.sas7bdat` (SAS)
+]
+
+---
+
+# Common Import Issues I
+
+- Most common: *"where the hell is my data file"??*
+
+- Recall `R` looks for files to `read_*()` in the default working directory (check what it is with `getwd()`, change it with `setwd()`)
+
+- You can tell `R` where this data is by making the `path` a part of the file's name when importing
+    - Use `..` to "move up one folder"
+    - Use `/` to "enter a folder"
+
+- Either use an **absolute path** on your computer:
+```{r, eval=F}
+# Example
+
+df <- read_csv("C:/Documents and Settings/Ryan Safner/Downloads/my_data.csv")
+```
+
+- Or use a **relative path** *from* R's working directory
+```{r, eval=F}
+# Example
+# If working directory is Documents, but data is in Downloads, like so: 
+# 
+# Ryan Safner
+# |
+# |- Documents/
+# |- Downloads/
+# |- Photos/
+# |- Videos/
+df <- read_csv("../Downloads/my_data.csv")
+```
+
+---
+
+# Common Import Issues II
+
+- Most common: *"where the hell is my data file"??*
+
+- Recall `R` looks for files to `read_*()` in the default working directory (check what it is with `getwd()`, change it with `setwd()`)
+
+- You can tell `R` where this data is by making the `path` a part of the file's name when importing
+    - Use `..` to "move up one folder"
+    - Use `/` to "enter a folder"
+
+- Or use a **relative path** *from* R's working directory
+```{r, eval=F}
+# Example
+# If working directory is Documents, but data is in Downloads, like so: 
+# 
+# Ryan Safner
+# |
+# |- Documents/
+# |- Downloads/
+# |- Photos/
+# |- Videos/
+df <- read_csv("../Downloads/my_data.csv")
+```
+
+---
+
+# Common Import Issues III
+
+- **Suggestion** to make your data import easier: *Download and move files to R's working directory*
+
+- Your computer and working directory are different from mine (and others)
+
+- This is *not* a reproducible workflow!
+
+- We'll finally fix this next class with `R Projects`
+    - The working directory is set to the Project Folder by default
+    - Same for everyone on any computer!
+
+---
+
+class: inverse, center, middle
+# dplyr: wrangling data
 
@@
 
-- `dplyr` uses much more efficient and intuitive commands<sup>.magenta[1]</sup> to manipulate dataframes
-- Grammatically:
-    - `Base R` is all about passively running functions on nouns: `function(object)`
-    - `dplyr` is all about actively using verbs: `dplyr_verb(df, conditions)`<sup>.red[2]</sup>
+- `dplyr` uses more efficient & intuitive commands to manipulate tibbles
+
+- `Base R` grammar: passively running functions on nouns: `function(object)`
+- `dplyr` grammar: actively using verbs: `dplyr_verb(df, conditions)`<sup>.red[1]</sup>
+
 - Three great features:
@@
 1. Allows use of `%>%` pipe operator
-2. Input and output is always a dataframe<sup>.magenta[3]</sup>
+2. Input and output is always a `tibble`
 3. Shows the output from a manipulation, but does not save/overwrite as an object unless explicitly assigned to an object
@@
 
-.footnote[<sup>.magenta[1]</sup> Particularly when combined with `magrittr`'s pipe (`%>%`) command!
-
-<sup>.red[2]</sup> With the pipe, even simpler: `df %>% dplyr_verb(conditions)`
-
-<sup>.red[3]</sup> Technically a `tibble` with `tidyverse`!]
+.footnote[<sup>.red[1]</sup> With the pipe, even simpler: `df %>% dplyr_verb(conditions)`]
 
@@
 |----------|------|
-| `filter` | Keep only selected *observations* | 
-| `select` | Keep only selected *variables* |
-| `arrange` | Reorder rows (e.g. in numerical order) |
-| `mutate` | Create new variables | 
-| `recode` | Change a variable's values or categories/`factor` levels |
-| `summarize` | Collapse data into summary statistics| 
-| `group_by` | Perform any of the above functions by groups/categories |
-
-]
-
----
-
-# dplyr::filter
+| `filter()` | Keep only selected *observations* | 
+| `select()` | Keep only selected *variables* |
+| `arrange()` | Reorder rows (e.g. in numerical order) |
+| `mutate()` | Create new variables | 
+| `summarize()` | Collapse data into summary statistics| 
+| `group_by()` | Perform any of the above functions by groups/categories |
+
+]
+
+---
+
+class: inverse, center, middle
+# dplyr::filter(): select observations
+
+.center[
+![](https://www.dropbox.com/s/n9o9x29mlvyrr8q/filter.png?raw=1)
+]
+
+---
+
+# dplyr::filter()
 
@@
 filter(gapminder, continent=="Africa")
-
+```
+
+```{r, eval=F}
 # using the pipe
@@
 ```{r, echo=F}
-# syntax with the pipe 
-gapminder %>%
-  filter(continent=="Africa")
+gapminder %>%
+  filter(continent == "Africa")
 ```
@@
 # base syntax
-africa <- filter(gapminder, continent=="Africa")
-
+africa <- filter(gapminder,
+                 continent=="Africa")
+```
+
+```{r, eval=F}
 # using the pipe
@@
   filter(continent == "Africa")
-
+```
+
+```{r, eval=F}
 # look at new tibble
@@
 # base syntax
-gapminder <- filter(gapminder, continent=="Africa")
-
+gapminder <- filter(gapminder,
+                    continent=="Africa")
+```
+
+```{r, eval=F}
 # using the pipe
@@
 
-# dplyr conditionals
+# dplyr Conditionals
 
 - In many data wrangling contexts, you will want to select data .whisper[conditionally]
-    - In computer code, you are selecting only observations for which a set of logical conditions are `TRUE`
+
+- In computer code, you are selecting only observations for which a set of logical conditions are `TRUE`<sup>.red[1]</sup>
+
 - For example:
@@
     - `%in%`: is a member of some defined set $(\in)$
-    - `&`: AND<sup>1</sup>
+    - `&`: AND (commas also work instead)
     - `|`: OR
-
-.footnote[<sup>1</sup> Commas also work in `dplyr`.]
-
----
-
-# dplyr::filter with conditionals
-
---
+    - `!`: not
+
+.footnote[<sup>.red[1]</sup> See `?Comparison` and `?Base::Logic`.]
+
+---
+
+# dplyr::filter() with Conditionals
 
@@
 ```{r, eval=F}
-# look only at African observations in 1997
+# look only at African observations
+# in 1997
+gapminder %>%
+  filter(continent == "Africa",
+         year == 1997)
+```
+]
+
+--
+
+.right-plot[
+
+```{r, echo=F}
 gapminder %>%
@@
 
+---
+
+# dplyr::filter() with Conditionals II
+
+.left-code[
+
+```{r, eval=F}
+# look only at African observations 
+# or observations in 1997
+gapminder %>%
+  filter(continent == "Africa" |
+           year == 1997)
+```
+]
+
 --
@@
 gapminder %>%
-filter(continent == "Africa", year == 1997)
-```
-]
-
----
-
-# dplyr::filter with conditionals II
-
---
+filter(continent == "Africa" |
+       year == 1997)
+```
+]
+
+---
+
+# dplyr::filter() with Conditionals III
 
@@
 ```{r, eval=F}
-# look only at African observations or observations in 1997
-gapminder %>%
-  filter(continent == "Africa" | year == 1997)
+# look only at U.S. and U.K. 
+# observations in 2002
+gapminder %>%
+  filter(country %in% 
+           c("United States",
+             "United Kingdom"),
+         year == 2002)
 ```
@@
 gapminder %>%
-filter(continent == "Africa" | year == 1997)
-```
-]
-
----
-
-# dplyr::filter with conditionals III
-
---
-
-.left-code[
-
-```{r, eval=F}
-# look only at U.S. and U.K. observations in 2002
-gapminder %>%
   filter(country %in% c("United States", "United Kingdom"), year == 2002)
@@
 
---
-
-.right-plot[
-
-```{r, echo=F}
-gapminder %>%
-  filter(country %in% c("United States", "United Kingdom"), year == 2002)
-```
-]
-
----
-
-# dplyr::arrange I
+---
+
+class: inverse, center, middle
+# dplyr::arrange(): reorder observations
+
+---
+
+# dplyr::arrange() I
 
@@
 ```{r, eval=F}
-# order by smallest to largest population
+# order by smallest to largest pop
 # syntax without the pipe
 arrange(gapminder, pop)
-
+```
+
+```{r, eval=F}
 # using the pipe
@@
 
-# dplyr::arrange II
+# dplyr::arrange() II
 
@@
 ```{r, eval=F}
-# order by year, with the smallest to largest population in each year
+# order by year, with the smallest 
+# to largest pop in each year
 # syntax without the pipe
 arrange(gapminder, year, pop)
-
+```
+
+```{r, eval=F}
 # using the pipe
@@
 
-# dplyr::arrange III
+# dplyr::arrange() III
 
@@
 ```{r, eval=F}
-# order by largest to smallest population
+# order by largest to smallest pop
 # syntax without the pipe
 arrange(gapminder, desc(pop))
-
+```
+
+```{r, eval=F}
 # using the pipe
@@
 
-# dplyr::select I
+class: inverse, center, middle
+# dplyr::select(): select variables
+
+.center[
+![](https://www.dropbox.com/s/meq9tyvn4u5wcfb/select.png?raw=1)
+]
+
+---
+
+# dplyr::select() I
 
@@
 ```{r, eval=F}
-# keep only country, year, and population variables
+# keep only country, year, 
+# and population variables
 # syntax without the pipe
 select(gapminder, country, year, pop)
-
+```
+
+```{r, eval=F}
 # using the pipe
@@
 
-# dplyr::select II
+# dplyr::select() II
 
@@
 ```{r, eval=F}
-# keep all except gdpPercap
+# keep all *except* gdpPercap
 # syntax without the pipe
 select(gapminder, -gdpPercap)
-
+```
+
+```{r, eval=F}
 # using the pipe
@@
 
-# dplyr::select III
+# dplyr::select() III
 
@@
 select(gapminder, pop, everything())
-
+```
+
+```{r, eval=F}
 # using the pipe
@@
 
-# dplyr::select IV
+# dplyr::select() IV
 
@@
 ```{r}
-# keep country and all variables containing "per"
+# keep country and all variables 
+# containing "per"
 
@@
 
-# dplyr::rename
+class: inverse, center, middle
+# dplyr::rename(): rename variables
+
+---
+
+# dplyr::rename()
 
 - `rename` changes the name of a variable (column)
-    - Format: `new_name=old_name`
+    - Format: `new_name = old_name`
 --
@@
 rename(gapminder, GDP = gdpPercap)
-
+```
+
+```{r, eval=F}
 # using the pipe
@@
 
-# dplyr::mutate
+class: inverse, center, middle
+# dplyr::mutate(): create new variables
+
+.center[
+![](https://www.dropbox.com/s/1ktrsvya8pq3hh3/mutate.png?raw=1)
+]
+
+---
+
+# dplyr::mutate()
 
@@
 
-# dplyr::mutate II
+# dplyr::mutate() II
 
@@
 
-.left-code[
-
-```{r, eval=F}
-# create variable "europe" if country is in Europe
+.pull-left[
+
+```{r, eval=F}
+# create variable "europe" if country 
+# is in Europe
 # syntax without the pipe
-mutate(gapminder, europe = ifelse(continent == "Europe", 
+mutate(gapminder, 
+       europe = ifelse(continent == "Europe",
                        yes = "In Europe",
                        no = "Not in Europe"))
-
+```
+
+```{r, eval=F}
 # using the pipe
@@
 
-.right-plot[
+.pull-right[
 
@@
 gapminder %>%
+  select(1:3) %>% # hiding vars just for space constraints
   mutate(europe = ifelse(continent == "Europe",
@@
 
-# dplyr::mutate III
+# dplyr::mutate() III
 
@@
 # syntax without the pipe
-mutate(gapminder, pop_millions = pop / 1000000)
-
+mutate(gapminder,
+       pop_mil = pop / 1000000)
+```
+
+```{r, eval=F}
 # using the pipe
@@
 gapminder %>%
-  rename(pop_millions = pop / 1000000)
+  rename(pop_mil = pop / 1000000)
 ```
@@
 gapminder %>%
-  mutate(pop_millions = pop / 1000000)
-```
-]
-
----
-
-# dplyr::mutate IV
+  select(1:5) %>% # hiding vars just for space constraints
+  mutate(pop_mil = pop / 1000000)
+```
+]
+
+---
+
+# dplyr::mutate() IV
 
@@
 ```{r, eval=F}
-# create GDP variable from gdpPercap and pop
+# create GDP variable from gdpPercap 
+# and pop, in billions
 # syntax without the pipe
-mutate(gapminder, GDP = gdpPercap * pop)
-
+mutate(gapminder,
+       GDP = ((gdpPercap * pop)/1000000000))
+```
+
+```{r, eval=F}
 # using the pipe
@@
 gapminder %>%
-  rename(GDP = gdpPercap * pop)
+  mutate(GDP = ((gdpPercap * pop)/1000000000))
 ```
@@
 gapminder %>%
-  mutate(GDP = gdpPercap * pop)
-```
-]
-
----
-
-# dplyr::mutate V
+  select(c(country, continent, year, pop, gdpPercap)) %>% # hiding vars just for space constraints
+  mutate(GDP = ((gdpPercap * pop)/1000000000))
+```
+]
+
+---
+
+# dplyr::mutate() V
+
+- Change `class` of a variable inside `mutate()` with `as.*()`
+
+```{r}
+gapminder %>% head(., 2)
+
+# change year from an integer to a factor
+gapminder %>%
+  mutate(year = as.factor(year))
+```
+
+---
+
+# dplyr::mutate(): Multiple Variables
 
@@
 
---
+---
+
+# dplyr::transmute()
 
@@
 
---
-
-- Don't forget to assign the output to  new `tibble` if you want to "save" the new variables!
-
----
-
-# Exporting Tables with knitr::kable
-
----
-
-# dplyr::tally with group_by
-
-- `tally` provides counts, best used with `group_by` for categorical data
-
+---
+
+# dplyr::mutate(): Conditionals
+
+- Boolean, logical, and conditionals all work well in `mutate()`:
+
+```{r}
+gapminder %>%
+  select(country, year, lifeExp) %>%
+  mutate(long_1 = lifeExp > 70,
+         long_2 = ifelse(lifeExp > 70, "Long", "Short"))
+```
+
+---
+
+# dplyr::mutate(): order Aware
+
+- `mutate()` is order-aware, so you can chain multiple mutates that depend on previous mutates
+
+```{r}
+gapminder %>%
+  select(country, year, lifeExp) %>%
+  mutate(dog_years = lifeExp * 7,
+         comment = paste("Life expectancy in", country, "is", dog_years, "in dog years.", sep = " "))
+```
+
+---
+
+# dplyr::mutate(): case_when()
+
+- `case_when` creates a new variable with values that are conditional on values of other variables (e.g., "if/else")
+    - Last argument: `TRUE`: when 
+```{r}
+gapminder %>%
+  mutate(European = case_when(
+    continent == "Europe" ~ "Aye",
+    TRUE ~ "Nay"
+  ))
+```
+
+---
+
+# dplyr::mutate(): scoped I
+
+- "Scoped" variants of `mutate` that work on a subset of variables:
+    - `mutate_all()` affects every variable
+    - `mutate_at()` affects named or selected variables
+    - `mutate_if()` affects variables that meet a criteria
+
+```{r}
+# round all observations of numeric
+# variables to 2 digits
+gapminder %>%
+  mutate_if(is.numeric, round, digits = 2)
+```
+
+---
+
+# dplyr::mutate(): scoped II
+
+- "Scoped" variants of `mutate` that work on a subset of variables:
+    - `mutate_all()` affects every variable
+    - `mutate_at()` affects named or selected variables
+    - `mutate_if()` affects variables that meet a criteria
+
+```{r}
+# make all factor variables uppercase
+gapminder %>%
+  mutate_if(is.factor, toupper) 
+```
+
+---
+
+# dplyr::mutate()
+
+- Don't forget to assign the output to a new `tibble` (or overwrite original) if you want to "save" the new variables!
+
+---
+
+class: inverse, center, middle
+# dplyr::summarize(): create statistics
+
+.center[
+![](https://www.dropbox.com/s/d7hiv0fst5nhhzu/summarize.png?raw=1)
+]
+
+---
+
+# dplyr::summarize() I
+
+- `summarize`<sup>.red[1]</sup> outputs a tibble of desired summary statistics
+    - can name the statistic variable as if you were `mutate`-ing a new variable
+--
+
+.left-code[
+
+```{r, eval=F}
+# get average life expectancy
+# call it avg_LE
+summarize(gapminder,
+          avg_LE = mean(lifeExp))
+```
+
+```{r, eval=F}
+# using the pipe
+
+gapminder %>%
+  summarize(avg_LE = mean(lifeExp))
+```
+
+]
+
+--
+
+.right-plot[
+
+```{r, echo=F}
+gapminder %>%
+  summarize(avg_LE = mean(lifeExp))
+```
+
+]
+
+.footnote[<sup>.red[1]</sup> Also the more civilised non-U.S. English spelling `summarise` also works. `dplyr` was written by a Kiwi after all!
+]
+---
+
+# dplyr::summarize() II
+
+- Useful `summarize()` commands:
+
+| Command | Does |
+|---------|------|
+| `n()`<sup>.red[*]</sup>   | Number of observations |
+| `n_distinct()`<sup>.red[*]</sup> | Number of unique observations |
+| `sum()` | Sum all observations of a variable |
+| `mean()` | Average of all observations of a variable |
+| `median()` | 50<sup>th</sup> percentile of all observations of a variable |
+| `sd()`  | Standard deviation of all observations of a variable |
+
+.footnote[<sup>.red[*]</sup> Most commands require you to put a variable name inside the command's argument parentheses. These commands require nothing to be in parentheses!]
+
+---
+
+# dplyr::summarize() II
+
+- Useful `summarize()` commands (continued):
+
+| Command | Does |
+|---------|------|
+| `min()` | Minimum value of a variable |
+| `max()` | Maximum value of a variable |
+| `quantile(., 0.25)`<sup>.red[+]</sup> | Specified percentile (example `25`<sup>th</sup> percentile) of a variable |
+| `first()` | First value of a variable |
+| `last()` | Last value of a variable |
+| `nth(., 2)`<sup>.red[+]</sup> | Specified position of a variable (example `2`<sup>nd</sup>) |
+
+.footnote[<sup>.red[+]</sup> The `.` is where you would put your variable name.]
+
+---
+
+# dplyr::summarize() counts
+
+- Counts of a categorical variable are useful, and can be done a few different ways:
+
+--
+
+```{r}
+# summarize with n() gives size of current group, has no arguments
+gapminder %>%
+  summarize(amount = n()) # I've called it "amount"
+```
+--
+
+```{r}
+# count() is a dedicated command, counts observations by specified variable
+gapminder %>%
+  count(year) # counts how many observations per year
+```
+
+---
+
+# dplyr::summarize() Conditionally
+
+- Can do counts and proportions by conditions
+    - How many observations fit specified conditions (e.g. `TRUE`)
+    - Numeric objects: `TRUE=1` and `FALSE=0`
+        - `sum(x)` becomes the number of `TRUE`s in `x`
+        - `mean(x)` becomes the proportion
+
+--
+
+.pull-left[
+```{r}
+# How many countries have life expectancy
+# over 70 in 2007?
+gapminder %>%
+  filter(year=="2007") %>%
+  summarize(Over_70 = sum(lifeExp>70))
+```
+]
+--
+
+.pull-right[
+```{r}
+# What *proportion* of countries have life
+# expectancy over 70 in 2007?
+gapminder %>%
+  filter(year=="2007") %>%
+  summarize(Over_70 = mean(lifeExp>70))
+```
+]
+---
+
+# dplyr::summarize() Multiple Variables
+
+- Can `summarize()` multiple *variables* at once, separate by commas
+
+.left-code[
+
+```{r, eval=F}
+# get average life expectancy and GDP 
+# call each avg_LE, avg_GDP
+summarize(gapminder,
+          avg_LE = mean(lifeExp),
+          avg_GDP = mean(gdpPercap))
+```
+
+```{r, eval=F}
+# using the pipe
+
+gapminder %>%
+  summarize(avg_LE = mean(lifeExp),
+            avg_GDP = mean(gdpPercap))
+```
+
+]
+
+--
+
+.right-plot[
+
+```{r, echo=F}
+gapminder %>%
+  summarize(avg_LE = mean(lifeExp),
+            avg_GDP = mean(gdpPercap))
+```
+
+]
+
+---
+
+# dplyr::summarize() Multiple Statistics
+
+- Can `summarize()` multiple *statistics* of a variable at once, separate by commas
+
+.left-code[
+
+```{r, eval=F}
+# get count, mean, sd, min, max
+# of life Expectancy 
+summarize(gapminder,
+          obs = n(),
+          avg_LE = mean(lifeExp),
+          sd_LE = sd(lifeExp),
+          min_LE = min(lifeExp),
+          max_LE = max(lifeExp))
+```
+
+```{r, eval=F}
+# using the pipe
+
+gapminder %>%
+  summarize(obs = n(),
+          avg_LE = mean(lifeExp),
+          sd_LE = sd(lifeExp),
+          min_LE = min(lifeExp),
+          max_LE = max(lifeExp))
+```
+
+]
+
+--
+
+.right-plot[
+
+```{r, echo=F}
+gapminder %>%
+  summarize(obs = n(),
+          avg_LE = mean(lifeExp),
+          sd_LE = sd(lifeExp),
+          min_LE = min(lifeExp),
+          max_LE = max(lifeExp))
+```
+
+]
+
+---
+
+# dplyr::summarize() Multiple Statistics
+
+- "Scoped" versions of `summarize()` that work on a subset of variables
+    - `summarize_all()`: affects every variable
+    - `summarize_at()`: affects named or selected variables
+    - `summarize_if()`: affects variables that meet a criteria
+
+--
+
+.pull-left[
+
+```{r}
+# get the average of all
+# numeric variables 
+gapminder %>%
+  summarize_if(is.numeric,
+               funs(avg = mean))
+```
+]
+
+--
+.pull-right[
+```{r}
+# get mean and sd for
+# pop and lifeExp
+
+gapminder %>%
+  summarize_at(vars(pop, lifeExp),
+    funs("avg" = mean,
+         "std dev" = sd))
+```
+]
+
+---
+
+# dplyr::summarize() with group_by() I
+
+- If we have `factor` variables grouping a variable into categories, we can run `dplyr` verbs by group
+    - Particularly useful for `summarize()`
+
+- First define the group with `group_by()`
+
+--
+
+```{r}
+# get average life expectancy and gdp by continent
+
+gapminder %>%
+  group_by(continent) %>%
+  summarize(mean_life = mean(lifeExp),
+            mean_GDP = mean(gdpPercap))
+```
+
+---
+
+# dplyr::summarize() with group_by() II
+
+```{r}
+# track changes in average life expectancy and gdp over time
+
+gapminder %>%
+  group_by(year) %>%
+  summarize(mean_life = mean(lifeExp),
+            mean_GDP = mean(gdpPercap))
+```
+
+---
+
+# dplyr::summarize() with group_by() III
+
+- Can group observations by multiple variables (in proper order)
+
+```{r}
+# track changes in average life expectancy and gdp by continent over time
+
+gapminder %>%
+  group_by(continent, year) %>%
+  summarize(mean_life = mean(lifeExp),
+            mean_GDP = mean(gdpPercap))
+```
+
+---
+
+# Example: Piping Across Packages
+
+- `tidyverse` uses same grammar and design philosophy
+- .green[**Example**]: graphing change in average life expectancy by continent over time
+
+--
+.pull-left[
+.font80[
+```{r ex_plot, eval=F}
+gapminder %>%
+  group_by(continent, year) %>%
+  summarize(mean_life = mean(lifeExp),
+            mean_GDP = mean(gdpPercap)) %>%
+  # now pipe this tibble in as data for ggplot!
+  ggplot(data = ., # . stands in for stuff ^!
+         aes(x = year,
+             y = mean_life,
+             color = continent))+
+  geom_path(size=1)+
+  labs(x = "Year",
+       y = "Average Life Expectancy (Years)",
+       color = "Continent",
+       title = "Average Life Expectancy Over Time")+
+  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
+```
+]
+]
+
+--
+
+.pull-right[
+
+```{r, ref.label="ex_plot", fig.height=5, fig.retina=3, echo=F}
+
+```
+]
+
+---
+
+# dplyr: Other Useful Commands I
+
+- `tally` provides counts, best used with `group_by` for `factors`
+
+--
+
+.pull-left[
 ```{r}
@@
 ```
-
+]
+
+--
+
+.pull-right[
 ```{r}
@@
 ```
-
----
-
-# Advanced: tidyr and joining data
-
+]
+
+---
+
+# dplyr: Other Useful Commands II
+
+- `slice()` subsets rows by *position* instead of `filter`ing by *values*
+
+--
+
+```{r}
+gapminder %>%
+  slice(15:17) # see 15th through 17th observations
+```
+
+---
+
+# dplyr: Other Useful Commands III
+
+- `pull()` extracts a column from a `tibble` (just like `$`)
+
+--
+
+```{r}
+# Get all U.S. life expectancy observations
+gapminder %>%
+  filter(country == "United States") %>%
+  pull(lifeExp)
+```
+
+--
+
+```{r}
+# Get U.S. life expectancy in 2007
+gapminder %>%
+  filter(country == "United States" & year == 2007) %>%
+  pull(lifeExp)
+```
+
+---
+
+# dplyr: Other Useful Commands IV
+
+- `distinct()` shows the distinct values of a specified variable (recall `n_distinct()` inside `summarize()` just gives you the *number* of values)
+
+```{r}
+gapminder %>%
+  distinct(country)
+```
+
+---
+
+class: inverse, center, middle
+# tidyr: reshaping data
+
+.center[
+![](https://www.dropbox.com/s/pu5stir0u59ayae/tidyr2.png?raw=1)
+]
 ---
@@
 - `tidyr` helps clean data into more usable format for other `tidyverse` packages
-- .hl["tidy" data].footnote[This is the core and eponym of the `tidyverse`: all associated packages and functions use or require this data format!] are (an opinionated view of) data where
-
-1. Each .shout[variable] is in a .shout[column]
-2. Each .whisper[observation] is a .whisper[row]
-3. Each .onfire[value] is a .onfire[cell]
+- .shout["tidy" data]<sup>.red[1]</sup> are (an opinionated view of) data where
+
+1. Each .hl[variable] is in a .hl[column]
+2. Each .hl[observation] is a .hl[row]
+3. Each .hl[observational unit] forms a .hl[table]<sup>.red[2]</sup>
 
 - Un-tidy data needs to be tidied into this format
-    - You will spend less time fighting tools and more time on analysis!
-
-]
-
----
-
-# Tidying Data: `gather()` wide to long I
-
-.pull-left[
-
-```{r, echo=F}
+    - Spend less time fighting your tools and more time on analysis!
+
+]
+
+.footnote[<sup>.red[1]</sup> This is the core and eponym of the `tidyverse`: all associated packages and functions use or require this data format!
+
+<sup>.red[2]</sup> Alternatively, sometimes rule 3 is "every value is its own cell."]
+---
+
+# tidyr: Tidy Data
+
+- "tidy" data $\neq$ clean, perfect data
+
+> "Happy families are all alike; every unhappy family is unhappy in its own way." - Leo Tolstoy
+
+> "Tidy datasets are all alike, but every messy dataset is messy in its own way." - Hadley Wickham
+
+.center[
+![](https://www.dropbox.com/s/ukxnk8y9be3fyeb/tidydata1.png?raw=1)
+]
+
+---
+
+# tidyr::gather() wide to long I
+
+.pull-left[
+
+```{r}
+# make example untidy data 
 ex_wide<-tribble(
@@
 )
-```
-
-```{r}
 ex_wide
@@
 
-- Common source of "un-tidy" data: .l[Column headers are values, not variable names]
+- **Common source of "un-tidy" data**: .hl[Column headers are values, not variable names]
     - Column names are *values* of a `year` variable!
     - Each row represents *two* observations (one in 2000 and one in 2010)!
+]
+
+---
+
+# tidyr::gather() wide to long II
+
+.pull-left[
+
+```{r}
+# make example untidy data 
+ex_wide<-tribble(
+  ~"Country", ~"2000", ~"2010",
+  "United States", 140, 180,
+  "Canada", 102, 98,
+  "China", 111, 123
+)
+ex_wide
+```
+
+]
+
+.pull-right[
+
 - We need to `gather()` these columns into a new pair of variables
@@
 ---
-
-# Tidying Data: `gather()` wide to long II
+# tidyr::gather() wide to long III
 
@@
 
-# Tidying Data: `spread()` long to wide I
+# tidyr::spread() long to wide I
 
@@
 ```{r}
-ex_long
-```
-
-]
-
-.pull-right[
-
-- Common source of "un-tidy" data: .l[observations are scattered across multiple rows]
+ex_long # example I made (code hidden)
+```
+
+]
+
+.pull-right[
+
+- **Another common source of "un-tidy" data**: .hl[observations are scattered across multiple rows]
     - Each country has two rows per observation, one for `Cases` and one for `Population` (categorized by `type` of variable)
+]
+
+---
+
+# tidyr::spread() long to wide II
+
+.pull-left[
+
+```{r, echo=F}
+ex_long<-tribble(
+  ~"Country", ~"Year", ~"Type", ~"Count",
+  "United States", 2000, "Cases", 140,
+  "United States", 2000, "Population", 300,
+  "United States", 2010, "Cases", 180,
+  "United States", 2010, "Population", 310,
+  "Canada", 2000, "Cases", 102,
+  "Canada", 2000, "Population", 110,
+  "Canada", 2010, "Cases", 98,
+  "Canada", 2010, "Population", 121,
+  "China", 2000, "Cases", 111,
+  "China", 2000, "Population", 1201,
+  "China", 2010, "Cases", 123,
+  "China", 2010, "Population", 1241
+)
+
+```
+
+```{r}
+ex_long # example I made (code hidden)
+```
+
+]
+
+.pull-right[
+
 - We need to `spread()` these columns into a new pair of variables
@@
 
-# Tidying Data: `spread()` long to wide II
+# tidyr::spread() long to wide III
 
@@
 
+# tidyr
+
+.pull-left[
+.center[
+![](https://www.dropbox.com/s/buiczvm2wny72xz/tidyr-spread-gather.gif?raw=1)
+]
+]
+
+.footnote[<sup>*</sup> Image from Garrick Aden-Buie's excellent [tidyexplain](https://github.com/gadenbuie/tidyexplain)]
+
+---
+
+class: inverse, middle, center
+
 # Combining Datasets
 
+.center[
+![](https://www.dropbox.com/s/99sppuxkb3hdfx3/joinfull.png?raw=1)
+]
+---
+
+# Combining Datasets
+
 - Often, data doesn't come from just one source, but several sources
+
 - We can combine datasets into a single dataframe (tibble) using `dplyr` commands in several ways:
     1. `bind` dataframes together by row or by column
-        - `bind_rows()` adds observations (rows) to existing dataset<sup>.magenta[1]</sup>
-        - `bind_cols()` adds variables (columns) to existing dataset<sup>.magenta[2]</sup>
-    2. `join` two existing dataframes by designating a variable(s) as a `key` to match rows by identical values of that `key`
-
-.footnote[<sup>.magenta[1]</sup> Note the columns must be identical between the original dataset and the new observations
-
-<sup>.magenta[1]</sup> Note the rows must be identical between original dataset and new variable]
-
----
-
-# Two *Similar* Datasets
+        - `bind_rows()` adds observations (rows) to existing dataset<sup>1</sup>
+        - `bind_cols()` adds variables (columns) to existing dataset<sup>2</sup>
+    2. `join` two dataframes by designating variable(s) as `key` to match rows by identical values of that `key`
+
+.footnote[<sup>.red[1]</sup> Note the columns must be identical between the original dataset and the new observations
+
+<sup>.red[2]</sup> Note the rows must be identical between original dataset and new variable]
+
+---
+
+# Two *Similar* Datasets I
 
@@
     - New variables contain all the same observations as existing data
-- In this case, simply using `bind_*(old, new)` will work
-    - `bind_columns(old, new)` adds columns from `new` to `old` dataframe
-    - `bind_rows(old, new)` adds rows from `new` to `old` dataframe
+
+- In this case, simply using `bind_*(old_df, new_df)` will work
+    - `bind_columns(old_df, new_df)` adds columns from `new_df` to `old_df`
+    - `bind_rows(old_df, new_df)` adds rows from `new_df`  to `old_df`
+
+---
+
+# Two *Similar* Datasets II
+
+.pull-left[
+.center[
+`bind_columns()` (Variables)
+
+![](https://www.dropbox.com/s/b1xq2n5t61koxvo/bindcols.png?raw=1)
+]
+]
+
+.pull-right[
+.center[
+`bind_rows()` (Observations)
+
+![](https://www.dropbox.com/s/4wnon47pi6m19z6/bindrows.png?raw=1)
+]
+]
 
@@
 
-.footnote[<sup>*</sup> Images on all following slides come from Gaden Buie's excellent [tidyexplain](https://github.com/gadenbuie/tidyexplain)]
+.footnote[<sup>*</sup> Images on all following slides come from Garrick Aden-Buie's excellent [tidyexplain](https://github.com/gadenbuie/tidyexplain)]
 
@@
 ]
+
+---
+
+# Joining Two *Different* Datasets: Overview
+
+.center[
+![](https://www.dropbox.com/s/3ao0g9yfwg46mdw/dplyr-joins.png?raw=1)
+
+From [R Studio Cheatsheet: Data Wrangling](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
+]
+
+---
+
+# References
+
+- `tibble`
+    - [*R For Data Science*, Chapter 10: Tibbles](https://r4ds.had.co.nz/tibbles.html)
+- `readr` and importing data
+    - [*R For Data Science*, Chapter 11: Data Import](https://r4ds.had.co.nz/data-import.html)
+    - [R Studio Cheatsheet: Data Import](https://www.rstudio.com/resources/cheatsheets/#import)
+- `dplyr` and data wrangling
+    - [*R For Data Science*, Chapter 5: Data Transformation](https://r4ds.had.co.nz/tibbles.html)
+    - [R Studio Cheatsheet: Data Wrangling](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) ([New version](https://www.rstudio.com/resources/cheatsheets/#dplyr))
+- `tidyr` and tidying or reshaping data
+    - [*R For Data Science*, Chapter 12: Tidy Data](https://r4ds.had.co.nz/tidy-data.html)
+    - [R Studio Cheatsheet: Data Wrangling](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
+    - [R Studio Cheatsheet: Data Import](https://www.rstudio.com/resources/cheatsheets/#import)
+- joining data
+    - [*R For Data Science*, Chapter 13: Relational Data](https://r4ds.had.co.nz/relational-data.html)
+    - [R Studio Cheatsheet: Data Transformation](https://www.rstudio.com/resources/cheatsheets/#dplyr)

### 2019-09-10 content/class/05-class.Rmd
Link to all lesson 5 content!

---
title: "1.3 Meet R - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

*Tuesday, September 10, 2019*

# Overview

Today we will cover the heart of `tidyverse` and use it for ["data wrangling"](https://en.wikipedia.org/wiki/Data_wrangling). Today will again be a lot of content thrown at you, so you can look back at this as a reference all semester. Then we will do more practice problems.

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/05-slides.html)

# Practice Problems

Today you will be working on [R practice problems](/r/05-r-practice.html). Check back later for solutions.

# Statistics and Software Survey

I have posted ["answers"](/static/homeworks/00-preliminary-survey-answers.html) to the statistics survey.

# Additional Useful Information

Check back here!

### 2019-09-10 content/r/05-r-practice.Rmd
Make practice 5 look better

@@
 
-## 
-#### 1. Load `gapminder`. `select()` the variables `year`, `lifeExp`, `country`.
+## 1.
+Load `gapminder`. `select()` the variables `year`, `lifeExp`, `country`.
 
-## 
-#### 2. `select()` all variables *except* `pop`.
+## 2.
+`select()` all variables *except* `pop`.
 
-## 
-#### 3. `rename()` `continent` to `cont`.
+## 3.
+`rename()` `continent` to `cont`.
 
 ## 4. 
-#### `arrange()` by year.
+`arrange()` by year.
 
 ## 5. 
-#### `arrange()` by year, but in descending order.
+`arrange()` by year, but in descending order.
 
 ## 6. 
-#### `arrange()` by year, then by life expectancy.
+`arrange()` by year, then by life expectancy.
 
 ## 7. 
-#### `filter()` observations with `pop` greater than 1 billion.
+`filter()` observations with `pop` greater than 1 billion.
 
@@
 ## 9.
-#### Try out the pipe (`%`) if you haven't already, by chaining commands: `select()` your data to look only at `year`, `gdpPercap`, and `country` in the year `1997`, for countries that have a `gdpPercap` greater than 20,000, and `arrange()` them alphabetically.
+Try out the pipe (`%`) if you haven't already, by chaining commands: `select()` your data to look only at `year`, `gdpPercap`, and `country` in the year `1997`, for countries that have a `gdpPercap` greater than 20,000, and `arrange()` them alphabetically.
 
 ## 10.
-#### `mutate()` a new variable called `GDP` that is equal to `gdpPercap * pop`.
+`mutate()` a new variable called `GDP` that is equal to `gdpPercap * pop`.
 
 ## 11.
-#### `mutate()` a new population variable that is the `pop` in millions.
+`mutate()` a new population variable that is the `pop` in millions.
 
 ## 12.
-#### `summarize()` to get the average GDP per capita.
+`summarize()` to get the average GDP per capita.
 
 ## 13.
-#### Get the number of observations, average, minimum, maximum, and standard deviation for GDP per capita.
+Get the number of observations, average, minimum, maximum, and standard deviation for GDP per capita.
 
 ## 14.
-#### Get the average GDP per capita over time. Hint, first `group_by()` year.
+Get the average GDP per capita over time. Hint, first `group_by()` year.
 
 ## 15.
-#### Get the average GDP per capita by continent.
+Get the average GDP per capita by continent.
 
 ## 16.
-#### Get the average GDP per capita by year and by continent.^[Hint: do `year` first, if you do `continent` first, there are no years to group by!] Then save this as another `tibble` called `gdp`.
+Get the average GDP per capita by year and by continent.^[Hint: do `year` first, if you do `continent` first, there are no years to group by!] Then save this as another `tibble` called `gdp`.
 Create a `ggplot` of a `line` graph of average continent GDP over time using the `gdp` data.
@@
 ## 17.
-#### Try it again all in one command with the pipe `%>%`. Instead of saving the data as `gdp`, pipe it right into `ggplot`!^[Hint: You can use `.` as a placeholder.]
+Try it again all in one command with the pipe `%>%`. Instead of saving the data as `gdp`, pipe it right into `ggplot`!^[Hint: You can use `.` as a placeholder.]
 
@@
 ## 18. 
+Import the data with `read_csv()` and assign it to `tibble` called `majors`.^[One way to avoid error messages is to move it to the same file as R's working directory, which again you can determine with `getwd()`.] The first argument of this command the name of the original file, in quotes.^[If the file is in a different folder, the argument is the full path in quotes.]
 
-#### Import the data with `read_csv()` and assign it to `tibble` called `majors`.^[One way to avoid error messages is to move it to the same file as R's working directory, which again you can determine with `getwd()`.] The first argument of this command the name of the original file, in quotes.^[If the file is in a different folder, the argument is the full path in quotes.]
-
-## 19. Look at the data with `glimpse()`. This is a suped-up version of `str()` in `tidyverse`.
+## 19.
+Look at the data with `glimpse()`. This is a suped-up version of `str()` in `tidyverse`.
 
@@
 ## 21.
-### Which major has the lowest unemployment rate?
+Which major has the lowest unemployment rate?
 
 ## 22. 
-#### What are the top 3 majors that have the highest percentage of women?
+What are the top 3 majors that have the highest percentage of women?
 
 ### 23. 
-#### Make a boxplot of `Median` wage by `Major_Category`.^[You won't be able to read the labels easily, so add `theme(axis.text.x=element_text(angle=45, hjust=1)` to angle `x`-axis labels (and move them down by 1)]
+Make a boxplot of `Median` wage by `Major_Category`.^[You won't be able to read the labels easily, so add `theme(axis.text.x=element_text(angle=45, hjust=1)` to angle `x`-axis labels (and move them down by 1)]
 
 ### 24.
-#### Which major category is the least popular in this sample?^[Hint: use `group_by` first.]
+Which major category is the least popular in this sample?^[Hint: use `group_by` first.]
 
 ### 25. 
-#### Is there a systematic difference between STEM majors and non-STEM majors? First define:
+Is there a systematic difference between STEM majors and non-STEM majors? First define:
 

### 2019-09-10 content/r/05-r-practice.Rmd
Fix some typos in R practice 5

@@
 ## 1.
-Load `gapminder`. `select()` the variables `year`, `lifeExp`, `country`.
+Load `gapminder`. First, `select()` the variables `year`, `lifeExp`, `country`.
 
@@
 ## 8. 
-#### Of those, look only at `India`.
+Of those, look only at `India`.
 
@@
 
-# Example I: Majors
+# Example: the Economics of College Majors
 
@@
 ## 18. 
-Import the data with `read_csv()` and assign it to `tibble` called `majors`.^[One way to avoid error messages is to move it to the same file as R's working directory, which again you can determine with `getwd()`.] The first argument of this command the name of the original file, in quotes.^[If the file is in a different folder, the argument is the full path in quotes.]
+Import the data with `read_csv()` and assign it to an object called `majors`.^[One way to avoid error messages is to move it to the same file as R's working directory, which again you can determine with `getwd()`.] The first argument of this command is the name of the original file, in quotes.^[If the file is in a different folder, the argument is the full path in quotes.]
 
@@
 ## 20.
-What are all of the unique values of `Major`? How many are there?
+What are all of the *unique* values of `Major`? How many are there?
 
 ## 21.
-Which major has the lowest unemployment rate?
+Which major has the *lowest* unemployment rate?
 
@@
 ```
-Next, make a variable `stem`, for whether or not a `Major_category` is `"stem"` or `"not_stem"`.^[Hint: try out the `ifelse()` function which has three inputs: condition(s) for a variable(s), what to do if `TRUE` (the if), and what to if `FALSE` (the else).]
+Next, make a variable called `stem`, for whether or not a `Major_category` is `"stem"` or `"not_stem"`.^[Hint: try out the `ifelse()` function which has three inputs: condition(s) for a variable(s), what to do if `TRUE` (the if), and what to if `FALSE` (the else), i.e.
+
+```{r, eval=F}
+ifelse(my_conditions,
+       yes = do_this_if_TRUE,
+       no = do_this_if_FALSE)
+```
+]
 

### 2019-09-11 content/r/04-r-practice.Rmd
Fixed link to survey answers on class 5, added R practice 1.4 answers and links

@@
 ```
+
 - [<i class="fas fa-file-pdf"></i> `Download as PDF`](/r/04-r-practice.pdf)
+- [<i class="fas fa-external-link-square-alt"></i> `Answer Key (html)`](/r/04-r-practice-answers.html)
+- [<i class="fas fa-file-pdf"></i> `Answer Key (PDF)`](/r/04-r-practice-answers-pdf.pdf)
+- [<i class="fas fa-registered"></i> `R Script`](/r/04-r-practice-answers.R)
 

### 2019-09-11 static/r/04-r-practice-answers-pdf.Rmd
Fixed link to survey answers on class 5, added R practice 1.4 answers and links

--- 2019-09-06 static/r/04-r-practice.Rmd
@@
 ---
-title: "1.4 Data Visualization with ggplot2 - R Practice"
+title: "1.4 Data Visualization with ggplot2 - R Practice (Answers)"
 author: "Ryan Safner"
@@
 
-### 1.
+## 1.
 We will look at GDP per Capita and Life Expectancy using some data from the [gapminder project](http://gapminder.org).
@@
 
-### 2. Let's get a quick look at `gapminder` to see what we're dealing with.
+```{r}
+# first time only
+# install.packages("gapminder")
+
+# load gapminder
+library(gapminder)
+
+# get help
+?gapminder
+```
+
+## 2. Let's get a quick look at `gapminder` to see what we're dealing with.
 
@@
 
+```{r}
+str(gapminder)
+```
+
 b. What variables are there?
+
+```{r}
+# - country: a factor
+# - continent: a factor
+# - year: an integer
+# - lifeExp: a number
+# - gdpPercap: a number
+```
 
@@
 
+```{r}
+head(gapminder)
+```
+
 d. Get `summary` statistics of all variables.
+
+```{r}
+summary(gapminder)
+```
 
@@
 
-### 3. Let's make sure you can do some basic plots before we get into the `gg`. Use base `R`'s `hist()` function to plot a *histogram* of `gdpPercap`.
+## 3. Let's make sure you can do some basic plots before we get into the `gg`. Use base `R`'s `hist()` function to plot a *histogram* of `gdpPercap`.
 
-### 4. Use base `R`'s `boxplot()` function to plot a *boxplot* of `gdpPercap`.
+```{r}
+hist(gapminder$gdpPercap)
+```
 
-### 5. Now make it a *boxplot* by `continent`.^[Hint: use formula notation with `~`.]
+## 4. Use base `R`'s `boxplot()` function to plot a *boxplot* of `gdpPercap`.
 
-### 6. Now make a *scatterplot* of `gdpPercap` on the $x$-axis and `LifeExp` on the $y$-axis.
+```{r}
+boxplot(gapminder$gdpPercap)
+```
+
+## 5. Now make it a *boxplot* by `continent`.^[Hint: use formula notation with`~`.]
+
+```{r}
+boxplot(gapminder$gdpPercap~gapminder$continent)
+
+# alternate method
+# boxplot(gdpPercap~continent, data = gapminder)
+```
+
+## 6. Now make a *scatterplot* of `gdpPercap` on the $x$-axis and `LifeExp` on the $y$-axis.
+
+```{r}
+plot(gapminder$lifeExp~gapminder$gdpPercap)
+
+# alternate method
+# boxplot(lifeExp~gdpPercap, data = gapminder)
+```
 
@@
 
-### 7. Load the package `ggplot2` (you should have installed it previously. If not, install first with `install.packages("ggplot2")`).
+## 7. Load the package `ggplot2` (you should have installed it previously. If not, install first with `install.packages("ggplot2")`).
 
-### 8.
+```{r}
+# install if you don't have
+# install.packages("ggplot2")
+
+# load ggplot2 
+library(ggplot2)
+```
+
+## 8.
 Let's first make a `bar` graph to see how many countries are in each continent.
@@
 
-### 9. For quantitative data, we want a `histogram` to visualize the distribution of a variable. Make a `histogram` of `gdpPercap`. Your only `aes`thetic here is to map `gdpPercap` to `x`.
+```{r}
+ggplot(data = gapminder,
+       aes(x = continent))+
+  geom_bar()
+```
 
-### 10.
+## 9. For quantitative data, we want a `histogram` to visualize the distribution of a variable. Make a `histogram` of `gdpPercap`. Your only `aes`thetic here is to map `gdpPercap` to `x`.
+
+```{r}
+ggplot(data = gapminder,
+       aes(x = gdpPercap))+
+  geom_histogram()
+```
+
+## 10.
 Now let's try adding some color, specifically, add an `aes`thetic that maps `continent` to `fill`.^[In general, `color` refers to the outside borders of a `geom` (except points), `fill` is the interior of an object.]
 
-### 11. Instead of a `histogram`, change the `geom` to make it a `density` graph. To avoid overplotting, add `alpha=0.4` to the `geom` argument (alpha changes the *transparency* of a `fill`).
+```{r}
+ggplot(data = gapminder,
+       aes(x = gdpPercap,
+           fill = continent))+
+  geom_histogram()
+```
 
-### 12. Redo your plot from 11 for `lifeExp` instead of `gdpPercap`.
+## 11. Instead of a `histogram`, change the `geom` to make it a `density` graph. To avoid overplotting, add `alpha=0.4` to the `geom` argument (alpha changes the *transparency* of a `fill`).
 
-### 13. Now let's try a scatterplot for `lifeExp` (as `y`) on `gdpPercap` (as `x`). You'll need both for `aes`thetics. The `geom` here is `geom_point()`.
+```{r}
+ggplot(data = gapminder,
+       aes(x = gdpPercap,
+           fill = continent))+
+  geom_density(alpha=0.4)
+```
 
-### 14. Add some color by mapping `continent` to `color` in your `aes`thetics.
+## 12. Redo your plot from 11 for `lifeExp` instead of `gdpPercap`.
 
-### 15. Now let's try adding a regression line with `geom_smooth()`. Add this layer on top of your `geom_point()` layer.
+```{r}
+ggplot(data = gapminder,
+       aes(x = lifeExp,
+           fill = continent))+
+  geom_density(alpha=0.4)
+```
 
-### 16.
+## 13. Now let's try a scatterplot for `lifeExp` (as `y`) on `gdpPercap` (as `x`). You'll need both for `aes`thetics. The `geom` here is `geom_point()`.
+
+```{r}
+ggplot(data = gapminder,
+       aes(x = gdpPercap,
+           y = lifeExp))+
+  geom_point()
+```
+
+## 14. Add some color by mapping `continent` to `color` in your `aes`thetics.
+
+```{r}
+ggplot(data = gapminder,
+       aes(x = gdpPercap,
+           y = lifeExp,
+           color = continent))+
+  geom_point()
+```
+
+## 15. Now let's try adding a regression line with `geom_smooth()`. Add this layer on top of your `geom_point()` layer.
+
+```{r}
+ggplot(data = gapminder,
+       aes(x = gdpPercap,
+           y = lifeExp,
+           color = continent))+
+  geom_point()+
+  geom_smooth()
+```
+
+## 16.
 Did you notice that you got multiple regression lines (colored by continent)?
@@
 
-### 17. Now add an `aes`thetic to your `point`s to map `pop` to `size`.
+```{r}
+ggplot(data = gapminder,
+       aes(x = gdpPercap,
+           y = lifeExp))+
+  geom_point(aes(color = continent))+
+  geom_smooth()
+```
 
-### 18.
+## 17. Now add an `aes`thetic to your `point`s to map `pop` to `size`.
+
+```{r}
+ggplot(data = gapminder,
+       aes(x = gdpPercap,
+           y = lifeExp))+
+  geom_point(aes(color = continent, 
+                 size = pop))+
+  geom_smooth()
+```
+
+## 18.
 Change the color of the regression line to `"black"`.
@@
 
-### 19. Another way to separate out continents is with `facet`ing. Add `+facet_wrap(~continent)` to create subplots by `continent`.
+```{r}
+ggplot(data = gapminder,
+       aes(x = gdpPercap,
+           y = lifeExp))+
+  geom_point(aes(color = continent, 
+                 size = pop))+
+  geom_smooth(aes(color = "black"))
 
-### 20. Remove the `facet` layer. The `scale` is quite annoying for the `x`-axis, a lot of points are clustered on the lower level. Let's try changing the scale by adding a layer: `+scale_x_log10()`.
+# putting it inside aesthetics tries to map color to something
+# in the da ta called "black", since R can't find "black", 
+# it will produce some random color
 
-### 21.
+ggplot(data = gapminder,
+       aes(x = gdpPercap,
+           y = lifeExp))+
+  geom_point(aes(color = continent, 
+                 size = pop))+
+  geom_smooth(color = "black")
+
+# putting it outside aesthetics (correctly) sets color to black
+```
+
+## 19. Another way to separate out continents is with `facet`ing. Add `+facet_wrap(~continent)` to create subplots by `continent`.
+
+```{r}
+ggplot(data = gapminder,
+       aes(x = gdpPercap,
+           y = lifeExp))+
+  geom_point(aes(color = continent, 
+                 size = pop))+
+  geom_smooth(color = "black")+
+  facet_wrap(~continent)
+```
+
+## 20. Remove the `facet` layer. The `scale` is quite annoying for the `x`-axis, a lot of points are clustered on the lower level. Let's try changing the scale by adding a layer: `+scale_x_log10()`.
+
+```{r}
+ggplot(data = gapminder,
+       aes(x = gdpPercap,
+           y = lifeExp))+
+  geom_point(aes(color = continent, 
+                 size = pop))+
+  geom_smooth(color="black")+
+  scale_x_log10()
+```
+
+## 21.
 Now let's fix the labels by adding `+labs()`.
@@
 
-### 22.
+```{r}
+ggplot(data = gapminder,
+       aes(x = gdpPercap,
+           y = lifeExp))+
+  geom_point(aes(color = continent, 
+                 size = pop))+
+  geom_smooth(color="black")+
+  scale_x_log10()+
+  labs(x = "GDP per Capita",
+       y = "Life Expectancy",
+       color = "Continent",
+       size = "Population")
+```
+
+## 22.
 Now let's try subsetting by looking only at North America.
@@
 
-### 23. Try this again for the *whole* world, but just for observations in the year 2002.
+```{r}
+america<-gapminder[gapminder$continent=="Americas",]
+
+# verify this worked
+america
+
+ggplot(data = america,
+       aes(x = gdpPercap,
+           y = lifeExp))+
+  geom_point(aes(color = continent, 
+                 size = pop))+
+  geom_smooth()
+```
+
+## 23. Try this again for the *whole* world, but just for observations in the year 2002.
+
+```{r}
+gap_2002<-gapminder[gapminder$year==2002,]
+
+# verify this worked
+gap_2002
+
+ggplot(data = gap_2002,
+       aes(x = gdpPercap,
+           y = lifeExp))+
+  geom_point(aes(color = continent, 
+                 size = pop))+
+  geom_smooth()
+```

### 2019-09-11 static/r/04-r-practice-answers.Rmd
Fixed link to survey answers on class 5, added R practice 1.4 answers and links

--- 2019-09-11 static/r/04-r-practice-answers-pdf.Rmd
@@
 ## 1.
-We will look at GDP per Capita and Life Expectancy using some data from the [gapminder project](http://gapminder.org).
+#### We will look at GDP per Capita and Life Expectancy using some data from the [gapminder project](http://gapminder.org).
 There is a [handy package](https://github.com/jennybc/gapminder) called `gapminder` that uses a small snippet of this data for exploratory analysis.
@@
 
-## 2. Let's get a quick look at `gapminder` to see what we're dealing with.
+## 2
+#### Let's get a quick look at `gapminder` to see what we're dealing with.
 
-a. Get the `str`ucture of the `gapminder` data.
+### a.
+#### Get the `str`ucture of the `gapminder` data.
 
@@
 
-b. What variables are there?
+### b.
+#### What variables are there?
 
@@
 
-c. Look at the `head` of the dataset to get an idea of what the data looks like.
+### c.
+#### Look at the `head` of the dataset to get an idea of what the data looks like.
 
@@
 
-d. Get `summary` statistics of all variables.
+### d.
+#### Get `summary` statistics of all variables.
 
@@
 
-## 3. Let's make sure you can do some basic plots before we get into the `gg`. Use base `R`'s `hist()` function to plot a *histogram* of `gdpPercap`.
+## 3.
+#### Let's make sure you can do some basic plots before we get into the `gg`. Use base `R`'s `hist()` function to plot a *histogram* of `gdpPercap`.
 
@@
 
-## 4. Use base `R`'s `boxplot()` function to plot a *boxplot* of `gdpPercap`.
+## 4. 
+#### Use base `R`'s `boxplot()` function to plot a *boxplot* of `gdpPercap`.
 
@@
 
-## 5. Now make it a *boxplot* by `continent`.^[Hint: use formula notation with`~`.]
+## 5. 
+#### Now make it a *boxplot* by `continent`.^[Hint: use formula notation with`~`.]
 
@@
 
-## 6. Now make a *scatterplot* of `gdpPercap` on the $x$-axis and `LifeExp` on the $y$-axis.
+## 6. 
+#### Now make a *scatterplot* of `gdpPercap` on the $x$-axis and `LifeExp` on the $y$-axis.
 
@@
 
-## 7. Load the package `ggplot2` (you should have installed it previously. If not, install first with `install.packages("ggplot2")`).
+## 7. 
+#### Load the package `ggplot2` (you should have installed it previously. If not, install first with `install.packages("ggplot2")`).
 
@@
 ## 8. 
-Let's first make a `bar` graph to see how many countries are in each continent.
+#### Let's first make a `bar` graph to see how many countries are in each continent.
 The only `aes`thetic you need is to map `continent` to `x`.
@@
 
-## 9. For quantitative data, we want a `histogram` to visualize the distribution of a variable. Make a `histogram` of `gdpPercap`. Your only `aes`thetic here is to map `gdpPercap` to `x`.
+## 9. 
+#### For quantitative data, we want a `histogram` to visualize the distribution of a variable. Make a `histogram` of `gdpPercap`. Your only `aes`thetic here is to map `gdpPercap` to `x`.
 
@@
 ## 10. 
-Now let's try adding some color, specifically, add an `aes`thetic that maps `continent` to `fill`.^[In general, `color` refers to the outside borders of a `geom` (except points), `fill` is the interior of an object.]
+#### Now let's try adding some color, specifically, add an `aes`thetic that maps `continent` to `fill`.^[In general, `color` refers to the outside borders of a `geom` (except points), `fill` is the interior of an object.]
 
@@
 
-## 11. Instead of a `histogram`, change the `geom` to make it a `density` graph. To avoid overplotting, add `alpha=0.4` to the `geom` argument (alpha changes the *transparency* of a `fill`).
+## 11. 
+#### Instead of a `histogram`, change the `geom` to make it a `density` graph. To avoid overplotting, add `alpha=0.4` to the `geom` argument (alpha changes the *transparency* of a `fill`).
 
@@
 
-## 12. Redo your plot from 11 for `lifeExp` instead of `gdpPercap`.
+## 12. 
+#### Redo your plot from 11 for `lifeExp` instead of `gdpPercap`.
 
@@
 
-## 13. Now let's try a scatterplot for `lifeExp` (as `y`) on `gdpPercap` (as `x`). You'll need both for `aes`thetics. The `geom` here is `geom_point()`.
+## 13. 
+#### Now let's try a scatterplot for `lifeExp` (as `y`) on `gdpPercap` (as `x`). You'll need both for `aes`thetics. The `geom` here is `geom_point()`.
 
@@
 
-## 14. Add some color by mapping `continent` to `color` in your `aes`thetics.
+## 14. 
+#### Add some color by mapping `continent` to `color` in your `aes`thetics.
 
@@
 
-## 15. Now let's try adding a regression line with `geom_smooth()`. Add this layer on top of your `geom_point()` layer.
+## 15. 
+#### Now let's try adding a regression line with `geom_smooth()`. Add this layer on top of your `geom_point()` layer.
 
@@
 ## 16. 
-Did you notice that you got multiple regression lines (colored by continent)?
+#### Did you notice that you got multiple regression lines (colored by continent)?
 That's because we set a `global` `aes`thetic of mapping `continent` to `color`.
@@
 
-## 17. Now add an `aes`thetic to your `point`s to map `pop` to `size`.
+## 17. 
+#### Now add an `aes`thetic to your `point`s to map `pop` to `size`.
 
@@
 ## 18. 
-Change the color of the regression line to `"black"`.
+#### Change the color of the regression line to `"black"`.
 Try first by putting this inside an `aes()` in your `geom_smooth`, and try a second time by just putting it inside `geom_smooth` without an `aes()`.
@@
 
-## 19. Another way to separate out continents is with `facet`ing. Add `+facet_wrap(~continent)` to create subplots by `continent`.
+## 19. 
+#### Another way to separate out continents is with `facet`ing. Add `+facet_wrap(~continent)` to create subplots by `continent`.
 
@@
 
-## 20. Remove the `facet` layer. The `scale` is quite annoying for the `x`-axis, a lot of points are clustered on the lower level. Let's try changing the scale by adding a layer: `+scale_x_log10()`.
+## 20. 
+#### Remove the `facet` layer. The `scale` is quite annoying for the `x`-axis, a lot of points are clustered on the lower level. Let's try changing the scale by adding a layer: `+scale_x_log10()`.
 
@@
 ## 21. 
-Now let's fix the labels by adding `+labs()`.
+#### Now let's fix the labels by adding `+labs()`.
 Inside `labs`, make proper axes titles for `x`, `y`, and a `title` to the plot.
@@
 ## 22. 
-Now let's try subsetting by looking only at North America.
+#### Now let's try subsetting by looking only at North America.
 Take the `gapminder` dataframe and subset it to only look at `continent=="Americas"`).
@@
 
-## 23. Try this again for the *whole* world, but just for observations in the year 2002.
+## 23. 
+#### Try this again for the *whole* world, but just for observations in the year 2002.
 

### 2019-09-11 content/assignment/00-preliminary-survey.Rmd
Fix links to survey answers on survey page

--- 2019-09-06 content/assignment/00-preliminary-survey.Rmd
@@
 - [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/00-preliminary-survey.pdf)
-- [<i class="fas fa-external-link-square-alt"></i> `Answer Key`](/homeworks/00-preliminary-survey-answers-pdf.pdf)
-- [<i class="fas fa-file-pdf"></i> `Answer Key`](/homeworks/00-preliminary-survey-answers.html)
+- [<i class="fas fa-external-link-square-alt"></i> `Answer Key (html)`](/homeworks/00-preliminary-survey-answers.html)
+- [<i class="fas fa-file-pdf"></i> `Answer Key (PDF)`](/homeworks/00-preliminary-survey-answers-pdf.pdf)
 

### 2019-09-11 content/reading/06-reading.Rmd
Update schedule format, readings for 5 and 6

---
title: "1.6 Optimize Your Workflow: Projects and Markdown - Readings"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-fullnote-bibliography-no-bib.csl
---

# Strongly Recommended

- <i class="fas fa-external-link-square-alt"></i>[Chs. 8, 27, 28, 29, 30 (but especially 27)](http://r4ds.had.co.nz) in @R4DS

# Other Resources

## On R Markdown
- Xie, Yihui, J.J. Allaire, and Garett Grolemund, [*R Markdown: The Definitive Guide*](https://bookdown.org/yihui/rmarkdown/)
- RStudio, [*R Markdown Reference Guide*](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)
- RStudio, [*Overview of R Markdown*](https://rmarkdown.rstudio.com/)
- Rstudio, [*R Markdown Cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)

## On Optimized Workflow

- Healey, Kieran, [*The Plain Person’s Guide to Plain Text Social Science*](http://plain-text.co/)
- Bryan, Jenny and Jim Hester, [*Happy Git and GitHub for the useR*](https://happygitwithr.com/)

### 2019-09-12 content/class/06-class.Rmd
Update lesson 6 stuff

---
title: "1.6: Optimize Your Workflow: Projects, Markdown, and Git - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

*Tuesday, September 10, 2019*

# Overview

Today I will give you about half the class period to finish your practice problems and I can answer questions.

# Knitting on Trading Room Computers

Last year, IT and I were unable to get the Trading Room computers to knit to pdf. I may have a workaround. We will install a minimal version of $\LaTeX{}$ inside of R Studio. Run the following code:

```{r, eval=F}
tinytex::install_tinytex()
```

Then make a new `.Rmd` and try to `knit` it to `pdf`. It may install several packages (`knitr` and `rmarkdown` at least) and ask you a few prompts. It should then work.

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/05-slides.html)

# Practice Problems

Today you will be working on [R practice problems](/r/05-r-practice.html). Check back later for solutions.

# Statistics and Software Survey

I have posted ["answers"](/homeworks/00-preliminary-survey-answers.html) to the statistics survey.

# Additional Useful Information

Check back here!

### 2019-09-12 static/slides/06-slides.Rmd
Update lesson 6 stuff

---
title: "1.6: Optimize Workflow: Projects, Markdown, and Git"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(message=F,
                      warning=F)
library(tidyverse)

```

class: inverse, middle, center

.pull-left[

### [The Office Model](#4)

### [The Plain Text Model](#12)

### [R Markdown](#18)

### [Compiling Your Documents](#43)

### [R Projects](#46)

### [Version Control](#55)

### [Resources](#69)

]

.pull-right[

### [dplyr::select(): select variables](#60)

### [dplyr::rename(): rename variables](#73)

### [dplyr::mutate(): create new variables](#77)

### [dplyr::summarize(): create statistics](#98)

### [tidyr: reshaping data](#133)

### [dplyr: combining datasets](#134)

]

---

# Your Workflow Has a Lot of Moving Parts

.pull-left[

1. Writing text/documents

2. Managing citations and bibliographies

3. Performing data analysis

4. Making figures and tables

5. Saving files for future use

6. Monitoring changes in documents

7. Collaborating and sharing with others

8. Combining into a deliverable (report, paper, presentation, etc.)
]

.pull-right[
.center[
![](https://www.dropbox.com/s/xtg1neu883j6r10/workflow.jpg?raw=1)
]
]
---

class: inverse, center, middle
# The Office Model

---

# The Office Model I

.pull-left[

1. Writing text/documents

2. Managing citations and bibliographies

3. Performing data analysis

4. Making figures and tables

5. Saving files for future use

6. Monitoring changes in documents

7. Collaborating and sharing with others

8. Combining into a deliverable (report, paper, presentation, etc.)
]

.pull-right[
.center[
![](https://www.dropbox.com/s/9zu9f3x4rhec0dt/msofficemodel.png?raw=1)
]
]

---

# The Office Model II

.pull-left[

- A lot of **copy-pasting**

- A lot of...

.center[
![](https://www.dropbox.com/s/1b2gwpck4pdd7gl/wysiwygproblem.jpg?raw=1)
]
]

.pull-right[
.center[
![](https://www.dropbox.com/s/9zu9f3x4rhec0dt/msofficemodel.png?raw=1)
]
]

---

# The Office Model: A Short Horror Movie

.center[
<iframe width="560" height="315" src="https://www.youtube.com/embed/s3JldKoA0zw" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
]

---

# The Office Model: Mistakes

.pull-left[
.center[
![](https://www.dropbox.com/s/cqcy9l5czztgpu3/excelfail1.png?raw=1)

Source: [Science Magazine](https://www.sciencemag.org/news/2016/08/one-five-genetics-papers-contains-errors-thanks-microsoft-excel)
]
]

.pull-right[
.center[
![](https://www.dropbox.com/s/1a7mdlthfv6r8x8/excelfail2.png?raw=1)

Source: [Bloomberg](https://www.bloomberg.com/news/articles/2013-04-18/faq-reinhart-rogoff-and-the-excel-error-that-changed-history)
]

]

---

# The Office Model: Not Reproducible

.center[
<blockquote class="twitter-tweet"><p lang="en" dir="ltr">&quot;&#39;Reproducible research&#39; is a redundant term. &#39;Irreproducible research&#39; just used to be known as &#39;bullshit&#39;.&quot; - <a href="https://twitter.com/fperez_org?ref_src=twsrc%5Etfw">@fperez_org</a> ::slow clap::</p>&mdash; Kaitlin Thaney 💁🏻 (@kaythaney) <a href="https://twitter.com/kaythaney/status/464543147083968513?ref_src=twsrc%5Etfw">May 8, 2014</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script> 
]

---

# What I'm About to Show You

.center[
![](https://www.dropbox.com/s/ze75ntay9kp3yjf/drawanowl.jpg?raw=1)
]

---

# What I'm About to Show You

.pull-left[
- This is how I make my...
    - Research papers
    - Course documents
    - Websites
    - Slides and presentations

- I have not used any MS Office products since 2011 (good riddance!)

- **This stuff is optional**
    - If you like your office model, you can keep it
    - But this is what most people who take this course continue to use (R is only really if you have data work)

]

.pull-right[
.center[
![](https://www.dropbox.com/s/546o01icserlone/rmdout.png?raw=1)
]
]
---

class: inverse, center, middle

# The Plain Text Model

---

# The Plain Text Model I

.pull-left[

- Meet `R Markdown`, which can do *all of this* in one pipeline

1. Writing text/documents
2. Managing citations and bibliographies
3. Performing data analysis
4. Making figures and tables
5. Saving files for future use
6. Monitoring changes in documents
7. Collaborating and sharing with others
8. Combining into a deliverable (report, paper, presentation, etc.)
]

.pull-right[
.center[
![](https://www.dropbox.com/s/546o01icserlone/rmdout.png?raw=1)

![From [R Studio's R Markdown Cheatsheet](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)](images/rmd2.png)

]
]

---

# The Plain Text Model II

- .shout[Plain text] files: readable by *both* machines and humans
    - Understand how a document is structured and formatted via code and markup to text

- Focus entirely on the *actual writing of the content* instead of the formatting and aesthetics
    - You can still customize, but with precise commands instead of point, click, drag, guess, pray

- **Open Source**: free, useable forever, often very small file size
    - Proprietary software is a gamble - can you still open a `.doc` file from Microsoft Word 1997?

- **Automate and Minimize Errors**, especially in repetitive processes

- Can be used with **version control** (see below)

---

# Making Your Work Reproducible

> One day you will need to quit R, go do something else and return to your analysis the next day. One day you will be working on multiple analyses simultaneously that all use R and you want to keep them separate. One day you will need to bring data from the outside world into R and send numerical results and figures from R back out into the world. To handle these real life situations, you need to make two decisions:
What about your analysis is "real", i.e. what will you save as your lasting record of what happened?
Where does your analysis "live"? 
- Hadley Wickham, [R For Data Science](http://r4ds.had.co.nz/workflow-projects.html)

- We've talked about `.R` script files that let you "keep" commands

- What about output? Must you save and copy/paste to MS Word?

- No!

---

# Making Your Work Reproducible

- `R Markdown` files (`.Rmd`) is the "real" part of your analysis, *everything* can live in this plain-text file.

- Document text in `markdown`

- `R code` executed in "chunks"

- Plots and tables generated from `R code`

- Citations and bibliography automated with `.bib` file

---

# The Future of Science is Open Source Plain Text

.pull-left[
.center[
![](https://www.dropbox.com/s/lw9fouq57tpic9p/notebooks1.png?raw=1)

Source: [The Atlantic](https://www.theatlantic.com/science/archive/2018/04/the-scientific-paper-is-obsolete/556676/)
]
]

.pull-right[
.center[
![](https://www.dropbox.com/s/e8pnjaqp2xbovit/notebooks2.png?raw=1)

Source: [Paul Romer (2018 Economics Nobel)](https://paulromer.net/jupyter-mathematica-and-the-future-of-the-research-paper/)
]
]

---

class: inverse, center, middle
# R Markdown

---

# Creating an R Markdown Document I

.pull-left[

- `File -> New File -> R Markdown...`

- Outputs:
    - Document (what you'll use for most things)
    - Presentation (for making slides in various formats)
    - Shiny (an html and R based web app, advanced)
    - Templates (some built-in, other packages like `rticles` or `xaringan` add neat templates)

]

.pull-right[
.center[
![](https://www.dropbox.com/s/czklhgt5llrekjc/newrmarkdown.png?raw=1)
]
]

---

# Creating an R Markdown Document II

- `File -> New File -> R Markdown...`

- Documents:
    - `html`: renders a webpage, viewable in any browser
        - default, easiest to produce and share
        - can have interactive elements (gifs, animations, web apps)
        - requires internet connection to host and share (*you* can view offline)
    - `pdf`: renders a PDF document
        - most common document format around
        - requires `LaTeX` distribution to render (more on that soon)
    - `word`: create a Micosoft Word document
        - ...if you must

---

# Structure of an `R Markdown` Document

- Entire document is written in a single file:<sup>.red[1]<sup>

1. `YAML` header for metadata

2. Text of the document written with `markdown`

3. `R` chunks for data analysis, plots, figures, tables, statistics, as necessary

.footnote[<sup>.red[1]</sup> The one exception is for managing bibliographies, this requires an additional `.bib` file!]

---

# YAML Header I

- Top of a document contains the `YAML`<sup>.red[1]</sup> separated by three dashes `---` above and below

- Contains the **metadata** of the document, such as:

```{yaml}
title: "My Title"
author: "Ryan Safner"
date: "`r Sys.Date()`"
output: pdf_document
```

- `output` must be specified, everything else can be left blank, and other options can be added as necessary

- In most cases, you can safely ignore other things in the `yaml` until you are ready

- `output` must be 
.footnote[<sup>.red[1]</sup> YAML stands for "YAML Ain't Markup Language." Nerds love recursive acronyms.]

---

# YAML Header: Example from one of my research papers

```{yaml}
title: Distributing Patronage^[I would like to thank the Board of Associates of Hood College...]
subtitle: Intellectual Property in the Transition from Natural State to Open Access Order
date: \today
author: 
- Ryan Safner^[Hood College, Department of Economics and Business Administration; safner@hood.edu]

abstract: |
  | "This paper explores the emergence of the modern forms of copyright and patent in ...
  | *JEL Classification:* O30, O43, N43
  | *Keywords:* Copyright, intellectual property, economic history, freedom of the press, economic development

bibliography: patronage.bib
geometry: margin = 1in
fontsize: 12pt
mainfont: Fira Sans Condensed
output: 
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    fig_caption: yes

header-includes:
    - \usepackage{booktabs}
```

---

# R Chunks I

- You can create a "chunk" of `R` code with three backticks<sup>.red[1]</sup> above and below your code
  - After the first pair of backticks, signify the language of the code<sup>.red[2]</sup> inside braces, e.g:

.pull-left[

## Input

````
```{r, echo=T, eval=F}
2+2
```
````
]

.pull-right[

## Output

```{r, echo=T}
2+2
```

]

.footnote[<sup>.red[1]</sup> The key to the left of the #1 key.

<sup>.red[2]</sup> Yes that does mean you can use other coding languages!]

---

# R Chunks II

.pull-left[

## Input

````
```{r, echo=T, eval=F}
head(mpg, n=2)
```
````
]

.pull-right[

## Output

```{r, echo=T}
head(mpg, n=2)
```

]

---

# R Chunks III

.pull-left[

## Input

````
```{r, echo=T, eval=F}
library("ggplot2")
ggplot(data = mpg)+
  aes(x = displ)+
  geom_histogram()
```
````
]

.pull-right[

## Output

```{r, echo=T, fig.height=4}
library("ggplot2")
ggplot(data = mpg)+
  aes(x = displ)+
  geom_histogram()
```
]

---

# R Chunks Options

- You can add [additional options](this guide](https://yihui.name/knitr/options/)) inside the `{braces}` after `r`, some common options:

- **Name**: you can name your chunk for further reference later (not required)<sup>.red[1]</sup>

- `echo` set `=TRUE` to display the `R` code input; `=FALSE` shows will not show your code

- `eval` set `=TRUE` to run your code; `=FALSE` only displays your code without running it

- `fig` has a lot of options for displaying plot outputs (`fig.height`, `fig.width`, `fig.asp`, etc)

- `results` will format the output of a chunk in a certain way (used for advanced things we'll talk about later)

.footnote[<sup>.red[1]</sup> The chunk name is the only thing that needs to go right next to `r` without a comma. Everything else needs to go after a comma, e.g. `{r my_chunk, echo=F}`]

---

# R Inline Code I

- If you just want to display some code (or at least format it like code) in the middle of a sentence, `place between a backtick on either side.` That's what I do when I mention `tidyverse` or `gapminder`.

- If you want to actually *run* something from `R` and have it show up in the middle of a sentence, put `r` as the first character inside the backticks, and then run the actual code such as pi is equal to `r pi`.

.pull-left[

## Input

pi is equal to 
```{r, results='asis', echo=FALSE}
cat("`` `r pi` ``")
```

]

.pull-right[

## Output
pi is equal to `r pi`

]

---

# R Inline Code II

.pull-left[
# Code

  ```{r, eval=F, echo=T}
  library(gapminder)
  ```

The average GDP per capita is $`` `r
round(mean(gapminder$gdpPercap),2)` `` with a standard deviation of $`` `r
round(sd(gapminder$gdpPercap),2)` ``.  
]

.pull-right[
# Output

```{r, echo=F}
library(gapminder)
```

The average GDP per capita is $`r round(mean(gapminder$gdpPercap),2)` with a standard deviation of $`r round(sd(gapminder$gdpPercap),2)`.

]

---

# Writing Text with Markdown

- [Markdown](https://en.wikipedia.org/wiki/Markdown) is a lightweight markup language geared towards HTML (i.e. the internet)
    - [Markup languages](https://en.wikipedia.org/wiki/Markup_language) used to add commands about how to display plain text
- Very simple and intuitive
- Write normal text as usual in any word processor
- Change font styling with tags (asterisks): 
    - `*italics text*` creates *italics text*
    - `**bold text**` creates **bold text**

---

# Writing Text with Markdown: Lists

- Create an unordered list with lines of (- or + or * ), e.g.:

.pull-left[

## Input

```
- item 1
- item 2
  - item 2a
- item 3
```

]

.pull-right[

## Output

- item 1
- item 2
  - item 2a
- item 3

]
---

# Writing Text with Markdown: Headings and Comments

- Section headings are created with a `#` hashtag at the front of a line
    - e.g. `# My Title` creates a main heading
    - more hashtags create subheadings, e.g. `## My Sub-Heading`, `### My Sub-sub Heading`

- You can write comments in the file that will not print in the final version with `<!-- Unprinted comments here -->` (this comes from html)

---

# Writing Text with Markdown: Tables

.pull-left[

## Input

```markdown

| Header 1 | Header 2 | 
|----------|----------|
| Cell 1   | Cell 2   |
| Cell 3   | Cell 4   |
```

]

.pull-right[

| Header 1 | Header 2 | 
|----------|----------|
| Cell 1   | Cell 2   |
| Cell 3   | Cell 4   |

]

- For more complicated tables, there are other packages and techniques

---

# Writing Math I

- Add beautifully-formatted math with the `$` tag before and after the math, two `$$` before/after for a centered equation

- In-line math example: `$1^2=\frac{\sqrt{16}}{4}$` produces $1^2=\frac{\sqrt{16}}{4}$

- Centered-equation example:

.pull-left[

## Code

`$$` `\hat{\beta_1}=\frac{\displaystyle \sum_{i=1}^n (X_i-\bar{X})(Y_i-\bar{Y})}{\displaystyle \sum_{i=1}^n (X_i-\bar{X})^2}` `$$`

]

.pull-right[

## Output

$$\hat{\beta_1}=\frac{\displaystyle \sum_{i=1}^n (X_i-\bar{X})(Y_i-\bar{Y})}{\displaystyle \sum_{i=1}^n (X_i-\bar{X})^2}$$

]
---

# Writing Math II

- Math uses a (much older) language called [LaTeX](http://en.wikipedia.org/wiki/LaTeX), used by mathematicians, economists, and others to write papers and slides with perfect math and formatting
    - I used to use for everything before I found `R` and `markdown` 
    - Producing `pdf` or `html` output actually converts `markdown` files into $\TeX{}$ first! (See [the process described below](#process))
    - Much steeper learning curve, [a good cheatsheet](https://wch.github.io/latexsheet/latexsheet.pdf)
    - An extensive library of mathematical symbols, notation, formats, and ligatures, e.g.

---

# Writing Math III

| Input | Output |
|----|----|
| `$\alpha$` | $\alpha$ |
| `$\pi$` | $\pi$ | 
| `$\frac{1}{2}$` | $\frac{1}{2}$ | 
| `$\hat{x}$` | $\hat{x}$ | 
| `$\bar{y}$` | $\bar{y}$ | 
| `$x_{1,2}$` | $x_{1,2}$ | 
| `x^{a-1}$` | $x^{a-1}$ |
| `$\lim_{x \to \infty}$` | $\lim_{x \to \infty}$ | 
| `$A=\begin{bmatrix} a_{1,1} & a_{1,2} \\ a_{2,1} & a_{2,2} \\ \end{bmatrix}$` | $A=\begin{bmatrix} a_{1,1} & a_{1,2} \\ a_{2,1} & a_{2,2} \\ \end{bmatrix}$ |

- A great resource: [Wikibooks LaTeX Mathematics chapter](https://en.wikibooks.org/wiki/LaTeX/Mathematics)
- And for more about $\LaTeX{}$ in general: [Wikibooks LaTeX](https://en.wikibooks.org/wiki/LaTeX/)

---

# Citations, References, and Bibliography

- Manage your citations and bibliography automatically with `.bib` files 
- First create a `.bib` file to list all of your references in
    - You can do this in `R` via: File -> New File -> Text File (and save with `.bib` at the end)
    - See `examplebib.bib` in this repository used in this document 
    - At the top of your `YAML` header in the main document, add `bibliography: examplebib.bib` so `R` knows to pull references from this file 
    - For each reference, add information to a `.bib` file, like so:

---

# An Example .bib File

.pull-left[

```
@article{safner2016,
  author = {Ryan Safner},
  year = {2016},
  journal = {Journal of Institutional Economics},
  title = {Institutional Entrepreneurship, Wikipedia, 
           and the Opportunity of the Commons},
  volume = {12},
  number = {4},
  pages = {743-771}
}
```
]

.pull-right[
- A `.bib` file is a plain text file with entries like this

- Classes for `@article`, `@book`, `@collectedwork`, `@unpublished`, etc. 
    - Each will have different keys needed (e.g. `editor`, `publisher`, `address`)

- First term after the `{` is your **citation key** (e.g. `safner2016`)
]
---

# An Example .bib File

- Whenever you want to cite a work in your text, call up the **citation key** with `@`, like so: `@safner2016[]`, which produces @safner2016[]
    - You can customize citations, e.g.:
        - **All in parentheses**: `[@safner2016]` produces [@safner2016]
        - **No parentheses**: `@safner2016` produces @safner2016
        - **Date only**: `[-@safner2016]` produces [-@safner2016]
        - **Adding pages or comments**: `@safner2016[p.743-744 and passim]` produces @safner2016[p.743-744 and passim]
- BibTeX will automatically collect all works cited at the end and produce a **bibliography** according to a style you can choose

---

# Reference Management Software

- For more information and examples, see [R Studio's R Markdown Guide on Bibliographies](https://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html)
- Lot of programs can help you manage references and export complete `.bib` files to use with `R Markdown`
    - [Mendeley](https://www.mendeley.com) and [Zotero](https://www.zotero.org/) are free and cross-platform
    - I use [Papers](https://www.readcube.com/papers/) (Paid and Mac only)
    - Simplest program (what I use) that makes `.bib` files is [Bibdesk](https://bibdesk.sourceforge.io/)

---

# Plain-Text Editors

- Markdown files are [**plain text**](https://en.wikipedia.org/wiki/Plain_text) files and can be edited in *any* text editor
    - something as basic (and boring!) as "**Notepad**," for example
    - many good [text editors](https://en.wikipedia.org/wiki/Text_editor) out there

- Any good editor will have **syntax highlighting** and **coloring** when you use tags (like **bold**, *italic*, `code`, and `code #comments`).

.pull-left[
.center[
[Notepad++](https://notepad-plus-plus.org/)
![](https://www.dropbox.com/s/rya3ntafbuablgw/notepad%2B%2B.jpg?raw=1)
]
]

.pull-right[
.center[
[Sublime](www.sublimetext.com)
![](https://www.dropbox.com/s/3dajm2u2i9hdyzd/sublime.png?raw=1)
]

]

.footnote[Big shout outs also to [Typora](https://www.typora.io/) and [Ulysses](https://ulysses.app/) (Mac only) that both write (and preview) Markdown in a simple interface, no distractions.]

---

# R Studio is My Text Editor of Choice

- Honestly, I write **everything** in R Studio's text editor
    - Syntax highlighting
    - Actually can *run* R code, autocomplete, etc<sup>.red[1]</sup>
    - Can render the markdown to an output format: html, pdf, etc.

.footnote[<sup>.red[1]</sup> You can *write* R code in other text editors, but you can't *execute* them outside of *R Studio* (or the command line, but that's too advanced.) Same with actually rendering your markdown to an output (pdf, html, etc)]

---

# Tips with Markdown

- Empty space is *very important* in markdown

- Lines that begin with a space may not render properly

- Math that contains spaces *between* the dollar-signs may not render properly

- Moving from one type of content to another (e.g. a heading to a list to text to an equation to text) requires *blank lines between them* to work

- Here is a [great general tutorial on markdown syntax](https://www.markdowntutorial.com/)

---

class: inverse, center, middle
# Compiling Your Documents

---

# knitr

- When you are ready, you "compile" your markdown and code into an output format using: 
    - [`knitr`](https://yihui.name/knitr/)<sup>.red[1]</sup>, an R package that "`knit`s" your R code and markdown `.Rmd` into a `.md` file for:  
  - [pandoc](http://pandoc.org) is a "swiss-army knife" utility that can convert between *dozens* of document types

- All you need to do is click the `Knit` button at the top of the text editor!

.footnote[<sup>.red[1]</sup> `knitr` also relies on the `rmarkdown` package, which will probably be installed when you first `knit.`]
---

# knitr: What's Happening On Your Computer

.center[
![](https://www.dropbox.com/s/kzvb8n8fpf6lstz/pandoc1.png?raw=1)
]

---

class: inverse, center, middle
# R Projects

---

# R Projects I

- A `R Project` is a way of systematically organizing your `R` history, working directory, and related files in a single package
- Can easily be sent to others who can reproduce your work easily 
- Connects well with version control software like GitHub
- Can open multiple projects in multiple windows

---

# R Projects II

- Projects solve all of the following problems:
    1. Organizing your files (data, plots, text, citations, etc)
    2. Having an accessible working directory (for loading and saving data, plots, etc)
    3. Saving and reloading your commands history and preferences
    4. Sending files to collaborators, so they have the same working directory as you

---

# Creating a Project I

.center[
![](https://www.dropbox.com/s/h3oucgn03cl0o52/project1.png?raw=1)

]

---

# Creating a Project II

.pull-left[
.center[
![](https://www.dropbox.com/s/mfstq4g7enq6vxo/project2.png?raw=1)
]
]

.pull-right[

- In almost all cases, you simply want a `New Project`

- For more advanced uses, your project can be an `R Package` or a `Shiny Web Application`

- If you have other packages that create templates installed (as I do, in the previous image), they will also show up as options

]

---

# Creating a Project III

.pull-left[
.center[
![](https://www.dropbox.com/s/qx93q89x0yoojpk/project3.png?raw=1)
]
]

.pull-right[

- Enter a name for the project in the top field 
    - Also creates a folder on your computer with the name you enter into the field

- Choose the location of the folder on your computer

- Depending on if you have other packages or utilities installed (such as `git`, see below!), there may be additional options, do not check them unless you know what you are doing

- Bottom left checkbox allows you to open a new instance (window) of `R` just for this project (and keep existing windows open)
]

---

# Projects

.pull-left[
.center[
![:scale 50%](https://www.dropbox.com/s/266cs7tuuxgurfv/projectswitcher.png?raw=1)
]
Switch between each project (Window) on your computer (this is on a Mac).

]

.pull-right[
.center[
![:scale 40%](https://www.dropbox.com/s/evb7npimklwuyyk/projectlist.png?raw=1)
]
- At top right corner of RStudio
    - Click the button to the right of the name to open in a new window!

]

---

# Loading Others' Projects

.center[
![:scale 80%](https://www.dropbox.com/s/i1ijatt56ncxi0e/github%202.png?raw=1)

- This project is on [GitHub](http://github.com/ryansafner/workflow)
- Click the green button, download to your computer, open `.Rproj` file in R Studio
]

---

# A Good File Structure

.pull-left[
.center[
![](https://www.dropbox.com/s/kaby3i11svyj81b/projectfiles.png?raw=1)
]
]

.pull-right[
- Look through this on your own
- Read the `README` of this repository on GitHub for instructions (automatically shows on the main page)
- Look at the `Example_paper.Rmd`
    - Uses data from **Data** folder
    - Uses `.R` scripts from **Scripts** folder
    - Uses figures from **Figures** folder 
    - Uses `bibexample.bib` from **Bibliography** folder
- This file is `workflow.presentation.Rmd` in **Presentation** folder
]

---

class: inverse, center, middle

# Version Control

---

# Have You Done This?

.center[
![](https://www.dropbox.com/s/c4a2e8p176yb79c/phdfinal1.png?raw=1)

.source[[PhD Comics](http://phdcomics.com/comics/archive_print.php?comicid=1531)]
]

---

# Have You Done This?

.center[
![](https://www.dropbox.com/s/b5xu55iym9jjphd/phdfinal2.png?raw=1)

.source[[PhD Comics](http://phdcomics.com/comics/archive_print.php?comicid=1531)]
]
---

# Have You Done This?

.center[
![](https://www.dropbox.com/s/434uy4d8wiuop3a/phdfinal3.png?raw=1)

.source[[PhD Comics](http://phdcomics.com/comics/archive_print.php?comicid=1531)]
]

---

# Do You Want to Be Able To

.pull-left[

- Keep your files backed up

- Track changes

- Collaborate on the same files with others

- Edit files on one computer and then open and continue working on another?
]

.pull-right[
.center[
![](https://www.dropbox.com/s/aulqgvdy9u71hai/versioncontrol.png?raw=1)
]
]
---

# The Training-Wheels Version

.pull-left[
.center[
![](https://www.dropbox.com/s/lj4g0lvwkopd9vf/dropbox.png?raw=1)

[Dropbox.com](http://dropbox.com)
]
]
.pull-right[
- Register an account for free

- Set up a location on your computer for the `Dropbox/` folder

- Anything you put in this folder will sync to the cloud 
    - As soon as you change files, they *automatically* update and sync!
    - Can download any of these flies from the *website* on any device
    - Set this up on multiple computers so when you change a file on one, it updates on all the others!

]
---

# The Training-Wheels Version

.center[
![:scale 80%](https://www.dropbox.com/s/inlmf8znau7kbzz/mydropbox.png?raw=1)

My Dropbox - my life goes here
]
]

---

# The Expert Version

.pull-left[
.center[
![](https://www.dropbox.com/s/hm2bquswiqmk4ac/git.png?raw=1)

![](https://www.dropbox.com/s/5mmsx958d0l0a2j/github.png?raw=1)

]
]
.pull-right[
- `Git` is an "open source distributed version control system" *widely* used in the software development industry

- Track changes **on steroids**
    - Organize folders/files to track (a `"repository"`)
    - Take a snapshot of all of your files (a "`commit`") with "`comment`s"
    - `push` these to the cloud
    - `pull` changes to (other) computers as needed

- `GitHub` is a popular (not the only!) cloud destination for these repositories

]

---

# The Expert Version

.pull-left[
.center[
![](https://www.dropbox.com/s/hm2bquswiqmk4ac/git.png?raw=1)

![](https://www.dropbox.com/s/5mmsx958d0l0a2j/github.png?raw=1)

]
]
.pull-right[
- Shows history (`versions`) of files with comments
    - Can `fork` or `branch` repository into multiple versions at once
    - Good for "testing" things out without destroying old versions!
    - `revert` back to original versions as needed

]

---

# The Expert Version

.pull-left[
.center[
![:scale 75%](https://www.dropbox.com/s/0493d7zw6ixzq9i/gitbranches.png?raw=1)
]
]
.pull-right[
- Shows history (`versions`) of files with comments
    - Can `fork` or `branch` repository into multiple versions at once
    - Good for "testing" things out without destroying old versions!
    - `revert` back to original versions as needed

]

---

# The Expert Version

.center[
![:scale 70%](https://www.dropbox.com/s/egfojonn2s54dk2/basic-remote-workflow.png?raw=1)
]

---

# The Expert Version

- Requires *some* advanced set up, see [this excellent guide](http://happygitwithr.com/)

- R Studio integrates git and github commands nicely

---

# This Class on GitHub

.pull-left[
.center[
![:scale 100%](https://www.dropbox.com/s/arjk6zvbxaxqp8b/gitmetrics1.png?raw=1)
]
]

.pull-right[
.center[
![:scale 100%](https://www.dropbox.com/s/e9cbyzls160yvzt/gitmetrics2.png?raw=1)
]
]

.center[
[github.com/ryansafner/metricsF19](http://github.com/ryansafner/metricsF19)
]
---

# Most Packages Start on GitHub

.pull-left[
.center[
![:scale 100%](https://www.dropbox.com/s/urrisfpf9u39s2v/githubtidyverse.png?raw=1)

[github.com/tidyverse/tidyverse](https://github.com/tidyverse/tidyverse)
]
]

.pull-right[
.center[
![:scale 100%](https://www.dropbox.com/s/3a8hmg9hgcl9ojx/githubgapminder.png?raw=1)

[github.com/jennybc/gapminder](https://github.com/jennybc/gapminder)

]
]

---

# Resources

1. R Studio's [R Markdown Cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf) for a quick overview of R markdown 
2. R Studio's [Overview of R Markdown](https://rmarkdown.rstudio.com/) for some tutorials 
3. R Studio's [R Markdown Reference Guide](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf) for more specific options and issues
4. Kieran Healey's [The Plain Person's Guide to Plain Text Social Science](http://plain-text.co) on managing workflow with plain text files, R, and Git
5. Yihui Xie's (and coauthors) [R Markdown: the Definitive Guide](https://bookdown.org/yihui/rmarkdown/) on R Markdown syntax and customization options
6. Hadley Wickham's (and Garrett Grolemund) [R for Data Science](http://r4ds.had.co.nz/) on how to use R and R Markdown for data science work
7. Jenny Bryan's [Happy Git with R](http://happygitwithr.com/) on how to use git and GitHub with R as a version control system

### 2019-09-12 content/r/06-r-practice.Rmd
Add r practice 6 links

---
title: "1.5 Data Wrangling with tidyverse - R Practice"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

- [<i class="fab fa-markdown"></i> Download R Practice 4 answers as  `.Rmd`](/r/04-r-practice-answers.Rmd)
- [<i class="fab fa-markdown"></i> See my `.Rmd` for today's slides on github](https://github.com/ryansafner/metricsF19/blob/master/static/slides/06-slides.Rmd)

Download these and look through the markdown to see how they make the webpages or slides. Note the slides are a bit advanced - I use a lot of formatting!

### 2019-09-12 static/slides/06-slides.Rmd
Fix 06 slides

@@
 
-.pull-left[
-
 ### [The Office Model](#4)
@@
 ### [Resources](#69)
-
-]
-
-.pull-right[
-
-### [dplyr::select(): select variables](#60)
-
-### [dplyr::rename(): rename variables](#73)
-
-### [dplyr::mutate(): create new variables](#77)
-
-### [dplyr::summarize(): create statistics](#98)
-
-### [tidyr: reshaping data](#133)
-
-### [dplyr: combining datasets](#134)
-
-]
 
@@
 
-![From [R Studio's R Markdown Cheatsheet](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)](images/rmd2.png)
-
+From [R Studio's R Markdown Cheatsheet](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)(images/rmd2.png)
+]
 ]
@@
 .center[
-![](https://www.dropbox.com/s/czklhgt5llrekjc/newrmarkdown.png?raw=1)
+![](https://www.dropbox.com/s/ebf8lod67dqvaoe/newrmarkdown.png?raw=1)
 ]
@@
 
-# Structure of an `R Markdown` Document
+# Structure of an R Markdown Document
+
+.pull-left[
 
@@
 3. `R` chunks for data analysis, plots, figures, tables, statistics, as necessary
-
+]
+
+.pull-right[
+
+.center[
+![](https://www.dropbox.com/s/og99cl4vw761m1n/rmarkdownex1.png?raw=1)
+]
+]
 .footnote[<sup>.red[1]</sup> The one exception is for managing bibliographies, this requires an additional `.bib` file!]
@@
 
-- You can add [additional options](this guide](https://yihui.name/knitr/options/)) inside the `{braces}` after `r`, some common options:
+- You can add [additional options](https://yihui.name/knitr/options/) inside the `{braces}` after `r`, some common options:
 
@@
 # Writing Text with Markdown
+
+.pull-left[
 
@@
     - `**bold text**` creates **bold text** 
-
+]
+
+.pull-right[
+.center[
+![](https://www.dropbox.com/s/c2jq6hpqybksp6q/rmarkdownex2.png?raw=1)
+]
+]
 ---
@@
 - A great resource: [Wikibooks LaTeX Mathematics chapter](https://en.wikibooks.org/wiki/LaTeX/Mathematics)
-- And for more about $\LaTeX{}$ in general: [Wikibooks LaTeX](https://en.wikibooks.org/wiki/LaTeX/)
 
@@
 
-- First term after the `{` is your **citation key** (e.g. `safner2016`)
+- First input after the `@article` is your **citation key** (e.g. `safner2016`)
+    - Whenever you want to cite this article, you'll invoke this key
 ]
@@
     - You can customize citations, e.g.:
-        - **All in parentheses**: `[@safner2016]` produces [@safner2016]
-        - **No parentheses**: `@safner2016` produces @safner2016
-        - **Date only**: `[-@safner2016]` produces [-@safner2016]
-        - **Adding pages or comments**: `@safner2016[p.743-744 and passim]` produces @safner2016[p.743-744 and passim]
+        - **All in parentheses**: `[@safner2016]` produces: (Safner, 2016)
+        - **No parentheses**: `@safner2016` produces: Safner 2016
+        - **Date only**: `[-@safner2016]` produces (2016)
+        - **Adding pages or comments**: `@safner2016[p.743-744 and passim]` produces (Safner, 2016, p.743-744 and passim)
+
 - BibTeX will automatically collect all works cited at the end and produce a **bibliography** according to a style you can choose
+
+- We'll see more when we discuss writing your paper
 
@@
     - something as basic (and boring!) as "**Notepad**," for example
-    - many good [text editors](https://en.wikipedia.org/wiki/Text_editor) out there
+    - many good [text editors](https://en.wikipedia.org/wiki/Text_editor) out there, I like [Typora](https://www.typora.io/) or [Ulysses](https://ulysses.app/) (Mac only) for writing (and previewing) Markdown in a simple interface, with no distractions
 
@@
 
-.footnote[Big shout outs also to [Typora](https://www.typora.io/) and [Ulysses](https://ulysses.app/) (Mac only) that both write (and preview) Markdown in a simple interface, no distractions.]
-
 ---
@@
     - Syntax highlighting
-    - Actually can *run* R code, autocomplete, etc<sup>.red[1]</sup>
+    - Actually can *run* R code, autocomplete, etc
     - Can render the markdown to an output format: html, pdf, etc.
 
-.footnote[<sup>.red[1]</sup> You can *write* R code in other text editors, but you can't *execute* them outside of *R Studio* (or the command line, but that's too advanced.) Same with actually rendering your markdown to an output (pdf, html, etc)]
+- You can *write* R code in other text editors, but you can't *execute* them outside of *R Studio* (or the command line, but that's too advanced.) Same with actually rendering your markdown to an output (pdf, html, etc)
 
@@
 
+.pull-left[
 - When you are ready, you "compile" your markdown and code into an output format using:
+
 - [`knitr`](https://yihui.name/knitr/)<sup>.red[1]</sup>, an R package that "`knit`s" your R code and markdown `.Rmd` into a `.md` file for:
+
 - [pandoc](http://pandoc.org) is a "swiss-army knife" utility that can convert between *dozens* of document types
@@
 - All you need to do is click the `Knit` button at the top of the text editor!
-
+]
+
+.pull-right[
+.center[
+![](https://www.dropbox.com/s/kzvb8n8fpf6lstz/pandoc1.png?raw=1)
+
+]
+]
 .footnote[<sup>.red[1]</sup> `knitr` also relies on the `rmarkdown` package, which will probably be installed when you first `knit.`]
----
-
-# knitr: What's Happening On Your Computer
-
-.center[
-![](https://www.dropbox.com/s/kzvb8n8fpf6lstz/pandoc1.png?raw=1)
-]
-
 ---
@@
 .center[
-![:scale 80%](https://www.dropbox.com/s/i1ijatt56ncxi0e/github%202.png?raw=1)
-
+![:scale 70%](https://www.dropbox.com/s/i1ijatt56ncxi0e/github%202.png?raw=1)
+]
 - This project is on [GitHub](http://github.com/ryansafner/workflow)
 - Click the green button, download to your computer, open `.Rproj` file in R Studio
-]
 
@@
 
-- `GitHub` is a popular (not the only!) cloud destination for these repositories
+- [`GitHub`](http://github.com) is a popular (not the only!) cloud destination for these repositories
 
@@
 .center[
-![](https://www.dropbox.com/s/hm2bquswiqmk4ac/git.png?raw=1)
-
-![](https://www.dropbox.com/s/5mmsx958d0l0a2j/github.png?raw=1)
-
+![:scale 75%](https://www.dropbox.com/s/0493d7zw6ixzq9i/gitbranches.png?raw=1)
 ]
@@
 
-.pull-left[
-.center[
-![:scale 75%](https://www.dropbox.com/s/0493d7zw6ixzq9i/gitbranches.png?raw=1)
-]
-]
-.pull-right[
-- Shows history (`versions`) of files with comments
-    - Can `fork` or `branch` repository into multiple versions at once
-    - Good for "testing" things out without destroying old versions!
-    - `revert` back to original versions as needed
-
-]
-
----
-
-# The Expert Version
-
-.center[
-![:scale 70%](https://www.dropbox.com/s/egfojonn2s54dk2/basic-remote-workflow.png?raw=1)
+.center[
+![:scale 60%](https://www.dropbox.com/s/egfojonn2s54dk2/basic-remote-workflow.png?raw=1)
 ]

### 2019-09-13 content/class/06-class.Rmd
Updated slides and practice page for lesson 1.6

@@
 
-- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/05-slides.html)
+- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/06-slides.html)
 
@@
 
-Today you will be working on [R practice problems](/r/05-r-practice.html). Check back later for solutions.
-
-# Statistics and Software Survey
-
-I have posted ["answers"](/homeworks/00-preliminary-survey-answers.html) to the statistics survey.
+Today's "[practice problems](/r/06-r-practice.html)" are a series of raw `.Rmd` files that I've used for various things, so you can see how they work.
 
@@
 
-Check back here!
+See the last 2 slides and the readings page for my workflow, my suggestions to you, and a series of links to more resources and readings.

### 2019-09-13 content/r/06-r-practice.Rmd
Updated slides and practice page for lesson 1.6

@@
 ---
-title: "1.5 Data Wrangling with tidyverse - R Practice"
+title: "1.6 Optimize Workflow: Projects, Markdown, and Git - R Practice"
 editor_options: 
@@
 
-- [<i class="fab fa-markdown"></i> Download R Practice 4 answers as  `.Rmd`](/r/04-r-practice-answers.Rmd)
+There's currently no actual *practice* problems, but you should start taking a look at some actual `.Rmd` files to see how they work, if you are interested:
+
+- [<i class="fas fa-external-link-square-alt"></i> Workflow R Project on GitHub](http://github.com/ryansafner/workflow)
+
+Look at my `Workflow` R project on GitHub (first link).
+Also, see the 2nd-to-last slide in the slides (updated)!
+Follow the instructions to download and open the `.RProj` file on your computer (in RStudio).
+Look around the files (in the bottom-right panel), and open `Example_paper.Rmd` to see how you might write a paper in `R Markdown`.
+
+If you are curious, below are two other examples.
+The first is the raw `.Rmd` file I wrote that created the (`html`) answer key for R practice 4 (on `ggplot2`).
+The second is the code for today's slides, written in [`Xaringan`](http://github.com/yihui/xaringan) (a special, `html`-based slides package), hosted on github.
+Note the slides are quite advanced - I use a lot of html and css formatting to make them pretty!
+
+- [<i class="fab fa-markdown"></i> `04-r-practice-answers.Rmd`](/r/04-r-practice-answers.Rmd)
 - [<i class="fab fa-markdown"></i> See my `.Rmd` for today's slides on github](https://github.com/ryansafner/metricsF19/blob/master/static/slides/06-slides.Rmd)
-
-Download these and look through the markdown to see how they make the webpages or slides. Note the slides are a bit advanced - I use a lot of formatting!

### 2019-09-13 static/slides/06-slides.Rmd
Updated slides and practice page for lesson 1.6

@@
 
+# My Workflow (that I suggest to you)
+
+1. Create a new repository on Github.
+2. Start a New R Project in R Studio (link it to the github repository - [see guide](http://happygitwithr.com))
+3. Create a logical file system ([see example](https://github.com/ryansafner/workflow)), such as:
+```
+project # folder on my computer (the new working directory)
+|
+|- Data/ # folder for data files 
+|- Scripts/ # folder .R code
+|- Bibliography/ # folder for .bib files
+|- Figures/ # folder to plots and figures to
+|- paper.Rmd # write document here
+```
+4. Write document in `paper.Rmd`, loading/saving files from/to various folders in project
+5. Knit document to `pdf` or `html`.
+6. Occasionally, `stage` and `commit` changes with a description, `push` to GitHub.
+
+---
+
 # Resources

### 2019-09-13 content/assignment/01-problem-set.Rmd
Update slides 6, class 6, reading 7, add problem set 1 files, r practice 5 answers

---
title: "Problem Set 1"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

*Due by Thursday, September 19, 2019*

- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/01-problem-set.pdf)
- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/01-r-problem-set.Rmd)

# Instructions

For this problem set, you may type up your answers to each question (save plots as images and include them with your answers) and also turn in a (commented!) *.R* file of commands for each relevant question. You may email or turn in a paper copy. It is not relevant for this particular problem set, but in the future, I will allow handwritten answers for math/theory questions.

Alternatively, you may download the `.Rmd` file, do the homework in markdown, and email to me a single `knit`ted `html` or `pdf` file (and be sure that it shows all of your code).

To minimize confusion, I suggest creating a new `R Project` (e.g. `hw1`) and storing any data and plots in that folder on your computer. See my [example workflow](https://github.com/ryansafner/workflow).

You may work together (and I highly encourage that) but you must turn in your own answers. I grade homeworks 70% for completion, and for the remaining 30%, pick one question to grade for accuracy - so it is best that you try every problem, even if you are unsure how to complete it accurately.

# The Popularity of Baby Names

Install and load the package `babynames`. Get help for `?babynames` to see what the data includes.

## 1. 
### a. 
What are the top 5 boys names for 2017, and what *percent* of overall names is each?

### b.
What are the top 5 *girls* names, and what *percent* of overall names is each?

## 2. 
Make two barplots, of these top 5 names, one for each sex. Map `aes`thetics `x` to `name` and `y` to `prop`^[Or `percent`, if you made that variable, as I did.] and use `geom_col` (since you are declaring a specific `y`, otherwise you could just use `geom_bar()` and just an `x`.)

## 3. 
Find your name.^[If your name isn't in there :(, pick a random name.] `count` by `sex` how many babies since 1880 were named your name.^[Hint: if you do this, you'll get the number of *rows* (years) there are in the data. You want to add the number of babies in each row (`n`), so inside `count`, add `wt=n` to weight the count by `n`.] Also add a variable for the percent of each sex.

## 4.
Make a line graph of the number of babies with your name over time, `color`ed by `sex`.

## 5.
### a.
Make a table of the most common name for boys by year between 1980-2017.^[Hint: once you've got all the right conditions, you'll get a table with a lot of data. You only want to `slice` the `1`st row for each table.]

### b.
Now do the same for girls.

## 6. Now let's graph the evolution of the most common names since 1880.
### a.
First, find out what are the top 10 *overall* most popular names for boys and for girls. You may want to create two vectors, each with these top 5 names.

### b.
Now make two `line`graphs of these 5 names over time, one for boys, and one for girls.

## 7.
**Bonus (hard!)**: What are the 10 most common "gender-neutral" names?^[This is hard to define. For our purposes, let's define this as names where between 48 and 52% of the babies with the name are Male.]

# Political and Economic Freedom Around the World

For the remaining questions, we'll look at the relationship between Economic Freedom and Political Freedom in countries around the world today. Our data for economic freedom comes from the [Fraser Institute](https://www.fraserinstitute.org/economic-freedom/dataset?geozone=world&year=2016&page=dataset), and our data for political freedom comes from [Freedom House](https://freedomhouse.org/content/freedom-world-data-and-resources).

## 8. 
Download these two datasets that I've cleaned up a bit:^[If you want, try downloading them from the websites yourself!]

- [<i class="fas fa-table"></i> `econfreedom.csv`](http://metricsf19.classes.ryansafner.com/data/econfreedom.csv)
- [<i class="fas fa-table"></i> `freedomhouse2018.csv`](http://metricsf19.classes.ryansafner.com/data/freedomhouse2018.csv)

Load them with `df<-read_csv("name_of_the_file.csv")` and save one as `econfreedom` and the other as `polfreedom`. Look at each `tibble` you've created.

## 9. 
The `polfreedom` dataset is still a bit messy. Let's overwrite it (or assign to something like `polfreedom2`) and select `Country.Territory` and `Total` (total freedom score) and rename `Country.Territory` to `Country`.

## 10.
Now we can try to merge these two datasets into one. Since they both have `Country` as a variable, we can merge these tibbles using `left_join(econfreedom, polfreedom, by="Country")`^[Note, if you saved as something else in question 9., use that instead of `polfreedom`!] and save this as a new tibble (something like `freedom`).

## 11.
Now make a scatterplot of Political Freedom (`total`)^[Feel free to `rename` these!] as `y` on Economic Freedom (`ef`) as `x` and `color` by `continent`.

## 12.
Let's do this again, but highlight some key countries. Pick three countries, and make a new tibble from `freedom` that is only the observations of those countries. Additionally, *install* and *load* a packaged called `ggrepel`^[This automatically adjusts labels so they don't cover points on a plot!] Next, redo your plot from question 11, but now add a layer: `geom_label_repel` and set its `data` to your three-country tibble, use same `aes`thetics as your overall plot, but be sure to add `label = ISO`, to use the ISO country code to label.^[You might also want to set a low `alpha` level to make sure the labels don't obscure other points!]

## 13.
Make another plot similar to 12, except this time use GDP per Capita (`gdp`) as `y`. Feel free to try to put a regression line with `geom_smooth()`!^[If you do, be sure to set its data to the full `freedom`, not just your three countries!]^[Those of you in my Development course, you just made my graphs from [Lesson 2](https://devf19.classes.ryansafner.com/slides/02-slides#23)!]

### 2019-09-13 content/r/05-r-practice.Rmd
Test fixing footnote in r practice 5
Update slides 6, class 6, reading 7, add problem set 1 files, r practice 5 answers

@@
 ```
+
+- [<i class="fas fa-external-link-square-alt"></i> `Answer Key (html)`](/r/05-r-practice-answers.html)
+- [<i class="fab fa-markdown"></i> `Answer Key (markdown)`](/r/05-r-practice-answers.Rmd)
+- [<i class="fas fa-registered"></i> `Answer Key (R Script)`](/r/05-r-practice-answers.R)
 
@@
 
-```{r, eval=F}
-ifelse(my_conditions,
+`ifelse(my_conditions,
        yes = do_this_if_TRUE,
-       no = do_this_if_FALSE)
-```
-]
+       no = do_this_if_FALSE)`]
 

### 2019-09-13 content/reading/07-reading.Rmd
Update slides 6, class 6, reading 7, add problem set 1 files, r practice 5 answers

--- 2019-09-11 content/reading/06-reading.Rmd
@@
 
-- <i class="fas fa-external-link-square-alt"></i>[Chs. 8, 27, 28, 29, 30 (but especially 27)](http://r4ds.had.co.nz) in @R4DS
+- <i class="fas fa-book"></i> @Bailey[Math and Probability Background Appendices A-I]
 
-# Other Resources
-
-## On R Markdown
-- Xie, Yihui, J.J. Allaire, and Garett Grolemund, [*R Markdown: The Definitive Guide*](https://bookdown.org/yihui/rmarkdown/)
-- RStudio, [*R Markdown Reference Guide*](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)
-- RStudio, [*Overview of R Markdown*](https://rmarkdown.rstudio.com/)
-- Rstudio, [*R Markdown Cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
-
-## On Optimized Workflow
-
-- Healey, Kieran, [*The Plain Person’s Guide to Plain Text Social Science*](http://plain-text.co/)
-- Bryan, Jenny and Jim Hester, [*Happy Git and GitHub for the useR*](https://happygitwithr.com/)
+Now that we return to the statistics, we will do a minimal overview of basic statistics and distributions. Review Bailey's appendices.

### 2019-09-13 static/homeworks/01-problem-set-pdf.Rmd
Update slides 6, class 6, reading 7, add problem set 1 files, r practice 5 answers

--- 2019-09-13 content/assignment/01-problem-set.Rmd
@@
 title: "Problem Set 1"
-editor_options: 
-  chunk_output_type: console
+author: "Ryan Safner"
+date: "ECON 480 - Fall 2019"
+output: pdf_document
 ---
-
-```{r message=FALSE, warning=FALSE, include=FALSE}
-library(tidyverse)
-library(pander)
-source(file.path(here::here(), "lib", "graphics.R"))
-```
-
-*Due by Thursday, September 19, 2019*
-
-- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/01-problem-set.pdf)
-- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/01-r-problem-set.Rmd)
 
@@
 
-## 1. 
-### a. 
-What are the top 5 boys names for 2017, and what *percent* of overall names is each?
+1. 
+a. What are the top 5 boys names for 2017, and what *percent* of overall names is each?
 
-### b.
-What are the top 5 *girls* names, and what *percent* of overall names is each?
+\vspace{3in}
 
-## 2. 
+b. What are the top 5 *girls* names, and what *percent* of overall names is each?
+
+\vspace{3in}
+
+2.
 Make two barplots, of these top 5 names, one for each sex.
@@
 
-## 3. 
+\clearpage
+
+3.
 Find your name.^[If your name isn't in there :(, pick a random name.] `count` by `sex` how many babies since 1880 were named your name.^[Hint: if you do this, you'll get the number of *rows* (years) there are in the data.
@@
 
-## 4.
-Make a line graph of the number of babies with your name over time, `color`ed by `sex`.
+\vspace{3in}
 
-## 5.
-### a.
+4. Make a line graph of the number of babies with your name over time, `color`ed by `sex`.
+
+\clearpage
+5.
+a.
 Make a table of the most common name for boys by year between 1980-2017.^[Hint: once you've got all the right conditions, you'll get a table with a lot of data.
@@
 
-### b.
-Now do the same for girls.
+\vspace{3in}
 
-## 6. Now let's graph the evolution of the most common names since 1880.
-### a.
-First, find out what are the top 10 *overall* most popular names for boys and for girls. You may want to create two vectors, each with these top 5 names.
+b. Now do the same for girls.
 
-### b.
-Now make two `line`graphs of these 5 names over time, one for boys, and one for girls.
+\clearpage
 
-## 7.
+6. Now let's graph the evolution of the most common names since 1880.
+a. First, find out what are the top 10 *overall* most popular names for boys and for girls. You may want to create two vectors, each with these top 5 names.
+
+b. Now make two `line`graphs of these 5 names over time, one for boys, and one for girls.
+
+\vspace{3in}
+
+7.
 **Bonus (hard!)**: What are the 10 most common "gender-neutral" names?^[This is hard to define.
 For our purposes, let's define this as names where between 48 and 52% of the babies with the name are Male.]
+
+\clearpage
 
@@
 
-## 8. 
-Download these two datasets that I've cleaned up a bit:^[If you want, try downloading them from the websites yourself!]
+8. Download these two datasets that I've cleaned up a bit:^[If you want, try downloading them from the websites yourself!]
 
@@
 
-Load them with `df<-read_csv("name_of_the_file.csv")` and save one as `econfreedom` and the other as `polfreedom`. Look at each `tibble` you've created.
+9. Load them with `df<-read_csv("name_of_the_file.csv")` and save one as `econfreedom` and the other as `polfreedom`. Look at each `tibble` you've created.
 
-## 9. 
+10.
 The `polfreedom` dataset is still a bit messy.
@@
 
-## 10.
+11.
 Now we can try to merge these two datasets into one.
@@
 
-## 11.
-Now make a scatterplot of Political Freedom (`total`)^[Feel free to `rename` these!] as `y` on Economic Freedom (`ef`) as `x` and `color` by `continent`.
+\vspace{3in}
 
-## 12.
+12. Now make a scatterplot of Political Freedom (`total`)^[Feel free to `rename` these!] as `y` on Economic Freedom (`ef`) as `x` and `color` by `continent`.
+
+\clearpage
+
+13.
 Let's do this again, but highlight some key countries.
@@
 
-## 13.
-Make another plot similar to 12, except this time use GDP per Capita (`gdp`) as `y`.
-Feel free to try to put a regression line with `geom_smooth()`!^[If you do, be sure to set its data to the full `freedom`, not just your three countries!]^[Those of you in my Development course, you just made my graphs from [Lesson 2](https://devf19.classes.ryansafner.com/slides/02-slides#23)!]
+\vspace{2in}
+
+14.
+Make another plot similar to f, except this time use GDP per Capita (`gdp`) as `y`.
+Feel free to try to put a regression line with `geom_smooth()`!^[If you do, be sure to set its data to the full `freedom`, not just your three countries!] Those of you in my Development course, you just made my graphs from [Lesson 2](https://devf19.classes.ryansafner.com/slides/02-slides#23)!

### 2019-09-13 static/homeworks/01-problem-set.Rmd
Update slides 6, class 6, reading 7, add problem set 1 files, r practice 5 answers

---
title: "Problem Set 1"
author: "YOUR NAME HERE" # put your name here! 
date: "ECON 480 - Fall 2019"
output: html_document # change to pdf if you'd like, this will make a webpage (you can email it and open in any browser)
---

<!--CLICK "KNIT" ABOVE TO RENDER TO HTML, PDF, OR WORD OUTPUT
In fact, try knitting right away and see what this produces!
-->

*Due by Thursday, September 19, 2019*

# Instructions

For this problem set, you may type up your answers to each question (save plots as images and include them with your answers) and also turn in a (commented!) *.R* file of commands for each relevant question. You may email or turn in a paper copy. It is not relevant for this particular problem set, but in the future, I will allow handwritten answers for math/theory questions.

Alternatively, you may download the `.Rmd` file, do the homework in markdown, and email to me a single `knit`ted `html` or `pdf` file (and be sure that it shows all of your code).

To minimize confusion, I suggest creating a new `R Project` (e.g. `hw1`) and storing any data and plots in that folder on your computer. See my [example workflow](https://github.com/ryansafner/workflow).

You may work together (and I highly encourage that) but you must turn in your own answers. I grade homeworks 70% for completion, and for the remaining 30%, pick one question to grade for accuracy - so it is best that you try every problem, even if you are unsure how to complete it accurately.

# The Popularity of Baby Names

Install and load the package `babynames`. Get help for `?babynames` to see what the data includes.

```{r}
# write your code here! 
```

## 1. 
### a. 
**What are the top 5 boys names for 2017, and what *percent* of overall names is each?**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 1-a}
# write your code here! 
```

### b.
**What are the top 5 *girls* names, and what *percent* of overall names is each?**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 1-b}
# write your code here! 
```

## 2. 
**Make two barplots, of these top 5 names, one for each sex. Map `aes`thetics `x` to `name` and `y` to `prop`^[Or `percent`, if you made that variable, as I did.] and use `geom_col` (since you are declaring a specific `y`, otherwise you could just use `geom_bar()` and just an `x`.)**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 2}
# write your code here! 
```

## 3. 
**Find your name.^[If your name isn't in there :(, pick a random name.] `count` by `sex` how many babies since 1880 were named your name.^[Hint: if you do this, you'll get the number of *rows* (years) there are in the data. You want to add the number of babies in each row (`n`), so inside `count`, add `wt=n` to weight the count by `n`.] Also add a variable for the percent of each sex.**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 3}
# write your code here! 
```

## 4.
**Make a line graph of the number of babies with your name over time, `color`ed by `sex`.**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 4}
# write your code here! 
```

## 5.
### a.
**Make a table of the most common name for boys by year between 1980-2017.^[Hint: once you've got all the right conditions, you'll get a table with a lot of data. You only want to `slice` the `1`st row for each table.]**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 5-a}
# write your code here! 
```

### b.
**Now do the same for girls.**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 5-b}
# write your code here! 
```

## 6. Now let's graph the evolution of the most common names since 1880.
### a.
**First, find out what are the top 10 *overall* most popular names for boys and for girls. You may want to create two vectors, each with these top 5 names.**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 6-a}
# write your code here! 
```

### b.
**Now make two `line`graphs of these 5 names over time, one for boys, and one for girls.**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 6-b}
# write your code here! 
```

## 7.
**_Bonus (hard!)_**: What are the 10 most common "gender-neutral" names?^[This is hard to define. For our purposes, let's define this as names where between 48 and 52% of the babies with the name are Male.]**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 7}
# write your code here! 
```

---

# Political and Economic Freedom Around the World

**For the remaining questions, we'll look at the relationship between Economic Freedom and Political Freedom in countries around the world today. Our data for economic freedom comes from the [Fraser Institute](https://www.fraserinstitute.org/economic-freedom/dataset?geozone=world&year=2016&page=dataset), and our data for political freedom comes from [Freedom House](https://freedomhouse.org/content/freedom-world-data-and-resources).**

## 8. 
**Download these two datasets that I've cleaned up a bit:^[If you want, try downloading them from the websites yourself!]**

- [<i class="fas fa-table"></i> `econfreedom.csv`](http://metricsf19.classes.ryansafner.com/data/econfreedom.csv)
- [<i class="fas fa-table"></i> `freedomhouse2018.csv`](http://metricsf19.classes.ryansafner.com/data/freedomhouse2018.csv)

**Load them with `df<-read_csv("name_of_the_file.csv")` and save one as `econfreedom` and the other as `polfreedom`. Look at each `tibble` you've created.**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 8}
# write your code here! 
```

## 9. 
**The `polfreedom` dataset is still a bit messy. Let's overwrite it (or assign to something like `polfreedom2`) and select `Country.Territory` and `Total` (total freedom score) and rename `Country.Territory` to `Country`.**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 9}
# write your code here! 
```

## 10.
**Now we can try to merge these two datasets into one. Since they both have `Country` as a variable, we can merge these tibbles using `left_join(econfreedom, polfreedom, by="Country")`^[Note, if you saved as something else in question 8., use that instead of `polfreedom`!] and save this as a new tibble (something like `freedom`).**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 10}
# write your code here! 
```

## 11.
**Now make a scatterplot of Political Freedom (`total`)^[Feel free to `rename` these!] as `y` on Economic Freedom (`ef`) as `x` and `color` by `continent`.**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 11}
# write your code here! 
```

## 12.
**Let's do this again, but highlight some key countries. Pick three countries, and make a new tibble from `freedom` that is only the observations of those countries. Additionally, *install* and *load* a packaged called `ggrepel`^[This automatically adjusts labels so they don't cover points on a plot!] Next, redo your plot from question 11, but now add a layer: `geom_label_repel` and set its `data` to your three-country tibble, use same `aes`thetics as your overall plot, but be sure to add `label = ISO`, to use the ISO country code to label.^[You might also want to set a low `alpha` level to make sure the labels don't obscure other points!]**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 12}
# write your code here! 
```

## 13.
**Make another plot similar to 12, except this time use GDP per Capita (`gdp`) as `y`. Feel free to try to put a regression line with `geom_smooth()`!^[If you do, be sure to set its data to the full `freedom`, not just your three countries!] Those of you in my Development course, you just made my graphs from [Lesson 2](https://devf19.classes.ryansafner.com/slides/02-slides#23)!**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 13}
# write your code here! 
```

### 2019-09-13 static/slides/06-slides.Rmd
Update slides 6, class 6, reading 7, add problem set 1 files, r practice 5 answers

@@
 
-1. Create a new repository on Github.
-2. Start a New R Project in R Studio (link it to the github repository - [see guide](http://happygitwithr.com))
+1. Create a new repository on Github.<sup>.red[*]</sup>
+2. Start a New R Project in R Studio (link it to the github repository<sup>.red[*]</sup> - [see guide](http://happygitwithr.com))
 3. Create a logical file system ([see example](https://github.com/ryansafner/workflow)), such as:
@@
 4. Write document in `paper.Rmd`, loading/saving files from/to various folders in project
+    - e.g. load data like `df<-read_csv("Data/my_data")`; save plots like `ggsave("Figures/p.png")`
 5. Knit document to `pdf` or `html`.
-6. Occasionally, `stage` and `commit` changes with a description, `push` to GitHub.
-
+6. Occasionally, `stage` and `commit` changes with a description, `push` to GitHub.<sup>.red[*]</sup>
+
+.footnote[<sup>.red[*]</sup> Optional and a bit advanced, remember this is _my_ workflow.]
 ---

### 2019-09-13 content/assignment/01-problem-set.Rmd
Minor fixes to problem set 1 page (links and formatting)

@@
 
-- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/01-problem-set.pdf)
-- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/01-r-problem-set.Rmd)
+- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/01-problem-set-pdf.pdf)
+- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/01-problem-set.Rmd)
 
@@
 
-## 6. Now let's graph the evolution of the most common names since 1880.
+## 6. 
+Now let's graph the evolution of the most common names since 1880.
 ### a.
@@
 Make another plot similar to 12, except this time use GDP per Capita (`gdp`) as `y`.
-Feel free to try to put a regression line with `geom_smooth()`!^[If you do, be sure to set its data to the full `freedom`, not just your three countries!]^[Those of you in my Development course, you just made my graphs from [Lesson 2](https://devf19.classes.ryansafner.com/slides/02-slides#23)!]
+Feel free to try to put a regression line with `geom_smooth()`!^[If you do, be sure to set its data to the full `freedom`, not just your three countries!] Those of you in my Development course, you just made my graphs from [Lesson 2](https://devf19.classes.ryansafner.com/slides/02-slides#23)!

### 2019-09-17 content/class/07-class.Rmd
Add class 7 content and revise schedule (separate 2.1 into 2.1 and 2.2, add problem set 2 and 3)

---
title: "2.1: Random Variables and Distributions - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

*Tuesday, September 17, 2019*

# Overview

Today we pick up where we left off from causality. We begin with an review and overview of using data and descriptive statistics. We want to quantify characteristics about samples as *statistics*, which we will later use to *infer* things about populations (which we will later *identify* causal relationships).

Next class will be on random variables and distributions. This full week is your crash course/review of basic statistics that we will need to start the "meat and potatoes" of this class: linear regression next Tuesday. As such, I'll give you a brief homework to review these statistical concepts (with minimal use of `R`).

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/07-slides.html)

# Problem Set

[Problem Set 1](/assignment/01-problem-set.html) is due Thursday September 19.

Problem set 2 (on classes 2.1-2.2) will be posted shortly, and is due by Tuesday September 24.

# Math Appendix

## The Summation Operator

### Definition

Many elementary propositions in econometrics (and statistics) involve the use of the sums of numbers. Mathematicians often use the summation operator (the greek letter $\Sigma$ --"sigma") as a shorthand, rather than writing everything out the long way. It will be worth your time to understand the summation operator, and some of its properties, and how these can provide shortcuts to proving more advanced theorems in econometrics.

Let $X$ be a random variable from which a sample of $n$ observations is observed, so we have a sequence $\{x_1, x_2,...,x_n\}$ i.e. $x_i, $ for $i=1,2,...,n$. Then the total sum of the observations $(x_1+x_2+...+x_n)$ can be represented as:

$$\sum_{i=1}^n x_i = x_1+x_2+...+x_n$$

- The term beneath $\Sigma$ is known as the "index," which tells us where to begin our adding (at the 1<sup>st</sup> individual $x$ term, $x_1$)
    - Note other letters, such as $j$, or $k$ may be used (especially if $i$ is defined elsewhere)
- The term above $\Sigma$ is the total number of $x$ terms we should add $(n)$
- Essentially, read $\displaystyle \sum_{i=1}^{n} x_i$ as "add up all the individual $x$ observations from the 1<sup>st</sup> $(x_1)$ to the final $n$\textsuperscript{th} $(x_n)$."

### Useful Properties of Summation Operators

**Rule 1**: The summation of a constant $k$ times a random variable $X_i$ is equal to the constant times the summation of that random variable:

$$\sum_{i=1}^n kX_i = k \sum^n_{i=1} X_i$$

Proof:

$$\begin{align*}
\sum_{i=1}^n kX_i &= k x_1 + kx_2 +...+ kx_n\\
&= k(x_1+x_2+...x_n)\\
&= k\sum_{i=1}^n X_i. \\
\end{align*}$$

**Rule 2**: The summation of a sum of two random variables is equal to the sum of their summations:

$$\sum_{i=1}^n (X_i+Y_i) = \sum_{i=1}^n X_i + \sum_{i=1}^n Y_i$$

Proof:

$$\begin{align*}
\sum_{i=1}^n (X_i+Y_i) &=(X_1+Y_1) + (X_2+Y_2) + ... (X_n+Y_n)\\
&=(X_1+X_2+...+X_n) + (Y_1+Y_2+...+Y_n)\\
&=\sum_{i=1}^n X_i + \sum_{i=1}^n Y_i\\
\end{align*}$$

**Rule 3**: The summation of constant over $n$ observations is the product of the constant and $n$:

$$\sum_{i=1}^n k = nk$$

Proof:

$$\sum_{i=1}^n k = \underbrace{k + k + ... + k}_{n \text{ times}} = nk$$

**Combining these 3 rules:** for the sum of a linear combination of a random variable ($a+bX$):

$$\sum_{i=1}^n (a+bX_i) = na+b\sum_{i=1}^n X_i$$

Proof: left to you as an exercise!

### Advanced: Useful Properties for Regression

There are some additional properties of summations that may not be immediately obvious, but will be quite essential in proving properties of linear regressions.

Using the properties above, we can describe the **mean**, **variance**, and **covariance** of random variables.^[For more beyond the mere definition, see the appendix on **Covariance and Correlation**]

First, define the mean of  a sequence $\{X_i: i=1,...,n\}$ and $\{Y_i: i=1,...,n\}$ as:

$$\bar{X}=\frac{1}{n}\sum^n_{i=1}X_i$$

Second, the variance of  $X$ is:

$$var(X)=\frac{1}{n}\sum^n_{i=1}(X_i-\bar{X})^2$$

Third, the covariance of $X$ and $Y$ is:

$$cov(X,Y)=\frac{1}{n}\sum^n_{i=1}(X_i-\bar{X})(Y_i-\bar{Y})$$

**Rule 4**: The sum of the deviations of observations of $X_i$ from its mean ($\bar{X}$) is 0:

$$\sum^n_{i=1} (X_i-\bar{X})=0$$

Proof:

$$\begin{align*}
	\sum^n_{i=1} (X_i-\bar{x}) &=\sum^n_{i=1}X_i-\sum^n_{i=1}\bar{X} && \\
	&=\sum^n_{i=1}X_i-n\bar{X} && \text{Since $\bar{x}$ is a constant}\\
	&=n\underbrace{\frac{\displaystyle\sum^n_{i=1}X_i}{n}}_{\bar{X}}-n\bar{X} && \text{Multiply the first term by }\frac{n}{n}=1\\ 
	&=n\bar{X}-n\bar{X}&& \text{By the definition of the mean }\bar{X}\\
	&=0 &&  \\
\end{align*}$$

**Rule 5**: The squared deviations of $X$ are equal to the product of $X$ times its deviations:

$$\sum^n_{i=1} (X_i-\bar{X})^2=\sum^n_{i=1} X_i(X_i-\bar{X})$$

Proof:

$$\begin{align*}
	\sum^n_{i=1} (X_i-\bar{X})^2&=\sum^n_{i=1} (X_i-\bar{X})(X_i-\bar{X})	&& \text{Expanding the square}\\
	&=\sum^n_{i=1} X_i(X_i-\bar{X})-\sum^n_{i=1}\bar{X}(X_i-\bar{X}) && \text{Breaking apart the first term} \\
	&=\sum^n_{i=1} X_i(X_i-\bar{X})-\bar{X}\sum^n_{i=1}(X_i-\bar{X}) && \text{Since }\bar{X} \text{ is constant, not depending on } i's\\
	&=\sum^n_{i=1} X_i(X_i-\bar{X})-\bar{X}(0) && \text{From rule 4} \\
	&=\sum^n_{i=1} X_i(X_i-\bar{X}) && \text{Remainder after multiplying by 0}\\
\end{align*}$$

**Rule 6**: The following summations involving $X$ and $Y$ are equivalent:

$$\sum^n_{i=1} Y_i(X_i-\bar{X})=\sum^n_{i=1}X_i(Y_i-\bar{Y})=\sum^n_{i=1} (X_i-\bar{X})(Y_i-\bar{Y})$$

Proof:

$$\begin{align*}
	\sum^n_{i=1} (X_i-\bar{X})(Y_i-\bar{Y}) &=\sum^n_{i=1} Y_i(X_i-\bar{X})-\sum^n_{i=1}\bar{Y}(X_i-\bar{X}) && \text{Breaking apart the second term} \\
	&=\sum^n_{i=1} Y_i(X_i-\bar{X})-\bar{Y}(0) && \text{From rule 4}\\
	&=\sum^n_{i=1} Y_i(X_i-\bar{X}) && \text{Remainder after multiplying by 0}\\
\end{align*}$$

equivalently:

$$\begin{align*}
	\sum^n_{i=1} (X_i-\bar{X})(Y_i-\bar{Y}) &=\sum^n_{i=1} X_i(Y_i-\bar{Y})-\sum^n_{i=1}\bar{X}(Y_i-\bar{Y}) && \text{Breaking apart the first term} \\
	&=\sum^n_{i=1} X_i(Y_i-\bar{Y})-\bar{X}(0) && \text{From rule 4}\\
	&=\sum^n_{i=1} X_i(Y_i-\bar{Y}) && \text{Remainder after multiplying by 0}\\
\end{align*}$$

### 2019-09-17 content/reading/07-reading.Rmd
Add class 7 content and revise schedule (separate 2.1 into 2.1 and 2.2, add problem set 2 and 3)

@@
 ---
-title: "1.6 Optimize Your Workflow: Projects and Markdown - Readings"
+title: "2.1 Data 101 and Descriptive Statistics - Readings"
 citeproc: false
@@
 
-- <i class="fas fa-book"></i> @Bailey[Math and Probability Background Appendices A-I]
+- <i class="fas fa-book"></i> @Bailey[Math and Probability Background Appendix A]
 
-Now that we return to the statistics, we will do a minimal overview of basic statistics and distributions. Review Bailey's appendices.
+Now that we return to the statistics, we will do a minimal overview of basic statistics and distributions. Review Bailey's appendices. Today, only A is really useful, the rest will come next
+
+Chapter 2 is optional, but will give you a good overview of using data.

### 2019-09-17 static/homeworks/01-problem-set-answers.Rmd
Add class 7 content and revise schedule (separate 2.1 into 2.1 and 2.2, add problem set 2 and 3)

---
title: "Problem Set 1 (Answers)"
author: "Ryan Safner"
date: "ECON 480 - Fall 2019"
output:
  html_document:
    df_print: paged
    theme: simplex
    toc: true 
    toc_depth: 3
    toc_float: true
    code_folding: show
    highlight: tango
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

*Due by Thursday, September 19, 2019*

- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/01-problem-set.pdf)
- [<i class="fab fa-markdown"></i> `04-r-practice-answers.Rmd`](/homeworks/01-r-problem-set.Rmd)

# Instructions

For this problem set, you may type up your answers to each question (save plots as images and include them with your answers) and also turn in a (commented!) *.R* file of commands for each relevant question. You may email or turn in a paper copy. It is not relevant for this particular problem set, but in the future, I will allow handwritten answers for math/theory questions.

Alternatively, you may download the `.Rmd` file, do the homework in markdown, and email to me a single `knit`ted `html` or `pdf` file (and be sure that it shows all of your code).

To minimize confusion, I suggest creating a new `R Project` (e.g. `hw1`) and storing any data and plots in that folder on your computer. See my [example workflow](https://github.com/ryansafner/workflow).

You may work together (and I highly encourage that) but you must turn in your own answers. I grade homeworks 70% for completion, and for the remaining 30%, pick one question to grade for accuracy - so it is best that you try every problem, even if you are unsure how to complete it accurately.

# The Popularity of Baby Names

**Install and load the package `babynames`. Get help for `?babynames` to see what the data includes.**

```{r}
library(tidyverse)

# install for first use
# install.packages("babynames")

# load package 
library(babynames)

# explore help
# ?babynames
```

## 1. 
### a. 
**What are the top 5 boys names for 2017, and what *percent* of overall names is each?**

```{r}
# save as a new tibble
top_5_boys_2017 <- babynames %>% # take data
  filter(sex=="M", # filter by males
         year==2017) %>% # and for 2007
  arrange(desc(n)) %>% # arrange in largest-to-smallest order of n (number)
  slice(1:5) %>% # optional, look only at first 5 rows
  mutate(percent = round(prop*100, 2)) # also optional, make a percent variable rounded to 2 decimals

# look at our new tibble
top_5_boys_2017
```

The top 5 names are

1. Liam (0.95%)
2. Noah (0.93%)
3. William (0.76%)
4. James (0.72%)
5. Logan (0.71%)

### b.
**What are the top 5 *girls* names, and what *percent* of overall names is each?**

```{r}
# save as a new tibble
top_5_girls_2017 <- babynames %>% # take data
  filter(sex=="F", # filter by females
         year==2017) %>% # and for 2007
  arrange(desc(n)) %>% # arrange in largest-to-smallest order of n (number)
  slice(1:5) %>% # optional, look only at first 5 rows
  mutate(percent = round(prop*100, 2)) # also optional, make a percent variable rounded to 2 decimals

# look at our new tibble
top_5_girls_2017
```

The top 5 names are

1. Emma (1.05%)
2. Olivia (0.99%)
3. Ava (0.85%)
4. Isabella (0.81%)
5. Sophia (0.79%)

## 2. 
**Make two barplots, of these top 5 names, one for each sex. Map `aes`thetics `x` to `name` and `y` to `prop`^[Or `percent`, if you made that variable, as I did.] and use `geom_col` (since you are declaring a specific `y`, otherwise you could just use `geom_bar()` and just an `x`.)**

```{r}
ggplot(data = top_5_boys_2017)+
  aes(x = reorder(name, n), #note this reorders the x variable from small to large n
      y = percent, # you can use prop if you didn't make a percent variable
      fill = name)+ # optional color!
  geom_col()+

  # now I'm just making it pretty
  scale_y_continuous(labels=function(x)paste(x,"%",sep=""))+ # optional, add percent signs
      labs(x = "Name",
         y = "Percent of All Babies With Name",
         title = "Most Popular Boys Names Since 1880",
         fill = "Boy's Name",
         caption = "Source: SSA")+
    theme_classic(base_family = "Fira Sans Condensed", base_size=16)+
  coord_flip()+ # rotate axes!
  theme(legend.position = "") # hide legend

```

```{r}
ggplot(data = top_5_girls_2017)+
  aes(x = reorder(name, n), #note this reorders the x variable from small to large n
      y = percent, # you can use prop if you didn't make a percent variable
      fill = name)+ # optional color!
  geom_col()+
  # now I'm just making it pretty
  scale_y_continuous(labels=function(x)paste(x,"%",sep=""))+ # optional, add percent signs
      labs(x = "Name",
         y = "Percent of All Girls With Name",
         title = "Most Popular Girls Names Since 1880",
         fill = "Girl's Name",
         caption = "Source: SSA")+
    theme_classic(base_family = "Fira Sans Condensed", base_size=16)+
  coord_flip()+ # rotate axes!
  theme(legend.position = "") # hide legend

```

## 3. 
**Find your name.^[If your name isn't in there :(, pick a random name.] `count` by `sex` how many babies since 1880 were named your name.^[Hint: if you do this, you'll get the number of *rows* (years) there are in the data. You want to add the number of babies in each row (`n`), so inside `count`, add `wt=n` to weight the count by `n`.] Also add a variable for the percent of each sex.**

```{r}
babynames %>%
  filter(name == "Ryan") %>%
  count(sex, wt=n) %>%
  mutate(percent = round((n/sum(n)*100),2))
```

## 4.
**Make a line graph of the number of babies with your name over time, `color`ed by `sex`.**
```{r}
babynames %>%
  filter(name == "Ryan") %>%
  ggplot(data = .)+
  aes(x = year,
      y = n,
      color = sex)+
  geom_line(size=1)+
  labs(x = "Year",
       y = "Number of Babies",
       title = "Popularity of Babies Named 'Ryan'",
       color = "Sex",
       caption = "Source: SSA")+
    theme_classic(base_family = "Fira Sans Condensed", base_size=16)
```

## 5.
### a.
**Make a table of the most common name for boys by year between 1980-2017.^[Hint: once you've got all the right conditions, you'll get a table with a lot of data. You only want to `slice` the `1`st row for each table.]**

```{r}
babynames %>%
  group_by(year) %>%
  filter(sex == "M",
         year>1980) %>%
  arrange(desc(n))%>%
  slice(1)
```

### b.
**Now do the same for girls.**

```{r}
babynames %>%
  group_by(year) %>%
  filter(sex == "F",
         year>1980) %>%
  arrange(desc(n))%>%
  slice(1)
```

## 6. Now let's graph the evolution of the most common names since 1880.
### a.
**First, find out what are the top 10 *overall* most popular names for boys and for girls. You may want to create two vectors, each with these top 5 names.**

```{r}
babynames %>%
  group_by(name) %>%
  filter(sex=="M") %>%
  summarize(total=sum(n)) %>%
  arrange(desc(total)) %>%
  slice(1:5)

top_boys_names<-c("James", "John", "Robert", "Michael", "William")
```

```{r}
babynames %>%
  group_by(name) %>%
  filter(sex=="F") %>%
  summarize(total=sum(n)) %>%
  arrange(desc(total)) %>%
  slice(1:5)

top_girls_names<-c("Mary", "Elizabeth", "Patricia", "Jennifer", "Linda")
```

### b.
**Now make two `line`graphs of these 5 names over time, one for boys, and one for girls.**

```{r}
babynames %>%
  group_by(year) %>%
  filter(sex == "M",
         name %in% top_boys_names) %>%
  ggplot(data = .,
         aes(x = year,
             y = prop,
             color = name))+
  geom_line()+
      labs(x = "Year",
         y = "Proportion of Babies with Name",
         title = "Most Popular Boys Names Since 1880",
         color = "Boy's Name",
         caption = "Source: SSA")+
    theme_classic(base_family = "Fira Sans Condensed", base_size=16)

```

```{r}
babynames %>%
  group_by(year) %>%
  filter(sex == "F",
         name %in% top_girls_names) %>%
  ggplot(data = .,
         aes(x = year,
             y = prop,
             color = name))+
  geom_line()+
    labs(x = "Year",
         y = "Proportion of Babies with Name",
         title = "Most Popular Girls Names Since 1880",
         color = "Girl's Name",
         caption = "Source: SSA")+
    theme_classic(base_family = "Fira Sans Condensed", base_size=16)
```

## 7.
**_Bonus (hard!)_: What are the 10 most common "gender-neutral" names?^[This is hard to define. For our purposes, let's define this as names where between 48 and 52% of the babies with the name are Male.]**

```{r}
babynames %>%
  mutate(male = ifelse(sex == "M", n, 0),
         female = ifelse(sex == "F", n, 0)) %>%
  group_by(name) %>%
    summarize(Male = sum(male),
              Female = sum(female))%>%
  filter(Male!=0,
         Female!=0) %>%
  mutate(perc_male = (Male/(Male+Female)*100)) %>%
  arrange(perc_male) %>%
  filter(perc_male > 48,
         perc_male < 52) %>%
  mutate(total = Male+Female) %>%
  arrange(desc(total)) %>%
  slice(1:10)
```

---

# Political and Economic Freedom Around the World

**For the remaining questions, we'll look at the relationship between Economic Freedom and Political Freedom in countries around the world today. Our data for economic freedom comes from the [Fraser Institute](https://www.fraserinstitute.org/economic-freedom/dataset?geozone=world&year=2016&page=dataset), and our data for political freedom comes from [Freedom House](https://freedomhouse.org/content/freedom-world-data-and-resources).**

## 8. 
**Download these two datasets that I've cleaned up a bit:^[If you want, try downloading them from the websites yourself!]**

- [<i class="fas fa-table"></i> `econfreedom.csv`](http://metricsf19.classes.ryansafner.com/data/econfreedom.csv)
- [<i class="fas fa-table"></i> `freedomhouse2018.csv`](http://metricsf19.classes.ryansafner.com/data/freedomhouse2018.csv)

**Load them with `df<-read_csv("name_of_the_file.csv")` and save one as `econfreedom` and the other as `polfreedom`. Look at each `tibble` you've created.**

```{r}
# note these file paths will be different for you
polfreedom<-read.csv(here::here("./static/data/freedomhouse2018.csv"))
econfreedom<-read.csv(here::here("./static/data/econfreedom.csv"))

polfreedom

econfreedom
```

## 9. 
**The `polfreedom` dataset is still a bit messy. Let's overwrite it (or assign to something like `polfreedom2`) and select `Country.Territory` and `Total` (total freedom score) and rename `Country.Territory` to `Country`.**

```{r}
polfreedom<-polfreedom %>%
  select(Country.Territory, Total) %>%
  rename(Country=Country.Territory)
```

## 10.
**Now we can try to merge these two datasets into one. Since they both have `Country` as a variable, we can merge these tibbles using `left_join(econfreedom, polfreedom, by="Country")`^[Note, if you saved as something else in question 9., use that instead of `polfreedom`!] and save this as a new tibble (something like `freedom`).**

```{r}
freedom<-left_join(econfreedom, polfreedom, by="Country")
```

## 11.
**Now make a scatterplot of Political Freedom (`total`)^[Feel free to `rename` these!] as `y` on Economic Freedom (`ef`) as `x` and `color` by `continent`.**

```{r, plot1, echo=F, fig.height=3, out.width="90%", fig.retina=3}
ggplot(data=freedom, aes(x=ef,y=Total))+
  geom_point(aes(color=continent))+
  xlab("Economic Freedom Score")+ylab("Political Freedom Score")+theme_bw()+labs(caption="Sources: Frasier Institute, Freedom House")+
    theme_classic(base_family = "Fira Sans Condensed", base_size=16)
```

## 12.
**Let's do this again, but highlight some key countries. Pick three countries, and make a new tibble from `freedom` that is only the observations of those countries. Additionally, *install* and *load* a packaged called `ggrepel`^[This automatically adjusts labels so they don't cover points on a plot!] Next, redo your plot from question 11, but now add a layer: `geom_label_repel` and set its `data` to your three-country tibble, use same `aes`thetics as your overall plot, but be sure to add `label = ISO`, to use the ISO country code to label.^[You might also want to set a low `alpha` level to make sure the labels don't obscure other points!]**

```{r, fig.retina=3}
# install.packages("ggrepel")
library(ggrepel)

interest<-filter(freedom, ISO %in% c("CHN", "NOR", "USA"))

ggplot(data=freedom, aes(x=ef,y=Total))+
  geom_point(aes(color=continent))+
  geom_label_repel(data=interest, aes(ef, Total, label=ISO,color=continent),alpha=0.6)+
  xlab("Economic Freedom Score")+ylab("Political Freedom Score")+theme_bw()+labs(caption="Sources: Frasier Institute, Freedom House")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=16)

```

## 13.
**Make another plot similar to 12, except this time use GDP per Capita (`gdp`) as `y`. Feel free to try to put a regression line with `geom_smooth()`!**^[If you do, be sure to set its data to the full `freedom`, not just your three countries!]. **Those of you in my Development course, you just made my graphs from [Lesson 2](https://devf19.classes.ryansafner.com/slides/02-slides#23)!**

```{r, fig.retina=3}
ggplot(data=freedom, aes(x=ef,y=gdp))+
  geom_point(aes(color=continent))+
  geom_smooth(data=freedom)+
  geom_label_repel(data=interest, aes(ef, Total, label=ISO,color=continent),alpha=0.6)+
  xlab("Economic Freedom Score")+ylab("Political Freedom Score")+theme_bw()+labs(caption="Sources: Frasier Institute, Freedom House")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=16)
```

### 2019-09-17 static/slides/07-slides.Rmd
Add class 7 content and revise schedule (separate 2.1 into 2.1 and 2.2, add problem set 2 and 3)

---
title: "2.1: Data 101 and Descriptive Statistics"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)

```

# Review From 1.2: Two Big Problems with Data

.pull-left[

- We want to use econometrics to **identify** causal relationships and make **inferences** about them

1. Problem for **identification**: .shout[endogeneity]

2. Problem for **inference**: .shout[randomness]

]

.pull-right[
.center[
![:scale 90%](https://www.dropbox.com/s/phhet82tcnuegxp/randomimage.jpg?raw=1)]
]

---

# Review from 1.2: Identification Problem: Endogeneity

.pull-left[

- An independent variable $(X)$ is .shout[exogenous] if its variation is *unrelated* to other factors that affect the dependent variable $(Y)$

- An independent variable $(X)$ is .shout[endogenous] if its variation is *related* to other factors that affect the dependent variable $(Y)$
]

.pull-right[

.center[
![](https://www.dropbox.com/s/v5vwsadw5vs448t/causality.jpg?raw=1)
]
]

---

# Review from 1.2: Inference Problem: Randomness

.pull-left[
- Data is random due to **natural sampling variation**

    - Taking one sample of a population will yield slightly different information than another sample of the same population

- Common in statistics, *easy to fix*

- **Inferential Statistics**: making claims about a wider population using sample data

    - We use common tools and techniques to deal with randomness
]

.pull-right[
.center[
![:scale 90%](https://www.dropbox.com/s/bsdtuddzjouwzr1/sampling.jpg?raw=1)
]
]

---

# The Two Problems: Where We're Heading...Ultimately

$$\text{Sample} \underbrace{\rightarrow}_{\text{statistical inference}} \text{Population} \underbrace{\rightarrow}_{\text{causal indentification}} \text{Unobserved Parameters}$$

- We want to .whisper[identify] causal relationships between **population** variables
    - Logically first thing to consider
    - **Endogeneity problem**

- We'll use **sample** *statistics* to .whisper[infer] something about population *parameters*
    - In practice, we'll only ever have a finite *sample distribution* of data
    - We *don't* know the *population distribution* of data
    - **Randomness problem**

---

class: inverse, center, middle
# Data 101

---

# Data 101

.pull-left[

- .hl[Data] are information with context
- .hl[Individuals] are the entities described by a set of data 
    - e.g. persons, households, firms, countries 
- .hl[Variables] are particular characteristics about an individual
    - e.g. age, income, profits, population, GDP, marital status, type of legal institutions
- .hl[Observations] or .hl[cases] are the separate individuals described by a collection of variables
    - e.g. for one individual, we have their age, sex, income, education, etc.  
    - individuals and observations are *not necessarily* the same: 
        - e.g. we can have separate observations on the same individual over time 
]

.pull-right[

![](https://www.dropbox.com/s/gmb5xihskovjrly/individual1.jpg?raw=1)

]

---

# Categorical Data

.pull-left[

- .hl[Categorial data] place an individual into one of several possible *categories*
    - e.g. sex, season, political party
    - may be responses to survey questions
    - can be quantitative (e.g. age, zip code)

- `R` calls these `factors` (we'll deal with them much later in the course)
]

.pull-right[

![](https://www.dropbox.com/s/pgdswagr7b348vu/categoricaldata.png?raw=1)

]

---

# Categorical Data: Visualizing I

.pull-left[

```{r, eval=F}
diamonds %>%
  count(cut) %>%
  mutate(frequency = n / sum(n))
```

```{r, echo=F, results="asis"}
diamonds %>%
  count(cut) %>%
  mutate(frequency = n / sum(n)) %>%
knitr::kable(., format = "html", caption = "Summary of diamonds by cut")
```

]

.pull-right[

- Good way to represent categorical data is with a .hl[frequency table]

- **Count (n)**: total number of individuals in a category

- **Frequency**: proportion of a category relative to all data

]

---

# Categorical Data: Visualizing II

.pull-left[

- Charts and graphs are *always* better ways to visualize data

- A .shout[bar graph] represents categories as bars, with lengths proportional to the count or relative frequency fo each category

```{r, echo=T, eval=F}
ggplot(diamonds, aes(x=cut,
                     fill=cut))+
  geom_bar()+
  guides(fill=F)+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
```

]

.pull-right[

```{r, fig.retina=3}
ggplot(diamonds, aes(x=cut, fill=cut))+
  geom_bar()+
  guides(fill=F)+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
```

]

---

# Categorical Data: Visualizing III

.pull-left[

- Avoid pie charts!

- People are *not* good at judging 2-d differences (angles, area)

- People *are* good at judging 1-d differences (length)

]

--

.pull-right[

.center[
![](https://www.dropbox.com/s/jk79mb94qohqekb/piechart.jpg?raw=1)
]

]

---

# Categorical Data: Visualizing IV

.pull-left[

- Maybe a *stacked bar chart*

```{r, echo=T, eval=F}
diamonds %>%
  count(cut) %>%
ggplot(., aes(x="", y=n, fill=cut))+
  geom_col()+
  geom_label(aes(x="", y=n, label=cut), position = position_stack(), color="white")+
  guides(fill=F)+
  theme_void()
```

]

.pull-right[

```{r, fig.retina=3}
diamonds %>%
  count(cut) %>%
ggplot(., aes(x="", y=n, fill=cut))+
  geom_col()+
  geom_label(aes(x="", y=n, label=cut), position = position_stack(), color="white")+
  guides(fill=F)+
  theme_void()

```

]

---

# Categorical Data: Visualizing IV

.pull-left[

- Maybe *lollipop chart*

```{r, echo=T, eval=F}
diamonds %>%
  count(cut) %>%
  mutate(cut_name = as.factor(cut)) %>%
ggplot(., aes(x = cut_name, y = n, color = cut))+
 geom_point(stat="identity",
            fill="black",
            size=12)  +
  geom_segment(aes(x = cut_name, y = 0,
                   xend = cut_name,
                   yend = n), size = 2)+
  geom_text(aes(label = n),color="white", size=3) +
  coord_flip()+
  labs(x = "Cut")+
  theme_classic(base_family = "Fira Sans Condensed",
                base_size=20)+
  guides(color = F)
```

]

.pull-right[

```{r, fig.retina=3}
diamonds %>%
  count(cut) %>%
  mutate(cut_name = as.factor(cut)) %>%
ggplot(., aes(x=cut_name, y=n, color=cut))+
 geom_point(stat='identity', fill="black", size=12)  +
  geom_segment(aes(x = cut_name, 
                   y = 0,
                   xend = cut_name,
                   yend = n),
               size = 2)+
  geom_text(aes(label = n),color="white", size=3) +
  coord_flip()+
  labs(x = "Cut")+
  theme_classic(base_family = "Fira Sans Condensed",
                base_size=20)+
  guides(color = F)
```

]

---

# Categorical Data: Visualizing IV

.pull-left[

- Maybe a *treemap*

```{r, echo=T, eval=F}
library(treemapify)
diamonds %>%
  count(cut) %>%
ggplot(., aes(area = n, fill = cut)) +
  geom_treemap() +
  guides(fill = FALSE) +
  geom_treemap_text(aes(label = cut), 
                    colour = "white",
                    place = "center",
                    grow = TRUE)
```

]

.pull-right[

```{r, fig.retina=3}
library(treemapify)
diamonds %>%
  count(cut) %>%
ggplot(., aes(area = n, fill = cut)) +
  geom_treemap() +
  guides(fill = FALSE) +
  geom_treemap_text(aes(label = cut), 
                    colour = "white", place = "center", grow = TRUE)
```

]

---

# Quantitative Data I

.pull-left[

- .hl[Quantitative variables] take on numerical values of equal units that describe an individual
    - Units: points, dollars, inches 
    - Context: GPA, prices, height

- We can mathematically manipulate *only* quantitative data
    - e.g. sum, average, standard deviation

]

.pull-right[

.image-75[![](https://www.dropbox.com/s/1e0mcehgzqgnfs8/mathoperations.jpg?raw=1)]

]

---

# Context is Key!

- How variables are classified depends on the *purpose* of collecting and using the data

--

.content-box-blue[
.blue[**Quick Check**]: What kind of data (categorical or quantitative) does each variable describe?
]

--

1. Age measured in years

--

2. Age measured in ranges (0-9 years, 10-19 years, 20-29 years, etc)

--

3. The date a purchase was made

--

4. Transaction ID

--

5. The amount of money spent on a Super Bowl ad

--

6. Customer ratings

--

7. The number of correct answers on an exam

---

# Discrete Data

.pull-left[

- .hl[Discrete data] are finite, with a countable number of alternatives

- Categorical: e.g. letter grades A, B, C, D, F

- Quantitative: integers, e.g. SAT Score, number of children

]

.pull-right[

![](https://www.dropbox.com/s/qfziu7r0wob0q1n/buildingblocks.jpeg?raw=1)
]

---

# Continuous Data

.pull-left[

- .hl[Continuous data] are infinitely divisible, with an uncountable number of alternatives
    - e.g. weights, temperature, GPA

- Many discrete variables may be treated as if they are continuous
    - e.g. SAT scores, wages

]

.pull-right[
.center[
![:scale 90%](https://www.dropbox.com/s/lgjpc9cxmrkio0x/continuous.png?raw=1)
]
]
---

# Discrete or Continuous?

.content-box-blue[
.blue[**Quick Check**]: What kind of data (discrete or continuous) does each variable describe?
]

--

1. Weight in pounds

--

2. Price in dollars

--

3. Grade (Letter)

--

4. Grade (Percentage)

--

5. Temperature

--

6. Amazon Star Rating

--

7. Number of customers

---

# Spreadsheets

.pull-left[

```{r, example.spreadsheet.setup, echo=FALSE}
ID<-c(1,2,3,4,5)
Name<-c("John","Emile","Natalya","Lakisha","Cheng")
Age<-c(23,18,28,31,36)
Sex<-c("Male","Male","Female","Female","Male")
Income<-c(41000,52600,48000,60200,81900)
example<-data.frame(ID,Name,Age,Sex,Income)
```

```{r, example.spreadsheet, results="asis", echo=FALSE}
knitr::kable(example, format="html")
```

]

.pull-right[

- The most common data structure we use is a .hl[spreadsheet]
    - Note, in *R*: a `data.frame` or `tibble`

- A **row** contains data about all variables for a single **individual**

- A **column** contains data about a single **variable** across all individuals

]

---

# Spreadsheets II

- It is common to use some notation like the following:

- Let $\{x_1, x_2, \cdots, x_n\}$ be a simple data series on variable $X$
    - $n$ individual observations
    - $x_i$ is the value of the $i$<sup>th</sup> observation for $i=1,2,\cdots, n$

--

.content-box-blue[
.blue[**Quick Check**]: Let $x$ represent the score on a homework assignment:
$$75, 100, 92, 87, 79, 0, 95$$

1. What is $n$?
2. What is $x_1$?
3. What is $x_6$?
]

---

# Datasets: Cross-Sectional

.pull-left[

```{r, example.spreadsheet2, echo=FALSE, results="asis"}
knitr::kable(example, format="html")
```

]

.pull-right[

- .hl[Cross-sectional data]: observations of individuals at a given point in time

- Each observation is a unique individual

- Simplest and most common data

- A "snapshot" to compare differences across individuals

]

---

# Datasets: Time-Series

.pull-left[

```{r, example.timeseries, echo=FALSE, results="asis"}
Year<-c(1950,1960,1970,1980,1985)
GDP<-c(8.2,9.9,10.2,12.4,13.6)
Unemployment<-c(0.06,0.04,0.08,0.08,0.06)
CPI<-c(100,118,130,190,196)
example.timeseries<-data.frame(Year,GDP,Unemployment,CPI)
knitr::kable(example.timeseries, format="html")
```

]

.pull-right[

- .hl[Time-series data]: observations of the same individuals over time

- Each observation is an individual-year

- Often used for macroeconomics, finance, and forecasting

- Unique challenges for time series

- A "moving picture" to see how individuals change over time

]

---

# Datasets: Panel

.pull-left[

```{r, example.panel.setup, echo=FALSE, results="asis"}
City<-c("Philadelphia","Philadelphia","D.C.","D.C.", "New York")
Year<-c(1986, 1990, 1986, 1990, 1986)
Murders<-c(5, 8, 2, 10, 3)
Population<-c(3.7,4.2,0.250,0.275,6.4)
Unemployment<-c(8.7,7.2,5.4,5.5,9.6)
example.panel<-data.frame(City,Year,Murders,Population,Unemployment)
knitr::kable(example.panel, format="html")
```

]

.pull-right[

- .hl[Panel], or .hl[longitudinal] dataset: a time-series for *each* cross-sectional entity

- Must be the *same* cross-sectional entities over time

- More common today for serious researchers

- Unique challenges for panel data

- A combination of "snapshot" comparisons and differences over time 
]

---

class: inverse, center, middle

# Descriptive Statistics

---

# Variables and Distributions

- Variables take on different values, we can describe a variable's .hl[distribution] (of these values)

- We want to *visualize* and *analyze* distributions to search for meaningful patterns using **statistics**

---

# Two Branches of Statistics

.pull-left[

- Two main branches of statistics:

1. .hl[Descriptive Statistics:] describes or summarizes the properties of a sample

2. .hl[Inferential Statistics:] infers properties about a larger population from the properties of a sample<sup>.red[1]</sup>

]

.pull-right[
.center[
![](https://www.dropbox.com/s/l2s6c0s55lsws6y/statsgraphs.jpg?raw=1)
]
]

.footnote[<sup>.red[1]</sup> We'll encounter inferential statistics mainly in the context of regression later.]

---

# Histograms

.pull-left[

- A common way to present a *quantitative* variable's distribution is a .hl[histogram]
    - The quantitative analog to the bar graph for a categorical variable

- Divide up values into **bins** of a certain size, and count the number of values falling within each bin, representing them visually as bars

]

.pull-right[

```{r, echo=F}
df<-tibble(x=rnorm(500,0,1))

ggplot(df, aes(x = x))+
  geom_histogram(color="white")+
  theme_bw(base_family = "Fira Sans Condensed", base_size=20)
```
]

---

# Histogram: Example

.pull-left[

.content-box-green[
.green[**Example**]: a class of 13 students takes a quiz (out of 100 points) with the following results:

$$\{ 0, 62, 66, 71, 71, 74, 76, 79, 83, 86, 88, 93, 95 \}$$
]
]

---

# Histogram: Example

.pull-left[

.content-box-green[
.green[**Example**]: a class of 13 students takes a quiz (out of 100 points) with the following results:

$$\{ 0, 62, 66, 71, 71, 74, 76, 79, 83, 86, 88, 93, 95 \}$$
]

```{r histogram-input, echo=T}
quizzes<-tibble(scores = c(0,62,66,71,71,74,76,79,83,86,88,93,95))
```
]

---

# Histogram: Example

.pull-left[

.content-box-green[
.green[**Example**]: a class of 13 students takes a quiz (out of 100 points) with the following results:

$$\{ 0, 62, 66, 71, 71, 74, 76, 79, 83, 86, 88, 93, 95 \}$$
]

```{r histogram-1, echo=T, eval=F}
h<-ggplot(quizzes,aes(x=scores))+
  geom_histogram(breaks = seq(0,100,10),
                 color = "black",
                 fill = "#56B4E9")+
  scale_x_continuous(breaks = seq(0,100,10))+
  labs(x = "Scores",
       y = "Number of Students")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
h
```
]

.pull-right[
.center[
```{r, ref.label="histogram-1", fig.retina=3}
```
]
]
---

# Descriptive Statistics

.pull-left[
- We are often interested in the *shape* or *pattern* of a distribution, particularly: 
    - Measures of **center**
    - Measures of **dispersion**
    - **Shape** of distribution 
]

.pull-right[
.center[
![](https://www.dropbox.com/s/l2s6c0s55lsws6y/statsgraphs.jpg?raw=1)
]
]

---

# Mode

- The .shout[mode] of a variable is simply its most frequent value

- A variable can have multiple modes

--

.content-box-green[
.green[**Example**]: a class of 13 students takes a quiz (out of 100 points) with the following results:

$$\{ 0, 62, 66, \mathbf{71}, \mathbf{71}, 74, 76, 79, 83, 86, 88, 93, 95 \}$$
]

---

# Mode

.pull-left[

- There is no dedicated `mode()` function in `R`, surprisingly

- A workaround in `dplyr`:

```{r mode, echo=T, eval=F}
quizzes %>%
  count(scores) %>%
  arrange(desc(n))
```

]

.pull-right[
```{r, ref.label="mode"}

```
]

---

# Multi-Modal Distributions

.pull-left[

- Looking at a histogram, the modes are the "peaks" of the distribution
    - Note: depends on how wide you make the bins!

- May be unimodal, bimodal, trimodal, etc

```{r, echo=T}
tibble(scores=c(0,33,33,33,33,35,62,66,71,71,74,76,79,83,86,88,93,95)) %>%
  count(scores) %>%
  arrange(desc(n))
```

]

.pull-right[
```{r, fig.retina=3}
tibble(scores=c(0,33,33,33,33,35,62,66,71,71,74,76,79,83,86,88,93,95)) %>%
  ggplot(data = .,
         aes(x=scores))+
  geom_histogram(breaks = seq(0,100,10),
                 color = "black",
                 fill = "#56B4E9")+
  scale_x_continuous(breaks = seq(0,100,10))+
  labs(x = "Scores",
       y = "Number of Students")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

---

# Symmetry and Skew I

.pull-left[

- A distribution is **symmetric** if it looks roughly the same on either side of the "center"

- The thinner ends (far left and far right) are called the **tails** of a distribution
]

.pull-right[
```{r, fig.retina=3}
symmetric<-data.frame(x=rnorm(500,25,10))

ggplot(symmetric,(aes(x)))+
    geom_histogram(stat="bin",bins=10,color="black",fill="green")+
    theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)

```
]

---

# Symmetry and Skew I

.pull-left[

- If one tail stretches farther than the other, distribution is **skewed** in the direction of the longer tail
]

.pull-right[
```{r, fig.retina=3}
skew.left<-data.frame(x=rbeta(500,5,1))

ggplot(skew.left,(aes(x)))+
    geom_histogram(stat="bin",bins=10,color="black",fill="indianred")+
    theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)

```
]

---

# Outliers

.pull-left[

- .shout[Outlier]: extreme value that does not appear part of the general pattern of a distribution

- Can strongly affect descriptive statistics

- Might be the most informative part of the data

- Could be the result of errors

- Should always be explored and discussed!

]

.pull-right[
```{r, fig.retina=3}
h
```
]

---

class: inverse, center, middle
# Measures of Center

---

# Arithmetic Mean (Population)

- The natural measure of the center of a *population*'s distribution is its "average" or .shout[arithmetic mean `\\((\mu)\\)`]

$$\mu=\frac{x_1+x_2+...+x_N}{N} = \frac{1}{N} \sum^N_{i=1} x_i$$

- For $N$ values of variable $x$, "mu" is the sum of all individual $x$ values $(x_i)$ from 1 to $N$, divided by the $N$ number of values<sup>.red[1]</sup>

- See today's class notes for more about the summation operator, $\displaystyle\Sigma$, it'll come up again!

.footnote[<sup>.red[1]</sup> Note the mean need not be an actual value of the data!]

---

# Arithmetic Mean (Sample)

- When we have a *sample*, we compute the .shout[sample mean `\\((\bar{x})\\)`]

$$\bar{x}=\frac{x_1+x_2+...+x_n}{n} = \frac{1}{n} \sum^n_{i=1} x_i$$

- For $n$ values of variable $x$, "x-bar" is the sum of all individual $x$ values $(x_i)$ divided by the $n$ number of values

--

.pull-left[

.content-box-green[
.green[**Example**]:

$$\{0, 62, 66, 71, 71, 74, 76, 79, 83, 86, 88, 93, 95\}$$

$$\begin{align*}
\bar{x}&=\frac{1}{13}(0+62+66+71+71+74+76+79+83+86+88+93+95)\\
\bar{x}&=\frac{944}{13}\\
\bar{x}&=72.62\\
\end{align*}$$

]
]

--

.pull-right[

```{r, echo=T}
quizzes %>%
  summarize(mean=mean(scores))
```
]

---

# Arithmetic Mean: Affected by Outliers

- If we drop the outlier (0)

--

.pull-left[

.content-box-green[
.green[**Example**]:

$$\{62, 66, 71, 71, 74, 76, 79, 83, 86, 88, 93, 95\}$$

$$\begin{align*}
\bar{x}&=\frac{1}{12}(62+66+71+71+74+76+79+83+86+88+93+95)
&=\frac{944}{12}\\
&=78.67\\
\end{align*}$$
]
]

--

.pull-right[

```{r, echo=T}
quizzes %>%
  filter(scores>0) %>%
  summarize(mean=mean(scores))
```
]

---

# Median

$$\{0, 62, 66, 71, 71, 74, \mathbf{76}, 79, 83, 86, 88, 93, 95\}$$

- The .shout[median] is the midpoint of the distribution
    - 50% to the left of the median, 50% to the right of the median

- Arrange values in numerical order
    - For odd $n$: median is middle observation
    - For even $n$: median is average of two middle observations

---

# Mean, Median, and Outliers

.center[
![](https://www.dropbox.com/s/znl2rdi3f19r6ap/meanoutliers.jpg?raw=1)
]

---

# Mean, Median, Symmetry, Skew I

.pull-left[

- A symmetric distribution has mean $\approx$ median

```{r}
symmetric<-tibble(x=c(1,2, 2, 3, 3, 3, 4, 4, 4, 4, 5, 5, 5, 6, 6, 7))
```

```{r, echo=T}
symmetric %>%
  summarize(mean = mean(x),
            median = median(x))
```
]

.pull-right[

```{r, fig.retina=3}
ggplot(symmetric,aes(x=x))+
  geom_histogram(breaks=seq(0,7,1),color="black",fill="#56B4E9")+
  geom_vline(aes(xintercept=mean(x)), color="red", linetype="dashed", size=1)+
  geom_vline(aes(xintercept=median(x)), color="green", linetype="dotted", size=2)+
  scale_x_continuous(breaks=seq(0,7,1))+
  labs(caption = "Mean: Red dashed line, Median: Green dotted line")+
      theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

---

# Mean, Median, Symmetry, Skew II

.pull-left[

- A left-skewed distribution has mean $<$ median

```{r}
leftskew<-data.frame(x=c(1,2,3,4,4,4,5,5,6,6,6,7,7))
```

```{r, echo=T}
leftskew %>%
  summarize(mean = mean(x),
            median = median(x))
```
]

.pull-right[

```{r, fig.retina=3}
ggplot(leftskew,aes(x=x))+
  geom_histogram(breaks=seq(0,7,1),color="black",fill="#56B4E9")+
  geom_vline(aes(xintercept=mean(x)), color="red", linetype="dashed", size=1)+
  geom_vline(aes(xintercept=median(x)), color="green", linetype="dotted", size=2)+
  scale_x_continuous(breaks=seq(0,7,1))+
  labs(caption = "Mean: Red dashed line, Median: Green dotted line")+
      theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

---

# Mean, Median, Symmetry, Skew III

.pull-left[

- A right-skewed distribution has mean $>$ median

```{r}
rightskew<-tibble(x=c(1,1,2,2,2,3,3,4,4,4,5,6,7))
```

```{r, echo=T}
rightskew %>%
  summarize(mean = mean(x),
            median = median(x))
```
]

.pull-right[

```{r, fig.retina=3}
ggplot(rightskew,aes(x=x))+
  geom_histogram(breaks=seq(0,7,1),color="black",fill="#56B4E9")+
  geom_vline(aes(xintercept=mean(x)), color="red", linetype="dashed", size=1)+
  geom_vline(aes(xintercept=median(x)), color="green", linetype="dotted", size=2)+
  scale_x_continuous(breaks=seq(0,7,1))+
  labs(caption = "Mean: Red dashed line, Median: Green dotted line")+
      theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

---

class: inverse, center, middle
# Measures of Spread

---

# Measures of Spread: Range

- The more *variation* in the data, the less helpful a measure of central tendency will tell us
- Beyond just the center, we also want to measure the spread
- Simplest metric is .shout[range] $=max-min$

---

# Measures of Spread: 5 Number Summary I

A common set of summary statistics about a distribution is known as the .shout["five number summary"]:

.pull-left[
.font90[
1. Minimum value
2. 25<sup>th</sup> percentile $(Q_1$, the median of the first 50% of data)
3. 50<sup>th</sup> percentile (median, $Q_2)$
4. 25<sup>th</sup> percentile $(Q_3$, the median of the last 50% of data)
5. Maximum value
]
]
.pull-right[

.font90[
```{r, echo=T}
# Base R summary command (includes Mean)
summary(quizzes$scores)

quizzes %>% # dplyr
  summarize(Min = min(scores),
            Q1 = quantile(scores, 0.25),
            Median = median(scores),
            Q3 = quantile(scores, 0.75),
            Max = max(scores))
```
]
]
---

# Measures of Spread: 5 Number Summary II

- The $n$<sup>th</sup> percentile of a distribution is the value that places $n$ percent of values beneath it

```{r, echo = T}
quizzes %>%
  summarize("37th percentile" = quantile(scores,0.37))
```

---

# Boxplots I

.pull-left[

- .shout[Boxplots] are a great way to visualize the 5 number summary

- Height of box: $Q_1$ to $Q_3$ (known as .onfire[interquartile range (IQR)], middle 50% of data)
- Line inside box: median (50<sup>th</sup> percentile)
- "Whiskers" identify data within $1.5 \times IQR$
- Points *beyond* whiskers are a common definition of **outliers**
]
.pull-right[

```{r}
quizzes %>%
  ggplot(data = .)+
  aes(x = "",
      y = scores)+
  geom_boxplot()+
    labs(x = "Quiz")+
      theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)

```

]

---

# Comparisons I

- Boxplots (and five number summaries) are great for comparing two distributions

.content-box-green[
.green[**Example**]:

$$\begin{align*} \text{Quiz 1}&: \{0, 62, 66, 71, 71, 74, 76, 79, 83, 86, 88, 93, 95\}	\\
			\text{Quiz 2}&: \{50, 62, 72, 73, 79, 81, 82, 82, 86, 90, 94, 98, 99\} \\ \end{align*}$$

]

```{r}
quizzes_new <- quizzes %>%
  rename(quiz_1 = scores) %>% # rename to Quiz 1
  mutate(student = 1:13) %>% # add students to track over 2 quizzes
  select(student, everything()) %>% # move students to front
  bind_cols(quiz_2 = c(50, 62, 72, 73, 79, 81, 82, 82, 86, 90, 94, 98, 99))
```

---

# Comparisons II

.pull-left[

```{r, echo=T}
quizzes_new %>% summary(quiz_1)

quizzes_new %>% summary(quiz_2)
```
]

.pull-right[

```{r}
quizzes_long<-gather(quizzes_new, key="student", value="score")

ggplot(data = quizzes_long)+
  aes(x = student,
      y = score,
      fill = student)+
  geom_boxplot()+
    labs(x = "Quiz")+
  scale_x_discrete(labels=c("Quiz 1", "Quiz 2"))+
  scale_fill_viridis_d()+
  guides(fill=F)+
      theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
```

]

---

# Aside: Making Nice Summary Tables I

- I don't like the options available for printing out summary statistics

- So I wrote my own `R` function that makes nice summary tables that uses `dplyr` and `tidyr`

- One day I will release as a package; until then the `.R` file is [saved here](/files/summaries.R)

```{r, echo=T}
# loads .R files to full functions from; make sure it's in working directory/Project!
source("../files/summaries.R")

# let's summarize "hwy" and "cty" from mpg dataset
mpg %>%
  summary_table(hwy, cty)
```

---

# Aside: Making Nice Summary Tables II

- And when `knit`ted in `R markdown`:

```{r, echo=T, results="asis"}
mpg %>%
  summary_table(hwy, cty) %>%
  knitr::kable(., format="html")
```

- We'll talk more about using `markdown` and making final products nicer when we discuss your paper project (have you forgotten?)

---

# Deviations

- Every observation .onfire[deviates] from the mean of the data: 
$$deviation_i = x_i-\mu	$$

- There are as many deviations as there are data points $(n)$
- We can measure the *average* or .shout[standard deviation] of a variable from its mean

---

# Variance (Population)

- The .shout[population variance `\\((\sigma^2)\\)`] of a *population* distribution measures the average of the *squared* deviations from the *population* mean

$$\sigma^2 = \frac{1}{N}\displaystyle\sum^N_{i=1} (x_i-\mu)^2$$

- Why do we square deviations?

- What are these units?

---

# Standard Deviation (Population)

- Square root the variance to get the .shout[population standard deviation `\\((\sigma)\\)`], the average deviation from the population mean (in same units as $x$)

$$\sigma=\sqrt{\sigma^2}=\sqrt{\frac{1}{N}\displaystyle\sum^N_{i=1} (x_i-\mu)^2	}$$

---

# Variance (Sample)

- The .shout[sample variance `\\((s^2)\\)`] of a *sample* distribution measures the average of the *squared* deviations from the *sample* mean

$$\sigma^2 = \frac{1}{n-1}\displaystyle\sum^n_{i=1} (x_i-\mu)^2$$

- Why do we divide by $n-1$?

---

# Standard Deviation (Sample)

- Square root the sample variance to get the .shout[sample standard deviation `\\((s)\\)`], the average deviation from the *sample* mean (in same units as $x$)

$$s=\sqrt{s^2}=\sqrt{\frac{1}{n-1}\displaystyle\sum^n_{i=1} (x_i-\mu)^2	}$$

---

# Sample Standard Deviation: Example

.content-box-green[
.green[**Example**]: Calculate the sample standard deviation for the following series:

$$\{2, 4, 6, 8, 10 \}$$

]

--

```{r, echo=T}
sd(c(2,4,6,8,10))
```

---

# The Steps to Calculate sd(), Coded I

```{r,echo =T}
#  first let's save our data in a tibble
sd_example<-tibble(x=c(2,4,6,8,10))

# first find the mean (just so we know)

sd_example %>%
  summarize(mean(x))

# now let's make some more columns:
sd_example <- sd_example %>%
  mutate(deviations = x-mean(x), # take deviations from mean
         deviations_sq = deviations^2) # square them

```

---

# The Steps to Calculate sd(), Coded II

.pull-left[
```{r, echo=T, eval=F}
sd_example # see what we made
```
]

.pull-right[
```{r}
sd_example
```
]
---

# The Steps to Calculate sd(), Coded III

.pull-left[
```{r, echo=T, eval=F}
sd_example %>%
  # sum the squared deviations
  summarize(sum_sq_devs = sum(deviations_sq), 
            # divide by n-1 to get variance
            variance = sum_sq_devs/(n()-1), 
            # square root to get sd
            std_dev = sqrt(variance)) 
```
]
.pull-right[
```{r}
sd_example %>%
  summarize(sum_sq_devs = sum(deviations_sq), 
            variance = sum_sq_devs/(n()-1),
            std_dev = sqrt(variance)) 
```

]

---

# Sample Standard Deviation: You Try

.content-box-blue[
.blue[**You Try**]: Calculate the sample standard deviation for the following series:

$$\{1, 3, 5, 7 \}$$

]

--

```{r, echo=T}
sd(c(1,3,5,7))
```

---

# Descriptive Statistics: Populations vs. Samples

.pull-left[
## Population parameters

- **Population size**: $N$

- **Mean**: $\mu$

- **Variance**: $\sigma^2=\frac{1}{N} \displaystyle\sum^N_{i=1} (x_i-\mu)^2$

- **Standard deviation**: $\sigma = \sqrt{\sigma^2}$
]

.pull-right[
## Sample statistics

- **Population size**: $n$

- **Mean**: $\bar{x}$

- **Variance**: $s^2=\frac{1}{n-1} \displaystyle\sum^n_{i=1} (x_i-\bar{x})^2$

- **Standard deviation**: $s = \sqrt{s^2}$

]

### 2019-09-17 static/slides/07-slides.Rmd
Tweaks to slides 7

@@
 
-# Data 101
+# Data 101 I
 
@@
 - .hl[Data] are information with context
+
 - .hl[Individuals] are the entities described by a set of data 
     - e.g. persons, households, firms, countries
+
+]
+
+.pull-right[
+
+![](https://www.dropbox.com/s/gmb5xihskovjrly/individual1.jpg?raw=1)
+
+]
+
+---
+
+# Data 101 I
+
+.pull-left[
+
 - .hl[Variables] are particular characteristics about an individual
     - e.g. age, income, profits, population, GDP, marital status, type of legal institutions
+
 - .hl[Observations] or .hl[cases] are the separate individuals described by a collection of variables
     - e.g. for one individual, we have their age, sex, income, education, etc.
+
 - individuals and observations are *not necessarily* the same: 
@@
 
-- .hl[Categorial data] place an individual into one of several possible *categories*
+- .hl[Categorical data] place an individual into one of several possible *categories*
     - e.g. sex, season, political party
@@
 
-- See today's class notes for more about the summation operator, $\displaystyle\Sigma$, it'll come up again!
+- See today's class notes for more about the .onfire[summation operator, `\\(\displaystyle\Sigma\\)`], it'll come up again!
 
@@
 ]
+
+--
+
 .pull-right[
@@
 
-- Height of box: $Q_1$ to $Q_3$ (known as .onfire[interquartile range (IQR)], middle 50% of data)
-- Line inside box: median (50<sup>th</sup> percentile)
-- "Whiskers" identify data within $1.5 \times IQR$
-- Points *beyond* whiskers are a common definition of **outliers**
+- **Height of box**: $Q_1$ to $Q_3$ (known as .onfire[interquartile range (IQR)], middle 50% of data)
+
+- **Line inside box**: median (50<sup>th</sup> percentile)
+
+- **"Whiskers"** identify data within $1.5 \times IQR$
+
+- Points *beyond* whiskers are **outliers**
+    - common definition: $Outlier >1.5 \times IQR$
 ]
@@
 ```{r, echo=T}
-quizzes_new %>% summary(quiz_1)
-
-quizzes_new %>% summary(quiz_2)
+quizzes_new %>% summary()
 ```
@@
 
-- So I wrote my own `R` function that makes nice summary tables that uses `dplyr` and `tidyr`
+- So I wrote my own `R function` that makes nice summary tables that uses `dplyr` and `tidyr`
 
@@
 ```{r, echo=T}
-# loads .R files to full functions from; make sure it's in working directory/Project!
-source("../files/summaries.R")
-
-# let's summarize "hwy" and "cty" from mpg dataset
+# loads .R files to full functions from; make sure it's in YOUR working directory/Project!
+source("../files/summaries.R") # MY path (this website's R Project)
+
+# let's summarize variables "hwy" and "cty" from *mpg* dataset
 mpg %>%
@@
 
-- Every observation .onfire[deviates] from the mean of the data: 
+- Every observation $i$ .onfire[deviates] from the mean of the data: 
 $$deviation_i = x_i-\mu	$$
@@
 - There are as many deviations as there are data points $(n)$
+
 - We can measure the *average* or .shout[standard deviation] of a variable from its mean
 
+- Before we get there...
+
 ---
@@
 
-- The .shout[population variance `\\((\sigma^2)\\)`] of a *population* distribution measures the average of the *squared* deviations from the *population* mean
+- The .shout[population variance `\\((\sigma^2)\\)`] of a *population* distribution measures the average of the *squared* deviations from the *population* mean $(\mu)$
 
@@
 
-- The .shout[sample variance `\\((s^2)\\)`] of a *sample* distribution measures the average of the *squared* deviations from the *sample* mean
-
-$$\sigma^2 = \frac{1}{n-1}\displaystyle\sum^n_{i=1} (x_i-\mu)^2$$
+- The .shout[sample variance `\\((s^2)\\)`] of a *sample* distribution measures the average of the *squared* deviations from the *sample* mean $(\bar{x})$
+
+$$\sigma^2 = \frac{1}{n-1}\displaystyle\sum^n_{i=1} (x_i-\bar{x})^2$$
 
@@
 
-$$s=\sqrt{s^2}=\sqrt{\frac{1}{n-1}\displaystyle\sum^n_{i=1} (x_i-\mu)^2	}$$
+$$s=\sqrt{s^2}=\sqrt{\frac{1}{n-1}\displaystyle\sum^n_{i=1} (x_i-\bar{x})^2	}$$
 

### 2019-09-19 content/assignment/02-problem-set.Rmd
Add class 8 content

--- 2019-09-13 content/assignment/01-problem-set.Rmd
@@
 
-# The Popularity of Baby Names
-
-Install and load the package `babynames`. Get help for `?babynames` to see what the data includes.
+# Theory and Concepts
 
 ## 1.
+
 ### a.
-What are the top 5 boys names for 2017, and what *percent* of overall names is each?
+
+In your own words, explain what (sample) standard deviation *means*.
 
 ### b.
-What are the top 5 *girls* names, and what *percent* of overall names is each?
 
-## 2. 
-Make two barplots, of these top 5 names, one for each sex.
-Map `aes`thetics `x` to `name` and `y` to `prop`^[Or `percent`, if you made that variable, as I did.] and use `geom_col` (since you are declaring a specific `y`, otherwise you could just use `geom_bar()` and just an `x`.)
-
-## 3. 
-Find your name.^[If your name isn't in there :(, pick a random name.] `count` by `sex` how many babies since 1880 were named your name.^[Hint: if you do this, you'll get the number of *rows* (years) there are in the data.
-You want to add the number of babies in each row (`n`), so inside `count`, add `wt=n` to weight the count by `n`.] Also add a variable for the percent of each sex.
-
-## 4.
-Make a line graph of the number of babies with your name over time, `color`ed by `sex`.
-
-## 5.
-### a.
-Make a table of the most common name for boys by year between 1980-2017.^[Hint: once you've got all the right conditions, you'll get a table with a lot of data.
-You only want to `slice` the `1`st row for each table.]
-
-### b.
-Now do the same for girls.
-
-## 6. 
-Now let's graph the evolution of the most common names since 1880.
-### a.
-First, find out what are the top 10 *overall* most popular names for boys and for girls. You may want to create two vectors, each with these top 5 names.
-
-### b.
-Now make two `line`graphs of these 5 names over time, one for boys, and one for girls.
-
-## 7.
-**Bonus (hard!)**: What are the 10 most common "gender-neutral" names?^[This is hard to define.
-For our purposes, let's define this as names where between 48 and 52% of the babies with the name are Male.]
-
-# Political and Economic Freedom Around the World
-
-For the remaining questions, we'll look at the relationship between Economic Freedom and Political Freedom in countries around the world today.
-Our data for economic freedom comes from the [Fraser Institute](https://www.fraserinstitute.org/economic-freedom/dataset?geozone=world&year=2016&page=dataset), and our data for political freedom comes from [Freedom House](https://freedomhouse.org/content/freedom-world-data-and-resources).
-
-## 8. 
-Download these two datasets that I've cleaned up a bit:^[If you want, try downloading them from the websites yourself!]
-
-- [<i class="fas fa-table"></i> `econfreedom.csv`](http://metricsf19.classes.ryansafner.com/data/econfreedom.csv)
-- [<i class="fas fa-table"></i> `freedomhouse2018.csv`](http://metricsf19.classes.ryansafner.com/data/freedomhouse2018.csv)
-
-Load them with `df<-read_csv("name_of_the_file.csv")` and save one as `econfreedom` and the other as `polfreedom`. Look at each `tibble` you've created.
-
-## 9. 
-The `polfreedom` dataset is still a bit messy.
-Let's overwrite it (or assign to something like `polfreedom2`) and select `Country.Territory` and `Total` (total freedom score) and rename `Country.Territory` to `Country`.
-
-## 10.
-Now we can try to merge these two datasets into one.
-Since they both have `Country` as a variable, we can merge these tibbles using `left_join(econfreedom, polfreedom, by="Country")`^[Note, if you saved as something else in question 9., use that instead of `polfreedom`!] and save this as a new tibble (something like `freedom`).
-
-## 11.
-Now make a scatterplot of Political Freedom (`total`)^[Feel free to `rename` these!] as `y` on Economic Freedom (`ef`) as `x` and `color` by `continent`.
-
-## 12.
-Let's do this again, but highlight some key countries.
-Pick three countries, and make a new tibble from `freedom` that is only the observations of those countries.
-Additionally, *install* and *load* a packaged called `ggrepel`^[This automatically adjusts labels so they don't cover points on a plot!] Next, redo your plot from question 11, but now add a layer: `geom_label_repel` and set its `data` to your three-country tibble, use same `aes`thetics as your overall plot, but be sure to add `label = ISO`, to use the ISO country code to label.^[You might also want to set a low `alpha` level to make sure the labels don't obscure other points!]
-
-## 13.
-Make another plot similar to 12, except this time use GDP per Capita (`gdp`) as `y`.
-Feel free to try to put a regression line with `geom_smooth()`!^[If you do, be sure to set its data to the full `freedom`, not just your three countries!] Those of you in my Development course, you just made my graphs from [Lesson 2](https://devf19.classes.ryansafner.com/slides/02-slides#23)!
+In your own words, explain how (sample) standard deviation is calculated. You may also write the formula, but it is not necessary.

### 2019-09-19 content/class/08-class.Rmd
Add class 8 content

---
title: "2.2: Random Variables and Distributions - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

*Thursday, September 19, 2019*

# Overview

Today we finish your crash course/review of basic statistics with random variables and distributions.

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/08-slides.html)

# Assignments: Problem Set 1 DUE and Problem Set 2

[Problem Set 1](/assignment/01-problem-set.html) is DUE

Problem set 2 (on classes 2.1-2.2) will be posted shortly, and is due by Thursday September 26.

# Appendix: Graphing Statistical and Mathematical Functions in R

Check back later today!

### 2019-09-19 static/slides/08-slides.Rmd
Add class 8 content

---
title: "2.2: Random Variables and Distributions"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))
```

class: inverse, center, middle
# Random Variables

---

# Experiments

.pull-left[

- An .onfire[experiment] is any procedure that can (in principle) be repeated infinitely and has a well-defined set of outcomes

.content-box-green[
.green[**Example**]: flip a coin 10 times

]
]

.pull-right[
.center[
![](https://www.dropbox.com/s/o0iunsic2clueo6/headsandtails.jpg?raw=1)
]
]

---

# Random Variables

.pull-left[

- A .onfire[random variable (RV)] takes on values that are unknown in advance, but determined by an experiment

- A numerical summary of a random outcome

.content-box-green[
.green[**Example**]: the number of heads from 10 coin flips

]
]

.pull-right[
.center[
![](https://www.dropbox.com/s/o0iunsic2clueo6/headsandtails.jpg?raw=1)
]
]

---

# Random Variables: Notation

- Random variable $X$ takes on individual values $(x_i)$ from a set of possible values

- Often capital letters to denote RV's
    - lowercase letters for individual values

.content-box-green[
.green[**Example**]: Let $X$ be the number of Heads from 10 coin flips. $\quad x_i \in \{0, 1, 2,...,10\}$
]

---

class: inverse, center, middle
# Random Variables

---

# Discrete Random Variables

.pull-left[

- A .shout[discrete random variable]: takes on a finite/countable set of possible values

.content-box-green[
.green[**Example**]: Let $X$ be the number of times your computer crashes this semester<sup>.red[1]</sup>, $x_i \in \{0, 1, 2, 3, 4\}$
]

]

.pull-right[

.center[
![](https://www.dropbox.com/s/uj0aa30d9ct4lgh/bluescreen.jpg?raw=1)
]
]

.footnote[<sup>.red[1]</sup> Please, back up your files!]

---

# Discrete Random Variables: Probability Distribution

.pull-left[
- .shout[Probability distribution] of a R.V. fully lists all the possible values of $X$ and their associated probabilities
]

.pull-right[

.green[**Example**]:

| $x_i$  | $P(X=x_i)$ |
|--------|------------|
| 0      | 0.80       |
| 1      | 0.10       |
| 2      | 0.06       |
| 3      | 0.03       |
| 4      | 0.01       |

]

---

# Discrete Random Variables: pdf

.pull-left[
.shout[Probability distribution function (pdf)] summarizes the possible outcomes of $X$ and their probabilities

- Notation: $f_X$ is the pdf of $X$:

$$f_X=p_i, \quad i=1,2,...,k$$

- For any real number $x_i$, $f(x_i)$ is the probablity that $X=x_i$

]

.pull-right[
.green[**Example**]:

| $x_i$  | $P(X=x_i)$ |
|--------|------------|
| 0      | 0.80       |
| 1      | 0.10       |
| 2      | 0.06       |
| 3      | 0.03       |
| 4      | 0.01       |

]

--

- What is $f(0)$?

--

- What is $f(3)$?

---

# Discrete Random Variables: pdf Graph

.pull-left[

```{r, echo=T, eval=F}
crashes<-tibble(number = c(0,1,2,3,4),
                prob = c(0.80, 0.10, 0.06, 0.03, 0.01))

ggplot(data = crashes)+
  aes(x = number,
      y = prob)+
  geom_col(fill="#0072B2")+
  labs(x = "Number of Crashes",
       y = "Probability")+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

]

.pull-right[

```{r, fig.retina=3}
crashes<-tibble(number = c(0,1,2,3,4),
                prob = c(0.80, 0.10, 0.06, 0.03, 0.01))

ggplot(data = crashes)+
  aes(x = number,
      y = prob)+
  geom_col(fill="#0072B2")+
  labs(x = "Number of Crashes",
       y = "Probability")+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

---

# Discrete Random Variables: cdf

.pull-left[
.shout[Cumulative distribution function (pdf)] lists probability $X$ will be *at most* (less than or equal to) a given value $x_i$

- Notation: $F_X=P(X \leq x_i)$

]

.pull-right[

.green[**Example**]:

| $x_i$  | $f(x)$ | $F(x)$ |
|--------|--------|--------|
| 0      | 0.80   | 0.80   |
| 1      | 0.10   | 0.90   |
| 2      | 0.06   | 0.96   |
| 3      | 0.03   | 0.99   |
| 4      | 0.01   | 1.00   |

]

--

- What is the probability your computer will crash *at most* once, $F(1)$?

--

- What about three times, $F(3)$?

---

# Discrete Random Variables: cdf Graph

.pull-left[

```{r, echo=T, eval=T}
crashes<-crashes %>%
  mutate(cum_prob = cumsum(prob))

crashes

```

```{r, echo=T, eval=F}
ggplot(data = crashes)+
  aes(x = number,
      y = cum_prob)+
  geom_col(fill="#0072B2")+
  labs(x = "Number of Crashes",
       y = "Probability")+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

]

.pull-right[

```{r, fig.retina=3}
crashes<-crashes %>%
  mutate(cum_prob = cumsum(prob))

ggplot(data = crashes)+
  aes(x = number,
      y = cum_prob)+
  geom_col(fill="#0072B2")+
  labs(x = "Number of Crashes",
       y = "Probability")+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

---

class: inverse, center, middle
# Expected Value and Variance

---

# Expected Value of a Random Variable

- .shout[Expected value] of a random variable $X$, written $E(X)$, is the long-run average value of $X$ "expected" after many repetitions

$$E(X)=\sum^k_{i=1} p_i x_i$$

--

- $E(X)=p_1x_1+p_2x_2+ \cdots +p_kx_k$

- A **probability-weighted average** of $X$, with each $x_i$ weighted by its associated probability $p_i$

- Also called the .shout["mean"] or .shout["expectation"] of $X$, always denoted either $E(X)$ or $\mu_X$

---

# Expected Value: Example I

.content-box-green[
.green[**Example**]: Suppose you lend your friend $100 at 10% interest. If the loan is repaid, you receive $110. You estimate that your friend is 99% likely to repay, but there is a default risk of 1% where you get nothing. What is the expected value of repayment?

]

---

# Expected Value: Example II

.green[**Example**]:

Let $X$ be a random variable that is described by the following pdf:
]

| $x_i$  | $P(X=x_i)$ |
|--------|------------|
| 1      | 0.50       |
| 2      | 0.25       |
| 3      | 0.15       |
| 4      | 0.10       |

Calculate $E(X)$.

---

# The Steps to Calculate E(X), Coded

```{r, echo=T}
# Make a Random Variable called X
X<-tibble(x_i=c(1,2,3,4), # values of X
          p_i=c(0.50,0.25,0.15,0.10)) # probabilities

X %>%
  summarize(expected_value = sum(x_i*p_i))
```

---

# Variance of a Random Variable

- The .shout[variance] of a random variable $X$, denoted $var(X)$ or $\sigma^2_X$ is:

$$\begin{align*}\sigma^2_X &= E[(X-\mu_X)^2]\\
&=\sum_{i=1}^n(x_i-\mu_X)^2p_i\\ \end{align*}$$

- This is the .onfire[expected value of the squared deviations from the mean]
    - i.e. the probability-weighted average of the squared deviations

---

# Standard Deviation of a Random Variable

- The .shout[standard deviation] of a random variable $X$, denoted $sd(X)$ or $\sigma_X$ is:

$$\sigma_X=\sqrt{\sigma_X^2}$$

---

# Standard Deviation: Example I

.content-box-green[
.green[**Example**]: What is the standard deviation of computer crashes?
]

| $x_i$  | $P(X=x_i)$ |
|--------|------------|
| 0      | 0.80       |
| 1      | 0.10       |
| 2      | 0.06       |
| 3      | 0.03       |
| 4      | 0.01       |

---

# The Steps to Calculate sd(X), Coded I

```{r, echo=T}
# get the expected value 
crashes %>%
  summarize(expected_value = sum(number*prob))

# save this for quick use
exp_value<-0.35

crashes_2 <- crashes %>%
  select(-cum_prob) %>% # we don't need the cdf
  # create new columns
  mutate(deviations = number - exp_value, # deviations from exp_value
         deviations_sq = deviations^2,
         weighted_devs_sq = prob * deviations^2) # square deviations
```

---

# The Steps to Calculate sd(X), Coded II

```{r, echo=T}
# look at what we made
crashes_2
```

---

# The Steps to Calculate sd(X), Coded III

```{r, echo=T}
# now we want to take the expected value of the squared deviations to get variance
crashes_2 %>%
  summarize(variance = sum(weighted_devs_sq), # variance
            sd = sqrt(variance)) # sd is square root
```

---

# Standard Deviation: Example II

.content-box-green[
.green[**Example**]: What is the standard deviation of the random variable we saw before?
]

| $x_i$  | $P(X=x_i)$ |
|--------|------------|
| 1      | 0.50       |
| 2      | 0.25       |
| 3      | 0.15       |
| 4      | 0.10       |

Hint: you already found it's expected value.

---

class: inverse, center, middle
# Continuous Random Variables

---

# Continuous Random Variables

.pull-left[

- .shout[Continuous random variables] can take on an uncountable (infinite) number of values

- So many values that the probability of any specific value is infinitely small:
$$P(X=x_i)\rightarrow 0$$

- Instead, we focus on a *range* of values it might take on

]

.pull-right[

.center[
![:scale 100%](https://www.dropbox.com/s/lgjpc9cxmrkio0x/continuous.png?raw=1)
]
]

---

# Continuous Random Variables: pdf I

.pull-left[
.shout[Probability *density* function (pdf)] of a continuous variable represents the probability between two values as the area under a curve

- The total area under the curve is 1

- Since $P(a)=0$ and $P(b)=0$, $P(a<X<b)=P(a \leq X \leq b)$

- See [today's class notes](/class/08-class) for how to graph math/stats functions in `ggplot`!
]

.pull-right[

```{r, fig.retina=3}

shade_0_to_2 <- function(x) {
    y <- dnorm(x, mean = 0, sd = 1)
    y[x < 0 | x >2] <- NA
    return(y)
}

ggplot(data = tibble(x=-4:4))+
  aes(x = x)+
  stat_function(fun = dnorm, size=2, color="blue")+
  stat_function(fun=shade_0_to_2, geom="area", fill="blue", alpha=0.5)+
  labs(x = "X",
       y = "Probability")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

.center[
$P(0 \leq X \leq 2)=$ area under the curve between 0 and 2
]
]

---

# Continuous Random Variables: pdf II

.pull-left[

- FYI using calculus:

$$P(a \leq X \leq b) = \int_a^b f(x) dx $$

- Complicated: software or (old fashioned!) probability tables to calculate

]

.pull-right[

```{r, fig.retina=3}

shade_0_to_2 <- function(x) {
    y <- dnorm(x, mean = 0, sd = 1)
    y[x < 0 | x >2] <- NA
    return(y)
}

ggplot(data = tibble(x=-4:4))+
  aes(x = x)+
  stat_function(fun = dnorm, size=2, color="blue")+
  stat_function(fun=shade_0_to_2, geom="area", fill="blue", alpha=0.5)+
  labs(x = "X",
       y = "Probability")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

.center[
$P(0 \leq X \leq 2)=$ area under the curve between 0 and 2
]
]

---

# Continuous Random Variables: cdf I

.pull-left[

- The .shout[cumulative density function (cdf)] describes the area under the pdf for all values less than or equal to (i.e. to the left of) a given value, $k$

$$P(X \leq k)$$

]

.pull-right[

```{r, fig.retina=3}

shade_up_to_2 <- function(x) {
    y <- pnorm(x, mean = 0, sd = 1)
    y[x < -4 | x >2] <- NA
    return(y)
}

ggplot(data = tibble(x=-4:4))+
  aes(x = x)+
  stat_function(fun = pnorm, size=2, color="blue")+
  stat_function(fun=shade_up_to_2, geom="area", fill="blue", alpha=0.5)+
  labs(x = "X",
       y = "Probability")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

.center[
$P(X \leq 2)=$ area under the curve to the left of 2
]
]

---

# Continuous Random Variables: cdf II

.pull-left[

- Note: to find the probability of values *greater* than or equal to (to the right of) a given value $k$:

$$P(X \geq k)=1-P(X \leq k)$$

$$P(X \leq k)$$

]

.pull-right[

```{r, fig.retina=3}

shade_2_up <- function(x) {
    y <- pnorm(x, mean = 0, sd = 1)
    y[x < 2 | x >4] <- NA
    return(y)
}

ggplot(data = tibble(x=-4:4))+
  aes(x = x)+
  stat_function(fun = pnorm, size=2, color="blue")+
  stat_function(fun=shade_2_up, geom="area", fill="blue", alpha=0.5)+
  labs(x = "X",
       y = "Probability")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

.center[
$P(X \geq 2)=$ area under the curve to the right of 2
]
]

---

class: inverse, center, middle
# The Normal Distribution

---

# The Normal Distribution I

.pull-left[

- The .shout[Gaussian] or .shout[normal distribution] is the most useful type of probability distribution

$$ X \sim N(\mu,\sigma)$$

- Continuous, symmetric, unimodal, with mean $\mu$ and standard deviation $\sigma$

]

.pull-right[

```{r, fig.retina=3}
ggplot(data = tibble(x=-4:4))+
  aes(x = x)+
  stat_function(fun = dnorm, size=2, color="blue")+
  labs(x = "X",
       y = "Probability")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

]

---

# The Normal Distribution II

.pull-left[

- FYI: The pdf of $X \sim N(\mu, \sigma)$ is 
$$P(X=k)=	\frac{1}{\sqrt{2\pi \sigma^2}}e^{-\frac{1}{2}\big(\frac{(k-\mu)}{\sigma}\big)^2}$$

- **Do not try and learn this**, we have software and (previously tables) to calculate pdfs and cdfs

]

.pull-right[

```{r, fig.retina=3}
normal_pdf<-ggplot(data = tibble(x=-4:4))+
  aes(x = x)+
  stat_function(fun = dnorm, size=2, color="blue")+
  labs(x = "X",
       y = "Probability")+
  scale_x_continuous(breaks=seq(-4,4,1))+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
normal_pdf
```

]

---

# The 68-95-99.7 Rule

.pull-left[

- .shout[68-95-99.7% empirical rule]: for a normal distribution:
]

.pull-right[

```{r}
shade68 <- function(x) {
    y <- dnorm(x, mean = 0, sd = 1)
    y[x < -1 | x > 1] <- NA
    return(y)
}

shade95 <- function(x) {
    y <- dnorm(x, mean = 0, sd = 1)
    y[x < -2 | x > 2] <- NA
    return(y)
}

shade99.7 <- function(x) {
    y <- dnorm(x, mean = 0, sd = 1)
    y[x < -3 | x > 3] <- NA
    return(y)
}

```

```{r, fig.retina=3}
normal_pdf
```
]

---

# The 68-95-99.7 Rule

.pull-left[

- .shout[68-95-99.7% empirical rule]: for a normal distribution:

- $P(\mu-1\sigma \leq X \leq \mu+1\sigma) \approx$ 68\%

]

.pull-right[

```{r, fig.retina=3}
normal_pdf + 
  stat_function(fun = shade68, geom="area", fill="#8856a7", alpha=0.6)+
  annotate("segment", x = -1, xend = 1, y = 0.15, yend = 0.15, colour = "#000000", size=0.5, linetype = 2, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))+
  annotate("text", x = 0, y = 0.16, label = "68%", color="#000000")
```
]

---

# The 68-95-99.7 Rule

.pull-left[

- .shout[68-95-99.7% empirical rule]: for a normal distribution:

- $P(\mu-1\sigma \leq X \leq \mu+1\sigma) \approx$ 68\%

- $P(\mu-2\sigma \leq X \leq \mu+2\sigma) \approx$ 95\%

]

.pull-right[

```{r, fig.retina=3}
normal_pdf + 
  stat_function(fun = shade99.7, geom="area", fill="#e0ecf4", alpha=0.6)+
  stat_function(fun = shade95, geom="area", fill="#9ebcda", alpha=0.6)+
  stat_function(fun = shade68, geom="area", fill="#8856a7", alpha=0.6)+
  annotate("text", x = 0, y = 0.11, label = "95%", color="#000000")+
  annotate("segment", x = -1, xend = 1, y = 0.15, yend = 0.15, colour = "#000000", size=0.5, linetype = 2, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))+
  annotate("segment", x = -2, xend = 2, y = 0.10, yend = 0.10, colour = "#000000", size=0.5, linetype = 2, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))+
  annotate("text", x = 0, y = 0.16, label = "68%", color="#000000")
```
]

---
# The 68-95-99.7 Rule

.pull-left[

- .shout[68-95-99.7% empirical rule]: for a normal distribution:

- $P(\mu-1\sigma \leq X \leq \mu+1\sigma) \approx$ 68\%

- $P(\mu-2\sigma \leq X \leq \mu+2\sigma) \approx$ 95\%

- $P(\mu-3\sigma \leq X \leq \mu+3\sigma) \approx$ 99.7\%
]

.pull-right[

```{r, fig.retina=3}
normal_pdf + 
  stat_function(fun = shade99.7, geom="area", fill="#e0ecf4", alpha=0.6)+
  stat_function(fun = shade95, geom="area", fill="#9ebcda", alpha=0.6)+
  stat_function(fun = shade68, geom="area", fill="#8856a7", alpha=0.6)+
  annotate("segment", x = -3, xend = 3, y = 0.05, yend = 0.05, colour = "#000000", size=0.5, linetype = 2, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))+
  annotate("text", x = 0, y = 0.06, label = "99.7%", color="#000000")+
  annotate("segment", x = -2, xend = 2, y = 0.10, yend = 0.10, colour = "#000000", size=0.5, linetype = 2, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))+
  annotate("text", x = 0, y = 0.11, label = "95%", color="#000000")+
  annotate("segment", x = -1, xend = 1, y = 0.15, yend = 0.15, colour = "#000000", size=0.5, linetype = 2, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))+
  annotate("text", x = 0, y = 0.16, label = "68%", color="#000000")
```
]

---

# The Standard Normal Distribution

.pull-left[

- The .onfire[standard] normal distribution has mean 0 and standard deviation 1

$$Z \sim N(0,1)$$

]

.pull-right[

```{r}
normal_pdf
```
]

---

# The Standard Normal cdf

.pull-left[

- The .onfire[standard] normal cdf

$$\Phi(k)=P(Z \leq k)$$

]

.pull-right[

```{r, fig.retina=3}
ggplot(data = tibble(x=-4:4))+
  aes(x = x)+
  stat_function(fun = pnorm, size=2, color="blue")+
  labs(x = "X",
       y = "Probability")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

]

---

# Standardizing Variables: Example I

.content-box-green[
.green[**Example**]: On August 8, 2011, the Dow dropped 634.8 points, sending shock waves through the financial community. Assume that during mid-2011 to mid-2012 the daily change for the Dow is normally distributed, with the mean daily change of 1.87 points and a standard deviation of 155.28 points. What is the $Z$-score?

]

--

$$Z = \frac{X - \mu}{\sigma}$$

--

$$Z = \frac{634.8-1.87}{155.28}$$

--

$$Z = -4.1$$

This is 4.1 standard deviations $(\sigma)$ beneath the mean, an *extremely* low probability event.

---

# Standardizing Variables: From X to Z I

.content-box-green[
.green[**Example**]: In the last quarter of 2015, a group of 64 mutual funds had a mean return of 2.4% with a standard deviation of 5.6%. These returns can be approximated by a normal distribution.

What percent of the funds would you expect to be earning between -3.2% and 8.0% returns?

]

--

Convert to standard normal to find $Z$-scores for 8 and $-3.2.$

$$P(-3.2 < X < 8)$$

--

$$P(\frac{-3.2-2.4}{5.6} < \frac{X-2.4}{5.6} < \frac{8-2.4}{5.6})$$

--

$$P(-1 < Z < 1)$$

--

$$P(X \pm 1\sigma)=0.68$$

---

# Standardizing Variables: From X to Z II

```{r, fig.height=3, fig.retina=3, fig.align="center"}

shade_neg3.2_to_8 <- function(x) {
    y <- dnorm(x, mean = 2.4, sd = 5.6)
    y[x < -3.2 | x >8] <- NA
    return(y)
}

ggplot(data = tibble(x=seq(-20,24.8,5.6)))+
  aes(x = x)+
  stat_function(fun = dnorm, args=list(mean = 2.4, sd = 5.6), size=2, color="blue")+
  stat_function(fun=shade_neg3.2_to_8, geom="area", fill="blue", alpha=0.5)+
  labs(x = "Mutual Funds Return % (X)",
       y = "Probability")+
  scale_x_continuous(breaks=seq(-20,24.8,5.6))+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

```{r, fig.height=3, fig.retina=3, fig.align="center"}

shade_neg1_to_1 <- function(x) {
    y <- dnorm(x, mean = 0, sd = 1)
    y[x < -1 | x >1] <- NA
    return(y)
}

ggplot(data = tibble(x=seq(-4,4,1)))+
  aes(x = x)+
  stat_function(fun = dnorm, size=2, color="green")+
  stat_function(fun=shade_neg1_to_1, geom="area", fill="green", alpha=0.5)+
  labs(x = "Standard Deviations (Z)",
       y = "Probability")+
  scale_x_continuous(breaks=seq(-4,4,1))+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

---

# Standardizing Variables: From X to Z III

.content-box-blue[
.blue[**You Try**]: In the last quarter of 2015, a group of 64 mutual funds had a mean return of 2.4% with a standard deviation of 5.6%. These returns can be approximated by a normal distribution.

1. What percent of the funds would you expect to be earning between -3.2% and 8.0% returns?

2. What percent of the funds would you expect to be earning 2.4% or less?

3. What percent of the funds would you expect to be earning between -8.8% and 13.6%?

4. What percent of the funds would you expect to be earning returns greater than 13.6%?
]

---

# Finding Z-score Probabilities I

- How do we actually find the probabilities for $Z-$scores?

--

.center[
![](https://www.dropbox.com/s/r1k1x5117rt1mc8/ztable.jpg?raw=1)
]

---

# Finding Z-score Probabilities II

.pull-left[

Probability to the **left** of $z_i$

$$P(Z \leq z_i)= \underbrace{\Phi(z_i)}_{\text{cdf of }z_i}$$

```{r, fig.retina=3, fig.align="center"}
shade_neg4_to_1 <- function(x) {
    y <- dnorm(x, mean = 0, sd = 1)
    y[x < -4 | x >1] <- NA
    return(y)
}

ggplot(data = tibble(x=-4:4))+
  aes(x = x)+
  stat_function(fun = dnorm, size=2, color="blue")+
  stat_function(fun=shade_neg4_to_1, geom="area", fill="blue", alpha=0.5)+
  labs(x = "X",
       y = "Probability")+
  scale_x_continuous(breaks=c(1),
                      label=c(expression(z[i])))+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)

```

]
.pull-right[

Probability to the **right** of $z_i$

$$P(Z \geq z_i)= 1-\underbrace{\Phi(z_i)}_{\text{cdf of }z_i}$$

```{r, fig.retina=3, fig.align="center"}
shade_1_to_4 <- function(x) {
    y <- dnorm(x, mean = 0, sd = 1)
    y[x < 1 | x >4] <- NA
    return(y)
}

ggplot(data = tibble(x=-4:4))+
  aes(x = x)+
  stat_function(fun = dnorm, size=2, color="blue")+
  stat_function(fun=shade_1_to_4, geom="area", fill="blue", alpha=0.5)+
  labs(x = "X",
       y = "Probability")+
  scale_x_continuous(breaks=c(1),
                      label=c(expression(z[i])))+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)

```

]

---

# Finding Z-score Probabilities III

Probability **between** $z_1$ and $z_2$

$$P(z_1 \geq Z \geq z_2)= \underbrace{\Phi(z_2)}_{\text{cdf of }z_2} - \underbrace{\Phi(z_1)}_{\text{cdf of }z_1}$$

```{r, fig.retina=3, fig.align="center"}
shade_neg1_to_1 <- function(x) {
    y <- dnorm(x, mean = 0, sd = 1)
    y[x < -1 | x >1] <- NA
    return(y)
}

ggplot(data = tibble(x=-4:4))+
  aes(x = x)+
  stat_function(fun = dnorm, size=2, color="blue")+
  stat_function(fun=shade_neg1_to_1, geom="area", fill="blue", alpha=0.5)+
  labs(x = "X",
       y = "Probability")+
  scale_x_continuous(breaks=c(-1,1),
                      label=c(expression(z[1]), expression(z[1])))+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)

```

---

# Finding Z-score Probabilities IV

.pull-left[

- `pnorm()` calculates `p`robabilities with a `norm`al distribution with arguments:
  - `mean = ` the mean
  - `sd = ` the standard deviation
  - `lower.tail = ` 
      - `TRUE` if looking at area to *LEFT* of value
      - `FALSE` if looking at area to *RIGHT* of value

]

.pull-right[
```{r, fig.retina=3, fig.align="center"}
normal_pdf
```

]

---

# Finding Z-score Probabilities IV

.pull-left[

.content-box-green[
.green[**Example**:] Let the distribution of grades be normal, with mean 75 and standard deviation 10.

]

- Probability a student gets **at least an 80**

```{r, echo=T}
pnorm(80, 
      mean = 75,
      sd = 10,
      lower.tail = FALSE) # looking to right
```
]

.pull-right[
```{r, fig.retina=3, fig.align="center"}
shade_80_up <- function(x) {
    y <- dnorm(x, mean = 75, sd = 10)
    y[x < 80] <- NA
    return(y)
}

ggplot(data = tibble(x=35:115))+
  aes(x = x)+
  stat_function(fun = dnorm, args=list(mean = 75, sd = 10), size=2, color="blue")+
  stat_function(fun=shade_80_up, geom="area", fill="blue", alpha=0.5)+
  labs(x = "X",
       y = "Probability")+
  scale_x_continuous(breaks = seq(35,115,5))+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)

```

]

---

# Finding Z-score Probabilities V

.pull-left[

.content-box-green[
.green[**Example**:] Let the distribution of grades be normal, with mean 75 and standard deviation 10.

]

- Probability a student gets **at most an 80**

```{r, echo=T}
pnorm(80, 
      mean = 75,
      sd = 10,
      lower.tail = TRUE) # looking to left
```
]

.pull-right[
```{r, fig.retina=3, fig.align="center"}
shade_up_to_80 <- function(x) {
    y <- dnorm(x, mean = 75, sd = 10)
    y[x > 80] <- NA
    return(y)
}

ggplot(data = tibble(x=35:115))+
  aes(x = x)+
  stat_function(fun = dnorm, args=list(mean = 75, sd = 10), size=2, color="blue")+
  stat_function(fun=shade_up_to_80, geom="area", fill="blue", alpha=0.5)+
  labs(x = "X",
       y = "Probability")+
  scale_x_continuous(breaks = seq(35,115,5))+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)

```

]

---

# Finding Z-score Probabilities VI

.pull-left[

.content-box-green[
.green[**Example**:] Let the distribution of grades be normal, with mean 75 and standard deviation 10.

]

- Probability a student gets **between a 65 and 85**

```{r, echo=T}
# subtract two left tails!
pnorm(85, # larger number first!
      mean = 75,
      sd = 10,
      lower.tail = TRUE)- # looking to left
  pnorm(65, # smaller number second!
        mean = 75,
        sd = 10,
        lower.tail = TRUE) #looking to left
```
]

.pull-right[
```{r, fig.retina=3, fig.align="center"}
shade_65_to_85 <- function(x) {
    y <- dnorm(x, mean = 75, sd = 10)
    y[x < 65 | x > 85] <- NA
    return(y)
}

ggplot(data = tibble(x=35:115))+
  aes(x = x)+
  stat_function(fun = dnorm, args=list(mean = 75, sd = 10), size=2, color="blue")+
  stat_function(fun=shade_65_to_85, geom="area", fill="blue", alpha=0.5)+
  labs(x = "X",
       y = "Probability")+
  scale_x_continuous(breaks = seq(35,115,5))+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)

```

]

---

### 2019-09-19 static/slides/08-slides.Rmd
Fix slides 8

@@
 class: inverse, center, middle
-# Random Variables
+# Discrete Random Variables
 
@@
 
-$$\begin{align*}\sigma^2_X &= E[(X-\mu_X)^2]\\
+$$\begin{align*}\sigma^2_X &= E[(x_i-\mu_X)^2]\\
 &=\sum_{i=1}^n(x_i-\mu_X)^2p_i\\ \end{align*}$$
@@
 
+.content-box-green[
+.green[**Example**]: P(0 \leq X \leq 2)
+]
+
 - See [today's class notes](/class/08-class) for how to graph math/stats functions in `ggplot`!
@@
 
-.center[
-$P(0 \leq X \leq 2)=$ area under the curve between 0 and 2
-]
 ]
@@
 
+.content-box-green[
+.green[**Example**]: P(X \leq 2)
+]
 ]
@@
 
-.center[
-$P(X \leq 2)=$ area under the curve to the left of 2
-]
 ]
@@
 
-$$P(X \leq k)$$
-
+.content-box-green[
+.green[**Example**]: P(X \geq 2) = 1 - P(X \leq 2)
+]
 ]
@@
   stat_function(fun = dnorm, size=2, color="blue")+
+  scale_x_continuous(breaks=seq(-4,4,1))+
   labs(x = "X",
@@
 
-- $P(\mu-1\sigma \leq X \leq \mu+1\sigma) \approx$ 68\%
+- $P(\mu-1\sigma \leq X \leq \mu+1\sigma) \approx$ 68%
 
@@
 
-- $P(\mu-1\sigma \leq X \leq \mu+1\sigma) \approx$ 68\%
-
-- $P(\mu-2\sigma \leq X \leq \mu+2\sigma) \approx$ 95\%
+- $P(\mu-1\sigma \leq X \leq \mu+1\sigma) \approx$ 68%
+
+- $P(\mu-2\sigma \leq X \leq \mu+2\sigma) \approx$ 95%
 
@@
 
-- $P(\mu-1\sigma \leq X \leq \mu+1\sigma) \approx$ 68\%
-
-- $P(\mu-2\sigma \leq X \leq \mu+2\sigma) \approx$ 95\%
-
-- $P(\mu-3\sigma \leq X \leq \mu+3\sigma) \approx$ 99.7\%
+- $P(\mu-1\sigma \leq X \leq \mu+1\sigma) \approx$ 68%
+
+- $P(\mu-2\sigma \leq X \leq \mu+2\sigma) \approx$ 95%
+
+- $P(\mu-3\sigma \leq X \leq \mu+3\sigma) \approx$ 99.7%
+
+- **68/95/99.7%** of observations fall within **1/2/3 standard deviations** of the mean
 ]
@@
 
-- The .onfire[standard] normal distribution has mean 0 and standard deviation 1
+- The .onfire[standard] normal distribution (often referred to as **Z**) has mean 0 and standard deviation 1
 
@@
 
-# Standardizing Variables: Example I
+# Standardizing Variables:
+
+.pull-left[
+
+- We can take any normal distribution (for any $\mu, \sigma)$ and **standardize** it to the standard normal distribution by taking the .shout[Z-score] of any value, $x_i$:
+
+$$Z=\frac{x_i-\mu}{\sigma}$$
+
+- Subtract any value by the distribution's mean and divide by standard deviation
+
+- $Z$: number of standard deviations $x_i$ value is away from the mean
+]
+
+.pull-right[
+```{r, fig.retina=3}
+normal_pdf
+```
+]
+
+# Standardizing Variables: Example II
 
@@
 
-```{r, fig.retina=3, fig.align="center"}
+```{r, fig.retina=3, fig.height=5, fig.align="center"}
 shade_neg4_to_1 <- function(x) {
@@
 
-```{r, fig.retina=3, fig.align="center"}
+```{r, fig.retina=3, fig.height=5, fig.align="center"}
 shade_1_to_4 <- function(x) {
@@
 
-```{r, fig.retina=3, fig.align="center"}
+```{r, fig.retina=3, fig.height=5, fig.align="center"}
 shade_neg1_to_1 <- function(x) {
@@
       sd = 10,
-      lower.tail = TRUE)- # looking to left
+      lower.tail = TRUE) - # looking to left, & SUBTRACT
   pnorm(65, # smaller number second!
@@
 ]
-
----

### 2019-09-20 content/assignment/02-problem-set.Rmd
Added problem set 2

@@
 ---
-title: "Problem Set 1"
+title: "Problem Set 2"
 editor_options: 
@@
 
-*Due by Thursday, September 19, 2019*
+*Due by Thursday, September 26, 2019*
 
-- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/01-problem-set-pdf.pdf)
-- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/01-problem-set.Rmd)
+- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/02-problem-set-pdf.pdf)
+- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/02-problem-set.Rmd)
 
@@
 
-For this problem set, you may type up your answers to each question (save plots as images and include them with your answers) and also turn in a (commented!) *.R* file of commands for each relevant question.
-You may email or turn in a paper copy.
-It is not relevant for this particular problem set, but in the future, I will allow handwritten answers for math/theory questions.
+For this problem set, you may submit handwritten answers on a plain sheet of paper, or download and type/handwrite on the PDF.
 
 Alternatively, you may download the `.Rmd` file, do the homework in markdown, and email to me a single `knit`ted `html` or `pdf` file (and be sure that it shows all of your code).
-
-To minimize confusion, I suggest creating a new `R Project` (e.g.
-`hw1`) and storing any data and plots in that folder on your computer.
-See my [example workflow](https://github.com/ryansafner/workflow).
 
@@
 
-## 1.
+## Question 1
 
-### a.
+### Part A 
+In your own words, explain the difference between endogeneity and exogeneity.^[Hint: you may want to look back at class 1.2 for this question.]
+
+### Part B
+In your own words, explain how conducting a randomized controlled trial helps to identify the causal connection between two variables.
+
+## Question 2
+
+### Part A
 
@@
 
-### b.
+### Part B
 
-In your own words, explain how (sample) standard deviation is calculated. You may also write the formula, but it is not necessary.
+In your own words, explain how (sample) standard deviation *is calculated.* You may also write the formula, but it is not necessary.
+
+# Problems
+
+For the remaining questions, you may use `R` to *verify*, but please calculate all sample statistics by hand and show all work.
+
+## Question 3
+
+Suppose you have a very small class of four students that all take a quiz. Their scores are reported as follows:
+
+$$\{83, 92, 72, 81\}$$
+
+## Part A 
+Calculate the median.
+
+## Part B
+Calculate the sample mean, $\bar{x}$.
+
+## Part C
+Calculate the sample standard deviation, $s$.
+
+## Part D
+Make or sketch a rough histogram of this data, with the size of each bin being 10 (i.e.
+70's, 80's, 90's, 100's).
+You can draw this by hand or use `R`.^[If you are using `ggplot`, you want to use `+geom_histogram(breaks=seq(start,end,by))` and add `+scale_x_continuous(breaks=seq(start,end,by))`.
+For each, it creates bins in the histogram, and ticks on the x axis by creating a `seq`uence starting at `start` (a number), ending at `end` (number), `by` a certain interval (i.e.
+by `10`s.).] Is this distribution roughly symmetric or skewed?
+What would we expect about the mean and the median?
+
+## Part E
+Suppose instead the person who got the 72 did not show up that day to class, and got a 0 instead. Recalculate the mean and median. What happened and why?
+
+## Question 4
+
+Suppose the probabilities of a visitor to Amazon’s website buying 0, 1, or 2 books are 0.2, 0.4, and 0.4 respectively.
+
+### Part A
+Calculate the *expected number* of books a visitor will purchase.
+
+### Part B 
+Calculate the *standard deviation* of book purchases.
+
+### Part C 
+**Bonus**: try doing this in `R` by making an initial dataframe of the data, and then making new columns to the "table" like we did in class.
+
+## Question 5
+
+Scores on the SAT (out of 1600) are approximately normally distributed with a mean of 500 and standard deviation of 100.
+
+### Part A
+
+What is the probability of getting a score between a 400 and a 600?
+
+### Part B
+
+What is the probability of getting a score between a 300 and a 700?
+
+### Part C
+
+What is the probability of getting *at least* a 700?
+
+### Part D
+
+What is the probability of getting *at most* a 700?
+
+### Part E
+
+What is the probability of getting exactly a 500?
+
+## Question 6
+
+Redo problem 5 by using the `pnorm()` command in `R`.^[Hint: This function has four arguments: 1.
+the value of the random variable, 2.
+the mean of the distribution, 3.
+the sd of the distribution, and 4.
+`lower.tail` `TRUE` or `FALSE`.]

### 2019-09-20 static/homeworks/02-problem-set-pdf.Rmd
Added problem set 2

--- 2019-09-20 content/assignment/02-problem-set.Rmd
@@
 title: "Problem Set 2"
-editor_options: 
-  chunk_output_type: console
+author: "Ryan Safner"
+date: "ECON 480 - Fall 2019"
+output: pdf_document
 ---
 
-```{r message=FALSE, warning=FALSE, include=FALSE}
-library(tidyverse)
-library(pander)
-source(file.path(here::here(), "lib", "graphics.R"))
-```
-
 *Due by Thursday, September 26, 2019*
-
-- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/02-problem-set-pdf.pdf)
-- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/02-problem-set.Rmd)
 
@@
 
-## Question 1
+1.
 
-### Part A 
-In your own words, explain the difference between endogeneity and exogeneity.^[Hint: you may want to look back at class 1.2 for this question.]
+a. In your own words, explain the difference between endogeneity and exogeneity.^[Hint: you may want to look back at class 1.2 for this question.]
+b. In your own words, explain how conducting a randomized controlled trial helps to identify the causal connection between two variables.
 
-### Part B
-In your own words, explain how conducting a randomized controlled trial helps to identify the causal connection between two variables.
+\clearpage
 
-## Question 2
+2.
 
-### Part A
+a. In your own words, explain what (sample) standard deviation *means*.
+b. In your own words, explain how (sample) standard deviation *is calculated.* You may also write the formula, but it is not necessary.
 
-In your own words, explain what (sample) standard deviation *means*.
-
-### Part B
-
-In your own words, explain how (sample) standard deviation *is calculated.* You may also write the formula, but it is not necessary.
+\clearpage
 
@@
 
-## Question 3
-
-Suppose you have a very small class of four students that all take a quiz. Their scores are reported as follows:
+3. Suppose you have a very small class of four students that all take a quiz. Their scores are reported as follows:
 
@@
 
-## Part A 
-Calculate the median.
-
-## Part B
-Calculate the sample mean, $\bar{x}$.
-
-## Part C
-Calculate the sample standard deviation, $s$.
-
-## Part D
+a. Calculate the median.
+b. Calculate the sample mean, $\bar{x}$.
+c. Calculate the sample standard deviation, $s$.
+d.
 Make or sketch a rough histogram of this data, with the size of each bin being 10 (i.e.
@@
 What would we expect about the mean and the median?
+e. Suppose instead the person who got the 72 did not show up that day to class, and got a 0 instead. Recalculate the mean and median. What happened and why?
 
-## Part E
-Suppose instead the person who got the 72 did not show up that day to class, and got a 0 instead. Recalculate the mean and median. What happened and why?
+\clearpage
 
-## Question 4
+4. Suppose the probabilities of a visitor to Amazon’s website buying 0, 1, or 2 books are 0.2, 0.4, and 0.4 respectively.
 
-Suppose the probabilities of a visitor to Amazon’s website buying 0, 1, or 2 books are 0.2, 0.4, and 0.4 respectively.
+a. Calculate the *expected number* of books a visitor will purchase.
+b. Calculate the *standard deviation* of book purchases.
+c. **Bonus**: try doing this in `R` by making an initial dataframe of the data, and then making new columns to the "table" like we did in class.
 
-### Part A
-Calculate the *expected number* of books a visitor will purchase.
+\clearpage
 
-### Part B 
-Calculate the *standard deviation* of book purchases.
+5. Scores on the SAT (out of 1600) are approximately normally distributed with a mean of 500 and standard deviation of 100.
 
-### Part C 
-**Bonus**: try doing this in `R` by making an initial dataframe of the data, and then making new columns to the "table" like we did in class.
+a. What is the probability of getting a score between a 400 and a 600?
+b. What is the probability of getting a score between a 300 and a 700?
+c. What is the probability of getting *at least* a 700?
+d. What is the probability of getting *at most* a 700?
+e. What is the probability of getting exactly a 500?
 
-## Question 5
+\clearpage
 
-Scores on the SAT (out of 1600) are approximately normally distributed with a mean of 500 and standard deviation of 100.
-
-### Part A
-
-What is the probability of getting a score between a 400 and a 600?
-
-### Part B
-
-What is the probability of getting a score between a 300 and a 700?
-
-### Part C
-
-What is the probability of getting *at least* a 700?
-
-### Part D
-
-What is the probability of getting *at most* a 700?
-
-### Part E
-
-What is the probability of getting exactly a 500?
-
-## Question 6
-
+6.
 Redo problem 5 by using the `pnorm()` command in `R`.^[Hint: This function has four arguments: 1.

### 2019-09-20 static/homeworks/02-problem-set.Rmd
Added problem set 2

---
title: "Problem Set 2"
author: "YOUR NAME HERE" # put your name here! 
date: "ECON 480 - Fall 2019"
output: html_document # change to pdf if you'd like, this will make a webpage (you can email it and open in any browser)
---

<!--CLICK "KNIT" ABOVE TO RENDER TO HTML, PDF, OR WORD OUTPUT
In fact, try knitting right away and see what this produces!
-->

*Due by Thursday, September 26, 2019*

# Instructions

For this problem set, you may submit handwritten answers on a plain sheet of paper, or download and type/handwrite on the PDF.

Alternatively, you may download the `.Rmd` file, do the homework in markdown, and email to me a single `knit`ted `html` or `pdf` file (and be sure that it shows all of your code).

You may work together (and I highly encourage that) but you must turn in your own answers. I grade homeworks 70% for completion, and for the remaining 30%, pick one question to grade for accuracy - so it is best that you try every problem, even if you are unsure how to complete it accurately.

# Theory and Concepts

Install and load the package `babynames`. Get help for `?babynames` to see what the data includes.

```{r}
# write your code here! 
```

## Question 1
### Part A
**In your own words, explain the difference between endogeneity and exogeneity.**^[Hint: you may want to look back at class 1.2 for this question.]

<!--WRITE YOUR ANSWERS BELOW -->

### Part B
**In your own words, explain how conducting a randomized controlled trial helps to identify the causal connection between two variables.**

<!--WRITE YOUR ANSWERS BELOW -->

## Question 2

### Part A

**In your own words, explain what (sample) standard deviation _means_.**

<!--WRITE YOUR ANSWERS BELOW -->

### Part B

**In your own words, explain how (sample) standard deviation _is calculated._ You may also write the formula, but it is not necessary.**

<!--WRITE YOUR ANSWERS BELOW -->

# Problems

**For the remaining questions, you may use `R` to *verify*, but please calculate all sample statistics by hand and show all work.**

## Question 3

**Suppose you have a very small class of four students that all take a quiz. Their scores are reported as follows:**

$$\{83, 92, 72, 81\}$$

## Part A 
**Calculate the median.**

<!--WRITE YOUR ANSWERS BELOW -->

## Part B 
**Calculate the sample mean, $\bar{x}$**

<!--WRITE YOUR ANSWERS BELOW -->

## Part C 
**Calculate the sample standard deviation, $s$**

<!--WRITE YOUR ANSWERS BELOW -->

## Part D
**Make or sketch a rough histogram of this data, with the size of each bin being 10 (i.e. 70's, 80's, 90's, 100's). You can draw this by hand or use `R`.**^[If you are using `ggplot`, you want to use `+geom_histogram(breaks=seq(start,end,by))` and add `+scale_x_continuous(breaks=seq(start,end,by))`. For each, it creates bins in the histogram, and ticks on the x axis by creating a `seq`uence starting at `start` (a number), ending at `end` (number), `by` a certain interval (i.e. by `10`s.).] **Is this distribution roughly symmetric or skewed? What would we expect about the mean and the median?**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 5-c}
# write your code here
```

## Part E

**Suppose instead the person who got the 72 did not show up that day to class, and got a 0 instead. Recalculate the mean and median. What happened and why?**

<!--WRITE YOUR ANSWERS BELOW -->

## Question 4

Suppose the probabilities of a visitor to Amazon’s website buying 0, 1, or 2 books are 0.2, 0.4, and 0.4 respectively.

### Part A
**Calculate the _expected number_ of books a visitor will purchase.**

<!--WRITE YOUR ANSWERS BELOW -->

### Part B
**Calculate the _standard deviation_ of book purchases.**

<!--WRITE YOUR ANSWERS BELOW -->

### Part C 
**BONUS: try doing this in `R` by making an initial dataframe of the data, and then making new columns to the "table" like we did in class.**

<!--WRITE YOUR ANSWERS BELOW -->

```{r 4-c}
# write your code here
```

## Question 5

Scores on the SAT (out of 1600) are approximately normally distributed with a mean of 500 and standard deviation of 100.

### Part A
**What is the probability of getting a score between a 400 and a 600?**

<!--WRITE YOUR ANSWERS BELOW -->

### Part B
**What is the probability of getting a score between a 300 and a 700?**

<!--WRITE YOUR ANSWERS BELOW -->

### Part C
**What is the probability of getting _at least_ a 700?**

<!--WRITE YOUR ANSWERS BELOW -->

### Part D
**What is the probability of getting _at most_ a 700?**

<!--WRITE YOUR ANSWERS BELOW -->

### Part E
**What is the probability of getting _exactly_ a 500?**

<!--WRITE YOUR ANSWERS BELOW -->

## Question 6

**Redo problem 5 by using the `pnorm()` command in `R`.**^[Hint: This function has four arguments: 1. the value of the random variable, 2. the mean of the distribution, 3. the sd of the distribution, and 4. `lower.tail` `TRUE` or `FALSE`.]

<!--WRITE YOUR ANSWERS BELOW -->

```{r 6}
# write your code here
```

### 2019-09-20 content/assignment/02-problem-set.Rmd
Minor formatting tweak to problem set 2

@@
 
-## Part A 
+### Part A 
 Calculate the median.
 
-## Part B
+### Part B
 Calculate the sample mean, $\bar{x}$.
 
-## Part C
+### Part C
 Calculate the sample standard deviation, $s$.
 
-## Part D
+### Part D
 Make or sketch a rough histogram of this data, with the size of each bin being 10 (i.e.
@@
 
-## Part E
+### Part E
 Suppose instead the person who got the 72 did not show up that day to class, and got a 0 instead. Recalculate the mean and median. What happened and why?

### 2019-09-21 content/class/08-class.Rmd
Added problem set 2
Update class 8 with appendix for stat graphs and functions

@@
 
-Problem set 2 (on classes 2.1-2.2) will be posted shortly, and is due by Thursday September 26.
+[Problem Set 2](/assignment/01-problem-set.html) (on classes 2.1-2.2) is due by Thursday September 26.
 
@@
 
-Check back later today!
+The `mosaic` package is useful for making and using mathematical functions in `R`.
+
+```{r}
+# install.packages("mosaic") install for first use
+library(mosaic) # load 
+```
+
+You can create custom mathematical functions using mosaic by defining an R `function()` with multiple arguments. As a simple example, make the function `f(x) = 2x+x^2` as follows:
+
+```{r}
+# store as a named function, I'll call it "my_function"
+my_function<-function(x){10*x-x^2}
+
+# look at it
+my_function
+
+```
+
+There are some notational requirements from `R` for making functions. Any coefficient in front of a variable (such as the 2 in `2x` must be explicitly multiplied by the variable, as in `10*x`.)
+
+To use the function, simply define what the input `(x)` is and run your named function on it:
+
+```{r}
+# f of 2 
+my_function(2)
+
+# f of 2 and 4
+my_function(c(2,4))
+
+# f of 2 through 7
+my_function(2:7)
+
+# ALTERNATIVELY
+# define x first and then run function on it
+
+x<-c(2,4)
+my_function(x)
+```
+
+## To Graph
+
+In `ggplot` there is a dedicated `stat_function()` (equivalent to a `geom_` layer) to graph mathematical and statistical functions.
+All that is needed is a `data.frame` of a range of `x` values to act as the source for `data`, and set `x` equal to those values for `aes`thetics.
+
+```{r, fig.retina=3}
+# x values are integers 1 through 10
+ggplot(data = data.frame(x = 1:10))+
+  aes(x = x)
+```
+
+Then we add the `stat_function`, where `fun =` is the most important argument where you define the to function to graph as your function created above, for example, our `my_function`.
+
+```{r, fig.retina=3}
+ggplot(data = data.frame(x = 1:10))+
+  aes(x = x)+
+  stat_function(fun = my_function) #<<
+```
+
+You can also adjust things like size, color, and line type.
+
+```{r, fig.retina=3}
+ggplot(data = data.frame(x = 1:10))+
+  aes(x = x)+
+  stat_function(fun = my_function, color = "blue", size = 2, linetype = "dashed") #<<
+```
+
+There are some standard statistical distributions built into R:
+
+**Prefixes**:
+
+| Action/Type | Prefix |
+|--------|--------|
+| random draw | `r` |
+| density (pdf) | `d` |  
+| cdf    | `p`    |
+| quantile (inverse cdf) | `q` |
+
+**Distributions**:
+
+| Distribution | Name |
+|--------|--------|
+| normal | `norm` |
+| uniform | `unif` |  
+| t    | `t`    |
+| binomial    | `binom`    |
+| negative binomial    | `nbinom`    |
+| hyper    | `hypergeometric`    |
+| weibull    | `weibull`    |
+| beta    | `beta`    |
+| gamma    | `gamma`    |
+
+Thus, what you want is a combination of the prefix and the distribution.
+
+### Some common examples:
+
+1. Take random draws from a normal distribution:
+
+```{r}
+rnorm(n = 10, # take 10 draws from a normal distribution with:
+      mean = 2, # mean of 2
+      sd = 1) # sd of 1
+```
+
+2. Get probability of a random variable being less than or equal to a value (cdf) from a normal distribution:
+
+```{r}
+# find probability of area to the LEFT of a number on pdf (note this = cdf of that number!)
+pnorm(q = 80, # number is 80 from a distribution where: 
+      mean = 200, # mean is 100
+      sd = 100, # sd is 100
+      lower.tail = TRUE) # looking to the LEFT in lower tail
+```
+
+3. Find the value of a distribution that is a specified percentile.
+
+```{r}
+# find the 38th percentile value
+qnorm(p = 0.38, # 38th percentile from a distribution where:
+      mean = 200, # mean is 200
+      sd = 100) # sd is 100
+```
+
+You can also graph these commonly used statistical functions by setting `fun = ` the named functions in your `stat_function()` layer.
+If you want to specify the mean and standard deviation, use `args = list()` to include the required arguments from the named function above (e.g.
+`dnorm` needs `mean` and `sd`).
+
+```{r, fig.retina=3}
+ggplot(data = data.frame(x = -400:600))+
+  aes(x = x)+
+  stat_function(fun = dnorm, args = list(mean = 200, sd = 200), color = "blue", size = 2, linetype = "dashed") #<<
+```
+
+If you don't include this, it will graph the standard distribution:
+
+```{r, fig.retina=3}
+ggplot(data = data.frame(x = -4:4))+
+  aes(x = x)+
+  stat_function(fun = dnorm, color = "blue", size = 2, linetype = "dashed") #<<
+```
+
+To add shading under a distribution, simply add a duplicate of the `stat_function()` layer, but add `geom="area"` to indicate the area beneath the function should be filled, and you can limit the domain of the `fill` with `xlim=c(start,end)`, where `start` and `end` are the x-values for the endpoints of the `fill.`
+
+```{r, fig.retina=3}
+# graph normal distribution and shade area between -2 and 2
+ggplot(data = data.frame(x = -4:4))+
+  aes(x = x)+
+  stat_function(fun = dnorm, color = "blue", size = 2, linetype = "dashed")+
+  stat_function(fun = dnorm, xlim = c(-2,2), geom = "area", fill = "green", alpha=0.5)#<<
+```
+
+Hence, here is one graph from my slides:
+
+```{r, fig.retina=3}
+
+ggplot(data = tibble(x=35:115))+
+  aes(x = x)+
+  stat_function(fun = dnorm, args = list(mean = 75, sd = 10), size = 2, color="blue")+
+  stat_function(fun = dnorm, args = list(mean = 75, sd = 10), geom = "area", xlim = c(65,85), fill="blue", alpha=0.5)+
+  labs(x = "X",
+       y = "Probability")+
+  scale_x_continuous(breaks = seq(35,115,5))+
+  theme_classic(base_family = "Fira Sans Condensed",
+           base_size=20)
+```

### 2019-09-21 content/class/08-class.Rmd
Clean up class notes appendix 8
More cleaning of class notes 8 appendix

@@
 
-```{r}
+```{r, message = F, warning = F}
 # install.packages("mosaic") install for first use
@@
 
-You can create custom mathematical functions using mosaic by defining an R `function()` with multiple arguments. As a simple example, make the function `f(x) = 2x+x^2` as follows:
+## Creating Mathematical Functions
+
+You can create custom mathematical functions using mosaic by defining an R `function()` with multiple arguments. As a simple example, make the function $f(x) = 10x-x^2$ as follows:
 
@@
 
-There are some notational requirements from `R` for making functions. Any coefficient in front of a variable (such as the 2 in `2x` must be explicitly multiplied by the variable, as in `10*x`.)
+There are some notational requirements from `R` for making functions. Any coefficient in front of a variable (such as the 10 in `10x` must be explicitly multiplied by the variable, as in `10*x`).
 
@@
 
-## To Graph
+## Graphing Mathematical Functions
 
@@
   aes(x = x)+
-  stat_function(fun = my_function) #<<
+  stat_function(fun = my_function) 
 ```
@@
   aes(x = x)+
-  stat_function(fun = my_function, color = "blue", size = 2, linetype = "dashed") #<<
+  stat_function(fun = my_function, color = "blue", size = 2, linetype = "dashed") 
 ```
 
-There are some standard statistical distributions built into R:
+## Bultin Statistical Functions
+
+There are some standard statistical distributions built into R. They require a combination of a specific prefix and a distribution.
 
@@
 | density (pdf) | `d` |  
-| cdf    | `p`    |
+| cumulative density (cdf)    | `p`    |
 | quantile (inverse cdf) | `q` |
@@
 
-| Distribution | Name |
+| Distribution | Name in R |
 |--------|--------|
-| normal | `norm` |
-| uniform | `unif` |  
-| t    | `t`    |
-| binomial    | `binom`    |
-| negative binomial    | `nbinom`    |
-| hyper    | `hypergeometric`    |
-| weibull    | `weibull`    |
-| beta    | `beta`    |
-| gamma    | `gamma`    |
+| Normal | `norm` |
+| Uniform | `unif` |  
+| Student's t    | `t`    |
+| Binomial    | `binom`    |
+| Negative binomial    | `nbinom`    |
+| Hypergeometric    | `hyper`    |
+| Weibull    | `weibull`    |
+| Beta    | `beta`    |
+| Gamma    | `gamma`    |
 
@@
 
-### Some common examples:
+## Some common examples:
 
@@
 
+## Graphing Statistical Functions
+
 You can also graph these commonly used statistical functions by setting `fun = ` the named functions in your `stat_function()` layer.
@@
   aes(x = x)+
-  stat_function(fun = dnorm, args = list(mean = 200, sd = 200), color = "blue", size = 2, linetype = "dashed") #<<
+  stat_function(fun = dnorm, args = list(mean = 200, sd = 200), color = "blue", size = 2, linetype = "dashed") 
 ```
@@
   aes(x = x)+
-  stat_function(fun = dnorm, color = "blue", size = 2, linetype = "dashed") #<<
+  stat_function(fun = dnorm, color = "blue", size = 2, linetype = "dashed") 
 ```
@@
   stat_function(fun = dnorm, color = "blue", size = 2, linetype = "dashed")+
-  stat_function(fun = dnorm, xlim = c(-2,2), geom = "area", fill = "green", alpha=0.5)#<<
+  stat_function(fun = dnorm, xlim = c(-2,2), geom = "area", fill = "green", alpha=0.5)
 ```

### 2019-09-24 content/class/09-class.Rmd
Add class 9 materials, class 10 reading

---
title: "2.3: OLS Linear Regression - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

*Tuesday, September 24, 2019*

# Overview

Today we start looking at *associations* between variables, which we will first attempt to quantify with measures like *covariance* and *correlation*. Then we turn to fitting a line to data via *linear regression*. We overview the basic regression model, the parameters and how they are derived, and see how to work with regressions in `R` with `lm` and the tidyverse package [`broom`](https://broom.tidyverse.org/).

We consider an extended example about class sizes and test scores, which comes from a (Stata) dataset from an old textbook that I used to use, [Stock and Watson, 2007](https://www.pearson.com/us/higher-education/product/Stock-Introduction-to-Econometrics-Update-3rd-Edition/9780133486872.html?tab=resources). Download and follow along with the data from today's example:^[Note this is a `.dta` Stata file. You will need to (install and) load the package `haven` to `read_dta()` Stata files into a dataframe.]

- [<i class="fas fa-table"></i> `caschool.dta`](http://metricsf19.classes.ryansafner.com/data/caschool.dta)

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/09-slides.html)

# Assignments: Problem Set 1 DUE and Problem Set 2

[Problem Set 1](/assignment/01-problem-set.html) answers will be posted shortly, and the graded assignment emailed back to you.^[Sorry I've been too busy to send emails confirming that I recieved your problem set!]

[Problem Set 2](/assignment/01-problem-set.html) (on classes 2.1-2.2) is due by Thursday September 26.

# Appendix: Covariance and Correlation

## Variance

Recall the variance of a random variable $X$, denoted $var(X)$ or $\sigma^2$, is the expected value (probability-weighted average) of the squared deviations of $X_i$ from it's mean (or expected value) $\bar{X}$ or $E(X)$.^[Note there will be a different in notation depending on whether we refer to a population (e.g. $\mu_{X}$) or to a sample (e.g. $\bar{X}$). As the overwhelming majority of cases we will deal with samples, I will use sample notation for means).]

$$\begin{align*}
\sigma_X^2&=E(X-E(X))\\
&=\sum^n_{i=1} (X_i-\bar{X})^2 p_i
\end{align*}$$

Note if all possible values of $X_i$ are equally likely (or we don't know the probabilities), we can write variance as a simple average of squared deviations from the mean:

$$\begin{align*}
\sigma_X^2&=\frac{1}{n}\sum^n_{i=1}(X_i-\bar{X})^2 	
\end{align*}$$

Variance has some useful properties:

**Property 1**: The variance of a constant is 0

$$var(c)=0 \text{ iff } P(X=c)=1$$

If a random variable takes the same value (e.g. 2) with probability 1.00, $E(2)=2$, so the average squared deviation from the mean is 0, because there are never any values other than 2.

**Property 2**: The variance is unchanged for a random variable plus/minus a constant

$$var(X\pm c)$$

Since the variance of a constant is 0.

**Property 3**: The variance of a scaled random variable is scaled by the square of the coefficient

$$var(aX)=a^2var(X)$$

**Property 4**: The variance of a linear transformation of a random variable is scaled by the square of the coefficient

$$var(aX+b)=a^2var(X)$$

## Covariance

For two random variables, $X$ and $Y$, we can measure their **covariance** (denoted $cov(X,Y)$ or $\sigma_{X,Y}$)^[Again, to be technically correct, $\sigma_{X,Y}$ refers to populations, $s_{X,Y}$ refers to samples, in line with population vs. sample variance and standard deviation. Recall also that sample estimates of variance and standard deviation divide by $n-1$, rather than $n$. In large sample sizes, this difference is negligible.] to quantify how they vary *together*. A good way to think about this is: when $X$ is above its mean, would we expect $Y$ to also be above its mean (and covary positively), or below its mean (and covary negatively). Remember, this is describing the *joint* probability distribution for two random variables.

$$\begin{align*}
\sigma_{X,Y}&=E\big[(X-\bar{X})(Y-\bar{Y})\big]
\end{align*}$$

Again, in the case of equally probable values for both $X$ and $Y$, covariance is sometimes written:

$$\begin{align*}
\sigma_{X,Y}&=\frac{1}{N}\sum_{i=1}^n(X-\bar{X})(Y-\bar{Y})
\end{align*}$$

Covariance also has a number of useful properties:

**Property 1**: The covariance of a random variable $X$ and a constant $c$ is 0

$$cov(X,c)=0$$

**Property 2**: The covariance of a random variable and itself is the variable's variance

$$\begin{align*}
	cov(X,X)&=var(X)\\
	\sigma_{X,X}&=\sigma^2_X\\
	\end{align*}$$

**Property 3**: The covariance of a two random variables $X$ and $Y$ each scaled by a constant $a$ and $b$ is the product of the covariance and the constants

$$cov(aX,bY)=a\times b \times cov(X,Y)$$

**Property 4**: If two random variables are independent, their covariance is 0

$$cov(X,Y)=0 \text{ iff } X \text{ and } Y \text{ are independent:}  E(XY)=E(X)\times E(Y)$$

## Correlation

Covariance, like variance, is often cumbersome, and the numerical value of the covariance of two random variables does not really mean much. It is often convenient to normalize the covariance to a decimal between $-1$ and 1. We do this by dividing by the product of the standard deviations of $X$ and $Y$. This is known as the **correlation coefficient** between $X$ and $Y$, denoted $corr(X,Y)$ or $\rho_{X,Y}$ (for populations) or $r_{X,Y}$ (for samples):

$$\begin{align*}	
r_{X,Y}&=\frac{cov(X,Y)}{sd(X)sd(Y)}\\
&=\frac{E\big[(X-\bar{X})(Y-\bar{Y})\big]}{\sqrt{E\big[X-\bar{X}\big]}\sqrt{E\big[Y-\bar{Y}\big]}}\\
&=\frac{\sigma_{X,Y}}{\sigma_X \sigma_Y}\\
\end{align*}$$

Note this also means that covariance is the product of the standard deviation of $X$ and $Y$ and their correlation coefficient:

$$\begin{align*}
\sigma_{X,Y}&=r_{X,Y}\sigma_X \sigma_Y	\\
cov(X,Y)&=corr(X,Y)\times sd(X) \times sd(Y)	\\
\end{align*}$$

Another way to reach the (sample) correlation coefficient is by finding the average joint $Z$-score of each pair of $(X_i,Y_i)$:

$$\begin{align*}	
r_{X,Y}&=\frac{1}{n}\frac{\displaystyle\sum^n_{i=1}(X_i-\bar{X})(Y_i-\bar{Y}))}{s_X s_Y} && \text{Definition of sample correlation}\\
&=\frac{1}{n}\displaystyle\sum^n_{i=1}\bigg(\frac{X_i-\bar{X}}{s_X}\bigg)\bigg(\frac{Y_i-\bar{Y}}{s_Y}\bigg) && \text{Breaking into separate sums} \\
&=\frac{1}{n}\displaystyle\sum^n_{i=1}(Z_X)(Z_Y) && \text{Recognize each sum is the z-score for that r.v.} \\
\end{align*}$$

Correlation has some useful properties that should be familiar to you:

- Correlation is between $-1$ and 1
- A correlation of -1 is a downward sloping straight line
- A correlation of 1 is an upward sloping straight line
- A correlation of 0 implies no relationship

# Calculating Correlation Example

We can calculate the correlation of a simple data set (of 4 observations) using `R` to show how correlation is calculated. We will use the $Z$-score method. Begin with a simple set of data in $(X_i, Y_i)$ points:

$$ (1,1), (2,2), (3,4), (4,9) $$

```{r, fig.retina=3, message = F, warning = F}
library(tidyverse)

corr_example<-tibble(x=c(1,2,3,4),
                         y=c(1,2,4,9))

ggplot(corr_example,aes(x=x,y=y))+geom_point()

corr_example %>%
  summarize(mean_x = mean(x), #find mean of x, its 2.5
            sd_x = sd(x), #find sd of x, its 1.291
            mean_y = mean(y), #find mean of y, its 4
            sd_y = sd(y)) #find sd of y, its 3.559

#take z score of x,y for each pair and multiply them

corr_example <- corr_example %>%
  mutate(z_product = ((x-mean(x))/sd(x)) * ((y-mean(y))/sd(y)))

corr_example %>%
  summarize(avg_z_product = sum(z_product)/(n()-1), # average z products over n-1
            actual_corr = cor(x,y), #compare our answer to actual cor() command!
            covariance = cov(x,y)) # just for kicks, what's the covariance? 
```

### 2019-09-24 static/homeworks/02-problem-set-answers.Rmd
Add class 9 materials, class 10 reading

--- 2019-09-20 static/homeworks/02-problem-set-pdf.Rmd
@@
 ---
-title: "Problem Set 2"
+title: "Problem Set 2 (Answers)"
 author: "Ryan Safner"
 date: "ECON 480 - Fall 2019"
-output: pdf_document
+output:
+  html_document:
+    df_print: paged
+    theme: simplex
+    toc: true 
+    toc_depth: 3
+    toc_float: true
+    code_folding: show
+    highlight: tango
 ---
 
-*Due by Thursday, September 26, 2019*
+```{r message=FALSE, warning=FALSE, include=FALSE}
+library(pander)
+source(file.path(here::here(), "lib", "graphics.R"))
+```
+
+*Due by Thursday, September 19, 2019*
+
+- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/01-problem-set.pdf)
+- [<i class="fab fa-markdown"></i> `04-r-practice-answers.Rmd`](/homeworks/01-r-problem-set.Rmd)
 
@@
 
-1.
+## 1.
 
-a. In your own words, explain the difference between endogeneity and exogeneity.^[Hint: you may want to look back at class 1.2 for this question.]
-b. In your own words, explain how conducting a randomized controlled trial helps to identify the causal connection between two variables.
+### a. 
+In your own words, explain the difference between endogeneity and exogeneity.^[Hint: you may want to look back at class 1.2 for this question.]
 
-\clearpage
+### b.
+In your own words, explain how conducting a randomized controlled trial helps to identify the causal connection between two variables.
 
-2.
+## 2.
 
-a. In your own words, explain what (sample) standard deviation *means*.
-b. In your own words, explain how (sample) standard deviation *is calculated.* You may also write the formula, but it is not necessary.
+### a.
 
-\clearpage
+In your own words, explain what (sample) standard deviation *means*.
+
+### b.
+
+In your own words, explain how (sample) standard deviation *is calculated.* You may also write the formula, but it is not necessary.
 
@@
 
-For the remaining questions, you may use `R` to *verify*, but please calculate all sample statistics by hand and show all work.
+## 3.
 
-3. Suppose you have a very small class of four students that all take a quiz. Their scores are reported as follows:
+Suppose you have a very small class of four students that all take a quiz. Their scores are reported as follows:
 
@@
 
-a. Calculate the median.
-b. Calculate the sample mean, $\bar{x}$.
-c. Calculate the sample standard deviation, $s$.
-d.
+For the remaining questions, you may use `R` to *verify*, but please calculate all sample statistics by hand and show all work.
+
+## a. 
+Calculate the median.
+```{r}
+median(c(83,92,72,81))
+```
+
+## b.
+Calculate the sample mean, $\bar{x}$.
+
+```{r}
+mean(c(83,92,72,81))
+```
+
+## c.
+Calculate the sample standard deviation, $s$.
+
+```{r}
+sd(c(83,92,72,81))
+```
+
+## d.
 Make or sketch a rough histogram of this data, with the size of each bin being 10 (i.e.
@@
 What would we expect about the mean and the median?
-e. Suppose instead the person who got the 72 did not show up that day to class, and got a 0 instead. Recalculate the mean and median. What happened and why?
 
-\clearpage
+```{r}
+df <- tibble(x = c(83,92,72,81))
 
-4. Suppose the probabilities of a visitor to Amazon’s website buying 0, 1, or 2 books are 0.2, 0.4, and 0.4 respectively.
+ggplot(data = df)+
+  aes(x = x)+
+  geom_histogram(breaks=seq(0,100,10), color="black")+
+  scale_x_continuous(breaks=seq(0,100,10))+
+  theme_classic()
+```
 
-a. Calculate the *expected number* of books a visitor will purchase.
-b. Calculate the *standard deviation* of book purchases.
-c. **Bonus**: try doing this in `R` by making an initial dataframe of the data, and then making new columns to the "table" like we did in class.
+## e.
+Suppose instead the person who got the 72 did not show up that day to class, and got a 0 instead. Recalculate the mean and median. What happened and why?
 
-\clearpage
+```{r}
+mean(c(83,92,0,81))
+median(c(83,92,0,81))
+```
 
-5. Scores on the SAT (out of 1600) are approximately normally distributed with a mean of 500 and standard deviation of 100.
+## 4.
 
-a. What is the probability of getting a score between a 400 and a 600?
-b. What is the probability of getting a score between a 300 and a 700?
-c. What is the probability of getting *at least* a 700?
-d. What is the probability of getting *at most* a 700?
-e. What is the probability of getting exactly a 500?
+Suppose the probabilities of a visitor to Amazon’s website buying 0, 1, or 2 books are 0.2, 0.4, and 0.4 respectively.
 
-\clearpage
+### a.
+Calculate the *expected number* of books a visitor will purchase.
 
-6.
+### b. 
+Calculate the *standard deviation* of book purchases.
+
+### c. 
+**Bonus**: try doing this in `R` by making an initial dataframe of the data, and then making new columns to the "table" like we did in class.
+
+## 5.
+
+Scores on the SAT (out of 1600) are approximately normally distributed with a mean of 500 and standard deviation of 100.
+
+### a.
+
+What is the probability of getting a score between a 400 and a 600?
+
+### b.
+
+What is the probability of getting a score between a 300 and a 700?
+
+### c.
+
+What is the probability of getting *at least* a 700?
+
+### d.
+
+What is the probability of getting *at most* a 700?
+
+### e.
+
+What is the probability of getting exactly a 500?
+
+## 6.
+
 Redo problem 5 by using the `pnorm()` command in `R`.^[Hint: This function has four arguments: 1.

### 2019-09-24 static/slides/09-slides.Rmd
Add class 9 materials, class 10 reading

---
title: "2.3: OLS Linear Regression"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))
```

class: inverse, center, middle
# Exploring Relationships

---

# Bivariate Data and Relationships

- We looked at single variables for descriptive statistics
- Most uses of statistics in economics and business investigate relationships *between* variables

.content-box-green[
.green[**Examples**]

- \# of police & crime rates
- healthcare spending & life expectancy
- government spending & GDP growth
- carbon dioxide emissions & temperatures
]

- We will begin with .whisper[bivariate] data for relationships between $X$ and $Y$
- Immediate aim is to explore .onfire[associations] between variables, quantified with .shout[correlation] and .shoutcaps[linear regression]
- Later we want to develop more sophisticated tools to argue for .onfire[causation]

---

# Bivariate Data: Spreadsheets I

.pull-left[

```{r econfreedomtable, echo=T, eval=F}
econfreedom <- read_csv("../data/econfreedom.csv")
head(econfreedom)
```

]

.pull-right[
```{r, echo=F, ref.label="econfreedomtable",out.width ="90%"}
```
]

- **Rows** are individual observations
- **Columns** are variables on all individuals

---

# Bivariate Data: Spreadsheets II

```{r echo=T}
econfreedom %>%
  glimpse()
```

---

# Bivariate Data: Spreadsheets III

```{r, echo=T, eval=F}
source("../files/summaries.R") # use my summary_table function

econfreedom %>%
    summary_table(ef, gdp)
```

```{r}
source("../files/summaries.R") # use my summary_table function

econfreedom %>%
    summary_table(ef, gdp) %>%
  knitr::kable(., format = "html")
```

---

# Bivariate Data: Scatterplots

.pull-left[

- The best way to visualize an association between two variables is with a .shout[scatterplot]

- Each point: pair of variable values $(x_i,y_i) \in X, Y$ for observation $i$

```{r econfreedom-scatter, eval=F, echo = T}
library("ggplot2")
ggplot(data = econfreedom)+
  aes(x = ef, y = gdp)+
  geom_point(aes(color = continent),
             size = 2)+
  labs(x = "Economic Freedom Index (2014)",
       y = "GDP per Capita (2014 USD)",
       color = "")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)+
  theme(legend.position = "bottom")
```
]

.pull-right[
```{r econfreedom-scatter-out, ref.label="econfreedom-scatter", fig.retina=3, fig.align="center"}
```

]

---

# Associations

.pull-left[

- Look for .onfire[association] between independent and dependent variables

1. **Direction**: is the trend positive or negative?

2. **Form**: is the trend linear, quadratic, something else, or no pattern?

3. **Strength**: is the association strong or weak?

4. **Outliers**: do any observations break the trends above?

]

.pull-right[
```{r, ref.label="econfreedom-scatter", fig.retina=3, fig.align="center"}
```

]

---

class: inverse, center, middle

# Quantifying Relationships

---

# Covariance

- For any two variables, we can measure their .shout[sample covariance, `\\(cov(X,Y)\\)`] or .shout[`\\(s_{X,Y}\\)`] to quantify how they vary *together*<sup>.red[1]</sup>

--

$$s_{X,Y}=E\big[(X-\bar{X})(Y-\bar{Y}) \big]$$

--

- Intuition: if $X$ is above its mean, would we expect $Y$:
    - to be *above* its mean also $(X$ and $Y$ covary *positively*)
    - to be *below* its mean $(X$ and $Y$ covary *negatively*)

--

- Covariance is a common measure, but the units are meaningless, thus we rarely need to use it so **don't worry about learning the formula**

.footnote[<sup>.red[1]</sup> Henceforth we limit all measures to *samples*, for convenience. Population covariance is denoted `\\(\sigma_{X,Y}\\)`]

---

# Covariance, in R

```{r, echo = T}
cov(econfreedom$ef,econfreedom$gdp)
```

```{r, echo = T}
econfreedom %>%
  summarize(cov = cov(ef,gdp))
```

---

# Correlation

- More convenient to *standardize* covariance into a more intuitive concept: .shout[correlation, `\\(\rho\\)`] or .shout[`\\(r\\)`] $\in [-1, 1]$

--

$$r_{X,Y}=\frac{s_{X,Y}}{s_X s_Y}=\frac{cov(X,Y)}{sd(X)sd(Y)}$$

--

- Simply weight covariance by the product of the standard deviations of $X$ and $Y$

--

- Alternatively, take the average<sup>.red[1]</sup> of the product of standardized ($Z$-scores for) each $(x_i,y_i)$ pair:<sup>.red[2]</sup>

--

$$\begin{align*}
r&=\frac{1}{n-1}\sum^n_{i=1}\bigg(\frac{x_i-\bar{X}}{s_X}\bigg)\bigg(\frac{y_i-\bar{Y}}{s_Y}\bigg)\\
r&=\frac{1}{n-1}\sum^n_{i=1}Z_XZ_Y\\
\end{align*}$$

.footnote[<sup>.red[1]</sup> Over `\\(n-1\\)`, since this is a *sample* statistic.

<sup>.red[2]</sup> See today's [class notes page](/class/09-class) for an example of the code for how to calculate correlation "by hand" in R using the second method.]

---

# Correlation: Interpretation

.pull-left[

- Correlation is standardized to 
$$-1 \leq r \leq 1$$
    - Negative values $\implies$ negative association
    - Positive values $\implies$ positive association
    - Correlation of 0 $\implies$ no association
    - As $|r| \rightarrow 1 \implies$ the stronger the association
    - Correlation of $|r|=1 \implies$ perfectly linear
]

.pull-right[

```{r, echo=F}
library("MASS")
library("gridExtra")
library("ggthemes")

# generated data for specified correlation: https://stackoverflow.com/questions/13291308/generate-numbers-with-specific-correlation
# colors chosen from http://colorbrewer2.org/#type=diverging&scheme=RdBu&n=6

# corr of -.75

dfn75 <- mvrnorm(50, mu = c(0,0), Sigma = matrix(c(1,-0.75,-0.75,1), ncol = 2),
               empirical = TRUE)
dfn75<-data.frame(dfn75)
dfn75 <- dfn75 %>%
  rename(y=X1,
         x=X2)

corrn75<-ggplot(dfn75, aes(x=x,
                   y=y))+
  geom_point(color="#b2182b")+
  theme_void()+
  labs(title="Correlation: -0.75")

# corr of -.5

dfn50 <- mvrnorm(50, mu = c(0,0), Sigma = matrix(c(1,-0.5,-0.5,1), ncol = 2),
               empirical = TRUE)
dfn50<-data.frame(dfn50)
dfn50 <- dfn50 %>%
  rename(y=X1,
         x=X2)

corrn50<-ggplot(dfn50, aes(x=x,
                   y=y))+
  geom_point(color="#ef8a62")+
  theme_void()+
  labs(title="Correlation: -0.50")

# corr of -.25

dfn25 <- mvrnorm(50, mu = c(0,0), Sigma = matrix(c(1,-0.25,-0.25,1), ncol = 2),
               empirical = TRUE)
dfn25<-data.frame(dfn25)
dfn25 <- dfn25 %>%
  rename(y=X1,
         x=X2)

corrn25<-ggplot(dfn25, aes(x=x,
                   y=y))+
  geom_point(color="#fddbc7")+
  theme_void()+
  labs(title="Correlation: -0.25")

# corr of 0.25

df25 <- mvrnorm(50, mu = c(0,0), Sigma = matrix(c(1,0.25,0.25,1), ncol = 2),
               empirical = TRUE)
df25<-data.frame(df25)
df25 <- df25 %>%
  rename(y=X1,
         x=X2)

corr25<-ggplot(df25, aes(x=x,
                   y=y))+
  geom_point(color="#d1e5f0")+
  theme_void()+
  labs(title="Correlation: 0.25")

# corr of .50

df50 <- mvrnorm(50, mu = c(0,0), Sigma = matrix(c(1,0.50,0.50,1), ncol = 2),
               empirical = TRUE)
df50<-data.frame(df50)
df50 <- df50 %>%
  rename(y=X1,
         x=X2)

corr50<-ggplot(df50, aes(x=x,
                   y=y))+
  geom_point(color="#67a9cf")+
  theme_void()+
  labs(title="Correlation: 0.50")

# corr of 0.75

df75 <- mvrnorm(50, mu = c(0,0), Sigma = matrix(c(1,0.75,0.75,1), ncol = 2),
               empirical = TRUE)
df75<-data.frame(df75)
df75 <- df75 %>%
  rename(y=X1,
         x=X2)

corr75<-ggplot(df75, aes(x=x,
                   y=y))+
  geom_point(color="#2166ac")+
  theme_void()+
  labs(title="Correlation: 0.75")

# arrange all on grid
grid.arrange(corrn25, corrn50, corrn75, corr25, corr50, corr75, ncol=3)

```
]

---

# Guess the Correlation!

.center[
![:scale 25%](https://www.dropbox.com/s/70opitsw7jjpfko/guessthecorrelation.png?raw=1)

[Guess the Correlation Game](http://guessthecorrelation.com/index.html)
]

---

# Correlation and Endogeneity

- Your Occasional Reminder: .shoutcaps[Correlation does not imply causation!]
    - I'll show you the difference in a few weeks (when we can actually talk about causation)

- If $X$ and $Y$ are strongly correlated, $X$ can still be **endogenous**!

- See today's class notes page for more on Covariance and Correlation

---

# Always Plot Your Data!

.center[
![](https://d2f99xq7vri1nk.cloudfront.net/DinoSequentialSmaller.gif)
]

---

class: inverse, center, middle

# Linear Regression

---

# Fitting a Line to Data

.pull-left[

- If an association appears linear, we can estimate the equation of a line that would "fit" the data

$$Y = a + bX$$

- Recall a linear equation describing a line contains:<sup>.red[1]</sup>
    - $a$: vertical intercept
    - $b$: slope

.footnote[<sup>.red[1]</sup> Note we'll use different symbols for `\\(a\\)` & `\\(b\\)`, the standard econometric notation.]

]

.pull-right[

```{r, fig.retina=3}
x <- runif(100,1,9)
e <- rnorm(100,0,2.5)
y <- 12-0.5*x+e
df <- tibble(x,y)

p <- ggplot(df, aes(x=x,y=y))+
  geom_point()+
  scale_x_continuous(breaks = seq(0,20,1),
                     limits = c(0,10),
                     expand = c(0,0))+
  scale_y_continuous(breaks = seq(0,20,2),
                     limits = c(0,20),
                     expand = c(0,0))+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)

p+geom_abline(slope=-1, intercept=14, color="red", size=1)
  #geom_smooth(method="lm", se=F)
```

]

---

# Fitting a Line to Data

.pull-left[

- If an association appears linear, we can estimate the equation of a line that would "fit" the data

$$Y = a + bX$$

- Recall a linear equation describing a line contains:<sup>.red[1]</sup>
    - $a$: vertical intercept
    - $b$: slope

- How do we choose the equation that *best* fits the data?

.footnote[<sup>.red[1]</sup> Note we'll use different symbols for `\\(a\\)` & `\\(b\\)`, the standard econometric notation.]

]

.pull-right[

```{r, fig.retina=3}
p+geom_abline(slope=-1, intercept=14, color="red", size=1)+
  geom_abline(slope=-0.125, intercept=10, color="orange", size=1)
  #geom_smooth(method="lm", se=F)
```

]

---

# Fitting a Line to Data

.pull-left[

- If an association appears linear, we can estimate the equation of a line that would "fit" the data

$$Y = a + bX$$

- Recall a linear equation describing a line contains:<sup>.red[1]</sup>
    - $a$: vertical intercept
    - $b$: slope

- How do we choose the equation that *best* fits the data?

- This process is called .shout[linear regression]

.footnote[<sup>.red[1]</sup> Note we'll use different symbols for `\\(a\\)` & `\\(b\\)`, the standard econometric notation.]

]

.pull-right[
```{r, fig.retina=3}
p+geom_abline(slope=-1, intercept=14, color="red", size=1)+
  geom_abline(slope=-0.125, intercept=10, color="orange", size=1)+
  geom_smooth(method="lm", se=F, color="green", size=2, fullrange = T)
```

]

---

# Population Linear Regression Model

- Linear regression lets us estimate the slope of the population regression line between $X$ and $Y$

- We can make **inferences** about the population slope coefficient
    - eventually, a causal interpretation
    - $\text{slope}=\frac{\Delta Y}{\Delta X}$: for a 1-unit change in $X$, how many units will this *cause* $Y$ to change?

---

# Class Size Example

.pull-left[

.content-box-green[
.green[**Example**]: What is the relationship between class size and educational performance?
]

]

.pull-right[

.center[
![](https://www.dropbox.com/s/80sb1duv6e2tazd/smallclass.jpg?raw=1)
]
]

---

# Class Size Example: Load the Data

```{r, echo=TRUE}
# install.packages("haven") # install for first use
library("haven") # load for importing .dta files
CASchool<-read_dta("../data/caschool.dta")
```

---

# Class Size Example: Look at the Data I

```{r, echo=T}
glimpse(CASchool)
```

---

# Class Size Example: Look at the Data II

```{r, results="asis"}
library(kableExtra)
head(CASchool) %>%
  knitr::kable(.,format="html") %>%
  kable_styling(font_size = 8)
```

---

# Class Size Example: Scatterplot

.pull-left[
```{r, school-scatter, eval=F, echo =T}
scatter<-ggplot(data = CASchool)+
  aes(x = str,
      y = testscr)+
  geom_point(color = "blue")+
  labs(x = "Student to Teacher Ratio",
       y = "Test Score")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size = 20)
scatter
```
]

.pull-right[
```{r, school-scatter-out, ref.label="school-scatter", fig.retina=3}
```
]

---

# Class Size Example: Slope I

.pull-left[

- If we *change* $(\Delta)$ the class size by an amount, what would we expect the *change* in test scores to be?

$$\beta = \frac{\text{change in test score}}{\text{change in class size}} = \frac{\Delta \text{test score}}{\Delta \text{class size}}$$

- If we knew $\beta$, we could say that changing class size by 1 student will change test scores by $\beta$

]

.pull-right[

.center[
![](https://www.dropbox.com/s/80sb1duv6e2tazd/smallclass.jpg?raw=1)
]
]

---

# Class Size Example: Slope II

.pull-left[

- Rearranging:

$$\Delta \text{test score} = \beta \times \Delta \text{class size}$$

]

.pull-right[

.center[
![](https://www.dropbox.com/s/80sb1duv6e2tazd/smallclass.jpg?raw=1)
]
]

# Class Size Example: Slope II

.pull-left[

- Rearranging:

$$\Delta \text{test score} = \beta \times \Delta \text{class size}$$

- Suppose $\beta=-0.6$. If we shrank class size by 2 students, our model predicts:

$$\begin{align*}
	\Delta \text{test score} &= -2 \times \beta\\
	\Delta \text{test score} &= -2 \times -0.6\\
	\Delta \text{test score}&= 1.2	\\
\end{align*}$$

]

.pull-right[

.center[
![](https://www.dropbox.com/s/80sb1duv6e2tazd/smallclass.jpg?raw=1)
]
]

---

# Class Size Example: Slope and Average Effect

.pull-left[

$$\text{test score} = \beta_0 + \beta_{1} \times \text{class size}$$

- The line relating class size and test scores has the above equation

- $\beta_0$ is the vertical-intercept, test score where class size is 0

- $\beta_{ClassSize}$ is the .onfire[slope] of the regression line

- This relationship only holds __on average__ for all districts in the population, *individual* districts are also affected by other factors

]

.pull-right[

.center[
![](https://www.dropbox.com/s/80sb1duv6e2tazd/smallclass.jpg?raw=1)
]
]

---

# Class Size Example: Marginal Effects

.pull-left[

- To get an equation that holds for *each* district, we need to include other factors

$$\text{test score} = \beta_0 + \beta_1 \text{class size}+\text{other factors}$$

- For now, we will ignore these until Unit 3

- Thus, $\beta_0 + \beta_1 \text{class size}$ gives the **average effect** of class sizes on scores

- Later, we will want to estimate the **marginal effect** (**causal effect**) of each factor on an individual district's test score, holding all other factors constant

]

.pull-right[

.center[
![](https://www.dropbox.com/s/80sb1duv6e2tazd/smallclass.jpg?raw=1)
]
]

---

# Econometric Models Overview

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + u$$

--

- $Y$ is the .shout[dependent variable] of interest
    - AKA "response variable," "regressand," "Left-hand side (LHS) variable"

--

- $X_1$ and $X_2$ are .shout[independent variables]
    - AKA "explanatory variables", "regressors," "Right-hand side (RHS) variables", "covariates"

--

- Our data consists of a spreadsheet of observed values of $(X_{1i}, X_{2i}, Y_i)$

--

- To model, we "regress Y on $X_1$ and $X_2$"

--

- $\beta_0$ and $\beta_1$ are .onfire[parameters] that describe the population relationships between the variables
    - unknown! to be estimated!

--

- $u$ is the random .shout[error term]
    - **'U'nobservable**, we can't measure it, and must model with assumptions about it

---

# The Population Regression Model

.pull-left[

- How do we draw a line through the scatterplot? We do not know the **"true"** $\beta_0$ or $\beta_1$

- We do have data from a *sample* of class sizes and test scores<sup>.red[1]</sup>

- So the real question is, .onfire[how can we estimate `\\(\beta_0\\)` and `\\(\beta_1\\)`?]
]

.pull-right[
```{r, echo=FALSE, fig.retina=3}
scatter
```
]

.footnote[<sup>.red[1]</sup> Data are student-teacher-ratio and average test scores on Stanford 9 Achievement Test for 5th grade students for 420 K-6 and K-8 school districts in California in 1999, (Stock and Watson, 2015: p. 141)]
---

class: inverse, center, middle

# Deriving OLS

---

# Deriving OLS

.pull-left[

- Suppose we have some data points

]

.pull-right[

```{r,echo =F}

x <- runif(25,0,10)
e <- rnorm(25,0,1)
y <- 2+0.5*x+e

df <- tibble(x,y)
p <- ggplot(df,
       aes(x = x,
           y = y))+
  geom_point()+
  scale_x_continuous(limits = c(0,10))+
  scale_y_continuous(limits = c(0,10))+
  #geom_smooth(method="lm")+
  theme_void()

p
```

]

---

# Deriving OLS

.pull-left[

- Suppose we have some data points
- We add a line

]

.pull-right[

```{r,echo =F}

p+geom_smooth(method="lm", se=F)
```

]

---

# Deriving OLS

.pull-left[

- Suppose we have some data points
- We add a line
- The .shout[residual], $\hat{u}$ of each data point is the difference between the .onfire[actual] and the .onfire[predicted] value of $Y$ given $X$:

$$u_i = Y_i - \hat{Y_i}$$

]

.pull-right[

```{r,echo =F}
library("broom")

reg <- lm(y ~ x, data = df)

reg_tidy <- tidy(reg)
reg_aug <- augment(reg)

p2 <- ggplot(reg_aug,
       aes(x = x,
           y = y))+
  geom_point()+
  geom_segment(aes(x=x, xend=x, y=y, yend=.fitted), color="red", linetype="dashed")+
  geom_smooth(method="lm", se=F)+
  scale_x_continuous(limits = c(0,10))+
  scale_y_continuous(limits = c(0,10))+
  theme_void()

p2
```

]

---

# Deriving OLS

.pull-left[

- Suppose we have some data points
- We add a line
- The .shout[residual], $\hat{u}$ of each data point is the difference between the .onfire[actual] and the .onfire[predicted] value of $Y$ given $X$:

$$u_i = Y_i - \hat{Y_i}$$

- We square each residual
]

.pull-right[

```{r,echo =F}
library("broom")

p2+geom_rect(aes(xmin=x, xmax=x+.resid, ymin=y, ymax=y-.resid),alpha=0.25,color="red", fill="red", linetype="dashed")
  #geom_tile(aes(x=x, y=y, width=.resid, height=.resid),alpha=0.125,fill="red",color="red")+
```

]

---

# Deriving OLS

.pull-left[

- Suppose we have some data points
- We add a line
- The .shout[residual], $\hat{u}$ of each data point is the difference between the .onfire[actual] and the .onfire[predicted] value of $Y$ given $X$:

$$u = Y_i - \hat{Y_i}$$

- We square each residual
- Add all of these up: .shout[Sum of Squared Errors (SSE)]

$$SSE = \sum^n_{i=1} u_i^2$$

]

.pull-right[

```{r,echo=F}
p2+geom_rect(aes(xmin=x, xmax=x+.resid, ymin=y, ymax=y-.resid),alpha=0.25,color="red", fill="red", linetype="dashed")
  #geom_tile(aes(x=x, y=y, width=.resid, height=.resid),alpha=0.125,fill="red",color="red")+
```

]

---

# Deriving OLS

.pull-left[

- Suppose we have some data points
- We add a line
- The .shout[residual], $\hat{u}$ of each data point is the difference between the .onfire[actual] and the .onfire[predicted] value of $Y$ given $X$:

$$u_i = Y_i - \hat{Y_i}$$

- We square each residual
- Add all of these up: .shout[Sum of Squared Errors (SSE)]

$$SSE = \sum^n_{i=1} u_i^2$$

- .onfire[The line of best fit *minimizes* SSE]

]

.pull-right[

```{r,echo =F}
p2+geom_rect(aes(xmin=x, xmax=x+.resid, ymin=y, ymax=y-.resid),alpha=0.25,color="red", fill="red", linetype="dashed")
  #geom_tile(aes(x=x, y=y, width=.resid, height=.resid),alpha=0.125,fill="red",color="red")+
```

]

---

### Corplots

```{r}
library(corrplot) # see more at https://github.com/taiyun/corrplot
library(RColorBrewer) # for color scheme used here
M <-cor(mtcars) # need to make a correlation matrix with cor(); can only include numeric variables
corrplot(M, type="upper", 
         method = "circle", # number for showing correlation coefficient
         order="alphabet", 
         col=brewer.pal(n=8, name="RdBu"))
```

---

# *O*rdinary *L*east *S*quares Estimators

- The .shout[Ordinary Least Squares (OLS) estimators] of the unknown population parameters $\beta_0$ and $\beta_1$, solve the calculus problem:

$$\min_{\beta_0, \beta_1} \sum^n_{i=1}[\underbrace{Y_i-(\underbrace{\beta_0+\beta_1 X_i}_{\hat{Y_i}})}_{u}]^2$$

--

- Intuitively, OLS estimators .onfire[minimize the average squared distance between the actual values `\\((Y_i)\\)` and the predicted values `\\((\hat{Y}_i)\\)` along the estimated regression line]

---

# The OLS Regression Line

- The .shout[OLS regression line] or .shout[sample regression line] is the linear function constructed using the OLS estimators:

$$\hat{Y_i}=\hat{\beta_0}+\hat{\beta_1}X_i$$

--

- $\hat{\beta_0}$ and $\hat{\beta_1}$ ("beta 0 hat" & "beta 1 hat") are the .shout[OLS estimators] of population parameters $\beta_0$ and $\beta_1$ using sample data

--

- The .shout[predicted value] of Y given X, based on the regression, is $E(Y_i|X_i)=\hat{Y_i}$

--

- The .onfire[residual] or .onfire[prediction error] for the $i^{th}$ observation is the difference between observed $Y_i$ and its predicted value, $\hat{u_i}=Y_i-\hat{Y_i}$

---

# The OLS Regression Estimators

- The solution to the SSE minimization problem yields:.footnote[See **Class Info Page** for proofs.]

--

$$\hat{\beta}_0=\bar{Y}-\hat{\beta_1}\bar{X}$$

--

$$\hat{\beta}_1=\frac{\displaystyle\sum^n_{i=1}(X_i-\bar{X})(Y_i-\bar{Y})}{\displaystyle\sum^n_{i=1}(X_i-\bar{X})^2}=\frac{s_{XY}}{s^2_X}= \frac{cov(X,Y)}{var(X)}$$

---

class: inverse, center, middle

# Our Class Size Example in R

---

# Class Size Scatterplot (Again)

.pull-left[

```{r school-scatter-2, echo=T, eval=F}
scatter
```

- There is some true (unknown) population relationship: 
$$\text{test score}=\beta_0+\beta_1 \times str$$

- $\beta_1=\frac{\Delta \text{test score}}{\Delta \text{str}}= ??$

]

.pull-right[
```{r, ref.label="school-scatter-2", fig.retina=3}
```

]

---

# Class SIze Scatterplot with Regression Line

.pull-left[

```{r school-scatter-line, echo=T, eval=F}
scatter+
  geom_smooth(method = "lm", color = "red") #<<
```

]

.pull-right[
```{r school-scatter-line-out, echo=F}
scatter+
  geom_smooth(method = "lm", color = "red") #<<
```

]

---

# OLS in R

.left-code[
```{r, echo=T}
# run regression of testscr on str
school_reg <- lm(testscr ~ str, 
                 data = CASchool)
```
]

.right-plot[
- Format for regression is `lm(y ~ x, data = df)`
- `y` is dependent variable (listed first!)
- `~` means "modeled by"
- `x` is the independent variable
- `df` is the dataframe where the data is stored
]

---

# OLS in R II

.left-code[
```{r, echo=T}
# look at reg object
school_reg
```
]
.right-plot[
- Stored as an `lm` object called `school_reg`, a `list` object
]

---

# OLS in R III

.left-code[

```{r, echo=T}
summary(school_reg) # get full summary
```

]

.right-plot[

- Looking at the `summary`, there's a lot of information here!

- These objects are cumbersome, come from a much older, pre-`tidyverse` epoch of `base R`

- Luckily, we now have `tidy` ways of working with regressions!
]

---

# Tidy OLS in R: broom

.left-code[

```{r, echo=T}
# load packages
library(broom)

# tidy regression output
tidy(school_reg)
```
]

.right-plot[

- The `broom` package allows us to tidy up regression objects

- The `tidy()` function creates a tidy tibble of regression output
]

---

# OLS in R IV

.left-code[

```{r, echo=T, eval=F}
# load packages
library(broom)

# tidy regression output
tidy(school_reg,
     conf.int = TRUE) # add CI's #<<
```

```{r, echo=F}
tidy(school_reg, conf.int = TRUE)
```
]

.right-plot[

- The `broom` package allows us to tidy up regression objects

- The `tidy()` function creates a tidy tibble of regression output

- Add `conf.int = TRUE` option to include confidence intervals for each coefficient
]

---

# More broom Tools: glance

- `glance()` shows us a lot of overall regression statistics and diagnostics
    - We'll interpret these in the next lecture and beyond

```{r, echo=T}
# look at regression statistics and diagnostics
glance(school_reg)
```

---

# More broom Tools: augment

.left-code[

```{r, echo=T}
augment(school_reg)
```
]

.right-plot[

- `augment()` creates useful new variables in the stored `lm` object
    - `.fitted` are the fitted (predicted) values from the model, i.e. $\hat{Y}_i$
    - `.resid` are the residuals (errors) from the model, i.e. $\hat{u}_i$
]

---

# Class Size Regression Result I

- Using OLS, we find:
$$\widehat{\text{test score}}=689.9-2.28 \times str$$

---

# Class Size Regression Result II

- There's a great package called `equatiomatic` that prints this equation in `markdown` or $\LaTeX$.

```{r, results="asis"}
library(equatiomatic)
extract_eq(school_reg, use_coefs = TRUE,coef_digits = 2,fix_signs = TRUE)
```

--

Here was my code:

```{r, echo = T}
# install.packages("equatiomatic") # install for first use
library(equatiomatic)
extract_eq(school_reg, 
           use_coefs = TRUE,
           coef_digits = 2,
           fix_signs = TRUE)
```

- In `R` chunk in `R markdown`, set `{r, results="asis"}` to print this raw output to be rendered

---

# Class Size Regression: A Data Point

.pull-left[
- One district in our sample is Richmond, CA:

```{r, echo = T}
CASchool %>%
  filter(district=="Richmond Elementary") %>%
  dplyr::select(district, testscr, str) # another package masks "select"
```

- Predicted value: $\widehat{\text{Test Score}}_{\text{Richmond}}=698-2.28(22) \approx 648$

- Residual: $\hat{u}_{Richmond}=672-648 \approx 24$

]

.pull-right[
```{r, echo=FALSE, fig.retina=3}

Richmond <- CASchool %>%
  filter(district=="Richmond Elementary")

ggplot(data = CASchool,
       aes(x = str,
           y = testscr))+
  geom_point(color="blue")+
  geom_smooth(method="lm", color="red")+
  geom_point(data = Richmond, color="magenta")+
  geom_text(data = Richmond, color="magenta", label="Richmond", vjust=-1)+
  geom_segment(x=22,y=672.4, xend=22,yend=652.4,linetype=2, color="magenta")+ #connect Richmond to regression line to show residual
  labs(x = "Student to Teacher Ratio",
       y = "Test Score")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size = 20)
```
]

### 2019-09-24 static/slides/09-slides.Rmd
Tweak slides 9

@@
 
-- We will begin with .whisper[bivariate] data for relationships between $X$ and $Y$
-- Immediate aim is to explore .onfire[associations] between variables, quantified with .shout[correlation] and .shoutcaps[linear regression]
+- We will begin with **bivariate** data for relationships between $X$ and $Y$
+- Immediate aim is to explore .onfire[associations] between variables, quantified with .shout[correlation] and .shout[linear regression]
 - Later we want to develop more sophisticated tools to argue for .onfire[causation]
@@
 ```{r, echo = T}
+# base R 
 cov(econfreedom$ef,econfreedom$gdp)
@@
 ```{r, echo = T}
+# dplyr
+
 econfreedom %>%
@@
 
-- Alternatively, take the average<sup>.red[1]</sup> of the product of standardized ($Z$-scores for) each $(x_i,y_i)$ pair:<sup>.red[2]</sup>
+- Alternatively, take the average<sup>.red[1]</sup> of the product of standardized $(Z$-scores for) each $(x_i,y_i)$ pair:<sup>.red[2]</sup>
 
@@
 
-.footnote[<sup>.red[1]</sup> Over `\\(n-1\\)`, since this is a *sample* statistic.
+.footnote[<sup>.red[1]</sup> Over `\\((n-1)\\)`, since this is a *sample* statistic.
 
@@
 
+# Correlation and Covariance in R
+
+```{r, echo = T}
+# Base r: cov or cor(df$x, df$y)
+
+cov(econfreedom$ef, econfreedom$gdp)
+
+cor(econfreedom$ef, econfreedom$gdp)
+
+# dplyr
+
+econfreedom %>%
+  summarize(covariance = cov(ef, gdp),
+            correlation = cor(ef, gdp))
+
+```
+---
+
+# Correlation and Covariance in R I
+
+- `corrplot` is a great package (install and then load) to **visualize** correlations in data
+
+```{r, echo=T, eval=F}
+library(corrplot) # see more at https://github.com/taiyun/corrplot
+library(RColorBrewer) # for color scheme used here
+library(gapminder) # for gapminder data
+
+# need to make a corelation matrix with cor(); can only include numeric variables
+gapminder_cor<- gapminder %>%
+  dplyr::select(gdpPercap, pop, lifeExp)
+
+# make a correlation table with cor (base R)
+gapminder_cor_table<-cor(gapminder_cor)
+
+# view it
+gapminder_cor_table
+```
+
+---
+
+# Correlation and Covariance in R II
+
+.pull-left[
+
+```{r, echo=T, eval=F}
+
+corrplot(gapminder_cor_table, type="upper", 
+         method = "circle", # number for showing correlation coefficient
+         order="alphabet", 
+         col=brewer.pal(n=8, name="RdBu"))
+]
+
+--
+
+.pull-right[
+
+```{r, fig.retina=3}
+library(corrplot) # see more at https://github.com/taiyun/corrplot
+library(RColorBrewer) # for color scheme used here
+library(gapminder) # for gapminder data
+
+# need to make a corelation matrix with cor(); can only include numeric variables
+gapminder_cor<- gapminder %>%
+  dplyr::select(gdpPercap, pop, lifeExp)
+
+# make a correlation table with cor (base R)
+gapminder_cor_table<-cor(gapminder_cor)
+
+corrplot(gapminder_cor_table, type="upper", 
+         method = "circle", # number for showing correlation coefficient
+         order="alphabet", 
+         col=brewer.pal(n=8, name="RdBu"))
+
+```
+]
+
+---
+
 # Correlation and Endogeneity
 
-- Your Occasional Reminder: .shoutcaps[Correlation does not imply causation!]
+.pull-left[
+
+- Your Occasional Reminder: .shout[Correlation does not imply causation!]
     - I'll show you the difference in a few weeks (when we can actually talk about causation)
@@
 - See today's class notes page for more on Covariance and Correlation
+
+]
+
+.pull-right[
+
+.center[
+![](https://www.dropbox.com/s/v5vwsadw5vs448t/causality.jpg?raw=1)
+]
+]
 
@@
 - We can make **inferences** about the population slope coefficient
+
 - eventually, a causal interpretation
+
 - $\text{slope}=\frac{\Delta Y}{\Delta X}$: for a 1-unit change in $X$, how many units will this *cause* $Y$ to change?
@@
 
+---
+
 # Class Size Example: Slope II
@@
 
-- $\beta_{ClassSize}$ is the .onfire[slope] of the regression line
+- $\beta_{1}$ is the .onfire[slope] of the regression line
 
@@
 - So the real question is, .onfire[how can we estimate `\\(\beta_0\\)` and `\\(\beta_1\\)`?]
+
+.footnote[<sup>.red[1]</sup> Data are student-teacher-ratio and average test scores on Stanford 9 Achievement Test for 5th grade students for 420 K-6 and K-8 school districts in California in 1999, (Stock and Watson, 2015: p.
+141)]
+
 ]
@@
 
-.footnote[<sup>.red[1]</sup> Data are student-teacher-ratio and average test scores on Stanford 9 Achievement Test for 5th grade students for 420 K-6 and K-8 school districts in California in 1999, (Stock and Watson, 2015: p.
-141)]
 ---
@@
 
-$$u = Y_i - \hat{Y_i}$$
+$$u_i = Y_i - \hat{Y_i}$$
 
@@
 
-### Corplots
-
-```{r}
-library(corrplot) # see more at https://github.com/taiyun/corrplot
-library(RColorBrewer) # for color scheme used here
-M <-cor(mtcars) # need to make a correlation matrix with cor(); can only include numeric variables
-corrplot(M, type="upper", 
-         method = "circle", # number for showing correlation coefficient
-         order="alphabet", 
-         col=brewer.pal(n=8, name="RdBu"))
-```
-
----
-
-# *O*rdinary *L*east *S*quares Estimators
+# *O* rdinary *L* east *S* quares Estimators
 
@@
 
-- The solution to the SSE minimization problem yields:.footnote[See **Class Info Page** for proofs.]
+- The solution to the SSE minimization problem yields:<sup>.red[1]</sup>
 
@@
 
+.footnote[<sup>.red[1]</sup> See [tomorrow's class notes page](/class/10-class) for proofs.]
 ---

### 2019-09-24 static/slides/09-slides.Rmd
Fix slides 9 problems

@@
 
-.footnote[<sup>.red[1]</sup> Over `\\((n-1)\\)`, since this is a *sample* statistic.
+.footnote[<sup>.red[1]</sup> Over `\\(n-1\\)`, since this is a *sample* statistic.
 
@@
          col=brewer.pal(n=8, name="RdBu"))
+```
 ]
@@
 
-# Tidy OLS in R: broom
+# Tidy OLS in R: broom I
 
@@
 
-# OLS in R IV
+# Tidy OLS in R: broom II
 
@@
 ```{r, echo=T}
+# add regression-based values to data
 augment(school_reg)
@@
 - `augment()` creates useful new variables in the stored `lm` object
-    - `.fitted` are the fitted (predicted) values from the model, i.e. $\hat{Y}_i$
-    - `.resid` are the residuals (errors) from the model, i.e. $\hat{u}_i$
+    - `.fitted` are fitted (predicted) values from model, i.e. $\hat{Y}_i$
+    - `.resid` are residuals (errors) from model, i.e. $\hat{u}_i$
 ]
@@
 # install.packages("equatiomatic") # install for first use
-library(equatiomatic)
-extract_eq(school_reg, 
-           use_coefs = TRUE,
-           coef_digits = 2,
-           fix_signs = TRUE)
+library(equatiomatic) # load it
+extract_eq(school_reg, # regression lm object
+           use_coefs = TRUE, # use names of variables
+           coef_digits = 2, # round to 2 digits
+           fix_signs = TRUE) # fix negatives (instead of + -)
 ```
@@
   filter(district=="Richmond Elementary") %>%
-  dplyr::select(district, testscr, str) # another package masks "select"
+  dplyr::select(district, testscr, str)
 ```

### 2019-09-25 static/homeworks/01-problem-set-answers.Rmd
Add problem set 1 answers, add broom logo in slides 2.3

@@
 
-*Due by Thursday, September 19, 2019*
-
-- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/01-problem-set.pdf)
-- [<i class="fab fa-markdown"></i> `04-r-practice-answers.Rmd`](/homeworks/01-r-problem-set.Rmd)
-
-# Instructions
-
-For this problem set, you may type up your answers to each question (save plots as images and include them with your answers) and also turn in a (commented!) *.R* file of commands for each relevant question.
-You may email or turn in a paper copy.
-It is not relevant for this particular problem set, but in the future, I will allow handwritten answers for math/theory questions.
-
-Alternatively, you may download the `.Rmd` file, do the homework in markdown, and email to me a single `knit`ted `html` or `pdf` file (and be sure that it shows all of your code).
-
-To minimize confusion, I suggest creating a new `R Project` (e.g.
-`hw1`) and storing any data and plots in that folder on your computer.
-See my [example workflow](https://github.com/ryansafner/workflow).
-
-You may work together (and I highly encourage that) but you must turn in your own answers.
-I grade homeworks 70% for completion, and for the remaining 30%, pick one question to grade for accuracy - so it is best that you try every problem, even if you are unsure how to complete it accurately.
+Answers may be longer than I would deem sufficient on an exam.
+Some might vary slightly based on points of interest, examples, or personal experience.
+These suggested answers are designed to give you both the answer and a short explanation of *why* it is the answer.
 
@@
   arrange(desc(n)) %>% # arrange in largest-to-smallest order of n (number)
-  slice(1:5) %>% # optional, look only at first 5 rows
+  slice(1:5) %>% # optional, look only at first 5 rows; head(., n=5) also works
   mutate(percent = round(prop*100, 2)) # also optional, make a percent variable rounded to 2 decimals
@@
   arrange(desc(n))%>%
-  slice(1)
+  slice(1) # take first row only
 ```
@@
 
+# make a vector of the names (we'll need this for our graph below)
 top_boys_names<-c("James", "John", "Robert", "Michael", "William")
@@
 
+# make a vector of the names (we'll need this for our graph below)
 top_girls_names<-c("Mary", "Elizabeth", "Patricia", "Jennifer", "Linda")
@@
 
+There's a lot to this, so I'll break this up step by step and show you what happens at each major step.
+
+We want to find the names where 48% to 52% of the babies with the name are male, as I defined in the footnote.
+First let's `mutate` a variable to figure out how many babies with a particular name are male.
+
+To do this, we'll need to make a two variables to count the number of `male`s and `female`s of each name each year. We'll use the `ifelse()` function for each:
+
+1. Make a `male` variable where, for each name in each year, if `sex=="M"`, then count the number of males (`n`) that year, otherwise set it equal to `0`.
+2. Make a `female` variable where, for each name in each year, if `sex=="F"`, then count the number of females (`n`) that year, otherwise set it equal to `0`.
+
+```{r}
+babynames %>%
+  mutate(male = ifelse(sex == "M", n, 0),
+         female = ifelse(sex == "F", n, 0))
+```
+
+Now with this variable, we want to count the total number of males and females with each name over the entire dataset.
+Let's first `group_by(name)` so we'll get one row for every name.
+We will `summarize()` and take the `sum` of our `male` and of our `female` variables.
+
+```{r}
+babynames %>%
+  mutate(male = ifelse(sex == "M", n, 0),
+         female = ifelse(sex == "F", n, 0)) %>%
+  group_by(name) %>%
+    summarize(Male = sum(male),
+              Female = sum(female))
+```
+
+Now, we want to figure out what *fraction* of each name is Male or Female.
+It doesn't matter which we do here, I'll do Male.
+`mutate()` a new variable I'll call `perc_male` for the percent of the name being for Male babies.
+It takes the summed variables we made before, and takes the fraction that are Male, multiplying by 100 to get percents (which isn't necessary, but is easy to read).
+
 ```{r}
@@
          Female!=0) %>%
+  mutate(perc_male = (Male/(Male+Female)*100))
+```
+
+Right now, it's still in alphabetical order. We want to arrange it by `perc_male`, and more importantly, we want `perc_male` to be between 48 and 52, so let's `filter` accordingly:
+
+```{r}
+babynames %>%
+  mutate(male = ifelse(sex == "M", n, 0),
+         female = ifelse(sex == "F", n, 0)) %>%
+  group_by(name) %>%
+    summarize(Male = sum(male),
+              Female = sum(female))%>%
+  mutate(perc_male = (Male/(Male+Female)*100)) %>%
+  arrange(perc_male) %>%
+  filter(perc_male > 48,
+         perc_male < 52)
+```
+
+This gives us a lot of names, all falling between 48% and 52% male.
+But we want the most popular names that are in this range.
+So let's finally `mutate` a new variable called `total` that simply adds the number of `Male` and `Female` babies with a name.
+Then let's `arrange` our results by `desc(total)` to get the largest first, and then `slice(1:10)` to get the top 10 only.
+
+```{r}
+babynames %>%
+  mutate(male = ifelse(sex == "M", n, 0),
+         female = ifelse(sex == "F", n, 0)) %>%
+  group_by(name) %>%
+    summarize(Male = sum(male),
+              Female = sum(female))%>%
   mutate(perc_male = (Male/(Male+Female)*100)) %>%
@@
 
+I am creating this document for/from the website, so these are all stored in a folder called `data`, one folder up from my current folder, `homeworks`.
+To get there, I go up one folder (`..`) and move to `data`, where these `csv` files are stored.
+
+I suggest you either keep the data in the same folder as your `R` working directory (always check with `getwd()`), or create an R Project and store the data files in that same folder.
+
 ```{r}
+# import data with read_csv from readr
+
 # note these file paths will be different for you
-polfreedom<-read.csv(here::here("./static/data/freedomhouse2018.csv"))
-econfreedom<-read.csv(here::here("./static/data/econfreedom.csv"))
+polfreedom<-read_csv("../data/freedomhouse2018.csv")
+econfreedom<-read_csv("../data/econfreedom.csv")
 
+# look at each dataframe
 polfreedom
-
 econfreedom
@@
 **The `polfreedom` dataset is still a bit messy.
-Let's overwrite it (or assign to something like `polfreedom2`) and select `Country.Territory` and `Total` (total freedom score) and rename `Country.Territory` to `Country`.**
+Let's overwrite it (or assign to something like `polfreedom2`) and select `Country/Territory` and `Total` (total freedom score) and rename `Country.Territory` to `Country`.**
 
@@
 polfreedom<-polfreedom %>%
-  select(Country.Territory, Total) %>%
-  rename(Country=Country.Territory)
+  select(`Country/Territory`, Total) %>%
+  rename(Country=`Country/Territory`)
 ```
@@
 Since they both have `Country` as a variable, we can merge these tibbles using `left_join(econfreedom, polfreedom, by="Country")`^[Note, if you saved as something else in question 9., use that instead of `polfreedom`!] and save this as a new tibble (something like `freedom`).**
+
+This one is a bit advanced to explain (but see the last few slides of 1.5 for more), so just copy what I gave you!
 
@@
 ```{r, fig.retina=3}
-# install.packages("ggrepel")
-library(ggrepel)
+# install.packages("ggrepel") install for first use 
+library(ggrepel) # load
 

### 2019-09-25 static/slides/09-slides.Rmd
Add problem set 1 answers, add broom logo in slides 2.3

@@
 
-.left-code[
+.left-column[
+.center[
+![](https://www.dropbox.com/s/qpe1604cf7s9r0g/rbroom.png?raw=1)
+]
+
+]
+.right-column[
+
+- The `broom` package allows us to *tidy* up regression objects<sup>.red[1]</sup>
+
+- The `tidy()` function creates a *tidy* `tibble` of regression output
 
@@
 ```
-]
-
-.right-plot[
-
-- The `broom` package allows us to tidy up regression objects
-
-- The `tidy()` function creates a tidy tibble of regression output
-]
-
+
+]
+
+.footnote[<sup>.red[1]</sup> See more at [broom.tidyverse.org](https://broom.tidyverse.org/).]
 ---
@@
 
-.left-code[
-
-```{r, echo=T, eval=F}
+.left-column[
+.center[
+![](https://www.dropbox.com/s/qpe1604cf7s9r0g/rbroom.png?raw=1)
+]
+
+]
+.right-column[
+
+- The `broom` package allows us to *tidy* up regression objects<sup>.red[1]</sup>
+
+- The `tidy()` function creates a *tidy* `tibble` of regression output
+
+```{r, echo=T}
 # load packages
@@
 
-# tidy regression output
+# tidy regression output (with confidence intervals!)
 tidy(school_reg,
-     conf.int = TRUE) # add CI's #<<
-```
-
-```{r, echo=F}
-tidy(school_reg, conf.int = TRUE)
-```
-]
-
-.right-plot[
-
-- The `broom` package allows us to tidy up regression objects
-
-- The `tidy()` function creates a tidy tibble of regression output
-
-- Add `conf.int = TRUE` option to include confidence intervals for each coefficient
-]
+     conf.int = TRUE) #<<
+```
+
+]
+
+.footnote[<sup>.red[1]</sup> See more at [broom.tidyverse.org](https://broom.tidyverse.org/).]
 

### 2019-09-25 static/homeworks/01-problem-set-answers.R

@@ -1,0 +1,252 @@
+# 1 -------------------
+library(tidyverse)
+
+# install for first use
+# install.packages("babynames")
+
+# load package 
+library(babynames)
+
+# explore help
+# ?babynames
+
+# a ------------------
+
+# save as a new tibble
+top_5_boys_2017 <- babynames %>% # take data
+  filter(sex=="M", # filter by males
+         year==2017) %>% # and for 2007
+  arrange(desc(n)) %>% # arrange in largest-to-smallest order of n (number)
+  slice(1:5) %>% # optional, look only at first 5 rows; head(., n=5) also works
+  mutate(percent = round(prop*100, 2)) # also optional, make a percent variable rounded to 2 decimals
+
+# look at our new tibble
+top_5_boys_2017
+
+# b -----------------
+
+# save as a new tibble
+top_5_girls_2017 <- babynames %>% # take data
+  filter(sex=="F", # filter by females
+         year==2017) %>% # and for 2007
+  arrange(desc(n)) %>% # arrange in largest-to-smallest order of n (number)
+  slice(1:5) %>% # optional, look only at first 5 rows; head(., n=5) also works
+  mutate(percent = round(prop*100, 2)) # also optional, make a percent variable rounded to 2 decimals
+
+# look at our new tibble
+top_5_girls_2017
+
+# 2 ------------------
+
+# for boys
+ggplot(data = top_5_boys_2017)+
+  aes(x = reorder(name, n), #note this reorders the x variable from small to large n
+      y = percent, # you can use prop if you didn't make a percent variable
+      fill = name)+ # optional color!
+  geom_col()+
+  
+  # now I'm just making it pretty
+  scale_y_continuous(labels=function(x)paste(x,"%",sep=""))+ # optional, add percent signs
+  labs(x = "Name",
+       y = "Percent of All Babies With Name",
+       title = "Most Popular Boys Names Since 1880",
+       fill = "Boy's Name",
+       caption = "Source: SSA")+
+  theme_classic(base_family = "Fira Sans Condensed", base_size=16)+
+  coord_flip()+ # rotate axes!
+  theme(legend.position = "") # hide legend
+
+# for girls 
+ggplot(data = top_5_girls_2017)+
+  aes(x = reorder(name, n), #note this reorders the x variable from small to large n
[…]

### 2019-09-26 content/class/09-class.Rmd
Nicer formatting to problem set 1 answer key html, and link to it in class pages

@@
 
-[Problem Set 1](/assignment/01-problem-set.html) answers will be posted shortly, and the graded assignment emailed back to you.^[Sorry I've been too busy to send emails confirming that I recieved your problem set!]
+[Problem Set 1](/assignment/01-problem-set.html) answers are posted, and the graded assignment emailed back to you.^[Sorry I've been too busy to send emails confirming that I recieved your problem set!]
 

### 2019-09-26 static/homeworks/01-problem-set-answers.Rmd
Nicer formatting to problem set 1 answer key html, and link to it in class pages

@@
 
-## 1. 
-### a. 
+## Question 1
+
+### Part A 
 **What are the top 5 boys names for 2017, and what *percent* of overall names is each?**
@@
 
-### b.
+### Part B
 **What are the top 5 *girls* names, and what *percent* of overall names is each?**
@@
 
-## 2. 
+## Question 2
 **Make two barplots, of these top 5 names, one for each sex.
@@
 
-## 3. 
+## Question 3
 **Find your name.^[If your name isn't in there :(, pick a random name.] `count` by `sex` how many babies since 1880 were named your name.^[Hint: if you do this, you'll get the number of *rows* (years) there are in the data.
@@
 
-## 4.
+## Question 4
 **Make a line graph of the number of babies with your name over time, `color`ed by `sex`.**
@@
 
-## 5.
-### a.
+## Question 5
+### Part A
 **Make a table of the most common name for boys by year between 1980-2017.^[Hint: once you've got all the right conditions, you'll get a table with a lot of data.
@@
 
-### b.
+### Part B
 **Now do the same for girls.**
@@
 
-## 6. Now let's graph the evolution of the most common names since 1880.
-### a.
+## Question 6
+**Now let's graph the evolution of the most common names since 1880.**
+### Part A
 **First, find out what are the top 10 *overall* most popular names for boys and for girls. You may want to create two vectors, each with these top 5 names.**
@@
 
-### b.
+### Part B
 **Now make two `line`graphs of these 5 names over time, one for boys, and one for girls.**
@@
 
-## 7.
+## Question 7
 **_Bonus (hard!)_: What are the 10 most common "gender-neutral" names?^[This is hard to define.
@@
 
----
-
 # Political and Economic Freedom Around the World
@@
 
-## 8. 
+## Question 8
 **Download these two datasets that I've cleaned up a bit:^[If you want, try downloading them from the websites yourself!]**
@@
 
-## 9. 
+## Question 9 
 **The `polfreedom` dataset is still a bit messy.
@@
 
-## 10.
+## Question 10
 **Now we can try to merge these two datasets into one.
@@
 
-## 11.
+## Question 11
 **Now make a scatterplot of Political Freedom (`total`)^[Feel free to `rename` these!] as `y` on Economic Freedom (`ef`) as `x` and `color` by `continent`.**
@@
 
-## 12.
+## Question 12
 **Let's do this again, but highlight some key countries.
@@
 
-## 13.
+## Question 13
 **Make another plot similar to 12, except this time use GDP per Capita (`gdp`) as `y`.

### 2019-09-26 static/homeworks/01-problem-set-answers.Rmd
Update problem set 1 answers and R file

@@
 babynames %>%
-  group_by(year) %>%
+  group_by(year) %>% # we want one observation per year
   filter(sex == "M",
-         year>1980) %>%
-  arrange(desc(n))%>%
+         year>1979) %>% # or >==1980
+  arrange(desc(n))%>% # start with largest n first
   slice(1) # take first row only
@@
 babynames %>%
-  group_by(year) %>%
+  group_by(year) %>% # we want one observation per year
   filter(sex == "F",
-         year>1980) %>%
-  arrange(desc(n))%>%
+         year>1979) %>% # or >==1980
+  arrange(desc(n))%>% # start with largest n first
   slice(1) # take first row only
@@
 **Now let's graph the evolution of the most common names since 1880.**
+
 ### Part A
+
 **First, find out what are the top 10 *overall* most popular names for boys and for girls. You may want to create two vectors, each with these top 5 names.**
@@
 babynames %>%
-  group_by(name) %>%
+  group_by(name) %>% # we want one row per name
   filter(sex=="M") %>%
-  summarize(total=sum(n)) %>%
-  arrange(desc(total)) %>%
+  summarize(total=sum(n)) %>% # add upp all of the n's for all years for each name
+  arrange(desc(total)) %>% # list largest total first
   slice(1:5)
@@
 top_boys_names<-c("James", "John", "Robert", "Michael", "William")
+
+# you could alternatively add a command, %>% pull(name) to the first chunk of code, and it would do the same thing, but we'd want to save it, for example:
+
+babynames %>%
+  group_by(name) %>% # we want one row per name
+  filter(sex=="M") %>%
+  summarize(total=sum(n)) %>% # add upp all of the n's for all years for each name
+  arrange(desc(total)) %>% # list largest total first
+  slice(1:5) %>%
+  pull(name)
+
 ```
@@
 babynames %>%
-  group_by(name) %>%
+  group_by(name) %>% # we want one row per name
   filter(sex=="F") %>%
-  summarize(total=sum(n)) %>%
-  arrange(desc(total)) %>%
+  summarize(total=sum(n)) %>% # add upp all of the n's for all years for each name
+  arrange(desc(total)) %>% # list largest total first
   slice(1:5)
@@
 top_girls_names<-c("Mary", "Elizabeth", "Patricia", "Jennifer", "Linda")
+
+# you could alternatively add a command, %>% pull(name) to the first chunk of code, and it would do the same thing, but we'd want to save it
 ```

### 2019-09-26 static/homeworks/01-problem-set-answers.R

@@ -101,19 +101,19 @@ babynames %>%
 # a -----------------
 
 babynames %>%
-  group_by(year) %>%
+  group_by(year) %>% # we want one observation per year
   filter(sex == "M",
-         year>1980) %>%
-  arrange(desc(n))%>%
+         year>1979) %>% # or >==1980
+  arrange(desc(n))%>% # start with largest n first
   slice(1) # take first row only
 
 # b -----------------
 
 babynames %>%
-  group_by(year) %>%
+  group_by(year) %>% # we want one observation per year
   filter(sex == "F",
-         year>1980) %>%
-  arrange(desc(n))%>%
+         year>1979) %>% # or >==1980
+  arrange(desc(n))%>% # start with largest n first
   slice(1) # take first row only
 
 
@@ -121,28 +121,39 @@ babynames %>%
 
 # a -----------------
 
-# for boys 
 babynames %>%
-  group_by(name) %>%
+  group_by(name) %>% # we want one row per name
   filter(sex=="M") %>%
-  summarize(total=sum(n)) %>%
-  arrange(desc(total)) %>%
-  slice(1:5)
+  summarize(total=sum(n)) %>% # add upp all of the n's for all years for each name
+  arrange(desc(total)) %>% # list largest total first
+  slice(1:5) 
 
 # make a vector of the names (we'll need this for our graph below)
 top_boys_names<-c("James", "John", "Robert", "Michael", "William")
 
+# you could alternatively add a command, %>% pull(name) to the first chunk of code, and it would do the same thing, but we'd want to save it, for example:
+
+babynames %>%
+  group_by(name) %>% # we want one row per name
+  filter(sex=="M") %>%
+  summarize(total=sum(n)) %>% # add upp all of the n's for all years for each name
+  arrange(desc(total)) %>% # list largest total first
+  slice(1:5) %>%
+  pull(name)
+
 # for girls
 
 babynames %>%
-  group_by(name) %>%
+  group_by(name) %>% # we want one row per name
   filter(sex=="F") %>%
-  summarize(total=sum(n)) %>%
-  arrange(desc(total)) %>%
+  summarize(total=sum(n)) %>% # add upp all of the n's for all years for each name
+  arrange(desc(total)) %>% # list largest total first
   slice(1:5)
 
 # make a vector of the names (we'll need this for our graph below)
 top_girls_names<-c("Mary", "Elizabeth", "Patricia", "Jennifer", "Linda")
+
[…]

### 2019-09-26 static/homeworks/01-problem-set-answers.R
Format better for problem set 1 answers

@@ -82,9 +82,16 @@ babynames %>%
 
 # 4 -----------------
 
+# note here I'm going to wrangle the data and then pipe it directly into ggplot
+# you can wrangle the data and save it as a different tibble, then use THAT tibble
+# for your (data = ...) command in ggplot
+
+# first wrangle data
 babynames %>%
   filter(name == "Ryan") %>%
-  ggplot(data = .)+
+  
+  # now we pipe into ggplot
+  ggplot(data = .)+ # the "." is a placeholder for the stuff above!
   aes(x = year,
       y = n,
       color = sex)+
@@ -131,7 +138,10 @@ babynames %>%
 # make a vector of the names (we'll need this for our graph below)
 top_boys_names<-c("James", "John", "Robert", "Michael", "William")
 
-# you could alternatively add a command, %>% pull(name) to the first chunk of code, and it would do the same thing, but we'd want to save it, for example:
+# you could alternatively add a command, 
+# %>% pull(name) to the first chunk of code, 
+# and it would do the same thing, but we'd want to save it, 
+# for example:
 
 babynames %>%
   group_by(name) %>% # we want one row per name

### 2019-09-26 content/class/10-class.Rmd
Add class 10 content

---
title: "2.4: OLS: Goodness of Fit and Bias - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

*Tuesday, September 24, 2019*

# Overview

Today we start looking at *associations* between variables, which we will first attempt to quantify with measures like *covariance* and *correlation*. Then we turn to fitting a line to data via *linear regression*. We overview the basic regression model, the parameters and how they are derived, and see how to work with regressions in `R` with `lm` and the tidyverse package [`broom`](https://broom.tidyverse.org/).

We consider an extended example about class sizes and test scores, which comes from a (Stata) dataset from an old textbook that I used to use, [Stock and Watson, 2007](https://www.pearson.com/us/higher-education/product/Stock-Introduction-to-Econometrics-Update-3rd-Edition/9780133486872.html?tab=resources). Download and follow along with the data from today's example:^[Note this is a `.dta` Stata file. You will need to (install and) load the package `haven` to `read_dta()` Stata files into a dataframe.]

- [<i class="fas fa-table"></i> `caschool.dta`](http://metricsf19.classes.ryansafner.com/data/caschool.dta)

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/09-slides.html)

# Assignments: Problem Set 1 DUE and Problem Set 2

[Problem Set 1](/assignment/01-problem-set.html) answers will be posted shortly, and the graded assignment emailed back to you.^[Sorry I've been too busy to send emails confirming that I recieved your problem set!]

[Problem Set 2](/assignment/01-problem-set.html) (on classes 2.1-2.2) is due by Thursday September 26.

# Appendix: OLS Estimators

## Deriving the OLS Estimators

The population linear regression model is:

$$Y_i=\beta_0+\beta_1 X_i + u _i$$

The errors $(u_i)$ are unobserved, but for candidate values of $\hat{\beta_0}$ and $\hat{\beta_1}$, we can obtain an estimate of the residual. Algebraically, the error is:

$$\hat{u_i}=    Y_i-\hat{\beta_0}-\hat{\beta_1}X_i$$

Recall our goal is to find $\hat{\beta_0}$ and $\hat{\beta_1}$ that *minimizes* the sum of squared errors (SSE):

$$SSE= \sum^n_{i=1} \hat{u_i}^2$$

So our minimization problem is:

$$\min_{\hat{\beta_0}, \hat{\beta_1}} \sum^n_{i=1} (Y_i-\hat{\beta_0}-\hat{\beta_1}X_i)^2$$

Using calculus, we take the partial derivatives and set it equal to 0 to find a minimum. The first order conditions are:

$$\begin{align*}
\frac{\partial SSE}{\partial \hat{\beta_0}}&=-2\displaystyle\sum^n_{i=1} (Y_i-\hat{\beta_0}-\hat{\beta_1} X_i)=0\\
\frac{\partial SSE}{\partial \hat{\beta_1}}&=-2\displaystyle\sum^n_{i=1} (Y_i-\hat{\beta_0}-\hat{\beta_1} X_i)X_i=0\\
\end{align*}$$

### Finding $\hat{\beta_0}$

Working with the first FOC, divide both sides by $-2$:

$$\displaystyle\sum^n_{i=1} (Y_i-\hat{\beta_0}-\hat{\beta_1} X_i)=0$$

Then expand the summation across all terms and divide by $n$:

$$\underbrace{\frac{1}{n}\sum^n_{i=1} Y_i}_{\bar{Y}}-\underbrace{\frac{1}{n}\sum^n_{i=1}\hat{\beta_0}}_{\hat{\beta_0}}-\underbrace{\frac{1}{n}\sum^n_{i=1} \hat{\beta_1} X_i}_{\hat{\beta_1}\bar{X}}=0$$

Note the first term is $\bar{Y}$, the second is $\hat{\beta_0}$, the third is $\hat{\beta_1}\bar{X}$.^[From the [rules about summation operators](https://metricsf19.classes.ryansafner.com/class/07-class/#the-summation-operator), we define the mean of a random variable $X$ as $\bar{X}=\frac{1}{n}\displaystyle\sum_{i=1}^n X_i$. The mean of a constant, like $\beta_0$ or $\beta_1$ is itself.]

So we can rewrite as: 
$$\bar{Y}-\hat{\beta_0}-\beta_1=0$$

Rearranging:

$$\hat{\beta_0}=\bar{Y}-\bar{X}\beta_1$$

### Finding $\hat{\beta_1}$

To find $\hat{\beta_1}$, take the second FOC and divide by $-2$:

$$\displaystyle\sum^n_{i=1} (Y_i-\hat{\beta_0}-\hat{\beta_1} X_i)X_i=0$$

From the formula for $\hat{\beta_0}$, substitute in for $\hat{\beta_0}$:

$$\displaystyle\sum^n_{i=1} \bigg(Y_i-[\bar{Y}-\hat{\beta_1}\bar{X}]-\hat{\beta_1} X_i\bigg)X_i=0$$

Combining similar terms:

$$\displaystyle\sum^n_{i=1} \bigg([Y_i-\bar{Y}]-[X_i-\bar{X}]\hat{\beta_1}\bigg)X_i=0$$

Distribute $X_i$ and expand terms into the subtraction of two sums (and pull out $\hat{\beta_1}$ as a constant in the second sum:

$$\displaystyle\sum^n_{i=1} [Y_i-\bar{Y}]X_i-\hat{\beta_1}\displaystyle\sum^n_{i=1}[X_i-\bar{X}]X_i=0$$

Move the second term to the righthand side:

$$\displaystyle\sum^n_{i=1} [Y_i-\bar{Y}]X_i=\hat{\beta_1}\displaystyle\sum^n_{i=1}[X_i-\bar{X}]X_i$$

Divide to keep just $\hat{\beta_1}$ on the right:

$$\frac{\displaystyle\sum^n_{i=1} [Y_i-\bar{Y}]X_i}{\displaystyle\sum^n_{i=1}[X_i-\bar{X}]X_i}=\hat{\beta_1}$$

Note that from the [rules about summation operators](https://metricsf19.classes.ryansafner.com/class/07-class/#the-summation-operatorproperties):

$$\displaystyle\sum^n_{i=1} [Y_i-\bar{Y}]X_i=\displaystyle\sum^n_{i=1} (Y_i-\bar{Y})(X_i-\bar{X})$$

and:

$$\displaystyle\sum^n_{i=1} [X_i-\bar{X}]X_i=\displaystyle\sum^n_{i=1} (X_i-\bar{X})(X_i-\bar{X})=\displaystyle\sum^n_{i=1}(X_i-\bar{X})^2$$

Plug in these two facts:

$$\frac{\displaystyle\sum^n_{i=1} (Y_i-\bar{Y})(X_i-\bar{X})}{\displaystyle\sum^n_{i=1}(X_i-\bar{X})^2}=\hat{\beta_1}$$

## Algebraic Properties of OLS Estimators

The OLS residuals $\hat{u}$ and predicted values $\hat{Y}$ are chosen by the minimization problem to satisfy:

1. The expected value (average) error is 0: 
$$E(u_i)=\frac{1}{n}\displaystyle \sum_{i=1}^n \hat{u_i}=0$$

2. The covariance between $X$ and the errors is 0:
$$\hat{\sigma}_{X,u}=0$$

Note the first two properties imply strict **exogeneity**. That is, this is only a valid model if $X$ and $u$ are not correlated.

3. The expected predicted value of $Y$ is equal to the expected value of $Y$:
$$\bar{\hat{Y}}=\frac{1}{n} \displaystyle\sum_{i=1}^n \hat{Y_i} = \bar{Y}$$

4. Total sum of squares is equal to the explained sum of squares plus sum of squared errors:
$$\begin{align*}TSS&=ESS+SSE\\
\sum_{i=1}^n (Y_i-\bar{Y})^2&=\sum_{i=1}^n (\hat{Y_i}-\bar{Y})^2 + \sum_{i=1}^n {u}^2\\
\end{align*}$$

Recall $R^2$ is $\frac{ESS}{TSS}$ or $1-SSE$

5. The regression line passes through the point $(\bar{X},\bar{Y})$, i.e. the mean of $X$ and the mean of $Y$.

## Bias in $\hat{\beta_1}$

Begin with the formula we derived for $\hat{\beta_1}$:

$$\hat{\beta_1}=\frac{\displaystyle \sum^n_{i=1} (Y_i-\bar{Y})(X_i-\bar{X})}{\displaystyle\sum^n_{i=1} (X_i-\bar{X})^2}$$

Recall from **Rule 6** of summations, we can rewrite the numerator as

$$\begin{align*}
	=&\displaystyle \sum^n_{i=1} (Y_i-\bar{Y})(X_i-\bar{X})\\
	=& \displaystyle \sum^n_{i=1} Y_i(X_i-\bar{X})\\
\end{align*}$$

$$\hat{\beta_1}=\frac{\displaystyle \sum^n_{i=1} Y_i(X_i-\bar{X})}{\displaystyle\sum^n_{i=1} (X_i-\bar{X})^2}$$

We know the true population relationship is expressed as:

$$Y_i=\beta_0+\beta_1 X_i+u_i$$

Substituting this in for $Y_i$ in equation 2:

$$\hat{\beta_1}=\frac{\displaystyle \sum^n_{i=1} (\beta_0+\beta_1X_i+u_i)(X_i-\bar{X})}{\displaystyle\sum^n_{i=1} (X_i-\bar{X})^2}$$
Breaking apart the sums in the numerator:

$$\hat{\beta_1}=\frac{\displaystyle \sum^n_{i=1} \beta_0(X_i-\bar{X})+\displaystyle \sum^n_{i=1} \beta_1X_i(X_i-\bar{X})+\displaystyle \sum^n_{i=1} u_i(X_i-\bar{X})}{\displaystyle\sum^n_{i=1} (X_i-\bar{X})^2}$$

We can simplify equation 4 using **Rules 4 and 5** of summations

1. The first term in the numerator $\left[\displaystyle \sum^n_{i=1} \beta_0(X_i-\bar{X})\right]$ has the constant $\beta_0$, which can be pulled out of the summation. This gives us the summation of deviations, which add up to 0 as per **Rule 4**:

$$\begin{align*}
\displaystyle \sum^n_{i=1} \beta_0(X_i-\bar{X})&= \beta_0 \displaystyle \sum^n_{i=1} (X_i-\bar{X})\\
&=\beta_0 (0)\\
&=0\\
\end{align*}$$

2. The second term in the numerator  $\left[\displaystyle \sum^n_{i=1} \beta_1X_i(X_i-\bar{X})\right]$ has the constant $\beta_1$, which can be pulled out of the summation. Additionally, **Rule 5** tells us $\displaystyle \sum^n_{i=1} X_i(X_i-\bar{X})=\displaystyle \sum^n_{i=1}(X_i-\bar{X})^2$:

$$\begin{align*}
\displaystyle \sum^n_{i=1} \beta_1X_1(X_i-\bar{X})&= \beta_1 \displaystyle \sum^n_{i=1} X_i(X_i-\bar{X})\\
&=\beta_1\displaystyle \sum^n_{i=1}(X_i-\bar{X})^2\\
\end{align*}$$

When placed back in the context of being the numerator of a fraction, we can see this term simplifies to just $\beta_1$:

$$\begin{align*}
	\frac{\beta_1\displaystyle \sum^n_{i=1}(X_i-\bar{X})^2}{\displaystyle\sum^n_{i=1} (X_i-\bar{X})^2} &=\frac{\beta_1}{1} \times \frac{\displaystyle \sum^n_{i=1}(X_i-\bar{X})^2}{\displaystyle\sum^n_{i=1} (X_i-\bar{X})^2}\\
	&=\beta_1	\\
\end{align*}$$

Thus, we are left with:

$$\hat{\beta_1}=\beta_1+\frac{\displaystyle \sum^n_{i=1} u_i(X_i-\bar{X})}{\displaystyle\sum^n_{i=1} (X_i-\bar{X})^2}$$

Now, take the expectation of both sides:

$$E[\hat{\beta_1}]=E\left[\beta_1+\frac{\displaystyle \sum^n_{i=1} u_i(X_i-\bar{X})}{\displaystyle\sum^n_{i=1} (X_i-\bar{X})^2} \right]$$

We can break this up, using properties of expectations. First, recall $E[a+b]=E[a]+E[b]$, so we can break apart the two terms.

$$E[\hat{\beta_1}]=E[\beta_1]+E\left[\frac{\displaystyle \sum^n_{i=1} u_i(X_i-\bar{X})}{\displaystyle\sum^n_{i=1} (X_i-\bar{X})^2} \right]$$

Second, the true population value of $\beta_1$ is a constant, so $E[\beta_1]=\beta_1$.

Third, since we assume $X$ is also "fixed" and not random, the variance of $X$, $\displaystyle\sum_{i=1}^n (X_i-\bar{X})$, in the denominator, is just a constant, and can be brought outside the expectation.

$$E[\hat{\beta_1}]=\beta_1+\frac{E\left[\displaystyle \sum^n_{i=1} u_i(X_i-\bar{X})\right]	}{\displaystyle\sum^n_{i=1} (X_i-\bar{X})^2}$$

Thus, the properties of the equation are primarily driven by the expectation $E\bigg[\displaystyle \sum^n_{i=1} u_i(X_i-\bar{X})\bigg]$. We now turn to this term.

Use the property of summation operators to expand the numerator term:

$$\begin{align*}
	\hat{\beta_1}&=\beta_1+\frac{\displaystyle \sum^n_{i=1} u_i(X_i-\bar{X})}{\displaystyle\sum^n_{i=1} (X_i-\bar{X})^2} \\
		\hat{\beta_1}&=\beta_1+\frac{\displaystyle \sum^n_{i=1} (u_i-\bar{u})(X_i-\bar{X})}{\displaystyle\sum^n_{i=1} (X_i-\bar{X})^2} \\
\end{align*}$$

Now divide the numerator and denominator of the second term by $\frac{1}{n}$. Realize this gives us the covariance between $X$ and $u$ in the numerator and variance of $X$ in the denominator, based on their respective definitions.

$$\begin{align*}
	\hat{\beta_1}&=\beta_1+\cfrac{\frac{1}{n}\displaystyle \sum^n_{i=1} (u_i-\bar{u})(X_i-\bar{X})}{\frac{1}{n}\displaystyle\sum^n_{i=1} (X_i-\bar{X})^2} \\
	\hat{\beta_1}&=\beta_1+\cfrac{cov(X,u)}{var(X)} \\
	\hat{\beta_1}&=\beta_1+\cfrac{s_{X,u}}{s_X^2} \\
\end{align*}$$

By the Zero Conditional Mean assumption of OLS, $s_{X,u}=0$.

Alternatively, we can express the bias in terms of correlation instead of covariance:

$$E[\hat{\beta_1}]=\beta_1+\cfrac{cov(X,u)}{var(X)}$$

From the definition of correlation:

$$\begin{align*}
	 corX,u)&=\frac{cov(X,u)}{s_X s_u}\\
	 corX,u)s_Xs_u &=cov(X,u)\\
\end{align*}$$

Plugging this in:

$$\begin{align*}
E[\hat{\beta_1}]&=\beta_1+\frac{cov(X,u)}{var(X)} \\
E[\hat{\beta_1}]&=\beta_1+\frac{\big[corX,u)s_xs_u\big]}{s^2_X} \\
E[\hat{\beta_1}]&=\beta_1+\frac{corX,u)s_u}{s_X} \\
E[\hat{\beta_1}]&=\beta_1+corX,u)\frac{s_u}{s_X} \\
\end{align*}$$

## Proof of the Unbiasedness of $\hat{\beta_1}$

Begin with equation:

$$\hat{\beta_1}=\frac{\sum Y_iX_i}{\sum X_i^2}$$

Substitute for $Y_i$:

$$\hat{\beta_1}=\frac{\sum (\beta_1 X_i+u_i)X_i}{\sum X_i^2}$$

Distribute $X_i$ in the numerator:

$$\hat{\beta_1}=\frac{\sum \beta_1 X_i^2+u_iX_i}{\sum X_i^2}$$

Separate the sum into additive pieces:

$$\hat{\beta_1}=\frac{\sum \beta_1 X_i^2}{\sum X_i^2}+\frac{u_i X_i}{\sum X_i^2}$$

$\beta_1$ is constant, so we can pull it out of the first sum:

$$\hat{\beta_1}=\beta_1 \frac{\sum X_i^2}{\sum X_i^2}+\frac{u_i X_i}{\sum X_i^2}$$

Simplifying the first term, we are left with:

$$\hat{\beta_1}=\beta_1 +\frac{u_i X_i}{\sum X_i^2}$$

Now if we take expectations of both sides:

$$E[\hat{\beta_1}]=E[\beta_1] +E\left[\frac{u_i X_i}{\sum X_i^2}\right]$$

$\beta_1$ is a constant, so the expectation of $\beta_1$ is itself.

$$E[\hat{\beta_1}]=\beta_1 +E\bigg[\frac{u_i X_i}{\sum X_i^2}\bigg]$$

Using the properties of expectations, we can pull out $\frac{1}{\sum X_i^2}$ as a constant:

$$E[\hat{\beta_1}]=\beta_1 +\frac{1}{\sum X_i^2} E\bigg[\sum u_i X_i\bigg]$$

Again using the properties of expectations, we can put the expectation inside the summation operator (the expectation of a sum is the sum of expectations):

$$E[\hat{\beta_1}]=\beta_1 +\frac{1}{\sum X_i^2}\sum E[u_i X_i]$$

Under the exogeneity condition, the correlation between $X_i$ and $u_i$ is 0.

$$E[\hat{\beta_1}]=\beta_1$$

### 2019-09-26 static/slides/10-slides.Rmd
Add class 10 content

---
title: "2.4: OLS: Goodness of Fit and Bias"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))
```

```{r regression-setup, echo=F, results="hide"}

library(haven)
CASchool<-read_dta("../data/caschool.dta")

# run regression of testscr on str
school_reg <- lm(testscr ~ str, 
                 data = CASchool)

library(broom)
school_reg_tidy <- tidy(school_reg,
     conf.int = TRUE) # add confidence intervals
CASchool_aug <- augment(school_reg)

library(equatiomatic)
extract_eq(school_reg, use_coefs = TRUE,coef_digits = 2,fix_signs = TRUE)

```

class: inverse, center, middle
# Goodness of Fit

---

# Models

> "All models are wrong. But some are useful." - George Box

--

# All of Statistics:

$$Observed_i = Model_i + Error_i$$

---

# Goodness of Fit

- How well does a line fit data? How tightly clustered around the line are the data points?

- Quantify .onfire[how much variation in `\\(Y_i\\)` is "explained" by the model]

--

$$\underbrace{Y_i}_{Observed}=\underbrace{\widehat{Y_i}}_{Model}+\underbrace{\hat{u}}_{Error}$$

- Recall OLS estimators are chosen specifically to minimize the .shout[Sum of Squared Errors (SSE)]: $\left(\displaystyle \sum^n_{i=1}\hat{u_i}^2\right)$

---

# Goodness of Fit: $R^2$

- Primary measure<sup>.red[1]</sup> is .shout[regression R-squared], the fraction of variation in $Y$ explained by variation in predicted values

$$R^2 = \frac{\text{ variation in }\widehat{Y_i}}{\text{variation in }Y_i}$$

.footnote[<sup>.red[1]</sup> Sometimes called the .shout["coefficient of determination"]]

---

# Goodness of Fit: $R^2$ Formula

$$R^2 =\frac{ESS}{TSS}$$

--

- .shout[Explained Sum of Squares (ESS)]:<sup>.red[1]</sup> sum of squared deviations of *predicted* values from their mean<sup>.red[2]</sup>

--

$$ESS= \sum^n_{i=1}(\hat{Y_i}-\bar{Y})^2$$

--

- .shout[Total Sum of Squares (TSS)]: sum of squared deviations of *observed* values from their mean

--

$$TSS= \sum^n_{i=1}(Y_i-\bar{Y})^2$$

.footnote[

<sup>1</sup> Sometimes called Model Sum of Squares (MSS) or Regression Sum of Squares (RSS) in other textbooks

<sup>2</sup> It can be shown that $\bar{\hat{Y_i}}=\bar{Y}$]

---

# Goodness of Fit: $R^2$ Formula II

- Equivalently, the complement of the fraction of *unexplained* variation in $Y_i$

$$R^2=1-\frac{SSE}{TSS}$$

- Equivalently, the square of the correlation coefficient between $X$ and $Y$:

--

$$R^2=(r_{X,Y})^2$$

---

# Calculating $R^2$ in R I

- If we wanted to calculate it manually:

```{r, echo=T}
# as squared correlation coefficient

# Base R
cor(CASchool$testscr, CASchool$str)^2

# dplyr
CASchool %>%
  summarize(r_sq = cor(testscr,str)^2)
```

---

# Calculating $R^2$ in R II

- Recall `broom`'s `augment()` command makes a lot of new regression-based values, such as:
    - `.fitted`: predicted values $(\hat{Y_i})$
    - `.resid`: residuals $(\hat{u_i})$

```{r, echo=T}
library(broom)
school_reg %>%
  augment() %>%
  head(., n=5) # show first 5 values
```

---

# Calculating $R^2$ in R III

- We can calculate `R` as the ratio of variances in model vs. actual (i.e. akin to $\frac{ESS}{TSS}$)

```{r, echo = T}
# as ratio of variances
school_reg %>%
  augment() %>%
  summarize(r_sq = var(.fitted)/var(testscr)) # variance of *predicted* testscr over variance of *actual* teestscr
```

---

# Goodness of Fit: Standard Error of the Regression

- The .shout[Standard Error of the Regression], $\hat{\sigma}$ or $\hat{\sigma}_{u}$ is an estimator of the standard deviation of $u_i$

--

$$\hat{\sigma_u}=\sqrt{\frac{SSE}{n-2}}$$

--

- Measures the **average size of the residuals** (distance between a data point and the regression line)
    - **Degrees of Freedom correction** of $n-2$: we use up 2 df to first calculate $\hat{\beta_0}$ and $\hat{\beta_1}$!

---

# Calculating SER in R

.pull-left[

```{r, echo=T, eval=F}
school_reg %>%
  augment() %>%
  summarize(SSE = sum(.resid^2),
            df = n()-1-1,
            SER = sqrt(SSE/df))
```
]

.pull-right[

```{r, echo=F}
school_reg %>%
  augment() %>%
  summarize(SSE = sum(.resid^2),
            df = n()-1-1,
            SER = sqrt(SSE/df))
```
]

--

In large samples (where $n-2$ is very close to $n)$, SER converges to just the standard deviation of the residuals

.pull-left[

```{r, echo=T, eval =F}
school_reg %>%
  augment() %>%
  summarize(sd_resid = sd(.resid))
```
]

.pull-right[
```{r, echo=F}
school_reg %>%
  augment() %>%
  summarize(sd_resid = sd(.resid))
```

]

---

# Goodness of Fit: Looking at R I

.left-code[
- `summary()` command in `Base R` gives:
  - `Multiple R-squared`
  - `Residual standard error` (SER)
    - Calculated with a df of $n-2$

```{r, echo=T, eval=F}
# Base R
summary(school_reg)
```
]

.right-plot[
```{r}
summary(school_reg)
```
]

---

# Goodness of Fit: Looking at R II

```{r, echo=T}
# using broom
library(broom)
glance(school_reg)
```

--

- `r.squared` is `0.05` $\implies$ about 5% of variation in `testscr` is explained by our model
- `sigma` (SER) is `18.6` $\implies$ average test score is about 18.6 points above/below our model's prediction

--

```{r, echo=T}
# extract it if you want with $r.squared
glance(school_reg)$r.squared
```

---

class: inverse, center, middle

# Bias: The Sampling Distributions of the OLS Estimators

---

# Recall: The Two Big Problems with Data

.pull-left[

- We use econometrics to .onfire[identify] causal relationships and make .onfire[inferences] about them

1. Problem for .onfire[identification]: .shout[endogeneity]
    - $X$ is **exogenous** if its variation is *unrelated* to other factors $(u)$ that affect $Y$
    - $X$ is **endogenous** if its variation is *related* to other factors $(u)$ that affect $Y$

2. Problem for .onfire[inference]: .shout[randomness]
    - Data is random due to **natural sampling variation**
    - Taking one sample of a population will yield slightly different information than another sample of the same population

]

.pull-right[
.center[
![:scale 55%](https://www.dropbox.com/s/v5vwsadw5vs448t/causality.jpg?raw=1)

![:scale 55%](https://www.dropbox.com/s/phhet82tcnuegxp/randomimage.jpg?raw=1)]
]

---

# Distributions of the OLS Estimators

- OLS estimators $(\hat{\beta_0}$ and $\hat{\beta_1})$ are computed from a finite (specific) sample of data

- Our OLS model contains **2 sources of randomness**:

--

- .onfire[*Modeled* randomness]: $u$ includes all factors affecting $Y$ *other* than $X$
    - different samples will have different values of those other factors $(u_i)$

--

- .onfire[*Sampling* randomness]: different samples will generate different OLS estimators
    - Thus, $\hat{\beta_0}, \hat{\beta_1}$ are *also* **random variables**, with their own <span class=shout>sampling distribution</span>

---

# Inferential Statistics and Sampling Distributions

.pull-left[

- **Inferential statistics** analyzes a **sample** to make inferences about a much larger (unobservable) **population**

- .shout[Population]: all possible individuals that match some well-defined criterion of interest (people, firms, cities, etc)
  - Characteristics about (relationships between variables describing) populations are called **parameters**

- .shout[Sample]: some portion of the population of interest to *represent the whole*
  - Samples examine part of a population to generate **statistics** used to estimate population parameters

- We almost never can directly study the population, so we *model* it with our samples 
]

.pull-right[
.center[
![](https://www.dropbox.com/s/17cb7ceqipgi8fs/citymodel.jpg?raw=1)
]
]

---

# Sampling Basics

.content-box-green[
.green[**Example**]: Suppose you randomly select 100 people and ask how many hours they spend on the internet each day. You take the mean of your sample, and it comes out to 5.4 hours. 
]

--

- 5.4 hours is a **sample statistic** describing the sample
- We are more interested in the corresponding **parameter**	of the relevant population (e.g. all Americans)

--

- If we take another sample of $n=100$ people, would we get the same number?

--

- Roughly, but probably not exactly

- .shout[Sampling variability] describes the effect of a statistic varying somewhat from sample to sample
    - This is normal, not the result of any error or bias!

---

# I.I.D. Samples

.pull-left[

- If we collect many samples, and each sample is randomly drawn from the population (and then replaced), then the distribution of samples is said to be .onfire[independently and identically distributed (i.i.d.)]

- Each sample is **independent** of each other sample (due to replacement)

- Each sample comes from the **identical** underlying population distribution

]

.pull-right[
.center[
![:scale 90%](https://www.dropbox.com/s/bsdtuddzjouwzr1/sampling.jpg?raw=1)
]
]

---

# The Sampling Distribution of OLS Estimators

.pull-left[
- Calculating OLS estimators for a sample makes the OLS estimators *themselves* random variables:

- Draw of $i$ is random $\implies$ value of each $(X_i,Y_i)$ is random $\implies$ $\hat{\beta_0},\hat{\beta_1}$ are random

- Taking different samples will create different values of $\hat{\beta_0},\hat{\beta_1}$

- Therefore, $\hat{\beta_0},\hat{\beta_1}$ each have a .shout[sampling distribution] across different samples

]

.pull-right[

.center[

![:scale 75%](https://www.dropbox.com/s/phhet82tcnuegxp/randomimage.jpg?raw=1)]

]

---

# The Central Limit Theorem

- .shout[Central Limit Theorem (CLT)]: if we collect samples of size $n$ from the same population and generate a sample statistic (e.g. OLS estimator), then with large enough $n$, the distribution of the sample statistic is approximately normal IF 
    1. $n \geq 30$
    2. Samples come from a *known* normal distribution $\sim N(\mu,\sigma)$

--

- If neither of these are true, we have other methods (coming shortly!)

- One of the most fundamental principles in all of statistics

- Allows for virtually all testing of statistical hypotheses $\rightarrow$ estimating probabilities of values on a normal distribution

---

# The Sampling Distribution of $\hat{\beta_1}$ I

.pull-left[

- The CLT allows us to approximate the sampling distributions of $\hat{\beta_0}$ and $\hat{\beta_1}$ as normal
    - Generally agreed for $n>100$

- We care about $\hat{\beta_1}$ (slope) since it has economic meaning, rarely about $\hat{\beta_0}$ (intercept)

$$\hat{\beta_1} \sim N(E[\hat{\beta_1}], \sigma_{\hat{\beta_1}})$$

]

.pull-right[

```{r, fig.retina=3}
beta_dist<-ggplot(data = tibble(x=-4:4))+
  aes(x = x)+
  stat_function(fun = dnorm, size=2, color="blue")+
  geom_segment(aes(x=0,xend=0, y=0, yend=0.4), linetype="dashed")+
  scale_x_continuous(breaks = 0,
                     labels = expression(E(hat(beta[1]))))+
  labs(x = expression(hat(beta[1])),
       y = "Probability")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
beta_dist
```

]
---

# The Sampling Distribution of $\hat{\beta_1}$ II

.pull-left[

$$\hat{\beta_1} \sim N(E[\hat{\beta_1}], \sigma_{\hat{\beta_1}})$$

- We want to know:

1. $E[\hat{\beta_1}]$; what is the **center** of the distribution? (today)

2. $\sigma_{\hat{\beta_1}}$; how **precise** is our estimate? (next class)

]

.pull-right[

```{r, fig.retina=3}
beta_dist
```

]

---

class: inverse, center, middle
# Bias and Exogeneity

---

# Assumptions about Errors I

.pull-left[

- In order to talk about $E[\hat{\beta_1}]$, we need to talk about $u$

- Recall: $u$ is a random variable, and we can never measure the error term

]

.pull-right[

.center[
![](https://www.dropbox.com/s/95rh9ow982y5htb/error.png?raw=1)
]
]

---

# Assumptions about Errors II

.pull-left[

- We make .shout[4 critical **assumptions** about `\\(u\\)`]:

1. The expected value of the residuals is 0
$$E[u]=0$$

]

.pull-right[

.center[
![](https://www.dropbox.com/s/95rh9ow982y5htb/error.png?raw=1)
]
]

---

# Assumptions about Errors II

.pull-left[

- We make .shout[4 critical **assumptions** about `\\(u\\)`]:

1. The expected value of the residuals is 0
$$E[u]=0$$

]

.pull-right[

.center[
![](https://www.dropbox.com/s/95rh9ow982y5htb/error.png?raw=1)
]
]

---

# Assumptions about Errors II

.pull-left[

- We make .shout[4 critical **assumptions** about `\\(u\\)`]:

1. The expected value of the residuals is 0
$$E[u]=0$$

2. The variance of the residuals is constant, written:
$$var(u)=\sigma^2_{u}$$

]

.pull-right[

.center[
![](https://www.dropbox.com/s/95rh9ow982y5htb/error.png?raw=1)
]
]

---

# Assumptions about Errors II

.pull-left[

- We make .shout[4 critical **assumptions** about `\\(u\\)`]:

1. The expected value of the residuals is 0
$$E[u]=0$$

2. The variance of the residuals is constant, written:
$$var(u)=\sigma^2_{u}$$

3. Errors are not correlated across observations: 
$$cor(u_i,u_j)=0 \quad \forall i \neq j$$

]

.pull-right[

.center[
![](https://www.dropbox.com/s/95rh9ow982y5htb/error.png?raw=1)
]
]

---

# Assumptions about Errors II

.pull-left[

- We make .shout[4 critical **assumptions** about `\\(u\\)`]:

1. The expected value of the residuals is 0
$$E[u]=0$$

2. The variance of the residuals is constant, written:
$$var(u)=\sigma^2_{u}$$

3. Errors are not correlated across observations: 
$$cor(u_i,u_j)=0 \quad \forall i \neq j$$

4. There is no correlation between $X$ and the error term: 
$$cor(X, u)=0 \text{ or } E[u|X]=0$$

]

.pull-right[

.center[
![](https://www.dropbox.com/s/95rh9ow982y5htb/error.png?raw=1)
]
]

---

# Assumptions 1 and 2: Errors are i.i.d.

.pull-left[

1. The expected value of the residuals is 0
$$E[u]=0$$

2. The variance of the residuals is constant, written:
$$var(u)=\sigma^2_{u}$$

- The first two assumptions $\implies$ errors are **i.i.d.**, drawn from the same distribution with mean 0 and variance $\sigma^2_{u}$

]

.pull-right[

.center[
![](https://www.dropbox.com/s/95rh9ow982y5htb/error.png?raw=1)
]
]

---

# Assumption 2: Heteroskedasticity

.pull-left[

- The variance of the residuals is constant, written:
$$var(u)=\sigma^2_{u}$$

- Assumption 2 implies that errors are ".shout[homoskedastic]:
 they have the same variance across $X$

- Often this assumption is violated: errors may be ".shout[heteroskedastic]:
 they do not have the same variance across $X$

- This *is* a problem for **inference**, but we have a simple fix for this (coming shortly)

]

.pull-right[

.center[
![](https://www.dropbox.com/s/95rh9ow982y5htb/error.png?raw=1)
]
]

---

# Assumption 3: No Serial Correlation

.pull-left[

- Errors are not correlated across observations: 
$$cor(u_i,u_j)=0 \quad \forall i \neq j$$

- For simple cross-sectional data, this is rarely an issue

- Time-series & panel data nearly always contain .shout[serial correlation] or .shout[autocorrelation] between errors

- e.g. "this week's sales look a lot like last weel's sales, which look like...etc"

- There are fixes to deal with autocorrelation (coming much later)

]

.pull-right[

.center[
![](https://www.dropbox.com/s/95rh9ow982y5htb/error.png?raw=1)
]
]

---

# Assumption 4: The Zero Conditional Mean Assumption

.pull-left[

- There is no correlation between $X$ and the error term: 
$$cor(X, u)=0$$

- .onfire[This is the absolute killer assumption, because it assumes **exogeneity**]

- Often called the .shout[Zero Conditional Mean] assumption: 
$$E[u|X]=0$$

- "Does knowing $X$ give me any useful information about $u$?"
    - If yes: model is **endogenous**, **biased** and **not-causal**!

]

.pull-right[

.center[
![](https://www.dropbox.com/s/95rh9ow982y5htb/error.png?raw=1)
]
]

---

# Exogeneity and Unbiasedness

- $\hat{\beta_1}$ is .onfire[unbiased] iff no systematic difference, on average, between sample values of $\hat{\beta_1}$ and true population $\beta_1$, i.e.

$$E[\hat{\beta_1}]=\beta_1$$

--

- Does *not* mean any sample gives us $\hat{\beta_1}=\beta_1$, only the **estimation procedure** will, *on average*, yield the correct value

- Random errors above and below the true value cancel out (so that on average, $E[\hat{u}|X]=0)$

---

# Sidenote: Estimators of Statistics I

- In statistics, an .shout[estimator] is simply a rule that for calculating a statistic (often about a wider population parameter)

--

.content-box-green[
.green[**Example**]:
We want to estimate the average height (H) of U.S. adults (population) and have a random sample of 100 adults.]

--

- Calculate the mean height of our sample $(\bar{H})$ to estimate the true mean height of the population $(\mu_H)$

- $\bar{H}$ is an **estimator** of $\mu_H$

--

- There are many estimators we *could* use to estimate $\mu_H$
    - How about using the first value in our sample: $H_1$

---

# Sidenote: Estimators of Statistics II

.pull-left[

- What makes one estimator (e.g. $\bar{H}$) better than another (e.g. $H_1$)?<sup>.red[1]</sup>

1. .onfire[Biasedness]: does the estimator give us the true parameter *on average*?

2. .onfire[Efficiency]: an estimator with a smaller variance is better
]

.pull-right[
.center[
![](https://www.dropbox.com/s/3adddurpkp2k22o/biasvariability.png?raw=1)
]
]

.footnote[<sup>.red[1]</sup> Technically, we also care about <span class="onfire">consistency</span>: minimizing uncertainty about the correct value. The Law of Large Numbers, similar to CLT, permits this. We don't need to get too advanced about probability in this class.]

---

# Exogeneity and Unbiasedness I

.pull-left[

- $\mathbf{\hat{\beta_1}}$ is the .shout[Best Linear Unbiased Estimator (BLUE)] estimator of  $\mathbf{\beta_1}$ *when $X$ is exogenous*<sup>.red[1]</sup>

- No systematic difference, on average, between sample values of $\hat{\beta_1}$ and the true population $\beta_1$:

$$E[\hat{\beta_1}]=\beta_1$$

- Does *not* mean that each sample gives us $\hat{\beta_1}=\beta_1$, only the estimation *procedure* will, on average, yield the correct value

]

.pull-right[
.center[
![](https://www.dropbox.com/s/3adddurpkp2k22o/biasvariability.png?raw=1)
]
]

.footnote[<sup>.red[1]</sup> The proof for this is known as the [Gauss-Markov Theorem](https://en.wikipedia.org/wiki/Gauss%E2%80%93Markov_theorem). See today's [class notes](/class/10-class) for a simplified proof.]

---

# Exogeneity and Unbiasedness II

- Recall, an .shout[exogenous] variable $(X)$ is unrelated to other factors affecting $Y$, i.e.:

$$cor(X,u)=0$$

--

- Again, this is called the **Zero Conditional Mean Assumption**

$$E(u|X)=0$$

--

- For any known value of $X$, the expected value of $u$ is 0

- Knowing the value of $X$ must tell us *nothing* about the value of $u$ (anything else relevant to $Y$ other than $X$)

- We can then confidently assert causation: $X \rightarrow Y$

---

# Endogeneity and Bias

- Nearly all independent variables are .shout[endogenous], they **are** related to the error term $u$
$$cor(X,u)\neq 0$$

--

.content-box-green[
.green[**Example**]: Suppose we estimate the following relationship:

$$\text{Violent crimes}_t=\beta_0+\beta_1\text{Ice cream sales}_t+u_t$$

- We find $\hat{\beta_1}>0$

- Does this mean Ice cream sales$\rightarrow$Violent crimes?
]

---

# Endogeneity and Bias III

- The true expected value of $\hat{\beta_1}$ is actually:<sup>.red[1]</sup>

$$E[\hat{\beta_1}]=\beta_1+cor(X,u)\frac{\sigma_u}{\sigma_X}$$

--

- Takeaways:

--

1. If $X$ is exogenous: $cor(X,u)=0$, we're just left with $\beta_1$

--

2. The larger $cor(X,u)$ is, larger .onfire[bias]: $\left(E[\hat{\beta_1}]-\beta_1 \right)$

--

3. We can "sign" the direction of the bias based on $cor(X,u)$
    - **Positive** $cor(X,u)$ overestimates the true $\beta_1$ $(\hat{\beta_1}$ is too high)
    - **Negative** $cor(X,u)$ underestimates the true $\beta_1$ $(\hat{\beta_1}$ is too low)

.footnote[<sup>.red[1]</sup> See today's [class notes](/class/10-class) for proof.]

---

# Endogeneity and Bias: Example I

.content-box-green[
.green[**Example**]: 
$$wages_i=\beta_0+\beta_1 education_i+u$$

- Is this an accurate reflection of $educ \rightarrow wages$?

- Does $E[u|education]=0$?

- What would $E[u|education]>0$ mean?

]

---

# Endogeneity and Bias: Example II

.content-box-green[
.green[**Example**]: 
$$\text{per capita cigarette consumption}=\beta_0+\beta_1 \text{State cig tax rate}+u 	$$

- Is this an accurate reflection of $tax \rightarrow cons$?

- Does $E[u|tax]=0$?

- What would $E[u|tax]>0$ mean?

]

---

# Exogeneity and RCTs

.pull-left[

- Think about an idealized randomized controlled experiment

- Subjects randomly assigned to treatment or control group
    - Implies knowing whether someone is treated $(X)$ tells us nothing about their personal characteristics $(u)$

    - Random assignment makes $u$ independent of $X$, so

$$cor(X,u)=0\text{ and } E[u|X]=0$$

]

.pull-right[
.center[
![:scale 80%](https://www.dropbox.com/s/wqjisnpxaoaa2yy/rct.png?raw=1)
]

]

### 2019-09-26 content/class/10-class.Rmd
Update class 10; tweak 10 slides

@@
 
-Today we start looking at *associations* between variables, which we will first attempt to quantify with measures like *covariance* and *correlation*.
-Then we turn to fitting a line to data via *linear regression*.
-We overview the basic regression model, the parameters and how they are derived, and see how to work with regressions in `R` with `lm` and the tidyverse package [`broom`](https://broom.tidyverse.org/).
+Today we continue looking at basic OLS regression.
+We will cover how to measure if a regression line is a good fit (using $R^2$ and $\sigma_u$ or SER), and whether OLS estimators are *biased*.
+These will depend on four critical *assumptions about $u$.*
 
-We consider an extended example about class sizes and test scores, which comes from a (Stata) dataset from an old textbook that I used to use, [Stock and Watson, 2007](https://www.pearson.com/us/higher-education/product/Stock-Introduction-to-Econometrics-Update-3rd-Edition/9780133486872.html?tab=resources).
+In doing so, we begin an ongoing exploration into inferential statistics, which will finally become clear in another week.
+The most confusing part is recognizing that there is a *sampling distribution of each OLS estimator*.
+We want to measure the center of that sampling distribution, to see if the estimator is *biased*.
+Next class we will measure the spread of that distribution.
+
+We continue the extended example about class sizes and test scores, which comes from a (Stata) dataset from an old textbook that I used to use, [Stock and Watson, 2007](https://www.pearson.com/us/higher-education/product/Stock-Introduction-to-Econometrics-Update-3rd-Edition/9780133486872.html?tab=resources).
 Download and follow along with the data from today's example:^[Note this is a `.dta` Stata file.
@@
 
-- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/09-slides.html)
+- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/10-slides.html)
 
-# Assignments: Problem Set 1 DUE and Problem Set 2
+# Assignments: Problem Set 2 DUE
 
-[Problem Set 1](/assignment/01-problem-set.html) answers will be posted shortly, and the graded assignment emailed back to you.^[Sorry I've been too busy to send emails confirming that I recieved your problem set!]
-
-[Problem Set 2](/assignment/01-problem-set.html) (on classes 2.1-2.2) is due by Thursday September 26.
+[Problem Set 2](/assignment/01-problem-set.html) is DUE. Answers will be posted later today.
 

### 2019-09-26 static/slides/10-slides.Rmd
Update class 10; tweak 10 slides

@@
   augment() %>%
-  summarize(r_sq = var(.fitted)/var(testscr)) # variance of *predicted* testscr over variance of *actual* teestscr
+  summarize(r_sq = var(.fitted)/var(testscr)) # var. of *predicted* testscr over var. of *actual* testscr
 ```
@@
 - The CLT allows us to approximate the sampling distributions of $\hat{\beta_0}$ and $\hat{\beta_1}$ as normal
-    - Generally agreed for $n>100$
 

### 2019-09-27 content/class/10-class.Rmd
Add problem set 2 answers, update reference to include class appendices, fix class 10 appendix errors

@@
 $$\begin{align*}
-	 corX,u)&=\frac{cov(X,u)}{s_X s_u}\\
-	 corX,u)s_Xs_u &=cov(X,u)\\
+	 cor(X,u)&=\frac{cov(X,u)}{s_X s_u}\\
+	 cor(X,u)s_Xs_u &=cov(X,u)\\
 \end{align*}$$
@@
 E[\hat{\beta_1}]&=\beta_1+\frac{cov(X,u)}{var(X)} \\
-E[\hat{\beta_1}]&=\beta_1+\frac{\big[corX,u)s_xs_u\big]}{s^2_X} \\
-E[\hat{\beta_1}]&=\beta_1+\frac{corX,u)s_u}{s_X} \\
-E[\hat{\beta_1}]&=\beta_1+corX,u)\frac{s_u}{s_X} \\
+E[\hat{\beta_1}]&=\beta_1+\frac{\big[cor(X,u)s_xs_u\big]}{s^2_X} \\
+E[\hat{\beta_1}]&=\beta_1+\frac{cor(X,u)s_u}{s_X} \\
+E[\hat{\beta_1}]&=\beta_1+cor(X,u)\frac{s_u}{s_X} \\
 \end{align*}$$
@@
 
-Begin with equation:
+Begin with equation:^[Admittedly, this is a simplified version where $\hat{\beta_0}=0$, but there is no loss of generality in the results.]
 

### 2019-09-27 content/reference.Rmd
Add class 3 content
Add problem set 2 answers, update reference to include class appendices, fix class 10 appendix errors

@@
 
+# List of Math and R Appendices
+
+The following list contains links to all appendices spread across the Class Notes pages:
+
+- **R**: [Installing R and R Studio](https://metricsf19.classes.ryansafner.com/class/03-class/#installing-r-and-r-studio) (Class 1.3)
+- **R**: [R Packages](https://metricsf19.classes.ryansafner.com/class/03-class/#r-packages) (Class 1.3)
+- **R**: [Getting Help for R](https://metricsf19.classes.ryansafner.com/class/03-class/#getting-help-for-r) (Class 1.3)
+- **R**: [Other Useful Commands to Know](https://metricsf19.classes.ryansafner.com/class/03-class/#other-useful-commands-to-know) (Class 1.3)
+- **R**: [Style Guide for Coding](https://metricsf19.classes.ryansafner.com/class/03-class/#style-guide-for-coding) (Class 1.3)
+- **Math**: [The Summation Operator](https://metricsf19.classes.ryansafner.com/class/07-class/#the-summation-operator) (Class 2.1)
+- **R**: [Creating Mathematical Functions](https://metricsf19.classes.ryansafner.com/class/08-class/#creating-mathematical-functions) (Class 2.2)
+- **R**: [Graphing Mathematical Functions](https://metricsf19.classes.ryansafner.com/class/08-class/#graphing-mathematical-functions) (Class 2.2)
+- **R**: [Bultin Statistical Functions](https://metricsf19.classes.ryansafner.com/class/08-class/#bultin-statistical-functions) (Class 2.2)
+- **R**: [Graphing Statistical Functions](https://metricsf19.classes.ryansafner.com/class/08-class/#graphing-statistical-functions) (Class 2.2)
+- **Math**: [Variance](https://metricsf19.classes.ryansafner.com/class/09-class/#variance) (Class 2.3)
+- **Math**: [Covariance](https://metricsf19.classes.ryansafner.com/class/09-class/#covariance) (Class 2.3)
+- **Math**: [Correlation](https://metricsf19.classes.ryansafner.com/class/09-class/#correlation) (Class 2.3)
+- **R**: [Calculating Correlation Example](https://metricsf19.classes.ryansafner.com/class/09-class/#calculating-correlation-example) (Class 2.3)
+- **Math**: [Deriving the OLS Estimators](https://metricsf19.classes.ryansafner.com/class/10-class/#deriving-the-ols-estimators) (Class 2.4)
+- **Math**: [Algebraic Properties of OLS Estimators](https://metricsf19.classes.ryansafner.com/class/10-class/#algebraic-properties-of-ols-estimators) (Class 2.4)
+- **Math**: [Bias in $\hat{\beta_1}$](https://metricsf19.classes.ryansafner.com/class/10-class/#bias-in-hatbeta_1) (Class 2.4)
+- **Math**: [Proof of the Unbiasedness of  $\hat{\beta_1}$](https://metricsf19.classes.ryansafner.com/class/10-class/#proof-of-the-unbiasedness-of-hatbeta_1) (Class 2.4)
+
 # Installing R and R Studio
@@
 Additionally, you can turn to the community.
+
+# Built in Datasets
 

### 2019-09-27 static/homeworks/02-problem-set-answers.Rmd
Add problem set 2 answers, update reference to include class appendices, fix class 10 appendix errors

--- 2019-09-20 static/homeworks/02-problem-set-pdf.Rmd
@@
 ---
-title: "Problem Set 2"
+title: "Problem Set 2 (Answers)"
 author: "Ryan Safner"
 date: "ECON 480 - Fall 2019"
-output: pdf_document
+output:
+  html_document:
+    df_print: paged
+    theme: simplex
+    toc: true 
+    toc_depth: 3
+    toc_float: true
+    code_folding: show
+    highlight: tango
 ---
 
-*Due by Thursday, September 26, 2019*
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE,
+                      message = FALSE,
+                      warning = FALSE)
+```
 
-# Instructions
-
-For this problem set, you may submit handwritten answers on a plain sheet of paper, or download and type/handwrite on the PDF.
-
-Alternatively, you may download the `.Rmd` file, do the homework in markdown, and email to me a single `knit`ted `html` or `pdf` file (and be sure that it shows all of your code).
-
-You may work together (and I highly encourage that) but you must turn in your own answers.
-I grade homeworks 70% for completion, and for the remaining 30%, pick one question to grade for accuracy - so it is best that you try every problem, even if you are unsure how to complete it accurately.
+Answers may be longer than I would deem sufficient on an exam.
+Some might vary slightly based on points of interest, examples, or personal experience.
+These suggested answers are designed to give you both the answer and a short explanation of *why* it is the answer.
 
@@
 
-1.
+## Question 1
 
-a. In your own words, explain the difference between endogeneity and exogeneity.^[Hint: you may want to look back at class 1.2 for this question.]
-b. In your own words, explain how conducting a randomized controlled trial helps to identify the causal connection between two variables.
+### Part A
 
-\clearpage
+**In your own words, explain the difference between endogeneity and exogeneity.**^[Hint: you may want to look back at class 1.2 for this question.]
 
-2.
+An **exogenous** model is one where the independent variable $(X)$ is not associated with any other factors that affect the dependent variable $(Y)$.
+If a model is truly exogenous, we can estimate the **causal effect** of $X$ on $Y$.
 
-a. In your own words, explain what (sample) standard deviation *means*.
-b. In your own words, explain how (sample) standard deviation *is calculated.* You may also write the formula, but it is not necessary.
+An **endogenous** model is one where the independent variable $(X)$ *is* associated with any other factors that affect the dependent variable $(Y)$.
+If a model is endogenous, we have no accurately estimated the causal effect of $X$ on $Y$, since other factors are getting entangled with $X$ and $Y$.
 
-\clearpage
+### Part B
+
+**In your own words, explain how conducting a randomized controlled trial helps to identify the causal connection between two variables.**
+
+Randomized controlled experiments are where a pool of subjects representative of a population are randomly assigned into a treatment group (or into one of a number of groups given different levels of a treatment) or into a control group.
+The treatment group(s) is(are) given the treatment(s), the control group is given nothing (perhaps a placebo, though this is not always necessary), and then the average results of the two groups are compared to measure the true average effect of treatment.
+
+The key is that the assignment must be random,  which controls for all factors that potentially determine the outcome (e.g.
+when measuring individual outcomes, their height, family background, income, race, age, etc).
+If subjects are randomly assigned, then knowing anything about the individual (e.g.
+age, height, etc) tells us *nothing* about whether or not they got the treatment(s).
+The *only* thing that separates a member of the treatment group(s) from the control group is whether or not they were assigned to treatment.
+This ensures that the average person in the treatment group(s) looks like the average person in the control group, and that we are truly comparing apples to apples, rather than apples to oranges.
+
+## Question 2
+
+### Part A
+
+**In your own words, explain what (sample) standard deviation *means*.**
+
+Sample standard deviation measures the average deviation (distance) of any given value of a variable from the variable's mean.
+
+### Part B
+
+**In your own words, explain how (sample) standard deviation *is calculated.* You may also write the formula, but it is not necessary.**
+
+The formula is $$sd(X)=\sqrt{\frac{\displaystyle \sum^n_{i=1} (x_i-\bar{X})^2}{n-1}}$$
+
+It helps to consider the calculation as a series of steps:
+
+1. Find the mean of $X$, $(\bar{X})$
+2. Subtract the mean from each value of $X$ in the data, to get deviations, $(x_i-\bar{X})$
+3. Square the deviations to ensure they are all positive, $(x_i-\bar{X})^2$
+4. Take the average of the squared deviations: add them all up and divide by $n-1$, $\frac{1}{n}\displaystyle \sum^n_{i=1} (x_i-\bar{X})^2$
+5.
+What you have in step 4 is variance (measured in units of $X^2)$, square root to get standard deviation (measured in original units of $X$): $\sqrt{\frac{1}{n}\displaystyle \sum^n_{i=1} (x_i-\bar{X})^2}$
+
+Note on Step 4: because this is a *sample*, we have do deal with **degrees of freedom (df)** loss.
+We use up one df to calculate the mean, $\bar{x}$, which is needed before calculating variance or standard deviation.
+Hence, instead of averaging like normal: $\frac{1}{n} \sum x_i$, we need to divide by $n-1$.
 
@@
 
-For the remaining questions, you may use `R` to *verify*, but please calculate all sample statistics by hand and show all work.
+## Question 3
 
-3. Suppose you have a very small class of four students that all take a quiz. Their scores are reported as follows:
+Suppose you have a very small class of four students that all take a quiz. Their scores are reported as follows:
 
@@
 
-a. Calculate the median.
-b. Calculate the sample mean, $\bar{x}$.
-c. Calculate the sample standard deviation, $s$.
-d.
-Make or sketch a rough histogram of this data, with the size of each bin being 10 (i.e.
+For the remaining questions, you may use `R` to *verify*, but please calculate all sample statistics by hand and show all work.
+
+## Part A 
+**Calculate the median.**
+
+Arrange the values in numerical order from smallest to largest.
+Find the value in the middle (i.e.
+an equal number of values are on either side); possibly by crossing-out one number on either side at a time (like in Elementary School).
+
+$$\underline{72}, 81, 83, \underline{92}$$
+
+Since we have an even number of observations, we have two numbers in the middle, $81$ and $83$, so we must take the average of them:
+
+$$\frac{81+83}{2}=82$$
+
+Using R:
+
+```{r}
+median(c(83,92,72,81))
+```
+
+## Part B
+**Calculate the sample mean, $\bar{x}$.**
+
+$$\begin{align*}\bar{x}&=\frac{1}{n} \sum^n_{i=1} x_i\\
+\bar{x}&=\frac{72+81+83+92}{4}\\
+\bar{x}&=\frac{328}{4}\\
+\bar{x}&=82\\
+\end{align*}$$
+
+Using R:
+
+```{r}
+mean(c(83,92,72,81))
+```
+
+## Part C
+
+**Calculate the sample standard deviation, $s$.**
+
+My suggestion is to use the "table" method, and follow the 5 steps described in problem 2b.
+
+| $x_i$ | $x_i-\bar{X}$ | $(x_i-\bar{X})^2$ |
+|-------|--------------:|------------------:|
+| $72$  | $-10$         | $100$ |
+| $81$  | $-1$          | $1$   |
+| $83$  | $1$           | $1$   |
+| $92$  | $10$          | $100$ |
+| $\sum$ |              | $202$ | 
+| $\frac{1}{3} \times \sum$ |  | $\approx 67.33$ | 
+| $\mathbf{\sqrt{\frac{1}{3} \times \sum}}$ | | $\approx \mathbf{8.21}$ |
+
+In R:
+
+```{r}
+sd(c(83,92,72,81))
+```
+
+## Part D
+**Make or sketch a rough histogram of this data, with the size of each bin being 10 (i.e.
 70's, 80's, 90's, 100's).
@@
 by `10`s.).] Is this distribution roughly symmetric or skewed?
-What would we expect about the mean and the median?
-e. Suppose instead the person who got the 72 did not show up that day to class, and got a 0 instead. Recalculate the mean and median. What happened and why?
+What would we expect about the mean and the median?**
 
-\clearpage
+```{r}
+# load tidyverse (for tibble and ggplot2)
+library(tidyverse)
 
-4. Suppose the probabilities of a visitor to Amazon’s website buying 0, 1, or 2 books are 0.2, 0.4, and 0.4 respectively.
+# make a dataframe of our data, 
+# called df
+# one variable in it, called quiz
 
-a. Calculate the *expected number* of books a visitor will purchase.
-b. Calculate the *standard deviation* of book purchases.
-c. **Bonus**: try doing this in `R` by making an initial dataframe of the data, and then making new columns to the "table" like we did in class.
+df <- tibble(quiz = c(83,92,72,81))
 
-\clearpage
+# use this as our data for plot
+ggplot(data = df)+
+  aes(x = quiz)+
+  geom_histogram(breaks=seq(0,100,10), # make bins of size 10 between 0 and 100
+                 color = "white", # color is for borders
+                 fill = "blue")+ # fill is for area
+  scale_x_continuous(breaks=seq(0,100,10))+ # have x axix ticks same as breaks 
+  theme_classic()
+```
 
-5. Scores on the SAT (out of 1600) are approximately normally distributed with a mean of 500 and standard deviation of 100.
+We can see it is roughly symmetric. We would therefore expect the mean and the median to be approximately the same (which parts A and B showed was true).
 
-a. What is the probability of getting a score between a 400 and a 600?
-b. What is the probability of getting a score between a 300 and a 700?
-c. What is the probability of getting *at least* a 700?
-d. What is the probability of getting *at most* a 700?
-e. What is the probability of getting exactly a 500?
+## Part E
 
-\clearpage
+**Suppose instead the person who got the 72 did not show up that day to class, and got a 0 instead. Recalculate the mean and median. What happened and why?**
 
-6.
+$$\underline{0}, 81, 83, \underline{92}$$
+
+$$\frac{81+83}{2}=82$$
+
+Replacing the 72 with a 0, and keeping the same number of observations does not change the median!
+
+$$\begin{align*}\bar{x}&=\frac{1}{n} \sum^n_{i=1} x_i\\
+\bar{x}&=\frac{0+81+83+92}{4}\\
+\bar{x}&=\frac{256}{4}\\
+\bar{x}&=64\\
+\end{align*}$$
+
+The mean is pulled down significantly by the outlier.
+
+In R: 
+```{r}
+mean(c(83,92,0,81))
+median(c(83,92,0,81))
+```
+
+If we were to look at the histogram, it would be skewed, and the mean would be lower than the mean:
+
+```{r}
+# make new tibble called df_2
+df_2 <- tibble(quiz = c(83,92,0,81)) # replace 72 with 0
+
+# use this as our data for plot
+ggplot(data = df_2)+
+  aes(x = quiz)+
+  geom_histogram(breaks=seq(0,100,10), # make bins of size 10 between 0 and 100
+                 color = "white", # color is for borders
+                 fill = "blue")+ # fill is for area
+  geom_vline(aes(xintercept = median(quiz)), size = 1, color = "green", linetype = "dashed")+ # green dashed line is median
+  geom_label(aes(x = median(quiz), y = 1), label = "Median", color = "green")+ # label median line
+  geom_vline(aes(xintercept = mean(quiz)), size = 1, color = "red", linetype = "dotted")+ # green dashed line is mean
+  geom_label(aes(x = mean(quiz), y = 1), label = "Mean", color = "red")+ # label mean line
+  scale_x_continuous(breaks=seq(0,100,10))+ # have x axix ticks same as breaks 
+  theme_classic()
+```
+
+## Question 4
+
+Suppose the probabilities of a visitor to Amazon’s website buying 0, 1, or 2 books are 0.2, 0.4, and 0.4 respectively.
+
+### Part A
+**Calculate the *expected number* of books a visitor will purchase.**
+
+Define $X$ to be the number of books a visitor to Amazon's website purchases. The $pdf$ of $X$ is as follows:
+
+| $x_i$ | $P(X=x_i)$ |
+|-------|------------|
+| 0     | 0.20       |
+| 1     | 0.40       |
+| 2     | 0.40       |
+
+The expected value of $X$ is the probability weighted average of $X$:
+
+$$\begin{align*}E(X)&=\sum^n_{i=1} p_ix_i\\
+&=0.2(0)+(0.4)1+(0.4)2\\
+&=0+0.4+0.8\\
+&=1.2\\
+\end{align*}$$
+
+### Part B
+
+**Calculate the *standard deviation* of book purchases.**
+
+The formula(s) for standard deviation of a random variable is:
+
+$$\sigma_X = sd(X) = \sqrt{E[(X-E[X])^2]} = \sqrt{\sum^n_{i=1}p_i(x_i-E[X])^2}$$
+
+I suggest using the table method, again. Working from the inside out of the formula, the steps are:
+
+1. Find the expected value of $X$, $E[X]$.
+2. Subtract the expected value from each value of $X$ in the data, to get deviations, $(x_i-E[X])$
+3. Square the deviations to ensure they are all positive, $(x_i-E[X])^2$
+4.
+Take the probability-weighted average of the squared deviations: multiply each squared deviation by the probability of its associated $x$ value and add them all up $\displaystyle \sum^n_{i=1} p_i(x_i-E[X])^2$
+5. What you have in step 4 is variance (measured in units of $X^2)$, square root to get standard deviation (measured in original units of $X$): $\sqrt{\displaystyle \sum^n_{i=1} p_i(x_i-E[X])^2}$
+
+| $x_i$ | $P(X=x_i)$ | $x_i-E[X]$ | $(x_i-E[X])^2$ | $p_i(x_i-E[X])^2$ |
+|-------|------------|------------|----------------|-------------------|
+| 0     | 0.20       | $-1.20$    | $1.44$         | $0.288$           |
+| 1     | 0.40       | $-0.20$    | $0.04$         | $0.016$           |
+| 2     | 0.40       | $0.80$     | $0.64$         | $0.256$           |
+| $\sum$ |  |  |  | $0.560$ |
+| $sqrt{\sum}$| | | | $0.748$ |
+
+### Part C
+**Bonus: try doing this in `R` by making an initial dataframe of the data, and then making new columns to the "table" like we did in class.**
+
+```{r}
+# make a dataframe called "amazon" of # of books and associated probabilities
+amazon<-tibble(books = c(0,1,2),
+               prob = c(0.2,0.4,0.4))
+
+# look at it 
+amazon
+
+# find expected value
+amazon %>%
+  summarize(exp_value = sum(books*prob))
+
+# it's 1.2, let's save exp_value
+
+exp_value <- 1.2
+
+# make new columns: devs, devs_sq, p_weight_devs_sq
+# save to new tibble
+amazon_table<-amazon %>%
+  mutate(devs = books - exp_value,
+         devs_sq = devs^2,
+         p_weight_devs_sq = prob*devs^2)
+
+# look at the tibble
+amazon_table
+
+# now let's take these and summarize 
+amazon_table %>%
+  summarize(var = sum(p_weight_devs_sq), # variance
+            sd = sqrt(var)) # sqrt to get sd, confirm its same!
+```
+
+## Question 5
+
+Scores on the SAT (out of 1600) are approximately normally distributed with a mean of 500 and standard deviation of 100.
+
+### Part A
+
+**What is the probability of getting a score between a 400 and a 600?**
+
+Let random variable $S$ be the score earned on the SAT.
+
+Convert thse numbers to $Z$-scores.
+
+$$\begin{align*}P(400  \leq  S  \leq 600)&=P\big(\frac{400-500}{100} \leq \frac{S - 500}{100} \leq \frac{600-500}{100}\big)\\
+&=P\big(-1  \leq  Z  \leq 1\big)\\
+&\approx 0.68\\
+\end{align*}$$
+
+Using the 68-95-99.7 rule: about 68% of the values fall within one standard deviation $(\pm 1$ $Z$-score) of the mean.
+
+You don't need to draw the pdf, but it helps to visualize what we're looking for, and how converting to $Z$-scores helps:
+
+```{r, fig.height=3, fig.retina=3, fig.align="center"}
+# see class 2.3 notes on how to graph and shade stats graphs
+
+# it helps to first figure out where the x-axis ticks should be
+# show about 4 standard deviations above and below the mean (mu +/- 4*sd)
+# then have ticks in intervals of one sd
+
+# in this case, with mean 500 and sd 100, it should be seq(100,900,100)
+
+s_plot<-ggplot(data = tibble(scores=seq(from = 100,
+                           to = 900,
+                           by = 100)))+
+  aes(x = scores)+
+  stat_function(fun = dnorm,
+                args = list(mean = 500, sd = 100),
+                size = 2, color = "blue")+
+  labs(x = "SAT Scores (out of 1600)",
+       y = "Probability")+
+  scale_x_continuous(breaks=seq(from  = 100,
+                           to = 900,
+                           by = 100))+
+  theme_classic(base_family = "Fira Sans Condensed",
+           base_size=20)
+
+s_plot+stat_function(fun = dnorm,
+                args = list(mean = 500, sd = 100),
+                geom = "area",
+                xlim = c(400,600),
+                size = 2, fill = "blue", alpha = 0.5)
+```
+
+```{r, fig.height=3, fig.retina=3, fig.align="center"}
+Z<-ggplot(data = tibble(Z=seq(from = -4,
+                           to = 4,
+                           by = 1)))+
+  aes(x = Z)+
+  stat_function(fun = dnorm,
+                size = 2, color = "blue")+
+  labs(x = "Z-Scores",
+       y = "Probability")+
+  scale_x_continuous(breaks=seq(from  = -4,
+                           to = 4,
+                           by = 1))+
+  theme_classic(base_family = "Fira Sans Condensed",
+           base_size=20)
+
+Z+stat_function(fun = dnorm,
+                geom = "area",
+                xlim = c(-1,1),
+                size = 2, fill = "blue", alpha = 0.5)
+```
+
+### Part B
+
+**What is the probability of getting a score between a 300 and a 700?**
+
+Convert thse numbers to $Z$-scores.
+
+$$\begin{align*}P(300  \leq  S  \leq 700)&=P\big(\frac{300-500}{100} \leq \frac{S - 500}{100} \leq \frac{700-500}{100}\big)\\
+&=P\big(-2  \leq  Z  \leq 2\big)\\
+&\approx 0.95\\
+\end{align*}$$
+
+Using the 68-95-99.7 rule: about 95% of the values fall within two standard deviations $(\pm 2$ $Z$-score) of the mean.
+
+```{r, fig.height=3, fig.retina=3, fig.align="center"}
+s_plot+stat_function(fun = dnorm, args = list(mean = 500, sd = 100), geom = "area", xlim = c(300,700), size = 2, fill = "blue", alpha = 0.5)
+
+Z+stat_function(fun = dnorm, geom = "area", xlim = c(-2,2), size = 2, fill = "blue", alpha = 0.5)
+```
+
+### Part C
+
+**What is the probability of getting *at least* a 700?**
+
+We saw in part B that $Z$ for 700 is 2.
+
+Using the 68-95-99.7 rule, we know about 95% of the values fall within two standard deviations $(\pm 2$ $Z$-score) of the mean.
+That means that 5% of values fall *beyond* a $Z$ score of $\pm 2$, or 2.5% in each direction.
+We only want the right-tail (probability above $Z=2$, so 2.5%).
+
+```{r, fig.height=3, fig.retina=3, fig.align="center"}
+s_plot+stat_function(fun = dnorm, args = list(mean = 500, sd = 100), geom = "area", xlim = c(700,900), size = 2, fill = "blue", alpha = 0.5)
+
+Z+stat_function(fun = dnorm, geom = "area", xlim = c(2,4), size = 2, fill = "blue", alpha = 0.5)
+```
+
+### Part D
+
+**What is the probability of getting *at most* a 700?**
+
+We saw in Part C that the area to the right of $Z=2$ is 2.5%, so the remaining 97.5% falls to the left of $Z=2$.
+
+```{r, fig.height=3, fig.retina=3, fig.align="center"}
+s_plot+stat_function(fun = dnorm, args = list(mean = 500, sd = 100), geom = "area", xlim = c(100,700), size = 2, fill = "blue", alpha = 0.5)
+
+Z+stat_function(fun = dnorm, geom = "area", xlim = c(-4,2), size = 2, fill = "blue", alpha = 0.5)
+```
+
+### Part E
+
+**What is the probability of getting exactly a 500?**
+
+This is a trick question! For a continuous random variable, the probability of any one specific value is 0.
+
+## Question 6
+
 Redo problem 5 by using the `pnorm()` command in `R`.^[Hint: This function has four arguments: 1.
@@
 `lower.tail` `TRUE` or `FALSE`.]
+
+## Part A
+
+Look back to the graphs of the pdfs in Question 5 to visualize what we are looking for.
+
+`pnorm` converts $X$ to $Z$ and takes the **standard normal cdf**, $\Phi$ of a variable, i.e.
+
+$$\Phi(k)=P(Z \geq k)$$
+
+That means, it calculates the probability of everything up to (to the *left* of) that value on the **pdf.** So, if you want to calculate the area between values $j$ and $k$, you need to take the cdf of the larger number and subtract the cdf of the smaller number:
+
+$$P(j \leq Z \leq k) = \Phi(k)-\Phi(j)$$
+
+In this case (in $Z$-scores):
+
+$$P(-1 \leq Z \leq 1) = \Phi(1)-\Phi(-1)$$
+
+```{r}
+pnorm(600, mean = 500, sd = 100, lower.tail = TRUE) - pnorm(400, mean = 500, sd = 100, lower.tail = TRUE)
+```
+
+---
+
+## Part B
+
+In this case (in $Z$-scores):
+
+$$P(-2 \leq Z \leq 2) = \Phi(2)-\Phi(-2)$$
+
+```{r}
+pnorm(700, mean = 500, sd = 100, lower.tail = TRUE) - pnorm(300, mean = 500, sd = 100, lower.tail = TRUE)
+```
+
+---
+
+## Part C
+
+In this case (in $Z$-scores):
+
+$$P(Z \geq 2) = 1-P(Z \leq 2) = 1- \Phi(2)$$
+
+```{r}
+1- pnorm(700, mean = 500, sd = 100, lower.tail = TRUE)
+```
+
+---
+
+## Part D
+
+In this case (in $Z$-scores):
+
+$$P(Z \leq 2) = \Phi(2)$$
+
+```{r}
+pnorm(700, mean = 500, sd = 100, lower.tail = TRUE)
+```
+
+We can see that the 68-95-99.7 rule is *close*, but not *exactly* equal to the true probabilities of $Z$. As such, it's just a good rule of thumb!
+
+## Part E
+
+Again, the probability is 0, no need to do anything here.

### 2019-09-27 static/homeworks/02-problem-set-answers.Rmd
Fix n -> n-1 in problem set 2 answers!

@@
 3. Square the deviations to ensure they are all positive, $(x_i-\bar{X})^2$
-4. Take the average of the squared deviations: add them all up and divide by $n-1$, $\frac{1}{n}\displaystyle \sum^n_{i=1} (x_i-\bar{X})^2$
+4. Take the average of the squared deviations: add them all up and divide by $n-1$, $\frac{1}{n-1}\displaystyle \sum^n_{i=1} (x_i-\bar{X})^2$
 5.
-What you have in step 4 is variance (measured in units of $X^2)$, square root to get standard deviation (measured in original units of $X$): $\sqrt{\frac{1}{n}\displaystyle \sum^n_{i=1} (x_i-\bar{X})^2}$
+What you have in step 4 is variance (measured in units of $X^2)$, square root to get standard deviation (measured in original units of $X$): $\sqrt{\frac{1}{n-1}\displaystyle \sum^n_{i=1} (x_i-\bar{X})^2}$
 

### 2019-10-01 content/r/11-r-practice.Rmd
Add class 11 content

---
title: "2.4 OLS: Precision and Diagnostics - R Practice (Answers)"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

Our [data](https://github.com/fivethirtyeight/data/tree/master/congress-trump-score) comes from fivethirtyeight's [Trump Congress tracker](https://projects.fivethirtyeight.com/congress-trump-score/).

- [View Project on R Studio Cloud](https://rstudio.cloud/project/575252)

## Question 1
Download and read in (`read_csv`) the data.

## Question 2
Look at the data with `glimpse()`.

## Question 3
We want to see *how does the percentage that a member of Congress' agrees with President Trump (`agree_pct`) depend on the result of the 2016 Presidential election in their district (`net_trump_vote`)*? First, plot a scatterplot of `agree_pct` on `net_trump_vote`. Add a regression line with an additional layer of `geom_smooth(method="lm")`.

## Question 4
Find the correlation between `agree_pct` and `net_trump_vote`.

## Question 5
We weant to predict the following model:

$$\widehat{\text{agree_pct}}= \hat{\beta_0}+\hat{\beta_1}\text{net_trump_vote}$$

Run a regression, and save it as an object. Now get a `summary()` of it.

### Part A
What is $\hat{\beta_0}$? What does it mean in the context of our question?

### Part B
What is $\hat{\beta_1}$? What does it mean in the context of our question?

### Part C
What is $R^2$? What does it mean?

### Part D
What is the $SER$? What does it mean?

## Question 6
We can look at regression outputs in a tidier way, with the `broom` package.

### Part A
Install and then load `broom`.

### Part B
Run the function `tidy()` on your regression object (saved in question 5). Save this as an object and then look at it.

### Part C
Run the `glance()` function on your original regression object. What does it show you?

### Part D
Now run the `augment()` function on your original regression object, and save this as an object. Look at it. What does it show you?

## Question 7
Now let's start looking at the residuals of the regression.

### Part A
Take the augmented regression object from Question 6-D and use it as the source of your data to create a histogram, where $x$ is `.resid`. Does it look roughly normal?

### Part B
Take the augmented regression object and make a residual plot, which is a scatterplot where `x` is the normal `x` variable, and `y` is the `.resid`. Feel free to add a horizontal line at 0 with `geom_hline(yintercept=0)`.

## Question 8
Now let's try presenting your results in a regression table. Install and load `huxtable`, and run the `huxreg()` command. Your main input is your regression object you saved in Question 5. Feel free to customize the output of this table (see the slides).

## Question 9
Now let's check for heteroskedasticity.

### Part A
Looking at the scatterplot and residual plots in Questions 3 and 7B, do you think the errors are heteroskedastic or homoskedastic?

### Part B
Install and load the `lmtest` package and run `bptest` on your regression object. Was the data heteroskedastic or homoskedastic?

### Part C
Now let's make some heteroskedasticity-robust standard errors. Install and load the `estimatr` package and use the `lm_robust()` command (instead of the `lm()` command) to run a new regression (and save it). Make sure you add `se_type="stata"` inside the command to calculate robust SEs. Look at it. What changes?

### Part D
Now let's see this in a nice regression table. Use `huxreg()` again, but add both your original regression and your regression saved in part C. Notice any changes?

## Question 10
Now let's check for outliers.

### Part A
Just looking at the scatterplot in Question 3, do you see any outliers?

### Part B
Install and load the `car` package. Run the `outlierTest` command on your regression object. Does it detect any outliers?

### Part C
Look in your original data to match this outlier with an observation. Hint: use the `slice()` command, as the outlier test gave you an observation number!

## Question 11 (Optional)

This data is still a bit messy. We'd also probably like to plot our scatterplots with colors for Republican and Democratic party. Or sort by the House and the Senate.

### 2019-10-01 static/slides/10-slides.Rmd
Add class 11 content

@@
 
-2. The variance of the residuals is constant, written:
-$$var(u)=\sigma^2_{u}$$
+2. The variance of the residuals over $X$ is constant, written:
+$$var(u|X)=\sigma^2_{u}$$
 
@@
 
-# Assumption 2: Heteroskedasticity
+# Assumption 2: Homoskedasticity
 
@@
 
-- The variance of the residuals is constant, written:
-$$var(u)=\sigma^2_{u}$$
+- The variance of the residuals over $X$ is constant, written:
+$$var(u|X)=\sigma^2_{u}$$
 
-- Assumption 2 implies that errors are ".shout[homoskedastic]:
+- Assumption 2 implies that errors are ".shout[homoskedastic]:"
  they have the same variance across $X$
 
-- Often this assumption is violated: errors may be ".shout[heteroskedastic]:
+- Often this assumption is violated: errors may be ".shout[heteroskedastic]:"
  they do not have the same variance across $X$

### 2019-10-01 static/slides/11-slides.Rmd
Add class 11 content

---
title: "2.5: OLS: Precision and Diagnostics"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))
```

```{r regression-setup, echo=F, results="hide"}

library(haven)
CASchool<-read_dta("../data/caschool.dta")

# run regression of testscr on str
school_reg <- lm(testscr ~ str, 
                 data = CASchool)

library(broom)
school_reg_tidy <- tidy(school_reg,
     conf.int = TRUE) # add confidence intervals
CASchool_aug <- augment(school_reg)

library(equatiomatic)
extract_eq(school_reg, use_coefs = TRUE,coef_digits = 2,fix_signs = TRUE)

```

# The Sampling Distribution of $\hat{\beta_1}$

.pull-left[

$$\hat{\beta_1} \sim N(E[\hat{\beta_1}], \sigma_{\hat{\beta_1}})$$

1. $E[\hat{\beta_1}]$; the **center** of the distribution (last class)
    - $E[\hat{\beta_1}]=\beta_1$<sup><span class="red">1</span></sup>

]

.pull-right[

```{r, fig.retina=3}
beta_dist_1<-ggplot(data = tibble(x=-4:4))+
  aes(x = x)+
  stat_function(fun = dnorm, size=2, color="blue")+
  geom_label(x=1, y=dnorm(1), label=expression(sigma[hat(beta[1])]==1), color="blue")+
  geom_segment(aes(x=0,xend=0, y=0, yend=0.4), linetype="dashed")+
  scale_x_continuous(breaks = 0,
                     labels = expression(E(hat(beta[1]))))+
  labs(x = expression(hat(beta[1])),
       y = "Probability")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
beta_dist_1
```

]

---

# The Sampling Distribution of $\hat{\beta_1}$

.pull-left[

$$\hat{\beta_1} \sim N(E[\hat{\beta_1}], \sigma_{\hat{\beta_1}})$$

1. $E[\hat{\beta_1}]$; the **center** of the distribution (last class)
    - $E[\hat{\beta_1}]=\beta_1$<sup><span class="red">1</span></sup>

2. $\sigma_{\hat{\beta_1}}$; how **precise** is our estimate? (today)
    - <span class="shout">Variance $\sigma^2_{\hat{\beta_1}}$</span> or <span class="shout">standard error<sup><span class="red">+</span></sup> $\sigma_{\hat{\beta_1}}$</span>

]

.pull-right[

```{r, fig.retina=3}
beta_dist_1+
  stat_function(fun = dnorm, args=list(mean = 0, sd = 2), size=2, color="red")+
  geom_label(x=2, y=dnorm(2,0,2), label=expression(sigma[hat(beta[1])]==2), color="red")

```

]
.footnote[<sup>.red[1]</sup> Under the 4 assumptions about $u$ (particularly, $cor(X,u)=0)$.

<sup>.red[+]</sup> Standard **"error"** is the analog of standard *deviation* when talking about <br>
the *sampling distribution* of a sample statistic (such as $\bar{X}$ or $\hat{\beta_1})$.
]
---

class: inverse, center, middle

# Variation in $\hat{\beta_1}$

---

# What Affects Variation in $\hat{\beta_1}$

.pull-left[

$$var(\hat{\beta_1})=\frac{(SER)^2}{n \times var(X)}$$

$$se(\hat{\beta_1})=\sqrt{var(\hat{\beta_1})} = \frac{SER}{\sqrt{n} \times sd(X)}$$

]

.pull-right[

- Variation in $\hat{\beta_1}$ is affected by three things:

1. **Goodness of fit of the model $(SER)$**<sup>.red[+]</sup>
    - Larger $SER$ $\rightarrow$ larger $var(\hat{\beta_1})$
2. **Sample size, $n$**
    - Larger $n$ $\rightarrow$ smaller $var(\hat{\beta_1})$
3. **Variance of $X$**
    - Larger $var(X)$ $\rightarrow$ smaller $var(\hat{\beta_1})$

]

.footnote[<sup>.red[+]</sup> Recall from last class, the **S**tandard **E**rror of the **R**egression `\\(\hat{\sigma_u} = \sqrt{\frac{\sum \hat{u_i}^2}{n-2}}\\)`]

---

# Variation in $\hat{\beta_1}$: Goodness of Fit

.pull-left[

```{r, fig.retina=3}
data_1<-tibble(x=rnorm(50,5,1),
               u=rnorm(50,1,1),
               y=3+x+u)

sd_x_1<-lm(y~x, data = data_1) %>%
  tidy() %>%
  slice(2) %>% # get second row (which is x coefficient, beta 1)
  pull(std.error)  %>%
  round(.,2) %>%
  as.character()

beta0_1<-lm(y~x, data = data_1) %>%
  tidy() %>%
  slice(1) %>% # get first row (which is intercept, beta 0)
  pull(estimate)  %>%
  round(.,2) %>%
  as.character()

beta1_1<-lm(y~x, data = data_1) %>%
  tidy() %>%
  slice(2) %>% # get second row (which is x coefficient, beta 1)
  pull(estimate)  %>%
  round(.,2) %>%
  as.character()

ser_1<-lm(y~x, data = data_1) %>%
  glance() %>%
  pull(sigma)  %>%
  round(.,2) %>%
  as.character()

ggplot(data = data_1)+
  aes(x = x,
      y = y)+
  geom_point(color="blue")+
  geom_smooth(method = "lm", color = "red")+
  geom_text(aes(x=7,y=6), label=parse(text=paste('~hat(Y)==', beta0_1, '~+', beta1_1, '~X')))+
  geom_text(aes(x=7,y=5), label=parse(text=paste('~SER==', ser_1)))+
  geom_text(aes(x=7,y=4), label=parse(text=paste('~SE(hat(beta[1])) ==', sd_x_1)))+
  #geom_text(aes(x=5,y=13.5), label=bquote(sigma[hat(beta[1])] == .(sd_x_1)))+
  scale_x_continuous(breaks=seq(2,8,2),
                     limits=c(2,8))+
  scale_y_continuous(breaks=seq(3,15,3),
                     limits=c(3,15))+
  labs(x = "X",
       y = "Y",
       title = "Model With Better Fit",
       subtitle = expression(paste("Lower SER lowers variation in ", hat(beta[1]))))+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

```{r, fig.retina=3}
data_2<-tibble(x=data_1$x,
               u=rnorm(50,1,3),
               y=3+x+u)

sd_x_2<-lm(y~x, data = data_2) %>%
  tidy() %>%
  slice(2) %>% # get second row (which is x coefficient, beta 1)
  pull(std.error)  %>%
  round(.,2) %>%
  as.character()

beta0_2<-lm(y~x, data = data_2) %>%
  tidy() %>%
  slice(1) %>% # get first row (which is intercept, beta 0)
  pull(estimate)  %>%
  round(.,2) %>%
  as.character()

beta1_2<-lm(y~x, data = data_2) %>%
  tidy() %>%
  slice(2) %>% # get second row (which is x coefficient, beta 1)
  pull(estimate)  %>%
  round(.,2) %>%
  as.character()

ser_2<-lm(y~x, data = data_2) %>%
  glance() %>%
  pull(sigma)  %>%
  round(.,2) %>%
  as.character()

ggplot(data = data_2)+
  aes(x = x,
      y = y)+
  geom_point(color="blue")+
  geom_smooth(method = "lm", color = "red")+
  geom_text(aes(x=7,y=6), label=parse(text=paste('~hat(Y)==', beta0_2, '~+', beta1_2, '~X')))+
  geom_text(aes(x=7,y=5), label=parse(text=paste('~SER==', ser_2)))+
  geom_text(aes(x=7,y=4), label=parse(text=paste('~SE(hat(beta[1])) ==', sd_x_2)))+
  scale_x_continuous(breaks=seq(2,8,2),
                     limits=c(2,8))+
  scale_y_continuous(breaks=seq(3,15,3),
                     limits=c(3,15))+
  labs(x = "X",
       y = "Y",
       title = "Model With Worse Fit",
       subtitle = expression(paste("Higher SER raises variation in ", hat(beta[1]))))+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
```

]

---

# Variation in $\hat{\beta_1}$: Sample Size

.pull-left[

```{r, fig.retina=3}
# reuse data_1

ggplot(data = data_1)+
  aes(x = x,
      y = y)+
  geom_point(color="blue")+
  geom_smooth(method = "lm", color = "red")+
  geom_text(aes(x=7,y=6), label=parse(text=paste('~hat(Y)==', beta0_1, '~+', beta1_1, '~X')))+
  geom_text(aes(x=7,y=5), label=parse(text=paste('~SER==', ser_1)))+
  geom_text(aes(x=7,y=4), label=parse(text=paste('~SE(hat(beta[1])) ==', sd_x_1)))+
  #geom_text(aes(x=5,y=13.5), label=bquote(sigma[hat(beta[1])] == .(sd_x_1)))+
  scale_x_continuous(breaks=seq(2,8,2),
                     limits=c(2,8))+
  scale_y_continuous(breaks=seq(3,15,3),
                     limits=c(3,15))+
  labs(x = "X",
       y = "Y",
       title = "Model With Fewer Observations",
       subtitle = expression(paste("Smaller n raises variation in ", hat(beta[1]))))+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

```{r, fig.retina=3}
data_3<-tibble(x=rnorm(100,5,1),
               u=rnorm(100,1,1),
               y=3+x+u)

sd_x_3<-lm(y~x, data = data_3) %>%
  tidy() %>%
  slice(2) %>% # get second row (which is x coefficient, beta 1)
  pull(std.error)  %>%
  round(.,2) %>%
  as.character()

beta0_3<-lm(y~x, data = data_3) %>%
  tidy() %>%
  slice(1) %>% # get first row (which is intercept, beta 0)
  pull(estimate)  %>%
  round(.,2) %>%
  as.character()

beta1_3<-lm(y~x, data = data_3) %>%
  tidy() %>%
  slice(2) %>% # get second row (which is x coefficient, beta 1)
  pull(estimate)  %>%
  round(.,2) %>%
  as.character()

ser_3<-lm(y~x, data = data_3) %>%
  glance() %>%
  pull(sigma)  %>%
  round(.,2) %>%
  as.character()

ggplot(data = data_3)+
  aes(x = x,
      y = y)+
  geom_point(color="blue")+
  geom_smooth(method = "lm", color = "red")+
  geom_text(aes(x=7,y=6), label=parse(text=paste('~hat(Y)==', beta0_3, '~+', beta1_3, '~X')))+
  geom_text(aes(x=7,y=5), label=parse(text=paste('~SER==', ser_3)))+
  geom_text(aes(x=7,y=4), label=parse(text=paste('~SE(hat(beta[1])) ==', sd_x_3)))+
  scale_x_continuous(breaks=seq(2,8,2),
                     limits=c(2,8))+
  scale_y_continuous(breaks=seq(3,15,3),
                     limits=c(3,15))+
  labs(x = "X",
       y = "Y",
       title = "Model With More Observations",
       subtitle = expression(paste("Larger n lowers variation in ", hat(beta[1]))))+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
```

]

---

# Variation in $\hat{\beta_1}$: Variation in $X$

.pull-left[

```{r, fig.retina=3}
# reuse data_3

sd_X_3<-data_3 %>%
  summarize(sd_x_3=sd(x)) %>%
  round(.,2) %>%
  as.character()

ggplot(data = data_3)+
  aes(x = x,
      y = y)+
  geom_point(color="blue")+
  geom_smooth(method = "lm", color = "red")+
  geom_text(aes(x=7,y=6), label=parse(text=paste('~hat(Y)==', beta0_1, '~+', beta1_1, '~X')))+
  geom_text(aes(x=7,y=5), label=parse(text=paste('~SER==', ser_1)))+
  geom_text(aes(x=7,y=4), label=parse(text=paste('~SD(X) ==', sd_X_3)))+
  geom_text(aes(x=7,y=3), label=parse(text=paste('~SE(hat(beta[1])) ==', sd_x_1)))+
  #geom_text(aes(x=5,y=13.5), label=bquote(sigma[hat(beta[1])] == .(sd_x_1)))+
  scale_x_continuous(breaks=seq(2,8,2),
                     limits=c(2,8))+
  scale_y_continuous(breaks=seq(3,15,3),
                     limits=c(3,15))+
  labs(x = "X",
       y = "Y",
       title = "Model With More Variation in X",
       subtitle = expression(paste("Larger ", var(X), " lowers variation in ", hat(beta[1]))))+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

```{r, fig.retina=3}
#subset data_3

data_4<-data_3 %>%
  filter(x>4.5, x<5.5)

sd_X_4<-data_4 %>%
  summarize(sd_x_4=sd(x)) %>%
  round(.,2) %>%
  as.character()

sd_x_4<-lm(y~x, data = data_4) %>%
  tidy() %>%
  slice(2) %>% # get second row (which is x coefficient, beta 1)
  pull(std.error)  %>%
  round(.,2) %>%
  as.character()

beta0_4<-lm(y~x, data = data_4) %>%
  tidy() %>%
  slice(1) %>% # get first row (which is intercept, beta 0)
  pull(estimate)  %>%
  round(.,2) %>%
  as.character()

beta1_4<-lm(y~x, data = data_4) %>%
  tidy() %>%
  slice(2) %>% # get second row (which is x coefficient, beta 1)
  pull(estimate)  %>%
  round(.,2) %>%
  as.character()

ser_4<-lm(y~x, data = data_4) %>%
  glance() %>%
  pull(sigma)  %>%
  round(.,2) %>%
  as.character()

ggplot(data = data_4)+
  aes(x = x,
      y = y)+
  geom_point(color="blue")+
  geom_point(data = data_3, aes(x = x, y = y), color="blue", alpha=0.3)+
  geom_smooth(method = "lm", color = "red")+
  geom_text(aes(x=7,y=6), label=parse(text=paste('~hat(Y)==', beta0_4, '~+', beta1_4, '~X')))+
  geom_text(aes(x=7,y=5), label=parse(text=paste('~SER==', ser_4)))+
  geom_text(aes(x=7,y=4), label=parse(text=paste('~SD(X) ==', sd_X_4)))+
  geom_text(aes(x=7,y=3), label=parse(text=paste('~SE(hat(beta[1])) ==', sd_x_4)))+
  scale_x_continuous(breaks=seq(2,8,2),
                     limits=c(2,8))+
  scale_y_continuous(breaks=seq(3,15,3),
                     limits=c(3,15))+
  labs(x = "X",
       y = "Y",
       title = "Model With Less Variation in X",
       subtitle = expression(paste("Smaller ", var(X), " raises variation in ", hat(beta[1]))))+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
```

]

---

class: inverse, center, middle

# Presenting Regression Results

---

# Our Class Size Regression: Base R

.left-code[

```{r, echo=T}
summary(school_reg) # get full summary
```

]

.right-plot[

- How can we present all of this information in a tidy way?

]

---

# Our Class Size Regression: Broom I

- `broom`'s `tidy()` gives us a little bit neater table

```{r, echo=T}
library(broom)
tidy(school_reg)
```

---

# Our Class Size Regression: Broom II

- `broom`'s `glance()` gives us summary statistics about the regression

```{r, echo=T}
library(broom)
glance(school_reg)
```

---

# Presenting Regressions in a Table

.pull-left[

- Professional journals and papers often have a regression table, including:
  - Estimates of $\hat{\beta_0}$ and $\hat{\beta_1}$
  - Standard errors of $\hat{\beta_0}$ and $\hat{\beta_1}$ (often below, in parentheses)
  - Indications of statistical significance (often with asterisks)
  - Measures of regression fit: $R^2$, $SER$, etc

- Later: multiple columns for multiple models
]

.pull-right[
.font80[
.regtable[
```{r}
library(huxtable)
huxreg("Test Score" = school_reg,
       coefs = c("Intercept" = "(Intercept)",
                 "STR" = "str"),
       statistics = c("N" = "N",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```
]
]
]

---

# Regression Output with huxtable I

.pull-left[

- You will need to first `install.packages("huxtable")`

- Load with `library(huxtable)`

- Command: `huxreg()`

- Main argument is the name of your `lm` object

- Default output is fine, but often we want to customize a bit

```{r, echo=T, eval=F}
# install.packages("huxtable")
library(huxtable)

huxreg(school_reg)
```

]

.pull-right[

.font80[
.regtable[
```{r}
huxreg(school_reg)
```
]
]
]

---

# Regression Output with huxtable II

- Can give title to each column

```{r, echo=T, eval=F}
"Test Score" = school_reg
```

--

- Can change name of coefficients from default

```{r, echo=T, eval=F}
coefs = c("Intercept" = "(Intercept)",
          "STR" = "str")
```

--

- Decide what statistics to include, and rename them

```{r, echo=T, eval=F}
statistics = c("N" = "N",
               "R-Squared" = "r.squared",
               "SER" = "sigma")
```

--

- Choose how many decimal places to round to

```{r, echo=T, eval=F}
number_format = 2
```

---

# Regression Output with huxtable III

.pull-left[

```{r, echo=T, eval=F}
huxreg("Test Score" = school_reg,
       coefs = c("Intercept" = "(Intercept)",
                 "STR" = "str"),
       statistics = c("N" = "N",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```

]

--

.pull-left[
.font80[
.regtable[
```{r}
huxreg("Test Score" = school_reg,
       coefs = c("Intercept" = "(Intercept)",
                 "STR" = "str"),
       statistics = c("N" = "N",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```
]
]
]

---

# Regression Outputs

- `huxtable` is one package you can use
    - See [here for more options](https://cran.r-project.org/web/packages/huxtable/vignettes/huxtable.html)

- I used to only use [`stargazer`](https://cran.r-project.org/web/packages/stargazer/index.html), but as it was originally meant for STATA, it has limits and problems
    - A great [cheetsheat](http://jakeruss.com/cheatsheets/stargazer.html) by my friend Jake Russ

---

class: inverse, center, middle

# Diagnostics about Regression

---

# Diagnostics: Residuals I

- We often look at the residuals of a regression to get more insight about its **goodness of fit** and its **bias**

- Recall `broom`'s `augment` creates some useful new variables
    - `.fitted` are fitted (predicted) values from model, i.e. $\hat{Y}_i$
    - `.resid` are residuals (errors) from model, i.e. $\hat{u}_i$

---

# Diagnostics: Residuals II

- Often a good idea to store in a new object (so we can make some plots)

```{r, echo=T}
aug_reg<-augment(school_reg)
aug_reg %>%
  head()
aug_reg<-augment(school_reg)
aug_reg %>% head()
```

---

# Recap: Assumptions about Errors

.pull-left[

- Recall the .shout[4 critical **assumptions** about `\\(u\\)`]:

1. The expected value of the residuals is 0
$$E[u]=0$$

2. The variance of the residuals over $X$ is constant, written:
$$var(u|X)=\sigma^2_{u}$$

3. Errors are not correlated across observations: 
$$cor(u_i,u_j)=0 \quad \forall i \neq j$$

4. There is no correlation between $X$ and the error term: 
$$cor(X, u)=0 \text{ or } E[u|X]=0$$

]

.pull-right[

.center[
![](https://www.dropbox.com/s/95rh9ow982y5htb/error.png?raw=1)
]
]

---

# Assumptions 1 and 2: Errors are i.i.d.

- Assumptions 1 and 2 assume that errors are coming from the same (*normal*) distribution
$$u \sim N(0, \sigma_u)$$
  - Assumption 1: $E[u]=0$
  - Assumption 2: $sd(u|X)=\sigma_u$
      - virtually always unknown...

- We often can visually check by plotting a **histogram** of $u$

---

# Plotting Residuals

.pull-left[

```{r residual-hist, echo=T, eval=F}
ggplot(data = aug_reg)+
  aes(x = .resid)+
  geom_histogram(color="white")+
  labs(x = expression(paste("Residual, ", hat(u))))+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

--

.pull-right[

```{r, ref.label="residual-hist", fig.retina=3}
```

]

---

# Plotting Residuals

.pull-left[

```{r residual-hist-2, echo=T, eval=F}
ggplot(data = aug_reg)+
  aes(x = .resid)+
  geom_histogram(color="white")+
  labs(x = expression(paste("Residual, ", hat(u))))+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
```

- Just to check:

```{r, echo=T}
aug_reg %>%
  summarize(e_u = mean(.resid),
            sd_u = sd(.resid))
```

]

--

.pull-right[

```{r, ref.label="residual-hist-2", fig.retina=3}
```

]

---

# Residual Plot

.pull-left[

- We often plot a .shout[residual plot] to see any odd patterns about residuals
    - $x$-axis are $X$ values (`str`)
    - $y$-axis are $u$ values (`.resid`)

```{r residual-plot, eval=F}
ggplot(data = aug_reg)+
  aes(x = str,
      y = .resid)+
  geom_point(color="blue")+
  geom_hline(aes(yintercept = 0), color="red")+
  labs(x = "Student to Teacher Ratio",
       y = expression(paste("Residual, ", hat(u))))+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

--

.pull-right[

```{r, ref.label="residual-plot", fig.retina=3}
```

]

---

class: inverse, center, middle

# Problem: Heteroskedasticity

---

# Homoskedasticity

.pull-left[

- ".shout[Homoskedasticity]:" variance of the residuals over $X$ is constant, written:
$$var(u|X)=\sigma^2_{u}$$

- Knowing the value of $X$ does not affect the variance (spread) of the errors

]

.pull-right[

.center[
![](https://www.dropbox.com/s/95rh9ow982y5htb/error.png?raw=1)
]
]

---

# Heteroskedasticity I

.pull-left[

- ".shout[Heteroskedasticity]:" variance of the residuals over $X$ is *NOT* constant:
$$var(u|X) \neq \sigma^2_{u}$$

- **This does not cause $\hat{\beta_1}$ to be biased**, but it does cause the standard error of $\hat{\beta_1}$ to be incorrect

- This **does** cause a problem for .onfire[inference]!
]

.pull-right[

.center[
![](https://www.dropbox.com/s/95rh9ow982y5htb/error.png?raw=1)
]
]

---

# Heteroskedasticity II

- Recall the formula for the standard error of $\hat{\beta_1}$:

$$se(\hat{\beta_1})=\sqrt{var(\hat{\beta_1})} = \frac{SER}{\sqrt{n} \times sd(X)}$$

- This actually *assumes* homoskedasticity

---

# Heteroskedasticity III

- Under heteroskedasticity, the standard error of $\hat{\beta_1}$ mutates to:

$$se(\hat{\beta_1})=\sqrt{\frac{\displaystyle\sum^n_{i=1}(X_i-\bar{X})^2\hat{u}^2}{\big[\displaystyle\sum^n_{i=1}(X_i-\bar{X})^2\big]^2}}$$

- This is a .shout[heteroskedasticity-robust] (or just .shout["robust"]) method of calculating $se(\hat{\beta_1})$

- Don't learn formula, **do learn what heteroskedasticity is and how it affects our model!**

---

# Visualizing Heteroskedasticity I

.pull-left[

- Our original scatterplot with regression line
]

.pull-right[

```{r, fig.height=3.5, fig.retina=3}
ggplot(data = CASchool)+
  aes(x = str,
      y = testscr)+
  geom_point(color="blue")+
  geom_smooth(method="lm", color="red")+
  labs(x = "Student to Teacher Ratio",
       y = "Test Score")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=16)
```
]

---

# Visualizing Heteroskedasticity I

.pull-left[

- Our original scatterplot with regression line

- Does the spread of the errors change over different values of $str$?
    - No: homoskedastic
    - Yes: heteroskedastic
]

.pull-right[

```{r, fig.height=3.5, retina=3}
ggplot(data = CASchool)+
  aes(x = str,
      y = testscr)+
  geom_point(color="blue")+
  geom_vline(xintercept=seq(14,26,2), linetype="dashed", size=1)+
  geom_smooth(method="lm", color="red")+
  labs(x = "Student to Teacher Ratio",
       y = "Test Score")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=16)
```
]

---

# Visualizing Heteroskedasticity I

.pull-left[

- Our original scatterplot with regression line

- Does the spread of the errors change over different values of $str$?
    - No: homoskedastic
    - Yes: heteroskedastic
]

.pull-right[

```{r, fig.height=3.5, retina=3}
ggplot(data = CASchool)+
  aes(x = str,
      y = testscr)+
  geom_point(color="blue")+
  geom_vline(xintercept=seq(14,26,2), linetype="dashed", size=1)+
  geom_smooth(method="lm", color="red")+
  labs(x = "Student to Teacher Ratio",
       y = "Test Score")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=16)
```

```{r, fig.height=3.5, retina=3}
aug_reg_het<-aug_reg %>%
  mutate(range = case_when(str>=14 & str<16 ~ "14 to 16",
                           str>=16 & str<18 ~ "16 to 18",
                           str>=18 & str<20 ~ "18 to 20",
                           str>=20 & str<22 ~ "20 to 22",
                           str>=22 & str<24 ~ "22 to 24",
                           str>=24 & str<26 ~ "24 to 26"),
         range = factor(range, levels = c("14 to 16", "16 to 18", "18 to 20", "20 to 22", "22 to 24", "24 to 26"))) # needs to be a factor in order to plot in order

aug_reg_het_sigmas<-aug_reg_het %>%
  group_by(range) %>%
  summarize(sigmas = as.character(round(sd(.resid),2))) %>%
  slice(1:6) # remove NA row 7

ggplot(data = aug_reg_het)+
  aes(x = .resid)+
  geom_density(aes(fill=range), alpha=0.5)+
  geom_label(data=aug_reg_het_sigmas,
             aes(x=0,y=0.005,
                 label=sigmas,
                 color=range),size=5)+
                 #label=parse(text=paste('~sigma==', sigmas))))+
  facet_grid(~range)+
  guides(fill = F,
         color = F)+
  labs(x = "Residuals by STR",
       y = "Test Score")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=16)
```

]

---

# More Obvious Heteroskedasticity

.pull-left[

- Visual cue: data is "fan-shaped"
    - Data points are closer to line in some areas
    - Data points are more spread from line in other areas
]

.pull-right[

```{r, fig.height=3.5, fig.retina=3}
het_data<-tibble(x = runif(500,0,10),
                 y = rnorm(500,x,x))

ggplot(data = het_data)+
  aes(x = x,
      y = y)+
  geom_point(color="blue")+
  geom_smooth(method="lm", color="red")+
  scale_x_continuous(breaks=seq(0,10,1))+
    labs(x = "X",
       y = "Y")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)

```
]

---

# More Obvious Heteroskedasticity

.pull-left[

- Visual cue: data is "fan-shaped"
    - Data points are closer to line in some areas
    - Data points are more spread from line in other areas
]

.pull-right[

```{r, fig.height=3.5, fig.retina=3}
ggplot(data = het_data)+
  aes(x = x,
      y = y)+
  geom_point(color="blue")+
  geom_smooth(method="lm", color="red")+
  geom_vline(xintercept=seq(0,10,2), linetype="dashed", size=1)+
  scale_x_continuous(breaks=seq(0,10,1))+
    labs(x = "X",
       y = "Y")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)

het_reg<-lm(y~x, data = het_data)
aug_het_reg<-het_reg %>% augment()

aug_het_reg<-aug_het_reg %>%
  mutate(range = case_when(x>=0 & x<2 ~ "0 to 2",
                           x>=2 & x<4 ~ "2 to 4",
                           x>=4 & x<6 ~ "4 to 6",
                           x>=6 & x<8 ~ "6 to 8",
                           x>=8 & x<=10 ~ "8 to 10"),
         range = factor(range, levels = c("0 to 2", "2 to 4", "4 to 6", "6 to 8", "8 to 10"))) # needs to be a factor to be in order in the plot

aug_het_reg_sigmas<-aug_het_reg %>%
  group_by(range) %>%
  summarize(sigmas = as.character(round(sd(.resid),2))) %>%
  slice(1:6) # remove NA row 7

#facets$new = factor(temp$type, levels=c("T","F","P"), labels=c("T","F","P"))

ggplot(data = aug_het_reg)+
  aes(x = .resid)+
  geom_density(aes(fill=range), alpha=0.5)+
  geom_label(data=aug_het_reg_sigmas,
             aes(x=0,y=0.3,
                 label=sigmas,
                 color=range),size=5)+
                 #label=parse(text=paste('~sigma==', sigmas))))+
  facet_grid(~range)+
  guides(fill = F,
         color = F)+
  labs(x = "Residuals by Range of X",
       y = "Density")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)

```
]

---

# What Might Cause Heteroskedastic Errors?

.pull-left[

$$\widehat{wage_i}=\hat{\beta_0}+\hat{\beta_1}educ_i$$

```{r}
library(wooldridge)
wage_reg<-lm(wage~educ, data=wage1)
huxreg("Wage" = wage_reg,
       coefs = c("Intercept" = "(Intercept)",
                 "Years of Schooling" = "educ"),
       statistics = c("N" = "N",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```
]

.pull-right[

```{r, fig.height=3.5, fig.retina=3}

wage_plot<-ggplot(data = wage1)+
  aes(x = educ,
      y = wage)+
  geom_point(color="blue")+
  geom_smooth(method="lm", color="red")+
  scale_x_continuous(breaks=seq(0,20,2))+
  scale_y_continuous(labels=scales::dollar)+
    labs(x = "Years of Schooling",
       y = "Hourly Wage")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
wage_plot
```
]

---

# What Might Cause Heteroskedastic Errors?

.pull-left[

$$\widehat{wage_i}=\hat{\beta_0}+\hat{\beta_1}educ_i$$

```{r}
library(wooldridge)
wage_reg<-lm(wage~educ, data=wage1)
huxreg("Wage" = wage_reg,
       coefs = c("Intercept" = "(Intercept)",
                 "Years of Schooling" = "educ"),
       statistics = c("N" = "N",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```
]

.pull-right[

```{r, fig.height=3.5, fig.retina=3}

wage_reg <- lm(wage ~ educ, data = wage1)
aug_wage_reg<-wage_reg %>%
  augment()
wage_plot+geom_vline(xintercept=c(0,4,8,12,16,18), linetype="dashed", size=1)

```

```{r, fig.height=3.5, fig.retina=3}
aug_wage_reg<-aug_wage_reg %>%
  mutate(range = case_when(educ>=0 & educ<4 ~ "0 to 4",
                           educ>=4 & educ<8 ~ "4 to 8",
                           educ>=8 & educ<12 ~ "8 to 12",
                           educ>=12 & educ<16 ~ "12 to 16",
                           educ>=16 & educ<20 ~ "16 to 18"),
         range = factor(range, levels = c("0 to 4", "4 to 8", "8 to 12", "12 to 16", "16 to 18"))) # needs to be a factor to be in order in the plot

aug_wage_reg_het_sigmas<-aug_wage_reg %>%
  group_by(range) %>%
  summarize(sigmas = as.character(round(sd(.resid),2))) %>%
  slice(1:6) # remove NA row 7

#facets$new = factor(temp$type, levels=c("T","F","P"), labels=c("T","F","P"))

ggplot(data = aug_wage_reg)+
  aes(x = .resid)+
  geom_density(aes(fill=range), alpha=0.5)+
  geom_label(data=aug_wage_reg_het_sigmas,
             aes(x=5,y=0.3,
                 label=sigmas,
                 color=range),size=5)+
                 #label=parse(text=paste('~sigma==', sigmas))))+
  facet_grid(~range)+
  guides(fill = F,
         color = F)+
  labs(x = "Residuals by Years of Schooling",
       y = "Density")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)

```

---

# Detecting Heteroskedasticity I

- Several tests to check if data is heteroskedastic 
- One common test is **Breusch-Pagan test** 
- Can use `bptest()` with `lmtest` package in `R`
    - $H_0$: homoskedastic
    - If $p$-value < 0.05, reject $H_0\implies$ heteroskedastic

--

.pull-left[
```{r,eval=F, echo=T}
# install.packages("lmtest")
library("lmtest")
bptest(school_reg)
```
]

.pull-right[

```{r}
library("lmtest") #load lmtest package, install if first time
bptest(school_reg)
```
]

---

# Detecting Heteroskedasticity II

- How about our wage regression?

.pull-left[
```{r,eval=F, echo=T}
# install.packages("lmtest")
library("lmtest")
bptest(wage_reg)
```
]

.pull-right[

```{r}
library("lmtest") #load lmtest package, install if first time
bptest(wage_reg)
```
]

---

# Fixing Heteroskedasticity I

- Heteroskedasticity is easy to fix with software that can calculate .shout[robust] standard errors (using the more complicated formula above)

- Easiest method is to use `estimatr` package
  - `lm_robust()` command (instead of `lm`) to run regression
  - set `se_type="stata"` to calculate robust SEs using the formula above

```{r}
#install.packages("estimatr")
library(estimatr)

school_reg_robust <-lm_robust(testscr ~ str, data = CASchool,
                              se_type = "stata")

school_reg_robust
```

---

# Fixing Heteroskedasticity II

.pull-left[
```{r, echo=T, eval=F}
library(huxtable)
huxreg("Normal" = school_reg,
       "Robust" = school_reg_robust,
       coefs = c("Intercept" = "(Intercept)",
                 "STR" = "str"),
       statistics = c("N" = "N",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)

```
]

.pull-right[
.font80[
.regtable[
```{r}
huxreg("Normal" = school_reg,
       "Robust" = school_reg_robust,
       coefs = c("Intercept" = "(Intercept)",
                 "STR" = "str"),
       statistics = c("N" = "N",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```
]
]
]

---

# Assumption 3: No Serial Correlation

.pull-left[

- Errors are not correlated across observations: 
$$cor(u_i,u_j)=0 \quad \forall i \neq j$$

- For simple cross-sectional data, this is rarely an issue

- Time-series & panel data nearly always contain .shout[serial correlation] or .shout[autocorrelation] between errors

- Errors may be .shout[clustered]
    - **by group**: e.g. all observations from Maryland, all observations from Virginia, etc.
    - **by time**: GDP in 2006 around the world, GDP in 2008 around the world, etc.

- We'll deal with these fixes when we talk about panel data (or time-series if necessary)

]

.pull-right[

.center[
![](https://www.dropbox.com/s/95rh9ow982y5htb/error.png?raw=1)
]
]

---

class: inverse, center, middle
# Outliers

---

# Outliers Can Bias OLS! I

.pull-left[

- .shout[Outliers] can affect the slope (and intercept) of the line and add .shout[bias]
    - May be result of human error (measurement, transcribing, etc)
    - May be meaningful and accurate

- In any case, compare how including/dropping outliers affects regression and always discuss outliers!

]

.pull-right[

```{r, fig.retina=3}
#Just to simplify the dataset, let's make a new, smaller data frame with just the variables: observation, district, testscr, and str 
CA.simple<-subset(CASchool,select=c(observat,district,testscr,str))

#Now let's create two crazy outliers 
outlier<-c(421,"Terrible School 1",800,30)
outlier2<-c(422,"Terrible School 2",850,28)
outlier3<-c(423,"Terrible School 2",820,29)

#And add them to a new dataframe 
CA.outlier<-rbind(CA.simple,outlier,outlier2,outlier3)
CA.outlier$str<-as.numeric(CA.outlier$str) #dont know why it changes class to character, change to numeric
CA.outlier$testscr<-as.numeric(CA.outlier$testscr)
outliers<-subset(CA.outlier, testscr>790)

#Create a scatterplot with the outlier 
ggplot(CA.simple, aes(str,testscr))+
  geom_point(color="blue")+ #the normal data points
  geom_smooth(data=CA.simple,method=lm, color="red")+ #the regression line
  geom_point(data=outliers, color="magenta")+ #the outliers (in magenta)
  geom_smooth(data=CA.outlier, method=lm,color="purple")+ #the regression line with outliers
  xlab("Student to Teacher Ratio")+ylab("Test Score")+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)

```
]
---

# Outliers Can Bias OLS! II

.font80[
.regtable[
```{r}
school_outlier_reg<-lm(testscr~str, data=CA.outlier)

huxreg("No Outliers" = school_reg,
       "Outliers" = school_outlier_reg,
       coefs = c("Intercept" = "(Intercept)",
                 "STR" = "str"),
       statistics = c("N" = "N",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```
]
]

---

# Detecting Outliers

- The `car` package has an `outlierTest` command to run on the regression

```{r, echo=T}
library("car")

# Use Bonferonni test 
outlierTest(school_outlier_reg) # will point out which obs #s seem outliers
```

### 2019-10-03 content/class/11-class.Rmd
Add 12 class content

---
title: "2.5: OLS: Precision and Diagnostics - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

*Tuesday, October 2, 2019*

# Overview

Last class and this class we are looking at the *sampling distibution* of OLS estimators (particularly $\hat{\beta_1})$. Last class we looked at what the *center* of the distribution was - the true $\beta_1$ - so long as the assumptions about $u$ hold:

- When $cor(X,u)=0$, $X$ is *exogenous* and the OLS estimators are *unbiased*.
- What $cor(X,u)\neq 0$, $X$ is *endogenous* and the OLS estimators are *biased*.

Today we continue looking at the *sampling distibution* by determining the variation in $\hat{beta_1}$ (it's variance or its standard error^[The square root of variance, as always!]). We look at the formula and see the three major determinants of variation in $\hat{\beta_1}$:

1. Goodness of fit of the regression $(SER$ or $\hat{\sigma_u}$
2. Sample size $n$
3. Variation in $X$

We also look at the diagnostics of a regression by looking at its residuals $(\hat{u_i})$ for anomalies. We focus on the problem of *heteroskedasticity* (where the variation in $\hat{u_i])$ changes over the range of $X$, which violates assumption 2 (errors are homoskedastic): how to detect it, test it, and fix it with some packages. We also look at outliers, which can bias the regression. Finally, we also look at how to present regression results.

We continue the extended example about class sizes and test scores, which comes from a (Stata) dataset from an old textbook that I used to use, [Stock and Watson, 2007](https://www.pearson.com/us/higher-education/product/Stock-Introduction-to-Econometrics-Update-3rd-Edition/9780133486872.html?tab=resources). Download and follow along with the data from today's example:^[Note this is a `.dta` Stata file. You will need to (install and) load the package `haven` to `read_dta()` Stata files into a dataframe.]

- [<i class="fas fa-table"></i> `caschool.dta`](http://metricsf19.classes.ryansafner.com/data/caschool.dta)

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/11-slides.html)

# Practice Problems

Today you will be working on [R practice problems](/r/11-r-practice.html). Check back later for solutions.

# New Packages Mentioned

- `broom`: for tidy regression outputs, summary statistics, and adding $\hat{Y_i}$ and $\hat{u_i}$ into the dataframe
- `huxtable`: to present regression output in a table with `huxreg()`
- `lmtest`: for testing for heteroskedasticity in errors with `bptest()`
- `car`: for testing for outliers with `outlierTest()`
- `estimatr`: for calculating robust standard errors with `lm_robust()`

# Assignments: Problem Set 2 Answers

[Problem Set 2](/assignment/01-problem-set.html) answers are posted.

### 2019-10-03 content/class/12-class.Rmd
Add 12 class content

---
title: "2.6: Statistical Inference I - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

*Thursday, October 3, 2019*

# Overview

We begin with some more time for you to work on the [R Practice]() from last class. I will post the answer key after class.

This class and the next class are about inferential statistics: using *statistics* calculated from a *sample* of data to **infer** the true (and unmeasurable) *parameters* that describe a *population.* In doing so, we can run *hypothesis tests* on our sample to determine a *point* estimate of a parameter, or construct a *confidence interval* from our sample to cast a range for the true parameter.

This is standard principles of statistics - you hopefully should have learned it before. If it has been a while (or never) since your last statistics class, this is one of the hardest concepts to understand at first glance. I recommend [Khan Academy](https://www.khanacademy.org/math/statistics-probability)^[From sampling distributions through significance tests, for this. Though the whole class is helpful!] or Google for these concepts, as *every* statistics class will cover them in the standard way.

That being said, this semester I am *not* going to cover them in the standard way (see the appendix today below for an overview of the standard way). I think it will be more intuitive if I *show* you where these concepts come from by *simulation* as opposed to theoretical sampling distributions.

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/12-slides.html)

# Practice Problems

Today you will be finishing up [R practice problems](/r/11-r-practice.html) from last class. Check back later for solutions.

# New Packages Mentioned

- `broom`: for tidy regression outputs, summary statistics, and adding $\hat{Y_i}$ and $\hat{u_i}$ into the dataframe
- `huxtable`: to present regression output in a table with `huxreg()`
- `lmtest`: for testing for heteroskedasticity in errors with `bptest()`
- `car`: for testing for outliers with `outlierTest()`
- `estimatr`: for calculating robust standard errors with `lm_robust()`

# Assignments: Problem Set 2 Answers

[Problem Set 2](/assignment/01-problem-set.html) answers are posted.

# Appendix: Inferential Statistics

# Inferential Statistics

## Differences Between What We Learned in Class and Classical Statistics

In class, you learned the basics behind inferential statistics--p-values, confidence intervals, hypothesis testing--via *empirical simulation* of many samples permuted from our existing data. We took our sample, ran 1,000 simulations by permutation of our sample without replacement^[That is, for each simulation, we randomly selected observations from our existing sample to be in the simulation, and then did *not* put that observation back in the pool to possibly be selected again.], calculated the statistic ($\hat{\beta}_0$, the slope) of each simulation; this gave us a (sampling) distribution of our sample statistics, and then found the probability on that distribution that we would observe our *actual* statistic in our actual data -- this is the $p$-value.

Classically, before the use of computers that could run and visualize 1,000s of simulations within seconds,^[Even when *I* was in graduate school, 2011--2015]  inferential statistics was taught using *theoretical* distributions. **Essentially, we calculate a test-statistic by normalizing our finding against a theoretical (null) sampling distribution of our sample statistic, and find p-values by estimating the probability of observing that statistic on that theoretical distribution.** These distributions are almost always *normal* or *normal-like* distributions. The distribution that we almost always use in econometrics is the (Student's) $t$-distribution.

Furthermore, testing the null hypothesis $H_0: \,\beta_1=0$, is not the only type of hypothesis test, nor is the slope the only statistic we can test. In fact, there are many different types of hypothesis tests that are well-known and well-used, we focused entirely on regression (since that is the largest tool of the course).

This appendix will give you more background on the theory of inferential statistics, and is more in line with what you may have learned in earlier statistics courses.

## Inferential Statistics Basics

### Prerequisites

It is important to remember several statistical distributions, tools, and facts. Most of them have to do with the **normal distribution**. If $X$ is normally distributed with mean $\mu$ and standard deviation $\sigma$:

$$X \sim N(\mu, \sigma)$$

```{r}
library(tidyverse)

# plot a theoretical normal distribution without any data

# our "geom" here is the "stat_function()" command, which draws statistical functions

normal.pdf <- ggplot(data = data.frame(x = -4:4), # our "data" is just a sequence of x from -4 to 4
                     aes(x = x))+
  stat_function(fun = dnorm, # dnorm is the normal distributions
                args = list(mean = 0, sd = 1), # set mean = 0 and sd = 1
                color="blue")+
  scale_x_continuous(breaks = seq(-4,4,1))+
    xlab("X (or Z-score)")+ylab("p(X)")+
  theme_light()

normal.pdf
```

Then recall the **68-95-99.7 empirical rule**, that:

- $P(\mu-1 \sigma \leq X \leq \mu+ 1\sigma) \approx 0.68$
- $P(\mu-2 \sigma \leq X \leq \mu+ 2\sigma) \approx 0.95$
- $P(\mu-3 \sigma \leq X \leq \mu+ 3\sigma) \approx 0.997$

Again, in English: "68% of the observations fall within 1 standard deviation of the mean; 95% fall within 2 standard deviations of the mean, and 99.7% fall within 3 standard deviations of the mean."

If we have the **standard normal distribution** with mean 0 and standard deviation 1:

```{r, echo=F}

# write functions to shade portions of distribution

shade68 <- function(x) {
    y <- dnorm(x, mean = 0, sd = 1)
    y[x < -1 | x > 1] <- NA
    return(y)
}

shade95 <- function(x) {
    y <- dnorm(x, mean = 0, sd = 1)
    y[x < -2 | x > 2] <- NA
    return(y)
}

shade99.7 <- function(x) {
    y <- dnorm(x, mean = 0, sd = 1)
    y[x < -3 | x > 3] <- NA
    return(y)
}

normal.pdf + 
  stat_function(fun = shade99.7, geom="area", fill="#e0ecf4", alpha=0.6)+
  stat_function(fun = shade95, geom="area", fill="#9ebcda", alpha=0.6)+
  stat_function(fun = shade68, geom="area", fill="#8856a7", alpha=0.6)+
  annotate("segment", x = -3, xend = 3, y = 0.15, yend = 0.15, colour = "#000000", size=0.5, linetype = 2, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))+
  annotate("text", x = 0, y = 0.16, label = "99.7%", color="#000000")+
  annotate("segment", x = -2, xend = 2, y = 0.20, yend = 0.20, colour = "#000000", size=0.5, linetype = 2, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))+
  annotate("text", x = 0, y = 0.21, label = "95%", color="#000000")+
  annotate("segment", x = -1, xend = 1, y = 0.25, yend = 0.25, colour = "#000000", size=0.5, linetype = 2, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))+
  annotate("text", x = 0, y = 0.26, label = "68%", color="#000000")
```

Again, we can *standardize* any normally-distributed random variable by finding the **Z-score** of each observation:

$$Z=\frac{X_i - \mu}{\sigma}$$

This ensures the mean will be 0 and standard deviation will be 1. Thus, $Z$ is the number of standard deviations above $(+)$ or below $(-)$ the mean an observation is.

We can use $Z$-scores to find the probability of any range of observations of $X$ occuring in the distribution.

```{r, echo = TRUE}
pnorm(-2, mean = 0, sd = 1, lower.tail = TRUE) # area to left of -2
pnorm(2, mean = 0, sd = 1, lower.tail = TRUE) # area to left of 2
pnorm(2, mean = 0, sd = 1, lower.tail = FALSE) # area to RIGHT of 2
pnorm(2, mean = 0, sd = 1, lower.tail = TRUE)- pnorm(-2, mean = 0, sd = 1, lower.tail = TRUE) # area between -2 and 2
```

### The Central Limit Theorem

Inferential statistics can be summarized in 2 sentences:

> There are unknown **parameters** that describe a **population** distribution that we want to know. We use **statistics** that describe a **sample** to *estimate* the population parameters.

Recall there is an element of randomness in our sample statistics due to **sampling variability**. For example, if we take the mean of one sample, $\bar{x}$, and then take the mean of a *different* sample, the $\bar{x}$'s will be *slightly different*. We can concieve of a distribution of $\bar{x}$'s across many different samples, and this is called the **sampling distribution** of the statistic $(\bar{x})$.

Via the sampling distribution, the sample statistic $(\bar{X})$ itself is distributed with

- mean $E[\bar{x}]=\mu_X$ (the true population mean)
- standard deviation $\sigma_{\bar{x}}$

**Central Limit Theorem**: *with large enough sample size ($n\geq30$), the sampling distribution of a sample statistic is approximately normal*^[If samples are i.i.d. (independently and identically distributed if they are drawn from the same population randomly and then replaced) we don't even need to know the population distribution to assume normality]

Thus, the **sampling distribution of the sample mean** ($\bar{x}$):

$$\bar{X} \sim \left(\mu_X, \frac{\sigma_X}{\sqrt{n}}\right)$$

The second term we call the **standard error of the sample mean**^[Instead of the "standard deviation". "Standard error" refers to the sampling variability of a sample statistic, and is always talking about a sampling distribution.]. Note that it takes the true standard deviation of the population ($\sigma_X$)^[Which we need to know! We often do not know it!] and divides it by the square root of the sample size, $\sqrt{n}$.

Thus **if we know the true population standard deviation** ($\sigma_X$) then we can simply use the normal distribution for confidence intervals and hypothesis tests of a sample statistic. Since we often do not, we need to use another distribution for inferential statistics, often the $t$-distribution.

### If We Don't Know $\sigma$: The Student's $t$-Distribution

We rarely, if ever, know the true population standard deviation for variable $X$, $\sigma_X$. Additionally, we sometimes have sample sizes of $n < 30$. If either of these conditions are true, we cannot use leverage the Central Limit Theorem and simplify with a standard normal distribution. Instead of the normal distribution, we use a **Student's t-Distribution**^["Student" was the penname of William Sealy Gosset, who has one of the more interesting stories in statistics. He worked for Guiness in Ireland testing the quality of beer. He found that with small sample sizes, normal distributions did not yield accurate results. He came up with a more accurate distribution, and since Guiness would not let him publish his findings, published it under the pseudonym of "Student."]

$t$ is functionally equivalent to the idea of a $Z$-score, with some slight modifications:

$$t = \cfrac{\bar{x}-\mu}{\left(\frac{s}{\sqrt{n}}\right)}$$
- $\bar{x}$ is our estimated statistic (e.g. sample mean)
- $\mu$ is the true population parametner (e.g. population mean)
- $s$ is the sample standard deviation
- $n$ is the sample size

$t$-scores similarly measure the number of standard deviations an observation is above or below the mean.

The other main difference between normal distributions/$Z$-scores and $t$ distributions /$t$-scores is that $t$ distributions have $n-1$ degrees of freedom.^[**Degrees of freedom**, $df$ are the number of independent values used for the calculation of a statistic minus the number of other statistics used as intermediate steps. For sample standard deviation $s$, we use $n$ deviations $(x_i-\bar{x})$ and 1 parameter $(\bar{x})$, hence $df=n-1$]

$$t \sim t_{n-1}$$

```{r, echo=F}
library("viridis")
ggplot(data = data.frame(x = -4:4), # our "data" is just a sequence of x from -4 to 4
                     aes(x = x))+
  stat_function(fun = dt,
                args = list(df=1),
                aes(color="df=1"))+
  stat_function(fun = dt,
                args = list(df=10),
                aes(color="df=10"))+
  stat_function(fun = dt,
                args = list(df=20),
                aes(color="df=25"))+
  stat_function(fun = dt,
                args = list(df=50),
                aes(color="df=50"))+
  scale_x_continuous(breaks = seq(-4,4,1))+
  xlab("t")+ylab("p(X)")+
  scale_color_viridis_d()+
  labs(x = "t",
       y = "p(t)",
       color = "df")+
  theme_light()
```

The standard $t$-distribution looks normal-ish, with a mean of 0, but with more area in the tails of the distribution. The exact shape depends on the degrees of freedom $(n-1)$. As $\uparrow n$, $\uparrow df$, the $t$-distribution approximates a normal distribution.

By convention, in regression we *always* use $t$-distributions for confidence intervals and hypothesis tests. For nearly all of the confidence intervals and hypothesis tests below, we functionally replace $Z$ with $t$.

## Confidence Intervals

A **confidence interval** describes the range of estimates for a population parameter in the form:

$$(\text{estimate} - \text{margin of error}, \, \text{estimate} + \text{margin of error})$$

Our *confidence level* is $1-\alpha$

- $\alpha$ again is the "significance level", the probability that the true population parameter is *not* within our confidence interval^[Equivalently, $\alpha$ is the probability of a **Type I error**: a false positive finding where we incorrectly reject a null hypothesis when it the null hypothesis is in fact true.]
- Typical confidence levels: 90%, 95%, 99%

> A confidence interval tells us that if we were to conduct many samples, ($1-\alpha$)% would contain the true population parameter within the interval

To construct a confidence interval, we do the following:

1. **Calculate the sample statistic.**

2. **Find $Z$-score that corresponds to desired confidence level.**^[Of course, if we don't know the population $\sigma$, we need to use the $t$-distribution and find critical $t$-scores instead of $Z$-scores. See above.] We need to find what are called the **"critical values"** of $Z$, which we will call $Z_{0.5\alpha}$ on the normal distribution that puts ($1-\alpha$) probability between $\pm Z_{0.5\alpha}$ and $0.5\alpha$ in each of the tails of the distribution. The distribution would look like this:

```{r, echo=F}
shadeabstract <- function(x) {
    y <- dnorm(x, mean = 0, sd = 1)
    y[x < -2 | x > 2] <- NA
    return(y)
}

conf_int_abstract <- ggplot(data = data.frame(x = -4:4), # our "data" is just a sequence of x from -4 to 4
                     aes(x = x))+
  stat_function(fun = dnorm, # dnorm is the normal distributions
                args = list(mean = 0, sd = 1), # set mean = 0 and sd = 1
                color="blue")+
  stat_function(fun = shadeabstract, geom="area", fill="skyblue", alpha=0.6)+
  annotate("segment", x = -2, xend = 2, y = 0.15, yend = 0.15, colour = "#000000", size=0.5, linetype = 2, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))+
  annotate("text", x = 0, y = 0.16, label = expression(1-alpha), color="#000000")+
  annotate("segment", x = -4, xend = -2, y = 0.1, yend = 0.1, color = "#000000", size=0.5, linetype = 2, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))+
  annotate("text", x = -3, y = 0.11, label = expression(0.5*alpha), color="#000000")+
  annotate("segment", x = 2, xend = 4, y = 0.1, yend = 0.1, color = "#000000", size=0.5, linetype = 2, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))+
  annotate("text", x = 3, y = 0.11, label = expression(0.5*alpha), color="#000000")+
  scale_x_continuous(breaks = c(-2,2),
                     labels = c(expression(-Z[0.5*alpha]),expression(Z[0.5*alpha])))+
    xlab("Z")+ylab("p(X)")+
  theme_light()

conf_int_abstract
```

- The confidence interval between the two $Z$-scores $-Z_{0.5\alpha}$ and $Z_{0.5\alpha}$ contains the desired $(1-\alpha)%$ of observations
- The area beyond each $Z$-score contains $0.5\alpha$% of observations in each direction, for a total of $\alpha$% beyond the critical values of $Z$

Note that the image above is abstract. So for example, if we wanted a (typical) **95% confidence interval** with $\alpha=0.05$, the critical value(s) of $Z$ are $\pm 1.96$^[Note this is the precise value behind the rule of thumb that 95% of observations fall within 2 standard deviations of the mean!], and looking on the distribution:

```{r, echo=F}
shade1.96 <- function(x) {
    y <- dnorm(x, mean = 0, sd = 1)
    y[x < -1.96 | x > 1.96] <- NA
    return(y)
}

conf_int_abstract <- ggplot(data = data.frame(x = -4:4), # our "data" is just a sequence of x from -4 to 4
                     aes(x = x))+
  stat_function(fun = dnorm, # dnorm is the normal distributions
                args = list(mean = 0, sd = 1), # set mean = 0 and sd = 1
                color="blue")+
  stat_function(fun = shade1.96, geom="area", fill="skyblue", alpha=0.6)+
  annotate("segment", x = -1.96, xend = 1.96, y = 0.15, yend = 0.15, colour = "#000000", size=0.5, linetype = 2, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))+
  annotate("text", x = 0, y = 0.16, label = "0.95", color="#000000")+
  annotate("segment", x = -4, xend = -1.96, y = 0.1, yend = 0.1, color = "#000000", size=0.5, linetype = 2, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))+
  annotate("text", x = -3, y = 0.11, label = "0.025", color="#000000")+
  annotate("segment", x = 1.96, xend = 4, y = 0.1, yend = 0.1, color = "#000000", size=0.5, linetype = 2, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="both", type="closed"))+
  annotate("text", x = 3, y = 0.11, label = "0.025", color="#000000")+
  scale_x_continuous(breaks = c(-1.96,1.96),
                     labels = c(-1.96, 1.96))+
    xlab("Z")+ylab("p(X)")+
  theme_light()

conf_int_abstract
```

The critical values of $Z$ are often given in $Z$-tables, which you can find in classical statistics textbooks or online. Critical values  of $Z$ for common confidence intervals values are well known:

| Confidence Level | $\alpha$ | $\pm Z_{0.5\alpha}$ |
|------------------|----------|---------------------|
| 90%              | 0.10     | $\pm 1.645$         |
| 95%              | 0.05     | $\pm 1.96$          |
| 99%              | 0.01     | $\pm 2.58$          |

3. **Calculate the margin of error (MOE)**

The margin of error is the critical value of $Z$ times the standard error of the estimate ($\sigma$).

$$MOE=Z_{0.5\alpha}\sigma$$

4. **Construct the confidence interval**

The confidence interval is simply our estimate plus and minus the margin of error.

$$CI = \left([\bar{x}-Z_{0.5}\sigma_{\bar{x}}], \, [\bar{x}+Z_{0.5}\sigma_{\bar{x}}]\right)$$

5. Intepret the confidence interval in the context of the problem

> "We estimate with [1-alpha]% confidence that the true [population parameter] is between [lowerbound] and [upperbound]".

## Hypothesis Testing

### 2019-10-03 static/homeworks/03-problem-set-answers.Rmd
Add 12 class content

--- 2019-09-20 static/homeworks/02-problem-set-pdf.Rmd
@@
 
-1.
+## Question 1
 
-a. In your own words, explain the difference between endogeneity and exogeneity.^[Hint: you may want to look back at class 1.2 for this question.]
-b. In your own words, explain how conducting a randomized controlled trial helps to identify the causal connection between two variables.
+### Part A
+**In your own words, explain the difference between endogeneity and exogeneity.**^[Hint: you may want to look back at class 1.2 for this question.]
 
-\clearpage
+The $R^2$ is a measure of how well the OLS regression line "fits" our observed data points.
+It is a measure of the share of the total variation in $Y$ (TSS) that is explained by the variation from our model (ESS), where:
 
-2.
+$$R^2=\frac{ESS}{TSS} = \frac{\displaystyle \sum^n_{i=1} (\hat{Y_i}-\bar{Y})^2}{\displaystyle \sum^n_{i=1}(Y_i-\bar{Y})^2}$$
 
-a. In your own words, explain what (sample) standard deviation *means*.
-b. In your own words, explain how (sample) standard deviation *is calculated.* You may also write the formula, but it is not necessary.
+Equivalently, you can also estimate $R^2$ as:
 
-\clearpage
+$$R^2=1-\frac{SSE}{TSS} = 1 - \left(\frac{\displaystyle \sum^n_{i=1} \hat{u_i}^2}{\displaystyle \sum^n_{i=1}(Y_i-\bar{Y})^2}\right)$$
+
+or, most simply:
+
+$$R^2=[cor(X,Y)]^2$$
+
+The closer $R^2$ is to 1, the better the fit, the closer to 0, the poorer the fit. Low $R^2$ tells us that there are better models, including more variables, that explain the variation in Y.
+
+### Part B
+
+**In your own words, explain how conducting a randomized controlled trial helps to identify the causal connection between two variables.**
+
+## 2.
+
+### a.
+
+In your own words, explain what (sample) standard deviation *means*.
+
+### b.
+
+In your own words, explain how (sample) standard deviation *is calculated.* You may also write the formula, but it is not necessary.
 

### 2019-10-03 static/r/11-practice-answers.Rmd
Add 12 class content

---
title: "2.4 OLS: Precision and Diagnostics - R Practice (Answers)"
author: "Ryan Safner"
date: "ECON 480 - Fall 2019"
output:
  html_document:
    df_print: paged
    theme: simplex
    toc: true 
    toc_depth: 3
    toc_float: true
    code_folding: show
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting Set Up

Before we begin, start a new file with `File` $\rightarrow$ `New File` $\rightarrow$ `R Script`. As you work through this sheet in the console in `R`, also add (copy/paste) your commands that work into this new file. At the end, save it, and run to execute all of your commands at once.

First things first, load `tidyverse`.

```{r}
library("tidyverse")
```

```{r}
congress<-read_csv("../data/congress-trump-score.csv")
```

```{r}
congress %>%
  count(party)
```

```{r}
glimpse(congress)
```

```{r}
congress %>%
  count(congress)
```

```{r}
congress_tidy <- congress %>%
  filter(congress!=0) %>%
  mutate(agree_pct = agree_pct * 100)
```

```{r}
congress_tidy %>%
  mutate(pol_party = case_when(
    party %in% c("Democrat", "D") ~ "Democrat",
    party %in% c("Republican", "R") ~ "Republican",
    party %in% c("Independent", "I") ~ "Independent"
  )) %>%
  count(pol_party)

congress_tidy <- congress_tidy %>%
  mutate(pol_party = case_when(
    party %in% c("Democrat", "D") ~ "Democrat",
    party %in% c("Republican", "R") ~ "Republican",
    party %in% c("Independent", "I") ~ "Independent"
  ))

```

# Warm Up to `dplyr` With `gapminder`

## 
#### 1. Load `gapminder`. `select()` the variables `year`, `lifeExp`, `country`.

```{r}

party_colors <- c("Democrat" = "blue", "Republican" = "red", "Independent" = "gray")

ggplot(data = congress_tidy)+
  aes(x = net_trump_vote, y=agree_pct)+
  geom_point(aes(color = pol_party))+
  geom_smooth(method="lm", color = "black")+
  scale_colour_manual("Parties", values = party_colors)+
  theme_classic()
```

```{r}
congress_tidy %>%
  summarize(cor = cor(net_trump_vote, agree_pct))

congress_cor<-congress_tidy %>%
  select(agree_pct, net_trump_vote)
```

```{r}
reg<-lm(agree_pct ~ net_trump_vote, data = congress_tidy)
summary(reg)
```

```{r}
library(broom)
reg %>%
  tidy()
```

```{r, results="asis"}
library(equatiomatic)
extract_eq(reg, use_coefs = TRUE,coef_digits = 2,fix_signs = TRUE)
```

```{r}
reg %>%
  glance()
```

```{r}
reg_aug<-reg %>%
  augment()
```

```{r}
reg_aug %>%
  head()
```

```{r}
ggplot(data = reg_aug)+
  aes(x = net_trump_vote, y=.resid)+
  geom_point()+
  geom_hline(yintercept = 0, color = "red")+
  theme_classic()
```

```{r}
ggplot(data = reg_aug)+
  aes(x = .resid)+
  geom_histogram(color="white")+
  theme_classic()
```

```{r}
congress %>%
  count(chamber)
```

```{r}
ggplot(data = reg_aug)+
  aes(x = .resid)+
  geom_density()
```

```{r}
library(car)
outlierTest(reg)
```

```{r}
congress_tidy %>%
  slice(1070)
```

```{r}
reg_dfbetas<-as.tibble(dfbetas(reg))#

reg_dfbetas
```

```{r}
reg_dfbetas %>%
  arrange(desc(net_trump_vote))
```

# Check for Heteroskedasticity

```{r}
library(lmtest)
bptest(reg)
```

### 2019-10-03 static/r/11-r-practice-rmd.Rmd
Add 12 class content

--- 2019-10-01 content/r/11-r-practice.Rmd
@@
 ---
-title: "2.4 OLS: Precision and Diagnostics - R Practice (Answers)"
-editor_options: 
-  chunk_output_type: console
+title: "2.4 OLS: Precision and Diagnostics - R Practice"
+author: "YOUR NAME"
+date: "`r Sys.Date()`"
+output:
+  html_document:
+    df_print: paged
+    #theme: 
+    toc: true 
+    toc_depth: 3
+    toc_float: true
+    code_folding: show
+    highlight: tango
 ---
 
-```{r message=FALSE, warning=FALSE, include=FALSE}
-library(tidyverse)
-library(pander)
-source(file.path(here::here(), "lib", "graphics.R"))
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
 ```
 
-Our [data](https://github.com/fivethirtyeight/data/tree/master/congress-trump-score) comes from fivethirtyeight's [Trump Congress tracker](https://projects.fivethirtyeight.com/congress-trump-score/).
+# Set Up
+To minimize confusion, I suggest creating a new `R Project` (e.g. `regression_practice`) and storing any data in that folder on your computer.
 
-- [View Project on R Studio Cloud](https://rstudio.cloud/project/575252)
+Alternatively, I have made a project in R Studio Cloud that you can use (and not worry about trading room computer limitations), with the data already inside (you will still need to assign it to an object).
+
+- [<i class="fas fa-cloud"></i> View Project on R Studio Cloud](https://rstudio.cloud/project/575252)
 
 ## Question 1
+Our [data](https://github.com/fivethirtyeight/data/tree/master/congress-trump-score) comes from fivethirtyeight's [Trump Congress tracker](https://projects.fivethirtyeight.com/congress-trump-score/).
 Download and read in (`read_csv`) the data.
+
+- [<i class="fas fa-table"></i> `congress-trump-score.csv`](/data/congress-trump-score.csv)
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 Look at the data with `glimpse()`.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 4
 Find the correlation between `agree_pct` and `net_trump_vote`.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part A
 What is $\hat{\beta_0}$? What does it mean in the context of our question?
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part C
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part D
 What is the $SER$? What does it mean?
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part B
 Run the function `tidy()` on your regression object (saved in question 5). Save this as an object and then look at it.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part D
 Now run the `augment()` function on your original regression object, and save this as an object. Look at it. What does it show you?
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part B
@@
 Feel free to add a horizontal line at 0 with `geom_hline(yintercept=0)`.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 9
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part B
 Install and load the `lmtest` package and run `bptest` on your regression object. Was the data heteroskedastic or homoskedastic?
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part D
 Now let's see this in a nice regression table. Use `huxreg()` again, but add both your original regression and your regression saved in part C. Notice any changes?
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part C
-Look in your original data to match this outlier with an observation. Hint: use the `slice()` command, as the outlier test gave you an observation number!
+Look in your original data to match this outlier with an observation. Hint: use the `slice()` command, as the outlier test gave you an observation (row) number!
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
-This data is still a bit messy. We'd also probably like to plot our scatterplots with colors for Republican and Democratic party. Or sort by the House and the Senate.
+This data is still a bit messy.
+Let's check in on your `tidyverse` skills again!
+For example, we'd probably like to plot our scatterplots with colors for Republican and Democratic party.
+Or plot by the House and the Senate.
+
+### Part A
+First, the variable `congress` (session of Congress) seems a bit off. Get a `count()` of `congress`.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
+### Part B
+Let's get rid of the `0` values for `congress` (someone made a mistake coding this, probably).
+Also, while we're at it, let's take `agree_pct` and `mutate` a variable that is a proper percentage (i.e.
+`*100`).
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
+### Part C
+The variable `party` is also quite a mess.
+`count()` by `party` to see.
+Then let's `mutate` a variable to make a better measure of political party - just `"Republican"`, `"Democrat"`, and `"Independent"`.
+Try doing this with the `case_when()` command (as your `mutate` formula).^[The syntax for `case_when()` is to have a series of `condition ~ "Outcome"`, separated by commas.
+For example, one condition is to assign both `"Democrat"` and `"D"` to `"Democrat"`, as in `party %in% c("Democrat", "D") ~ "Democrat"`.
+You could also do this with a few `ifelse()` commands, but that's a bit more awkward.] When you're done `count()` by your new party variable to make sure it worked.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
+### Part D
+Now plot a scatterplot (same as Question 3) and set `color` to your party variable.
+Notice `R` uses its own default colors, which don't match to the actual colors these political parties use!
+Make a vector where you define the party colors as follows: `party_colors <- c("Democrat" = "blue", "Republican" = "red", "Independent" = "gray")`.
+Then, run your plot again, adding the following line to customize the colors `+scale_colour_manual("Parties", values = party_colors)`.^[`"Parties"` is the title that will show up on the legend, feel free to edit it, or remove the legend with another layer `+guides(color = F)`.]
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
+### Part E
+Now facet your scatterplot by `chamber`.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---

### 2019-10-03 static/slides/12-slides.Rmd
Add 12 class content

---
title: "2.6: Inferential Statistics I"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))
```

```{r regression-setup, echo=F, results="hide"}

library(haven)
CASchool<-read_dta("../data/caschool.dta")

# run regression of testscr on str
school_reg <- lm(testscr ~ str, 
                 data = CASchool)

library(broom)
school_reg_tidy <- tidy(school_reg,
     conf.int = TRUE) # add confidence intervals
CASchool_aug <- augment(school_reg)

library(equatiomatic)
extract_eq(school_reg, use_coefs = TRUE,coef_digits = 2,fix_signs = TRUE)

```

# The Sampling Distribution of $\hat{\beta_1}$

.pull-left[

$$\hat{\beta_1} \sim N(E[\hat{\beta_1}], \sigma_{\hat{\beta_1}})$$

1. $E[\hat{\beta_1}]$; the **center** of the distribution (2 classes ago)
    - $E[\hat{\beta_1}]=\beta_1$<sup><span class="red">1</span></sup>

2. $\sigma_{\hat{\beta_1}}$; how **precise** is our estimate? (last class)
    - Variance $\sigma^2_{\hat{\beta_1}}$ or standard error $\sigma_{\hat{\beta_1}}$

]

.pull-right[

```{r, fig.retina=3}
ggplot(data = tibble(x=-4:4))+
  aes(x = x)+
  stat_function(fun = dnorm, size=2, color="blue")+
  geom_label(x=1, y=dnorm(1), label=expression(sigma[hat(beta[1])]==1), color="blue")+
  geom_segment(aes(x=0,xend=0, y=0, yend=0.4), linetype="dashed")+
  scale_x_continuous(breaks = 0,
                     labels = expression(E(hat(beta[1]))))+
  labs(x = expression(hat(beta[1])),
       y = "Probability")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)+
  stat_function(fun = dnorm, args=list(mean = 0, sd = 2), size=2, color="red")+
  geom_label(x=2, y=dnorm(2,0,2), label=expression(sigma[hat(beta[1])]==2), color="red")

```

]
.footnote[<sup>.red[1]</sup> Under the 4 assumptions about $u$ (particularly, $cor(X,u)=0)$.
]

---

# Recall: The Two Big Problems with Data

.pull-left[

- We use econometrics to .onfire[identify] causal relationships and make .onfire[inferences] about them

1. Problem for .onfire[identification]: .shout[endogeneity]
    - $X$ is **exogenous** if its variation is *unrelated* to other factors $(u)$ that affect $Y$
    - $X$ is **endogenous** if its variation is *related* to other factors $(u)$ that affect $Y$

2. Problem for .onfire[inference]: .shout[randomness]
    - Data is random due to **natural sampling variation**
    - Taking one sample of a population will yield slightly different information than another sample of the same population

]

.pull-right[
.center[
![:scale 55%](https://www.dropbox.com/s/v5vwsadw5vs448t/causality.jpg?raw=1)

![:scale 55%](https://www.dropbox.com/s/phhet82tcnuegxp/randomimage.jpg?raw=1)]
]

---

# Distributions of the OLS Estimators

- OLS estimators $(\hat{\beta_0}$ and $\hat{\beta_1})$ are computed from a finite (specific) sample of data

- Our OLS model contains **2 sources of randomness**:

--

- .onfire[*Modeled* randomness]: $u$ includes all factors affecting $Y$ *other* than $X$
    - different samples will have different values of those other factors $(u_i)$

--

- .onfire[*Sampling* randomness]: different samples will generate different OLS estimators
    - Thus, $\hat{\beta_0}, \hat{\beta_1}$ are *also* **random variables**, with their own <span class=shout>sampling distribution</span>

---

# Inferential Statistics and Sampling Distributions

.pull-left[

- **Inferential statistics** analyzes a **sample** to make inferences about a much larger (unobservable) **population**

- .shout[Population]: all possible individuals that match some well-defined criterion of interest (people, firms, cities, etc)
  - Characteristics about (relationships between variables in) populations are called **parameters**

- .shout[Sample]: some portion of the population of interest to *represent the whole*
  - Samples generate **statistics** used to *estimate* population parameters

- We almost never can directly study the population, so we *model* it with our samples 
]

.pull-right[
.center[
![](https://www.dropbox.com/s/17cb7ceqipgi8fs/citymodel.jpg?raw=1)
]
]

---

# Inference in Econometrics: The Big Picture

$$\text{Sample} \underbrace{\rightarrow}_{\text{statistical inference}} \text{Population} \underbrace{\rightarrow}_{\text{causal indentification}} \text{Unobserved Parameters}$$

- We want to .whisper[identify] causal relationships between **population** variables
    - Logically first thing to consider
    - **Endogeneity problem**

- We'll use **sample** *statistics* to .whisper[infer] something about population *parameters*
    - In practice, we'll only ever have a finite *sample distribution* of data
    - We *don't* know the *population distribution* of data
    - **Randomness problem**

---

# Two Methods of Statistical Inference

1. .shout[Estimation]: use our sample data to construct a point estimate of a population parameter and subject it to a .shout[hypothesis test]

--

2. .shout[Confidence interval]: use our sample data to construct a *range* for the population parameter

--

- First method is more common, but second is still widely acknowledged

- Both will give you similar results
    - Tradeoff of accuracy vs. precision

- Note *statistical* inference is different than *causal* inference!

---

class: inverse, center, middle

# Hypothesis Testing

---

# Estimation and Hypothesis Testing I

- We have already used statistics to **estimate** a relationship between $X$ and $Y$
    - OLS estimators $\hat{\beta_0}$ and $\hat{\beta_1}$ of the true population $\beta_0$ and $\beta_1$

- We want to **test** if these estimates are **statistically significant** and they describe the population
    - This is the "bread and butter" of inferential statistics and the purpose of regression

--

.content-box-green[
.green[**Examples**]:
- Does reducing class size actually improve test scores?
- Do more years of education increase your wages? 
- Is the gender wage gap between men and women really $0.77? 
]

--

- .whisper[All modern science is built upon statistical hypothesis testing, so understand it well!]

---

# Estimation and Hypothesis Testing II

- Note, we can test a lot of hypotheses about a lot of population parameters, e.g.
    - A population mean $\mu$ 
      - <span class="green">**Example**: average height of adults</span>
    - A population proportion $p$
      - <span class="green">**Example**: percent of voters who voted for Trump</span>
    - A difference in population means $\mu_A-\mu_B$
      - <span class="green">**Example**: difference in average wages of men vs. women</span>
    - A difference in population proportions $p_A-p_B$
      - <span class="green">**Example**: difference in percent of patients reporting symptoms of drug A vs B</span>
    - See all the possibilities in glorious detail in today's Class Notes

- We will focus only on hypotheses about the .onfire[population regression slope] $(\hat{\beta}_1)$, i.e. the .onfire[causal effect]<sup>.red[1]</sup> of $X$ on $Y$

.footnote[<sup>1</sup> With a model this simple, it's almost certainly **not** causal, but this is the ultimate direction we are heading...]

---

# Null and Alternative Hypotheses I

- All scientific inquiries begin with a .shout[null hypothesis] $(H_0)$ that proposes a specific value of a population parameter
    - Notation: add a subscript 0: $\beta_{1,0}$ (or $\mu_0$, $p_0$, etc)

--

- We suggest an .shout[alternative hypothesis] $(H_a)$, often the one we hope to verify
    - Note, can be multiple alternative hypotheses: $H_1, H_2, \ldots , H_n$

--

- Ask: .onfire["Does our data (sample) give us sufficient evidence to reject `\\(H_0\\)` in favor of `\\(H_a\\)`?"]
    - Note: **the test is *always* about** $\mathbf{H_0}$! 
    - See if we have sufficient evidence to reject the status quo

---

# Null and Alternative Hypotheses II

- Null hypothesis assigns a value (or a range) to a population parameter
    - e.g. $\beta_1=2$ or $\beta_1 \leq 20$
    - **Most common null hypothesis is $\beta_1=0$** $\implies$ $X$ has no effect on $Y$ (no slope for a line)
    - Note: always an equality!

--

- Alternative hypothesis must mathematically *contradict* the null hypothesis
    - e.g. $\beta_1 \neq 2$ or $\beta_1 > 20$ or $\beta_1 \neq 0$
    - Note: always an inequality!

--

- Alternative hypotheses come in two forms:
    1. **One-sided alternative**: $\beta_1 >H_0$ or $\beta_1< H_0$
    2. **Two-sided alternative**: $\beta_1 \neq H_0$
        - Note this means either $\beta_1 < H_0$ or $\beta_1 > H_0$

---

# Components of a Valid Hypothesis Test

- All statistical hypothesis tests have the following components:

1. A .onfire[null hypothesis], $H_0$**

2. An .onfire[alternative hypothesis], $H_a$**

3. A .onfire[test statistic] to determine if we reject $H_0$ when the statistic reaches a "critical value"
    - Beyond the critical value is the "rejection region", sufficient evidence to reject $H_0$

4. A .onfire[conclusion] whether or not to reject $H_0$ in favor of $H_a$

---

# Type I and Type II Errors I

.pull-left[

- Any sample statistic (e.g. $\hat{\beta_1}$) will rarely be exactly equal to the hypothesized population parameter (e.g. $\beta_1$)

- Difference between observed statistic and true paremeter could be because:

1. Parameter is *not* the hypothesized value $(H_0$ is *false*)

2. Parameter is truly the hypothesized value $(H_0$ is *true*) but *sampling variability* gave us a different estimate

- .onfire[We cannot distinguish between these two possibilities with any certainty]

]

.pull-right[
.center[
![](https://www.dropbox.com/s/95rh9ow982y5htb/error.png?raw=1)
]
]
---

# Type I and Type II Errors II

.pull-left[

- We can interpret our estimates probabilistically as commiting one of two types of error:

1. .shout[Type I error (false positive)]: rejecting $H_0$ when it is in fact true
    - Believing we found an important result when there is truly no relationship

2. .shoutcaps[Type II error (false negative)]: failing to reject $H_0$ when it is in fact false
    - Believing we found nothing when there was truly a relationship to find
]

.pull-right[
.center[
![](https://www.dropbox.com/s/95rh9ow982y5htb/error.png?raw=1)
]
]

---

# Type I and Type II Errors III

```{r, echo=FALSE, results="asis", eval=F}
library(kableExtra)
tribble(
  ~"", ~"", ~"Null is True", ~"Null is False",
  "Judgment", "Reject Null", "Type I Error", "Correct",
  "Judgment", "Don't Reject Null", "Correct", "Type II Error"
  ) %>%
  knitr::kable(.,format="html") %>%
  kable_styling(full_width = F) %>%
  add_header_above(c(" " = 2, "Truth" = 2)) %>%
  column_spec(c(1,2), bold=T) %>%
  collapse_rows(columns = 1, latex_hline = "major", valign = "middle")
```

.regtable[
```{r, echo=FALSE, results="asis"}
library(kableExtra)
tribble(
  ~"", ~"", ~"Null is True", ~"Null is False",
  "Judgment", "Reject Null", "TYPE I ERROR", "CORRECT",
  "Judgment", "Reject Null", "(False +)", "(True +)",
  "Judgment", "Don't Reject Null", "CORRECT", "TYPE II ERROR",
  "Judgment", "Don't Reject Null", "(True -)", "(False -)"
  ) %>%
  knitr::kable(.,format="html") %>%
  kable_styling(full_width = F) %>%
  add_header_above(c(" " = 2, "Truth" = 2)) %>%
  column_spec(c(1,2), bold=T) %>%
  collapse_rows(columns = c(1,2), latex_hline = "major", valign = "middle")

```
]

- Depending on context, committing one type of error may be more serious than the other

---

# Type I and Type II Errors IV

.regtable[
```{r, echo=FALSE, results="asis"}
library(kableExtra)
tribble(
  ~"", ~"", ~"Defendant is Innocent", ~"Defendant is Guilty",
  "Judgment", "Convict", "TYPE I ERROR", "CORRECT",
  "Judgment", "Convict", "(False +)", "(True +)",
  "Judgment", "Acquit", "CORRECT", "TYPE II ERROR",
  "Judgment", "Acquit", "(True -)", "(False -)"
  ) %>%
  knitr::kable(.,format="html") %>%
  kable_styling(full_width = F) %>%
  add_header_above(c(" " = 2, "Truth" = 2)) %>%
  column_spec(c(1,2), bold=T) %>%
  collapse_rows(columns = c(1,2), latex_hline = "major", valign = "middle")

```
]

- Anglo-American common law *presumes* defendant is innocent: $H_0$

--

- Jury judges whether the evidence presented against the defendant is plausible *assuming the defendant were in fact innocent*

--

- If highly improbable: sufficient evidence to reject $H_0$ and convict
    - Beyond a "reasonable doubt" that the defendant is innocent

---

# Type I and Type II Errors V

.left-column[

.center[
![:scale 80%](https://www.dropbox.com/s/bufqsl9ub03iru3/blackstone.png?raw=1)

William Blackstone

(1723-1780)

]
]

.right-column[

> "It is better that ten guilty persons escape than that one innocent suffer."

- Type I error is worse than a Type II error in law!

]

.source[Blackstone, William, 1765-1770, *Commentaries on the Laws of England*]

---

# Type I and Type II Errors VI

.content-box-blue[
.blue[**Let's Check**:]
For each of the following scenarios, identify the Type I error, Type II error, and which error is of greater consequence:
]

--
1. $H_0:$ the victim of an automobile accident is alive when he arrives at the Emergency Room

--

2. $H_0$: a rock climber's equipment is safe

--

3. $H_0$: a woman is not pregnant

--

4. $H_0$: a highway project will cost no more than $10 million

--

5. $H_0$: an experimental cancer drug has a cure rate of at least 75%

---

# Significance Level, $\alpha$, and Confidence Level $1-\alpha$

- The .shout[significance level, `\\(\alpha\\)`], is the probability of a **Type I error**

$$\alpha=P(\text{Reject } H_0 | H_0 \text{ is true})$$

--

- The .shout[confidence level] is defined as .shout[`\\((1-\alpha)\\)`]

- We often specify in advance an $\alpha$-level (0.10, 0.05, 0.01) with associated confidence level (90%, 95%, 99%)

--

- The probability of a **Type II error** is defined as $\beta$:

$$\beta=P(\text{Don't reject } H_0 | H_0 \text{ is false})$$

---

# $\alpha$ and $\beta$

.regtable[
```{r, echo=FALSE, results="asis"}
tribble(
  ~"", ~"", ~"Null is True", ~"Null is False",
  "Judgment", "Reject Null", "Type I Error", "Correct",
  "Judgment", "Reject Null", "alpha", "1-beta",
  "Judgment", "Don't Reject Null", "Correct", "Type II Error",
  "Judgment", "Don't Reject Null", "1-alpha", "beta"
  ) %>%
  knitr::kable(.,format="html") %>%
  kable_styling(full_width = F) %>%
  add_header_above(c(" " = 2, "Truth" = 2)) %>%
  column_spec(c(1,2), bold=T) %>%
  collapse_rows(columns = c(1,2), latex_hline = "major", valign = "middle")
```
]

---

# Power and p-values

- The statistical .shout[power of the test] is $1-\beta$, the probability of correctly rejecting $H_0$ when $H_0$ is in fact false (e.g. not convicting an innocent person)

$$\text{Power} = 1- \beta = P(\text{Reject }H_0|H_0 \text{ is false})$$

--

- The .shout[`\\(p\\)`-value] or .shout[significance probability] is the probability that, given the null hypothesis is true, the test statistic from a random sample will be at least as extreme as the test statistic of our sample

$$p(|H_0)$$

- If $p<\alpha$, the results are "**statistically significant**"

---

# p-Values and Statistical Significance

- After running our test, we need to make a *decision* between the competing hypotheses
- Compare $p$-value with *pre-determined* $\alpha$ (commonly, $\alpha=0.05$, 95% confidence level)
    - If $p<\alpha$: **statistically significant** evidence sufficient to *reject* $H_0$ in favor of $H_a$
    - If $p \geq \alpha$: *insufficient* evidence to reject $H_0$
        - Note this does **not** mean $H_0$ is true! We merely have *failed* to *reject* $H_0$

---

class: inverse, center, middle

# Digression: p-Values and the Philosophy of Science

---

# Hypothesis Testing and the Philosophy of Science I

.left-column[
.center[
![:scale 75%](https://www.dropbox.com/s/kssn3fwxgu8k457/rafisher.jpg?raw=1)

Sir Ronald A. Fisher

(1890&mdash;1962)
]
]

.right-column[

> "The null hypothesis is never proved or established, but is possibly disproved, in the course of experimentation. Every experiment may be said to exist only in order to give the facts a chance of disproving the null hypothesis."

1931, *The Design of Experiments*
]

---

# Hypothesis Testing and the Philosophy of Science I

.pull-left[

- Modern philosophy of science is largely based off of hypothesis testing and .whisper[falsifiability], which form the "Scientific Method"<sup>1</sup>
- For something to be "scientific", it must be .whisper[falsifiable], or at least .whisper[testable]
- Hypotheses can be *corroborated* with evidence, but always *tentative* until falsified by data in suggesting an alternative hypothesis
- .onfire["All swans are white"] is a hypothesis rejected upon discovery of a single black swan
]

.pull-right[

.center[
.polaroid[![](https://www.dropbox.com/s/d7i7m77iui5sdb8/blackswan.jpg?raw=1)]
]
]
.footnote[<sup>1</sup> Note: economics is a very different kind of "science" with a different methodology!]

---

# Hypothesis Testing I

- Hypothesis testing, confidence intervals, and p-values are probably the hardest thing to understand in statistics

.center[

<iframe src="https://fivethirtyeight.abcnews.go.com/video/embed/56150342" width="640" height="360" scrolling="no" style="border:none;" allowfullscreen></iframe>

[Fivethirtyeight: Not Even Scientists Can Easily Explain P-values](https://fivethirtyeight.com/features/not-even-scientists-can-easily-explain-p-values/)]

---

# Hypothesis Testing II

- Rigorous course on statistics (**ECMG 212** or **MATH 112**) will spend weeks going through different types of tests:
    - Sample mean; difference of means
    - Proportion; difference of proportions
    - Z-test vs t-test
    - 1 sample vs. 2 samples
    - $\chi^2$ test

- Fortunately, some statisticians realized ["**there is only one test**"](https://allendowney.blogspot.com/2011/05/there-is-only-one-test.html) and coded it that way

---

# Elements of a Hypothesis Test

1. Calculate a statistic, $\delta$<sup>1</sup>, from a sample of data
2. Simulate a world where $\delta$ is null
3. Examine $\delta$ in the null world
4. Calculate the probability that $\delta$ could exist in the null world
5. Decide if $\delta$ is statistically significant

.footnote[<sup>1</sup> $\delta$ can stand in for any test-statistic in any hypothesis test! For our purposes, $\delta$ is the slope of our regression sample, $\hat{\beta}_1$.]

---

# Elements of a Hypothesis Test

.center[
![:scale 80%](https://www.dropbox.com/s/jenilao59zrbswq/onetest.png?raw=1)

[Alan Downey: "There is still only one test"](https://allendowney.blogspot.com/2016/06/there-is-still-only-one-test.html)
]

---

# Hypothesis Testing with the infer Package I

- R naturally runs the following hypothesis test on any regression as part of `lm()`:

$$\begin{align*}
H_0: \; & \beta_1=0\\
H_1: \; & \beta_1 \neq 0
\end{align*}$$

- `infer` allows you to run through these steps manually to understand the process

---

# Hypothesis Testing with the infer Package II

1. `specify()` a model

--

2. `hypothesize()` the null

--

3. `generate()` simulations of the null world

--

4. `calculate()`

--

5. `visualize()`

---

# Hypothesis Testing with the infer Package III

.center[
![:scale 65%](https://www.dropbox.com/s/zz3h7wq1tr08dus/infer0jpeg.jpeg?raw=1)
]

---

# Hypothesis Testing with the infer Package III

.center[
![:scale 65%](https://www.dropbox.com/s/25axifo95oz5zs7/infer1jpeg.jpeg?raw=1)
]

---

# Hypothesis Testing with the infer Package III

.center[
![:scale 65%](https://www.dropbox.com/s/d9bnryxznfwe4fc/infer2jpeg.jpeg?raw=1)
]

---

# Hypothesis Testing with the infer Package III

.center[
![:scale 65%](https://www.dropbox.com/s/pc8733aurnouvsr/infer3jpeg.jpeg?raw=1)
]

---

# Hypothesis Testing with the infer Package III

.center[
![:scale 65%](https://www.dropbox.com/s/4wojnmworf406ku/infer4jpeg.jpeg?raw=1)
]

---

# Hypothesis Testing with the infer Package III

.center[
![:scale 65%](https://www.dropbox.com/s/6328hg5t03jbd0w/infer5jpeg.jpeg?raw=1)
]

---

# Test Statistic

- .whisper[Test statistic]: measure of how far what we observed in our sample is from what we would expect if the null hypothesis was true
    - Calculate from a distribution
- .whisper[Rejection region]: if the test statistic reaches a .onfire["critical value"], then we reject the null hypothesis

---

# Statistical Significance and Economic Significance

## Is a finding *significant* ?

## Is a finding *meaningful* ?

---

# How Big is "Big"?

.center[
![:scale 40%](https://www.dropbox.com/s/qtd2uwvr52jvkcu/drwhocontext.jpg?raw=1)
]

---

# Imagine a Null World

.center[
![:scale 70%](https://www.dropbox.com/s/x7b5azt5algjodg/parallelworld.jpeg?raw=1)
]

---

# Imagine a Null World

.pull-left[

# Our Sample

```{r, echo=F}
tidy(school_reg)
```
]

.pull-right[

# Another Sample

```{r, echo = F}
library(infer)
CASchool %>%
  specify(testscr ~ str) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1, type = "permute") %>%
  lm(testscr ~ str, data = .) %>%
  tidy()
```
]

---

# 1000 Parallel Universes

.left-code[

```{r infer-example, echo = T, eval = F}
library(infer)
simulations <- CASchool %>%
  specify(testscr ~ str) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "slope")

simulations %>%
  rename(sample = replicate,
         slope = stat)
```
]

.right-plot[
```{r ref.label="infer-example", echo = F}
```

]

---

# 1000 Parallel Universes

```{r echo = F}
plot <- simulations %>%
  visualize()+
  theme_minimal()+
  labs(x = "Slope",
       y = "Samples")

plot
```

---

# 1000 Parallel Universes

```{r}
obs_slope <- school_reg_tidy %>% 
  filter(term=="str") %>%
  pull(estimate) # from dplyr

plot2 <- simulations %>%
  visualize(obs_stat = obs_slope)

plot2+
  shade_p_value(obs_stat = obs_slope, 
                direction = "two_sided")+
  theme_minimal()+
  labs(x = "Slope",
       y = "Samples")
```

---

# 1000 Parallel Universes

```{r}
# get confidence interval values

school_reg_CI<- school_reg %>%
  tidy(., conf.int=T) %>%
  filter(term=="str") %>%
  select(conf.low, conf.high)
```

```{r}
plot2+
  shade_confidence_interval(endpoints = school_reg_CI)+
  theme_minimal()+
  labs(x = "Slope",
       y = "Samples")
```

---

# p-Value

```{r}
simulations %>%
  get_p_value(obs_stat = obs_slope, direction = "both")
```

> There's a 0% chance of seeing a slope at least as large as $-2.28$ in the null world where there is no relationship.

---

# What R Does

```{r}
tidy(school_reg)
```

- R calculates the $p$-value the "old-fashioned way", using a theoretical distribution instead of simulation
    - Generates a test statistic (`statistic`) and compares it to a critical value of the test statistic<sup>1</sup>
$$\text{test statistic} = \frac{\text{estimate} - \text{null hypothesis}}{\text{standard error of estimate}}$$

.footnote[<sup>1</sup> Rule of thumb: test statistic of 2 or larger is generally significant at the 95% level. This comes directly from the empirical rule that 95% of observations occur within 2 standard deviations of the mean! See the **class info** page for more.]

---

# What R Does II

```{r}
tidy(school_reg)
```

$$t = \frac{-2.28-0}{0.48} = -4.75$$
- Intepretation like $Z$-scores: number of standard deviations away from the center<sup>1</sup>
- $p$-value is the probability of getting a test statistic at least as large (in magnitude) as this one if the null hypothesis were true<sup>2</sup>

.footnote[<sup>1</sup> The "center" here is the null hypothesis value i.e. a slope of 0.
<sup>2</sup> Think of our simulated null world(s), the center of the sampling distribution was a slope of 0.]

---

# Confidence Intervals

```{r}
tidy(school_reg, conf.int=TRUE)
```

- We can also generate a .shout[confidence interval] for $\hat{\beta}_1$ that contain the true population slope ($\beta_1$) with a specified probability
    - Note confidence level $=1-\alpha$; commonly 95% ($\alpha=0.05$)
- .whisper[We are 95% confident that the true slope is between $-3.22$ and $-1.34$.]
    - Note if the confidence interval contains 0, *cannot* reject the null hypothesis of $\beta_1=0$! 
- See **class info** page for more on confidence intervals and how they are constructed

---

```{r, results="asis"}
equatiomatic::extract_eq(school_reg, use_coefs = TRUE,coef_digits = 2,fix_signs = TRUE)
```

---

# Testing Reg Output: huxtable

.font80[
.regtable[
```{r, echo=F}
library(huxtable)

hr <- huxreg("Model 1" = school_reg, # can name the columns in quotes
       coefs = c("Intercept" = "(Intercept)", # can rename the coefficients
                 "STR" = "str"),
       statistics = c("N" = "nobs", # statistics to show for regression, taken from glance()
                      "R Squared" = "r.squared",
                      "Adj. R Squared" = "adj.r.squared",
                      "SER" = "sigma",
                      "F" = "statistic"),
       number_format = 2, # number of decimals to round to
       bold_signif = 0.05) # can bold coeffs by significance level

hr
```
]
]

---

### 2019-10-03 content/class/12-class.Rmd
There's no practice for class 12, shorten slides 12

@@
 
-- `broom`: for tidy regression outputs, summary statistics, and adding $\hat{Y_i}$ and $\hat{u_i}$ into the dataframe
-- `huxtable`: to present regression output in a table with `huxreg()`
-- `lmtest`: for testing for heteroskedasticity in errors with `bptest()`
-- `car`: for testing for outliers with `outlierTest()`
-- `estimatr`: for calculating robust standard errors with `lm_robust()`
+- `infer`: for simulation for statistical inference
 
@@
 
-5. Intepret the confidence interval in the context of the problem
+5. **Intepret the confidence interval in the context of the problem**
 
 > "We estimate with [1-alpha]% confidence that the true [population parameter] is between [lowerbound] and [upperbound]".
-
-## Hypothesis Testing

### 2019-10-03 content/reading/12-reading.Rmd
There's no practice for class 12, shorten slides 12

---
title: "2.6 Inferential Statistics I - Readings"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-fullnote-bibliography-no-bib.csl
---

# Strongly Recommended

- <i class="fas fa-book"></i> [Ch. 4] in @Bailey
- [<i class="fas fa-external-link-square-alt"></i> Ch. 10 (optionally 8-9) in *Modern Dive* ](https://moderndive.netlify.com/10-inference-for-regression.html)
- [<i class="fab fa-youtube"></i> *Last Week Tonight with John Oliver*: Scientific Studies](https://www.youtube.com/watch?v=0Rnq1NpHdmw)
- [<i class="fas fa-external-link-square-alt"></i> Visualizing Probability and Inference ](https://seeing-theory.brown.edu/frequentist-inference/index.html)

Bailey teaches inferential statistics in the classical way (with reference to theoretical $Z$ and $t$ distributions, and $Z$ and $t$ tests). This is all valid. Again, you may wish to brush up with [Khan Academy](https://www.khanacademy.org/math/statistics-probability)^[From sampling distributions through significance tests, for this. Though the whole class is helpful!].

The latter "book" (also free online, like R4DS) uses the `infer` package to run simulations for inferential statistics. Chapter 10 is focused on regression (but I also recommend the chapters leading up to it, which are on inferential statistics broadly, using this method).

Also be sure to watch the excellent and hilarious discussion of the limits and misuses of scientific studies and statistical significance $(p$-values) in the *Last Week Tonight* clip.

The final link is a great website for visualizing basic statistic concepts like probability, distributions, confidence intervals, hypothesis tests, central limit theorem, and regression.

### 2019-10-03 static/slides/12-slides.Rmd
There's no practice for class 12, shorten slides 12

@@
 
-# Distributions of the OLS Estimators
+# Recall: Distributions of the OLS Estimators
 
@@
 
-# Inferential Statistics and Sampling Distributions
+# Recall: Inferential Statistics and Sampling Distributions
 
@@
 
-- We almost never can directly study the population, so we *model* it with our samples 
 ]
@@
 
-# Inference in Econometrics: The Big Picture
+# Recall: Inference in Econometrics: The Big Picture
 
@@
 
-1. A .onfire[null hypothesis], $H_0$**
+1. A .onfire[null hypothesis, `\\(H_0\\)`]
 
-2. An .onfire[alternative hypothesis], $H_a$**
+2. An .onfire[alternative hypothesis, `\\(H_a\\)`]
 
@@
 
-2. .shoutcaps[Type II error (false negative)]: failing to reject $H_0$ when it is in fact false
+2. .shout[Type II error (false negative)]: failing to reject $H_0$ when it is in fact false
     - Believing we found nothing when there was truly a relationship to find
@@
 
-.content-box-blue[
-.blue[**Let's Check**:]
-For each of the following scenarios, identify the Type I error, Type II error, and which error is of greater consequence:
+.center[
+![](https://www.dropbox.com/s/wf6i1462dadmvrw/errorspregnant.png?raw=1)
 ]
-
---
-1. $H_0:$ the victim of an automobile accident is alive when he arrives at the Emergency Room
-
---
-
-2. $H_0$: a rock climber's equipment is safe
-
---
-
-3. $H_0$: a woman is not pregnant
-
---
-
-4. $H_0$: a highway project will cost no more than $10 million
-
---
-
-5. $H_0$: an experimental cancer drug has a cure rate of at least 75%
 
@@
 - The .shout[confidence level] is defined as .shout[`\\((1-\alpha)\\)`]
-
-- We often specify in advance an $\alpha$-level (0.10, 0.05, 0.01) with associated confidence level (90%, 95%, 99%)
+  - Specify *in advance* an $\alpha$-level (0.10, 0.05, 0.01) with associated confidence level (90%, 95%, 99%)
 
@@
   ~"", ~"", ~"Null is True", ~"Null is False",
-  "Judgment", "Reject Null", "Type I Error", "Correct",
-  "Judgment", "Reject Null", "alpha", "1-beta",
-  "Judgment", "Don't Reject Null", "Correct", "Type II Error",
-  "Judgment", "Don't Reject Null", "1-alpha", "beta"
+  "Judgment", "Reject Null", "TYPE I ERROR", "CORRECT",
+  "Judgment", "Reject Null", "(alpha)", "(1-beta)",
+  "Judgment", "Don't Reject Null", "CORRECT", "TYPE II ERROR",
+  "Judgment", "Don't Reject Null", "(1-alpha)", "(beta)"
   ) %>%
@@
 
-- The statistical .shout[power of the test] is $1-\beta$, the probability of correctly rejecting $H_0$ when $H_0$ is in fact false (e.g. not convicting an innocent person)
+- The statistical .shout[power of the test] is $(1-\beta)$: the probability of correctly rejecting $H_0$ when $H_0$ is in fact false (e.g. not convicting an innocent person)
 
@@
 
-$$p(|H_0)$$
-
-- If $p<\alpha$, the results are "**statistically significant**"
+$$p(\delta_i \geq \delta^*|H_0 \text{ is true})$$
+    - where $\delta$ represents some test statistic
+    - $\delta_i$ is the test statistic we observe in our sample
+    - $\delta^*$ is some **critical value** of the test statistic
+    - More on this in a bit
 
@@
 - After running our test, we need to make a *decision* between the competing hypotheses
+
 - Compare $p$-value with *pre-determined* $\alpha$ (commonly, $\alpha=0.05$, 95% confidence level)
@@
 
-- Modern philosophy of science is largely based off of hypothesis testing and .whisper[falsifiability], which form the "Scientific Method"<sup>1</sup>
-- For something to be "scientific", it must be .whisper[falsifiable], or at least .whisper[testable]
+- Modern philosophy of science is largely based off of hypothesis testing and .onfire[falsifiability], which form the "Scientific Method"<sup>1</sup>
+
+- For something to be "scientific", it must be .whisper[falsifiable], or at least .onfire[testable]
+
 - Hypotheses can be *corroborated* with evidence, but always *tentative* until falsified by data in suggesting an alternative hypothesis
-- .onfire["All swans are white"] is a hypothesis rejected upon discovery of a single black swan
+
+> **"All swans are white"** is a hypothesis rejected upon discovery of a single black swan 
 ]
@@
 
-# Hypothesis Testing I
+# Hypothesis Testing and p-Values
 
@@
 
-# Hypothesis Testing II
+# Hypothesis Testing: Which Test? I
 
-- Rigorous course on statistics (**ECMG 212** or **MATH 112**) will spend weeks going through different types of tests:
+- Rigorous course on statistics ([**ECMG 212**](http://ryansafner.com/courses/ecmg212) or **MATH 112**) will spend weeks going through different types of tests:
     - Sample mean; difference of means
@@
 
-- Fortunately, some statisticians realized ["**there is only one test**"](https://allendowney.blogspot.com/2011/05/there-is-only-one-test.html) and coded it that way
+- See today's [class notes](/class/12-class) page for more
 
@@
 
-# Elements of a Hypothesis Test
+# Hypothesis Testing: Which Test? II
+
+.center[
+![:scale 80%](https://www.dropbox.com/s/3f33cnn595x2nsj/hypothesistestflowchart.png?raw=1)
+]
+
+---
+
+# There is Only One Test
+
+- Fortunately, some statisticians realized ["**there is only one test**"](https://allendowney.blogspot.com/2011/05/there-is-only-one-test.html) and some clever folks built a nice `R` package called `infer`
 
 1. Calculate a statistic, $\delta$<sup>1</sup>, from a sample of data
+
 2. Simulate a world where $\delta$ is null
+
 3. Examine $\delta$ in the null world
+
 4. Calculate the probability that $\delta$ could exist in the null world
+
 5. Decide if $\delta$ is statistically significant
 
-.footnote[<sup>1</sup> $\delta$ can stand in for any test-statistic in any hypothesis test! For our purposes, $\delta$ is the slope of our regression sample, $\hat{\beta}_1$.]
+.footnote[<sup>1</sup> `\\(\delta\\)` can stand in for any test-statistic in any hypothesis test! For our purposes, `\\(\delta\\)` is the slope of our regression sample, `\\(\hat{\beta}_1\\)`.]
 
@@
 ]
-
----
-
-# Test Statistic
-
-- .whisper[Test statistic]: measure of how far what we observed in our sample is from what we would expect if the null hypothesis was true
-    - Calculate from a distribution
-- .whisper[Rejection region]: if the test statistic reaches a .onfire["critical value"], then we reject the null hypothesis
-
----
-
-# Statistical Significance and Economic Significance
-
-## Is a finding *significant* ?
-
-## Is a finding *meaningful* ?
-
----
-
-# How Big is "Big"?
-
-.center[
-![:scale 40%](https://www.dropbox.com/s/qtd2uwvr52jvkcu/drwhocontext.jpg?raw=1)
-]
-
----
-
-# Imagine a Null World
-
-.center[
-![:scale 70%](https://www.dropbox.com/s/x7b5azt5algjodg/parallelworld.jpeg?raw=1)
-]
-
----
-
-# Imagine a Null World
-
-.pull-left[
-
-# Our Sample
-
-```{r, echo=F}
-tidy(school_reg)
-```
-]
-
-.pull-right[
-
-# Another Sample
-
-```{r, echo = F}
-library(infer)
-CASchool %>%
-  specify(testscr ~ str) %>%
-  hypothesize(null = "independence") %>%
-  generate(reps = 1, type = "permute") %>%
-  lm(testscr ~ str, data = .) %>%
-  tidy()
-```
-]
-
----
-
-# 1000 Parallel Universes
-
-.left-code[
-
-```{r infer-example, echo = T, eval = F}
-library(infer)
-simulations <- CASchool %>%
-  specify(testscr ~ str) %>%
-  hypothesize(null = "independence") %>%
-  generate(reps = 1000, type = "permute") %>%
-  calculate(stat = "slope")
-
-simulations %>%
-  rename(sample = replicate,
-         slope = stat)
-```
-]
-
-.right-plot[
-```{r ref.label="infer-example", echo = F}
-```
-
-]
-
----
-
-# 1000 Parallel Universes
-
-```{r echo = F}
-plot <- simulations %>%
-  visualize()+
-  theme_minimal()+
-  labs(x = "Slope",
-       y = "Samples")
-
-plot
-```
-
----
-
-# 1000 Parallel Universes
-
-```{r}
-obs_slope <- school_reg_tidy %>% 
-  filter(term=="str") %>%
-  pull(estimate) # from dplyr
-
-plot2 <- simulations %>%
-  visualize(obs_stat = obs_slope)
-
-plot2+
-  shade_p_value(obs_stat = obs_slope, 
-                direction = "two_sided")+
-  theme_minimal()+
-  labs(x = "Slope",
-       y = "Samples")
-```
-
----
-
-# 1000 Parallel Universes
-
-```{r}
-# get confidence interval values
-
-school_reg_CI<- school_reg %>%
-  tidy(., conf.int=T) %>%
-  filter(term=="str") %>%
-  select(conf.low, conf.high)
-```
-
-```{r}
-plot2+
-  shade_confidence_interval(endpoints = school_reg_CI)+
-  theme_minimal()+
-  labs(x = "Slope",
-       y = "Samples")
-```
-
----
-
-# p-Value
-
-```{r}
-simulations %>%
-  get_p_value(obs_stat = obs_slope, direction = "both")
-```
-
-> There's a 0% chance of seeing a slope at least as large as $-2.28$ in the null world where there is no relationship.
-
----
-
-# What R Does
-
-```{r}
-tidy(school_reg)
-```
-
-- R calculates the $p$-value the "old-fashioned way", using a theoretical distribution instead of simulation
-    - Generates a test statistic (`statistic`) and compares it to a critical value of the test statistic<sup>1</sup>
-$$\text{test statistic} = \frac{\text{estimate} - \text{null hypothesis}}{\text{standard error of estimate}}$$
-
-.footnote[<sup>1</sup> Rule of thumb: test statistic of 2 or larger is generally significant at the 95% level.
-This comes directly from the empirical rule that 95% of observations occur within 2 standard deviations of the mean!
-See the **class info** page for more.]
-
----
-
-# What R Does II
-
-```{r}
-tidy(school_reg)
-```
-
-$$t = \frac{-2.28-0}{0.48} = -4.75$$
-- Intepretation like $Z$-scores: number of standard deviations away from the center<sup>1</sup>
-- $p$-value is the probability of getting a test statistic at least as large (in magnitude) as this one if the null hypothesis were true<sup>2</sup>
-
-.footnote[<sup>1</sup> The "center" here is the null hypothesis value i.e. a slope of 0.
-<sup>2</sup> Think of our simulated null world(s), the center of the sampling distribution was a slope of 0.]
-
----
-
-# Confidence Intervals
-
-```{r}
-tidy(school_reg, conf.int=TRUE)
-```
-
-- We can also generate a .shout[confidence interval] for $\hat{\beta}_1$ that contain the true population slope ($\beta_1$) with a specified probability
-    - Note confidence level $=1-\alpha$; commonly 95% ($\alpha=0.05$)
-- .whisper[We are 95% confident that the true slope is between $-3.22$ and $-1.34$.]
-    - Note if the confidence interval contains 0, *cannot* reject the null hypothesis of $\beta_1=0$! 
-- See **class info** page for more on confidence intervals and how they are constructed
-
----
-
-```{r, results="asis"}
-equatiomatic::extract_eq(school_reg, use_coefs = TRUE,coef_digits = 2,fix_signs = TRUE)
-```
-
----
-
-# Testing Reg Output: huxtable
-
-.font80[
-.regtable[
-```{r, echo=F}
-library(huxtable)
-
-hr <- huxreg("Model 1" = school_reg, # can name the columns in quotes
-       coefs = c("Intercept" = "(Intercept)", # can rename the coefficients
-                 "STR" = "str"),
-       statistics = c("N" = "nobs", # statistics to show for regression, taken from glance()
-                      "R Squared" = "r.squared",
-                      "Adj. R Squared" = "adj.r.squared",
-                      "SER" = "sigma",
-                      "F" = "statistic"),
-       number_format = 2, # number of decimals to round to
-       bold_signif = 0.05) # can bold coeffs by significance level
-
-hr
-```
-]
-]
-
----

### 2019-10-03 static/slides/12-slides.Rmd
Tweak slides 12 more

@@
 
-- Fortunately, some statisticians realized ["**there is only one test**"](https://allendowney.blogspot.com/2011/05/there-is-only-one-test.html) and some clever folks built a nice `R` package called `infer`
+- Fortunately, some clever statisticians realized ["**there is only one test**"](https://allendowney.blogspot.com/2011/05/there-is-only-one-test.html) and built a nice `R` package called `infer`
 
-1. Calculate a statistic, $\delta$<sup>1</sup>, from a sample of data
+1. **Calculate** a statistic, $\delta_i$<sup>1</sup>, from a sample of data
 
-2. Simulate a world where $\delta$ is null
+2. **Simulate** a world where $\delta$ is null $(H_0)$
 
-3. Examine $\delta$ in the null world
+3. **Examine** the distribution of $\delta$ across the null world
 
-4. Calculate the probability that $\delta$ could exist in the null world
+4. **Calculate** the probability that $\delta_i$ could exist in the null world
 
-5. Decide if $\delta$ is statistically significant
+5. **Decide** if $\delta_i$ is statistically significant
 
@@
 
-- `infer` allows you to run through these steps manually to understand the process
+- `infer` allows you to run through these steps manually to understand the process:
 
----
-
-# Hypothesis Testing with the infer Package II
+--
 
@@
 
-4. `calculate()`
+4. `calculate()` the $p$-value
 
@@
 
-5. `visualize()`
+5. `visualize()` with a histogram (optional)
 
@@
 
-# Hypothesis Testing with the infer Package III
+# Hypothesis Testing with the infer Package II
 

### 2019-10-03 static/slides/12-slides.Rmd
Fix p-value definition!

@@
 
-$$p(\delta_i \geq \delta^*|H_0 \text{ is true})$$
+$$p(\delta \geq \delta_i|H_0 \text{ is true})$$
   - where $\delta$ represents some test statistic
   - $\delta_i$ is the test statistic we observe in our sample
-  - $\delta^*$ is some **critical value** of the test statistic
   - More on this in a bit

### 2019-10-04 content/class/11-class.Rmd
Fix slides 11, add practice 11 answers, edit class 12 page

@@
 
-*Tuesday, October 2, 2019*
+*Tuesday, October 1, 2019*
 
@@
 [Problem Set 2](/assignment/01-problem-set.html) answers are posted.
+
+# Appendix: Robust Standard Errors in R
+
+This, since I started using `huxtable` instead of another package (`stargazer`) to make regression tables, I have gone all in on `estimatr`'s `lm_robust()` option to calculate robust standard errors.
+Before this, there were some other methods that I had to resort to.
+You can read about that in [this blog post](http://ryansafner.com/post/replicating-statas-robust-option-for-ols-standard-errors-in-r).

### 2019-10-04 content/class/12-class.Rmd
Add problem set 3 mention in class 12
Fix slides 11, add practice 11 answers, edit class 12 page

@@
 
-*Thursday, October 3, 2019*
+*Thursday, October 3, 2019* and *Tuesday, October 8, 2019*
 
@@
 
-We begin with some more time for you to work on the [R Practice]() from last class. I will post the answer key after class.
+We begin with some more time for you to work on the [R Practice](/r/11-r-practice) from last class. The answer key is now posted.
 
-This class and the next class are about inferential statistics: using *statistics* calculated from a *sample* of data to **infer** the true (and unmeasurable) *parameters* that describe a *population.* In doing so, we can run *hypothesis tests* on our sample to determine a *point* estimate of a parameter, or construct a *confidence interval* from our sample to cast a range for the true parameter.
+This class will be spread over 2 days, and is about inferential statistics: using *statistics* calculated from a *sample* of data to **infer** the true (and unmeasurable) *parameters* that describe a *population.* In doing so, we can run *hypothesis tests* on our sample to determine a *point* estimate of a parameter, or construct a *confidence interval* from our sample to cast a range for the true parameter.
 
@@
 
-Today you will be finishing up [R practice problems](/r/11-r-practice.html) from last class. Check back later for solutions.
+You will be finishing up [R practice problems](/r/11-r-practice.html) from last class. Answers are posted.
 
@@
 
-# Assignments: Problem Set 2 Answers
+# Assignments: Problem Set 3 Due Thursday October 10
 
-[Problem Set 2](/assignment/01-problem-set.html) answers are posted.
+Problem Set 3 will be posted later and is due Thursday September 19.
 

### 2019-10-04 content/r/11-r-practice.Rmd
Add link to data in practice 11
Fix slides 11, add practice 11 answers, edit class 12 page

@@
 ---
-title: "2.4 OLS: Precision and Diagnostics - R Practice (Answers)"
+title: "2.4 OLS: Precision and Diagnostics - R Practice"
 editor_options: 
@@
 
-Our [data](https://github.com/fivethirtyeight/data/tree/master/congress-trump-score) comes from fivethirtyeight's [Trump Congress tracker](https://projects.fivethirtyeight.com/congress-trump-score/).
+# Set Up
+To minimize confusion, I suggest creating a new `R Project` (e.g. `regression_practice`) and storing any data in that folder on your computer.
 
-- [View Project on R Studio Cloud](https://rstudio.cloud/project/575252)
+Alternatively, I have made a project in R Studio Cloud that you can use (and not worry about trading room computer limitations), with the data already inside (you will still need to assign it to an object).
+
+- [<i class="fas fa-cloud"></i> View Project on R Studio Cloud](https://rstudio.cloud/project/575252)
+- [<i class="fab fa-markdown"></i> `Download Markdown`](/r/11-r-practice-rmd.Rmd)
+
+- [<i class="fas fa-external-link-square-alt"></i> `Answer Key (html)`](/r/11-r-practice-answers.html)
+- [<i class="fab fa-markdown"></i> `Answer Key (markdown)`](/r/11-r-practice-answers.Rmd)
 
 ## Question 1
+Our [data](https://github.com/fivethirtyeight/data/tree/master/congress-trump-score) comes from fivethirtyeight's [Trump Congress tracker](https://projects.fivethirtyeight.com/congress-trump-score/).
 Download and read in (`read_csv`) the data.
+
+- [<i class="fas fa-table"></i> `congress-trump-score.csv`](/data/congress-trump-score.csv)
 
@@
 ### Part C
-Look in your original data to match this outlier with an observation. Hint: use the `slice()` command, as the outlier test gave you an observation number!
+Look in your original data to match this outlier with an observation. Hint: use the `slice()` command, as the outlier test gave you an observation (row) number!
 
@@
 
-This data is still a bit messy. We'd also probably like to plot our scatterplots with colors for Republican and Democratic party. Or sort by the House and the Senate.
+This data is still a bit messy.
+Let's check in on your `tidyverse` skills again!
+For example, we'd probably like to plot our scatterplots with colors for Republican and Democratic party.
+Or plot by the House and the Senate.
+
+### Part A
+First, the variable `congress` (session of Congress) seems a bit off. Get a `count()` of `congress`.
+
+### Part B
+Let's get rid of the `0` values for `congress` (someone made a mistake coding this, probably).
+Also, while we're at it, let's take `agree_pct` and `mutate` a variable that is a proper percentage (i.e.
+`*100`).
+
+### Part C
+The variable `party` is also quite a mess.
+`count()` by `party` to see.
+Then let's `mutate` a variable to make a better measure of political party - just `"Republican"`, `"Democrat"`, and `"Independent"`.
+Try doing this with the `case_when()` command (as your `mutate` formula).^[The syntax for `case_when()` is to have a series of `condition ~ "Outcome"`, separated by commas.
+For example, one condition is to assign both `"Democrat"` and `"D"` to `"Democrat"`, as in `party %in% c("Democrat", "D") ~ "Democrat"`.
+You could also do this with a few `ifelse()` commands, but that's a bit more awkward.] When you're done `count()` by your new party variable to make sure it worked.
+
+### Part D
+Now plot a scatterplot (same as Question 3) and set `color` to your party variable.
+Notice `R` uses its own default colors, which don't match to the actual colors these political parties use!
+Make a vector where you define the party colors as follows: `party_colors <- c("Democrat" = "blue", "Republican" = "red", "Independent" = "gray")`.
+Then, run your plot again, adding the following line to customize the colors `+scale_colour_manual("Parties", values = party_colors)`.^[`"Parties"` is the title that will show up on the legend, feel free to edit it, or remove the legend with another layer `+guides(color = F)`.]
+
+### Part E
+Now facet your scatterplot by `chamber`.

### 2019-10-04 static/slides/11-slides.Rmd
Fixed slides 11
Fix slides 11, add practice 11 answers, edit class 12 page

@@
                  "STR" = "str"),
-       statistics = c("N" = "N",
+       statistics = c("N" = "nobs",
                       "R-Squared" = "r.squared",
@@
 ```{r, echo=T, eval=F}
-statistics = c("N" = "N",
+statistics = c("N" = "nobs",
                "R-Squared" = "r.squared",
@@
 
-4. There is no correlation between $X$ and the error term: 
+4. No correlation between $X$ and the error term: 
 $$cor(X, u)=0 \text{ or } E[u|X]=0$$
@@
 aug_reg %>%
-  summarize(e_u = mean(.resid),
+  summarize(E_u = mean(.resid),
             sd_u = sd(.resid))
@@
 ]
-
---
 
@@
   theme_bw(base_family = "Fira Sans Condensed",
-           base_size=20)
+           base_size=16)
 
@@
                  "Years of Schooling" = "educ"),
-       statistics = c("N" = "N",
+       statistics = c("N" = "nobs",
                       "R-Squared" = "r.squared",
@@
 
-```
-
-```{r, fig.height=3.5, fig.retina=3}
 aug_wage_reg<-aug_wage_reg %>%
@@
 ```
-
+]
 ---
@@
 
-```{r}
+```{r, echo = T}
 #install.packages("estimatr")

### 2019-10-04 content/assignment/03-problem-set.Rmd
Remove excess data (from Heiss' class template), and add problem set 3 page

---
title: "Problem Set 3"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

*Due by Thursday, October 10, 2019*

- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/02-problem-set-pdf.pdf)
- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/03-problem-set.Rmd)

# Instructions

For this problem set, you may submit handwritten answers on a plain sheet of paper, or download and type/handwrite on the PDF.

Alternatively, you may download the `.Rmd` file, do the homework in markdown, and email to me a single `knit`ted `html` or `pdf` file (and be sure that it shows all of your code).

You may work together (and I highly encourage that) but you must turn in your own answers. I grade homeworks 70% for completion, and for the remaining 30%, pick one question to grade for accuracy - so it is best that you try every problem, even if you are unsure how to complete it accurately.

# Theory and Concepts

## Question 1
In your own words, describe what exogeneity and endogeneity mean, and how they are related to bias. What can we learn about the bias if we know $X$ is endogenous?

## Question 2
In your own words, describe what $R^2$ means. How do we calculate it, what does it tell us, and how do we interpret it?

## Question 3
In your own words, describe what the standard error of the regression ($SER$) means. How do we calculate it, what does it tell us, and how do we interpret it?

## Question 4
In your own words, describe what homoskedasticity and heteroskedasticity mean: both in ordinary English, and in terms of the graph of the OLS regression line.

## Question 5
In your own words, describe what the variation in $\hat{\beta_1}$ (either variance or standard error) means, or is measuring. What three things determine the variation, and in what way?

## Question 6
A researcher is interested in examining the impact of illegal music downloads on commercial music sales. The author collects data on commercial sales of the top 500 singles from 2017 ($Y$) and the number of downloads from a web site that allows `file sharing' ($X$). The author estimates the following model

$$\text{music sales}_i = \beta_0+\beta_1 \text{illegal downloads}_i + u_i$$

The author finds a large, positive, and statistically significant estimate of $\hat{\beta_1}$. The author concludes these results demonstrate that illegal downloads actually *boost* music sales. Is this an unbiased estimate of the impact of illegal music on sales? Why or why not? Do you expect the estimate to overstate or understate the true relationship between illegal downloads and sales?

## Question 6
A pharmaceutical company is interested in estimating the impact of a new drug on cholesterol levels. They enroll 200 people in a clinical trial. People are randomly assigned the treatment group or into the control group. Half of the people are given the new drug and half the people are given a sugar pill with no active ingredient. To examine the impact of dosage on reductions in cholesterol levels, the authors of the study regress the following model:

$$\text{cholesterol level}_i = \beta_0+\beta_1 \text{dosage level}_i + u_i$$

For people in the control group, dosage level$_i=0$ and for people in the treatment group, dosage level$_i$ measures milligrams of the active ingredient. In this case, the authors find a large, negative, statistically significant estimate of $\hat{\beta_1}$. Is this an unbiased estimate of the impact of dosage on change in cholesterol level? Why or why not? Do you expect the estimate to overstate or understate the true relationship between dosage and cholesterol level?

# Theory Problems

For the following questions, please *show all work* and explain answers as necessary. You may lose points if you only write the correct answer. You may use `R` to *verify* your answers, but you are expected to reach the answers in this section "manually."

## Question 7

A researcher wants to estimate the relationship between average weekly earnings $(AWE$, measured in dollars) and $Age$ (measured in years) using a simple OLS model. Using a random sample of college-educated full-time workers aged 25-65 yields the following:

$$\widehat{AWE} = 696.70+9.60\ \, Age$$

### Part A
Interpret what $\hat{\beta_0}$ means in this context.

### Part B
Interpret what $\hat{\beta_1}$ means in this context.

### Part C
The $R^2=0.023$ for this regression. What are the units of the $R^2$, and what does this mean?

### Part D
The $SER, \, \hat{\sigma_u}=624.1$ for this regression. What are the units of the SER in this context, and what does it mean? Is the SER large in the context of this regression?

### Part E
Suppose Maria is 20 years old. What is her predicted $\widehat{AWE}$?

### Part F
Suppose the data shows her *actual* $AWE$ is $430. What is her residual? Is this a relatively good or a bad prediction?^[Hint: compare your answer here to your answer in Part D.]

### Part G
What does the error term, $\hat{u_i}$ represent in this case? What might individuals have different values of $u_i$?

### Part H
Do you think that $Age$ is exogenous? Why or why not? Would we expect $\hat{\beta_1}$ to be too *large* or too *small*?

## Question 8

Suppose a researcher is interested in estimating a simple linear regression model:

$$Y_i=\beta_0+\beta_1X_i+u_i$$
In a sample of 48 observations, she generates the following descriptive statistics:

- $\bar{X}=30$
- $\bar{Y}=63$
- $\displaystyle\sum^n_{i=1}(X_i-\bar{X})^2= 6900$
- $\displaystyle\sum^n_{i=1}(Y_i-\bar{Y})^2= 29000$
- $\displaystyle\sum^n_{i=1}(X_i-\bar{X})(Y_i-\bar{Y})=13800$
- $\displaystyle\sum^n_{i=1}\hat{u}^2=1656$

### Part A
What is the OLS estimate of $\hat{\beta_1}$?

### Part B
What is the OLS estimate of $\hat{\beta_0}$?

### Part C
Suppose the OLS estimate of $\hat{\beta_1}$ has a standard error of $0.072$. Could we probably reject a null hypothesis of $H_0: \beta_1=0$ at the 95% level?

### Part D
Calculate the $R^2$ for this model. How much variation in $Y$ is explained by our model?

### Part E
How large is the average residual?

# R Questions

For the following problems, please attach/write the answers to each question on the same document as the previous problems, but also include a printed/attached (and commented!) *.R* script file of your commands to answer the questions.

## Question 9

- [<i class="fas fa-table"></i> `mlbattend.csv`](http://metricsf19.classes.ryansafner.com/data/mlbattend.csv)

Download the `MLBattend` dataset. This data contains data on attendance at major league baseball games for all 32 MLB teams from the 1970s-2000. We want to answer the following question:

> "How big is home-field advantage in baseball? Does a team with higher attendance at home games over their season have score more runs over their season?"

### Part A

Clean up the data a bit by making a new variable to measure home attendance in millions. This will make it easier to interpret your regression later on.

### Part B

Plot a scatterplot of Runs Scored (`y`) on Home Attendance (`x`). Add a regression line.

### Part C

Run a regression of Runs Scored on Home Attendance. What are $\beta_0$ and $\hat{\beta_1}$? Interpret them in the context of our question.

### Part D

Write out the estimated regression equation.

### Part E

Make a regression table of the output.

### Part F

Make a histogram of the residuals. Do they look roughly normal?

### Part G

Make a residual plot.

### Part H

Test the regression for heteroskedasticity. Are the errors homoskedastic or heteroskedastic? Generate robust standard errors. Make a regression output table, with one column with regular standard errors and another with robust standard errors.

### Part I

Are any results statistically significant or not?

### Part J

Test the data for outliers. If there are any, identify which team(s) and season(s) are outliers.

### Part K

Use the `infer` package to simulate 1000 additional samples^[`generate(reps = 1000, type = "permute")`] to calculate the slope^[`calculate(stat = "slope")`] under the null hypothesis that there is no connection between attendance and runs.^[`hypothesize(null = "independence")`], and save this as an object (it's a `tibble`).

### Part L
Save the (value of) our sample $\hat{\beta_1}$ from your regression in Part C as an object. Make a histogram of the simulated slopes, and plot our sample slope on that histogram, shading the $p$-value.^[You can use `ggplot2` to plot a histogram in the normal way and add a `geom_vline()`, setting `xintercept` equal to your saved object with the sample $\hat{\beta_1}$ value. *Alternatively*, you can use `infer` to pipe your `tibble` of simulations into `visualize()`, and inside `visualize()` set `obs_stat` equal to your saved $\hat{\beta_1}$ object. Regardless of which method you use, then add `+shade_p_value()`. Inside this, set `obs_stat` equal to your saved slope, and add `direction = "two_sided"`.]

### Part M
Get the confidence interval for your slope estimate,^[`tidy()` your original regression, with `, conf.int = TRUE` inside the command, then `select(conf.low, conf.high)` and `filter` by your `x` variable. Save this as an object.] and then make a histogram of the simulated slopes (like part L), but instead, add `+shade_confidence_interval()`.^[Inside of this, set `endpoints` equal to the object you just made with the low and high confidence interval values.]

### 2019-10-05 content/assignment/03-problem-set.Rmd
Add problem set 3 materials

@@
 
-- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/02-problem-set-pdf.pdf)
+- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/03-problem-set-pdf.pdf)
 - [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/03-problem-set.Rmd)
@@
 ## Question 6
+In your own words, describe what a $p$-value means, and how it is used to establish statistical significance.
+
+## Question 7
 A researcher is interested in examining the impact of illegal music downloads on commercial music sales.
@@
 
-## Question 6
+## Question 8
 A pharmaceutical company is interested in estimating the impact of a new drug on cholesterol levels.
@@
 
-## Question 7
+## Question 9
 
@@
 
-$$\widehat{AWE} = 696.70+9.60\ \, Age$$
+$$\widehat{AWE} = 696.70+9.60 \, Age$$
 
@@
 
-## Question 8
+## Question 10
 
@@
 
-For the following problems, please attach/write the answers to each question on the same document as the previous problems, but also include a printed/attached (and commented!) *.R* script file of your commands to answer the questions.
+Answer the following questions using `R`.
+When necessary, please write answers in the same document (knitted `Rmd` to `html` or `pdf`, typed `.doc(x)`, or handwritten) as your answers to the above questions.
+Be sure to include (email or print an `.R` file, or show in your knitted `markdown`) your code and the outputs of your code with the rest of your answers.
 
-## Question 9
+## Question 11
 
@@
 
-Make a histogram of the residuals. Do they look roughly normal?
+Now let's start running some diagnostics of the regression. Make a histogram of the residuals. Do they look roughly normal?
 
@@
 
-Are any results statistically significant or not?
+Test the data for outliers. If there are any, identify which team(s) and season(s) are outliers.
 
@@
 
-Test the data for outliers. If there are any, identify which team(s) and season(s) are outliers.
+What is the marginal effect of home attendance on runs scored? Is this statistically significant? Why or why not?
 
@@
 
-Use the `infer` package to simulate 1000 additional samples^[`generate(reps = 1000, type = "permute")`] to calculate the slope^[`calculate(stat = "slope")`] under the null hypothesis that there is no connection between attendance and runs.^[`hypothesize(null = "independence")`], and save this as an object (it's a `tibble`).
+Now we'll try out the `infer` package to understand the $p$-value and confidence interval for our observed slope in our regression model.
+Save the (value of) our sample $\hat{\beta_1}$ from your regression in Part C as an object.
+Then, install and load the `infer` package, and then calculate the slope^[`calculate(stat = "slope")`] under the null hypothesis that there is no connection between attendance and runs.^[`hypothesize(null = "independence")`] for 1000 additional simulated samples^[`generate(reps = 1000, type = "permute")`], and save this as an object (it's a `tibble`).
+Then, use this to `get_p_value()`^[Set `obs_stat` equal to your observed slope, and set `direction = "two_sided"`].
+Compare to the $p$-value given by `lm()` and `tidy()` above.
 
 ### Part L
-Save the (value of) our sample $\hat{\beta_1}$ from your regression in Part C as an object.
 Make a histogram of the simulated slopes, and plot our sample slope on that histogram, shading the $p$-value.^[You can use `ggplot2` to plot a histogram in the normal way and add a `geom_vline()`, setting `xintercept` equal to your saved object with the sample $\hat{\beta_1}$ value.
 *Alternatively*, you can use `infer` to pipe your `tibble` of simulations into `visualize()`, and inside `visualize()` set `obs_stat` equal to your saved $\hat{\beta_1}$ object.
-Regardless of which method you use, then add `+shade_p_value()`.
+*Regardless* of which method you use, add `+shade_p_value()`.
 Inside this, set `obs_stat` equal to your saved slope, and add `direction = "two_sided"`.]
@@
 ### Part M
-Get the confidence interval for your slope estimate,^[`tidy()` your original regression, with `, conf.int = TRUE` inside the command, then `select(conf.low, conf.high)` and `filter` by your `x` variable.
-Save this as an object.] and then make a histogram of the simulated slopes (like part L), but instead, add `+shade_confidence_interval()`.^[Inside of this, set `endpoints` equal to the object you just made with the low and high confidence interval values.]
+Get the 95% confidence interval for your slope estimate,^[`tidy()` your original regression, with `conf.int = TRUE` inside the command, then `select(conf.low, conf.high)` and `filter` by your `x` variable.
+Save this as an object.] and then make a histogram of the simulated slopes (like part L), but instead, add `+shade_confidence_interval()`.^[Inside of this, set `endpoints` equal to the object you just made with the low and high confidence interval values.] Compare this to what you get with `tidy()` above.

### 2019-10-05 static/homeworks/03-problem-set-pdf.Rmd
Add problem set 3 materials

--- 2019-10-05 content/assignment/03-problem-set.Rmd
@@
 title: "Problem Set 3"
-editor_options: 
-  chunk_output_type: console
+author: "ECON 480 - Fall 2019"
+date: "Due by Thursday, October 10, 2019"
+output: pdf_document
 ---
 
-```{r message=FALSE, warning=FALSE, include=FALSE}
-library(tidyverse)
-library(pander)
-source(file.path(here::here(), "lib", "graphics.R"))
-```
-
-*Due by Thursday, October 10, 2019*
-
-- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/03-problem-set-pdf.pdf)
-- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/03-problem-set.Rmd)
+<!--CLICK "KNIT" ABOVE TO RENDER TO HTML, PDF, OR WORD OUTPUT
+In fact, try knitting right away and see what this produces!
+-->
 
@@
 
-## Question 1
-In your own words, describe what exogeneity and endogeneity mean, and how they are related to bias. What can we learn about the bias if we know $X$ is endogenous?
+1. In your own words, describe what exogeneity and endogeneity mean, and how they are related to bias. What can we learn about the bias if we know $X$ is endogenous?
 
-## Question 2
-In your own words, describe what $R^2$ means. How do we calculate it, what does it tell us, and how do we interpret it?
+\vspace{3in}
 
-## Question 3
-In your own words, describe what the standard error of the regression ($SER$) means. How do we calculate it, what does it tell us, and how do we interpret it?
+2. In your own words, describe what $R^2$ means. How do we calculate it, what does it tell us, and how do we interpret it?
 
-## Question 4
-In your own words, describe what homoskedasticity and heteroskedasticity mean: both in ordinary English, and in terms of the graph of the OLS regression line.
+\clearpage
 
-## Question 5
-In your own words, describe what the variation in $\hat{\beta_1}$ (either variance or standard error) means, or is measuring. What three things determine the variation, and in what way?
+3. In your own words, describe what the standard error of the regression ($SER$) means. How do we calculate it, what does it tell us, and how do we interpret it?
 
-## Question 6
-In your own words, describe what a $p$-value means, and how it is used to establish statistical significance.
+\vspace{3in}
 
-## Question 7
+4. In your own words, describe what homoskedasticity and heteroskedasticity mean: both in ordinary English, and in terms of the graph of the OLS regression line.
+
+\clearpage
+
+5. In your own words, describe what the variation in $\hat{\beta_1}$ (either variance or standard error) means, or is measuring. What three things determine the variation, and in what way?
+
+\vspace{3in}
+
+6. In your own words, describe what a $p$-value means, and how it is used to establish statistical significance.
+
+\clearpage
+
+7.
 A researcher is interested in examining the impact of illegal music downloads on commercial music sales.
@@
 
-The author finds a large, positive, and statistically significant estimate of $\hat{\beta_1}$.
+> The author finds a large, positive, and statistically significant estimate of $\hat{\beta_1}$.
 The author concludes these results demonstrate that illegal downloads actually *boost* music sales.
@@
 
-## Question 8
+\vspace{3in}
+
+8.
 A pharmaceutical company is interested in estimating the impact of a new drug on cholesterol levels.
@@
 
-For people in the control group, dosage level$_i=0$ and for people in the treatment group, dosage level$_i$ measures milligrams of the active ingredient.
+> For people in the control group, dosage level$_i=0$ and for people in the treatment group, dosage level$_i$ measures milligrams of the active ingredient.
 In this case, the authors find a large, negative, statistically significant estimate of $\hat{\beta_1}$.
@@
 Do you expect the estimate to overstate or understate the true relationship between dosage and cholesterol level?
+
+\clearpage
 
@@
 
-## Question 9
-
+9.
 A researcher wants to estimate the relationship between average weekly earnings $(AWE$, measured in dollars) and $Age$ (measured in years) using a simple OLS model.
@@
 
-### Part A
-Interpret what $\hat{\beta_0}$ means in this context.
+a. Interpret what $\hat{\beta_0}$ means in this context.
 
-### Part B
-Interpret what $\hat{\beta_1}$ means in this context.
+\vspace{1.5in}
 
-### Part C
-The $R^2=0.023$ for this regression. What are the units of the $R^2$, and what does this mean?
+b. Interpret what $\hat{\beta_1}$ means in this context.
 
-### Part D
-The $SER, \, \hat{\sigma_u}=624.1$ for this regression. What are the units of the SER in this context, and what does it mean? Is the SER large in the context of this regression?
+\vspace{1.5in}
 
-### Part E
-Suppose Maria is 20 years old. What is her predicted $\widehat{AWE}$?
+c. The $R^2=0.023$ for this regression. What are the units of the $R^2$, and what does this mean?
 
-### Part F
-Suppose the data shows her *actual* $AWE$ is $430. What is her residual? Is this a relatively good or a bad prediction?^[Hint: compare your answer here to your answer in Part D.]
+\vspace{1.5in}
 
-### Part G
-What does the error term, $\hat{u_i}$ represent in this case? What might individuals have different values of $u_i$?
+d. The $SER, \, \hat{\sigma_u}=624.1$ for this regression. What are the units of the SER in this context, and what does it mean? Is the SER large in the context of this regression?
 
-### Part H
-Do you think that $Age$ is exogenous? Why or why not? Would we expect $\hat{\beta_1}$ to be too *large* or too *small*?
+\clearpage
 
-## Question 10
+e. Suppose Maria is 20 years old. What is her predicted $\widehat{AWE}$?
 
-Suppose a researcher is interested in estimating a simple linear regression model:
+\vspace{1.5in}
+
+f. Suppose the data shows her *actual* $AWE$ is $430. What is her residual? Is this a relatively good or a bad prediction?^[Hint: compare your answer here to your answer in Part D.]
+
+\vspace{1.5in}
+
+g. What does the error term, $\hat{u_i}$ represent in this case? What might individuals have different values of $u_i$?
+
+\vspace{1.5in}
+
+h. Do you think that $Age$ is exogenous? Why or why not? Would we expect $\hat{\beta_1}$ to be too *large* or too *small*?
+
+\clearpage
+
+10. Suppose a researcher is interested in estimating a simple linear regression model:
 
@@
 
-### Part A
-What is the OLS estimate of $\hat{\beta_1}$?
+a. What is the OLS estimate of $\hat{\beta_1}$?
 
-### Part B
-What is the OLS estimate of $\hat{\beta_0}$?
+\vspace{2in}
 
-### Part C
-Suppose the OLS estimate of $\hat{\beta_1}$ has a standard error of $0.072$. Could we probably reject a null hypothesis of $H_0: \beta_1=0$ at the 95% level?
+b. What is the OLS estimate of $\hat{\beta_0}$?
 
-### Part D
-Calculate the $R^2$ for this model. How much variation in $Y$ is explained by our model?
+\vspace{2in}
 
-### Part E
-How large is the average residual?
+c. Suppose the OLS estimate of $\hat{\beta_1}$ has a standard error of $0.072$. Could we probably reject a null hypothesis of $H_0: \beta_1=0$ at the 95% level?
+
+\clearpage
+
+d. Calculate the $R^2$ for this model. How much variation in $Y$ is explained by our model?
+
+\vspace{2in}
+
+e. How large is the average residual?
+
+\clearpage
 
@@
 
-## Question 11
-
-- [<i class="fas fa-table"></i> `mlbattend.csv`](http://metricsf19.classes.ryansafner.com/data/mlbattend.csv)
-
-Download the `MLBattend` dataset. This data contains data on attendance at major league baseball games for all 32 MLB teams from the 1970s-2000. We want to answer the following question:
+11. Download the `MLBattend` dataset. This data contains data on attendance at major league baseball games for all 32 MLB teams from the 1970s-2000. We want to answer the following question:
 
@@
 
-### Part A
-
-Clean up the data a bit by making a new variable to measure home attendance in millions. This will make it easier to interpret your regression later on.
-
-### Part B
-
-Plot a scatterplot of Runs Scored (`y`) on Home Attendance (`x`). Add a regression line.
-
-### Part C
-
-Run a regression of Runs Scored on Home Attendance. What are $\beta_0$ and $\hat{\beta_1}$? Interpret them in the context of our question.
-
-### Part D
-
-Write out the estimated regression equation.
-
-### Part E
-
-Make a regression table of the output.
-
-### Part F
-
-Now let's start running some diagnostics of the regression. Make a histogram of the residuals. Do they look roughly normal?
-
-### Part G
-
-Make a residual plot.
-
-### Part H
-
+a. Clean up the data a bit by making a new variable to measure home attendance in millions. This will make it easier to interpret your regression later on.
+b. Plot a scatterplot of Runs Scored (`y`) on Home Attendance (`x`). Add a regression line.
+c. Run a regression of Runs Scored on Home Attendance. What are $\beta_0$ and $\hat{\beta_1}$? Interpret them in the context of our question.
+d. Write out the estimated regression equation.
+e. Make a regression table of the output.
+f. Now let's start running some diagnostics of the regression. Make a histogram of the residuals. Do they look roughly normal?
+g.
+Make a residual plot.h.
 Test the regression for heteroskedasticity.
@@
 Make a regression output table, with one column with regular standard errors and another with robust standard errors.
-
-### Part I
-
-Test the data for outliers. If there are any, identify which team(s) and season(s) are outliers.
-
-### Part J
-
-What is the marginal effect of home attendance on runs scored? Is this statistically significant? Why or why not?
-
-### Part K
-
+i. Test the data for outliers. If there are any, identify which team(s) and season(s) are outliers.
+j. What is the marginal effect of home attendance on runs scored? Is this statistically significant? Why or why not?
+k.
 Now we'll try out the `infer` package to understand the $p$-value and confidence interval for our observed slope in our regression model.
@@
 Compare to the $p$-value given by `lm()` and `tidy()` above.
-
-### Part L
+l.
 Make a histogram of the simulated slopes, and plot our sample slope on that histogram, shading the $p$-value.^[You can use `ggplot2` to plot a histogram in the normal way and add a `geom_vline()`, setting `xintercept` equal to your saved object with the sample $\hat{\beta_1}$ value.
@@
 Inside this, set `obs_stat` equal to your saved slope, and add `direction = "two_sided"`.]
-
-### Part M
+m.
 Get the 95% confidence interval for your slope estimate,^[`tidy()` your original regression, with `conf.int = TRUE` inside the command, then `select(conf.low, conf.high)` and `filter` by your `x` variable.
-Save this as an object.] and then make a histogram of the simulated slopes (like part L), but instead, add `+shade_confidence_interval()`.^[Inside of this, set `endpoints` equal to the object you just made with the low and high confidence interval values.] Compare this to what you get with `tidy()` above.
+Save this as an object.] and then make a histogram of the simulated slopes (like part L), but instead, add `+shade_confidence_interval()`.^[Inside of this, set `endpoints` equal to the object you just made with the low and high confidence interval values.] Compare this to what you get with `tidy()` above.**

### 2019-10-05 static/homeworks/03-problem-set.Rmd
Add problem set 3 materials

--- 2019-10-05 static/homeworks/03-problem-set-pdf.Rmd
@@
 title: "Problem Set 3"
-author: "ECON 480 - Fall 2019"
-date: "Due by Thursday, October 10, 2019"
-output: pdf_document
+author: "YOUR NAME HERE" # put your name here! 
+date: "ECON 480 - Fall 2019"
+output: html_document # change to pdf if you'd like, this will make a webpage (you can email it and open in any browser)
 ---
@@
 -->
+
+*Due by Thursday, October 10, 2019*
 
@@
 
-1. In your own words, describe what exogeneity and endogeneity mean, and how they are related to bias. What can we learn about the bias if we know $X$ is endogenous?
+## Question 1
+**In your own words, describe what exogeneity and endogeneity mean, and how they are related to bias. What can we learn about the bias if we know $X$ is endogenous?**
 
-\vspace{3in}
+<!--WRITE YOUR ANSWERS BELOW -->
 
-2. In your own words, describe what $R^2$ means. How do we calculate it, what does it tell us, and how do we interpret it?
+## Question 2
+**In your own words, describe what $R^2$ means. How do we calculate it, what does it tell us, and how do we interpret it?**
 
-\clearpage
+<!--WRITE YOUR ANSWERS BELOW -->
 
-3. In your own words, describe what the standard error of the regression ($SER$) means. How do we calculate it, what does it tell us, and how do we interpret it?
+## Question 3
 
-\vspace{3in}
+**In your own words, describe what the standard error of the regression ($SER$) means. How do we calculate it, what does it tell us, and how do we interpret it?**
 
-4. In your own words, describe what homoskedasticity and heteroskedasticity mean: both in ordinary English, and in terms of the graph of the OLS regression line.
+<!--WRITE YOUR ANSWERS BELOW -->
 
-\clearpage
+## Question 4
 
-5. In your own words, describe what the variation in $\hat{\beta_1}$ (either variance or standard error) means, or is measuring. What three things determine the variation, and in what way?
+**In your own words, describe what homoskedasticity and heteroskedasticity mean: both in ordinary English, and in terms of the graph of the OLS regression line.**
 
-\vspace{3in}
+<!--WRITE YOUR ANSWERS BELOW -->
 
-6. In your own words, describe what a $p$-value means, and how it is used to establish statistical significance.
+## Question 5
 
-\clearpage
+**In your own words, describe what the variation in $\hat{\beta_1}$ (either variance or standard error) means, or is measuring. What three things determine the variation, and in what way?**
 
-7.
-A researcher is interested in examining the impact of illegal music downloads on commercial music sales.
+<!--WRITE YOUR ANSWERS BELOW -->
+
+## Question 6
+
+**In your own words, describe what a $p$-value means, and how it is used to establish statistical significance.**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+## Question 7
+
+**A researcher is interested in examining the impact of illegal music downloads on commercial music sales.
 The author collects data on commercial sales of the top 500 singles from 2017 ($Y$) and the number of downloads from a web site that allows `file sharing' ($X$).
-The author estimates the following model
+The author estimates the following model**
 
@@
 
-> The author finds a large, positive, and statistically significant estimate of $\hat{\beta_1}$.
+**The author finds a large, positive, and statistically significant estimate of $\hat{\beta_1}$.
 The author concludes these results demonstrate that illegal downloads actually *boost* music sales.
@@
 Why or why not?
-Do you expect the estimate to overstate or understate the true relationship between illegal downloads and sales?
+Do you expect the estimate to overstate or understate the true relationship between illegal downloads and sales?**
 
-\vspace{3in}
+<!--WRITE YOUR ANSWERS BELOW -->
 
-8.
-A pharmaceutical company is interested in estimating the impact of a new drug on cholesterol levels.
+## Question 8
+
+**A pharmaceutical company is interested in estimating the impact of a new drug on cholesterol levels.
 They enroll 200 people in a clinical trial.
@@
 Half of the people are given the new drug and half the people are given a sugar pill with no active ingredient.
-To examine the impact of dosage on reductions in cholesterol levels, the authors of the study regress the following model:
+To examine the impact of dosage on reductions in cholesterol levels, the authors of the study regress the following model:**
 
@@
 
-> For people in the control group, dosage level$_i=0$ and for people in the treatment group, dosage level$_i$ measures milligrams of the active ingredient.
+**For people in the control group, dosage level$_i=0$ and for people in the treatment group, dosage level$_i$ measures milligrams of the active ingredient.
 In this case, the authors find a large, negative, statistically significant estimate of $\hat{\beta_1}$.
@@
 Why or why not?
-Do you expect the estimate to overstate or understate the true relationship between dosage and cholesterol level?
+Do you expect the estimate to overstate or understate the true relationship between dosage and cholesterol level?**
 
-\clearpage
+<!--WRITE YOUR ANSWERS BELOW -->
 
@@
 
-For the following questions, please *show all work* and explain answers as necessary.
+**For the following questions, please *show all work* and explain answers as necessary.
 You may lose points if you only write the correct answer.
-You may use `R` to *verify* your answers, but you are expected to reach the answers in this section "manually."
+You may use `R` to *verify* your answers, but you are expected to reach the answers in this section "manually."**
 
-9.
-A researcher wants to estimate the relationship between average weekly earnings $(AWE$, measured in dollars) and $Age$ (measured in years) using a simple OLS model.
-Using a random sample of college-educated full-time workers aged 25-65 yields the following:
+## Question 9
+
+**A researcher wants to estimate the relationship between average weekly earnings $(AWE$, measured in dollars) and $Age$ (measured in years) using a simple OLS model.
+Using a random sample of college-educated full-time workers aged 25-65 yields the following:**
 
@@
 
-a. Interpret what $\hat{\beta_0}$ means in this context.
+### Part A
 
-\vspace{1.5in}
+**Interpret what $\hat{\beta_0}$ means in this context.**
 
-b. Interpret what $\hat{\beta_1}$ means in this context.
+<!--WRITE YOUR ANSWERS BELOW -->
 
-\vspace{1.5in}
+### Part B
 
-c. The $R^2=0.023$ for this regression. What are the units of the $R^2$, and what does this mean?
+**Interpret what $\hat{\beta_1}$ means in this context.**
 
-\vspace{1.5in}
+<!--WRITE YOUR ANSWERS BELOW -->
 
-d. The $SER, \, \hat{\sigma_u}=624.1$ for this regression. What are the units of the SER in this context, and what does it mean? Is the SER large in the context of this regression?
+### Part C
 
-\clearpage
+**The $R^2=0.023$ for this regression. What are the units of the $R^2$, and what does this mean?**
 
-e. Suppose Maria is 20 years old. What is her predicted $\widehat{AWE}$?
+<!--WRITE YOUR ANSWERS BELOW -->
 
-\vspace{1.5in}
+### Part D
 
-f. Suppose the data shows her *actual* $AWE$ is $430. What is her residual? Is this a relatively good or a bad prediction?^[Hint: compare your answer here to your answer in Part D.]
+**The $SER, \, \hat{\sigma_u}=624.1$ for this regression. What are the units of the SER in this context, and what does it mean? Is the SER large in the context of this regression?**
 
-\vspace{1.5in}
+<!--WRITE YOUR ANSWERS BELOW -->
 
-g. What does the error term, $\hat{u_i}$ represent in this case? What might individuals have different values of $u_i$?
+### Part E
 
-\vspace{1.5in}
+**Suppose Maria is 20 years old. What is her predicted $\widehat{AWE}$?**
 
-h. Do you think that $Age$ is exogenous? Why or why not? Would we expect $\hat{\beta_1}$ to be too *large* or too *small*?
+<!--WRITE YOUR ANSWERS BELOW -->
 
-\clearpage
+### Part F
 
-10. Suppose a researcher is interested in estimating a simple linear regression model:
+**Suppose the data shows her *actual* $AWE$ is $430. What is her residual? Is this a relatively good or a bad prediction?^[Hint: compare your answer here to your answer in Part D.]**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+### Part G
+
+**What does the error term, $\hat{u_i}$ represent in this case? What might individuals have different values of $u_i$?**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+### Part H
+
+**Do you think that $Age$ is exogenous? Why or why not? Would we expect $\hat{\beta_1}$ to be too *large* or too *small*?**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+## Question 10
+
+**Suppose a researcher is interested in estimating a simple linear regression model:**
 
 $$Y_i=\beta_0+\beta_1X_i+u_i$$
-In a sample of 48 observations, she generates the following descriptive statistics:
+**In a sample of 48 observations, she generates the following descriptive statistics:**
 
@@
 
-a. What is the OLS estimate of $\hat{\beta_1}$?
+### Part A
 
-\vspace{2in}
+**What is the OLS estimate of $\hat{\beta_1}$?**
 
-b. What is the OLS estimate of $\hat{\beta_0}$?
+<!--WRITE YOUR ANSWERS BELOW -->
 
-\vspace{2in}
+### Part B
 
-c. Suppose the OLS estimate of $\hat{\beta_1}$ has a standard error of $0.072$. Could we probably reject a null hypothesis of $H_0: \beta_1=0$ at the 95% level?
+**What is the OLS estimate of $\hat{\beta_0}$?**
 
-\clearpage
+<!--WRITE YOUR ANSWERS BELOW -->
 
-d. Calculate the $R^2$ for this model. How much variation in $Y$ is explained by our model?
+### Part C
 
-\vspace{2in}
+**Suppose the OLS estimate of $\hat{\beta_1}$ has a standard error of $0.072$. Could we probably reject a null hypothesis of $H_0: \beta_1=0$ at the 95% level?**
 
-e. How large is the average residual?
+<!--WRITE YOUR ANSWERS BELOW -->
 
-\clearpage
+### Part D
+
+**Calculate the $R^2$ for this model. How much variation in $Y$ is explained by our model?**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+### Part E
+
+**How large is the average residual?**
+
+<!--WRITE YOUR ANSWERS BELOW -->
 
@@
 
-11. Download the `MLBattend` dataset. This data contains data on attendance at major league baseball games for all 32 MLB teams from the 1970s-2000. We want to answer the following question:
+## Question 11
+
+- [<i class="fas fa-table"></i> `mlbattend.csv`](http://metricsf19.classes.ryansafner.com/data/mlbattend.csv)
+
+Download the `MLBattend` dataset. This data contains data on attendance at major league baseball games for all 32 MLB teams from the 1970s-2000. We want to answer the following question:
 
@@
 
-a. Clean up the data a bit by making a new variable to measure home attendance in millions. This will make it easier to interpret your regression later on.
-b. Plot a scatterplot of Runs Scored (`y`) on Home Attendance (`x`). Add a regression line.
-c. Run a regression of Runs Scored on Home Attendance. What are $\beta_0$ and $\hat{\beta_1}$? Interpret them in the context of our question.
-d. Write out the estimated regression equation.
-e. Make a regression table of the output.
-f. Now let's start running some diagnostics of the regression. Make a histogram of the residuals. Do they look roughly normal?
-g.
-Make a residual plot.h.
-Test the regression for heteroskedasticity.
+### Part A
+
+**Clean up the data a bit by making a new variable to measure home attendance in millions. This will make it easier to interpret your regression later on.**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+### Part B
+
+**Plot a scatterplot of Runs Scored (`y`) on Home Attendance (`x`). Add a regression line.**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+### Part C
+
+**Run a regression of Runs Scored on Home Attendance. What are $\beta_0$ and $\hat{\beta_1}$? Interpret them in the context of our question.**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+### Part D
+
+**Write out the estimated regression equation.**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+### Part E
+
+**Make a regression table of the output.**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+### Part F
+
+**Now let's start running some diagnostics of the regression. Make a histogram of the residuals. Do they look roughly normal?**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+### Part G
+
+**Make a residual plot.**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+### Part H
+
+**Test the regression for heteroskedasticity.
 Are the errors homoskedastic or heteroskedastic?
 Generate robust standard errors.
-Make a regression output table, with one column with regular standard errors and another with robust standard errors.
-i. Test the data for outliers. If there are any, identify which team(s) and season(s) are outliers.
-j. What is the marginal effect of home attendance on runs scored? Is this statistically significant? Why or why not?
-k.
-Now we'll try out the `infer` package to understand the $p$-value and confidence interval for our observed slope in our regression model.
+Make a regression output table, with one column with regular standard errors and another with robust standard errors.**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+### Part I
+
+**Test the data for outliers. If there are any, identify which team(s) and season(s) are outliers.**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+### Part J
+
+**What is the marginal effect of home attendance on runs scored? Is this statistically significant? Why or why not?**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+### Part K
+
+**Now we'll try out the `infer` package to understand the $p$-value and confidence interval for our observed slope in our regression model.
 Save the (value of) our sample $\hat{\beta_1}$ from your regression in Part C as an object.
@@
 Then, use this to `get_p_value()`^[Set `obs_stat` equal to your observed slope, and set `direction = "two_sided"`].
-Compare to the $p$-value given by `lm()` and `tidy()` above.
-l.
-Make a histogram of the simulated slopes, and plot our sample slope on that histogram, shading the $p$-value.^[You can use `ggplot2` to plot a histogram in the normal way and add a `geom_vline()`, setting `xintercept` equal to your saved object with the sample $\hat{\beta_1}$ value.
+Compare to the $p$-value given by `lm()` and `tidy()` above.**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+### Part L
+
+**Make a histogram of the simulated slopes, and plot our sample slope on that histogram, shading the $p$-value.^[You can use `ggplot2` to plot a histogram in the normal way and add a `geom_vline()`, setting `xintercept` equal to your saved object with the sample $\hat{\beta_1}$ value.
 *Alternatively*, you can use `infer` to pipe your `tibble` of simulations into `visualize()`, and inside `visualize()` set `obs_stat` equal to your saved $\hat{\beta_1}$ object.
 *Regardless* of which method you use, add `+shade_p_value()`.
-Inside this, set `obs_stat` equal to your saved slope, and add `direction = "two_sided"`.]
-m.
-Get the 95% confidence interval for your slope estimate,^[`tidy()` your original regression, with `conf.int = TRUE` inside the command, then `select(conf.low, conf.high)` and `filter` by your `x` variable.
+Inside this, set `obs_stat` equal to your saved slope, and add `direction = "two_sided"`.]**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+### Part M
+
+**Get the 95% confidence interval for your slope estimate,^[`tidy()` your original regression, with `conf.int = TRUE` inside the command, then `select(conf.low, conf.high)` and `filter` by your `x` variable.
 Save this as an object.] and then make a histogram of the simulated slopes (like part L), but instead, add `+shade_confidence_interval()`.^[Inside of this, set `endpoints` equal to the object you just made with the low and high confidence interval values.] Compare this to what you get with `tidy()` above.**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```

### 2019-10-05 content/assignment/03-problem-set.Rmd
Sneak in part b to find correlation in HW 3 problem 11...

@@
 
-Plot a scatterplot of Runs Scored (`y`) on Home Attendance (`x`). Add a regression line.
+Get the correlation between Runs Scored and Home Attendance.
 
@@
 
-Run a regression of Runs Scored on Home Attendance. What are $\beta_0$ and $\hat{\beta_1}$? Interpret them in the context of our question.
+Plot a scatterplot of Runs Scored (`y`) on Home Attendance (`x`). Add a regression line.
 
@@
 
-Write out the estimated regression equation.
+Run a regression of Runs Scored on Home Attendance. What are $\beta_0$ and $\hat{\beta_1}$? Interpret them in the context of our question.
 
@@
 
-Make a regression table of the output.
+Write out the estimated regression equation.
 
@@
 
-Now let's start running some diagnostics of the regression. Make a histogram of the residuals. Do they look roughly normal?
+Make a regression table of the output.
 
@@
 
+Now let's start running some diagnostics of the regression. Make a histogram of the residuals. Do they look roughly normal?
+
+### Part H
+
 Make a residual plot.
 
-### Part H
+### Part I
 
@@
 
-### Part I
+### Part J
 
@@
 
-### Part J
+### Part K
 
@@
 
-### Part K
+### Part L
 
@@
 
-### Part L
+### Part M
+
 Make a histogram of the simulated slopes, and plot our sample slope on that histogram, shading the $p$-value.^[You can use `ggplot2` to plot a histogram in the normal way and add a `geom_vline()`, setting `xintercept` equal to your saved object with the sample $\hat{\beta_1}$ value.
@@
 
-### Part M
+### Part N
+
 Get the 95% confidence interval for your slope estimate,^[`tidy()` your original regression, with `conf.int = TRUE` inside the command, then `select(conf.low, conf.high)` and `filter` by your `x` variable.

### 2019-10-05 static/homeworks/03-problem-set-pdf.Rmd
Sneak in part b to find correlation in HW 3 problem 11...
Problem set 3 question 11 part L, refers to regression in part C, now its part D

@@
 a. Clean up the data a bit by making a new variable to measure home attendance in millions. This will make it easier to interpret your regression later on.
-b. Plot a scatterplot of Runs Scored (`y`) on Home Attendance (`x`). Add a regression line.
-c. Run a regression of Runs Scored on Home Attendance. What are $\beta_0$ and $\hat{\beta_1}$? Interpret them in the context of our question.
-d. Write out the estimated regression equation.
-e. Make a regression table of the output.
-f. Now let's start running some diagnostics of the regression. Make a histogram of the residuals. Do they look roughly normal?
-g.
-Make a residual plot.h.
+b. Get the correlation between Runs Scored and Home Attendance.
+c. Plot a scatterplot of Runs Scored (`y`) on Home Attendance (`x`). Add a regression line.
+d. Run a regression of Runs Scored on Home Attendance. What are $\beta_0$ and $\hat{\beta_1}$? Interpret them in the context of our question.
+e. Write out the estimated regression equation.
+f. Make a regression table of the output.
+g. Now let's start running some diagnostics of the regression. Make a histogram of the residuals. Do they look roughly normal?
+h. Make a residual plot.
+i.
 Test the regression for heteroskedasticity.
@@
 Make a regression output table, with one column with regular standard errors and another with robust standard errors.
-i. Test the data for outliers. If there are any, identify which team(s) and season(s) are outliers.
-j. What is the marginal effect of home attendance on runs scored? Is this statistically significant? Why or why not?
-k.
+j. Test the data for outliers. If there are any, identify which team(s) and season(s) are outliers.
+k. What is the marginal effect of home attendance on runs scored? Is this statistically significant? Why or why not?
+l.
 Now we'll try out the `infer` package to understand the $p$-value and confidence interval for our observed slope in our regression model.
-Save the (value of) our sample $\hat{\beta_1}$ from your regression in Part C as an object.
+Save the (value of) our sample $\hat{\beta_1}$ from your regression in Part D as an object.
 Then, install and load the `infer` package, and then calculate the slope^[`calculate(stat = "slope")`] under the null hypothesis that there is no connection between attendance and runs.^[`hypothesize(null = "independence")`] for 1000 additional simulated samples^[`generate(reps = 1000, type = "permute")`], and save this as an object (it's a `tibble`).
@@
 Compare to the $p$-value given by `lm()` and `tidy()` above.
-l.
+m.
 Make a histogram of the simulated slopes, and plot our sample slope on that histogram, shading the $p$-value.^[You can use `ggplot2` to plot a histogram in the normal way and add a `geom_vline()`, setting `xintercept` equal to your saved object with the sample $\hat{\beta_1}$ value.
@@
 Inside this, set `obs_stat` equal to your saved slope, and add `direction = "two_sided"`.]
-m.
+n.
 Get the 95% confidence interval for your slope estimate,^[`tidy()` your original regression, with `conf.int = TRUE` inside the command, then `select(conf.low, conf.high)` and `filter` by your `x` variable.

### 2019-10-08 static/slides/12-slides.Rmd
Update slides 12

@@
 ---
-title: "2.6: Inferential Statistics I"
+title: "2.6: Inference for Regression"
 subtitle: "ECON 480 · Econometrics · Fall 2019"
@@
 ]
+
+---
+
+# Classical Statistical Inference: Critical Values of Test Statistic
+
+- .whisper[Test statistic `\\((\delta)\\)`]: measures **how far what we observed in our sample $(\hat{\beta_1})$ is from what we would expect if the null hypothesis were true $(\beta_1=0)$**
+    - Calculated from a sampling distribution of the estimator (i.e. $\hat{\beta_1})$
+    - In econometrics, we use $t$-distributions which have $n-k-1$ degrees of freedom<sup>.red</sup>
+
+- .whisper[Rejection region]: if the test statistic reaches a .onfire["critical value"] of $\delta$, then we **reject** the null hypothesis
+
+.footnote[<sup>.red[1]</sup> Again, see today's [class notes](/class/12-class) for more on the t-distribution.
+`\\(k\\)` is the number of independent variables our model has, in this case, with just one `\\(X\\)`, `\\(k=1\\)`.
+We use two degrees of freedom to calculate `\\(\hat{\beta_0}\\)` and `\\(\hat{\beta_1}\\)`, hence we have `\\(n-2\\)` df.]
+
+---
+
+class: inverse, center, middle
+
+# Simulating the Sampling Distribution with *infer*
+
+---
+
+# Imagine a Null World, where $H_0$ is True
+
+.center[
+![:scale 70%](https://www.dropbox.com/s/x7b5azt5algjodg/parallelworld.jpeg?raw=1)
+
+Our world, and a world where $\beta_1=0$ by assumption.
+]
+
+---
+
+# Comparing the Worlds I
+
+- From that null world where $H_0: \, \beta_1=0$ is true, we **simulate** another sample and calculate OLS estimators again
+
+--
+
+.pull-left[
+
+# Our Sample
+
+```{r, echo=F}
+tidy(school_reg)
+```
+]
+
+--
+
+.pull-right[
+
+# Another Sample
+
+```{r, echo = F}
+library(infer)
+CASchool %>%
+  specify(testscr ~ str) %>%
+  hypothesize(null = "independence") %>%
+  generate(reps = 1, type = "permute") %>%
+  lm(testscr ~ str, data = .) %>%
+  tidy()
+```
+]
+
+---
+
+# Comparing the Worlds II
+
+- From that null world where $H_0: \, \beta_1=0$ is true, let's **simulate 1,000** samples and calculate slope $(\hat{\beta_1})$ for each
+
+```{r}
+CASchool %>%
+  specify(testscr ~ str) %>%
+  hypothesize(null = "independence") %>%
+  generate(reps = 1000, type = "permute") %>%
+  calculate(stat = "slope")%>%
+  rename(sample = replicate,
+         slope = stat)
+```
+
+---
+
+# Prepping the *infer*  Pipeline
+
+- Before I show you how to do this, let's first save our estimated slope from our *actual* sample
+    - We'll want this later!
+
+```{r}
+# save as obs_slope
+sample_slope <- school_reg_tidy %>% # this is the regression tidied with broom
+  filter(term=="str") %>%
+  pull(estimate)
+
+# confirm what it is
+sample_slope
+```
+
+---
+# The *infer*  Pipeline: Specify
+
+.center[
+![:scale 65%](https://www.dropbox.com/s/25axifo95oz5zs7/infer1jpeg.jpeg?raw=1)
+]
+
+---
+
+# The *infer*  Pipeline: Specify
+
+.left-code[
+### Specify
+```{r, echo=T, eval=F}
+data %>%
+  specify(y ~ x)
+```
+]
+
+.right-plot[
+- Take our data and pipe it into the `specify()` function, which is essentially a `lm()` function for regression (for our purposes)
+
+```{r, echo=T, eval=F}
+CASchool %>%
+  specify(testscr ~ str)%>%
+  head(.,3)
+```
+
+```{r}
+CASchool %>%
+  specify(testscr ~ str)%>%
+  head(.,3)
+```
+]
+
+- Note nothing happens yet
+
+---
+
+# The *infer*  Pipeline: Hypothesize
+
+.center[
+![:scale 65%](https://www.dropbox.com/s/d9bnryxznfwe4fc/infer2jpeg.jpeg?raw=1)
+]
+
+---
+
+# The *infer*  Pipeline: Hypothesize
+
+.left-code[
+### Specify
+### Hypothesize
+```{r, echo=T, eval=F}
+%>% hypothesize(null = "independence")
+```
+]
+
+.right-plot[
+- Describe what the null hypothesis is here
+- In `infer`'s language, we are hypothesizing that `str` and `testscr` are `independent` $(\beta_1=0)$<sup>.red[1]</sup>
+
+```{r, echo=T, eval=F}
+CASchool %>%
+  specify(testscr ~ str) %>%
+  hypothesize(null = "independence")
+```
+
+```{r}
+CASchool %>%
+  specify(testscr ~ str) %>%
+  hypothesize(null = "independence") %>%
+  head(.,3)
+```
+]
+
+- Note nothing happens yet
+
+.footnote[<sup>.red[1]</sup> `type` can be either `point` (for specific point estimates for a single variable, such as a sample mean, `\\((\bar{x})\\)`, or `independence` (for hypotheses about two samples or a relationship between variables).
+See more [here](https://moderndive.netlify.com/9-hypothesis-testing.html).]
+---
+
+# The *infer*  Pipeline: Generate I
+
+.center[
+![:scale 65%](https://www.dropbox.com/s/pc8733aurnouvsr/infer3jpeg.jpeg?raw=1)
+]
+
+---
+
+# The *infer*  Pipeline: Generate I
+
+.left-code[
+### Specify
+### Hypothesize
+### Generate
+```{r, echo=T, eval=F}
+%>% generate(reps = n, type = "permute")
+```
+]
+
+.right-plot[
+- Now the magic starts, as we run a number of simulated samples<sup>.red[1]</sup>
+- Set the number of `reps` and set the `type` equal to `"permute"`
+
+```{r, echo=T, eval=F}
+i %>%
+  generate(reps = 1000, type = "permute")
+```
+
+]
+
+.footnote[<sup>.red[1]</sup> Note for spacing on the slide, I saved the previous code as `i` and pipe it into the remainder.]
+
+---
+
+# The *infer*  Pipeline: Generate II
+
+.left-code[
+### Specify
+### Hypothesize
+### Generate
+```{r, echo=T, eval=F}
+%>% generate(reps = n, type = "permute")
+```
+]
+
+.right-plot[
+
+```{r}
+i<-CASchool %>%
+  specify(testscr ~ str) %>%
+  hypothesize(null = "independence")
+i %>%
+  generate(reps = 1000, type = "permute")
+```
+]
+
+---
+
+# The *infer*  Pipeline: Generate III
+
+.left-code[
+### Specify
+### Hypothesize
+### Generate
+```{r, echo=T, eval=F}
+%>% generate(reps = n, type = "permute")
+```
+]
+
+.right-plot[
+
+- There are two types of simulations we can run:<sup>.red[1]</sup>
+- `"bootstrap"` takes a random draw of our *existing* sample's observations (of the same number of observations) *with replacement*
+  - this approximates a sampling distribution
+- `"permute"` is a `bootstrap` *without* replacement
+]
+
+.footnote[<sup>.red[1]</sup> You can do either of these in base R with `sample()`, which has 3 arguments: a vector to sample from, `size` (number of obs), and `replace` equal to `TRUE` or `FALSE`.
+See more for `infer`  [here](https://moderndive.netlify.com/9-hypothesis-testing.html).]
+
+---
+
+# The *infer*  Pipeline: Calculate I
+
+.center[
+![:scale 65%](https://www.dropbox.com/s/4wojnmworf406ku/infer4jpeg.jpeg?raw=1)
+]
+
+---
+
+# The *infer*  Pipeline: Calculate I
+
+.left-code[
+### Specify
+### Hypothesize
+### Generate
+### Calculate
+```{r, echo=T, eval=F}
+%>% calculate(stat = "")
+```
+]
+
+.right-plot[
+- We `calculate` sample statistics for each of the 1,000 `replicate` samples
+
+- In our case, calculate the slope, $(\hat{\beta_1})$ for each `replicate`
+
+```{r, echo=T, eval=F}
+i %>%
+  generate(reps = 1000, type = "permute") %>%
+  calculate(stat = "slope")
+```
+
+- Other `stat`s for calculation: `"mean"`, `"median"`, `"prop"`, `"diff in means"`, `"diff in props"`, etc. (see [package information](https://infer.netlify.com/))
+]
+
+---
+
+# The *infer*  Pipeline: Calculate II
+
+.left-code[
+### Specify
+### Hypothesize
+### Generate
+### Calculate
+```{r, echo=T, eval=F}
+%>% calculate(stat = "")
+```
+]
+
+.right-plot[
+```{r, echo=F}
+i %>%
+  generate(reps = 1000, type = "permute") %>%
+  calculate(stat = "slope")
+```
+]
+
+---
+
+# The *infer*  Pipeline: Get p Value
+
+.left-code[
+### Specify
+### Hypothesize
+### Generate
+### Calculate
+### Get p Value
+```{r, echo=T, eval=F}
+%>% get_p_value(obs stat = "",
+                direction = "both")
+```
+]
+
+.right-plot[
+
+- We can calculate the .shout[`\\(p\\)`-value]
+    - the probability of seeing a value at least as large as our `sample_slope` (`r round(sample_slope, 3)`) in our simulated null distribution
+
+- onfire[Two-sided alternative] $H_a: \beta_1 \neq 0$, we double the raw $p$-value
+
+```{r, echo=F}
+simulations <- i %>%
+  generate(reps = 1000, type = "permute") %>%
+  calculate(stat = "slope")
+```
+
+```{r, echo=T}
+simulations %>%
+  get_p_value(obs_stat = sample_slope,
+              direction = "both")
+```
+]
+
+.footnote[<sup>.red[+]</sup> Note here I saved the results of our previous code as `simulations` for spacing.]
+
+---
+
+# The *infer*  Pipeline: Get Confidence Interval
+
+.left-code[
+### Specify
+### Hypothesize
+### Generate
+### Calculate
+### Get confidence interval
+```{r, echo=T, eval=F}
+%>% get_ci(
+  level = 0.95,
+  type = "se",
+  point_estimate = "")
+```
+]
+
+.right-plot[
+
+- We can calculate the .shout[confidence interval] for $\beta_1$ from our `sample_slope` $\hat{\beta_1}$ of `r round(sample_slope, 3)`
+    - $CI_{0.95}= \hat{\beta_1} \pm 2 \times se(\hat{\beta_1})$
+    - Specify confidence level $(1-\alpha)$, often $0.95$
+    - Supply our original estimate for `point_estimate`
+
+```{r, echo=T}
+simulations %>%
+  get_confidence_interval(level = 0.95,
+                          type = "se",
+                          point_estimate = sample_slope)
+```
+]
+
+---
+
+# The *infer*  Pipeline: Visualize I
+
+.center[
+![:scale 65%](https://www.dropbox.com/s/6328hg5t03jbd0w/infer5jpeg.jpeg?raw=1)
+]
+
+---
+
+# The *infer*  Pipeline: Visualize I
+
+.left-code[
+### Specify
+### Hypothesize
+### Generate
+### Calculate
+### Visualize
+```{r, echo=T, eval=F}
+%>% visualize()
+```
+]
+
+.right-plot[
+
+- Make a histogram of our null distribution of $\beta_1$
+    - Note it is centered at $\beta_1=0$ because that's $H_0$!
+
+```{r, echo=T, fig.height=4, fig.retina=3, fig.align="center"}
+simulations %>%
+ visualize()
+```
+]
+
+---
+
+# The *infer*  Pipeline: Visualize II
+
+.left-code[
+### Specify
+### Hypothesize
+### Generate
+### Calculate
+### Visualize
+```{r, echo=T, eval=F}
+%>% visualize()
+```
+]
+
+.right-plot[
+
+- Add our `sample_slope` to show our finding on the null distr.
+
+```{r, echo=T, fig.height=4, fig.retina=3, fig.align="center"}
+simulations %>%
+ visualize(obs_stat = sample_slope)
+```
+]
+
+---
+
+# The *infer*  Pipeline: Visualize p-value
+
+.left-code[
+### Specify
+### Hypothesize
+### Generate
+### Calculate
+### Visualize
+```{r, echo=T, eval=F}
+%>% visualize()+
+  shade_p_value()
+```
+]
+
+.right-plot[
+
+- Add `shade_p_value` to see what $p$ is
+
+```{r, echo=T, fig.height=4, fig.retina=3, fig.align="center"}
+simulations %>%
+ visualize(obs_stat = sample_slope)+
+  shade_p_value(obs_stat = sample_slope,
+                direction = "two_sided")
+```
+]
+
+---
+
+# The *infer*  Pipeline: Visualize Confidence Intervals
+
+.left-code[
+### Specify
+### Hypothesize
+### Generate
+### Calculate
+### Visualize
+```{r, echo=T, eval=F}
+%>% visualize()+
+  shade_ci()
+```
+]
+
+.right-plot[
+
+- To shade confidence interval, we first need a vector of what they are
+  - I've saved the outputted `tibble` of them from 4 slides ago as `ci_values`
+
+```{r}
+ci_values<-simulations %>%
+  get_confidence_interval(level = 0.95,
+                          type = "se",
+                          point_estimate = sample_slope)
+```
+
+```{r, echo=T, fig.height=4, fig.retina=3, fig.align="center"}
+simulations %>%
+ visualize(obs_stat = sample_slope)+
+  shade_confidence_interval(ci_values)
+```
+]
+
+---
+
+# The *infer*  Pipeline: Visualize is a Wrapper of ggplot
+
+- `infer`'s `visualize()` function is just a wrapper function for `ggplot()`
+  - you can take your `simulations` `tibble` and just `ggplot` a normal histogram
+
+--
+
+.pull-left[
+```{r visualize-gg, echo=T, eval=F}
+simulations %>%
+  ggplot(data = .)+
+  aes(x = stat)+
+  geom_histogram(color="white", fill="indianred")+
+  geom_vline(xintercept = sample_slope,
+             color = "blue",
+             size = 2,
+             linetype = "dashed")+
+  labs(x = expression(paste("Distribution of ", hat(beta[1]), " under ", H[0], " that ", beta[1]==0)),
+       y = "Samples")+
+    theme_classic(base_family = "Fira Sans Condensed",
+           base_size=20)
+```
+]
+
+--
+
+.pull-right[
+```{r, ref.label="visualize-gg", fig.height=5, fig.retina=3, fig.align="center"}
+
+```
+]
+
+---
+
+# What R Does: Classical Statistical Inference I
+
+.pull-left[
+
+- R does things the old-fashioned way, using a *theoretical* null distribution instead of *simulation*
+
+- A .shout[`\\(t\\)`-distribution] with $n-k-1$ df<sup>1</sup>
+
+- Calculate a $t$-statistic for $\hat{\beta_1}$:
+
+$$\text{test statistic} = \frac{\text{estimate} - \text{null hypothesis}}{\text{standard error of estimate}}$$
+
+]
+
+.pull-right[
+
+```{r, fig.retina=3, fig.align="center"}
+t_ex<-ggplot(data = tibble(x=-5:5))+
+  aes(x = x)+
+  stat_function(fun = dt, args=list(df = 418), size=2, color="blue")+
+  #stat_function(fun=dt, geom="area", fill="blue", xlim=c(-5,4.75), alpha=0.5)+
+  #stat_function(fun=dt, geom="area", fill="blue", xlim=c(4.75,5), alpha=0.5)+
+  geom_vline(xintercept = c(-4.75, 4.75), size = 1, color = "red", linetype = "dashed")+
+  geom_rect(xmin=-5,xmax=-4.75, ymin=0, ymax=0.4, fill="red", alpha=0.3)+
+  geom_rect(xmin=4.75,xmax=5, ymin=0, ymax=0.4, fill="red", alpha=0.3)+
+  labs(x = "t",
+       y = "Probability")+
+  scale_x_continuous(breaks = seq(-5,5,1),
+                     lim = c(-5,5),
+                     expand = c(0,0))+
+  scale_y_continuous(breaks = seq(0,0.4,0.1),
+                     lim = c(0,0.4),
+                     expand = c(0,0))+
+  theme_classic(base_family = "Fira Sans Condensed",
+           base_size=20)
+t_ex
+```
+
+]
+
+.footnote[<sup>1</sup> $k$ is the number of $X$ variables.]
+
+---
+
+# What R Does: Classical Statistical Inference II
+
+.pull-left[
+
+$$\text{test statistic} = \frac{\text{estimate} - \text{null hypothesis}}{\text{standard error of estimate}}$$
+
+- $t$ has the same interpretation as $Z$, number of std. dev. away from the distribution's center<sup>1</sup>
+
+- Compares to a critical value of $t^*$ (determined by $\alpha$ & $n-k-1$)
+    - For 95% confidence, $\alpha=0.05$, $t^* \approx 2$<sup><span class="red">2</span></sup>
+]
+
+.pull-right[
+
+```{r, fig.retina=3, fig.align="center"}
+t_ex
+```
+
+]
+
+.footnote[<sup>1</sup> Think of our simulated distribution, the center was 0.
+
+<sup>2</sup> The 68-95-99.7% empirical rule!]
+
+---
+
+# What R Does: Classical Statistical Inference III
+
+.pull-left[
+
+$$\begin{align*}
+t &= \frac{\hat{\beta_1}-\beta_{1,0}}{se(\hat{\beta_1})}\\
+t &= \frac{-2.28-0}{0.48}\\
+t &= -4.75\\ \end{align*}$$
+
+- Our sample slope is 4.75 standard deviations below the mean under $H_0$
+
+- $p$-value: prob. of a test statistic at least as large (in magnitude) as ours if the null hypothesis were true<sup>1</sup>
+  - $p$-value is **2-sided** for $H_a: \beta_1 \neq 0$
+]
+
+.pull-right[
+
+```{r, fig.retina=3, fig.align="center"}
+t_ex
+```
+
+]
+
+.footnote[<sup>1</sup> Think of our simulated distribution, the center was 0.]
+
+---
+
+# 1-Sided vs. 2-Sided p-values I
+
+.pull-left[
+
+$H_a: \beta_1<0$
+
+$p$-value: $Prob(t < t_i)$
+
+```{r, fig.retina=3, fig.height=5, fig.align="center"}
+ggplot(data = tibble(x=-4:4))+
+  aes(x = x)+
+  stat_function(fun = dnorm, size=2, color="blue")+
+  stat_function(fun=dnorm, xlim=c(-4,-2), geom="area", fill="blue", alpha=0.5)+
+  labs(x = "t",
+       y = "Probability")+
+  scale_x_continuous(breaks=c(-2),
+                      label=c(expression(-t[i])))+
+  theme_classic(base_family = "Fira Sans Condensed",
+           base_size=20)
+
+```
+
+]
+.pull-right[
+
+$H_a: \beta_1>0$
+
+$p$-value: $Prob(t > t_i)$
+
+```{r, fig.retina=3, fig.height=5, fig.align="center"}
+ggplot(data = tibble(x=-4:4))+
+  aes(x = x)+
+  stat_function(fun = dnorm, size=2, color="blue")+
+  stat_function(fun=dnorm, xlim=c(2,4), geom="area", fill="blue", alpha=0.5)+
+  labs(x = "t",
+       y = "Probability")+
+  scale_x_continuous(breaks=c(2),
+                      label=c(expression(t[i])))+
+  theme_classic(base_family = "Fira Sans Condensed",
+           base_size=20)
+```
+
+]
+
+---
+
+# 1-Sided vs. 2-Sided p-values I
+
+$H_a: \beta_1 \neq 0$
+
+$p$-value: $2 \times Prob(t > |t_i|)$
+
+```{r, fig.retina=3, fig.height=5, fig.align="center"}
+ggplot(data = tibble(x=-4:4))+
+  aes(x = x)+
+  stat_function(fun = dnorm, size=2, color="blue")+
+  stat_function(fun=dnorm, xlim=c(-4,-2), geom="area", fill="blue", alpha=0.5)+
+  stat_function(fun=dnorm, xlim=c(2,4), geom="area", fill="blue", alpha=0.5)+
+  labs(x = "t",
+       y = "Probability")+
+  scale_x_continuous(breaks=c(-2,2),
+                      label=c(expression(-t[i]), expression(t[i])))+
+  theme_classic(base_family = "Fira Sans Condensed",
+           base_size=20)
+```
+
+---
+
+# Hypothesis Tests in Regression Output I
+
+```{r, echo = T}
+summary(school_reg)
+```
+
+---
+
+# Hypothesis Tests in Regression Output II
+
+- In `broom`'s `tidy()` (with confidence intervals)
+
+```{r, echo = T}
+tidy(school_reg, conf.int=TRUE)
+```
+
+---
+
+# Conclusions
+
+- .whisper[We have sufficient evidence to reject `\\(H_0\\)` in favor of our alternative hypothesis, that there *is a relationship* between class size and test scores].
+    -
+
+--
+
+- .whisper[We are 95% confident that the true relationship is between `\\(-3.22\\)` and `\\(-1.34\\)`.]
+
+---
+
+# Hypothesis Testing and Confidence Intervals: Relationship
+
+- Confidence intervals are all *two-sided* by nature
+$$CI_{0.95}=\left(\left[\hat{\beta_1}-2 \times se(\hat{\beta_1})\right], \, \left[\hat{\beta_1}+2 \times se(\hat{\beta_1}\right]) \right)$$
+
+- Hypothesis test $(t$-test) of $H_0: \, \beta_1=0$ computes a $t$-value of<sup>.red[1]</sup>
+$$t=\frac{\hat{\beta_1}}{se(\hat{\beta_1})}$$
+and $p$<0.05 when $t\geq2$
+
+- .onfire[If a confidence interval contains the $H_0$ value (i.e. $0$, for our test), then *we fail to reject* $H_0$.]
+
+.footnote[<sup>.red[1]</sup> Technically the `\\(t\\)`-value is `\\(\frac{\hat{\beta_1}-\beta_{1,0}}{se(\hat{\beta_1})}\\)`, but our `\\(H_0\\)` is often that `\\(beta_{1,0}=0\\)`, such that the formula reduces to the one above.]
+
+---
+
+# Significance In Regression Tables
+
+.pull-left[
+.font80[
+.regtable[
+```{r}
+library(huxtable)
+huxreg("Test Score" = school_reg,
+       coefs = c("Intercept" = "(Intercept)",
+                 "STR" = "str"),
+       statistics = c("N" = "nobs",
+                      "R-Squared" = "r.squared",
+                      "SER" = "sigma"),
+       number_format = 2)
+```
+]
+]
+]
+
+.pull-right[
+
+- Statistical significance is shown by asterisks, common (but not always!) standard:
+    - 1 asterisk: significant at $\alpha=0.10$
+    - 2 asterisks: significant at $\alpha=0.05$
+    - 3 asterisks: significant at $\alpha=0.01$
+
+- Rare, but sometimes regression tables include $p$-values for estimates
+]

### 2019-10-08 static/slides/12-slides.Rmd
Tweak slides 12
Tweaks to slides 12 (formatting fixes, and some more about p-values)

@@
 
-- We want to .whisper[identify] causal relationships between **population** variables
+- We want to .onfire[identify] causal relationships between **population** variables
     - Logically first thing to consider
@@
 
-- We'll use **sample** *statistics* to .whisper[infer] something about population *parameters*
+- We'll use **sample** *statistics* to .onfire[infer] something about population *parameters*
     - In practice, we'll only ever have a finite *sample distribution* of data
@@
 
-- .whisper[All modern science is built upon statistical hypothesis testing, so understand it well!]
+- .onfire[All modern science is built upon statistical hypothesis testing, so understand it well!]
 
@@
 
-- Modern philosophy of science is largely based off of hypothesis testing and .onfire[falsifiability], which form the "Scientific Method"<sup>1</sup>
-
-- For something to be "scientific", it must be .whisper[falsifiable], or at least .onfire[testable]
+- Modern philosophy of science is largely based off of hypothesis testing and .onfire[falsifiability], which form the "Scientific Method"<sup>.red[1]</sup>
+
+- For something to be "scientific", it must be .onfire[falsifiable], or at least .onfire[testable]
 
@@
 
-- .whisper[Test statistic `\\((\delta)\\)`]: measures **how far what we observed in our sample $(\hat{\beta_1})$ is from what we would expect if the null hypothesis were true $(\beta_1=0)$**
+- .onfire[Test statistic `\\((\delta)\\)`]: measures **how far what we observed in our sample $(\hat{\beta_1})$ is from what we would expect if the null hypothesis were true $(\beta_1=0)$**
     - Calculated from a sampling distribution of the estimator (i.e. $\hat{\beta_1})$
-    - In econometrics, we use $t$-distributions which have $n-k-1$ degrees of freedom<sup>.red</sup>
-
-- .whisper[Rejection region]: if the test statistic reaches a .onfire["critical value"] of $\delta$, then we **reject** the null hypothesis
+    - In econometrics, we use $t$-distributions which have $n-k-1$ degrees of freedom<sup>.red[1]</sup>
+
+- .onfire[Rejection region]: if the test statistic reaches a .onfire["critical value"] of $\delta$, then we **reject** the null hypothesis
 
@@
 
-```{r}
+```{r, echo=T}
 # save as obs_slope
@@
 CASchool %>%
-  specify(testscr ~ str)%>%
-  head(.,3)
+  specify(testscr ~ str)
 ```
@@
 ```
-]
-
-- Note nothing happens yet
+
+]
 
@@
 See more [here](https://moderndive.netlify.com/9-hypothesis-testing.html).]
+
 ---
@@
 ```{r, echo=T, eval=F}
-%>% generate(reps = n, type = "permute")
+%>% generate(reps = n,
+             type = "permute")
 ```
@@
 i %>%
-  generate(reps = 1000, type = "permute")
+  generate(reps = 1000,
+           type = "permute")
 ```
@@
 
-- onfire[Two-sided alternative] $H_a: \beta_1 \neq 0$, we double the raw $p$-value
+- .onfire[Two-sided alternative] $H_a: \beta_1 \neq 0$, we double the raw $p$-value
 
@@
 
-.footnote[<sup>1</sup> $k$ is the number of $X$ variables.]
+.footnote[<sup>1</sup> `\\(k\\)` is the number of `\\(X\\)` variables.]
 
@@
   stat_function(fun=dnorm, xlim=c(2,4), geom="area", fill="blue", alpha=0.5)+
-  labs(x = "t",
+  labs(x = expression(paste("Test Statistic t, (deviation from ", H[0], " if ", H[0], " were True)")),
        y = "Probability")+
@@
 
-- .whisper[We have sufficient evidence to reject `\\(H_0\\)` in favor of our alternative hypothesis, that there *is a relationship* between class size and test scores].
-    -
-
---
-
-- .whisper[We are 95% confident that the true relationship is between `\\(-3.22\\)` and `\\(-1.34\\)`.]
+$$\begin{align*}
+H_0: \,& \beta_1=0\\
+H_a: \, & \beta_a \neq 0\\
+\end{align*}$$
+
+- Because the hypothesis test's $p$-value $<$ $\alpha$ (0.05)...
+
+- .onfire[We have sufficient evidence to reject `\\(H_0\\)` in favor of our alternative hypothesis. Our sample suggests that there *is a relationship* between class size and test scores].
+
+--
+
+- Using the confidence intervals:
+
+- .onfire[We are 95% confident that the true marginal effect of class size on test scores is between `\\(-3.22\\)` and `\\(-1.34\\)`.]
 
@@
 
-- .onfire[If a confidence interval contains the $H_0$ value (i.e. $0$, for our test), then *we fail to reject* $H_0$.]
-
-.footnote[<sup>.red[1]</sup> Technically the `\\(t\\)`-value is `\\(\frac{\hat{\beta_1}-\beta_{1,0}}{se(\hat{\beta_1})}\\)`, but our `\\(H_0\\)` is often that `\\(beta_{1,0}=0\\)`, such that the formula reduces to the one above.]
+--
+
+- .whisper[If a confidence interval contains the `\\(H_0\\)` value (i.e. `\\(0\\)`, for our test), then *we fail to reject* `\\(H_0\\)`.]
+
+.footnote[<sup>.red[1]</sup> Since our null hypothesis is that $\beta_{1,0}=0$, the test statistic simplifies to this neat fraction.]
+
+---
+
+# Common Misconceptions about p-values
+
+- All of the following are **FALSE** interpretations of $p$ (and below are reasons why each is wrong)
+
+1. $p$ **is the probability that the alternative hypothesis is false**
+  - We can never *prove* an alternative hypothesis, only tentatively reject a null hypothesis
+
+--
+
+2. $p$ **is the probability that the null hypothesis is true**
+  - We're not *proving* the $H_0$ is false, only saying that it's very unlikely that if $H_0$ were true, we'd obtain a slope as rare as our sample's slope
+
+--
+
+3. $p$ **is the probability that our observed effects were produced purely by random chance**
+  - $p$ is computed under a specific model (think about our null world) that *assumes* $H_0$ is true
+
+--
+
+4. $p$ **tells us how significant our finding is**
+  - $p$ tells us nothing about the *size* or the *real world significance* of any effect deemed "statistically significant"
+  - it only tells us that the slope is statistically significantly different from 0 (if $H_0$ is $\beta_1=0)$
+
+---
+
+# Abusing p-Values I
+
+.pull-left[
+.center[
+![:scale 60%](https://www.dropbox.com/s/9nv7ewnkokuek3c/smbc1623-1.png?raw=1)
+]
+]
+
+--
+
+.pull-right[
+.center[
+![:scale 60%](https://www.dropbox.com/s/5xdh79z3uwg5n3b/smbc1623-2.png?raw=1)
+]
+]
+
+.source[Source: [SMBC](http://www.smbc-comics.com/?id=1623)]
+
+---
+
+# Abusing p-Values II
+
+.pull-left[
+
+.center[
+![](https://www.dropbox.com/s/70gxtqm7tmag454/asalogo.jpg?raw=1)
+]
+]
+
+.pull-right[
+
+>"The widespread use of 'statistical significance' (generally interpreted as $(p \leq 0.05)$ as a license for making a claim of a scientific finding (or implied truth) leads to considerable distortion of the scientific process."
+
+]
+
+.source[Wasserstein, Ronald L.
+and Nicole A.
+Lazar, (2016), ["The ASA's Statement on p-Values: Context, Process, and Purpose](http://www.tandfonline.com/doi/full/10.1080/00031305.2016.1154108)," *The American Statistician* 30(2): 129-133]
+
+---
+
+# p-value Clarification
+
+- Again, .shout[p-value is the probability that, assuming the null hypothesis is true, we obtain (by pure random chance) a test statistic at least as extreme as the one we estimated for our sample]
+
+- A low p-value means either (and we can't distinguish which):
+    1. $H_0$ is true and a highly improbable event has occurred OR
+    2. $H_0$ is false
 

### 2019-10-10 content/assignment/01-exam.Rmd
Add paper projec and midterm contents

---
title: "Midterm Exam"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

*Thursday, October 17*

# Concepts Study Guide

- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/oldexams/midtermconcepts.pdf)

# "Practice" Exam

- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/oldexams/01-old-exam.pdf)

# Formula Sheet (On Exam)

- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/oldexams/midtermformulas.pdf)

# Exam Information

**The exact number and type of questions may change.** Note both the concepts study guide and the

**You may not have anything with you** for the exam (no notes, etc) **except a calculator.** I write questions such that you can get a perfect score without a calculator, and answers should typically be simple whole numbers. However, I understand that if nothing else, calculators are moral support and you can use them. I will provide simple calculators if you need to borrow one, as well as extra paper.

**There will be no use of R on the exam**. However, you *should* be familiar with certain outputs of R that I can ask you questions about (such as regression output from `lm()`).

**There will be few, if any, calculation questions**. Questions are more about *interpretation* of concepts and interpretting numbers in the context of various regressions and research questions.

Students that require accommodations and students that know they will be absent must confirm with me any arrangements at least a week *in advance* of the Exam.

# My Advice

Make sure you do all of the homework problems and learn from the answer keys to the homeworks, as well as the in-class practice problems. While some of the questions should be novel applications, conceptual questions on homeworks will get you in the right headspace to think about answering a question on an exam.

### 2019-10-10 content/assignment/paper-project.Rmd
Add paper projec and midterm contents

---
title: "Empirical Research Paper Project"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

> "How can I know what I think until I see what I say?" --- E. M. Forster on Writing

> "No economist has achieved scientific success as a result of a statistically significant coefficient. Massed observations, clever common sense, elegant theorems, new policies, sagacious economic reasoning, historical perspective, relevant accounting, these have all led to scientific success. Statistical significance has not." --- McCloskey and Ziliak, "The Cult of Statistical Significance" (1996: 112)

- [<i class="fas fa-external-link-square-alt"></i> Writing and Reading Empirical Papers Slides](/slides/paper-slides.html)

# Overview

Your task is to write a research paper on any topic in political economy of your choosing so long as it has an empirical component. The purpose of this paper is threefold: (1) to demonstrate to me that you have mastered the material of this course, (2) to develop your writing, communication, and data analysis skills, and (3) to help you develop, strengthen, and/or modify your own views by grappling with them in writing. As a reminder, this paper constitutes 25% of your final course grade.

I will spend a significant portion of one class discussing more about the writing process, and provide a detailed guide to help you choose a topic, craft arguments, and write a good paper.

I am *NOT* looking for a survey of existing research, a series of block quotes showing what some economist said about X, a regurgitation of my lectures, a list of pros and cons with a last minute conclusion, a book review, and so on. I also do *NOT* want you to compress *everything* about econometrics you learned from this class into a single paper. You should only use those insights that are *relevant* to your topic and your argument (it may only be one or two things!).

I am looking for a paper that attempts to answer a specific research question of interest to economists by using data analysis. You should be able to summarize your paper in one or two sentences -- the specific research question your paper addresses, your method for answering it, and your results. It should be a *reasonably* original paper (it is difficult with limited knowledge, time, and data to offer something truly original), and it should be your own take on the topic. **I expect you to, at the very least, have one multivariate regression used to reach your conclusions. I do not expect or require you to find statistically significant results.**

Please note that, while you are not required to, I highly recommend discussing your ideas with me over email or in person. I will also read any drafts you would like to submit to me early, and provide you with helpful comments, subject to my own availability. I will stop accepting drafts to provide comments by November 30.

# Length, References, & Mechanics

I hesitate to give formal length requirements, because most students will write the bare minimum, and also because different papers have different optimal lengths. What truly matters is that your paper is long enough to say what you need to say, to say it well, and to say it briefly. *Ceterus paribus*, if papers $A$ and $B$ say the same thing, but paper $B$ says it in half the length (without sacrificing key arguments), paper $B$ is a better paper. You will also find that simply by including the necessary components of an empirical paper (plots, tables, etc.), your page count will by necessity increase on its own.

Since you of course still want to know what a good length will be, my *rule of thumb* is that your paper should be about **$\mathbf{10\pm 5}$ pages** (size 12 font, double spaced, 1" margins). This will depend on the topic you have chosen to written on, your data, and your own personal writing style.

I would also like you to use scholarly references, that is, articles from economics journals and cite them properly. I do not have a minimum requirement of the amount of references, but I expect you to have at least **2-5 scholarly references**, depending on your paper topic and thesis. I am not particularly picky about exactly how you format your citations or bibliography, just please be consistent, and do not use footnotes or endnotes (only because they annoy me). I *suggest* the APA author-year-page in-text citation format that is fairly standard in economics journals, i.e.: "The division of labor is limited by the extent of the market," (Smith 1776: 27). Look at my slides or my handouts for a suggested bibliography style. If you use `.bib` files, the default formatting is fine.

# Sources for Inspiration

Here is a list of a few places you might consider to dig up some information on your topic, or to help you find a topic. Just be sure to read and cite the actual sources that these secondary sources cite.

- Major news outlets (e.g. CNN, Wall Street Journal, New York Times, USA Today, Huffington Post, the Guardian) and wherever you may find your news (e.g. reddit, Buzzfeed, etc) both for current events, and also their Opinion/Commentary sections for other op-eds to learn from and/or critique
- Wikipedia -- no seriously, it is the first place I go to learn about a new topic. Just be sure to actually investigate the underlying research and use that for your references!
- Economics Podcasts
    - [Econtalk](http://econtalk.org) -- a fantastic podcast series by Russ Roberts that interviews famous economists, philosophers, businesspeople, and other figures who have an impact on the world of ideas	
    - [Freakonomics (Radio)](http://freakonomics.com/) -- another great podcast by one of the authors of the famed *Freakonomics* books about intermediate-level economic topics, often an in-depth series of interviews on a major issue 
    - [NPR Planet Money](http://npr.org/sections/money) -- another good podcast series on economics topics and current events, much shorter and more 10,000-foot level approach than Econtalk or Freakonomics
- Popular Economists' Blogs/Blogs on Economics:
    - [Marginal Revolution](http://marginalrevolution.com) -- Tyler Cowen & Alex Tabarrok (head and shoulders above the rest!)
    - [Cafe Hayek](http://cafehayek.org) -- Don Boudreaux (libertarian-leaning, mostly just Don on trade and micro-policy)
    - [EconLog](http://econlog.org) -- Bryan Caplan, David Henderson (libertarian-leaning, good analysis)
    - [The Conscience of a Liberal](http://krugman.blogs.nytimes.com/) -- Paul Krugman at New York Times (strong left-wing politics)
    - [The Grumpy Economist](https://johnhcochrane.blogspot.com/) -- John Cochrane (Chicago School approach)
    - [Greg Mankiw's Blog](http://gregmankiw.blogspot.com) -- Greg Mankiw (moderate conservative, New Keynesian approach)
    - [Undercover Economist](http://timharford.com/articles/undercovereconomist/) -- Tim Harford (British, non-political, very easy to understand)
    - [Chris Blattman](https://chrisblattman.com/) -- Chris Blattman (great on economic development, poverty, and conflict in poor countries)
    - [Slate Star Codex](http://slatestarcodex.com) -- "Scott Alexander" (a pseudonym, apparently a Medical Student, but one of the most lucid social science blogs ever written)
    - [Fivethirtyeight](https://fivethirtyeight.com/) -- Nate Silver & co. (a journalist, but a *leader* on using data and statistics in social science and journalism)
    - [Andrew Gelman](http://andrewgelman.com) -- Andrew Gelman (a statistician, but another *leader* on using data and statistics for social science)

# Data Sources

While it is one thing to find a topic to write on, it is an altogether different animal to find data to use to test empirical research questions. You will find out quickly that the constraint to writing an empirical paper is not the set of topics or questions to write on (though that is often a challenge itself!), but the data available to use.

Depending on the topic, you can also collect your own data, and many times you will want to create a custom dataset by simply combining data from different sources.

**General Databases of Datasets**

- [Google Database Search](https://toolbox.google.com/datasetsearch)
- [Kaggle](https://www.kaggle.com/datasets)
- [Harvard Law School: Find a Database](https://hls.harvard.edu/library/research/find-a-database/#hlsnav-databases-for-legal-and-law-related-research)

Here is a list of good data sources depending on the types of topics you might be interested in writing on:

- National and State Accounts Data: [Bureau of Economic Analysis](http://www.bea.doc.gov/)
- Labor Market and Price Data: [Bureau of Labor Statistics](http://www.bls.gov/)
- Macroeconomic Data: [Federal Reserve Economic Data (FRED)](https://fred.stlouisfed.org/), [World Development Indicators (World Bank)](https://datacatalog.worldbank.org/dataset/world-development-indicators), [Penn World Table](https://www.rug.nl/ggdc/productivity/pwt/)
- International Data: [NationMaster.com](http://www.nationmaster.com/), [Doing Business](http://rru.worldbank.org/DoingBusiness/), [CIESIN](ttp://www.ciesin.org/data.html)
- Census Data: [U.S. Census Bureau](http://www.census.gov/)
- Sports Data: [Spotrac](http://spotrac.com), [Rodney Fort's Sports Data](https://sites.google.com/site/rodswebpages/codes)
- Data Clearing House: [Stat USA](http://www.stat-usa.gov/), [Fedstats](http://www.fedstats.gov/), [Statistical Abstract of the United States](http://www.census.gov/prod/www/statistical-abstract-us.html), [Resources for Economists](http://rfe.org) 
- Political and Social Data: [ICPSR](http://www.icpsr.umich.edu/), [Federal Election Commission](http://fec.gov), [Poole and Rosenthal Roll Call Data (Voting ideology)](http://voteview.uh.edu/default_nomdata.htm), [Archigos Data on Political Leaders](https://www.rochester.edu/college/faculty/hgoemans/data.htm), [Library of Congress: Thomas (Legislation)](http://thomas.loc.gov/), [Iowa Electronic Markets (Prediction Markets)](http://www.biz.uiowa.edu/iem/)
- War and Violence Data: [Correlates of War](http://www.correlatesofwar.org/data-sets)
- State Level Data: [Correlates of State Policy](http://ippsr.msu.edu/public-policy/correlates-state-policy)
- Health Data: [Centers for Disease Control](http://www.cdc.gov/scientific.htm), [CDC Wonder System](http://wonder.cdc.gov/DataSets.shtml)
- Crime Data: [Bureau of Justice Statistics](http://www.ojp.usdoj.gov/bjs/)
- Education Data: [National Center for Education Statistics](http://www.nces.ed.gov/)
- Environmental Data: [EPA](http://www.epa.gov/ceisweb1/ceishome/)
- Religion Data: [American Religion Data Archiva (ARDA)](http://www.arda.tm/)
- Financial Data: [Financial Data Finder](http://www.cob.ohio-state.edu/cgi-bin/DB_Search/db_search.cgi?setup_file=finance.setup.cgi){Financial Data Finder}
- Philanthropy Data: [The Urban Institute](http://nccs.urban.org/)

Most government agencies and non-profit organizations have publicly available data on the issue you are concerned with, and will be your first place to look.

# Grading Rubric and Deadlines

While it may be possible for many papers in your college career, is not a paper you can write the last minute and do well on. To ensure that you do not get too far behind, I have split the assignment into stages that are due at different intervals over the semester. Note that your topic can and may change depending on what you are able to find and work with. **The hardest part is finding data that allows you to test a research question**. It is primarily for this reason that writing an empirical paper on what you want is very difficult.

| Assignment | Points | Due Date | Description |
|------------|--------|----------|-------------|
| Abstract | 5 | October 29 | Short summary of your ideas |
| Data Description | 10 | November 14 | Description of data sources, and some summary statistics |
| Literature Review | 10 | December 5 | 1-3 paragraphs on 2-3 scholarly sources |
| Presentation | 5 | December 10/12 | Short presentation of your project so far |
| Final Paper Due | 70 | December 17 | Email to me paper, data, code |

- **Abstract**: write a short paragraph (3-6 sentences) summarizing: what rough topic you want to look at, a specific research question that you think you can get data to test, where you think you might be able to get some data. It's okay that much of this is speculative and you might change your mind or do very different things later!

- **Data Description**: describe what data you managed to find (a few paragraphs): where is/are the dataset(s) from? What variables are included, and how are they measured (e.g. what units, categories, etc)? Give us some summary statistics of the data (a table would be nice, some scatterplots and histograms would be nice), are there any interesting patterns?

- **Literature Review**: find 2-3 scholarly sources (ideally scholarly journal articles) that discuss your research question (or related topics), especially if they have empirical findings. Explain what these sources found, how they found it, and how your paper relates to this literature. Don't worry about being original!

- **Presentation**: a 5-10 minute presentation of what you have so far. Different people will be at different stages, since your paper is not due yet! Use this as an opportunity to get feedback from me and your classmates, that might help you finish your project. Using some slides to show tables, models, and plots is highly recommended.

- **Final Paper Due**: *email* the full paper to me (see below for what should be in the email).

The remaining 70% for the final product are broken down as follows:

| Category | Points |
|----------|--------|
| Persuasiveness | 10 |
| Clarity | 10 |
| Econometric Validity | 20 |
| Economic Soundness | 20 |
| Organization | 5 |
| References | 5 |
| TOTAL | 70 |

- **Persuasiveness**: How persuasive is your argument? Would a reasonably educated college-level reader who is familiar with economics and statistics but not necessarily this course find themselves understanding your argument and agreeing with you? [Write for an audience wider than just members of this class. Therefore, don't use terms, sources, or "inside jokes" that only other students in this class (and no one else) would understand.] Remember, your goal is not to convince *me* (though you may), your goal is to convince *any educated reader*, and I grade you the probability that this is likely. You are the lawyer, I am the judge, and your audience is the jury. 
- **Clarity** How clear is your paper? Is it clear what your research question is, how you answer it, and what your results are? Can you summarize these in a sentence or two? Are there confusing passages, excessive jargon or passive voice, or irrelevant arguments and examples?
- **Econometric Validity**: Is your econometric model sensible and plausible? Do you adequately address the assumptions and limitations of your model? Do you use appropriate data? Do you adequately describe your data, identify patterns, aberrations, and clearly generate testable hypotheses from your data? Would someone else, given your data, be able to replicate your findings? 
- **Economic Soundness**: Do you place your empirical question in a broader context of applying economic principles? Do you describe relevant economic policies, institutions, relationships, and/or other relevant economic principles to the questions you are asking? Are your theories and hypotheses about relationships between data plausible and intuitive? Do you makes sense of your hypotheses and the results of your analysis and connect them to sound economic principles? Is your question, strategy, and/or results of interest to economists? **You will lose up to 20 points for papers that have no economic content.**
- **Organization:** Is your paper organized? Have you presented your separate arguments/examples in a logical order? Is it clear when you are moving on from one section to another? Is it clear when and where you are summarizing and concluding? Is your paper presented according to professional norms (e.g. summary statistics tables, regression tables, etc)? 
- **References:** Does your paper use multiple scholarly references? Does it properly cite them in the text for main ideas borrowed and for direct quotations (if applicable)? Are they consistently listed at the end in a references section? 
- **Style:** Is your paper interesting and easy to read? Does it engage the reader? Is it written in active voice? This is somewhat subjective, and hence, the smallest portion of your grade.

## Emailing Your Final Version to Me

When you send your final email (by Thursday December 17), it should contain the following files:

1. **Your final paper as a `.pdf`.** It should include an abstract and bibliography and all tables and figures contained within it. 
2. **The (commented!) code used for your data analysis** (i.e. loading data, making tables, making plots, running regressions). These can be either `.R` files: one or multiple (one-per-task) are equally fine OR a `.Rmd` file. I want to know *how* you reached the results you got! **Reproducibility is the goal!**
3. **Your data used**, in whatever original format you found it (e.g. `.csv`, `.xlsx`, `.dta`)

Again, you are not obligated to use `R Markdown` to write your paper. Microsoft Word is fine.

### 2019-10-10 static/slides/paper-slides.Rmd
Add paper projec and midterm contents

---
title: "Writing and Reading Empirical Papers"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))
```

# Finding a Research Question

- A paper needs a **specific research question** that you will ask and provide evidence towards a *clear*, *quantifiable* answer

- Good research questions are:

1. **A claim** about something
		- Capital punishment is the most efficient deterrent for violent crimes.
		- Women are paid, on average, $0.33 less than men performing the same work.

2. **As specific as possible**, given the length constraints
    - Do candidates that spend more money than their opponents tend to win Congressional races?

3. **Testable**, with data that can provide *some* evidence one way or another
    - One study will never be "the" *definitive proof* of something, only evidence suggesting one conclusion over others

---

# Structure of an Empirical Paper

1. Introduction
2. Literature Review
3. Theory/Model 
4. Data Description
5. Empirical Model
6. Results/Implications
7. Bibliography

---

# Introduction

- Get to your research question ASAP!  Make it the first sentence even.
- Hook your reader
    - Who cares?  Why is this important?  Why is this relevant? How does this affect people?
		- Statistics and background information can often help

--

.content-box-green[
.green[**Example**]: As a student writing an empirical research paper, does writing a longer paper attain a higher grade?]

---

# Introduction II

- State your research question clearly and quickly

--

- Do NOT write a "blog post" about how you became interested in the question, or all the work (and dead-ends) that led you to your answer
    - **Nobody cares about the labor pains, they just want to see the baby!**

--

- Outline:
    - Why your question matters
    - How do you answer the question in this paper?
    - What approach do you use? What data?
    - What are your most important results?

---

# Introduction III

.content-box-green[
.green[**Example**]: I estimate the relationship between paper length and grades by using a simple OLS regression using sample data collected from previous classes. I find that there is a strong positive effect, that students who write longer papers earn higher grades. On average, for every additional page written, grades improve by 2 points.
]

---

# Introduction IV

- Most people do not write enough in their introductions

- Consider the incentives of a (skimming) reader pressed for time
    - If someone only skims your intro, what do you want them to know??

- My rough suggestion: make your introduction about 15-20% of your paper:

| Paper Length | Intro Length |
|--------------|--------------|
| 5 pages | 1-1.5 pages |
| 10 pages | 2-2.5 pages |
| 30 pages | 5 pages |

---

# Literature Review

- **Literature Review** can be summarized into the introduction or given its' own section (debatable)

- No work is totally original
    - What have other relevant researchers written and discovered about your topic?
    - What data and models did they use? What did they find? 
    - How does your paper connect (new data? new model? different angle?) and stand apart from what's been done?

---

# Theory

- These are economics papers, so you must describe some **economic theory** behind the question you are asking and answering
- Most scholarly papers have a formal economic model, which then generates predictions that they test for with data
- You do not need a model, but you do need to discuss economic principles or concepts that are relevant
    - Often there may be multiple theories that might conflict, or our expectations might not be clear (these are the best papers!)
    - There may be a significant tradeoff between competing goals, values, or expectations

---

# Theory II

.content-box-green[
.green[**Example**:] Students that write longer papers likely place higher value on their work and dedicate more resources towards improving its quality, resulting in higher grades.

However, some students hope or believe that longer papers earn higher grades, and will simply put extra low quality filler in their paper to inflate the length. These students likely earn lower grades as a result.

]

---

# Data I

- Describe your data sources
    - Who collected or compiled the data and how? 
        - e.g. government agencies, businesses, nonprofits, social surveys, etc.
    - If you collected your own data, what was your procedure?

---

# Data II

- Describe the data itself
    - What are your variables? What--specifically, and in English--does each measure?
    - How many observations do you have?
    - If you transformed your variables--how and why?
        - e.g. recoded into categories or dummies
        - e.g. took logs or rescaled units

---

# Data III

- Rule I: Show your data! Show us basic summary statistics and any patterns
    - your judgment: we don't want or need to see *everything* -- what do you think is interesting or important?
    - Plots $>$ Tables $>$ Nothing
- Good ideas to *always* have:
    1. A table of variables and their description
    2. A table of summary statistics of variables
    3. Plots of most important variables and interactions (histograms, boxplots, scatterplots, etc)

---

# Data: Variables

.pull-left[

| Variable | Description |
|----------|-------------|
| Grade    | Grade on paper assignment (0-100) |
| Pages    | Number of pages written |
| Final | Final grade for student in class |
| Gender | Gender of student |
| Class | Class in which paper was assigned |
| School | School of class taught |
| Year | Year of class |
| Time | Time of day class met |

- All variables were collected from assignments that I gave over the 2013-2019 period at three colleges I taught at

]

.pull-right[
```{r}
library(tidyverse)
papers<-read_csv("../data/paperlengthsregcsv.csv")
source("../files/summaries.R")
```

```{r, fig.retina=3, fig.align="center", out.width="60%"}
papers<-papers %>%
  mutate(Sex=as.factor(Sex),
         Sex=recode(Sex, `F` = "Female", M = "Male")) %>%
  mutate(Female = ifelse(Sex=="Female",1,0),
         Morning = ifelse(Time=="Morning",1,0),
         Hood=ifelse(School=="Hood",1,0),
         Econometrics=ifelse(Class=="Econometrics",1,0))

papers_cor <- papers %>%
  select_if(is.numeric) %>%
  cor()

library(corrplot)

corrplot(papers_cor, type="upper", 
         method = "number", # number for showing correlation coefficient
         order="original")
```

]
---

# Data: Summary Statistics of Quantitative Variables

```{r}
summary_table(papers, Pages, Grade, Final, Year, Female, Morning, Hood, Econometrics) %>%
  knitr::kable(., format="html")
```

---

# Data: Counts of Categorical Variables I

.pull-left[

.pull-left[
```{r}
papers %>%
  count(Year) %>%
  knitr::kable(., format="html")
```

]
.pull-right[
```{r}
papers %>%
  count(Sex) %>%
  knitr::kable(., format="html")
```

]
]

.pull-right[

.pull-left[
```{r}
papers %>%
  count(Morning) %>%
  knitr::kable(., format="html")
```

]

.pull-right[
```{r}
papers %>%
  count(Class) %>%
  knitr::kable(., format="html")
```
]

]

---

# Data: Counts of Categorical Variables II

```{r}
papers %>%
  count(School) %>%
  knitr::kable(., format="html")
```

---

# Data: Histogram I

```{r, fig.retina=3, fig.align="center", out.width="60%"}
pages_hist<-ggplot(data = papers)+
  aes(x = Pages)+
  geom_histogram(color="white", fill="indianred",breaks=seq(0,24,2))+
  scale_x_continuous(breaks=seq(0,24,2))+
  scale_y_continuous(breaks=seq(0,50,10))+
  labs(x = "Number of Pages Written",
       y = "Number of Papers")+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
pages_hist
```

---

# Data: Histogram II

```{r, fig.retina=3, fig.align="center", out.width="60%"}
pages_hist+facet_wrap(~Sex)
```

---

# Data: Scatterplot I

```{r, fig.retina=3, fig.align="center", out.width="60%"}

papers<-papers %>%
  mutate(Metrics = ifelse(Econometrics==1, "Econometrics", "Other"))

ggplot(data = papers)+
  aes(x = Pages)+
  geom_histogram(color="white", fill="indianred",breaks=seq(0,24,2))+
  scale_x_continuous(breaks=seq(0,24,2))+
  scale_y_continuous(breaks=seq(0,50,10))+
  labs(x = "Number of Pages Written",
       y = "Number of Papers")+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)+
  facet_wrap(~Metrics)
```

---

# Data: Scatterplot II

```{r, fig.retina=3, fig.align="center", out.width="60%"}

ggplot(data = papers)+
  aes(x = Pages,
      y = Grade)+
  geom_jitter(color="blue")+
  geom_smooth(method="lm", color="red")+
  scale_x_continuous(breaks=seq(0,24,2),
                     limits=c(0,24),
                     expand=c(0,0))+
  scale_y_continuous(breaks=seq(0,100,10),
                     limits=c(0,120),
                     expand=c(0,0))+
  labs(x = "Number of Pages Written",
       y = "Paper Grade")+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

---

# Data: Scatterplot III

```{r, fig.retina=3, fig.align="center", out.width="60%"}

ggplot(data = subset(papers, Grade>0))+
  aes(x = Pages,
      y = Grade)+
  geom_jitter(color="blue")+
  geom_smooth(method="lm", color="red")+
  scale_x_continuous(breaks=seq(0,24,2),
                     limits=c(0,24),
                     expand=c(0,0))+
  scale_y_continuous(breaks=seq(0,100,10),
                     limits=c(0,120),
                     expand=c(0,0))+
  labs(x = "Number of Pages Written",
       y = "Paper Grade")+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

---

# Data: Scatterplot IV

```{r, fig.retina=3, fig.align="center", out.width="60%"}

ggplot(data = subset(papers, Grade>0))+
  aes(x = Pages,
      y = Grade)+
  geom_jitter(color="blue")+
  geom_smooth(method="lm", color="red")+
  scale_x_continuous(breaks=seq(0,24,2),
                     limits=c(0,24),
                     expand=c(0,0))+
  scale_y_continuous(breaks=seq(0,100,10),
                     limits=c(0,120),
                     expand=c(0,0))+
  labs(x = "Number of Pages Written",
       y = "Paper Grade")+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)+
  facet_grid(~Sex)
```

---

# Data: Scatterplot V

```{r, fig.retina=3, fig.align="center", out.width="60%"}

ggplot(data = subset(papers, Grade>0))+
  aes(x = Pages,
      y = Grade)+
  geom_jitter(color="blue")+
  geom_smooth(method="lm", color="red")+
  scale_x_continuous(breaks=seq(0,24,2),
                     limits=c(0,24),
                     expand=c(0,0))+
  scale_y_continuous(breaks=seq(0,100,10),
                     limits=c(0,120),
                     expand=c(0,0))+
  labs(x = "Number of Pages Written",
       y = "Paper Grade")+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)+
  facet_grid(~Metrics)
```

---

# Empirical Model I

- Describe your empirical model and your **identification strategy** 
    - for most of you, just OLS and trying to include as many controls to remove omitted variable bias

- Why did you pick certain variables?

- How do you battle endogeneity?

- Hypothesize your expected size and magnitude of key variables
    - Give some **economic intution** behind what we would expect!

---

# Empirical Model II

.content-box-green[
.green[**Example**]:
$$Grade_i=Length_i+Final_i+Female_i+Morning_i+Hood_i+Metrics_i+u_i$$
]

- $Length$ is the most important variable we care about

- $Length$ probably endogenous, correlated with other Grade-determining factors:
    - Why we included $Final$, $Female$, $Morning$, $Hood$, and $Metrics$

- We probably expect $Length$ to be positive and small

---

# Empirical Model III

- Describe the **limitations of your model**
    - Every paper, even Nobel prize-winning ones, have limitations and problems!
    - Limited and/or poor quality data
    - Endogeneity, simultaneous causation, omitted variable bias

.content-box-green[
.green[**Example**]:
The model likely suffers from endogeneity, as how many pages a student writes is likely to be positively correlated with personal attributes like dilligence, conscientiousness, and intelligence, which themselves are likely positively correlated with the grade of the paper. Thus, we have likely \emph{over}stated the effect of page length on paper grades.Furthermore, we are unable to measure other variables that make page length endogenous, such as the topic that was chosen. Some topics lend themselves to shorter or longer papers and may have better or worse data that make it easier or difficult to run a clean empirical test.
]

---

# Empirical Model IV

- Are your results **robust** across different model specifications?
    - Do the size(s) of the marginal effect(s) you care about change or reverse direction? Become/lose significance?
- At minimum, you must run several models, including a multivariate regression
    - Run several variations of your model with and without controls (e.g. just $Y$ and $X$, $Y$ and $X_1$ and $X_2$, etc.)
    - Check for nonlinearities: polynomials, logs, etc.

---

# Results I

- Print a table(s) of your regression(s) results (`huxtable` is ideal)

- Interpret your data
    - What does a marginal (1 unit) change in $X$ mean for $Y$, a 1% change, etc? 
    - Is each coefficient statistically significant (at 10, 5, or 1% levels)?

---

```{r}
papers0<-papers %>%
  filter(Grade>0)

hood<-papers %>%
  filter(School=="Hood")

metrics<-papers %>%
  filter(Class=="Econometrics")

basicreg<-lm(Grade~Pages, data=papers)
no0reg<-lm(Grade~Pages, data=papers0)
basicmetricsreg<-lm(Grade~Pages, data=metrics)
finalreg<-lm(Grade~Pages+Final, data=papers0)
sexreg<-lm(Grade~Pages+Final+Female, data=papers0)
timereg<-lm(Grade~Pages+Final+Female+Morning, data=papers0)
hoodreg<-lm(Grade~Pages+Final+Female+Morning, data=hood)
metricsreg<-lm(Grade~Pages+Final+Female+Morning, data=metrics)

```

```{r}
library(huxtable)
huxreg("Baseline" = basicreg,
       "No Os" = no0reg,
       "Econometrics Only" = basicmetricsreg,
       "With Final Grades" = finalreg,
       "Controls" = timereg,
       "Hood Only" = hoodreg,
       "Econometrics Only" = metricsreg,
       coefs = c("Constant" = "(Intercept)",
                 "Length" = "Pages",
                 "Course Grade" = "Final",
                 "Female" = "Female",
                 "Morning" = "Morning"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```

---

# Results: Interpretation!

- Are your estimates **economically significant**?

- How big is "big"?

> No economist has achieved scientific success as a result of a statistically significant coefficient. Massed observations, clever common sense, elegant theorems, new policies, sagacious economic reasoning, historical perspective, relevant accounting, these have all led to scientific success. Statistical significance has not.'' 	McCloskey & Ziliak (1996: 112)

---

# Results: Interpretation!

.center[
![:scale 90%](https://www.dropbox.com/s/qtd2uwvr52jvkcu/drwhocontext.jpg?raw=1)
]

---

# Results: Interpretation!

.content-box-green[
.green[**Example**:] I find that for every additional page written, we can expect a paper's grade to increase by about a point or less, after controlling for other factorssuch as Final grade (proxying as a measure of overall diligence and intelligence), sex, and time of day. In the most relevant sample, econometrics students, the effect is even smaller, only about a third of a point increase for every additional page written. This small effect is statistically significant at the 10% level only.

However, we should not make much of these results due to the likely endogeneity of Pages due to unobserved factors such as topic and quality of writing, which clearly would matter much both for length and for grade. *It would be poor advice to recommend students simply to write long papers to earn a higher grade.*
]

---

# Results: Implications

- Describe several *implications* of your paper
    - Policy implications
    - Proposals for new research
    - Effects on current understanding
    - What else should we try to found out to answer the question better?

---

# Don't Get Discouraged

.pull-left[
.center[
![:scale 90%](https://www.dropbox.com/s/akegbq42i02gl8y/sciencerage.png?raw=1)
]
]

--

.pull-right[
.center[
![:scale 50%](https://www.dropbox.com/s/y3fvprtxyuvki24/sciencerage1.png?raw=1)
]
]

---

# Don't Get Discouraged

.left-column[
.center[
![:scale 80%](https://www.dropbox.com/s/udfdftpvalagsyj/einstein.png?raw=1)

Albert Enstein

(1870-1924)

]
]

.right-column[

> "If we knew what it was we were looking for, we wouldn't call it research, would we?"

]

---

# Deadlines and Reminders (From the Assignment Page)

| Assignment | Points | Due Date | Description |
|------------|--------|----------|-------------|
| Abstract | 5 | October 29 | Short summary of your ideas |
| Data Description | 10 | November 14 | Description of data sources, and some summary statistics |
| Literature Review | 10 | December 5 | 1-3 paragraphs on 2-3 scholarly sources |
| Presentation | 5 | December 10/12 | Short presentation of your project so far |
| Final Paper Due | 70 | December 17 | Email to me paper, data, code |

- note for each stage (except the Final Paper), it's **more than okay that your final topics, data, etc will change!**
- for each assignment, I will take 1 point off for every 24 hours it is late

---

# Grading of Final Paper (From the Assignment Page)

| Category | Points |
|----------|--------|
| Persuasiveness | 10 |
| Clarity | 10 |
| Econometric Validity | 20 |
| Economic Soundness | 20 |
| Organization | 5 |
| References | 5 |
| TOTAL | 70 |

---

# Submitting your Final Paper

When you send your final email (by Thursday December 20), it should contain the following files:

1. **Your final paper as a `.pdf`.** It should include an abstract and bibliography and all tables and figures.

2. **The (commented!) code used for your data analysis** (i.e. loading data, making tables, making plots, running regressions)
  - **either `.R` files** OR a **`.Rmd` file.** I want to know *how* you reached the results you got! **Reproducibility is the goal!**

3. **Your data used**, in whatever original format you found it (e.g. `.csv`, `.xlsx`, `.dta`)

### 2019-10-10 static/slides/midterm-review.Rmd
Add midterm review questions, fix paper project slides

---
title: "Midterm Review Questions"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))
```
# Question 1

What does **endogenous** mean, in words? What about statistically?

---

# Question 2

If a regression is biased (from endogeneity), what can we learn about the bias?

---

# Question 3

What does heteroskedasticity mean? Does heteroskedasticity **bias** $\hat{\beta_1}$?

---

# Question 4

Is this data likely heteroskedastic or homoskedastic?

```{r}
library(tidyverse)
het_data<-tibble(x = runif(500,0,2),
                 y = rnorm(500,x,x))

ggplot(data = het_data)+
  aes(x = x,
      y = y)+
  geom_point(color="blue")+
  geom_smooth(method="lm", color="red")+
  scale_x_continuous(breaks=seq(0,10,1))+
    labs(x = "X",
       y = "Y")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=20)
```

---

# Question 5

- What three things impact the variation of $\hat{\beta_1}$? How?

---

# Question 6

- What are the four assumptions we make about the error term?
- Which is most important?

---

# Question 7

$$Wages_i=\hat{\beta_0}+\hat{\beta_1}Education+u_i$$

- What is in $u_i$?

- Is $\hat{\beta_1}$ likely biased?

---

# Question 8

- What does $R^2$ measure? What does it mean? How do we calculate it?

---

# Question 9

- What does $\sigma_u$ (SER) measure? What does it mean?

---

# Question 10

- Interpret all of these numbers (except Adjusted R-squared and the last line):

```{r}
summary(lm(y~x, data = het_data))
```

---

# Question 11

- Interpret all of these numbers (except Adjusted R-squared and the last line):

```{r}
reg<-lm(y~x, data = het_data)
library(huxtable)
huxreg(reg,        statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"))
```

---

# Question 12

- Suppose $Y$ is normally distributed with a mean of 10 and a standard error of 5. What is the probability that $Y$ is between 5 and 15?

---

# Question 13

- Explain what a $Z$-score means.

---

# Question 14

- Explain what a $p$-value means. :)

---

# Question 15

- We run the following hypothesis test at $\alpha=0.05$:

$$\begin{align*}
H_0: \, & \beta_1=0\\
H_1: \, & \beta_1 \neq 0 \\
\end{align*}$$

- Is this test one-sided or two-sided?

- We find the $p$-value is 0.02. What is our conclusion? Be specific and precise in your wording!

### 2019-10-10 static/slides/paper-slides.Rmd
Add midterm review questions, fix paper project slides

@@
 1. Introduction
+
 2. Literature Review
+
 3. Theory/Model
+
 4. Data Description
+
 5. Empirical Model
+
 6. Results/Implications
+
 7. Bibliography
@@
 - Get to your research question ASAP!  Make it the first sentence even.
+
 - Hook your reader
@@
 I find that there is a strong positive effect, that students who write longer papers earn higher grades.
-On average, for every additional page written, grades improve by 2 points.
+On average, for every additional page written, grades improve by less than a point.
+These results are robust to a number of different model specifications and controls.
 ]
@@
 - These are economics papers, so you must describe some **economic theory** behind the question you are asking and answering
+
 - Most scholarly papers have a formal economic model, which then generates predictions that they test for with data
+
 - You do not need a model, but you do need to discuss economic principles or concepts that are relevant
@@
 
-- Rule I: Show your data! Show us basic summary statistics and any patterns
-    - your judgment: we don't want or need to see *everything* -- what do you think is interesting or important?
-    - Plots $>$ Tables $>$ Nothing
+- .shout[Show your data!] Show us basic summary statistics and any patterns
+    - Use your judgment: .onfire[we don't want or need to see *everything*]
+    - What do you think is *interesting* or *important*?
+    - Plots $>$ Tables $>$ Words $>$ Nothing
+
 - Good ideas to *always* have:
-    1. A table of variables and their description
-    2. A table of summary statistics of variables
+    1. A table(s) of variables and their description
+    2. A table(s) of summary statistics of variables
+    3. A table of correlations of key variables
     3. Plots of most important variables and interactions (histograms, boxplots, scatterplots, etc)
@@
 
-```{r, fig.retina=3, fig.align="center", out.width="60%"}
+```{r, fig.retina=3, fig.align="center", out.width="100%"}
 papers<-papers %>%
@@
 papers %>%
-  count(Morning) %>%
+  count(Time) %>%
   knitr::kable(., format="html")
@@
 
----
-
-# Data: Counts of Categorical Variables II
-
 ```{r}
@@
 
-```{r, fig.retina=3, fig.align="center", out.width="60%"}
+```{r, fig.retina=3, fig.align="center", out.width="50%"}
 pages_hist<-ggplot(data = papers)+
@@
 
-```{r, fig.retina=3, fig.align="center", out.width="60%"}
+```{r, fig.retina=3, fig.align="center", out.width="50%"}
 pages_hist+facet_wrap(~Sex)
@@
 
-```{r, fig.retina=3, fig.align="center", out.width="60%"}
+```{r, fig.retina=3, fig.align="center", out.width="50%"}
 
@@
     - Do the size(s) of the marginal effect(s) you care about change or reverse direction? Become/lose significance?
+
 - At minimum, you must run several models, including a multivariate regression
@@
 .center[
-![:scale 90%](https://www.dropbox.com/s/qtd2uwvr52jvkcu/drwhocontext.jpg?raw=1)
+![:scale 45%](https://www.dropbox.com/s/qtd2uwvr52jvkcu/drwhocontext.jpg?raw=1)
 ]

### 2019-10-10 static/slides/midterm-review.Rmd
Tweak review slides

@@
 
-- What are the four assumptions we make about the error term?
-- Which is most important?
+What are the four assumptions we make about the error term?
+
+Which is most important?
 
@@
 
-- What is in $u_i$?
+What is in $u_i$?
 
-- Is $\hat{\beta_1}$ likely biased?
+Is $\hat{\beta_1}$ likely biased?
 
@@
 
-- What does $R^2$ measure? What does it mean? How do we calculate it?
+What does $R^2$ measure? What does it mean? How do we calculate it?
 
@@
 
-- What does $\sigma_u$ (SER) measure? What does it mean?
+What does $\sigma_u$ (SER) measure? What does it mean?
 
@@
 
-- Interpret all of these numbers (except Adjusted R-squared and the last line):
+Interpret all of these numbers (except Adjusted R-squared and the last line):
 
@@
 
-- Interpret all of these numbers (except Adjusted R-squared and the last line):
+Interpret all of these numbers:
 
@@
 
-- Suppose $Y$ is normally distributed with a mean of 10 and a standard error of 5. What is the probability that $Y$ is between 5 and 15?
+Suppose $Y$ is normally distributed with a mean of 10 and a standard error of 5. What is the probability that $Y$ is between 5 and 15?
 
@@
 
-- Explain what a $Z$-score means.
+Explain what a $Z$-score means.
 
@@
 
-- Explain what a $p$-value means. :)
+Explain what a $p$-value means. :)
 
@@
 
-- We run the following hypothesis test at $\alpha=0.05$:
+We run the following hypothesis test at $\alpha=0.05$:
 
@@
 
-- Is this test one-sided or two-sided?
+Is this test one-sided or two-sided?
 
-- We find the $p$-value is 0.02. What is our conclusion? Be specific and precise in your wording!
+We find the $p$-value is 0.02. What is our conclusion? Be specific and precise in your wording!

### 2019-10-12 content/assignment/03-problem-set.Rmd
Problem set 3 question 11 part L, refers to regression in part C, now its part D
Add problem set 3 answers

@@
 - [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/03-problem-set.Rmd)
+
+**ANSWERS:**
+
+- [<i class="fas fa-file-pdf"></i> `html`](/homeworks/03-problem-set-answers.html)
+- [<i class="fab fa-markdown"></i> `Markdown`](/homeworks/03-problem-set-answers.Rmd)
+- [<i class="fas fa-registered"></i> `R Script`](/homeworks/03-problem-set-answers.R)
 
@@
 Now we'll try out the `infer` package to understand the $p$-value and confidence interval for our observed slope in our regression model.
-Save the (value of) our sample $\hat{\beta_1}$ from your regression in Part C as an object.
+Save the (value of) our sample $\hat{\beta_1}$ from your regression in Part D as an object.
 Then, install and load the `infer` package, and then calculate the slope^[`calculate(stat = "slope")`] under the null hypothesis that there is no connection between attendance and runs.^[`hypothesize(null = "independence")`] for 1000 additional simulated samples^[`generate(reps = 1000, type = "permute")`], and save this as an object (it's a `tibble`).

### 2019-10-12 static/homeworks/03-problem-set-answers.Rmd
Add problem set 3 answers

--- 2019-09-20 static/homeworks/02-problem-set-pdf.Rmd
@@
 ---
-title: "Problem Set 2"
+title: "Problem Set 3 (Answers)"
 author: "Ryan Safner"
 date: "ECON 480 - Fall 2019"
-output: pdf_document
+output:
+  html_document:
+    df_print: paged
+    #theme: simplex
+    toc: true 
+    toc_depth: 3
+    toc_float: true
+    code_folding: show
+    highlight: tango
 ---
 
-*Due by Thursday, September 26, 2019*
+```{r message=FALSE, warning=FALSE, include=FALSE}
+library(pander)
+source(file.path(here::here(), "lib", "graphics.R"))
+```
 
-# Instructions
-
-For this problem set, you may submit handwritten answers on a plain sheet of paper, or download and type/handwrite on the PDF.
-
-Alternatively, you may download the `.Rmd` file, do the homework in markdown, and email to me a single `knit`ted `html` or `pdf` file (and be sure that it shows all of your code).
-
-You may work together (and I highly encourage that) but you must turn in your own answers.
-I grade homeworks 70% for completion, and for the remaining 30%, pick one question to grade for accuracy - so it is best that you try every problem, even if you are unsure how to complete it accurately.
+Answers may be longer than I would deem sufficient on an exam.
+Some might vary slightly based on points of interest, examples, or personal experience.
+These suggested answers are designed to give you both the answer and a short explanation of *why* it is the answer.
 
@@
 
-1.
+## Question 1
+**In your own words, describe what exogeneity and endogeneity mean, and how they are related to bias. What can we learn about the bias if we know $X$ is endogenous?**
 
-a. In your own words, explain the difference between endogeneity and exogeneity.^[Hint: you may want to look back at class 1.2 for this question.]
-b. In your own words, explain how conducting a randomized controlled trial helps to identify the causal connection between two variables.
+The OLS estimators $\hat{\beta_0}$ and $\hat{\beta_1}$ are unbiased estimates of the true population parameters $\beta_0$ and $\beta_1$ if and only if $X$ is *exogenous.* That is to say, if $cor(X,u)=0$ (i.e.
+there is no correlation between $X$ and any unobserved variable that affects $Y$), then $E[\hat{\beta_1}]=\beta_1$.
 
-\clearpage
+If $X$ *is* correlated with the error term, then $X$ is *endogenous.* The true expected value of the OLS estimator is 
+$$E[\hat{\beta_1}]=\beta_1+cor(X,u)\frac{\sigma_u}{\sigma_X}$$
 
-2.
+The bias is $\big(E[\hat{\beta_1}]-\beta_1\big)$, i.e.
+the difference between average estimated sample slope and the `true' population slope, so we can determine first the *size* of the bias based on how large $cor(X,u)$ is.
+The stronger the correlation, the larger the bias.
 
-a. In your own words, explain what (sample) standard deviation *means*.
-b. In your own words, explain how (sample) standard deviation *is calculated.* You may also write the formula, but it is not necessary.
+Second, we can determine the *direction* of the bias depending on the sign of $cor(X,u)$.
 
-\clearpage
+- If $X$ and $u$ are positively correlated (move in the same direction), we know that we have *overstated* the true effect of $\Delta X$ on $\Delta Y$, since a change in $Y$ is picking up both a change in $X$ and a further change (in the same direction as $X$) in the unobserved $u$.
 
-# Problems
+- If the correlation is negative (move in opposite directions), we know that we have *understated* the true effect of $\Delta X$ on $\Delta Y$, since a change in $Y$ is picking up both a change in $X$ that is dampened by a change in the opposite direction of $u$.
 
-For the remaining questions, you may use `R` to *verify*, but please calculate all sample statistics by hand and show all work.
+## Question 2
+**In your own words, describe what $R^2$ means. How do we calculate it, what does it tell us, and how do we interpret it?**
 
-3. Suppose you have a very small class of four students that all take a quiz. Their scores are reported as follows:
+The $R^2$ is a measure of how well the OLS regression line "fits" our observed data points.
+It is the proportion of the total variation in $Y$ (TSS) that is *explained* by the variation from our model (ESS):
 
-$$\{83, 92, 72, 81\}$$
+$$\begin{align*}
+	R^2&=\frac{ESS}{TSS}= \frac{\sum (\hat{Y_i}-\bar{Y})^2}{\sum(Y_i-\bar{Y})^2}
+\end{align*}$$
 
-a. Calculate the median.
-b. Calculate the sample mean, $\bar{x}$.
-c. Calculate the sample standard deviation, $s$.
-d.
-Make or sketch a rough histogram of this data, with the size of each bin being 10 (i.e.
-70's, 80's, 90's, 100's).
-You can draw this by hand or use `R`.^[If you are using `ggplot`, you want to use `+geom_histogram(breaks=seq(start,end,by))` and add `+scale_x_continuous(breaks=seq(start,end,by))`.
-For each, it creates bins in the histogram, and ticks on the x axis by creating a `seq`uence starting at `start` (a number), ending at `end` (number), `by` a certain interval (i.e.
-by `10`s.).] Is this distribution roughly symmetric or skewed?
-What would we expect about the mean and the median?
-e. Suppose instead the person who got the 72 did not show up that day to class, and got a 0 instead. Recalculate the mean and median. What happened and why?
+Equivalently, it can be found by subtracting the proportion of *unexplained* variation in $Y$ (SSE/TSS) from 1:
 
-\clearpage
+$$R^2=1-\frac{SSE}{TSS}=1-\frac{\sum (u_i)^2}{\sum (Y_i-\bar{Y})^2}$$
 
-4. Suppose the probabilities of a visitor to Amazon’s website buying 0, 1, or 2 books are 0.2, 0.4, and 0.4 respectively.
+This is because $\frac{SSE+ESS}{TSS}=1$. Finally, $R^2$ is the square of the correlation coefficient between $X$ and $Y$, $R^2=(r_{X,Y})^2$
 
-a. Calculate the *expected number* of books a visitor will purchase.
-b. Calculate the *standard deviation* of book purchases.
-c. **Bonus**: try doing this in `R` by making an initial dataframe of the data, and then making new columns to the "table" like we did in class.
+The closer $R^2$ is to 1, the better the fit, the closer to 0, the poorer the fit.
 
-\clearpage
+## Question 3
+In your own words, describe what the standard error of the regression ($SER$) means. How do we calculate it, what does it tell us, and how do we interpret it?
 
-5. Scores on the SAT (out of 1600) are approximately normally distributed with a mean of 500 and standard deviation of 100.
+SER $(\hat{\sigma_u}$) is the average size of the error (or residual), $\hat{u_i}$, that is, the average distance from the regression line to the actual data value for $Y$ at a given $X$.
+The goal of OLS is to minimize this (well, technically minimize the *sum* of *squared* errors!).
 
-a. What is the probability of getting a score between a 400 and a 600?
-b. What is the probability of getting a score between a 300 and a 700?
-c. What is the probability of getting *at least* a 700?
-d. What is the probability of getting *at most* a 700?
-e. What is the probability of getting exactly a 500?
+$$\begin{align*}
+	SER&=\sqrt{\frac{1}{n-2}\sum \hat{u_i}^2}\\
+	SER &=\sqrt{\frac{SSE}{n-2}}\\
+\end{align*}$$
 
-\clearpage
+We calculate it by squaring the residuals (to get a positive distance) and taking the mean of them by adding them all up and dividing by $n-2$, and then taking the square root to return to normal (non-squared) units.
 
-6.
-Redo problem 5 by using the `pnorm()` command in `R`.^[Hint: This function has four arguments: 1.
-the value of the random variable, 2.
-the mean of the distribution, 3.
-the sd of the distribution, and 4.
-`lower.tail` `TRUE` or `FALSE`.]
+We divide by $n-2$ rather than by $n$ due to the degrees of freedom correction for calculating two prior statistics with our data already, $\hat{\beta_0}$ and $\hat{\beta_1}$.
+
+## Question 4
+**In your own words, describe what homoskedasticity and heteroskedasticity mean: both in ordinary English, and in terms of the graph of the OLS regression line.**
+
+*Homoskedasticity* means the errors are distributed with the same variance for all levels of X. Knowing anything about X will not tell us anything about the distribution of errors at that level of X.
+
+*Heteroskedasticity* means the errors are distributed differently for different levels of X. So, at different levels of X, there will be much more or much less variation in the residuals.
+
+## Question 5
+**In your own words, describe what the variation in $\hat{\beta_1}$ (either variance or standard error) means, or is measuring. What three things determine the variation, and in what way?**
+
+The variation of $\hat{\beta_1}$ (either it's variance or standard error) is a measure of how *precise* our estimate is.
+This idea comes from the sampling distribution of $\hat{\beta_1}$, since it is a random variable: if we were to take other samples and calculate the slope of a regression line $\hat{\beta_1}$ for each, the estimate would vary from sample to sample.
+
+The standard error of $\hat{\beta_1}$ (square this to get variance) is:
+
+$$se(\hat{\beta_1})=\frac{\sigma_u}{\sqrt{n} \times se(X)}$$
+
+The three things that affect it are:
+
+1. Goodness of Fit of the Regression $(\sigma_u)$ or $SER$. The worse the fit, the higher the $SER$, and the worse the precision (higher standard error) of $\hat{\beta_1}$.
+
+2. Sample size, $n$: the more data, the better the precision (lower standard error) of $\hat{\beta_1}$.
+
+3. Standard error of $X$: the more variation (spread) in $X$-values, the better the precision (lower standard error) of $\hat{\beta_1}$.
+
+See the graphs in [slides 6-8](https://metricsf19.classes.ryansafner.com/slides/11-slides#6) of class 2.5 for more.
+
+## Question 6
+**In your own words, describe what a $p$-value means, and how it is used to establish statistical significance.**
+
+The $p$-value is the probability that, if the null hypothesis were true, of observing a test statistic at least as extreme as the one found in our sample.
+Specifically, if $H_0: \beta_1$, it is the probability of getting a sample slope at least as extreme as the one in our sample, if the slope were truly 0.^[Note in the classic sense, the $p$-value is actually measuring the probability of a *test statistic* $(t)$ being at least as extreme as ours.
+The test statistic essentially standardizes our sample statistic $(\hat{\beta_1})$ so that it measures standard deviations from the null-hypothesized value (i.e.
+0), much like a $Z$-score.]
+
+$$Prob(\delta \geq \delta_i|H_0\text{ is true})$$
+
+where $\delta$ is a test-statistic and $\delta_i$ is the test statistic we obtained from our sample.
+
+Another way to interpret this is that the $p$-value is the probability we commit a Type I error: the probability that, if the null hypothesis were true, we falsely reject it from our sample evidence.
+
+Be careful, the $p$-value is not the probability that our alternative hypothesis is true given our findings (commonly believed)!
+In fact it is basically the opposite, the probability of our findings being valid given the null hypothesis!
+
+## Question 7
+**A researcher is interested in examining the impact of illegal music downloads on commercial music sales.
+The author collects data on commercial sales of the top 500 singles from 2017 ($Y$) and the number of downloads from a web site that allows `file sharing' ($X$).
+The author estimates the following model**
+
+$$\text{music sales}_i = \beta_0+\beta_1 \text{illegal downloads}_i + u_i$$
+
+**The author finds a large, positive, and statistically significant estimate of $\hat{\beta_1}$.
+The author concludes these results demonstrate that illegal downloads actually *boost* music sales.
+Is this an unbiased estimate of the impact of illegal music on sales?
+Why or why not?
+Do you expect the estimate to overstate or understate the true relationship between illegal downloads and sales?**
+
+Does knowing the amount of illegal downloads an artist has convey any information about other variables that affect music sales?
+In other words, we are asking if $E[u|X]=0$ (or more simply, $cor(X,u)=0)$.
+
+It is likely that artists and songs that are the most heavily pirated are the most popular ones, and also are likely have very high music sales.
+Economists say piracy is like a tax on success--it happens more to those who are already successful and less to those who are still trying to make it big.
+
+In any case, illegal downloads is probably endogenous.
+Since there is likely a positive correlation between music sales and popularity (in the error term), and popularity is also positively correlated with music sales, it is likely that we are *overstating* the effect of illegal downloads on sales.
+In other words, $\hat{\beta_1}$ is also picking up the positive effect of popular songs, and is too large.
+The true estimate of $\beta_1$ is likely much lower than measured.
+
+## Question 8
+**A pharmaceutical company is interested in estimating the impact of a new drug on cholesterol levels.
+They enroll 200 people in a clinical trial.
+People are randomly assigned the treatment group or into the control group.
+Half of the people are given the new drug and half the people are given a sugar pill with no active ingredient.
+To examine the impact of dosage on reductions in cholesterol levels, the authors of the study regress the following model:**
+
+$$\text{cholesterol level}_i = \beta_0+\beta_1 \text{dosage level}_i + u_i$$
+
+**For people in the control group, dosage level$_i=0$ and for people in the treatment group, dosage level$_i$ measures milligrams of the active ingredient.
+In this case, the authors find a large, negative, statistically significant estimate of $\hat{\beta_1}$.
+Is this an unbiased estimate of the impact of dosage on change in cholesterol level?
+Why or why not?
+Do you expect the estimate to overstate or understate the true relationship between dosage and cholesterol level?**
+
+Does knowing whether (or how much) a person was treated convey any information about other characteristics that affect cholesterol level (in $u_i$)? Again, we are asking if $E[u|X]=0$ or $cor(X, u)=0$
+
+In this case, the answer is clearly no, the equations do hold and treatment is exogenous.
+$X_i$ is determined by random assignment, some people assigned a the drug, and some nothing at all.
+But, the important fact is that dosage is determined randomly so we expect that on average it will not be correlated with $u$.
+In this case, $E[\hat{\beta_1}]=\beta_1$, $\hat{\beta_1}$ is unbiased.
+
+# Theory Problems
+
+For the following questions, please *show all work* and explain answers as necessary.
+You may lose points if you only write the correct answer.
+You may use `R` to *verify* your answers, but you are expected to reach the answers in this section "manually."
+
+## Question 9
+
+A researcher wants to estimate the relationship between average weekly earnings $(AWE$, measured in dollars) and $Age$ (measured in years) using a simple OLS model.
+Using a random sample of college-educated full-time workers aged 25-65 yields the following:
+
+$$\widehat{AWE} = 696.70+9.60\ \, Age$$
+
+### Part A
+**Interpret what $\hat{\beta_0}$ means in this context.**
+
+$\hat{\beta_0}$ is 696.70.
+This is the vertical intercept of the regression line.
+It means that a person that is 0 years old earns a $696.70 per week on average.
+This is often nonsensical, so we don't often care about the economic meaning of the intercept.
+
+### Part B
+**Interpret what $\hat{\beta_1}$ means in this context.**
+
+$\hat{\beta_1}$ is 9.60 This is the slope of the regression line.
+It means that for every year older a person is, they can expect their wages to increase by $9.60, on average.
+This is the marginal effect of Age on AWE (and the causal effect if this model were exogenous).
+
+### Part C
+**The $R^2=0.023$ for this regression. What are the units of the $R^2$, and what does this mean?**
+
+$R^2$ has no units, it is the proportion of variation in $AWE$ that is explained by our model, between 0 and 1.
+This model explains only 2.3% of the variation in $AWE$, meaning this model is poor, and the line does not fit the data points well.
+
+### Part D
+**The $SER, \, \hat{\sigma_u}=624.1$ for this regression. What are the units of the SER in this context, and what does it mean? Is the SER large in the context of this regression?**
+
+$SER$ is measured in the same units as the dependent variable, $AWE$, so it is measured in dollars.
+It is the average error or residual for an individual, the difference (in dollars) between OLS' predicted $\widehat{AWE}$ for that person, and their true $AWE$ in the data.
+This SER is quite big, $624 in average weekly earnings.
+
+### Part E
+**Suppose Maria is 20 years old. What is her predicted $\widehat{AWE}$?**
+
+$$\begin{align*}
+\widehat{AWE}_{Maria}&=696.70+9.60(20)\\
+&=888.70\\
+\end{align*}$$
+
+### Part F
+**Suppose the data shows her *actual* $AWE$ is $430. What is her residual? Is this a relatively good or a bad prediction?**^[Hint: compare your answer here to your answer in Part D.]
+
+$$\begin{align*}
+\widehat{u}_{Maria}&=Y_{Maria}-\widehat{Y}_{Maria}\\
+&=430-888.70\\
+&=-458.70\\
+\end{align*}$$
+
+This actually a relatively good prediction, as it is much lower than the average prediction error (SER), which was 624.1.
+
+### Part G
+**What does the error term, $u_i$ represent in this case? What might individuals have different values of $u_i$?**
+
+The error term represents *all* factors *other* than age that affects an individual's average weekly earnings.
+This could include things like experience, ability, job type, education level, conscienciousness etc.
+
+### Part H
+**Do you think that $Age$ is exogenous? Why or why not? Would we expect $\hat{\beta_1}$ to be too *large* or too *small*?**
+
+It's very unlikely that $Age$ is exogenous.
+Knowing someone's age likely gives us information about $u$: we can guess about their experience or level of education (they are likely higher for older people), and most of these positively affect wages.
+Thus, we have probably *overstimated* the effect of age on earnings (i.e.
+$\hat{\beta_1}$), and the true $\beta_1$ is likely smaller.
+
+## Question 10
+
+Suppose a researcher is interested in estimating a simple linear regression model:
+
+$$Y_i=\beta_0+\beta_1X_i+u_i$$
+In a sample of 48 observations, she generates the following descriptive statistics:
+
+- $\bar{X}=30$
+- $\bar{Y}=63$
+- $\displaystyle\sum^n_{i=1}(X_i-\bar{X})^2= 6900$
+- $\displaystyle\sum^n_{i=1}(Y_i-\bar{Y})^2= 29000$
+- $\displaystyle\sum^n_{i=1}(X_i-\bar{X})(Y_i-\bar{Y})=13800$
+- $\displaystyle\sum^n_{i=1}\hat{u}^2=1656$
+
+### Part A
+**What is the OLS estimate of $\hat{\beta_1}$?**
+
+The formula for $\hat{\beta_1}=\frac{\displaystyle\sum^n_{i=1}(X_i-\bar{X})(Y_i-\bar{Y})}{\displaystyle\sum^n_{i=1}(X_i-\bar{X})^2}=\frac{cov(X,Y)}{var(X)}	= \frac{13800}{6900}=2$
+
+### Part B
+**What is the OLS estimate of $\hat{\beta_0}$?**
+
+The formula for $\hat{\beta_0}=\bar{Y}-\hat{\beta_1}\bar{X}=63-30(2)=3$
+
+### Part C
+**Suppose the OLS estimate of $\hat{\beta_1}$ has a standard error of $0.072$. Could we probably reject a null hypothesis of $H_0: \beta_1=0$ at the 95% level?**
+
+Yes, we could reject the null hypothesis as the estimate of $\hat{\beta_1}=2$ is more than 2 times its standard error of 0.072. The test-statistic would actually be
+
+$$\begin{align*}
+t&=\frac{\hat{\beta_1}-\beta_{1,0}}{se(\hat{\beta_1})}\\
+t&=\frac{2-0}{0.072}\\
+t&\approx 27.78\end{align*}$$
+
+This is well beyond the critical value needed to reject $H_0$, and the $p$-value would be basically 0.
+
+### Part D
+**Calculate the $R^2$ for this model. How much variation in $Y$ is explained by our model?**
+
+We know TSS (4\textsuperscript{th} bullet point) and SSE (last bullet point).
+
+$$\begin{align*}
+R^2&=1-\frac{SSE}{TSS}\\
+&=1-\frac{1656}{29000}\\
+&=1-0.057\\
+&=0.943\\
+\end{align*}$$
+
+This model explains 94.3% of the variation in $Y_i$.
+
+### Part E
+**How large is the average residual?**
+
+We need to find the standard error of the regression (SER), but luckily we know the SSE (last bullet point)
+
+$$\begin{align*}
+SER &= \sqrt{\frac{SSE}{n-2}}\\
+&=\sqrt{\frac{1656}{48-2}}\\
+&=\sqrt{36}\\
+&= 6\\
+\end{align*}$$
+
+This tells us the average residual is 36 (units of $Y$).
+
+# R Questions
+
+For the following problems, please attach/write the answers to each question on the same document as the previous problems, but also include a printed/attached (and commented!) *.R* script file of your commands to answer the questions.
+
+## Question 11
+
+Download the `MLBattend` dataset from Blackboard. This data contains data on attendance at major league baseball games for all 32 MLB teams from the 1970s-2000.
+
+```{r}
+# first load tidyverse 
+library(tidyverse)
+
+# import data
+mlb<-read_csv("../data/MLBAttend.csv")
+```
+
+### Part A
+
+**Clean up the data a bit by making a new variable to measure home attendance in millions. This will make it easier to interpret your regression later on.**
+
+```{r}
+# make home attendance variable in millions
+mlb<-mlb %>%
+  mutate(home_attend_mil=home_attend/1000000)
+```
+
+### Part B
+
+**Get the correlation between Runs Scored and Home Attendance.**
+
+```{r}
+# summarize and get correlation
+mlb %>%
+  summarize(Correlation = cor(runs_scored, home_attend_mil))
+```
+
+### Part C
+
+**Plot a scatterplot of Runs Scored (`y`) on Home Attendance (`x`). Add a regression line.**
+
+```{r}
+# create scatterplot with regression line 
+scatter<-ggplot(data = mlb)+
+  aes(x = home_attend_mil,
+      y = runs_scored)+
+  geom_point()+
+  geom_smooth(method="lm")
+scatter
+```
+
+### Part D
+
+**Run a regression of Runs Scored on Home Attendance. What are $\beta_0$ and $\hat{\beta_1}$? Interpret them in the context of our question.**
+
+```{r}
+# run regression, save as reg
+reg<-lm(runs_scored ~ home_attend_mil, data = mlb)
+
+# get summary of reg
+summary(reg)
+```
+
+```{r}
+# Here I'm going to save beta 0 hat and beta 1 hat
+# as objects to call up in the text of the markdown document
+# We'll need broom and tidy() first
+library(broom)
+reg_tidy<-tidy(reg)
+
+reg_tidy
+
+# first beta 0 hat
+
+beta_0_hat<-reg_tidy %>%
+  filter(term=="(Intercept)") %>% # look at intercept row
+  pull(estimate) %>% # extract beta 0 hat
+  round(., 2) # round to 2 decimal places
+
+beta_0_hat # check
+
+# now beta 1 hat
+
+beta_1_hat<-reg_tidy %>%
+  filter(term=="home_attend_mil") %>% # look at X-variable row
+  pull(estimate) %>% # extract beta 1 hat
+  round(., 2) # round to 2 decimal places
+
+beta_1_hat
+```
+
+$\hat{\beta_0}=$ `r beta_0_hat`
+$\hat{\beta_1}$: `r beta_1_hat`
+
+Teams that have a home attendance of 0 over their season score 572.618 runs. For every additional 1 million fans attending at home games over the season, a team score s68.798 more runs.
+
+### Part E
+
+**Write out the estimated regression equation.**
+
+$$\widehat{\text{Runs scored}_i}=572.618-68.798 \text{ Home attendance (mil)}$$
+
+```{r}
+# if you are using markdown, try out the equatiomatic package
+#install.packages("equatiomatic")
+library(equatiomatic)
+extract_eq(reg, # the regression
+           use_coefs = TRUE, # use the estimated numbers
+           coef_digits = 3, # how many digits to show
+           fix_signs = TRUE) # fix negatives
+```
+
+### Part F
+
+**Make a regression table of the output.**
+
+```{r}
+# load huxtable
+library(huxtable)
+huxreg(reg, # this is sufficient, the rest is decoration
+       coefs = c("Constant" = "(Intercept)",
+                 "Home Attendance (Millions)" = "home_attend_mil"),
+       statistics = c("N" = "nobs",
+                      "R-Squared" = "r.squared",
+                      "SER" = "sigma"),
+       number_format = 3)
+```
+
+### Part G
+
+**Now let's start running some diagnostics of the regression. Make a histogram of the residuals. Do they look roughly normal?**
+
+```{r}
+# here we need broom's augment() command to add residuals to the data
+
+# load broom
+library(broom)
+# augment the regression, save as reg_aug
+reg_aug<-reg %>%
+  augment()
+
+# now we use this as the data in our histogram plot in ggplot, (x is .resid)
+ggplot(data = reg_aug)+
+  aes(x = .resid)+
+  geom_histogram(color="white")
+```
+
+They do look roughly normal.
+
+### Part H
+
+**Make a residual plot.**
+
+```{r}
+# this is another plot from reg_aug, where x is home attendance and y is .fitted
+ggplot(data = reg_aug)+
+  aes(x = home_attend_mil,
+      y = .resid)+
+  geom_point()+
+  geom_hline(yintercept=0, color="red")
+```
+
+### Part I
+
+**Test the regression for heteroskedasticity.
+Are the errors homoskedastic or heteroskedastic?
+Generate robust standard errors.
+Make a regression output table, with one column with regular standard errors and another with robust standard errors.**
+
+```{r}
+# this requires the lmtest package for the bptest() command
+
+# install.packages("lmtest")
+# load lmtest
+library(lmtest)
+bptest(reg)
+```
+
+The null hypothesis $H_0$ is that the errors are homoskedastic. The $p$-value for this test is very large, so we *cannot* reject the null hypothesis.
+
+This is good, it means the errors are homoskedastic, and our OLS estimators' standard errors are accurate and do not need to be corrected for heteroskedasticity.
+
+### Part J
+
+**Test the data for outliers. If there are any, identify which team(s) and season(s) are outliers.**
+
+```{r}
+# this requires the car package for the outlierTest() command
+
+# install.packages("car")
+# load car
+
+library(car)
+outlierTest(reg)
+```
+
+This test detected one outlier, which is observation (row) number 816. Let's look it up:
+
+```{r}
+mlb %>%
+  slice(816)
+```
+
+The Toronto Blue Jays' 1981 season is an outlier. Just for kicks, let's point it out on the scatterplot.
+
+```{r}
+outlier<-mlb %>%
+  slice(816)
+
+library(ggrepel)
+scatter+ # our scatterplot saved from part c
+  geom_point(data = outlier,
+             aes(x = home_attend_mil,
+                 y = runs_scored),
+             color = "red")+
+  geom_text_repel(data = outlier,
+             aes(x = home_attend_mil,
+                 y = runs_scored),
+             label = "1981 Blue Jays",
+             color = "red")
+```
+
+### Part K
+
+**What is the marginal effect of home attendance on runs scored? Is this statistically significant? Why or why not?**
+
+This is an interpretation question, no need to calculate anything.
+The marginal effect is $\hat{\beta_1}$: for every 1 additional million fans attending home games over a team's season, the team scores 69 more runs.
+
+Looking back at the regression output in part c, the $t$-score for the hypothesis test $H_0: \, \beta_1=0$, $H_1: \, \beta_1 \neq 0$ is 16.45, yielding a very very small $p$-value.
+We have sufficient evidence to reject $H_0$ in favor of our alternative hypothesis, that there is a relationship between home attendance and runs scored over a season.
+
+### Part L
+
+**Now we'll try out the `infer` package to understand the $p$-value and confidence interval for our observed slope in our regression model.
+Save the (value of) our sample $\hat{\beta_1}$ from your regression in Part D as an object.
+Then, install and load the `infer` package, and then calculate the slope^[`calculate(stat = "slope")`] under the null hypothesis that there is no connection between attendance and runs.^[`hypothesize(null = "independence")`] for 1000 additional simulated samples^[`generate(reps = 1000, type = "permute")`], and save this as an object (it's a `tibble`).
+Then, use this to `get_p_value()`^[Set `obs_stat` equal to your observed slope, and set `direction = "two_sided"`].
+Compare to the $p$-value given by `lm()` and `tidy()` above.**
+
+```{r}
+# save slope from our regression as sample_slope
+
+sample_slope<-reg_tidy %>%
+  filter(term=="home_attend_mil") %>%
+  pull(estimate)
+
+# double check it worked
+sample_slope
+# note I could also just use the beta_1_hat I optionally made in Part D
+
+# make 1000 simulations of sample slopes under null hypothesis that slope = 0
+
+#install.packages(infer)
+library(infer) # load infer
+slope_simulations<-mlb %>% # save as a tibble
+  specify(runs_scored ~ home_attend_mil) %>% # our lm model
+  hypothesize(null = "independence") %>% # null hypothesis, slope = 0 (X and Y independent)
+  generate(reps = 1000, type = "permute") %>% # make 1000 permutations
+  calculate(stat = "slope") # calculate the sample slope of each permutation
+
+# make sure it worked
+slope_simulations %>%
+  head(., n = 10) # there are a LOT! I only print the first 10 for space
+
+slope_simulations %>%
+  get_p_value(obs_stat = sample_slope,
+              direction = "both")
+```
+
+The $p$-value is basically 0. According to the regression output in part D, it is smaller than 0.00000000000000002.
+
+### Part M
+
+**Make a histogram of the simulated slopes, and plot our sample slope on that histogram, shading the $p$-value.^[You can use `ggplot2` to plot a histogram in the normal way and add a `geom_vline()`, setting `xintercept` equal to your saved object with the sample $\hat{\beta_1}$ value.
+*Alternatively*, you can use `infer` to pipe your `tibble` of simulations into `visualize()`, and inside `visualize()` set `obs_stat` equal to your saved $\hat{\beta_1}$ object.
+*Regardless* of which method you use, add `+shade_p_value()`.
+Inside this, set `obs_stat` equal to your saved slope, and add `direction = "two_sided"`.]**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# now make a histogram of this
+slope_simulations %>%
+  visualize(obs_stat = sample_slope)+
+  shade_p_value(obs_stat = sample_slope, # set the obs_stat equal to our saved slope from above
+                direction = "both") # two-sided test, shade both sides
+```
+
+```{r}
+# infer no longer shades the p-value, so I'll do it manually
+
+slope_simulations %>%
+  ggplot(data = .)+
+  aes(x = stat)+ # slope from simulations
+  geom_histogram(color = "white", # the histogram
+                 fill = "indianred")+
+  # add vertical line for our sample slope
+  geom_vline(xintercept = sample_slope,
+             color = "red",
+             size = 2)+
+  # add label
+  geom_label(x = sample_slope,
+             y = 200,
+             color = "red",
+             label = expression(paste("Our ", hat(beta[1]))))+
+  # add "shading" for p-value's two sides
+  # right side
+  geom_rect(xmin=68.798,
+            xmax=100,
+            ymin=0,
+            ymax=300,
+            color = "red",
+            alpha=0.5)+
+  # left side
+  geom_rect(xmin=-100,
+            xmax=-68.798,
+            ymin=0,
+            ymax=300,
+            color = "red",
+            alpha=0.5)+
+  scale_x_continuous(seq(-75,75,25))+
+  labs(x = expression(paste("Distribution of ", hat(beta[1]), " under ", H[0], " that ", beta[1]==0)),
+       y = "Samples")+
+    theme_classic(base_family = "Fira Sans Condensed",
+           base_size=20)
+```
+
+Note this is the sampling distribution of $\hat{\beta_1}$ under the null hypothesis (the true $\beta_1=0)$.
+Values on the horizontal axis are values of $\hat{\beta_1}$, *not* the number of standard deviations away from the null hypothesis.
+
+What the test-statistic does is standardize this distribution much like we would standardize a distribution to the standard normal distribution via calculating $Z$-scores.
+
+Let me visualize what would happen (and this is what R does to calculate the `t value` and associationed `p`-value.)
+
+```{r}
+slope_simulations %>%
+  summarize(mean = mean(stat), # get the mean slope of the simulated distribution of slopes
+            se = (sd(stat))) # get the standard error of the slope 
+```
+
+```{r}
+# standardize slopes to t-statistics
+tstatistics<-slope_simulations %>%
+  mutate(tscores = ((stat - mean(stat))/sd(stat)))
+
+our_t<-((sample_slope-mean(slope_simulations$stat))/sd(slope_simulations$stat))
+
+# what is our t-statistic?
+our_t
+
+# now plot t-statistics %>%
+ggplot(data = tstatistics)+
+  aes(x = tscores)+ # slope from simulations
+  geom_histogram(color = "white", # the histogram
+                 fill = "indianred")+
+  # add vertical line for our sample slope's t-statistic
+  geom_vline(xintercept = our_t,
+             color = "red",
+             size = 2)+
+  #add label
+  geom_label(x = our_t,
+             y = 50,
+             color = "red",
+             label = "Our t")
+```
+
+This is off slightly (see the t-statistic R calculated in part C), but it gives you an idea.
+Our test-statistic is somewhere around 14-16, this means it is 14-16 standard deviations away from 0, so again, a very very low p-value!
+
+### Part N
+
+**Get the 95% confidence interval for your slope estimate,^[`tidy()` your original regression, with `conf.int = TRUE` inside the command, then `select(conf.low, conf.high)` and `filter` by your `x` variable.
+Save this as an object.] and then make a histogram of the simulated slopes (like part L), but instead, add `+shade_confidence_interval()`.^[Inside of this, set `endpoints` equal to the object you just made with the low and high confidence interval values.] Compare this to what you get with `tidy()` above.**
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# save as a tibble called "ci_values"
+ci_values<-slope_simulations %>%
+  get_confidence_interval(level = 0.95,
+                          type = "se",
+                          point_estimate = sample_slope)
+
+# see what we made
+ci_values
+
+# we'll use this for the endpoints in the shade_ci() command
+
+slope_simulations %>%
+  visualize(obs_stat = sample_slope)+
+  shade_confidence_interval(endpoints = ci_values)
+```

### 2019-10-12 static/homeworks/03-problem-set-answers.Rmd
Update answer key 3 graph

@@
                  fill = "indianred")+
+  # add "shading" for p-value's two sides
+  # right side
+  geom_rect(xmin=68.798,
+            xmax=100,
+            ymin=0,
+            ymax=450,
+            fill = "pink",
+            alpha=0.4)+
+  # left side
+  geom_rect(xmin=-100,
+            xmax=-68.798,
+            ymin=0,
+            ymax=450,
+            fill = "pink",
+            alpha=0.4)+
   # add vertical line for our sample slope
@@
              label = expression(paste("Our ", hat(beta[1]))))+
-  # add "shading" for p-value's two sides
-  # right side
-  geom_rect(xmin=68.798,
-            xmax=100,
-            ymin=0,
-            ymax=300,
-            color = "red",
-            alpha=0.5)+
-  # left side
-  geom_rect(xmin=-100,
-            xmax=-68.798,
-            ymin=0,
-            ymax=300,
-            color = "red",
-            alpha=0.5)+
-  scale_x_continuous(seq(-75,75,25))+
+  scale_x_continuous(breaks=c(-75,-50,-25,0,25,50,75),
+                     limits=c(-75,75),
+                     expand=c(0,0))+
+  scale_y_continuous(limits=c(0,450),
+                       expand=c(0,0))+
   labs(x = expression(paste("Distribution of ", hat(beta[1]), " under ", H[0], " that ", beta[1]==0)),

### 2019-10-24 static/slides/14-slides.Rmd
Add class 14 content

---
title: "3.2: Multivariate OLS Estimators: Bias, Precision, and Fit"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))
```

```{r regression-setup, echo=F, results="hide"}

library(haven)
CASchool<-read_dta("../data/caschool.dta")

# run regression of testscr on str
school_reg <- lm(testscr ~ str, 
                 data = CASchool)

library(broom)
school_reg_tidy <- tidy(school_reg,
     conf.int = TRUE) # add confidence intervals
CASchool_aug <- augment(school_reg)

library(equatiomatic)
extract_eq(school_reg, use_coefs = TRUE,coef_digits = 2,fix_signs = TRUE)

```

class: inverse, center, middle

# The Multivariate OLS Estimators

---

# The Multivariate OLS Estimators

- By analogy, we still focus on the .shout[ordinary least squares (OLS) estimators] of the unknown population parameters $\beta_0, \beta_1, \beta_2, \cdots, \beta_k$ which solves:

$$\min_{\hat{\beta_0}, \hat{\beta_1}, \hat{\beta_2}, \cdots, \hat{\beta_k}} \sum^n_{i=1}\left[\underbrace{Y_i-(\hat{\beta_0}+\hat{\beta_1}X_{1i}+\hat{\beta_2}X_{2i}+\cdots+ \hat{\beta_k}X_{ki})}_{u_i}\right]^2$$

- Again, OLS estimators are chosen to **minimize** the **sum of squared errors (SSE)**
  - i.e. the sum of the squared distances between the actual values of $Y_i$ and the predicted values $\hat{Y_i}$

---

# The Multivariate OLS Estimators: FYI

.content-box-red[
.red[**Math FYI]**: in linear algebra terms, a regression model with $n$ observations of $k$ independent variables:

$$\mathbf{Y} = \mathbf{X \beta}+\mathbf{u}$$

$$\underbrace{\begin{pmatrix}
			y_1\\
			y_2\\
			\vdots \\
			y_n\\
		\end{pmatrix}}_{\mathbf{Y}_{(n \times 1)}}
		=
		\underbrace{\begin{pmatrix}
	x_{1,1} & x_{1,2} & \cdots & x_{1,n}\\
	x_{2,1} & x_{2,2} & \cdots & x_{2,n}\\
	\vdots & \vdots & \ddots & \vdots\\
	x_{k,1} & x_{k,2} & \cdots & x_{k,n}\\ 
\end{pmatrix}}_{\mathbf{X}_{(n \times k)}}
		\underbrace{\begin{pmatrix}
\beta_1\\
\beta_2\\
\vdots \\
\beta_k \\	
\end{pmatrix}}_{\mathbf{\beta}_{(k \times 1)}}
+
		\underbrace{\begin{pmatrix}
			u_1\\
			u_2\\
			\vdots \\
			u_n \\
		\end{pmatrix}}_{\mathbf{u}_{(n \times 1)}}$$

]

--

- The OLS estimator for $\beta$ is $\hat{\beta}=(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y}$

--

- Appreciate that I am saving you from such sorrow

---

class: center, middle, inverse

# The Sampling Distribution of $\hat{\beta_j}$

---

# The Sampling Distribution of $\hat{\beta_j}$ I

.pull-left[

- For any individual $\beta_j$, it has a sampling distribution:

$$\hat{\beta_j} \sim N \left(E[\hat{\beta_j}], \;se(\hat{\beta_j})\right)$$

- We want to know its sampling distribution's:
  - **Center**: $E[\hat{\beta_j}]$; what is the *expected value* of our estimator?
  - **Spread**: $se(\hat{\beta_j})$; how *precise* is our estimator?
]

.pull-right[

```{r, fig.retina=3}
beta_dist<-ggplot(data = tibble(x=-4:4))+
  aes(x = x)+
  stat_function(fun = dnorm, size=2, color="blue")+
  geom_segment(aes(x=0,xend=0, y=0, yend=0.4), linetype="dashed")+
  scale_x_continuous(breaks = 0,
                     labels = expression(E(hat(beta[j]))))+
  labs(x = expression(hat(beta[j])),
       y = "Probability")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
beta_dist
```
]

---

class: inverse, center, middle

# The Expected Value of $\hat{\beta_j}$: Bias

---

# Exogeneity and Unbiasedness

- As before, $E[\hat{\beta_j}]=\beta_j$ when $X_j$ is .onfire[exogenous] (i.e. $cor(X_j, u)=0$)

- We know the true $E[\hat{\beta_j}]=\beta_j+\underbrace{cor(X_j,u)\frac{\sigma_u}{\sigma_{X_j}}}_{\text{O.V. Bias}}$

- If $X_j$ is .shout[endogenous] (i.e. $cor(X_j, u)\neq 0$), contains **omitted variable bias**

- We can now try to *quantify* the omitted variable bias

---

# Measuring Omitted Variable Bias I

- Suppose the **_true_ population model** of a relationship is:

$$Y_i=\beta_0+\beta_1 X_{1i}+\beta_2 X_{2i}+u_i$$

- What happens when we run a regression and **omit** $X_{2i}$?

--

- Suppose we estimate the following .onfire[omitted regression] of just $Y_i$ on $X_{1i}$ (omitting $X_{2i})$:<sup>.red[+]</sup>

$$Y_i=\alpha_0+\alpha_1 X_{1i}+\nu_i$$

.footnote[<sup>.red[+]</sup> Note: I am using `\\(\alpha\\)`'s and `\\(\nu_i\\)` only to denote these are different estimates than the **true** model `\\(\beta\\)`'s and `\\(u_i\\)`]

---

# Measuring Omitted Variable Bias II

- .onfire[**Key Question**:] are $X_{1i}$ and $X_{2i}$ correlated?

--

- Run an .onfire[auxiliary regression] of $X_{2i}$ on $X_{1i}$ to see:<sup>.red[+]</sup>

$$X_{2i}=\delta_0+\delta_1 X_{1i}+\tau_i$$

--

- If $\delta_1=0$, then $X_{1i}$ and $X_{2i}$ are *not* linearly related

- If $|\delta_1|$ is very big, then $X_{1i}$ and $X_{2i}$ are strongly linearly related

.footnote[<sup>.red[+]</sup> Note: I am using `\\(\delta\\)`'s and `\\(\tau\\)` to differentiate estimates for this model.]

---

# Measuring Omitted Variable Bias III

- Now substitute our .onfire[auxiliary regression] between $X_{2i}$ and $X_{1i}$ into the .shout[*true* model]:
  - We know $X_{2i}=\delta_0+\delta_1 X_{1i}+\tau_i$

$$\begin{align*}
	Y_i&=\beta_0+\beta_1 X_{1i}+\beta_2 X_{2i}+u_i	\\
\end{align*}$$

---

# Measuring Omitted Variable Bias III

- Now substitute our .onfire[auxiliary regression] between $X_{2i}$ and $X_{1i}$ into the .shout[*true* model]:
  - We know $X_{2i}=\delta_0+\delta_1 X_{1i}+\tau_i$

$$\begin{align*}
	Y_i&=\beta_0+\beta_1 X_{1i}+\beta_2 X_{2i}+u_i	\\
	Y_i&=\beta_0+\beta_1 X_{1i}+\beta_2 \big(\delta_0+\delta_1 X_{1i}+\tau_i \big)+u_i	\\
\end{align*}$$

---
# Measuring Omitted Variable Bias III

- Now substitute our .onfire[auxiliary regression] between $X_{2i}$ and $X_{1i}$ into the .shout[*true* model]:
  - We know $X_{2i}=\delta_0+\delta_1 X_{1i}+\tau_i$

$$\begin{align*}
	Y_i&=\beta_0+\beta_1 X_{1i}+\beta_2 X_{2i}+u_i	\\
	Y_i&=\beta_0+\beta_1 X_{1i}+\beta_2 \big(\delta_0+\delta_1 X_{1i}+\tau_i \big)+u_i	\\
Y_i&=(\beta_0+\beta_2 \delta_0)+(\beta_1+\beta_2 \delta_1)X_{1i}+(\beta_2 \tau_i+u_i)\\
\end{align*}$$

---

# Measuring Omitted Variable Bias III

- Now substitute our .onfire[auxiliary regression] between $X_{2i}$ and $X_{1i}$ into the .shout[*true* model]:
  - We know $X_{2i}=\delta_0+\delta_1 X_{1i}+\tau_i$

$$\begin{align*}
	Y_i&=\beta_0+\beta_1 X_{1i}+\beta_2 X_{2i}+u_i	\\
	Y_i&=\beta_0+\beta_1 X_{1i}+\beta_2 \big(\delta_0+\delta_1 X_{1i}+\tau_i \big)+u_i	\\
Y_i&=(\underbrace{\beta_0+\beta_2 \delta_0}_{\alpha_0})+(\underbrace{\beta_1+\beta_2 \delta_1}_{\alpha_1})X_{1i}+(\underbrace{\beta_2 \tau_i+u_i}_{\nu_i})\\
\end{align*}$$

- Now relabel each of the three terms as the OLS estimates $(\alpha$'s) and error $(\nu_i)$ from the .onfire[omitted regression], so we again have:

$$Y_i=\alpha_0+\alpha_1X_{1i}+\nu_i$$

--

- Crucially, this means that our OLS estimate for $X_{1i}$ in the .onfire[omitted regression] is:
$$\alpha_1=\beta_1+\beta_2 \delta_1$$

---

# Measuring Omitted Variable Bias IV

.center[
$\alpha_1=$.green[`\\(\beta_1\\)`] $+$ .red[`\\(\beta_2\\)`].purple[`\\(\delta_1\\)`]
]

- The **Omitted Regression** OLS estimate for $X_{1i}$ picks up *both*:

1. .green[The true effect of `\\(X_{1}\\)` on `\\(Y_i\\)`: `\\((\beta_1)\\)`]

2. .red[The true effect of `\\(X_{2}\\)` on `\\(Y_i\\)`: `\\((\beta_2)\\)`]
  - As pulled through .purple[the relationship between `\\(X_1\\)` and `\\(X_2\\)`: `\\((\delta_1)\\)`]

--

- Recall our conditions for omitted variable bias from some variable $Z_i$:

1. $\mathbf{Z_i}$ **must be a determinant of $Y_i$** $\implies$ .red[`\\(\beta_2 \neq 0\\)`]

2. $\mathbf{Z_i}$ **must be correlated with $X_i$** $\implies$ .purple[`\\(\delta_1 \neq 0\\)`]

--

- Otherwise, if $Z_i$ does not fit these conditions, $\alpha_1=\beta_1$ and the .onfire[omitted regression] is *unbiased*!

---

# Measuring Omitted Variable Bias in Our Class Size Example I

.pull-left[

```{r}
true<-lm(testscr~str+el_pct, data=CASchool)
summary(true)
```
]

.pull-right[

- The "True" Regression $(Y_i$ on $X_{1i}$ and $X_{2i})$

$$\widehat{\text{Test Score}_i}=686.03-1.10\text{ STR}_i-0.65\text{ %EL}_i$$
]

---

# Measuring Omitted Variable Bias in Our Class Size Example II

.pull-left[

```{r}
omitted<-lm(testscr~str, data=CASchool)
summary(omitted)
```
]

.pull-right[

- The "Omitted" Regression $(Y_{i}$ on just $X_{1i})$

$$\widehat{\text{Test Score}_i}=698.93-2.28\text{ STR}_i$$
]

---

# Measuring Omitted Variable Bias in Our Class Size Example III

.pull-left[

```{r}
aux<-lm(el_pct~str, data=CASchool)
summary(aux)
```
]

.pull-right[

- The "Auxiliary" Regression $(X_{2i}$ on $X_{1i})$

$$\widehat{\text{%EL}_i}=-19.85+1.81\text{ STR}_i$$
]

---

# Measuring Omitted Variable Bias in Our Class Size Example IV

.pull-left[
.center[
"True" Regression

$$\widehat{\text{Test Score}_i}=686.03-1.10\text{ STR}_i-0.65\text{ %EL}$$

"Omitted" Regression

$\widehat{\text{Test Score}_i}=698.93$ .blue[`\\(-2.28\\)`] $\text{ STR}_i$

"Auxiliary" Regression

$$\widehat{\text{%EL}_i}=-19.85+1.81\text{ STR}_i$$

]
]

.pull-right[

- Omitted Regression estimate for $\alpha_1$ on STR is .blue[`\\(-2.28\\)`]
]

---

# Measuring Omitted Variable Bias in Our Class Size Example IV

.pull-left[
.center[
"True" Regression

$\widehat{\text{Test Score}_i}=686.03$ .green[`\\(-1.10\\)`] $\text{ STR}_i-0.65\text{ %EL}$

"Omitted" Regression

$\widehat{\text{Test Score}_i}=698.93$ .blue[`\\(-2.28\\)`] $\text{ STR}_i$

"Auxiliary" Regression

$\widehat{\text{%EL}_i}=-19.85+1.81\text{ STR}_i$

]
]

.pull-right[

- Omitted Regression estimate for $\alpha_1$ on STR is .blue[`\\(-2.28\\)`]

.center[
$\alpha_1=$.green[`\\(\beta_1\\)`] $+$ .red[`\\(\beta_2\\)`].purple[`\\(\delta_1\\)`]
]

- .green[The true effect of STR on Test Score: -1.10]

]

---

# Measuring Omitted Variable Bias in Our Class Size Example IV

.pull-left[
.center[
"True" Regression

$\widehat{\text{Test Score}_i}=686.03$ .green[`\\(-1.10\\)`] $\text{ STR}_i$ .red[`\\(-0.65\\)`] $\text{ %EL}$

"Omitted" Regression

$\widehat{\text{Test Score}_i}=698.93$ .blue[`\\(-2.28\\)`] $\text{ STR}_i$

"Auxiliary" Regression

$\widehat{\text{%EL}_i}=-19.85+1.81\text{ STR}_i$

]
]

.pull-right[

- Omitted Regression estimate for $\alpha_1$ on STR is .blue[`\\(-2.28\\)`]

.center[
$\alpha_1=$.green[`\\(\beta_1\\)`] $+$ .red[`\\(\beta_2\\)`].purple[`\\(\delta_1\\)`]
]

- .green[The true effect of STR on Test Score: -1.10]

- .red[The true effect of %EL on Test Score: -0.65]

]

---

# Measuring Omitted Variable Bias in Our Class Size Example IV

.pull-left[
.center[
"True" Regression

$\widehat{\text{Test Score}_i}=686.03$ .green[`\\(-1.10\\)`] $\text{ STR}_i$ .red[`\\(-0.65\\)`] $\text{ %EL}$

"Omitted" Regression

$\widehat{\text{Test Score}_i}=698.93$ .blue[`\\(-2.28\\)`] $\text{ STR}_i$

"Auxiliary" Regression

$\widehat{\text{%EL}_i}=-19.85+$ .purple[1.81] $\text{ STR}_i$

]
]

.pull-right[

- Omitted Regression estimate for $\alpha_1$ on STR is .blue[`\\(-2.28\\)`]

.center[
$\alpha_1=$.green[`\\(\beta_1\\)`] $+$ .red[`\\(\beta_2\\)`].purple[`\\(\delta_1\\)`]
]

- .green[The true effect of STR on Test Score: -1.10]

- .red[The true effect of %EL on Test Score: -0.65]

- .purple[The relationship between STR and %EL: 1.81]
]

---

# Measuring Omitted Variable Bias in Our Class Size Example IV

.pull-left[
.center[
"True" Regression

$\widehat{\text{Test Score}_i}=686.03$ .green[`\\(-1.10\\)`] $\text{ STR}_i$ .red[`\\(-0.65\\)`] $\text{ %EL}$

"Omitted" Regression

$\widehat{\text{Test Score}_i}=698.93$ .blue[`\\(-2.28\\)`] $\text{ STR}_i$

"Auxiliary" Regression

$\widehat{\text{%EL}_i}=-19.85+$ .purple[1.81] $\text{ STR}_i$

]
]

.pull-right[

- Omitted Regression estimate for $\alpha_1$ on STR is .blue[`\\(-2.28\\)`]

.center[
$\alpha_1=$.green[`\\(\beta_1\\)`] $+$ .red[`\\(\beta_2\\)`].purple[`\\(\delta_1\\)`]
]

- .green[The true effect of STR on Test Score: -1.10]

- .red[The true effect of %EL on Test Score: -0.65]

- .purple[The relationship between STR and %EL: 1.81]

- So, for the .onfire[omitted regression]:

.center[
.blue[`\\(\alpha_1 =-2.28\\)`] $=$.green[`\\(-1.10\\)`] $+$ .red[`\\(-0.65\\)`] $($ .purple[`\\(1.81\\)`] $)$

]

]

---

# Measuring Omitted Variable Bias in Our Class Size Example IV

.pull-left[
.center[
"True" Regression

$\widehat{\text{Test Score}_i}=686.03$ .green[`\\(-1.10\\)`] $\text{ STR}_i$ .red[`\\(-0.65\\)`] $\text{ %EL}$

"Omitted" Regression

$\widehat{\text{Test Score}_i}=698.93$ .blue[`\\(-2.28\\)`] $\text{ STR}_i$

"Auxiliary" Regression

$\widehat{\text{%EL}_i}=-19.85+$ .purple[1.81] $\text{ STR}_i$

]
]

.pull-right[

- Omitted Regression estimate for $\alpha_1$ on STR is .blue[`\\(-2.28\\)`]

.center[
$\alpha_1=$.green[`\\(\beta_1\\)`] $+$ .red[`\\(\beta_2\\)`].purple[`\\(\delta_1\\)`]
]

- .green[The true effect of STR on Test Score: -1.10]

- .red[The true effect of %EL on Test Score: -0.65]

- .purple[The relationship between STR and %EL: 1.81]

- So, for the .onfire[omitted regression]:

.center[
.blue[`\\(\alpha_1 =-2.28\\)`] $=$.green[`\\(-1.10\\)`] $+$ .red[`\\(-0.65\\)`] $($ .purple[`\\(1.81\\)`] $)$

- The **bias** is $($ .red[`\\(-0.65\\)`] $)($ .purple[`\\(1.81\\)`] $)=\mathbf{-1.18}$

]
]

---
class: inverse, center, middle
# Precision of $\hat{\beta_j}$

---

# Precision of $\hat{\beta_j}$ I

.pull-left[

- $\sigma_{\hat{\beta_j}}$; how **precise** are our estimates? (today)
    - <span class="shout">Variance $\sigma^2_{\hat{\beta_j}}$</span> or <span class="shout">standard error $\sigma_{\hat{\beta_j}}$</span>
]

.pull-right[

```{r, fig.retina=3}
beta_dist+
  geom_label(x=1, y=dnorm(1), label=expression(sigma[hat(beta[j])]==1), color="blue")+
  stat_function(fun = dnorm, args=list(mean = 0, sd = 2), size=2, color="red")+
  geom_label(x=2, y=dnorm(2,0,2), label=expression(sigma[hat(beta[j])]==2), color="red")
```
]

---

# Precision of $\hat{\beta_j}$ II

.pull-left[

$$var(\hat{\beta_j})=\underbrace{\frac{1}{1-R^2_j}}_{VIF} \times \frac{(SER)^2}{n \times var(X)}$$

$$se(\hat{\beta_j})=\sqrt{var(\hat{\beta_1})}$$

]

.pull-right[

- Variation in $\hat{\beta_j}$ is affected by **four** things now:

1. **Goodness of fit of the model $(SER)$**
    - Larger $SER$ $\rightarrow$ larger $var(\hat{\beta_j})$
2. **Sample size, $n$**
    - Larger $n$ $\rightarrow$ smaller $var(\hat{\beta_j})$
3. **Variance of $X$**
    - Larger $var(X)$ $\rightarrow$ smaller $var(\hat{\beta_j})$
4. **Variance Inflation Factor** $\frac{1}{(1-R^2_j)}$
    - Larger $VIF$, larger $var(\hat{\beta_j})$
    - This is the only new effect, explained in a moment

]

.footnote[<sup>.red[+]</sup> See Class 2.5 for a reminder of variation with just variable.]

---

# VIF and Multicollinearity I

- Two *independent* variables are .shout[multicollinear]:

$$cor(X_j, X_l) \neq 0 \quad \forall j \neq l$$

--

- .onfire[Multicollinearity between X variables does *not bias* OLS estimates]
  - Remember, we pulled another variable out of $u$ into the regression
  - If it were omitted, then it *would* cause omitted variable bias!

--

- .onfire[Multicollinearity does *increase the variance* of an estimate] by

$$VIF=\frac{1}{(1-R^2_j)}$$

---

# VIF and Multicollinearity II

$$VIF=\frac{1}{(1-R^2_j)}$$

- $R^2_j$ is the $R^2$ from an **auxiliary regression** of $X_j$ on all other regressors $(X$'s)

--

.content-box-green[
.green[**Example**]: Suppose we have a regression with three regressors $(k=3)$:

$$Y_i=\beta_0+\beta_1X_{1i}+\beta_2X_{2i}+\beta_3X_{3i}$$
]

--

- There will be three different $R^2_j$'s, one for each regressor:

$$\begin{align*}
R^2_1 \text{ for }  X_{1i}&=\gamma+\gamma X_{2i} + \gamma X_{3i}	\\
R^2_2 \text{ for }  X_{2i}&=\zeta_0+\zeta_1 X_{1i} + \zeta_2 X_{3i}	\\
R^2_3 \text{ for }  X_{3i}&=\eta_0+\eta_1 X_{1i} + \eta_2 X_{2i}	\\
  \end{align*}$$

---

# VIF and Multicollinearity III

$$VIF=\frac{1}{(1-R^2_j)}$$

- $R^2_j$ is the $R^2$ from an **auxiliary regression** of $X_j$ on all other regressors $(X$'s)

- The $R_j^2$ tells us .onfire[how much *other* regressors explain regressor `\\(X_j\\)`]

- **Key Takeaway**: If other $X$ variables explain $X_j$ well (high $R^2_J$), it will be harder to tell how *cleanly* $X_j \rightarrow Y_i$, and so $var(\hat{\beta_j})$ will be higher

---

# VIF and Multicollinearity IV

- Common to calculate the .shout[Variance Inflation Factor (VIF)] for each regressor:

$$VIF=\frac{1}{(1-R^2_j)}$$

- VIF quantifies the factor (scalar) by which $var(\hat{\beta_j})$ increases because of multicollinearity
  - e.g. increases 2x, 3x, etc.
--

- Baseline: $R^2_j=0$ $\implies$ *no* multicollinearity $\implies$ VIF = 1 (no inflation)

--

- Larger $R^2_j$ $\implies$ larger VIF
    - Rule of thumb: $VIF>10$ is worrisome

---

# VIF and Multicollinearity V

.pull-left[
```{r el-str-scatter, echo=T, eval=F}
ggplot(data=CASchool, aes(x=str,y=el_pct))+
  geom_point(color="blue")+
  geom_smooth(method="lm", color="red")+
  labs(x = "Student to Teacher Ratio",
       y = "Percentage of ESL Students")+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

```{r, echo=T}
CAcorr_ex<-subset(CASchool, select=c("testscr", "str", "el_pct"))

# Make a correlation table
cor(CAcorr_ex)
```

]

.pull-right[
```{r ref.label="el-str-scatter", fig.retina=3}

```

]

---

# VIF and Multicollinearity in R I

```{r, echo=T}
# our multivariate regression
elreg<-lm(testscr~str+el_pct,data=CASchool)

# use the "car" package for VIF function 
library("car")

# syntax: vif(lm.object)
vif(elreg)
```

--

- $var(\hat{\beta_1})$ on `str` increases by 1.036 times due to multicollinearity with `el_pct`
- $var(\hat{\beta_2})$ on `el_pct` increases by 1.036 times due to multicollinearity with `str`

---

# VIF and Multicollinearity in R II

- Let's calculate VIF manually to see where it comes from:

--

```{r,echo=T}
# run auxiliary regression of x2 on x1

auxreg<-lm(el_pct~str, data=CASchool)
# use broom package's tidy() command (cleaner)
library(broom) # load broom
tidy(auxreg) # look at reg output
```

---

# VIF and Multicollinearity in R III

```{r,echo=T}
glance(auxreg) # look at aux reg stats for R^2

# extract our R-squared from aux regression (R_j^2)
aux_r_squared<-glance(auxreg) %>%
  pull(r.squared)
aux_r_squared # look at it
```

---

# VIF and Multicollinearity in R IV

```{r,echo=T}
# calculate VIF manually

our_vif<-1/(1-aux_r_squared) # VIF formula 
our_vif
```

- Again, multicollinearity between the two $X$ variables inflates the variance on each by 1.036 times

---

# VIF and Multicollinearity: Another Example I

.content-box-green[
.green[**Example**:] For our Test Scores and Class Size example, what about district expenditures per student?
]

```{r, echo=T}
CAcorr2<-subset(CASchool, select=c("testscr", "str", "expn_stu"))

# Make a correlation table
corr2<-cor(CAcorr2)

# look at it
corr2
```

---

# VIF and Multicollinearity: Another Example II

.pull-left[
```{r exp-str-scatter, echo=T, eval=F}
ggplot(data=CASchool, aes(x=str,y=expn_stu))+
  geom_point(color="blue")+
  geom_smooth(method="lm", color="red")+
  scale_y_continuous(labels = scales::dollar)+
  labs(x = "Student to Teacher Ratio",
       y = "Expenditures per Student ($)")+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

]

.pull-right[
```{r ref.label="exp-str-scatter", fig.retina=3}

```

]

---

# VIF and Multicollinearity: Another Example III

1. $cor(\text{Test score, expn})\neq0$

2. $cor(\text{STR, expn})\neq 0$

--

- Omitting $expn$ will **bias** $\hat{\beta_1}$ on STR

--

- *Including* $expn$ will *not* bias $\hat{\beta_1}$ on STR, but *will* make it less precise (higher variance)

--

- Data tells us little about the effect of a change in $STR$ holding $expn$ constant
  - Hard to know what happens to test scores when high $STR$ AND high $expn$ and vice versa (*they rarely happen simultaneously*)!

---

# VIF and Multicollinearity: Another Example IV

.pull-left[

```{r, echo=T}
expreg<-lm(testscr~str+expn_stu, data=CASchool)
summary(expreg)
```

]

.pull-right[
```{r, echo=T}
vif(expreg)
```

- Including `expn_stu` increases variance of $\hat{\beta_1}$ by 1.62 times

]

---

# Multicollinearity Increases Variance

.pull-left[
```{r multioutput, echo=T, eval=F}
library(huxtable)
huxreg("Model 1" = school_reg,
       "Model 2" = expreg,
       coefs = c("Intercept" = "(Intercept)",
                 "Class Size" = "str",
                 "Expenditures per Student" = "expn_stu"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```

- We can see $SE(\hat{\beta_1})$ on `str` increases from 0.48 to 0.61 when we add `expn_stu`

]

.pull-right[
```{r, ref.label="multioutput"}
```

]

---

# Perfect Multicollinearity

- .shout[*Perfect* multicollinearity] is when a regressor is an exact linear function of (an)other regressor(s)

--

$$\widehat{Sales} = \hat{\beta_0}+\hat{\beta_1}\text{Temperature (C)} + \hat{\beta_2}\text{Temperature (F)}$$

--

$$\text{Temperature (F)}=32+1.8*\text{Temperature (C)}$$

--

- $cor(\text{temperature (F), temperature (C)})=1$

--

- $R^2_j=1$ is implying $VIF=\frac{1}{1-1}$ and $var(\hat{\beta_j})=0$!

--

- **This is fatal for a regression**
  - A logical impossiblity, *always caused by human error*

---

# Perfect Multicollinearity: Example

.content-box-green[
.green[**Example**:]

$$\widehat{TestScore_i} = \hat{\beta_0}+\hat{\beta_1}STR_i	+\hat{\beta_2}\%EL+\hat{\beta_3}\%ES$$

]

- %EL: the percentage of students learning English

- %ES: the percentage of students fluent in English

- $ES=100-EL$

- $|cor(ES, EL)|=1$

---

# Perfect Multicollinearity Example II

```{r, echo=T}
# generate %EF variable from %EL
CASchool_ex <- CASchool %>%
  mutate(ef_pct = 100 - el_pct)

CASchool_ex %>%
  summarize(cor = cor(ef_pct, el_pct))
```

---

# Perfect Multicollinearity Example III

.pull-left[
```{r multicol, echo=T, eval=F}
ggplot(data=CASchool_ex, aes(x=el_pct,y=ef_pct))+
  geom_point(color="blue")+
  scale_y_continuous(labels = scales::dollar)+
  labs(x = "Percent of ESL Students",
       y = "Percent of Non-ESL Students")+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

]

.pull-right[
```{r ref.label="multicol", fig.retina=3}

```

]

---

# Perfect Multicollinearity Example IV

.pull-left[
```{r, echo=T}
mcreg<-lm(testscr~str+el_pct+ef_pct, data=CASchool_ex)
summary(mcreg)
```

]

.pull-right[
- Note `R` *ignores* one of the multicollinear regressors (`ef_pct`) if you include both in a regression
]

---

class: inverse, center, middle

# A Summary of Multivariate OLS Estimator Properties

---

# A Summary of Multivariate OLS Estimator Properties

- $\hat{\beta_j}$ on $X_j$ is biased only if there is an omitted variable $(Z)$ such that: 
    1. $cor(Y,Z)\neq 0$
    2. $cor(X_j,Z)\neq 0$
    - If $Z$ is *included* and $X_j$ is collinear with $Z$, this does *not* cause a bias

- $var[\hat{\beta_j}]$ and $se[\hat{\beta_j}]$ measure precision of estimate:

--

$$var[\hat{\beta_j}]=\frac{1}{(1-R^2_j)}*\frac{SER^2}{n \times var[X_j]}$$

- VIF from multicollinearity: $\frac{1}{(1-R^2_j)}$
    - $R_j^2$ for auxiliary regression of $X_j$ on all other $X$'s
    - mutlicollinearity does not bias $\hat{\beta_j}$ but raises its variance 
    - *perfect* multicollinearity if $X$'s are linear function of others

---

class: inverse, center, middle

# Updated Measures of Fit

---

# (Updated) Measures of Fit

- Again, how well does a linear model fit the data?

- How much variation in $Y_i$ is "explained" by variation in the model ($\hat{Y_i}$)?

--

$$\begin{align*}
Y_i&=\hat{Y_i}+\hat{u_i}\\
\hat{u_i}&= Y_i-\hat{Y_i}\\
\end{align*}$$

---

# (Updated) Measures of Fit: SER

- Again, the .shout[Standard errror of the regression (SER)] estimates the standard error of $u$

$$SER=\frac{SSE}{n-\mathbf{k}-1}$$

- A measure of the spread of the observations around the regression line (in units of $Y$), the average "size" of the residual

- **Only new change:** divided by $n-\mathbf{k}-1$ due to use of $k+1$ degrees of freedom to first estimate $\beta_0$ and then all of the other $\beta$'s for the $k$ number of regressors<sup>.red[1]</sup>

.footnote[<sup>.red[1]</sup> Again, because your textbook defines $k$ as including the constant, the denominator would be $n-k$ instead of $n-k-1$.]

---

# (Updated) Measures of Fit: $R^2$

$$\begin{align*}
R^2&=\frac{ESS}{TSS}\\
&=1-\frac{SSE}{TSS}\\
&=(r_{X,Y})^2 \\ \end{align*}$$

- Again, $R^2$ is fraction of variation of the model $(\hat{Y_i}$ ("explained SS") to the variation of observations of $Y_i$ ("total SS")

---

# (Updated) Measures of Fit: Adjusted $\bar{R}^2$

- Problem: $R^2$ of a regression increases *every* time a new variable is added (it reduces SSE!)

- This does *not* mean adding a variable improves the fit of the model per se, $R^2$ gets **inflated**

--

-  We correct for this effect with the .shout[adjusted `\\(R^2\\)`]:

$$\bar{R}^2 = 1 - \frac{n-1}{n-k-1} \times \frac{SSE}{TSS}$$

- There are different methods to compute $\bar{R}^2$, and in the end, recall $R^2$ **was never very useful**, so don't worry about knowing the formula
  - Large sample sizes $(n)$ make $R^2$ and $\bar{R}^2$ very close

---

# In R (base)

.pull-left[
```{r, echo=F}
summary(elreg)
```
]

.pull-right[
- Base $R^2$ (`R` calls it `multiple R-squared`) went up 
- `Adjusted R-squared` went down 
]

---

# In R (broom)

.pull-left[
```{r, echo=T}
elreg %>%
  glance()
```
]

### 2019-10-29 content/r/15-practice.Rmd
Add class 15 content

---
title: "3.3 OLS: Model Specification Strategies - R Practice"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

# Set Up
To minimize confusion, I suggest creating a new `R Project` (e.g. `regression_practice`) and storing any data in that folder on your computer.

Alternatively, I have made a project in R Studio Cloud that you can use (and not worry about trading room computer limitations), with the data already inside (you will still need to assign it to an object).

- [<i class="fas fa-cloud"></i> View Project on R Studio Cloud](https://rstudio.cloud/project/662906)
- [<i class="fab fa-markdown"></i> `Download Markdown`](/r/15-r-practice-rmd.Rmd)

## Question 1

Download and read in (`read_csv`) the data below.

- [<i class="fas fa-table"></i> `congress-trump-score.csv`](/data/speeding-tickets.csv)
This data comes from a paper by Makowsky and Strattman (2009) that we will examine later. Even though state law sets a formula for tickets based on how fast a person was driving, police officers in practice often deviate from that formula. This dataset includes information on all traffic stops. An amount for the fine is given only for observations in which the police officer decided to assess a fine. There are a number of variables in this dataset, but the one's we'll look at are:

| Variable | Description |
|----------|-------------|
| `Amount` | Amount of fine (in dollars) assessed for speeding |
| `Age` | Age of speeding driver (in years) |
| `MPHover` | Miles per hour over the speed limit |

We want to explore who gets fines, and how much. We'll come back to the other variables (which are categorical) in this dataset in later lessons.

## Question 2
*How does the age of a driver affect the amount of the fine*? Make a scatterplot of the `Amount` of the fine and the driver's `Age`. Add a regression line with an additional layer of `geom_smooth(method="lm")`.

## Question 3
Find the correlation between `Amount` and `Age`.

## Question 4
We weant to predict the following model:

$$\widehat{\text{Amount}_i}= \hat{\beta_0}+\hat{\beta_1}\text{Age}_i$$

Run a regression, and save it as an object. Now get a `summary()` of it.

### Part A
Write out the estimated regression equation.

### Part B
What is $\hat{\beta_0}$? What does it mean in the context of our question?

### Part C
What is $\hat{\beta_1}$? What does it mean in the context of our question?

### Part D 
What is the marginal effect of age on amount?

## Question 5
Redo question 4 with the `broom` package. Try out `tidy()` and `glance()`. This is just to keep you versatile.

## Question 6
How big would the difference in expected fine be for two drivers, one 18 years old and one 40 years old?

## Question 7
Now run the regression again, controlling for speed (`MPHover`).

### Part A
Write the new regression equation.

### Part B
What is the marginal effect of `Age` on `Amount`? What happened to it?

### Part C
What is the marginal effect of `MPHover` on `Amount`?

### Part D
What is $\hat{\beta_0}$, and what does it mean?

### Part E
What is the adjusted $\bar{R}^2$? What does it mean?

## Question 8
Now suppose both the 18 year old and the 40 year old each went 10 MPH over the speed limit. How big would the difference in expected fine be for the two drivers?

## Question 9
How about the difference in expected fine between two 18 year olds, one who went 10 MPH over, and one who went 30 MPH over?

## Question 10
Use the `huxtable` package's `huxreg()` command to make a regression table of your two regressions: the one from question 4, and the one from question 7.

## Question 11
Are our two independent variables multicollinear? Do younger people tend to drive faster?

### Part A
Get the correlation between `Age` and `MPHover`.

### Part B
Make a scatterplot of `MPHover` on `Age`.

### Part C
Run an auxiliary regression of `MPHover` on `Age`.

### Part D
Interpret the coefficient on `Age`.

### Part E
Look at your regression table in question 10. What happened to the standard error on `Age`? Why (consider the formula for variation in $\hat{\beta_1})$

### Part F
Calculate the Variance Inflation Factor (VIF) using the `car` package's `vif()` command.^[Run it on your multivariate regression object from Question 7.]

### Part G
Calculate the VIF manually, using what you learned in this question.

## Question 12
Let's now think about the omitted variable bias. Suppose the "true" model is the one we ran from Question 7.

### Part A
Do you suppose that `MPHover` fits the two criteria for omitted variable bias?

### Part B
Look at the regression we ran in Question 4. Consider this the "omitted" regression, where we left out `MPHover`. Does our estimate of the marginal effect of `Age` on `Amount` overstate or understate the *true* marginal effect?

### Part C
Use the "true" model (Question 7), the "omitted" regression (Question 4), and our "auxiliary" regression (Question 11) to identify each of the following parameters that describe our biased estimate of the marginal effect of `Age` on `Amount`:^[See the notation from class 3.2.]
$$\alpha_1=\beta_1+\beta_2\delta1$$

### Part D
From your answer in part C, how large is the omitted variable bias from leaving out `MPHover`?

## Question 13
Make a coefficient plot of your coefficients from the regression in Question 7.

### 2019-10-29 static/r/15-r-practice-rmd.Rmd
Add class 15 content

--- 2019-10-29 content/r/15-practice.Rmd
@@
 ---
-title: "3.3 OLS: Model Specification Strategies - R Practice"
-editor_options: 
-  chunk_output_type: console
+title: "3.3 OLS: OLS: Model Specification Strategies - R Practice"
+author: "YOUR NAME"
+date: "`r Sys.Date()`"
+output:
+  html_document:
+    df_print: paged
+    #theme: 
+    toc: true 
+    toc_depth: 3
+    toc_float: true
+    code_folding: show
+    highlight: tango
 ---
 
-```{r message=FALSE, warning=FALSE, include=FALSE}
-library(tidyverse)
-library(pander)
-source(file.path(here::here(), "lib", "graphics.R"))
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
 ```
@@
 ## Question 1
-
 Download and read in (`read_csv`) the data below.
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 2
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 3
 Find the correlation between `Amount` and `Age`.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part A
+
 Write out the estimated regression equation.
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part B
+
 What is $\hat{\beta_0}$? What does it mean in the context of our question?
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part C
+
 What is $\hat{\beta_1}$? What does it mean in the context of our question?
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part D
+
 What is the marginal effect of age on amount?
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
+## Question 5
+Redo question 4 with the `broom` package. Try out `tidy()` and `glance()`. This is just to keep you versatile.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 6
+
 How big would the difference in expected fine be for two drivers, one 18 years old and one 40 years old?
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 7
+
 Now run the regression again, controlling for speed (`MPHover`).
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part B
 What is the marginal effect of `Age` on `Amount`? What happened to it?
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part D
 What is $\hat{\beta_0}$, and what does it mean?
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 8
+
 Now suppose both the 18 year old and the 40 year old each went 10 MPH over the speed limit. How big would the difference in expected fine be for the two drivers?
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 9
+
 How about the difference in expected fine between two 18 year olds, one who went 10 MPH over, and one who went 30 MPH over?
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 10
+
 Use the `huxtable` package's `huxreg()` command to make a regression table of your two regressions: the one from question 4, and the one from question 7.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part B
 Make a scatterplot of `MPHover` on `Age`.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part D
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part E
 Look at your regression table in question 10. What happened to the standard error on `Age`? Why (consider the formula for variation in $\hat{\beta_1})$
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part G
 Calculate the VIF manually, using what you learned in this question.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part B
+
 Look at the regression we ran in Question 4.
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part C
+
 Use the "true" model (Question 7), the "omitted" regression (Question 4), and our "auxiliary" regression (Question 11) to identify each of the following parameters that describe our biased estimate of the marginal effect of `Age` on `Amount`:^[See the notation from class 3.2.]
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part D
+
 From your answer in part C, how large is the omitted variable bias from leaving out `MPHover`?
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 Make a coefficient plot of your coefficients from the regression in Question 7.
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---

### 2019-10-29 static/slides/15-slides.Rmd
Add class 15 content

---
title: "3.3: Model Specification Strategies"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))
```

```{r regression-setup, echo=F, results="hide"}

library(haven)
CASchool<-read_dta("../data/caschool.dta")

# run regression of testscr on str
school_reg <- lm(testscr ~ str, 
                 data = CASchool)

library(broom)
school_reg_tidy <- tidy(school_reg,
     conf.int = TRUE) # add confidence intervals
CASchool_aug <- augment(school_reg)

library(equatiomatic)
extract_eq(school_reg, use_coefs = TRUE,coef_digits = 2,fix_signs = TRUE)

```

# Model Specification

- The big challenge in applied econometrics is choosing how to **specify a model** to regress

- Every dataset is different, every study has a different goal
    - there is no bright line rule, only a set of guidelines and skills that you can only learn by doing!

- But here are some helpful tips and frequent problems (and solutions)

---

# Model Specification: Process

**1. Identify your question of interest:** what do you want to know? What marginal effect(s) do you want to estimate?

--

**2. Think about possible sources of endogeneity:** what *other* variables would cause  .shout[omitted variable bias] if we left them out? Can we get data on them too? 
  - Again: must BOTH (1) affect $Y$ AND (2) be correlated with $X$
  - This requires much of your economic intuitions: $R^2$ and statistical measures cannot tell you everything!

--

**3. Run multiple models and check the robustness of your results:** does the size (or direction) of your marginal effect(s) of interest change as you change your model (i.e. add more variables)?

--

**4. Interpret your results**
  - Are they statistically significant? 
  - Regardless of *statistical* significance, are they **economically meaningful**? 
      - Why should we care? 
      - How big is "big"?

---

# Proxy Variables

- Ideally, we would want a randomized control experiment to assign individuals to treatment

--

- But with observational data, $u_i$ depends on other factors
    - *If* we can observe and measure these factors, then include them in the regression

--

- If we can't directly measure them, often we can include variables *correlated* with these variables to .shout[proxy] for the effects of them!

---

# Proxy Variables: Example I

.content-box-green[
.green[**Example**]: Consider test scores and class sizes again. What about learning opportunities outside of school?
]

--

- Probably a bias-causing omitted variable (affects test score and correlated with class size) but we can't measure it!

--

- Suppose we *can* measure a variable $V$, and significantly, $corr(V,Z)\neq 0$
  - e.g. we have data on the percent of students who get a free or subsidized lunch (`meal_pct`)

--

- This is a good .shout[proxy] for income-determined learning opportunities outside of school
  - `meal_pct` is correlated with Income (which is likely correlated with both class size and test score)
  - A good *indirect* measure of Income

---

# Proxy Variables: Example II

- We assumed we don't have data on average district income, we would expect `meal_pct` to be strongly negatively correlated with income

--

- Just kidding, we do have data on `avginc`, but we'll only use it to confirm our suspicion:

--

```{r, echo=T}
CASchool %>%
  select(testscr, str, el_pct, avginc,meal_pct) %>%
  cor()
```

--

- We can see `meal_pct` is strongly (negatively) correlated with income, as expected

---

# Proxy Variables: Example III

```{r, echo=T}
proxyreg<-lm(testscr~str+el_pct+meal_pct, data=CASchool)
summary(proxyreg)
```

---

# Interpreting Control Variables

$$\widehat{\text{Test Score}}_i=700.15-1.00 \, STR_i-0.122 \, \%EL_i - 0.547 \, meal\%_i$$

--

- Is `meal%` causal?

--

- Getting rid of programs in districts where a large percentage of students need them would boost test scores A LOT! (So probably not causal...)

--

- `meal%` likely correlated with other things in $u_i$ (like *outside learning opportunities*!). 
    - In fact, that's *exactly why* we included it as a variable!

--

- .whisper[We don't need the OLS estimate on `meal%` to be unbiased!]
    - **We only care about getting the estimate on `str` to be unbiased!**

---

# Control Variables

- A .shout[control variable] is a regressor variable **note** of interest, but included to hold factors constant that, if omitted, would bias the causal effect of interest

--

- The control variable may still be correlated with omitted causal factors in $u$
    - Estimators $(\hat{\beta}$'s) on control variables can be biased and that is OK!
    - So long as we have unbiased estimators $(\hat{\beta}$'s) on the regressors we do care about!

--

- Control variables allow us to proceed *as if* $X$ were randomly assigned

---

# Coefficient Plots I

- Sometimes it is helpful to make a **coefficient plot** to visualize the different marginal effects

--

- Take `broom` and `tidy()` your regression with `conf.int=TRUE` (and save as an object)

```{r, echo=T}
# run regression
reg<-lm(testscr~str+el_pct+meal_pct, data=CASchool)

# load broom
library(broom)
#  tidy regression with confidence intervals
# save as "reg_tidy"
reg_tidy <- tidy(reg, conf.int = TRUE)
```

---

# Coefficient Plots II

```{r, echo=T}
# look at it
reg_tidy

# intercept is too big relative to others! let's leave it out
reg_tidy_no_int<-reg_tidy %>%
  filter(term!="(Intercept)")
```

---

# Coefficient Plots III

.pull-left[

```{r coef-plot, echo=T, eval=F}
ggplot(data = reg_tidy_no_int)+
  aes(x = estimate,
      y = term,
      color = term)+
    geom_point()+ # point for estimate
    # now make "error bars" using conf. int.
    geom_segment(aes(x = conf.low,
                     xend = conf.high,
                     y=term,
                     yend=term))+
  # add line at 0
  geom_vline(xintercept=0, size=1, color="black", linetype="dashed")+
  scale_x_continuous(breaks=seq(-1.5,0.5,0.5),
                     limits=c(-1.5,0.5))+
  labs(x = expression(paste("Marginal effect, ", hat(beta[j]))),
       y = "Variable")+
  guides(color=F)+ # hide legend
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

```{r, ref.label="coef-plot"}

```
]

---

# Model Specification Warning: $R^2$

.red[Do NOT just try to maximize `\\(R^2\\)` or `\\(\bar{R}^2\\)`]

- A high $R^2$ or $\bar{R}^2$ means that the regressors explain variation in $Y$

- A high $R^2$ or $\bar{R}^2$ does *NOT* mean you have eliminated omitted variable bias

- A high $R^2$ or $\bar{R}^2$ does *NOT* mean you have an unbiased estimate of a causal effect

- A high $R^2$ or $\bar{R}^2$ does *NOT* mean included variables are statistically significant

---

# Model Specification: Goal

- We care mostly about measuring a causal effect

- This involves making the respective $\hat{\beta_j}$'s as unbiased as possible (because $\beta_j$ is the causal effect of $X_j \rightarrow Y$!

- You will not get high $R^2$'s in social science
  - even if you do, you are only explaining *variation* in $Y$ with the model, not necessarily the *causal effect*!

### 2019-10-29 content/r/15-practice.Rmd
Tweaks to R practice 15

@@
 
-- [<i class="fas fa-table"></i> `congress-trump-score.csv`](/data/speeding-tickets.csv)
+- [<i class="fas fa-table"></i> `speeding_tickets.csv`](/data/speeding_tickets.csv)
 This data comes from a paper by Makowsky and Strattman (2009) that we will examine later.
@@
 Use the "true" model (Question 7), the "omitted" regression (Question 4), and our "auxiliary" regression (Question 11) to identify each of the following parameters that describe our biased estimate of the marginal effect of `Age` on `Amount`:^[See the notation from class 3.2.]
-$$\alpha_1=\beta_1+\beta_2\delta1$$
+$$\alpha_1=\beta_1+\beta_2\delta_1$$
 

### 2019-10-29 static/r/15-r-practice-rmd.Rmd
Tweaks to R practice 15

@@
 
-- [<i class="fas fa-table"></i> `congress-trump-score.csv`](/data/speeding-tickets.csv)
+- [<i class="fas fa-table"></i> `speeding_tickets.csv`](/data/speeding-tickets.csv)
 This data comes from a paper by Makowsky and Strattman (2009) that we will examine later.

### 2019-10-29 static/r/15-r-practice-rmd.Rmd
Quick format tweak to practice 15 rmd

@@
 ### Part E
+
 Look at your regression table in question 10. What happened to the standard error on `Age`? Why (consider the formula for variation in $\hat{\beta_1})$
@@
 ### Part F
+
 Calculate the Variance Inflation Factor (VIF) using the `car` package's `vif()` command.^[Run it on your multivariate regression object from Question 7.]

### 2019-10-31 content/class/15-class.Rmd
Add class 16 content

---
title: "3.3: Model Specification Strategies - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

*Tuesday, October 29, 2019*

# Overview

Today we will simply talk about the process of specifying a model: how to pick $X$ variables to add to your regression. Recall, they should fit the two criteria for omitted variable bias. One additional issue we discuss is the use of **proxy** variables: where you know omitting a variable would cause bias, but you don't have data on it, just data on a variable *related* to it; and of **control** variables (including proxies), where the OLS estimator $(\hat{\beta_j})$ for an independent variable is *not causal*, but just helps remove bias on the independent variables you *do* care about being causal.

For the rest of class, we will work on practice problems for multivariate regression.

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/15-slides.html)

# Practice Problems

Today you will be working on [R practice problems](/r/15-r-practice.html). Check back later for solutions.

### 2019-10-31 content/class/16-class.Rmd
Add class 16 content

---
title: "3.4: Multivariate Regression Examples - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

*Thursday, October 31, 2019*

# Overview

Today we will continue working on practice problems from last class. Afterwards, we will explore some actual empirical economics papers that use (relatively) simple multivariate regressions. See them in today's [Readings page](/readings/16-reading).

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/16-slides.html)

# Practice Problems

Today you will finish working on [R practice problems](/r/15-r-practice.html) from last class. Check back later for solutions.

### 2019-10-31 content/reading/16-reading.Rmd
Add class 16 content

---
title: "3.4 Multivariate Regression Examples - Readings"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-fullnote-bibliography-no-bib.csl
---

# Primary Sources

- [<i class="fas fa-file-pdf"></i> Sacerdote (2001), "Peer Effects with Random Assignment: Results from Darmouth Roommates"](/readings/Sacerdote-2001.pdf)
- [<i class="fas fa-file-pdf"></i> Duggan and Levitt (2002), "Winning isn't Everything: Corruption in Sumo Wrestling"](/readings/Duggan-Levitt-2002.pdf)
- [<i class="fas fa-file-pdf"></i> Fisman and Miguel (2007), "Corruption, Norms, and Legal Enforcement: Evidence from Diplomatic Parking Tickets"](/readings/Fisman-Miguel-2007.pdf)

### 2019-10-31 static/slides/16-slides.Rmd
Add class 16 content

---
title: "3.4: Examples"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))
```

class: inverse, center, middle

# Sacerdote (2001) on Peer Effects

---

# Paper Motivation I

.pull-left[

- What determines student outcomes in college? (GPAs, fraternity enrollment, alcohol/drug use, etc)

- Effects of peer groups

]

.pull-right[
.center[
![](https://www.dropbox.com/s/bdy9gn83dgbc1s3/universitypeers.jpg?raw=1)
]
]
.source[Sacerdote, Bruce, (2001), "Peer Effects with Random Assignment: Results from Dartmouth Roommates" *Quarterly Journal of Economics* 116(2):681-704]

---

# Paper Motivation II

.pull-left[

- "Standard" way to estimate peer effects: regress student $i$'s outcomes/behavior on *other students'* outcomes/behavior

$$GPA_i = \beta_0+\beta_1 OwnBehavior_i + \beta_2 RoommateBehavior_i + u_i$$

]

.pull-right[
.center[
![](https://www.dropbox.com/s/bdy9gn83dgbc1s3/universitypeers.jpg?raw=1)
]
]
.source[Sacerdote, Bruce, (2001), "Peer Effects with Random Assignment: Results from Dartmouth Roommates" *Quarterly Journal of Economics* 116(2):681-704]

---

# Paper Motivation II

.pull-left[

- "Standard" way to estimate peer effects: regress student $i$'s outcomes/behavior on *other students'* outcomes/behavior

$$GPA_i = \beta_0+\beta_1 OwnBehavior_i + \beta_2 RoommateBehavior_i + u_i$$

- Problems with this approach: 
    1. Individuals **self-select** into peer groups
    2. If two roommates $A$ and $B$ influence each other, how do we disentangle causal effect of $B \rightarrow A$ vs. $A \rightarrow B$?
    3. Are peer effects actually driven by students' own backgrounds, or by their actual choices?

]

.pull-right[
.center[
![](https://www.dropbox.com/s/bdy9gn83dgbc1s3/universitypeers.jpg?raw=1)
]
]
.source[Sacerdote, Bruce, (2001), "Peer Effects with Random Assignment: Results from Dartmouth Roommates" *Quarterly Journal of Economics* 116(2):681-704]

---

# Paper Motivation II

.pull-left[

$$\begin{align*}
			corr(OwnBehavior, u) &\neq 0	\\
			corr(RoomateBehavior, u)& \neq 0 \\
			E[u|OwnBehavior,RoommateBehavior] & \neq 0 	\\
\end{align*}$$

]

.pull-right[
.center[
![](https://www.dropbox.com/s/bdy9gn83dgbc1s3/universitypeers.jpg?raw=1)
]
]
.source[Sacerdote, Bruce, (2001), "Peer Effects with Random Assignment: Results from Dartmouth Roommates" *Quarterly Journal of Economics* 116(2):681-704]

---

# Sacerdotes' Identification Strategy

.pull-left[

- Freshmen entering Dartmouth College are **randomly** assigned to dorms & roommates

- Removes self-selection of peer groups by shared characteristics

- **Random assignment**: roommate A's background characteristics are uncorrelated with roommate B's background characteristics

]

.pull-right[
.center[
![](https://www.dropbox.com/s/bdy9gn83dgbc1s3/universitypeers.jpg?raw=1)
]
]
.source[Sacerdote, Bruce, (2001), "Peer Effects with Random Assignment: Results from Dartmouth Roommates" *Quarterly Journal of Economics* 116(2):681-704]

---

# Relevant Institutions

- Freshmen entering Dartmouth College are randomly assigned to dorms & roommates
- Each incoming freshman fills out a questionnaire:
    1. I smoke. (Y/N)
    2. I like to listen to music while studying. (Y/N)
    3. I keep late hours. (Y/N)
    4. I am more neat than messy. (Y/N)
    5. I am (Male/Female). 	
- There are $2^5=32$ combinatorial possibilities of answers to the questions
- Students are assigned to roommates/dorms **at random**, conditional on their combination of answers to the 5 survey answers

---

# The Data

- Data from Dartmouth's database of students: history of dorm assignments & term-by-term academic performance

- Data on pre-treatment characteristics (SAT scores, high school class rank, private/public HS, home state, academic index)

- Outcome variables: GPA, time to graduation, frat membership, major choice, participation in athletics

- Survey of Incoming Freshman: if student drank beer in last year and expectation of graduating with honors

- Sample of 1589 students

- Create dummy variable for each block to control for covariates (we'll talk later about **dummy variables** and **fixed effects** like this)

---

# The Data: Summary Statistics I

.center[
![](https://www.dropbox.com/s/3fuhj7kbz0ptnnf/sacerdote1.png?raw=1)
]

---

# The Data: Summary Statistics II

.center[
![](https://www.dropbox.com/s/uhfupj0mg9p8m0h/sacerdote2.png?raw=1)
]

---

# The Data: Demonstrating Random Assignment

.center[
![](https://www.dropbox.com/s/bfw0v2wqnqw3j89/sacerdote3.png?raw=1)
]

---

# The Data: Demonstrating Random Assignment

.center[
![](https://www.dropbox.com/s/ull5y5x691hn54r/sacerdote3c.png?raw=1)
]

---

# Sacerdote's Empirical Model (Basically)

$$GPA_i=\beta_0+\beta_1ACA_i+\beta_2ACA_j+u_i$$

- Student $i$ and roommate $j$

- $ACA$: Index of academic performance (broken down into different metrics)

- Other outcomes of interest (besides $GPA$: graduation, major, fraternity, athlete

---

# Regression Results

.center[
![](https://www.dropbox.com/s/jwpt4n37qq0fr3h/sacerdote5.png?raw=1)
]

--

- For every 1 point increase (decrease) in your *roommate*'s GPA, *your* GPA increases (decreases) about 0.12 points

--

- If you would have been a 3.0 student with a 3.0 roommate, but you were assigned to a 2.0 roommate, your GPA would be 2.88

---

# Regression Results: Academic Performance

.center[
![](https://www.dropbox.com/s/d2dnjjxbivd9e8a/sacerdote6.png?raw=1)
]

---

# Regression Results: Social Outcomes

.center[
![](https://www.dropbox.com/s/1p0jdtio7qjxbc1/sacerdote9.png?raw=1)
]

---

# Conclusions

.pull-left[

- Peer effects are very strong!

- Important influences in Freshman year performance (GPA) and activities (joining a social organization)

- Not important for choosing a major

]

.pull-right[
.center[
![](https://www.dropbox.com/s/bdy9gn83dgbc1s3/universitypeers.jpg?raw=1)
]
]
.source[Sacerdote, Bruce, (2001), "Peer Effects with Random Assignment: Results from Dartmouth Roommates" *Quarterly Journal of Economics* 116(2):681-704]

---

.class: inverse, center, middle
# Duggan and Levitt (2002) on Corruption in Sumo Wrestling

---

# Paper Motivation I

.pull-left[

- 2000 year history, national sport of Japan, extremely ritualistic

- Japan is a country with low corruption (CPI: 75, Rank 18th best)

- Good data available

- Situation is ripe for cheating! So when/why does it happen?

]

.pull-right[
.center[
![](https://www.dropbox.com/s/iu3ua5wbh2ulv0d/sumo.jpg?raw=1)
]
]
.source[Duggan, Mark and Steven D. Levitt, (2002), "Winning isn't Everything: Corruption in Sumo Wrestling" *American Economic Review* 92(5):1594-1605]

---

# Relevant Institutions I

.pull-left[

- Tournaments (*basho*), 66 wrestlers (*rikishi*), 15 bouts each

- Wrestlers with 8+ wins (*kachi-koshi*) *move up* in rankings (*banzuke*)

- Those with a losing record ($<$8 wins) (*maki-koshi*) *fall* in rankings

]

.pull-right[
.center[
![](https://www.dropbox.com/s/iu3ua5wbh2ulv0d/sumo.jpg?raw=1)
]
]
.source[Duggan, Mark and Steven D. Levitt, (2002), "Winning isn't Everything: Corruption in Sumo Wrestling" *American Economic Review* 92(5):1594-1605]

---

# Relevant Institutions II

.pull-left[

- A marginal win generates a 2.5 rank increase

- But movement from 7 to 8 wins produces almost an 11 rank increase!

- Rank signals prestige, moving up a single rank is worth about $3,000/year

- Top 5th-10th ranked wrestlers make $250,000/year

]

.pull-right[
.center[
![](https://www.dropbox.com/s/23hzngfz77h7rtb/sumo1.png?raw=1)
]
]
.source[Duggan, Mark and Steven D. Levitt, (2002), "Winning isn't Everything: Corruption in Sumo Wrestling" *American Economic Review* 92(5):1594-1605]

---

# Relevant Institutions III

.pull-left[

- Consider 2 wrestlers: .red[A (8-6)] vs .blue[B (7-7)] going into final (15th) match

- Return to winning for .blue[B (7-7)] is much higher than for .red[A (8-6)]

- .red[A (8-6)] throws the match to .blue[B (7-7)], who must return the favor in later tournaments if .red[A] finds himself in the same 7-7 position

]

.pull-right[
.center[
![](https://www.dropbox.com/s/iu3ua5wbh2ulv0d/sumo.jpg?raw=1)
]
]
.source[Duggan, Mark and Steven D. Levitt, (2002), "Winning isn't Everything: Corruption in Sumo Wrestling" *American Economic Review* 92(5):1594-1605]

---

# The Data

- All official top-rank sumo matches from January 1989-January 2000

- Six tournaments per year, nearly 70 wrestlers per tournament

- Tournaments last 15 days with one match per wrestler

- 64,000 wrestler-matches

---

# The Theoretical vs. Actual Probability of Winning

.pull-left[
- Theoretical (binomial) probability of winning 8 times: 19.6%

- Actual probability (from data): 26%

- Much higher probability for 8 wins than it should be! (& lower for 7)
  - Could this be from rampant cheating!?

]

.pull-right[
.center[
![](https://www.dropbox.com/s/c8tf3yx6thuil89/sumo2.png?raw=1)
]
]

---

# The Model (Slightly Modified)

$$\text{Win}_{ijtd}=\beta_1 Bubble_{ijtd}+ \beta_2 Rankdiff_{ijt}+\lambda_{ij}+\delta_{it}+u_{ijtd}$$

- $Win=1$ if wrestler $i$ beats wrestler $j$ in tournament $t$ on day $d$

- $Bubble=1$ if wrestler ($i$) is on margin (7-7), -1 if opponent ($j$) is on margin, $=0$ if neither are on margin

- $Rankdiff$: difference in rank between wrestlers

- Wrestler $\lambda$ and tournament $\delta$ fixed effects

---

# Initial Results

.center[
![](https://www.dropbox.com/s/bamwd6eq97rugou/sumo4.png?raw=1)

- Frequency of rigging increases as tournament nears end (day 15)

- On day 15, 7-7 wrestlers on margin win 25% more often than they otherwise should
]

---

# Interpretation of Initial Results

- Two alternative hypotheses to explain results:
  1. Match rigging (corruption)
  2. Effort: wrestlers on margin (7-7) just fight harder! Wrestlers with 8 wins are more complacent (already made it)

- To test, look for evidence of **reciprocity** agreements over time 
  - If these tacit agreements to rig matches exist, wrestlers from stable A should have very high win rates when on the margin against wrestlers from stable B, and vice versa

---

# Interaction Effects

.pull-left[
.center[
![](https://www.dropbox.com/s/qnh3i0pwogiz1y6/sumo5.png?raw=1)
]
]

.pull-right[
- Last row (before $R^2$): wrestler's success strongly increases with overall success rates of playing wrestlers on the bubble from other stables

- For each 10% increase in success in other bubble matches between these two stables, the wrestler on the bubble is 2.7% more likely to win
]

---

# Reciprocity?

.center[
![](https://www.dropbox.com/s/z2l381t0yqaykki/sumo6.png?raw=1)
]

- Wrestlers who won a bubble match previously tend to do worse when playing same opponent - throw the match to them in reciprocity!

---

class: inverse, center, middle
# Fisman and Miguel (2007)

---

# Paper Motivation

.pull-left[

- What determines the level of corruption?
    1. Legal environment of country
    2. Culture or social norms

- How to identify the true source(s)?

]

.pull-right[
.center[
![](https://www.dropbox.com/s/92fk22dz2zctozu/rentseeking.png?raw=1)
]
]
.source[Fisman, Raymond and Edward Miguel, (2007), "Corruption, Norms, and Legal Enforcement: Evidence from Diplomatic Parking Tickets," *Journal of Political Economy* 115(6): 1020-1048]

---

# Relevant Institutions

.pull-left[

- U.N. Diplomats are given immunity from prosecution or lawsuits in the U.S.

- Reciprocal agreements with other countries, designed to protect diplomats in unfriendly environments

- Diplomatic license plates in NYC are identified, get ticket, but no way to enforce

- "The best free parking pass in town"

]

.pull-right[
.center[
![](https://www.dropbox.com/s/qk19adm49gy5plj/parkingticket.jpg?raw=1)

![](https://www.dropbox.com/s/qtgg9hkmpecm4dw/diplomatplates.jpg?raw=1)
]
]

.source[Fisman, Raymond and Edward Miguel, (2007), "Corruption, Norms, and Legal Enforcement: Evidence from Diplomatic Parking Tickets," *Journal of Political Economy* 115(6): 1020-1048]

---

# The Data

- Between 11/1987 and 12/2002, 150,000 unpaid parking tickets, fines totaling $18,000,000

- 30 Days to pay a fine, afterwards a 110% penalty. After 70 days, recorded as unpaid violation

- Individual violation-level data: license plate, name, country of origin, date & time of violation, fine, amount paid (if any)

- 43% were violations of "no standing/loading zone"

- 20% of cases, the car was registered to the diplomatic mission (not personal)

- Scale fines by the size of the country's mission

---

# Identification Strategy I

- Becker's (1968) rational crime model says with no punishment$\implies$ rational for all diplomats to never pay parking fines

- But large variation in data! Unpaid fines are strongly correlated with country's score on corruption index!

- Home country corruption norms are an important predictor of diplomats breaking the law
    - Low corruption countries' diplomats tend to pay the fine even where they are not legally compelled to
    - High corruption countries' diplomats rack up massive fines

---

# Identification Strategy II

.pull-left[
- **Natural experiment**: post-9/11, NYC began cracking down on enforcement

- Diplomats with 3+ unpaid parking tickets had diplomat plates revoked

- Led to immediate 98% reduction in unpaid parking tickets

- So enforcement matters as well as corruption norms
]

.pull-right[
.center[
![](https://www.dropbox.com/s/xe59fx3ced0cva9/parking1.png?raw=1)
]
]
---

# Data I

.center[
![](https://www.dropbox.com/s/5ndr02pwjvv6zr3/parking2.png?raw=1)
]

---

# Data II

.center[
![](https://www.dropbox.com/s/izkvkpu2izi78sr/parking3.png?raw=1)
]

---

# Empirical Model

$$\text{Unpaid Violations}=\beta_0+\beta_1\text{Corruption}+\beta_2\text{Enforcement}+\beta_3\text{Diplomats}+...+\beta_k\text{Controls}$$

---

# Results I

.center[
![](https://www.dropbox.com/s/rf6r3g66u3aaulr/parking4.png?raw=1)
]

---
# Results II

.center[
![](https://www.dropbox.com/s/2drdlc105j2i15m/parking7.png?raw=1)
]

---
# Results III

.center[
![](https://www.dropbox.com/s/t1s9e3xb3kkn428/parking5.png?raw=1)
]

### 2019-10-31 static/slides/16-slides.Rmd
Tweak slides 16 formatting

@@
 .center[
-![](https://www.dropbox.com/s/bfw0v2wqnqw3j89/sacerdote3.png?raw=1)
+![:scale 40%](https://www.dropbox.com/s/bfw0v2wqnqw3j89/sacerdote3.png?raw=1)
 ]
@@
 .center[
-![](https://www.dropbox.com/s/ull5y5x691hn54r/sacerdote3c.png?raw=1)
+![:scale 40%](https://www.dropbox.com/s/ull5y5x691hn54r/sacerdote3c.png?raw=1)
 ]
@@
 
-.class: inverse, center, middle
+class: inverse, center, middle
 # Duggan and Levitt (2002) on Corruption in Sumo Wrestling
@@
 ]
+
 .source[Duggan, Mark and Steven D. Levitt, (2002), "Winning isn't Everything: Corruption in Sumo Wrestling" *American Economic Review* 92(5):1594-1605]
@@
 .center[
-![](https://www.dropbox.com/s/bamwd6eq97rugou/sumo4.png?raw=1)
+![:scale 70%](https://www.dropbox.com/s/bamwd6eq97rugou/sumo4.png?raw=1)
 
@@
 .center[
-![](https://www.dropbox.com/s/qnh3i0pwogiz1y6/sumo5.png?raw=1)
+![:scale 60%](https://www.dropbox.com/s/qnh3i0pwogiz1y6/sumo5.png?raw=1)
 ]
@@
 .center[
-![](https://www.dropbox.com/s/z2l381t0yqaykki/sumo6.png?raw=1)
+![:scale 60%](https://www.dropbox.com/s/z2l381t0yqaykki/sumo6.png?raw=1)
 ]
@@
 ]
+
 ]
+
 .source[Fisman, Raymond and Edward Miguel, (2007), "Corruption, Norms, and Legal Enforcement: Evidence from Diplomatic Parking Tickets," *Journal of Political Economy* 115(6): 1020-1048]
@@
 .center[
-![](https://www.dropbox.com/s/5ndr02pwjvv6zr3/parking2.png?raw=1)
+![:scale 70%](https://www.dropbox.com/s/5ndr02pwjvv6zr3/parking2.png?raw=1)
 ]
@@
 .center[
-![](https://www.dropbox.com/s/izkvkpu2izi78sr/parking3.png?raw=1)
+![:scale 50%](https://www.dropbox.com/s/izkvkpu2izi78sr/parking3.png?raw=1)
 ]
@@
 .center[
-![](https://www.dropbox.com/s/rf6r3g66u3aaulr/parking4.png?raw=1)
+![:scale 40%](https://www.dropbox.com/s/rf6r3g66u3aaulr/parking4.png?raw=1)
 ]
@@
 .center[
-![](https://www.dropbox.com/s/2drdlc105j2i15m/parking7.png?raw=1)
+![:scale 70%](https://www.dropbox.com/s/2drdlc105j2i15m/parking7.png?raw=1)
 ]
@@
 .center[
-![](https://www.dropbox.com/s/t1s9e3xb3kkn428/parking5.png?raw=1)
+![:scale 70%](https://www.dropbox.com/s/t1s9e3xb3kkn428/parking5.png?raw=1)
 ]

### 2019-10-31 static/slides/16-slides.Rmd
Tweak slides 16 formatting

@@
 ---
-title: "3.4: Examples"
+title: "3.4: Multivariate Regression Examples"
 subtitle: "ECON 480 · Econometrics · Fall 2019"

### 2019-11-01 static/r/15-r-practice-answers.Rmd
Add R practice 15 answers

---
title: "3.3 Model Specification Strategies - R Practice (Answers)"
author: "Ryan Safner"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    #theme: 
    toc: true 
    toc_depth: 3
    toc_float: true
    code_folding: show
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set Up
To minimize confusion, I suggest creating a new `R Project` (e.g. `regression_practice`) and storing any data in that folder on your computer.

Alternatively, I have made a project in R Studio Cloud that you can use (and not worry about trading room computer limitations), with the data already inside (you will still need to assign it to an object).

- [<i class="fas fa-cloud"></i> View Project on R Studio Cloud](https://rstudio.cloud/project/662906)
- [<i class="fab fa-markdown"></i> `Download Markdown`](/r/15-r-practice-rmd.Rmd)

## Question 1
Download and read in (`read_csv`) the data below.

- [<i class="fas fa-table"></i> `speeding_tickets.csv`](/data/speeding-tickets.csv)
This data comes from a paper by Makowsky and Strattman (2009) that we will examine later. Even though state law sets a formula for tickets based on how fast a person was driving, police officers in practice often deviate from that formula. This dataset includes information on all traffic stops. An amount for the fine is given only for observations in which the police officer decided to assess a fine. There are a number of variables in this dataset, but the one's we'll look at are:

| Variable | Description |
|----------|-------------|
| `Amount` | Amount of fine (in dollars) assessed for speeding |
| `Age` | Age of speeding driver (in years) |
| `MPHover` | Miles per hour over the speed limit |

We want to explore who gets fines, and how much. We'll come back to the other variables (which are categorical) in this dataset in later lessons.

---

<!--ANSWER BELOW HERE-->

```{r}
library(tidyverse)
speed<-read_csv("../data/speeding_tickets.csv")
```

---

## Question 2
*How does the age of a driver affect the amount of the fine*? Make a scatterplot of the `Amount` of the fine and the driver's `Age`. Add a regression line with an additional layer of `geom_smooth(method="lm")`.

---

<!--ANSWER BELOW HERE-->

```{r}
ggplot(data = speed)+
  aes(x = Age,
      y = Amount)+
  geom_point()+
  geom_smooth(method = "lm")
```

---

## Question 3
Find the correlation between `Amount` and `Age`.

---

<!--ANSWER BELOW HERE-->

```{r}
# get summary of Amount
speed %>%
  select(Amount) %>%
  summary() # see there are 36,683 NA's!

# so we need to get correlation by dropping the NA's

# method 1: filter by non-NA observations for Amount
speed %>%
  select(Age, Amount) %>%
  filter(!is.na(Amount)) %>%
  cor()

# method 2: tell cor() to use "pairwise.complete.obs"
speed %>%
  select(Age, Amount) %>%
  cor(., use = "pairwise.complete.obs")

# note again I am using the tidyverse method of getting correlation

# you can also use summarize():
speed %>% summarize(my_cor = cor(Age, Amount, use = "pairwise.complete.obs"))

# or the Base R method:
cor(speed$Age, speed$Amount, use = "pairwise.complete.obs")
```

---

## Question 4
We want to predict the following model:

$$\widehat{\text{Amount}_i}= \hat{\beta_0}+\hat{\beta_1}\text{Age}_i$$

Run a regression, and save it as an object. Now get a `summary()` of it.

---

<!--ANSWER BELOW HERE-->

```{r}
reg1<-lm(Amount~Age, data = speed)
summary(reg1)
```

---

### Part A

Write out the estimated regression equation.

---

<!--ANSWER BELOW HERE-->

$$\widehat{\text{Amount}_i}=131.71-0.29 \, \text{Age}_i$$
```{r}
# if you want to use equatiomatic
library(equatiomatic)
extract_eq(reg1, use_coefs = TRUE, fix_signs = TRUE, coef_digits = 2)
```

---

### Part B

What is $\hat{\beta_0}$? What does it mean in the context of our question?

---

$\hat{\beta_0}=131.71$. It means when `Age` is 0, the fine `Amount` is expected to be $131.71.

---

### Part C

What is $\hat{\beta_1}$? What does it mean in the context of our question?

---

$\hat{\beta_1}=-0.29$. It means for every additional year old someone is, their expected `Amount` of the fine decreases by $0.29.

---

### Part D

What is the marginal effect of age on amount?

---

The marginal effect is $\hat{\beta_1}$, described in the previous part. Just checking to make sure you know it's the marginal effect!

---

## Question 5
Redo question 4 with the `broom` package. Try out `tidy()` and `glance()`. This is just to keep you versatile.

---

<!--ANSWER BELOW HERE-->

```{r}
library(broom)
# tidy() to get coefficients
reg1_tidy<-tidy(reg1)

reg1_tidy

# glance() at original reg for measures of fit
glance(reg1)
```

---

## Question 6

How big would the difference in expected fine be for two drivers, one 18 years old and one 40 years old?

---

<!--ANSWER BELOW HERE-->

18-year-old driver:

$$\begin{align*}
\widehat{Amount}_i &= \hat{\beta_0}+\hat{\beta_1}Age_i\\
&= 131.71-0.29(18)\\
&=126.49\\
\end{align*}$$

40-year-old driver:

$$\begin{align*}
\widehat{Amount}_i &= \hat{\beta_0}+\hat{\beta_1}Age_i\\
&= 131.71-0.29(40)\\
&=120.11\\
\end{align*}$$

The difference is $126.49-120.11=6.38$ only!

```{r}
# we can use R as a calculator and use the regression coefficients

# extract beta0 and save it
beta_0<-reg1_tidy %>%
  filter(term=="(Intercept)") %>%
  select(estimate)

# extract beta1 and save it
beta_1<-reg1_tidy %>%
  filter(term=="Age") %>%
  select(estimate)

# predicted amount for 18 year old
amount_18yr<-beta_0+beta_1*18

# predicted amount for 40 year old
amount_40yr<-beta_0+beta_1*40

# difference
amount_18yr-amount_40yr
```

---

## Question 7

Now run the regression again, controlling for speed (`MPHover`).

<!--ANSWER BELOW HERE-->

```{r}
reg2<-lm(Amount~Age+MPHover, data=speed)
summary(reg2)
```

---

### Part A
Write the new regression equation.

---

<!--ANSWER BELOW HERE-->

$$\widehat{\text{Amount}_i}=3.49+0.02 \, \text{Age}_i+6.89 \, \text{MPHover}_i$$

```{r}
# if you want to use equatiomatic
extract_eq(reg2, use_coefs = TRUE, fix_signs = TRUE, coef_digits = 2)
```

---

### Part B
What is the marginal effect of `Age` on `Amount`? What happened to it?

---

<!--ANSWER BELOW HERE-->

$\hat{\beta_1}=0.02$. For every additional year old someone is, *holding their speed constant*, their expected fine increases by $0.02. The magnitude of the effect shrank and it switched from negative to positive!

---

### Part C
What is the marginal effect of `MPHover` on `Amount`?

---

<!--ANSWER BELOW HERE-->

$\hat{\beta_2}=6.89$. For every additional MPH someone drives above the speed limit, *holding their Age constant*, their expected fine increases by $6.89

---

### Part D
What is $\hat{\beta_0}$, and what does it mean?

---

<!--ANSWER BELOW HERE-->

$\hat{\beta_0}=3.49$. This is the expected fine for a driver of `Age` $=0$ and `MPHover` $=0$.

---

### Part E
What is the adjusted $\bar{R}^2$? What does it mean?

---

It is $0.5026$. We can explain 50% of the variation in `Amount` from variation in our model.

---

## Question 8

Now suppose both the 18 year old and the 40 year old each went 10 MPH over the speed limit. How big would the difference in expected fine be for the two drivers?

---

<!--ANSWER BELOW HERE-->

18-year-old driver:

$$\begin{align*}
\widehat{Amount}_i &= \hat{\beta_0}+\hat{\beta_1}Age_i+\hat{\beta_2}MPHover_i\\
&= 3.49+0.02(18)+6.89(10)\\
&=72.75\\
\end{align*}$$

40-year-old driver:

$$\begin{align*}
\widehat{Amount}_i &= \hat{\beta_0}+\hat{\beta_1}Age_i+\hat{\beta_2}MPHover_i\\
&= 3.49+0.02(40)+6.89(10)\\
&=73.19\\
\end{align*}$$

The difference is $72.75-73.19=-0.44$ only!

```{r}
# we can use R as a calculator and use the regression coefficients

# first we need to tidy reg2
reg2_tidy<-tidy(reg2)

# extract beta0 and save it
multi_beta_0<-reg2_tidy %>%
  filter(term=="(Intercept)") %>%
  select(estimate)

# extract beta1 and save it
multi_beta_1<-reg2_tidy %>%
  filter(term=="Age") %>%
  select(estimate)

# extract beta2 and save it
multi_beta_2<-reg2_tidy %>%
  filter(term=="MPHover") %>%
  select(estimate)

# predicted amount for 18 year old going 10 MPH over
amount_18yr_10mph<-multi_beta_0+multi_beta_1*18+multi_beta_2*10

# predicted amount for 40 year old going 10 MPH over
amount_40yr_10mph<-multi_beta_0+multi_beta_1*40+multi_beta_2*10

# difference
amount_18yr_10mph-amount_40yr_10mph # close, we have some rounding error!
```

---

## Question 9

How about the difference in expected fine between two 18 year olds, one who went 10 MPH over, and one who went 30 MPH over?

---

<!--ANSWER BELOW HERE-->

18-year-old driver, 10 MPH over: we saw was $72.75.

18-year-old driver, 30 MPH over:

$$\begin{align*}
\widehat{Amount}_i &= \hat{\beta_0}+\hat{\beta_1}Age_i+\hat{\beta_2}MPHover_i\\
&= 3.49+0.02(18)+6.89(30)\\
&=210.55\\
\end{align*}$$

The difference is now $210.55-72.75=-137.80$!

```{r}
# we can use R as a calculator and use the regression coefficients

# predicted amount for 40 year old going 30 MPH over
amount_18yr_30mph<-multi_beta_0+multi_beta_1*18+multi_beta_2*30

# difference
amount_18yr_30mph-amount_18yr_10mph
```

So clearly the speed plays a *much* bigger role than age does!

---

## Question 10

Use the `huxtable` package's `huxreg()` command to make a regression table of your two regressions: the one from question 4, and the one from question 7.

---

<!--ANSWER BELOW HERE-->

```{r}
library(huxtable)
huxreg(reg1,reg2) # the default

# some customization
huxreg("Simple OLS"=reg1,
       "Mulitvariate OLS"=reg2,
       coefs = c("Constant" = "(Intercept)",
                 "Age" = "Age",
                 "MPH Over" = "MPHover"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)

```

Look at that change in $R^2$!

---

## Question 11
Are our two independent variables multicollinear? Do younger people tend to drive faster?

### Part A
Get the correlation between `Age` and `MPHover`.

---

<!--ANSWER BELOW HERE-->

```{r}
# same alternative methods as question 3

# fortunately no NA's here!

speed %>%
  select(Age, MPHover) %>%
  cor()
```

---

### Part B
Make a scatterplot of `MPHover` on `Age`.

---

<!--ANSWER BELOW HERE-->

```{r}
ggplot(data = speed)+
  aes(x = Age,
      y = MPHover)+
  geom_point()+
  geom_smooth(method = "lm")
```

---

### Part C
Run an auxiliary regression of `MPHover` on `Age`.

---

<!--ANSWER BELOW HERE-->

```{r}
reg_aux<-lm(MPHover~Age, data = speed)
summary(reg_aux)
```

---

### Part D
Interpret the coefficient on `Age`.

---

<!--ANSWER BELOW HERE-->

In our notation, this is $\hat{\delta_1}=-0.03$, which says for every 1 year older someone is, they drive 0.03 MPH less over the speed limit.

---

### Part E

Look at your regression table in question 10. What happened to the standard error on `Age`? Why (consider the formula for variation in $\hat{\beta_1})$

---

<!--ANSWER BELOW HERE-->

The formula for $var[\hat{\beta_1}]=\frac{1}{1-R^2_1} \times \frac{SER^2}{n \times var[Age]}$

The standard error of the regression (SER) went from 56.13 to 39.67!^[Note: you need to be sure to show `sigma` in your `huxreg()` or look at each of your regression results with broom's `glance()` or Base R's `summary()` output.] Adding the second variable created a much better fit, which lowered variance on the betas (and hence, standard error, the square root of variance).

Variance might have slightly increased to more than it otherwise would be, due to multicollinearity between `Age` and `MPHover`, which we investigate next.

---

### Part F

Calculate the Variance Inflation Factor (VIF) using the `car` package's `vif()` command.^[Run it on your multivariate regression object from Question 7.]

---

<!--ANSWER BELOW HERE-->

```{r}
library(car)
vif(reg2)
```

The VIF is 1.01 for $\hat{\beta_1}$ and $\hat{\beta_2}$. Variance increases only by 1% due to (weak) multicollinearity between `Age` and `MPHover`.

---

### Part G
Calculate the VIF manually, using what you learned in this question.

---

<!--ANSWER BELOW HERE-->

$VIF=\frac{1}{1-R^2_1}$, where $R^2_1$ comes from the auxiliary regression of `MPHover` on `Age` (Part C). The $R^2$ from that regression was $0.011$, so:

$$\begin{align*}
VIF&=\frac{1}{1-R^2_1}\\
&=\frac{1}{1-0.011}\\
&=\frac{1}{0.989}\\
& \approx 1.011\\
\end{align*}$$

```{r}
# we can do this in R

# first we need to extract and save the R-squared from the aux reg
# use broom's glance()

aux_r_sq<-glance(reg_aux)$r.squared %>%
  round(.,3) # round to 3 decimals

vif<-1/(1-aux_r_sq)
vif
```

Again, we have some slight rounding error.

---

## Question 12
Let's now think about the omitted variable bias. Suppose the "true" model is the one we ran from Question 7.

### Part A
Do you suppose that `MPHover` fits the two criteria for omitted variable bias?

---

<!--ANSWER BELOW HERE-->

1. `MPHover` probably affects `Amount`
2. $cor(MPHover, Age) \neq 0$

Possibly, though it is only weakly correlated with Age $(-0.10)$. Speed clearly has a big role to play in affecting and predicting fines, but probably does not add very much bias to the marginal effect of Age on Amount.

---

### Part B

Look at the regression we ran in Question 4. Consider this the "omitted" regression, where we left out `MPHover`. Does our estimate of the marginal effect of `Age` on `Amount` overstate or understate the *true* marginal effect?

---

<!--ANSWER BELOW HERE-->

Since there is negative correlation between `MPHover` and `Age`, $\hat{\beta_1}$ likely understates the effect of `Age` on `Amount.`

---

### Part C

Use the "true" model (Question 7), the "omitted" regression (Question 4), and our "auxiliary" regression (Question 11) to identify each of the following parameters that describe our biased estimate of the marginal effect of `Age` on `Amount`:^[See the notation from class 3.2.]
$$\alpha_1=\beta_1+\beta_2\delta_1$$

---

<!--ANSWER BELOW HERE-->

Using our notation from class, and the regressions from questions 4, 7, and 11, we have three regressions:

- **True Model**: $\widehat{\text{Amount}_i}=3.49+0.02 \, \text{Age}_i+6.89 \, \text{MPHover}_i$
- **Omitted Reg**: $\widehat{\text{Amount}_i}=131.71-0.29 \, \text{Age}_i$
- **Auxiliary Reg**: $\widehat{\text{MPHover}_i}=16.54-0.04 \, \text{Age}_i$

$$\begin{align*}
\hat{\alpha_1}&=\beta_1+\beta_2 \delta_1\\
-0.29 &= 0.02+6.89(-0.04)\\
\end{align*}$$

Where:

- $\beta_1=0.02$: the true marginal effect of `Age` on `Amount` (holding `MPHover` constant)
- $\beta_2=6.89$: the true marginal effect of `MPHover` on `Amount` (holding `Age` constant)
- $\delta_1=6.89$: the (auxiliary) effect of `MPHover` on `Age`

We have some slight rounding error, but this is the relationship.

---

### Part D

From your answer in part C, how large is the omitted variable bias from leaving out `MPHover`?

---

<!--ANSWER BELOW HERE-->

$$\begin{align*}
\hat{\alpha_1}&=\beta_1+\underbrace{\beta_2 \delta_1}_{Bias}\\
-0.29 &= 0.02+\underbrace{6.89(-0.04)}_{bias}\\
Bias &=-0.2756\\
\end{align*}$$

---

## Question 13
Make a coefficient plot of your coefficients from the regression in Question 7.

<!--ANSWER BELOW HERE-->

```{r}
# first we have to make sure to tidy our reg with confidence intervals!
reg2_tidy<-tidy(reg2, conf.int=TRUE)

ggplot(data = reg2_tidy)+
  aes(x = estimate,
      y = term,
      color = term)+
    geom_point()+ # point for estimate
    # now make "error bars" using conf. int.
    geom_segment(aes(x = conf.low,
                     xend = conf.high,
                     y=term,
                     yend=term))+
  # add line at 0
  geom_vline(xintercept=0, size=1, color="black", linetype="dashed")+
  #scale_x_continuous(breaks=seq(-1.5,0.5,0.5),
  #                   limits=c(-1.5,0.5))+
  labs(x = expression(paste("Marginal effect, ", hat(beta[j]))),
       y = "Variable")+
  guides(color=F)+ # hide legend
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

---

### 2019-11-05 content/class/17-class.Rmd
Add class 17 content

---
title: "3.5: Causal Inference and DAGs - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

*Tuesday, November 5, 2019*

# Overview

Today we return to a more nuanced discussion of causality, given what we now know about multivariate regression. We also introduce a new tool for thinking about simple causal models, the **directed acyclic graph (DAG)*. These are a hip new trend for thinking about causal inference, so new and trendy that they aren't really in any mainstream textbooks yet!

# New Packages Mentioned

- `ggdag`: for drawing DAGs in `ggplot`
- [`dagitty.net`](http://dagitty.net): not a package in R, but an online tool to help you work with DAGs (see the cheat sheets in today's [readings page](/reading/17-reading))

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/17-slides.html)

### 2019-11-05 content/reading/17-reading.Rmd
Add class 17 content

---
title: "3.5: Causal Inference and DAGs - Readings"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-fullnote-bibliography-no-bib.csl
---

DAGs are a trendy new concept in econometrics and causal inference, so much so that they have yet to find their way into any major econometrics textbook! There are some resources, however, that you can look to for understanding how they work (and I base much of my lecture off of them).

# Recommended Readings

- [<i class="fas fa-file-pdf"></i> Ch. 4 in Cunningham (2019), *Causal Inference, the Mixtape*](https://www.scunning.com/cunningham_mixtape.pdf)
- [<i class="fas fa-external-link-square-alt"></i> My blog post on "Econometrics, Data Science, and Causal Inference"](https://ryansafner.com/post/econometrics-data-science-and-causal-inference/)
- [<i class="fas fa-book"></i> Pearl and MacKenzie, (2018), *The Book of Why*](https://www.amazon.com/Book-Why-Science-Cause-Effect/dp/046509760X)
- [<i class="fas fa-external-link-square-alt"></i> Huntington-Klein (2019), Dagitty.net Cheat Sheet"](https://nickch-k.github.io/introcausality/Cheat%20Sheets/Dagitty_Cheat_Sheet.pdf)
- [<i class="fas fa-external-link-square-alt"></i> Huntington-Klein (2019), Causal Diagrams Cheat Sheet"](https://nickch-k.github.io/introcausality/Cheat%20Sheets/Causal_Diagrams_Cheat_Sheet.pdf)

The best book to get more into the philosophy of causality and the major origin of DAGs is Judea Pearl (and David McKenzie)'s *The Book of Why*. We owe much to Pearl.

The best instantiation of DAGs and causal inference into a "textbook" on econometrics and methods is Scott Cunningham's (open source!) *Causal Inference: The Mixtape* chapter on DAGs. Nick Huntington-Klein has some great lecture slides, and some cheat sheets on using Dagitty.net and understanding DAGs.

Finally, I have a blog post discussing the difference between econometrics, causal inference, and data science. The end touches on causality, DAGs, and Pearl.

### 2019-11-05 static/slides/17-slides.Rmd
Add class 17 content

---
title: "3.5: Causal Inference and DAGs"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))
```

# Causality

.pull-left[
- Statistics profession is obstinant that we cannot say anything about causality
  - Statisticians: causation is simply a subset of correlation, where correlation is $-1$ or $1$

- But you have to! It's how the human brain works!

- We can't concieve of (spurious) correlation without some causation

]

.pull-right[

.center[
![:scale 70%](https://www.dropbox.com/s/qp9m156rcxqp3nc/bookofwhy.jpg?raw=1)
]
]

---

background-image: url(https://www.dropbox.com/s/245ma3buawugeog/rctgoldstandard.png?raw=1)
background-size: cover

---

# RCTs are All the Rage

.pull-left[
.center[

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Professors Esther Duflo and Abhijit Banerjee, co-directors of MIT&#39;s <a href="https://twitter.com/JPAL?ref_src=twsrc%5Etfw">@JPAL</a>, receive congratulations on the big news this morning. They share in the <a href="https://twitter.com/hashtag/NobelPrize?src=hash&amp;ref_src=twsrc%5Etfw">#NobelPrize</a> in economic sciences “for their experimental approach to alleviating global poverty.”<br> <br>Photo: Bryce Vickmark <a href="https://t.co/NWeTrjR2Bq">pic.twitter.com/NWeTrjR2Bq</a></p>&mdash; Massachusetts Institute of Technology (MIT) (@MIT) <a href="https://twitter.com/MIT/status/1183752282988564480?ref_src=twsrc%5Etfw">October 14, 2019</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

]
]

.pull-right[
.center[

Source: [Vox (Oct 14, 2019)](https://www.vox.com/future-perfect/2019/10/14/20913928/nobel-prize-economics-duflo-banerjee-kremer)

![](https://www.dropbox.com/s/119oeqictaktgsw/voxnobelrcts.png?raw=1)

]
]

---

# But Not Everyone Agrees I

.left-column[

.center[
![:scale 75%](https://www.dropbox.com/s/z1mnxolnzhvuv27/deaton.jpg?raw=1)

Angus Deaton

Economics Nobel 2015
]

]
.right-column[

> The RCT is a useful tool, but I think that is a mistake to put method ahead of substance. I have written papers using RCTs. Like other methods of investigation, they are often useful, and, like other methods, they have dangers and drawbacks. Methodological prejudice can only tie our hands...**It is not true that an RCT, when feasible, will always do better than an observational study.** This should not be controversial [but] might still make some uncomfortable, particularly the second [statement]: (a) RCTs are affected by the same problems of inference and estimation that economists have faced using other methods, and (b) no RCT can ever legitimately claim to have established causality. **My theme is that RCTs have no special status, they have no exemption from the problems of inference that econometricians have always wrestled with, and there is nothing that they, and only they, can accomplish.**
]

.source[Deaton, Angus, 2019, "[Randomization in the Tropics Revisited: A Theme and Eleven Variations](https://scholar.princeton.edu/sites/default/files/deaton/files/deaton_randomization_revisited_v2_2019_01.pdf)," Working Paper]

---

# But Not Everyone Agrees II

.left-column[
.center[
![:scale 75%](https://www.dropbox.com/s/e6mbwn4oba8w7lx/pritchett.jpg?raw=1)

Lant Pritchett

]
]

.right-column[

> People keep saying that the recent Nobelists "studied global poverty." This is exactly wrong. They made a commitment to a method, not a subject, and their commitment to method prevented them from studying global poverty.

> At a conference at Brookings in 2008 Paul Romer (last years Nobelist) said: "You guys are like going to a doctor who says you have an allergy and you have cancer. With the skin rash we can divide you skin into areas and test variety of substances and identify with precision and some certainty the cause. Cancer we have some ideas how to treat it but there are a variety of approaches and since we cannot be sure and precise about which is best for you, we will ignore the cancer and not treat it."

]

.source[[Source](https://www.facebook.com/lant.pritchett/posts/10218688602381690)]

---

# But Not Everyone Agrees III

.left-column[

.center[
![:scale 75%](https://www.dropbox.com/s/z1mnxolnzhvuv27/deaton.jpg?raw=1)

Angus Deaton

Economics Nobel 2015
]

]
.right-column[

> "Lant Pritchett is so fun to listen to, sometimes you could forget that he is completely full of shit."

]

.source[[Source](https://medium.com/@ismailalimanik/lant-pritchett-the-debate-about-rcts-in-development-is-over-ec7a28a82c17)]

---

# RCTs and Evidence-Based Policy I

.pull-left[

- Programs *randomly* assign treatment to different individuals and measure causal effect of treatment

- **RAND Health Insurance Study**

- **Oregon Medicaid Expansion**

- **HUD's Moving to Opportunity**

- **Tennessee STAR**

]

.pull-right[
.center[
![](https://www.dropbox.com/s/j0kwzqb2ptusckt/madlaboratory.jpg?raw=1)
]
]

---

# RCTs and Evidence-Based Policy II

- Should we *ONLY* base policies on the evidence from Randomized Controlled Trials?

--

.pull-left[

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">|￣￣￣￣￣￣￣￣￣￣|<br> IF U DONT SMOKE,<br> U ALREADY <br> BELIEVE IN<br> CAUSAL INFERENCE<br> WITHOUT<br> RANDOMIZED TRIALS<br>|＿＿＿＿＿＿＿＿＿＿| <br> (\__/) ||<br> (•ㅅ•) ||<br> / 　 づ<a href="https://twitter.com/hashtag/HistorianSignBunny?src=hash&amp;ref_src=twsrc%5Etfw">#HistorianSignBunny</a> <a href="https://twitter.com/hashtag/Epidemiology?src=hash&amp;ref_src=twsrc%5Etfw">#Epidemiology</a></p>&mdash; Ellie Murray (@EpiEllie) <a href="https://twitter.com/EpiEllie/status/1017622949799571456?ref_src=twsrc%5Etfw">July 13, 2018</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

]

--

.pull-right[

.center[

Source: [British Medical Journal](https://www.bmj.com/content/363/bmj.k5094)

![](https://www.dropbox.com/s/9mimwcoamiv54tf/rctparachutesstudy.png?raw=1)
]

]

---

# RCTs and Evidence-Based Policy III

.pull-left[
.center[
![:scale 80%](https://www.dropbox.com/s/5ptmdhgeyerhr4a/rctparachutes1.jpg?raw=1)
]
]

--

.pull-right[
.center[
![:scale 80%](https://www.dropbox.com/s/3knd5wy8h4eyq1j/rctparachutes2.jpg?raw=1)
]

]

---

# Correlation and Causation I

.center[

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">&quot;Correlation implies casuation,&quot; the dean whispered as he handed me my PhD.<br><br>&quot;But then why-&quot;<br><br>&quot;Because if they knew, they wouldn&#39;t need us.&quot;</p>&mdash; David Robinson (@drob) <a href="https://twitter.com/drob/status/877976063942512640?ref_src=twsrc%5Etfw">June 22, 2017</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

]

---

# Correlation and Causation II

.center[

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">Causality isn&#39;t achieved; it&#39;s approached.</p>&mdash; John B. Holbein (@JohnHolbein1) <a href="https://twitter.com/JohnHolbein1/status/982635508089180161?ref_src=twsrc%5Etfw">April 7, 2018</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

]

---

# Correlation and Causation

.pull-left[
- .onfire[Correlation:]
  - Math & Statistics
  - Computers, AI, Machine learning can figure this out (even better than humans)

- .onfire[Causation:]
  - Philosophy, Intuition, Theory
  - .shout[Counterfactual thinking], unique to humans (vs. animals or machines)
  - Computers *cannot* yet figure this out

]

.pull-right[
.center[
![](https://www.dropbox.com/s/c91a06o91rf3e5h/causation.jpg?raw=1)
]
]

---

# Causation Requires Counterfactual Thinking

.pull-left[
.center[
![:scale 70%](https://www.dropbox.com/s/6n3dg3xpizsxpsj/ladderofcausation.png?raw=1)
]
]

.pull-right[
.center[
![:scale 70%](https://www.dropbox.com/s/qp9m156rcxqp3nc/bookofwhy.jpg?raw=1)
]
]

---

background-image: url(https://www.dropbox.com/s/zsx3pa4m51p82dj/twopaths.jpg?raw=1)
background-size: cover

---

# Recall: The Fundamental Problem of Causal Inference I

.left-column[

| John      | Maria     |
|-----------|-----------|
| ![](https://www.dropbox.com/s/5vh181yeg0p91t7/coldman2.jpg?raw=1) | ![](https://www.dropbox.com/s/7qa0gtuy29us4p2/femalejogger2.jpg?raw=1) |
| $Y_J^0=3$ | $Y_M^0=5$ | 
| $Y_J^1=4$ | $Y_M^1=5$ |
| $Y_J^1-Y_J^0=1$ | $Y_M^1-Y_M^0=0$ |
| $Y_J=(Y_J^1)=4$ | $Y_M=(Y_M^0)=5$ |

]

.right-column[

- Recall example from [class 1.2](/class/02-class)

- John will choose to buy health insurance

- Maria will choose to not buy health insurance

- Health insurance improves John's score by 1, has no effect on Maria's score

- Note, all we can observe in the data are their health outcomes *after* they have chosen (not) to buy health insurance: $Y_J=4$ and $Y_M=5$

- .whisper[*Observed* difference] between John and Maria: $$Y_J-Y_M=-1$$
]

---

# Recall: Counterfactuals

.left-column[

| John      | Maria     |
|-----------|-----------|
| ![](https://www.dropbox.com/s/5vh181yeg0p91t7/coldman2.jpg?raw=1) | ![](https://www.dropbox.com/s/7qa0gtuy29us4p2/femalejogger2.jpg?raw=1) |
| $Y_J=4$ | $Y_M=5$ |

This is all the data we *actually* observe
]

.right-column[
- Observed difference between John and Maria: 
$$Y_J-Y_M=\underbrace{Y^1_J-Y^0_M}_{=-1}$$

- Recall:
    - John has bought health insurance $(Y^1_J)$
    - Maria has not bought insurance $(Y^0_M)$

- We don't see the .shout[counterfactuals]: 
    - John's score *without* insurance
    - Maria score *with* insurance

]

---

# Causal Inference

.pull-left[

- We will seek to understand what causality *is* and how we can approach finding it

- We will also explore the different common .shout[research designs] meant to .whisper[identify] causal relationships

- **These skills**, more than supply & demand, constrained optimization models, ISLM, etc, **are the tools and comparative advantage of a modern research economist**
    - Why all big companies (especially in tech) have entire economics departments in them

]

.pull-right[

.center[
![](https://www.dropbox.com/s/yw8t5xsa8dgei71/coding.jpeg?raw=1)
]
]

---

# Clever Research Designs Identify Causality

```{r, fig.retina=3, fig.align="center", out.width="60%"}
ggplot(data = tibble(x=c(0,10),
                     y=c(0,10)))+
  aes(x = x,
      y = y)+
  geom_label(x=1,y=2, label="Correlation", size =10, color = "#fde0dd")+
  geom_label(x=5,y=2, label="Causation", size = 10, color = "#7a0177")+

  geom_label(x=1,y=4, label="Differences", size =5, color = "#fde0dd")+
  geom_label(x=1,y=3.5, label="Pre-Post", size =5, color = "#fde0dd")+
  geom_label(x=2,y=4.5, label="Multiple Regression", size =5, color = "#fbb4b9")+
  geom_label(x=2,y=3.5, label="Matching", size =5, color = "#fbb4b9")+
    geom_label(x=3,y=5, label="Fixed Effects", size =5, color = "#f768a1")+
    geom_label(x=3.5,y=4.5, label="Diff-in-Diff", size =5, color = "#c51b8a")+
  geom_label(x=4,y=3.5, label="Natural Experiments", size =5, color = "#c51b8a")+
  geom_label(x=3.5,y=4, label="Regression Discontinuity", size =5, color = "#c51b8a")+
  geom_label(x=5,y=4.5, label="RCTs", size =5, color = "#7a0177")+
  annotate("segment", x = 1, xend = 5, y = 3, yend = 3, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+
  scale_x_continuous(breaks=seq(0,6,1),
                     limits=c(0,6))+
  scale_y_continuous(breaks=seq(1,5,1),
                     limits=c(1,5))+
  theme_void()

```

---

# What Then IS Causation?

.pull-left[

- $X$ causes $Y$ if we can intervene and change $X$ without changing anything else, and $Y$ changes
- $Y$ "listens to" $X$
    - $X$ may not be the only thing that causes $Y$

]

.pull-right[
.center[
![:scale 80%](https://www.dropbox.com/s/6w66jcc9gmw3hw4/lightswitch.jpg?raw=1)
]
]

---

# What Then IS Causation?

.pull-left[

- $X$ causes $Y$ if we can intervene and change $X$ without changing anything else, and $Y$ changes
- $Y$ "listens to" $X$
    - $X$ may not be the only thing that causes $Y$

.content-box-green[
.green[**Example**]

If $X$ is a light switch, and $Y$ is a light:

- Flipping the switch $(X)$ causes the light to go on $(Y)$
- But NOT if the light is burnt out (No $Y$ despite $X$)
- OR if the light was already on $(Y$ without $X$)
]

]

.pull-right[
.center[
![:scale 80%](https://www.dropbox.com/s/6w66jcc9gmw3hw4/lightswitch.jpg?raw=1)
]
]

---

# Less-Clear Causal Claims

.content-box-green[
.green[**Example**]

- Getting a college degree raises your wages

]

--

.content-box-green[
.green[**Example**]

- Tariffs reduce the volume of international trade

]

---

# Non-Causal Claims

- All of the following have non-zero correlations. Are they causal?

.content-box-green[
.green[**Example**]

- Greater ice cream sales cause more violent crime
- Rooster crows cause the sun to rise in the morning

]

--

- Less clear:

.content-box-green[
.green[**Example**]

- Colds tend to clear up a few days after you take Vitamin-C
- The economy performs better or worse depending on the political party in power

]

---

# Counterfactuals

- The *sine qua non* of causal claims are .shout[counterfactuals]: what would $Y$ have been if $X$ had been different?
- It is **impossible** to make a counterfactual claim from data alone! Need a causal model of the data-generating process!

---

# Counterfactuals and RCTs

- Again, RCTs are invoked as the gold standard for their ability to make counterfactual claims:
- Treatment/intervention $(X)$ is *randomly assigned* to individuals, and then outcome $Y$ is measured

> If person i who recieved treatment *had not recieved* the treatment, we can predict what his outcome *would have been*

> If person j who did not recieve treatment *had recieved treatment*, we can predict what her outcome *would have been*

- We can say this because, on average, treatment and control groups are *the same before treatment*

---

# From RCTs to Causal Models

- RCTs are but the best-known method of a large, growing science of .shout[causal inference]
- We need a .shout[causal model] to describe the .shout[data-generating process (DGP)]
- Requires us to make some .whisper[assumptions]

---

class: inverse, center, middle

# Causal Diagrams

---

# Causal Diagrams/DAGs

.pull-left[
- A surprisingly simple, yet rigorous and powerful method of modeling is using a .shout[causal diagram] or .shout[Directed Acyclic Graph (DAG)]

- A simple series of nodes (variables) connected by arrows that indicate causal effects
    - Arrows can only go one way ("directed") and can never loop back ("acyclic")
]

.pull-right[
```{r, echo = F, warning = F, message = F, fig.retina = 2, fig.align = "center"}
library("ggdag")

dagify(Y ~ X, Y ~ M, M ~ X, Y ~ Z, X ~ U, Z ~ U) %>% 
  ggdag_parents("Y", layout = "circle", stylized = F)+theme_dag_blank()+
  theme(legend.position = "none")

```
]

---

# Drawing a DAG

- Causal diagram (DAG) is the *model* of what you think the data-generating process is

- Useful to help figure out how to *identify* particular relationships of interest

- Requires some common sense/economic intutition

- Remember, all models are wrong, we just need them to be *useful*!

---

# Drawing a DAG: Example

.pull-left[

- Suppose we have data on three variables
  - `IP`: how much a firm spends on IP lawsuits 
  - `tech`: whether a firm is in tech industry
  - `profit`: firm profits

- They are all correlated with each other, but what's are the causal relationships?

- We need our own **causal model** (from theory, intuition, etc) to sort
  - Data alone will not tell us!
]

.pull-right[

```{r, fig.retina=3, outwidth="90%", fig.align="center"}
library(ggdag)
#theme_set(theme_dag_blank())
library(gridExtra)

h1 <- dagify(profit ~ tech+ip,
             tech ~ ip,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE)+
  labs(title = "1")+
  theme_void()+theme(legend.position = "none")

h2 <- dagify(profit ~ tech+ip,
             ip ~ tech,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE)+
  labs(title = "2")+
  theme_void()+theme(legend.position = "none")

h3 <- dagify(profit ~ tech,
             ip ~ tech+profit,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE)+
  labs(title = "3")+
  theme_void()+theme(legend.position = "none")

h4 <- dagify(profit ~ ip,
             tech ~ ip+profit,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE)+
  labs(title = "4")+
  theme_void()+theme(legend.position = "none")

h5 <- dagify(tech ~ ip+profit,
             ip ~ profit,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE)+
  labs(title = "5")+
  theme_void()+theme(legend.position = "none")

h6 <- dagify(tech ~ profit,
             ip ~ profit+tech,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE)+
  labs(title = "6")+
  theme_void()+theme(legend.position = "none")

grid.arrange(h1, h2, h3, h4, h5, h6, ncol=2)

```
]

---

# Drawing a DAG:

.pull-left[

1. Consider all the variables likely to be important to the data-generating process (including variables we can't observe!)

2. For simplicity, combine some similar ones together or prune those that aren't very important

3. Consider which variables are likely to affect others, and draw arrows connecting them

4. Test some testable implications of the model (to see if we have a correct one!)

]

.pull-right[

.center[
![](https://www.dropbox.com/s/v5vwsadw5vs448t/causality.jpg?raw=1)
]
]

---

# Side Notes

.pull-left[

- Drawing an arrow requires a direction - making a statement about causality!

- *Omitting* an arrow makes an equally important statement too!
  - In fact, we will *need* omitted arrows to show causality!

- If two variables are correlated, but neither causes the other, it's likely they are both caused by another (perhaps unobserved) variable - add it!

- There should be no *cycles* or *loops* (if so, there's probably another missing variable, such as time)

]

.pull-right[
.center[
![](https://www.dropbox.com/s/v5vwsadw5vs448t/causality.jpg?raw=1)
]
]
---

# DAG Example I

.pull-left[

.content-box-green[
.green[**Example**]: what is the effect of education on wages?
]

- Education ("treatment" or "exposure")

- Wages ("outcome" or "response")
]

.pull-right[

```{r, echo = F}
dagify(wage~educ) %>% 
  ggdag_parents("wage", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# DAG Example I

.pull-left[

- What other variables are important?
  - Ability
  - Socioeconomic status
  - Demographics
  - Phys. Ed. requirements
  - Year of birth
  - Location
  - Schooling laws
  - Job connections
]

.pull-right[

```{r, echo = F}
dagify(wage~educ) %>% 
  ggdag_parents("wage", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# DAG Example I

.pull-left[

- In social science and complex systems, 1000s of variables could plausibly be in DAG!

- So simplify:
  - Ignore trivial things (Phys. Ed. requirement)
  - Combine similar variables (Socioeconomic status, Demographics, Location $\rightarrow$ Background)
]

.pull-right[

```{r, echo = F}
dagify(wage~educ) %>% 
  ggdag_parents("wage", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# DAG Example II

.pull-left[

- Background, Year of birth, Location, Compulsory schooling, all cause education

- Background, year of birth, location, job connections probably cause wages
]

.pull-right[

```{r, echo = F}
dagify(wage~educ+conx+year+bckg+loc,
       educ~bckg+year+loc+laws) %>%
  ggdag_parents("wage", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# DAG Example III

.pull-left[

- Background, Year of birth, Location, Compulsory schooling, all cause education

- Background, year of birth, location, job connections probably cause wages

- Job connections in fact is probably caused by education!

- Location and background probably both caused by unobserved factor `(u1)`
]

.pull-right[

```{r, echo = F}
dagify(wage~educ+conx+year+bckg+loc,
       educ~bckg+year+loc+laws,
       conx~educ,
       bckg~u1,
       loc~u1) %>% 
  ggdag_parents("wage", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# DAG Example IV

.pull-left[

- This is messy, but we have a causal model!

- Makes our assumptions **explicit**, and many of them are **testable**

- DAG suggests certain relationships that will *not* exist:
  - all relationships between `laws` and `conx` go through `educ`
  - so if we controlled for `educ`, then `cor(laws,conx)` should be zero!
]

.pull-right[

```{r, echo = F}
dagify(wage~educ+conx+year+bckg+loc,
       educ~bckg+year+loc+laws,
       conx~educ,
       bckg~u1,
       loc~u1) %>% 
  ggdag_parents("wage", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# Let the Computer Do It: Dagitty.net I

.pull-left[
.center[

![](https://www.dropbox.com/s/zhqgfyvk2x4z863/dagitty.png?raw=1)
]
]

.pull-right[

- [Dagitty.net](http://dagitty.net) is a great tool to make these and give you testable implications

- Click `Model -> New Model`

- Name your "exposure" variable $(X$ of interest) and "outcome" variable $(Y)$

]

---

# Let the Computer Do It: Dagitty.net II

.pull-left[
.center[

![](https://www.dropbox.com/s/qu6839emzoitb1c/dagittyex1.png?raw=1)
]
]

.pull-right[

- Click and drag to move nodes around

- Add a new variable by double-clicking

- Add an arrow by double-clicking one variable and then double-clicking on the target (do again to remove arrow)

]

---

# Let the Computer Do It: Dagitty.net III

.pull-left[
.center[

![](https://www.dropbox.com/s/a55b66p8i4znjto/dagittyex2.png?raw=1)
]
]

.pull-right[

- Tells you **how to identify your effect**! (upper right)

> Minimal sufficient adjustment sets containing background, location, year for estimating the total effect of educ on wage:
background, location, year

- Tells you **some testable implications** (middle right)
  - *conditional independencies*, take the last one:
  - `job_connections` $\perp$ `year` | `educ`
  - means: holding constant `educ`, there should be no correlation between `job_connections` and `year` - can test this with data!

]

---

# Implications

- In order to .onfire[identify] the effect of education on wages, our model implies we need to control for background, location, and year
  - note: we do *not* need to control for anything else!

--

- How to control for them?
  - many ways, we've learned one: multivariate regression:

--

$$wages=\beta_0+\beta_1 \, educ + \beta_2 \, background + \beta_3 \, location + \beta_4 \, year$$

--

- How did Dagitty know what we needed to do? Some simple rules (below)

---

# Causal Effect

.pull-left[
.center[

![](https://www.dropbox.com/s/o7h0zbpvej0zuf5/dagittycontrolled.png?raw=1)
]
]

.pull-right[

- If we control for `background`, `location`, and `year`, we can **identify** the causal effect of `educ` $\rightarrow$ `wage`.

]

---

# You Can Draw DAGs In R

.pull-left[

- New package: `ggdag`
- Arrows are made with formula notation: `Y~X` means "`Y` is caused by `X`"

```{r, echo = T, eval=F}
library(ggdag)
dagify(wage~educ+conx+year+bckg+loc,
       educ~bckg+year+loc+laws,
       conx~educ,
       bckg~u1,
       loc~u1) %>% 
  ggdag()+
  theme_dag()
```
]

.pull-right[

```{r, echo = F}
dagify(wage~educ+conx+year+bckg+loc,
       educ~bckg+year+loc+laws,
       conx~educ,
       bckg~u1,
       loc~u1) %>% 
  ggdag(stylized = FALSE)+
  theme_dag_blank()
```

]

---

class: inverse, center, middle

# DAG Rules

---

# DAGs I

.pull-left[

- Typical notation:

- $X$ is independent variable of interest
  - Epidemiology: .onfire["intervention"]

- $Y$ is dependent/.onfire["response"] variable

- Other variables use other letters
]

.pull-right[

```{r, echo = F, fig.retina = 2, fig.align="center"}
dagify(Y~X,
                coords=list(
                x=c(X=1,Y=2),
                y=c(X=1,Y=1)
              )) %>% 
  tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```
]

---

# DAGs and Causal Effects

.pull-left[

- Arrows indicate causal effect (in proper direction)

- Two types of causal effect:
    1. Direct effects $X \rightarrow Y$
]

.pull-right[

```{r, echo = F, fig.retina = 2, fig.align="center"}
dagify(Y~X,
                coords=list(
                x=c(X=1,Y=2),
                y=c(X=1,Y=1)
              )) %>% 
  tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# DAGs and Causal Effects

.pull-left[

- Arrows indicate causal effect (in proper direction)

- Two types of causal effect:

1. Direct effects $X \rightarrow Y$

2. Indirect effects $X \rightarrow M \rightarrow Y$ 
  - $M$ is a "mediator" variable, the mechanism by which $X$ affects $Y$
]

.pull-right[

```{r, echo = F, fig.retina = 2, fig.align="center"}
dagify(Y~M,
               M~X,
                coords=list(
                x=c(X=1,Y=2, M=1.5),
                y=c(X=1,Y=1, M=1)
              )) %>% 
  tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# DAGs and Causal Effects

.pull-left[

- Arrows indicate causal effect (in proper direction)

- Two types of causal effect:

1. Direct effects $X \rightarrow Y$

2. Indirect effects $X \rightarrow M \rightarrow Y$ 
  - $M$ is a "mediator" variable, the mechanism by which $X$ affects $Y$

3. You of course might have both! 
]

.pull-right[

```{r, echo = F, fig.retina = 2, fig.align="center"}
dagify(Y~M+X,
               M~X,
                coords=list(
                x=c(X=1,Y=2, M=1.5),
                y=c(X=1,Y=1, M=2 )
              )) %>% 
  tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# Confounders

.pull-left[

- $Z$ is a .shout[confounder] of $X \rightarrow Y$, $Z$ affects both $X$ and $Y$

- $cor(X,Y)$ is made up of two parts:
    - The value of $Y$ is affected by the value of $X$ $(X \rightarrow Y$, good!)
    - $Z$ affects both the values of $X$ and $Y$ (A spurious correlation between $X$ and $Y$, bad!)

- Failing to control for $Z$ will .whisper[bias] our estimate of the causal effect of $X \rightarrow Y$!

]

.pull-right[

```{r, echo = F}
dagify(Y~Z+X,
       X~Z,
       coords=list(
         x=c(X=1,Y=2, Z=1.5),
         y=c(X=1,Y=1, Z=2)
         )) %>% 
  #tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# Confounders

.pull-left[

- Confounders are the DAG-equivalent of .shout[omitted variable bias]

- Biased regression $Y=\beta_0+\beta_1 X$ leaving out $Z$

- $\hat{\beta}_1$ picks up both:
  - $X \rightarrow Y$
  - $X \leftarrow Z \rightarrow Y$
]

.pull-right[

```{r, echo = F}
dagify(Y~Z+X,
       X~Z,
       coords=list(
         x=c(X=1,Y=2, Z=1.5),
         y=c(X=1,Y=1, Z=2)
         )) %>% 
  #tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# Front Doors and Back Doors

.pull-left[

- With this graph, there are two paths that connect $X$ and $Y$<sup>1</sup>:

1. A .shout[causal "front-door" path] $X \rightarrow Y$
    - Good, what we want to measure

2. A .shout[non-causal "back-door" path] $X \leftarrow Z \rightarrow Y$
    - At least one causal arrow runs in the opposite direction
    - Bad, adds a confounding bias

]

.pull-right[

```{r, echo = F}
dagify(Y~Z+X,
       X~Z,
       coords=list(
         x=c(X=1,Y=2, Z=1.5),
         y=c(X=1,Y=1, Z=2)
         )) %>% 
  #tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

.footnote[<sup>1</sup> Regardless of the directions of the arrows!]

---

# Controlling I

.pull-left[

- If we can .onfire[control for] $Z$, we can *block* the back-door path $X \leftarrow Z \rightarrow Y$
    - Essentially, delete the arrow(s) connecting $X$, $Z$, and $Y$

- This would only leave the front-door, $X \rightarrow Y$

]

.pull-right[

```{r, echo = F}
dagify(Y~X,
       coords=list(
         x=c(X=1,Y=2),
         y=c(X=1,Y=1)
         )) %>% 
  #tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# Controlling II

.pull-left[

- How to "control for" $Z$? We want to remove its influence over determining the values of $X$ and $Y$

- Multiple methods exist, but we've done this with .shout[multivariate regression]

- At the simplest level, consider only looking at values of $X$ and $Y$ for all observations that have the *same* value of $Z$
  - removes causal influence $Z$ has on $X$ and on $Y$
  - leaves only the influence $X$ has on $Y$!

]

.pull-right[

```{r, echo = F}
dagify(Y~X,
       coords=list(
         x=c(X=1,Y=2),
         y=c(X=1,Y=1)
         )) %>% 
  #tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]
---

# Controlling III

.pull-left[

- Controlling for a single variable along a long causal path is sufficient to block that path!

- Causal path: $X \rightarrow Y$
- Backdoor path: $X \leftarrow A \rightarrow B \rightarrow C \rightarrow Y$

- It is sufficient to block this backdoor by controlling **either** $A$ **or** $B$ **or** $C$!

]

.pull-right[

```{r, echo = F}
dagify(Y~X+C,
       X~A,
       B~A,
       C~B,
       coords=list(
         x=c(X=1,A=1.25, B=1.5, C=1.75, Y=2),
         y=c(X=1,A=1.5,B=1.5,C=1.5,Y=1)
         )) %>% 
  #tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```
]
---

# The Back Door Criterion

.pull-left[

- To achieve proper .onfire[identification] of the causal effect:

- The .shout["back-door criterion"]: control for the minimal amount of variables sufficient to ensure that no back-door exists between $X$ and $Y$

]

.pull-right[

```{r, echo = F}
dagify(Y~Z+X,
       X~Z,
       coords=list(
         x=c(X=1,Y=2, Z=1.5),
         y=c(X=1,Y=1, Z=2)
         )) %>% 
  #tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# The Back Door Criterion

.pull-left[

- Implications of the Back-door criterion:

1. You *only* need to control for the variables that keep a back-door open, *not all other variables!*

.content-box-green[
.green[**Example**]:

- $X \rightarrow Y$ (front-door)
- $X \leftarrow A \rightarrow B \rightarrow Y$ (back-door)

- Need only control for $A$ *or* $B$ to block the one back-door path
- $Z$ has no effect on $X$, and therefore we don't need to control for it!

]

]

.pull-right[

```{r, echo = F}
dagify(Y~X+B+Z,
       X~A,
       B~A) %>% 
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# The Back Door Criterion: Colliders

.pull-left[

- A case where controlling for a variable actually *adds bias* is if that variable is known as a .shout["collider"].

.content-box-green[
.green[**Example**]:

- $X \rightarrow Y$ (front-door)
- $X \leftarrow A \rightarrow B \leftarrow C \rightarrow Y$ (back-door, but **blocked by collider!**)

- $B$ is a collider because both $A$ and $C$ point to it
]

- A collider automatically *blocks* a path
- If you control for the collider, it *opens* a path (bad idea if it's a backdoor path)

]

.pull-right[

```{r, echo = F}
dagify(Y~X+C,
       X~A,
       B~A+C,
       coords=list(
         x=c(X=1, A=1, B=1.5, C=2, Y=2),
         y=c(X=1, A=2, B=1.5, C=2, Y=1)
       )) %>% 
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# The Back Door Criterion: Colliders

.pull-left[

.content-box-green[
.green[**Example**]:

- $Flu$: getting the flu
- $Bus$: being hit by a bus
- $Hos$: being in the hospital

- Both $Flu$ and $Bus$ send you to $Hos$ (arrows)

- Knowing $Flu$ doesn't tell us about $Bus$ (no arrow)

- Conditional on being in $Hos$, negative association between $Flu$ and $Bus$ (spurious!) 
]

]

.pull-right[

```{r, echo = F}
dagify(Hos~Flu+Bus,
       coords=list(
         x=c(Flu=1, Hos=1.5, Bus=2),
         y=c(Flu=1, Hos=2, Bus=1)
       )) %>% 
  ggdag_parents("Hos", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# Collider Example: Gender Discrimination I

.pull-left[

.content-box-green[
.green[**Example**]: How much of wage disparities are caused by gender-based discrimination?

1. $gender \rightarrow discr \rightarrow wage$ (Causal path)
2. $gender \rightarrow discr \rightarrow occup \rightarrow wage$ (Causal path)
3. $discr \leftarrow gender \rightarrow occup \rightarrow wage$ (Open back door)
4. $discr \leftarrow gender \rightarrow occup \leftarrow abil \rightarrow wage$ (BLOCKED back door!)
5. $gender \rightarrow discrim \rightarrow occup \leftarrow abil \rightarrow wage$ (BLOCKED back door!)

- Should we control for `occup`?
]

]

.pull-right[

```{r, echo = F}
dagify(wage~abil+occup+discr,
       discr~gender,
       occup~gender+discr,
       exposure = "disc",
       outcome = "wage") %>% 
  ggdag_parents("wage", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# Collider Example: Gender Discrimination I

.pull-left[

- If leave occupation uncontrolled for: ignore non-discriminatory reasons different genders choose different occupations

- If we *control* for occupation:
  - opens backdoor paths 4 & 5!
  - creates spurious `cor(abil,discr)` where there wasn't one!
  - also closes causal path 2! (discriminatory reasons genders choose different occupations)

- **Can't identify causal effect** by controlling for `occup`!

- Research on this needs to get clever!

]

.pull-right[

```{r, echo = F}
dagify(wage~abil+occup+discr,
       discr~gender,
       occup~gender+discr,
       exposure = "disc",
       outcome = "wage") %>% 
  ggdag_parents("wage", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# The Front Door Criterion: Mediators

.pull-left[

- Another case where controlling for a variable actually *adds bias* is if that variable is known as a .shout["mediator"].

.content-box-green[
.green[**Example**]:

- $X \rightarrow M \rightarrow Y$ (front-door)
- $X \leftarrow A \rightarrow Y$ (back-door)
- $X \leftarrow B \rightarrow Y$ (back-door)

- Should we control for $M$? 
- If we did, this would block the front-door!

- If we can estimate $X \rightarrow M$ and $M \rightarrow Y$ (note, no back-doors to either of these!), we can estimate $X \rightarrow Y$
]

- This is the .whisper[front door method]

]

.pull-right[

```{r, echo = F}
dagify(Y~A+B+M,
       M~X,
       X~A+B,
       coords=list(
         x=c(X=1, A=1.5, B=1.5, M=1.5, Y=2),
         y=c(X=1, A=2, B=0, M=1, Y=1)
       )) %>% 
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

---

# Recap

.pull-left[

Thus, to achieve identification, control for the minimal amount of variables such that:

1. Ensure that no back-door path remains open
    - Close back-door paths by controlling for a variable along that path
    - Colliders along a path automatically close that path

2. Ensure that no front-door path is closed

]

.pull-right[

```{r, echo = F}
dagify(Y~A+B+M,
       M~X,
       X~A+B,
       coords=list(
         x=c(X=1, A=1.5, B=1.5, M=1.5, Y=2),
         y=c(X=1, A=2, B=0, M=1, Y=1)
       )) %>% 
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

]

### 2019-11-05 static/slides/17-slides.Rmd
Fix gender example in slides 17

@@
 
-1. $gender \rightarrow discr \rightarrow wage$ (Causal path)
-2. $gender \rightarrow discr \rightarrow occup \rightarrow wage$ (Causal path)
+1. $discr \rightarrow wage$ (Causal path)
+2. $discr \rightarrow occup \rightarrow wage$ (Causal path)
 3. $discr \leftarrow gender \rightarrow occup \rightarrow wage$ (Open back door)
 4. $discr \leftarrow gender \rightarrow occup \leftarrow abil \rightarrow wage$ (BLOCKED back door!)
-5. $gender \rightarrow discrim \rightarrow occup \leftarrow abil \rightarrow wage$ (BLOCKED back door!)
+5. $discr \rightarrow occup \leftarrow abil \rightarrow wage$ (BLOCKED back door!)
 
@@
        discr~gender,
-       occup~gender+discr,
+       occup~gender+discr+abil,
        exposure = "disc",
@@
 
-# Collider Example: Gender Discrimination I
-
-.pull-left[
-
-- If leave occupation uncontrolled for: ignore non-discriminatory reasons different genders choose different occupations
+# Collider Example: Gender Discrimination II
+
+.pull-left[
 
@@
 
-- **Can't identify causal effect** by controlling for `occup`!
+- **Can't identify causal effect** controlling for `occup`!
+  - *Could* if you could control for `occup` *and* `abil` (but perhaps `abil` is unobserved)
 

### 2019-11-07 content/r/17-r-practice.Rmd
Update class 17, add practice 17

---
title: "3.3 OLS: Model Specification Strategies - R Practice"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

# Set Up
To minimize confusion, I suggest creating a new `R Project` (e.g. `regression_practice`) and storing any data in that folder on your computer.

Alternatively, I have made a project in R Studio Cloud that you can use (and not worry about trading room computer limitations), with the data already inside (you will still need to assign it to an object).

- [<i class="fas fa-cloud"></i> View Project on R Studio Cloud](https://rstudio.cloud/project/692655)
- [<i class="fab fa-markdown"></i> `Download Markdown`](/r/17-r-practice-rmd.Rmd)

# Part A DAG Practice

For each of the following DAGs:^[You can use [`dagitty.net`](http://dagitty.net) to help you, but you should start trying to recognize these on your own!]

- i. Write out *all* of the causal pathways from `X` (treatment of interest) to `Y` (outcome of interest).
- ii. Write at least one testable implication of the model.^[i.e. what variable(s) should be independent of another? or what should be independent of another *once* controlling for other variables?]
- iii. Identify which variable(s) need to be controlled for to estimate the causal effect of `X` on `Y`.
- iv. Write out the regression equation (abstractly) for properly identifying the causal effect, based on part iii.

```{r, echo=F, message=F, warning=F}
# load to draw DAGs! 
library(tidyverse)
library(ggdag)
```

## Example A

```{r, fig.retina=3}
dagify(Y~X+Z,
       X~Z,
       exposure = "X",
       outcome = "Y") %>% 
  ggdag(stylized = FALSE, seed=1)+theme_dag_blank()+theme(legend.position = "none")
```

## Example B

```{r, fig.retina=3}
dagify(Y~X+M,
       M~X,
       exposure = "X",
       outcome = "Y") %>% 
  ggdag(stylized = FALSE, seed=1)+theme_dag_blank()+theme(legend.position = "none")
```

## Example C

```{r, fig.retina=3}
dagify(Y~X+Z+B,
       Z~B+A,
       X~A,
       exposure = "X",
       outcome = "Y") %>% 
  ggdag(stylized = FALSE, seed=1)+theme_dag_blank()+theme(legend.position = "none")
```

## Example D

```{r, fig.retina=3}
dagify(Y~X+B+C,
       Z~B+A,
       X~A,
       C~X,
       exposure = "X",
       outcome = "Y") %>% 
  ggdag(stylized = FALSE, seed=2)+theme_dag_blank()+theme(legend.position = "none")
```

## Example E

```{r, fig.retina=3}
dagify(Y~X+Z+A,
       Z~X+A+B,
       B~A,
       exposure = "X",
       outcome = "Y") %>% 
  ggdag(stylized = FALSE, seed=2)+theme_dag_blank()+theme(legend.position = "none")
```

# Part B Empirical Example

## Question 1

Install the `wooldridge` package if you do not already have it installed.^[This package contains datasets used in Jeffrey Wooldrige's *Introductory Econometrics: A Modern Approach* (the textbook that I used in *my* econometrics classes years ago!)] We will use the `bwght` data from `wooldridge`^[Which comes from The 1988 National Health Interview Survey., used in J. Mullahy (1997), “Instrumental-Variable Estimation of Count Data Models: Applications to Models of Cigarette Smoking Behavior,” *Review of Economics and Statistics* 79: 596-593.]

Let's just look at the following variables:

| Variable | Description |
|----------|-------------|
| `bwght` | Birth Weight (ounces) |
| `cigs` | Cigarettes smoked per day while pregnant (1988) |
| `motheduc` | Mother's education (number of years) |
| `cigprice` | Price of cigarette pack (1988) |
| `faminc` | Family's income in $1,000s (1988) |

> We want to explore how a mother smoking during pregnancy affects the baby's birthweight (which may have strong effects on outcomes over the child's life).

Just to be explicit about it, assign this as some dataframe (feel free to change the name), i.e.:

```{r}
# install.packages("wooldridge")
library(wooldridge)
bwght<-wooldridge::bwght
```

## Question 2
Make a correlation table for our variables listed above.^[Hints: `select()` these variables and then pipe them into `cor(., use="pairwise.complete.obs")` to use only observations for which there are data on each variable (to avoid `NA`'s).]

```{r}
library(corrplot)
bwght %>%
  select(bwght, cigs, faminc, cigprice, motheduc) %>%
  cor(., use="pairwise.complete.obs")
```

## Question 3
Consider the following causal model:

```{r}
library(ggdag)
dagify(bwght~cigs+inc,
       cigs~price+educ+inc,
       inc~educ,
       exposure = "cigs",
       outcome = "bwght") %>% 
  ggdag(stylized = FALSE, seed=1)+theme_dag_blank()+theme(legend.position = "none")
```

Note what we are hypothesizing:
  - `bwght` is caused by `cigs` and `inc`
  - `cigs` are caused by `price`, `educ`, and `inc`
  - `inc` is caused by `educ`
See also how this is written into the notation in R to make the DAG.

Create this model on [dagitty.net](htpp://dagitty.net).

## Question 4
What does `dagitty` tell us the testable implications of this causal model?

## Question 5
Test each implication given to you by `dagitty.`

- For independencies $(x \perp y)$: run a regression of $y$ on $x$.
- For *conditional* independencies $(x \per y | z, a)$: run a regression of $y$ on $x, z, a$.

For each, test against the null hypothesis that the relevant coefficient $(\hat{\beta_1})$ is 0 (i.e. $x$ and $y$ are indeed independent).

Does this causal model hold up well?

```{r}
lm(bwght~motheduc+cigs+faminc, data = bwght) %>% summary()
```

```{r}
lm(bwght~cigprice+cigs+faminc, data = bwght) %>% summary()
```

```{r}
lm(motheduc~cigprice, data = bwght) %>% summary()
```

```{r}
lm(faminc~cigprice, data = bwght) %>% summary()
```

## Question 6
List *all* of the possible pathways from `cigs` to `bwght`. Which are "front-doors" and which are "back-doors?" Are any blocked by colliders?

## Question 7
What is the minimal sufficient set of variables we need to control in order to causally identify the effect of `cigs` on `bwght`?

## Question 8
Estimate the causal effect by running the appropriate regression.^[Note, on `dagitty`, you can change a variable on the diagram to be "*adjusted*" (controlled for) by double-clicking it and then hitting the `A` key.]

## Question 9
We saw some effect between `faminc` and `cigprice`. Perhaps there are unobserved factors (such as the economy's performance) that affect both. Add an unobserved factor `u1` to your `dagitty` model.^[Note, on `dagitty`, you can make a variable be "*unobserved*" by double-clicking it and then hitting the `U` key.]

## Question 10
Perhaps our model is poorly specified. Maybe `motheduc` actually has a causal effect on `bwght`? Tweak your model from Question 9 on `dagitty` to add this potential relationship. What testable implications does this new model imply?

## Question 11
Test these implications appropriately, like you did in Question 5.

```{r}
lm(bwght~cigprice+cigs+motheduc+faminc, data = bwght) %>% summary()
```

```{r}
lm(motheduc~cigprice, data = bwght) %>% summary()
```

Does this model hold up well?

## Question 12
Under this new causal model, list *all* of the possible pathways from `cigs` to `bwght`. Which are "front-doors" and which are "back-doors?" Are any blocked by colliders?

## Question 13
Under this new causal model, what is the minimal sufficient set of variables we need to control in order to causally identify the effect of `cigs` on `bwght`?

## Question 14
Estimate the causal effect in this new model by running the appropriate regression. Compare your answers to those in question 8.

## Question 15
Try out drawing this model using the  `ggdag` package in R. See my DAG in question 3 for an example.

### 2019-11-07 static/r/17-r-practice.Rmd
Update class 17, add practice 17

--- 2019-11-07 content/r/17-r-practice.Rmd
@@
 ---
-title: "3.3 OLS: Model Specification Strategies - R Practice"
-editor_options: 
-  chunk_output_type: console
+title: "3.5: Causal Inference and DAGs - R Practice"
+author: "YOUR NAME"
+date: "`r Sys.Date()`"
+output:
+  html_document:
+    df_print: paged
+    #theme: 
+    toc: true 
+    toc_depth: 3
+    toc_float: true
+    code_folding: show
+    highlight: tango
 ---
 
-```{r message=FALSE, warning=FALSE, include=FALSE}
-library(tidyverse)
-library(pander)
-source(file.path(here::here(), "lib", "graphics.R"))
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
 ```
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
 ```{r}
-library(corrplot)
-bwght %>%
-  select(bwght, cigs, faminc, cigprice, motheduc) %>%
-  cor(., use="pairwise.complete.obs")
+# PUT CODE HERE
 ```
 
+---
+
 ## Question 3
+
 Consider the following causal model:
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 4
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 5
+
 Test each implication given to you by `dagitty.`
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
 ```{r}
-lm(bwght~motheduc+cigs+faminc, data = bwght) %>% summary()
+# PUT CODE HERE
 ```
 
-```{r}
-lm(bwght~cigprice+cigs+faminc, data = bwght) %>% summary()
-```
-
-```{r}
-lm(motheduc~cigprice, data = bwght) %>% summary()
-```
-
-```{r}
-lm(faminc~cigprice, data = bwght) %>% summary()
-```
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 7
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 8
 Estimate the causal effect by running the appropriate regression.^[Note, on `dagitty`, you can change a variable on the diagram to be "*adjusted*" (controlled for) by double-clicking it and then hitting the `A` key.]
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 Add an unobserved factor `u1` to your `dagitty` model.^[Note, on `dagitty`, you can make a variable be "*unobserved*" by double-clicking it and then hitting the `U` key.]
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
-## Question 11
-Test these implications appropriately, like you did in Question 5.
+---
+
+<!--ANSWER BELOW HERE-->
 
 ```{r}
-lm(bwght~cigprice+cigs+motheduc+faminc, data = bwght) %>% summary()
+# PUT CODE HERE
 ```
 
+---
+
+## Question 11
+Test these implications appropriately, like you did in Question 5. Does this model hold up well?
+
+---
+
+<!--ANSWER BELOW HERE-->
+
 ```{r}
-lm(motheduc~cigprice, data = bwght) %>% summary()
+# PUT CODE HERE
 ```
 
-Does this model hold up well?
+---
 
 ## Question 12
+
 Under this new causal model, list *all* of the possible pathways from `cigs` to `bwght`. Which are "front-doors" and which are "back-doors?" Are any blocked by colliders?
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 13
+
 Under this new causal model, what is the minimal sufficient set of variables we need to control in order to causally identify the effect of `cigs` on `bwght`?
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 14
+
 Estimate the causal effect in this new model by running the appropriate regression. Compare your answers to those in question 8.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 Try out drawing this model using the  `ggdag` package in R. See my DAG in question 3 for an example.
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```

### 2019-11-07 static/slides/17-slides.Rmd
Update class 17, add practice 17

@@
                y=c(tech=1, ip=2, profit=1)
-               )) %>% ggdag(stylized = FALSE)+
+               )) %>% ggdag(stylized = FALSE, seed=256)+
   labs(title = "1")+
@@
                y=c(tech=1, ip=2, profit=1)
-               )) %>% ggdag(stylized = FALSE)+
+               )) %>% ggdag(stylized = FALSE, seed=256)+
   labs(title = "2")+
@@
                y=c(tech=1, ip=2, profit=1)
-               )) %>% ggdag(stylized = FALSE)+
+               )) %>% ggdag(stylized = FALSE, seed=256)+
   labs(title = "3")+
@@
                y=c(tech=1, ip=2, profit=1)
-               )) %>% ggdag(stylized = FALSE)+
+               )) %>% ggdag(stylized = FALSE, seed=256)+
   labs(title = "4")+
@@
                y=c(tech=1, ip=2, profit=1)
-               )) %>% ggdag(stylized = FALSE)+
+               )) %>% ggdag(stylized = FALSE, seed=256)+
   labs(title = "5")+
@@
                y=c(tech=1, ip=2, profit=1)
-               )) %>% ggdag(stylized = FALSE)+
+               )) %>% ggdag(stylized = FALSE, seed=256)+
   labs(title = "6")+
@@
 dagify(wage~educ) %>% 
-  ggdag_parents("wage", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
+  ggdag_parents("wage", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
 ```
@@
        educ~bckg+year+loc+laws) %>%
-  ggdag_parents("wage", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
+  ggdag_parents("wage", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
 ```
@@
        loc~u1) %>% 
-  ggdag_parents("wage", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
+  ggdag_parents("wage", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
 ```
@@
        loc~u1) %>% 
-  ggdag(stylized = FALSE)+
+  ggdag(stylized = FALSE, seed=256)+
   theme_dag_blank()
@@
   tidy_dagitty() %>%
-  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
+  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
 ```
@@
   #tidy_dagitty() %>%
-  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
+  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
 ```
@@
        B~A) %>% 
-  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
+  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
 ```
@@
        )) %>% 
-  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
+  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
 ```
@@
        )) %>% 
-  ggdag_parents("Hos", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
+  ggdag_parents("Hos", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
 ```
@@
        outcome = "wage") %>% 
-  ggdag_parents("wage", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
+  ggdag_parents("wage", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
 ```
@@
        )) %>% 
-  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
+  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
+```
+
+]
+
+---
+
+# The Front Door Criterion: Mediators
+
+.pull-left[
+
+- Tobacco industry claimed that $cor(smoking, cancer)$ could be spurious due to a confounding `gene` that affects both!
+  - Smoking `gene` is unobservable
+
+- Suppose smoking causes `tar` buildup in lungs, which cause `cancer`
+
+- We should *not* control for `tar`, it's on the **front-door path**
+  - This is how scientific studies can relate smoking to cancer
+
+]
+
+.pull-right[
+
+```{r, echo = F}
+dagify(canc~tar+gene,
+       tar~smok,
+       smok~gene,
+       coords=list(
+         x=c(smok=1, tar=1.5, gene=1.5, canc=2),
+         y=c(smok=1, tar=1, gene=2, canc=1)
+       )) %>% 
+  ggdag(stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
 ```
@@
 
-1. Ensure that no back-door path remains open
-    - Close back-door paths by controlling for a variable along that path
-    - Colliders along a path automatically close that path
-
-2. Ensure that no front-door path is closed
+1. Ensure that **no back-door path remains open**
+    - Close back-door paths by *controlling* for any variable along that path
+    - Colliders along a path *automatically* close that path
+
+2. Ensure that **no front-door path is closed**
 
@@
        )) %>% 
-  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
-```
-
-]
+  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
+```
+
+]
+
+---
+
+# What are DAGs Good For?
+
+- Omitted variable bias
+
+- Help visualize how later methods (difference-in-difference, fixed effects, regression discontinuity, instrumental variables) help identify causal effectsp

### 2019-11-07 content/r/17-r-practice.Rmd
Fix link to practice 17, fix webpage for practice 17

@@
 ---
-title: "3.3 OLS: Model Specification Strategies - R Practice"
+title: "3.5: Causal Inference and DAGs - R Practice"
 editor_options: 
@@
 
-```{r, fig.retina=3}
+```{r, echo=F, fig.retina=3}
 dagify(Y~X+Z,
@@
 
-```{r, fig.retina=3}
+```{r, echo=F, fig.retina=3}
 dagify(Y~X+M,
@@
 
-```{r, fig.retina=3}
+```{r, echo=F, fig.retina=3}
 dagify(Y~X+Z+B,
@@
 
-```{r, fig.retina=3}
+```{r, echo=F, fig.retina=3}
 dagify(Y~X+B+C,
@@
 
-```{r, fig.retina=3}
+```{r, echo=F, fig.retina=3}
 dagify(Y~X+Z+A,
@@
 
-```{r}
-library(corrplot)
-bwght %>%
-  select(bwght, cigs, faminc, cigprice, motheduc) %>%
-  cor(., use="pairwise.complete.obs")
-```
-
 ## Question 3
@@
 
-```{r}
+```{r, echo=T}
 library(ggdag)
@@
 - For independencies $(x \perp y)$: run a regression of $y$ on $x$.
-- For *conditional* independencies $(x \per y | z, a)$: run a regression of $y$ on $x, z, a$.
+- For *conditional* independencies $(x \perp y | z, a)$: run a regression of $y$ on $x, z, a$.
 
@@
 Does this causal model hold up well?
-
-```{r}
-lm(bwght~motheduc+cigs+faminc, data = bwght) %>% summary()
-```
-
-```{r}
-lm(bwght~cigprice+cigs+faminc, data = bwght) %>% summary()
-```
-
-```{r}
-lm(motheduc~cigprice, data = bwght) %>% summary()
-```
-
-```{r}
-lm(faminc~cigprice, data = bwght) %>% summary()
-```
 
@@
 ## Question 11
-Test these implications appropriately, like you did in Question 5.
-
-```{r}
-lm(bwght~cigprice+cigs+motheduc+faminc, data = bwght) %>% summary()
-```
-
-```{r}
-lm(motheduc~cigprice, data = bwght) %>% summary()
-```
-
-Does this model hold up well?
+Test these implications appropriately, like you did in Question 5. Does this model hold up well?
 

### 2019-11-07 content/r/17-r-practice.Rmd
Final fixes to practice 17 content, slides 17

@@
 Note what we are hypothesizing:
+
 - `bwght` is caused by `cigs` and `inc`
@@
 - `inc` is caused by `educ`
+
 See also how this is written into the notation in R to make the DAG.

### 2019-11-07 static/slides/17-slides.Rmd
Final fixes to practice 17 content, slides 17

@@
 2. $discr \rightarrow occup \rightarrow wage$ (Causal path)
-3. $discr \leftarrow gender \rightarrow occup \rightarrow wage$ (Open back door)
+3. $discr \leftarrow gender \rightarrow occup \rightarrow wage$ (BLOCKED back door!)
 4. $discr \leftarrow gender \rightarrow occup \leftarrow abil \rightarrow wage$ (BLOCKED back door!)
@@
 - If we *control* for occupation:
-  - opens backdoor paths 4 & 5!
+  - opens backdoor paths 3, 4, 5!
   - creates spurious `cor(abil,discr)` where there wasn't one!
@@
 
-- **Can't identify causal effect** controlling for `occup`!
+- **Can't identify causal effect** controlling for `occup` alone!
   - *Could* if you could control for `occup` *and* `abil` (but perhaps `abil` is unobserved)
-
-- Research on this needs to get clever!
 
@@
 
-# The Front Door Criterion: Mediators
+# The Front Door Criterion: Mediators I
 
@@
 
-- If we can estimate $X \rightarrow M$ and $M \rightarrow Y$ (note, no back-doors to either of these!), we can estimate $X \rightarrow Y$
-]
-
-- This is the .whisper[front door method]
-
+]
 ]
@@
 
-# The Front Door Criterion: Mediators
+# The Front Door Criterion: Mediators II
+
+.pull-left[
+
+- Another case where controlling for a variable actually *adds bias* is if that variable is known as a .shout["mediator"].
+
+.content-box-green[
+.green[**Example**]:
+
+- If we control for $M$, this would block the front-door!
+
+- If we can estimate $X \rightarrow M$ and $M \rightarrow Y$ (note, no back-doors to either of these!), we can estimate $X \rightarrow Y$
+]
+
+- This is the .whisper[front door method]
+
+]
+
+.pull-right[
+
+```{r, echo = F}
+dagify(Y~A+B+M,
+       M~X,
+       X~A+B,
+       coords=list(
+         x=c(X=1, A=1.5, B=1.5, M=1.5, Y=2),
+         y=c(X=1, A=2, B=0, M=1, Y=1)
+       )) %>% 
+  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
+```
+
+]
+
+---
+
+# The Front Door Criterion: Mediators III
 
@@
 
-Thus, to achieve identification, control for the minimal amount of variables such that:
+Thus, to achieve .onfire[causal identification], control for the minimal amount of variables such that:
 
@@
 ]
-
----
-
-# What are DAGs Good For?
-
-- Omitted variable bias
-
-- Help visualize how later methods (difference-in-difference, fixed effects, regression discontinuity, instrumental variables) help identify causal effectsp

### 2019-11-12 content/class/18-class.Rmd
Add class 18 content

---
title: "3.6: Regression with Categorical Data - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

*Tuesday, November 12, 2019*

# Overview

Today we look at data that is categorical.

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/18-slides.html)

# Problem Set 4 Due Tuesday Nov 19

[Problm Set 4](/assignment/04-problem-set) (on classes 3.1-3.5) is due by Tuesday November 19.

### 2019-11-12 static/slides/18-slides.Rmd
Add class 18 content

---
title: "3.6: Regression with Categorical Data"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))

library(wooldridge)
wages<-wooldridge::wage1
wages<-wages %>%
  mutate(Gender = factor(ifelse(female==0,
                         "Male",
                         "Female")))
```

# Categorical Data

.pull-left[

- .hl[Categorical data] place an individual into one of several possible *categories*
    - e.g. sex, season, political party
    - may be responses to survey questions
    - can be quantitative (e.g. age, zip code)

- `R` calls these `factors` (we'll deal with them much later in the course)
]

.pull-right[

![](https://www.dropbox.com/s/pgdswagr7b348vu/categoricaldata.png?raw=1)

]

---

# Example Research Question

.pull-left[

.content-box-green[
.green[**Example**]: do men earn higher wages, on average, than women? If so, how much?
]

]

.pull-right[
.center[
![:scale 100%](https://www.dropbox.com/s/481uod9g3n1w7kr/genderpaygap.jpg?raw=1)
]
]

---

# The Pure Statistics of Comparing Group Means

.pull-left[

- Using basic statistics, we can test for a statistically significant difference in group means with a **t-test**<sup>.red[1]</sup>

- Define:

- .blue[`\\(Y_M\\)`]: average earnings of a sample of .blue[`\\(n_M\\)`] men

- .magenta[`\\(Y_W\\)`]: average earnings of a sample of .magenta[`\\(n_W\\)`] women

- **Difference** in group averages: $d=$ .blue[`\\(\bar{Y}_M\\)`] $-$ .magenta[`\\(\bar{Y}_W\\)`]

- The hypothesis test is:
  - $H_0: d=0$
  - $H_1: d \neq 0$

]

.pull-right[
.center[
![:scale 100%](https://www.dropbox.com/s/481uod9g3n1w7kr/genderpaygap.jpg?raw=1)
]
]

.footnote[ <sup>.red[1]</sup> See today's class notes page for this example]

---

# Comparing Groups with Regression

- In a regression, we can easily compare across groups via a .shout[dummy variable]<sup>.red[1]</sup>

- Dummy variable *only* $=0$ or $=1$, depending on if a condition is met

- Signifies whether an observation belongs to a category or not

.footnote[<sup>.red[1]</sup> Also called a .shout[binary variable] or .shout[dichotomous variable]]

--

.content-box-green[
.green[**Example**]:

$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Female_i \quad \quad \text{ where } Female_i =
 		 \begin{cases}
    		1 & \text{if person } i \text{ is } Female \\
   			0 & \text{if person } i \text{ is } Male\\
  		\end{cases}$$

]

--

- Again, $\hat{\beta_1}$ makes less sense as the "slope" of a line in this context

---

# Comparing Groups in Regression: Scatterplot

.pull-left[

- `Female` is our dummy $x$-variable

- Hard to see relationships because of **overplotting**

]

.pull-right[
```{r, fig.retina=3}
colors<-c(Female = "pink", Male = "blue")

ggplot(data = wages)+
  aes(x = female,
      y = wage)+
  geom_point(aes(color = Gender))+
  geom_smooth(method = "lm", color = "black")+
  scale_x_continuous(breaks=c(0,1),
                   labels=c(0,1))+
  scale_y_continuous(labels=scales::dollar)+
  labs(x = "Female",
       y = "Wage")+
  guides(color=F)+ # hide legend
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

---

# Comparing Groups in Regression: Scatterplot

.pull-left[

- `Female` is our dummy $x$-variable

- Hard to see relationships because of **overplotting**

- Use `geom_jitter()` instead of `geom_point()` to *randomly* nudge points
  - *Only* for plotting purposes, does not affect actual data, regression, etc.

]

.pull-right[
```{r, fig.retina=3}
colors<-c(Female = "pink", Male = "blue")

ggplot(data = wages)+
  aes(x = female,
      y = wage)+
  geom_jitter(aes(color = Gender),
              width=0.05)+
  geom_smooth(method = "lm", color = "black")+
  scale_x_continuous(breaks=c(0,1),
                   labels=c(0,1))+
  scale_y_continuous(labels=scales::dollar)+
  labs(x = "Female",
       y = "Wage")+
  guides(color=F)+ # hide legend
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

---

# Comparing Groups in Regression: Scatterplot

.pull-left[

- `Female` is our dummy $x$-variable

- Hard to see relationships because of **overplotting**

- Use `geom_jitter()` instead of `geom_point()` to *randomly* nudge points
  - *Only* for plotting purposes, does not affect actual data, regression, etc.

]

.pull-right[
```{r, fig.retina=3}
male_avg<-wages %>%
  filter(female==0) %>%
  summarize(mean(wage)) %>%
  pull() # get just a number (not a dataframe)

female_avg<-wages %>%
  filter(female==1) %>%
  summarize(mean(wage)) %>%
  pull() # get just a number (not a dataframe)

ggplot(data = wages)+
  aes(x = female,
      y = wage)+
  geom_jitter(aes(color = Gender),
              width=0.05)+
  geom_smooth(method = "lm", color = "black")+
  geom_hline(yintercept=male_avg,linetype="dashed",color="blue")+
  geom_label(x = 0.5, y=male_avg, label="Average for Men", color="blue")+
  geom_hline(yintercept=female_avg,linetype="dashed",color="red")+
  geom_label(x = 0.5, y=female_avg, label="Average for Women", color="red")+
  scale_x_continuous(breaks=c(0,1),
                   labels=c(0,1))+
  scale_y_continuous(labels=scales::dollar)+
  labs(x = "Female",
       y = "Wage")+
  guides(color=F)+ # hide legend
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

---

# Dummy Variables as Group Means

$$\hat{Y_i}=\hat{\beta_0}+\hat{\beta_1} D_i \quad \text{ where }D_i=\{0,1\}$$

--

- .red[When `\\(D_i=0\\)` (Control group):]
    - $\hat{Y_i}=\hat{\beta_0}$
    - $E[Y|D_i=0]=\hat{\beta_0}$ $\iff$ the mean of $Y$ when $D_i=0$
--

- .blue[When `\\(D_i=1\\)` (Treatment group):]
    - $\hat{Y_i}=\hat{\beta_0}+\hat{\beta_1} D_i$
    - $E[Y|D_i=1]=\hat{\beta_0}+\hat{\beta_1}$ $\iff$ the mean of $Y$ when $D_i=1$

--

- So the **difference** in group means:

$$\begin{align*}
		&=E[Y_i|D_i=1]-E[Y_i|D_i=0]\\
		&=(\hat{\beta_0}+\hat{\beta_1})-(\hat{\beta_0})\\
		&=\hat{\beta_1}\\
\end{align*}$$

---

# Dummy Variables as Group Means: Our Example

.pull-left[
.content-box-green[
.green[**Example**]:

$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Female_i$$

$$\text{where } Female_i =
 		 \begin{cases}
    		1 & \text{if } i \text{ is }Female \\
   			0 & \text{if } i \text{ is } Male\\
  		\end{cases}$$
]
]

.pull-right[

- Mean wage for men: 
]

---

# Dummy Variables as Group Means: Our Example

.pull-left[
.content-box-green[
.green[**Example**]:

$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Female_i$$

$$\text{where } Female_i =
 		 \begin{cases}
    		1 & \text{if } i \text{ is }Female \\
   			0 & \text{if } i \text{ is } Male\\
  		\end{cases}$$
]
]

.pull-right[

- Mean wage for men:
$$E[Wage|Female=0]=\hat{\beta_0}$$
]

---

# Dummy Variables as Group Means: Our Example

.pull-left[
.content-box-green[
.green[**Example**]:

$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Female_i$$

$$\text{where } Female_i =
 		 \begin{cases}
    		1 & \text{if } i \text{ is }Female \\
   			0 & \text{if } i \text{ is } Male\\
  		\end{cases}$$
]
]

.pull-right[

- Mean wage for men: 
$$E[Wage|Female=0]=\hat{\beta_0}$$

- Mean wage for women: 
]

---

# Dummy Variables as Group Means: Our Example

.pull-left[
.content-box-green[
.green[**Example**]:

$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Female_i$$

$$\text{where } Female_i =
 		 \begin{cases}
    		1 & \text{if } i \text{ is }Female \\
   			0 & \text{if } i \text{ is } Male\\
  		\end{cases}$$
]
]

.pull-right[

- Mean wage for men:
$$E[Wage|Female=0]=\hat{\beta_0}$$

- Mean wage for women:
$$E[Wage|Female=1]=\hat{\beta_0}+\hat{\beta_1}$$
]

---

# Dummy Variables as Group Means: Our Example

.pull-left[
.content-box-green[
.green[**Example**]:

$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Female_i$$

$$\text{where } Female_i =
 		 \begin{cases}
    		1 & \text{if } i \text{ is }Female \\
   			0 & \text{if } i \text{ is } Male\\
  		\end{cases}$$
]
]

.pull-right[

- Mean wage for men:
$$E[Wage|Female=0]=\hat{\beta_0}$$

- Mean wage for women:
$$E[Wage|Female=1]=\hat{\beta_0}+\hat{\beta_1}$$

- Difference in wage between men & women:
]

---

# Dummy Variables as Group Means: Our Example

.pull-left[
.content-box-green[
.green[**Example**]:

$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Female_i$$

$$\text{where } Female_i =
 		 \begin{cases}
    		1 & \text{if } i \text{ is }Female \\
   			0 & \text{if } i \text{ is } Male\\
  		\end{cases}$$
]
]

.pull-right[

- Mean wage for men: 
$$E[Wage|Female=0]=\hat{\beta_0}$$

- Mean wage for women: 
$$E[Wage|Female=1]=\hat{\beta_0}+\hat{\beta_1}$$

- Difference in wage between men & women: 
$$d = \hat{\beta_1}$$
]

---

# Comparing Groups in Regression: Scatterplot

.pull-left[

- `Female` is our dummy $x$-variable

- Hard to see relationships because of **overplotting**

- Use `geom_jitter()` instead of `geom_point()` to *randomly* nudge points
  - *Only* for plotting purposes, does not affect actual data, regression, etc.

]

.pull-right[
```{r, fig.retina=3}
ggplot(data = wages)+
  aes(x = female,
      y = wage)+
  geom_jitter(aes(color = Gender),
              width=0.05)+
  geom_smooth(method = "lm", color="black")+
  geom_hline(yintercept=male_avg,linetype="dashed",color="blue")+
  geom_label(x = 0.5, y=male_avg, label=expression(hat(beta[0])), color="blue")+
  geom_hline(yintercept=female_avg,linetype="dashed",color="red")+
  geom_label(x = 0.5, y=female_avg, label=expression(hat(beta[0])+hat(beta[1])), color="red")+
  geom_label(x = 0.75, y=5.5, label=expression(paste(hat(beta[1]),"= difference")), color="black")+
  scale_x_continuous(breaks=c(0,1),
                   labels=c(0,1))+
  scale_y_continuous(labels=scales::dollar)+
  labs(x = "Female",
       y = "Wage")+
  guides(color=F)+ # hide legend
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

---

# The Data

```{r, echo=T}
# from wooldridge package
library(wooldridge)

# save as a dataframe
wages<-wooldridge::wage1

head(wages)
```

---

# Get Group Averages & Std. Devs.

.pull-left[

```{r, echo=T}
# Summarize for Men

wages %>%
  filter(female==0) %>%
  summarize(mean = mean(wage),
            sd = sd(wage))
```

]

.pull-right[

```{r, echo=T}
# Summarize for Women

wages %>%
  filter(female==1) %>%
  summarize(mean = mean(wage),
            sd = sd(wage))
```

]

---

# Visualize Differences

```{r, fig.retina=3}
fill_colors<-c("0" = "blue", "1" = "red")

ggplot(data = wages)+
  aes(x = wage,
      fill = factor(female))+
  geom_density(alpha=0.5)+
  scale_x_continuous(labels=scales::dollar)+
  labs(x = "wage",
       y = "Density")+
  scale_fill_manual("Gender", labels=c("Men","Women"), values = fill_colors)+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

---

# The Regression I

```{r, echo=T}
femalereg<-lm(wage~female, data=wages)
summary(femalereg)
```

---

# The Regression II

```{r, echo=T}
library(broom)
tidy(femalereg)
```

---

# Dummy Regression vs. Group Means

.pull-left[

From tabulation of group means

| Gender | Avg. Wage | Std. Dev.   | $n$   |
|--------|-------------|-----------|-------|
| Female | $4.59$    | $2.33$      | $252$ |
| Male   | $7.10$    | $4.16$      | $274$ |
| Difference | $2.51$ | $0.30$ | $-$ |

From $t$-test of difference in group means

]

.pull-right[
```{r}
tidy(femalereg)
```

]

---

class: inverse, center, middle

# Recoding Dummies

---

# Recoding Dummies

.content-box-green[
.green[**Example**]:

- Suppose instead of $female$ we had used:

$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Male_i \quad \quad \text{ where } Male_i =
 		 \begin{cases}
    		1 & \text{if person } i \text{ is } Male \\
   			0 & \text{if person } i \text{ is } Female\\
  		\end{cases}$$

]

---

# Recoding Dummies with Data

```{r}
wages<-wages %>%
  mutate(male = ifelse(female == 0, # condition: is female equal to 0?
                       1, # if true: code as "1"
                       0)) # if false: code as "0"

# verify it worked
wages %>% 
  select(wage, female, male) %>%
  head()
```

---

# Scatterplot with Male

.pull-left[

- `Female` is our dummy $x$-variable

- Hard to see relationships because of **overplotting**

- Use `geom_jitter()` instead of `geom_point()` to *randomly* nudge points
  - *Only* for plotting purposes, does not affect actual data, regression, etc.

]

.pull-right[
```{r, fig.retina=3}
wages<-wages %>%
  mutate(Gender = factor(ifelse(female==0,
                         "Male",
                         "Female")))

ggplot(data = wages)+
  aes(x = male,
      y = wage)+
  geom_jitter(aes(color = Gender),
              width=0.05)+
  geom_smooth(method = "lm", color = "black")+
  geom_hline(yintercept=male_avg,linetype="dashed",color="blue")+
  geom_label(x = 0.5, y=male_avg, label="Average for Men", color="blue")+
  geom_hline(yintercept=female_avg,linetype="dashed",color="red")+
  geom_label(x = 0.5, y=female_avg, label="Average for Women", color="red")+
  scale_x_continuous(breaks=c(0,1),
                   labels=c(0,1))+
  scale_y_continuous(labels=scales::dollar)+
  labs(x = "Male",
       y = "Wage")+
  guides(color=F)+ # hide legend
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

---

# Dummy Variables as Group Means: With Male

.pull-left[
.content-box-green[
.green[**Example**]:

$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Male_i$$

$$\text{where } Male_i =
 		 \begin{cases}
    		1 & \text{if } i \text{ is } Male \\
   			0 & \text{if } i \text{ is } Female\\
  		\end{cases}$$
]
]

.pull-right[

- Mean wage for men: 
$$E[Wage|Male=1]=\hat{\beta_0}+\hat{\beta_1}$$

- Mean wage for women: 
$$E[Wage|Male=0]=\hat{\beta_0}$$

- Difference in wage between men & women: 
$$d = \hat{\beta_1}$$
]

---

# Scatterplot with Male

.pull-left[

]

.pull-right[
```{r, fig.retina=3}
ggplot(data = wages)+
  aes(x = male,
      y = wage)+
  geom_jitter(aes(color = Gender),
              width=0.05)+
  geom_smooth(method = "lm", color="black")+
  geom_hline(yintercept=male_avg,linetype="dashed",color="blue")+
  geom_label(x = 0.5, y=male_avg, label=expression(hat(beta[0])+hat(beta[1])), color="blue")+
  geom_hline(yintercept=female_avg,linetype="dashed",color="red")+
  geom_label(x = 0.5, y=female_avg, label=expression(hat(beta[0])), color="red")+
  geom_label(x = 0.75, y=5.5, label=expression(paste(hat(beta[1]),"= difference")), color="black")+
  scale_x_continuous(breaks=c(0,1),
                   labels=c(0,1))+
  scale_y_continuous(labels=scales::dollar)+
  labs(x = "Male",
       y = "Wage")+
  guides(color=F)+ # hide legend
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

---

# The Regression with Male I

```{r, echo=T}
malereg<-lm(wage~male, data=wages)
summary(malereg)
```

---

# The Regression with Male II

```{r, echo=T}
library(broom)
tidy(malereg)
```

---

# The Dummy Regression: Male or Female

.pull-left[

```{r}
library(huxtable)
huxreg(malereg, femalereg,
       coefs = c("Constant" = "(Intercept)",
                 "Female" = "female",
                 "Male" = "male"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```

]

.pull-right[

- Note it doesn't matter if we use `male` or `female`, males always earn $2.51 more than females

- Compare the constant (average for the $D=0$ group)

- Should you use `male` AND `female`? We'll come to that...

]

---

class: inverse, center, middle

# Categorical Variables (More than 2 Categories)

---

# Categorical Variables with More than 2 Categories

- A .shout[categorical variable] expresses membership in a category, where there is no ranking or hierarchy of the categories
    - We've looked at categorical variables with 2 categories only
    - e.g. Male/Female, Spring/Summer/Fall/Winter, Democratic/Republican/Independent

--

- Might be an .shout[ordinal variable] expresses rank or an ordering of data, but not necessarily their relative magnitude
    - e.g. Order of finalists in a competition (1st, 2nd, 3rd)
    - e.g. Highest education attained (1=elementary school, 2=high school, 3=bachelor's degree, 4=graduate degree)

---

# Using Categorical Variables in Regression I

.content-box-green[
.green[**Example**]: How do wages vary by region of the country? Let $Region_i=\{Northeast, \, Midwest, \, South, \, West\}$
]

--

- Can we run the following regression?

$$\widehat{Wages_i}=\hat{\beta_0}+\hat{\beta_1}Region_i$$

---

# Using Categorical Variables in Regression II

.content-box-green[
.green[**Example**]: How do wages vary by region of the country?

Code region numerically: $$Region_i=
\begin{cases}1 & \text{if } i \text{ is in }Northeast \\
   			2 & \text{if } i \text{ is in } Midwest\\
    		3 & \text{if } i \text{ is in } South \\
   			4 & \text{if } i \text{ is in } West\\ \end{cases}$$
]

--

- Can we run the following regression?

$$\widehat{Wages_i}=\hat{\beta_0}+\hat{\beta_1}Region_i$$

---

# Using Categorical Variables in Regression III

.content-box-green[
.green[**Example**]: How do wages vary by region of the country?

Create a dummy variable for *each* region:

- $Northeast_i = 1$ if $i$ is in Northeast, otherwise $=0$
- $Midwest_i = 1$ if $i$ is in Midwest, otherwise $=0$
- $South_i = 1$ if $i$ is in South, otherwise $=0$
- $West_i = 1$ if $i$ is in West, otherwise $=0$

]

--

- Can we run the following regression?

$$\widehat{Wages_i}=\hat{\beta_0}+\hat{\beta_1}Northeast_i+\hat{\beta_2}Midwest_i+\hat{\beta_3}South_i+\hat{\beta_4}West_i$$

--

- For every $i: \, Northeast_i+Midwest_i+South_i+West_i=1$!

---

# The Dummy Variable Trap

.content-box-green[
.green[**Example**]: $\widehat{Wages_i}=\hat{\beta_0}+\hat{\beta_1}Northeast_i+\hat{\beta_2}Midwest_i+\hat{\beta_3}South_i+\hat{\beta_4}West_i$

]

- If we include *all* possible categories, they are .onfire[perfectly multicollinear], an exact linear function of one another:

$$Northeast_i+Midwest_i+South_i+West_i=1 \quad \forall i$$

- This is known as the .shout[dummy variable trap], a common source of perfect multicollinearity

---

# The Reference Category

- To avoid the dummy variable trap, always omit one category from the regression, known as the .onfire[reference category]

- It does not matter which category we omit!

- .onfire[Coefficients on each dummy variable measure the *difference* between the *reference* category and each category dummy]

---

# The Reference Category: Example

.content-box-green[
.green[**Example**]: $\widehat{Wages_i}=\hat{\beta_0}+\hat{\beta_1}Northeast_i+\hat{\beta_2}Midwest_i+\hat{\beta_3}South_i$
]

- $West_i$ is omitted (arbitrarily chosen)

--

- $\hat{\beta_0}$:

---

# The Reference Category: Example

.content-box-green[
.green[**Example**]: $\widehat{Wages_i}=\hat{\beta_0}+\hat{\beta_1}Northeast_i+\hat{\beta_2}Midwest_i+\hat{\beta_3}South_i$
]

- $West_i$ is omitted (arbitrarily chosen)

- $\hat{\beta_0}$: average wage for $i$ in the West

--

- $\hat{\beta_1}$:

---

# The Reference Category: Example

.content-box-green[
.green[**Example**]: $\widehat{Wages_i}=\hat{\beta_0}+\hat{\beta_1}Northeast_i+\hat{\beta_2}Midwest_i+\hat{\beta_3}South_i$
]

- $West_i$ is omitted (arbitrarily chosen)

- $\hat{\beta_0}$: average wage for $i$ in the West

- $\hat{\beta_1}$: difference between West and Northeast

--

- $\hat{\beta_2}$:

---

# The Reference Category: Example

.content-box-green[
.green[**Example**]: $\widehat{Wages_i}=\hat{\beta_0}+\hat{\beta_1}Northeast_i+\hat{\beta_2}Midwest_i+\hat{\beta_3}South_i$
]

- $West_i$ is omitted (arbitrarily chosen)

- $\hat{\beta_0}$: average wage for $i$ in the West

- $\hat{\beta_1}$: difference between West and Northeast

- $\hat{\beta_2}$: difference between West and Midwest

--

- $\hat{\beta_3}$:

---

# The Reference Category: Example

.content-box-green[
.green[**Example**]: $\widehat{Wages_i}=\hat{\beta_0}+\hat{\beta_1}Northeast_i+\hat{\beta_2}Midwest_i+\hat{\beta_3}South_i$
]

- $West_i$ is omitted (arbitrarily chosen)

- $\hat{\beta_0}$: average wage for $i$ in the West

- $\hat{\beta_1}$: difference between West and Northeast

- $\hat{\beta_2}$: difference between West and Midwest

- $\hat{\beta_3}$: difference between West and South

---

# Dummy Variable Trap in R

```{r}
# create northeast dummy as the final category
wages$noreast<-ifelse((wages$northcen==0 & wages$south==0 & wages$west==0),1,0)
```

.font90[
```{r, echo=T}
summary(lm(wage~noreast+northcen+south+west, data=wages))
```
]

---

# Using Different Reference Categories in R

```{r}
# create northeast dummy as the final category
wages$noreast<-ifelse((wages$northcen==0 & wages$south==0 & wages$west==0),1,0)
```

```{r dummyregs, echo=T, eval=F}
# let's run 4 regressions, each one we omit a different region
no_noreast_reg<-lm(wage~northcen+south+west, data = wages)
no_northcen_reg<-lm(wage~noreast+south+west, data = wages)
no_south_reg<-lm(wage~noreast+northcen+west, data = wages)
no_west_reg<-lm(wage~noreast+northcen+south, data = wages)

# now make an output table
library(huxtable)
huxreg(no_noreast_reg,
       no_northcen_reg,
       no_south_reg,
       no_west_reg,
       coefs = c("Constant" = "(Intercept)",
                 "Northeast" = "noreast",
                 "Midwest" = "northcen",
                 "South" = "south",
                 "West" = "west"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 3)
```

---

# Using Different Reference Categories in R II

.pull-left[
.font90[
```{r, ref.label="dummyregs"}

```
]
]
.pull-right[

- Constant is alsways average wage for reference (omitted) region

- Compare coefficients between Midwest in (1) and Northeast in (2)...

- Compare coefficients between West in (3) and South in (4)...

- Does not matter which region we omit!
  - Same $R^2$, SER, coefficients give same results

]

---

# Dummy *Dependent* (Y) Variables

- In many contexts, we will want to have our *dependent* ($Y$) variable be a dummy variable

--

.content-box-green[
.green[**Example**]: 
$$\widehat{Admitted_i}=\hat{\beta_0}+\hat{\beta_1}GPA_i \quad \text{ where } Admitted_i =
 		 \begin{cases}
    		1 & \text{if } i \text{ is Admitted} \\
   			0 & \text{if } i \text{ is  Not Admitted}\\
  		\end{cases}$$
]

--

- A model where $Y$ is a dummy is called a .shout[linear probability model], as it measures the .onfire[probability of `\\(Y\\)` occuring `\\((=1)\\)` given the X's, i.e. `\\(P(Y_i=1|X_1, \cdots, X_k)\\)`]
    - e.g. the probability person $i$ is Admitted to a program with a given GPA

--

- Requires special tools to properly interpret and extend this (**logit**, **probit**, etc)

- Feel free to write papers that have dummy $Y$ variables (but you may have to ask me some more questions)!

### 2019-11-12 content/class/18-class.Rmd
Add r practice 17 answers

@@
 
-Today we look at data that is categorical.
+Today we look at how to use data that is categorical (i.e.
+variables that indicate an observation's membership in a particular group or category).
+We introduce them into regression models as **dummy variables** that can equal 0 or 1: where 1 indicates membership in a category, and 0 indicates non-membership.
+
+We also look at what happens when categorical variables have more than two values: for regression, we introduce a dummy variable for each possible category - but be sure to leave out one reference category to avoid the dummy variable trap.
 

### 2019-11-12 content/r/17-r-practice.Rmd
Add r practice 17 answers

@@
 
+- [<i class="fas fa-external-link-square-alt"></i> `Answer Key (html)`](/r/17-r-practice-answers.html)
+- [<i class="fab fa-markdown"></i> `Answer Key (markdown)`](/r/17-r-practice-answers.Rmd)
+
 # Part A DAG Practice
@@
 - i. Write out *all* of the causal pathways from `X` (treatment of interest) to `Y` (outcome of interest).
-- ii. Write at least one testable implication of the model.^[i.e. what variable(s) should be independent of another? or what should be independent of another *once* controlling for other variables?]
-- iii. Identify which variable(s) need to be controlled for to estimate the causal effect of `X` on `Y`.
-- iv. Write out the regression equation (abstractly) for properly identifying the causal effect, based on part iii.
+- ii. Identify which variable(s) need to be controlled for to estimate the causal effect of `X` on `Y`.
+- iii. Write out the regression equation (abstractly) for properly identifying the causal effect, based on part ii.
 

### 2019-11-12 static/r/17-r-practice-answers.Rmd
Add r practice 17 answers

--- 2019-11-12 content/r/17-r-practice.Rmd
@@
 ---
-title: "3.5: Causal Inference and DAGs - R Practice"
-editor_options: 
-  chunk_output_type: console
+title: "3.5: Causal Inference and DAGs - R Practice (Answers)"
+author: "Ryan Safner"
+date: "`r Sys.Date()`"
+output:
+  html_document:
+    df_print: paged
+    #theme: 
+    toc: true 
+    toc_depth: 3
+    toc_float: true
+    code_folding: show
+    highlight: tango
 ---
 
-```{r message=FALSE, warning=FALSE, include=FALSE}
-library(tidyverse)
-library(pander)
-source(file.path(here::here(), "lib", "graphics.R"))
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
 ```
@@
 - [<i class="fab fa-markdown"></i> `Download Markdown`](/r/17-r-practice-rmd.Rmd)
-
-- [<i class="fas fa-external-link-square-alt"></i> `Answer Key (html)`](/r/17-r-practice-answers.html)
-- [<i class="fab fa-markdown"></i> `Answer Key (markdown)`](/r/17-r-practice-answers.Rmd)
 
@@
 
+---
+
+i. Pathways:
+  1. $X \rightarrow Y$ (causal, front door)
+  2. $X \leftarrow Z \rightarrow Y$ (not causal, back door)
+
+ii. $Z$ needs to be controlled for, since it is opening a backdoor path.
+
+iii. $Y=\beta_0+\beta_1X+\beta_2 Z$
+
+![](images/Adagitty.png)
+
+---
+
 ## Example B
@@
 ```
+
+---
+
+i. Pathways:
+  1. $X \rightarrow Y$ (causal, front door)
+  2. $X \rightarrow M \rightarrow Y$ (causal, front door)
+
+ii. Nothing should be controlled for, since $M$ is a mediator, and part of the effect of $X$ on $Y$
+
+iii. $Y=\beta_0+\beta_1X$
+
+![](images/Bdagitty.png)
+
+---
 
@@
 ```
+
+---
+
+i. Pathways:
+  1. $X \rightarrow Y$ (causal, front door)
+  2. $X \leftarrow A \rightarrow Z \rightarrow Y$ (not causal, back door)
+  3. $X \leftarrow A \rightarrow Z \leftarrow B \rightarrow Y$ (not causal, back door)
+
+ii.
+Backdoor path 3 is closed by the collider at $Z$.
+Backdoor path 2 remains open, so we need to control for $A$.
+(If we blocked $Z$ to close path 2, that would open up backdoor path 3!) Only $A$ should be controlled for.
+
+iii. $Y=\beta_0+\beta_1X+\beta_2A$
+
+![](images/Cdagitty.png)
+
+---
 
@@
 
+---
+
+i. Pathways:
+  1. $X \rightarrow Y$ (causal, front door)
+  2. $X \rightarrow C \rightarrow Y$ (causal, front door)
+  3. $X \leftarrow A \rightarrow Z \rightarrow B \rightarrow Y$ (not causal, back door)
+
+ii. Path 2 is a front door path we want to leave open. Backdoor path 3 is closed by the collider at $Z$. Nothing needs to be controlled for!
+
+iii. $Y=\beta_0+\beta_1X$
+
+![](images/Ddagitty.png)
+
+---
+
 ## Example E
@@
 ```
+
+---
+
+i. Pathways:
+  1. $X \rightarrow Y$ (causal, front door)
+  2. $X \rightarrow Z \rightarrow Y$ (causal, front door)
+  3. $X \rightarrow Z \leftarrow A \rightarrow Y$ (not causal, back door)
+  4. $X \rightarrow Z \leftarrow B \rightarrow A \rightarrow Y$ (not causal, back door)
+
+ii. Path 2 is a front door path we want to leave open. Backdoor path 3 is closed by the collider at $Z$. Backdoor path 4 is closed by the collider at $Z$. We don't want to control for anything!
+
+iii. $Y=\beta_0+\beta_1X$
+
+![](images/Edagitty.png)
+
+---
 
@@
 
+---
+
 ## Question 2
 Make a correlation table for our variables listed above.^[Hints: `select()` these variables and then pipe them into `cor(., use="pairwise.complete.obs")` to use only observations for which there are data on each variable (to avoid `NA`'s).]
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+bwght %>%
+  select(bwght, cigs, motheduc, cigprice, faminc) %>%
+  cor(use = "pairwise.complete.obs")
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+
+```
+
+---
+
 ## Question 4
@@
 
+---
+
+![](images/4dagitty.png)
+
+See the middle box on the right on dagitty:
+
+1. $bwght \perp price \, | \, cigs, inc$: birthweight is independent of price, controlling for cigarettes and income
+
+2. $bwght \perp educ \, | \, cigs, inc$: birthweight is independent of education, controlling for cigarettes and income
+
+3. $inc \perp price$: income is independent of price
+
+4. $price \perp educ$: cigarette price is independent of education
+
+---
+
 ## Question 5
+
 Test each implication given to you by `dagitty.`
@@
 
+---
+
+### Implication 1:
+
+If we run a regression of `bwght` on `cigprice`, including `cigs` and `faminc` as controls, there should not not be a statistically significant coefficient on `cigprice` (i.e.
+there is no relationship between `cigprice` and `bwght` holding `cigs` and` faminc` constant):
+
+```{r}
+lm(bwght~cigprice+cigs+faminc, data = bwght) %>% summary()
+```
+
+The coefficient on `cigprice` is small and not statistically significant. This implication holds up well.
+
+### Implication 2:
+
+If we run a regression of `bwght` on `motheduc`, including `cigs` and `faminc` as controls, there should not not be a statistically significant coefficient on `cigprice` (i.e.
+there is no relationship between `motheduc` and `bwght` holding `cigs` and` faminc` constant):
+
+```{r}
+lm(bwght~motheduc+cigs+faminc, data = bwght) %>% summary()
+```
+
+### Implication 3:
+
+The model implies simply that there is no significant correlation between `faminc` and `cigprice`
+
+```{r}
+bwght %>%
+  select(faminc, cigprice) %>%
+  cor()
+```
+
+There is a fairly weak correlation. This implication mostly holds up.
+
+### Implication 4:
+
+The model implies simply that there is no significant correlation between `cigprice` and `motheduc`
+
+```{r}
+bwght %>%
+  select(cigprice, motheduc) %>%
+  cor(use="pairwise.complete.obs")
+```
+
+This is an even weaker correlation. This implication seems to holds up.
+
+---
+
 ## Question 6
 List *all* of the possible pathways from `cigs` to `bwght`. Which are "front-doors" and which are "back-doors?" Are any blocked by colliders?
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+1. $cigs \rightarrow bwght$ (causal, front door)
+2. $cigs \leftarrow faminc \rightarrow bwght$ (non-causal, back door)
+3. $cibs \leftarrow motheduc \rightarrow faminc \rightarrow bwght$ (non-causal, back door)
+
+There are no colliders on any path.
+
+---
 
@@
 
+---
+
+![](images/7dagitty.png)
+
+We simply need to control for `faminc`. This blocks the back door for both path 2 and path 3.
+
+---
+
 ## Question 8
 Estimate the causal effect by running the appropriate regression.^[Note, on `dagitty`, you can change a variable on the diagram to be "*adjusted*" (controlled for) by double-clicking it and then hitting the `A` key.]
+
+---
+
+We need to control only for `faminc`, so we put it into the regression to estimate:
+
+$$bwght_i=\beta_0+\beta_1 \, cigs_i+\beta_2 \,faminc_i$$
+<!--ANSWER BELOW HERE-->
+
+```{r}
+lm(bwght~cigs+faminc, data = bwght) %>% summary()
+```
+
+Controlling for income, each cigarette smoked while pregant will cause the birthweight to decrease by 0.46 ounces.
+
+---
 
@@
 Add an unobserved factor `u1` to your `dagitty` model.^[Note, on `dagitty`, you can make a variable be "*unobserved*" by double-clicking it and then hitting the `U` key.]
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+![](images/9dagitty.png)
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+See the middle box on the right on dagitty:
+
+1. $bwght \perp price \, | \, cigs, inc, educ$: birthweight is independent of price, controlling for cigarettes, income, and education
+
+2. $price \perp educ$: cigarette price is independent of education
+
+![](images/10dagitty.png)
+
+---
+
 ## Question 11
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+### Implication 1:
+
+If we run a regression of `bwght` on `cigprice`, including `cigs`, `faminc`, and `motheduc` as controls, there should not not be a statistically significant coefficient on `cigprice` (i.e.
+there is no relationship between `cigprice` and `bwght` holding `cigs`,` faminc`, and `motheduc` constant):
+
+```{r}
+lm(bwght~cigprice+cigs+faminc+motheduc, data = bwght) %>% summary()
+```
+
+The coefficient on `cigprice` is small and not statistically significant. This implication holds up well.
+
+### Implication 2:
+
+This is the same as implication 4 from question 5. Again, this holds up reasonably well.
+
+---
+
 ## Question 12
+
 Under this new causal model, list *all* of the possible pathways from `cigs` to `bwght`. Which are "front-doors" and which are "back-doors?" Are any blocked by colliders?
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+1. $cigs \rightarrow bwght$ (causal, front door)
+2. $cigs \leftarrow faminc \rightarrow bwght$ (non-causal, back door)
+3. $cigs \leftarrow cigprice \leftarrow u1 \rightarrow faminc \rightarrow bwght$ (non-causal, back door)
+4. $cigs \leftarrow motheduc \rightarrow bwght$ (non-causal, back door)
+5. $cigs \leftarrow motheduc \rightarrow faminc \rightarrow bwght$ (non-causal, back door)
+
+There are no colliders on any path.
+
+---
+
 ## Question 13
+
 Under this new causal model, what is the minimal sufficient set of variables we need to control in order to causally identify the effect of `cigs` on `bwght`?
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+We need to control for `faminc` and `motheduc`. This blocks the back door for paths 2, 3, 4, and 5.
+
+---
+
 ## Question 14
+
 Estimate the causal effect in this new model by running the appropriate regression. Compare your answers to those in question 8.
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+![](images/14dagitty.png)
+
+We need to control for `faminc` and `motheduc`, so we put them into the regression to estimate:
+
+$$bwght_i=\beta_0+\beta_1 \, cigs_i+\beta_2 \,faminc_i+ \beta_3 \, motheduc_i$$
+<!--ANSWER BELOW HERE-->
+
+```{r}
+lm(bwght~cigs+faminc+motheduc, data = bwght) %>% summary()
+```
+
+Controlling for income and education, each cigarette smoked while pregant will cause the birthweight to decrease by 0.46 ounces.
+It turns out there was no noticeable difference when we included education!
+
+---
+
 ## Question 15
+
 Try out drawing this model using the  `ggdag` package in R. See my DAG in question 3 for an example.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r, echo=T}
+library(ggdag)
+dagify(bwght~cigs+inc+educ,
+       cigs~price+educ+inc,
+       inc~educ+u1,
+       price~u1,
+       exposure = "cigs",
+       outcome = "bwght") %>% 
+  ggdag(stylized = FALSE, seed=1)+theme_dag_blank()+theme(legend.position = "none")
+```
+```

### 2019-11-12 static/slides/18-slides.Rmd
Add r practice 17 answers

@@
 
-- `Female` is our dummy $x$-variable
-
-- Hard to see relationships because of **overplotting**
-
-- Use `geom_jitter()` instead of `geom_point()` to *randomly* nudge points
-  - *Only* for plotting purposes, does not affect actual data, regression, etc.
-
 ]
@@
 
-Code region numerically: $$Region_i=
-\begin{cases}1 & \text{if } i \text{ is in }Northeast \\
+]
+
+Code region numerically: 
+$$Region_i=
+\begin{cases}1 & \text{if } i \text{ is in }Northeast\\
    			2 & \text{if } i \text{ is in } Midwest\\
     		3 & \text{if } i \text{ is in } South \\
-   			4 & \text{if } i \text{ is in } West\\ \end{cases}$$
-]
+   			4 & \text{if } i \text{ is in } West\\ 
+\end{cases}$$
 

### 2019-11-14 content/assignment/04-problem-set.Rmd
Add class 19 content

---
title: "Problem Set 4"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

*Due by Tuesday, November 19, 2019*

- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/04-problem-set-pdf.pdf)
- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/04-problem-set.Rmd)

# Instructions

For this problem set, you may submit handwritten answers on a plain sheet of paper, or download and type/handwrite on the PDF.

Alternatively, you may download the `.Rmd` file, do the homework in markdown, and email to me a single `knit`ted `html` or `pdf` file (and be sure that it shows all of your code).

You may work together (and I highly encourage that) but you must turn in your own answers. I grade homeworks 70% for completion, and for the remaining 30%, pick one question to grade for accuracy - so it is best that you try every problem, even if you are unsure how to complete it accurately.

# Theory and Concepts

## Question 1
In your own words, describe what omitted variable bias means. What are the two conditions for an omitted variable to cause a bias?

## Question 2
In your own words, describe what multicollinearity means. What is the cause, and what are the consequences of multicollinearity? How can we measure multicollinearity and its effects? What happens if multicollinearity is *perfect*?

## Question 3
In your own words, describe what a proxy variable is. When or why would we use a proxy variable, and what effects does it have on our model?

## Question 4
Explain how we use Directed Acyclic Graphs (DAGs) to depict a causal model: what are the two criteria that must hold for identifying a causal effect of $X$ on $Y$? When should we control a variable, and when should we *not* control a variable?

# Theory Problems

For the following questions, please *show all work* and explain answers as necessary. You may lose points if you only write the correct answer. You may use `R` to *verify* your answers, but you are expected to reach the answers in this section "manually."

## Question 5

Data were collected from a random sample of 220 home sales from a community in 2017.

$$\widehat{Price}=119.2+0.485 \, BDR+23.4 \, Bath+0.156 \, Hsize+0.002 \, Lsize+0.090 \, Age$$

- $Price$: selling price (in \$1,000s)
- $BDR$: number of bedrooms
- $Bath$: number of bathrooms
- $Hsize$: size of the house (in $ft^2$)
- $Lsize$: lot size (in $ft^2$)
- $Age$: age of the house (in years)

### Part A
Suppose that a homeowner converts part of an existing living space in her house to a new bathroom. What is the expected increase in the value of the house?

### Part B
Suppose a homeowner adds a new bathroom to her house, which also increases the size of the house by 100 square feet. What is the expected increase in the value of the house?

### Part C
Suppose the $R^2$ of this regression is 0.727. Calculate the adjusted $\bar{R}^2$.

### Part D
Suppose the following auxiliary regression for $BDR$ has an $R^2$ of 0.841.

$$\widehat{BDR}=\delta_0+\delta_1Bath+\delta_2Hsize+\delta_3Lsize+\delta_4Age$$

Calculate the Variance Inflation Factor for $BDR$ and explain what it means.

## Question 6

A researcher wants to investigate the effect of education on average hourly wages. Wage, education, and experience in the dataset have the following correlations:

|    | Wage | Education | Experience |
|----|------|-----------|------------|
| Wage | 1.0000 |    |    |
| Education | 0.4059 |    |    |
| Experience | 0.1129 | $-0.2995$ | 1.0000 |

She runs a simple regression first, and gets the results:

$$\widehat{\text{Wage}} = -0.9049 +  0.5414 \, Education$$

She runs another regression, and gets the results:

$$\widehat{\text{Experience}} = 35.4615 - 1.4681 \, Education$$

### Part A
If the true marginal effect of experience on wages (holding education constant) is 0.0701, calculate the omitted variable bias in the first regression caused by omitting experience. Does the estimate of $\hat{\beta_1}$ in the first regression overstate or understate the effect of education on wages?

### Part B
Knowing this, what would be the *true effect* of education on wages, holding experience constant?

### Part C
The $R^2$ for the second regression is 0.0897. If she were to run a better regression including both education and experience, how much would the variance of the coefficients on education and experience increase? Why?

# Question 7
Consider

# R Questions

Answer the following questions using `R`. When necessary, please write answers in the same document (knitted `Rmd` to `html` or `pdf`, typed `.doc(x)`, or handwritten) as your answers to the above questions. Be sure to include (email or print an `.R` file, or show in your knitted `markdown`) your code and the outputs of your code with the rest of your answers.

## Question 11

- [<i class="fas fa-table"></i> `heightwages.csv`](http://metricsf19.classes.ryansafner.com/data/heightwages.csv)

Download the `heightwages.csv` dataset. This data is a part of a larger dataset from the National Longitudinal Survey of Youth (NLSY) 1979 cohort: a nationally representative sample of 12,686 men and women aged 14-22 years old when they were first surveyed in 1979. They were subsequently interviewed every year through 1994 and then every other year afterwards. There are many included variables, but for now we will just focus on:

- `wage96`: Adult hourly wages (\$/hr) reported in 1996
- `height85`: Adult height (inches) reported in 1985
- `height81`: Adolescent height (inches) reported in 1981

> We want to figure out what is the effect of height on wages (e.g. do taller people earn more on average than shorter people?)

### Part A

Create a quick scatterplot between `height85` (as $X)$ amd `wage96` (as $Y)$.

### Part B

Regress wages on adult height. Write the equation of the estimated OLS regression. Interpret the coefficient on `height85`.

### Part C

How much would someone who is 5'10" be predicted to earn per hour, according to the model?

### Part D

Would adolescent height cause an omitted variable bias if it were left out? Explain using both your intuition, and some statistical evidence with `R`.

### Part E

Now add adolescent height to the regression, and write the new regression equation below, as before. Interpret the coefficient on `height85`.

### Part F

How much would someone who is 5'10" in 1985 and 4'8" in 1981 be predicted to earn, according to the model?

### Part G

What happened to the estimate on `height85` and its standard error?

### Part H

Is there multicollinearity between `height85` and `height81`? Explore with a scatterplot.^[Hint: to avoid overplotting, use `geom_jitter()` instead of `geom_point()` to get a better view of the data.]

### Part I

Quantify how much multicollinearity affects the variance of the OLS estimates on both heights.^[Hint: You'll need the `car` package.]

### Part J

Reach the same number as in part I by running an auxiliary regression.^[Hint: There's some missing `wage96` data that may give you a different answer, so `filter()` your data here by `!is.na(wage96)` before running this regression - this will include only observations for `wage96` that are not `NA`'s.]

### Part K

Make a regression table from part B and D using `huxtable`.

### 2019-11-14 content/class/19-class.Rmd
Add class 19 content

---
title: "3.7: Regression with Interaction Effects - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

*Thursday, November 14, 2019*

# Overview

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/19-slides.html)

# Problem Set 4 Due Thursday Nov 19

[Problm Set 4](/assignment/04-problem-set) (on classes 3.1-3.5) is due by Thursday November 21

### 2019-11-14 content/r/19-r-practice.Rmd
Add class 19 content

---
title: "3.7: Regression with Interaction Effects - R Practice"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

# Set Up
To minimize confusion, I suggest creating a new `R Project` (e.g. `regression_practice`) and storing any data in that folder on your computer.

Alternatively, I have made a project in R Studio Cloud that you can use (and not worry about trading room computer limitations), with the data already inside (you will still need to assign it to an object).

- [<i class="fas fa-cloud"></i> View Project on R Studio Cloud](https://rstudio.cloud/project/716555)
- [<i class="fab fa-markdown"></i> `Download Markdown`](/r/19-r-practice-rmd.Rmd)

## Question 1

We are returning to the speeding tickets data that we began to explore in [R Practice 15 on Multivariate Regression](http://metricsf19.classes.ryansafner.com/practice/15-practice). Download and read in (`read_csv`) the data below.

- [<i class="fas fa-table"></i> `speeding_tickets.csv`](/data/speeding-tickets.csv)

This data again comes from a paper by Makowsky and Strattman (2009) that we will examine later. Even though state law sets a formula for tickets based on how fast a person was driving, police officers in practice often deviate from that formula. This dataset includes information on all traffic stops. An amount for the fine is given only for observations in which the police officer decided to assess a fine. There are a number of variables in this dataset, but the one's we'll look at are:

| Variable | Description |
|----------|-------------|
| `Amount` | Amount of fine (in dollars) assessed for speeding |
| `Age` | Age of speeding driver (in years) |
| `MPHover` | Miles per hour over the speed limit |
| `Black` | Dummy $=1$ if driver was black, $=0$ if not |
| `Hispanic` | Dummy $=1$ if driver was Hispanic, $=0$ if not |
| `Female` | Dummy $=1$ if driver was female, $=0$ if not |
| `OutTown` | Dummy $=1$ if driver was not from local town, $=0$ if not |
| `OutState` | Dummy $=1$ if driver was not from local state, $=0$ if not |
| `StatePol` | Dummy $=1$ if driver was stopped by State Police, $=0$ if stopped by other (local) |

> We again want to explore who gets fines, and how much.

## Question 2

We will have to do a little bit of cleaning to get the data in a more usable form.

### Part A

Inspect the data with `str()` or `head()` or `glimpse()` to see what it looks like.

### Part B

What `class` of variable are `Black`, `Hispanic`, `Female`, `OutTown`, and `OutState`?^[Hint use the `class(df$variable)` command to ask what class something is, where `df` is your dataframe, and `variable` is the name of a variable.]

### Part C

### Part C

Notice that when importing the data from the `.csv` file, `R` interpretted these variables as `integer`, but we want them to be `factor` variables, to ensure `R` recognizes that there are two groups (categories), 0 and 1. Convert each of these variables into factors by reassigning it according to the format:

```{r, echo = T, eval = F}
df<-df %>%
  mutate(my_var=as.factor(my_var),
         my_var2=as.factor(my_var2))
```

where
- `df` is the name of your dataframe
- `my_var` and `my_var2` are example variables^[As a bonus, you can try doing this with just one conditional command: `mutate_if(is.integer, factor)`. See our [Data Wrangling slides](https://metricsf19.classes.ryansafner.com/slides/05-slides#95) for refreshers of all the fancy `mutate()` possibilities!]

### Part D

Confirm they are each now factors by checking their class again.

### Part E

Get a `summary()` of `Amount`. Note that there are a lot of `NA`'s (these are people that were stopped but did not recieve a fine)! `filter()` only those observations for which `Amount` is a positive number, and save this in your dataframe (assign and overwrite it, or make a new dataframe). Then double check that this worked by getting a `summary()` of `Amount` again.

## Question 3

Create a scatterplot between `Amount` (as Y) and Female (as X).^[Hint: Use `geom_jitter()` instead of `geom_point()` to plot the points, and play around with `width` settings inside `geom_jitter()`]

## Question 4

Now let's start looking at the distribution conditionally to find the different group means.

### Part A
Find the mean and standard deviation^[Hint: use the `summarrize()` command, once you have properly `filter()`ed the data.] of `Amount` for *male* drivers and again for *female* drivers.

### Part B

What is the difference between the average Amount for Males and Females?

### Part C

We did not go over how to do this in class, but you can run a **t-test for the difference in group means** to see if the difference is statistically significant. The syntax is similar for a regression:

`t.test(y~d, data=df)`

where: 
  - `y` is the variable we are testing
  - `d` is the dummy variable

Is there a statistically significant difference between Amount for male and female drivers?^[Hint: this is like any hypothesis test. A $t$-value needs to be large enough to be greater than a critical value of $t$. Check the $p$-value and see if it is less than our standard of $\alpha=0.05.$]

## Question 5

Now run the following regression to ensure we get the same result

$$\text{Amount}_i=\hat{\beta_0}+\hat{\beta_1}Female_i$$

### Part A

Write out the estimated regression equation.

### Part B

Use the regression coefficients to find

- (i) the average `Amount` for men
- (ii) the average `Amount` for women
- (iii) the difference in average `Amount` between men and women

## Question 6

Let's recode the sex variable.

### Part A

Make a new variable called `Male` and save it in your dataframe.^[Hint: `mutate()` and define `Male` equal to the `ifelse()` function, which has three arguments: 1. the condition to test, 2. what to define "Male" as when the condition is `TRUE`, and 3. what to define "Male" as when the condition is `FALSE`.]

### Part B

Run the same regression as in question 5, but use `Male` instead of `Female`.

### Part C

Write out the estimated regression equation.

### Part D

Use the regression coefficients to find

- (i) the average `Amount` for men
- (ii) the average `Amount` for women
- (iii) the difference in average `Amount` between men and women

## Question 7

Run a regression of `Amount` on `Male` and `Female`. What happens, and why?

## Question 8

`Age` probably has a lot to do with differences in fines, perhaps also age affects fines differences between males and females.

### Part A

Run a regression of `Amount` on `Age` and `Female.` How does the coefficient on `Female` change?

### Part B

Now let's see if the difference in fine between men and women are different depending on the driver's age. Run the regression again, but add an **interaction term** between `Female` and `Age` interaction term.

### Part C

Write out your estimated regression equation.

### Part D

Interpret the interaction effect. Is it statistically significant?

### Part E

Plugging in 0 or 1 as necessary, rewrite (on your paper) this regression as *two separate* equations, one for Males and one for Females.

### Part F

Let's try to visualize this. Make a scatterplot of `Age` (X) and `Amount` (Y) and include a regression line.

### Part G

Try adding to your base layer `aes()`, set `color=Female`. This will produce two lines and color the points by `Female`.^[Sometimes we may also need to remind `R` that `Female` is a factor with `as.factor(Female)`. We don't need to in *this* case because we already reset `Female` as a faction in question 1.]

### Part H

Add a facet layer to make two different scatterplots with an additional layer `+facet_wrap(~Female)`

## Question 9

Now let's look at the possible interaction between Sex (`Male` or `Female`) and whether a driver is from In-State or Out-of-State (`OutState`).

### Part A

Use `R` to examine the data and find the mean for (i) Males In-State, (ii) Males Out-of-State, (iii) Females In-State, and (iv) Females Out-of-State.

### Part B

Now run a regression of the following model:

$$\text{Amount}_i=\hat{\beta_0}+\hat{\beta_1}Female_i+\hat{\beta_2}OutState_i+\hat{\beta_3}Female_i*OutState_i$$

### Part C

Write out the estimated regression equation.

### Part D

What does each coefficient mean?

### Part E

Using the regression equation, what are the means for

- (i) Males In-State
- (ii) Males Out-of-State
- (iii) Females In-State
- (iv) Females Out-of-State?

Compare to your answers in part A.

## Question 10

Collect your regressions from questions 5, 6b, 8a, 8b, and 9b and output them in a regression table with `huxtable()`.

### 2019-11-14 static/r/19-r-practice-rmd.Rmd
Add class 19 content

--- 2019-11-14 content/r/19-r-practice.Rmd
@@
 title: "3.7: Regression with Interaction Effects - R Practice"
-editor_options: 
-  chunk_output_type: console
+author: "YOUR NAME"
+date: "`r Sys.Date()`"
+output:
+  html_document:
+    df_print: paged
+    #theme: 
+    toc: true 
+    toc_depth: 3
+    toc_float: true
+    code_folding: show
+    highlight: tango
 ---
 
-```{r message=FALSE, warning=FALSE, include=FALSE}
-library(tidyverse)
-library(pander)
-source(file.path(here::here(), "lib", "graphics.R"))
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
 ```
@@
 
-### Part C
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 Confirm they are each now factors by checking their class again.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 3
@@
 Create a scatterplot between `Amount` (as Y) and Female (as X).^[Hint: Use `geom_jitter()` instead of `geom_point()` to plot the points, and play around with `width` settings inside `geom_jitter()`]
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 What is the difference between the average Amount for Males and Females?
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 5
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part A
-
 Write out the estimated regression equation.
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part B
-
 Use the regression coefficients to find
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 6
-
 Let's recode the sex variable.
@@
 ### Part A
-
 Make a new variable called `Male` and save it in your dataframe.^[Hint: `mutate()` and define `Male` equal to the `ifelse()` function, which has three arguments: 1.
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part B
-
 Run the same regression as in question 5, but use `Male` instead of `Female`.
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part C
+Write out the estimated regression equation.
 
-Write out the estimated regression equation.
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 Run a regression of `Amount` on `Male` and `Female`. What happens, and why?
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part B
-
 Now let's see if the difference in fine between men and women are different depending on the driver's age.
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part C
-
 Write out your estimated regression equation.
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part D
-
 Interpret the interaction effect. Is it statistically significant?
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part E
+Plugging in 0 or 1 as necessary, rewrite (on your paper) this regression as *two separate* equations, one for Males and one for Females.
 
-Plugging in 0 or 1 as necessary, rewrite (on your paper) this regression as *two separate* equations, one for Males and one for Females.
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 Let's try to visualize this. Make a scatterplot of `Age` (X) and `Amount` (Y) and include a regression line.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part H
@@
 Add a facet layer to make two different scatterplots with an additional layer `+facet_wrap(~Female)`
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+----
+
 ### Part C
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part D
-
 What does each coefficient mean?
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part E
-
 Using the regression equation, what are the means for
@@
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ## Question 10
@@
 Collect your regressions from questions 5, 6b, 8a, 8b, and 9b and output them in a regression table with `huxtable()`.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```

### 2019-11-14 static/slides/19-slides.Rmd
Add class 19 content

---
title: "3.7: Regression with Interaction Effects"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))

library(wooldridge)
wages<-wooldridge::wage1
wages<-wages %>%
  mutate(Gender = factor(ifelse(female==0,
                         "Male",
                         "Female")))
```

# Interaction Effects

- Sometimes one $X$ variable might *interact* with another in determining $Y$

.content-box-green[
.green[**Example**]: Consider the gender pay gap again.

- *Experience* certainly affects wages

- Does experience affect *men*'s wages *differently* than it affects women's wages?
  - i.e. is there an .shout[interaction effect] between gender and experience?

- Note this is *NOT the same* as asking "do men earn more than women *with the same amount of experience*? (i.e. controlling for experience)
]

---

# Three Types of Interactions

- Depending on the types of variables paired, there are three possible types of interaction effects

- We will look at each in turn

--

1. Interaction between a **dummy** and a **continuous** variable:
$$Y_i=\beta_0+\beta_1X_i+\beta_2 D_i+\beta_3 \mathbf{(X_i \times D_i)}$$
--

2. Interaction between a **two dummy** variables:
$$Y_i=\beta_0+\beta_1D_{1i}+\beta_2 D_{2i}+\beta_3 \mathbf{(D_{1i} \times D_{2i})}$$
--

3. Interaction between a **two continuous** variables:
$$Y_i=\beta_0+\beta_1X_{1i}+\beta_2 X_{2i}+\beta_3 \mathbf{(X_{1i} \times X_{2i})}$$

---

class: inverse, center, middle

# Interactions Between a Dummy and Continuous Variable

---

# Interactions Between a Dummy and Continuous Variable I

- We can model an interaction by introducing a variable that is an .shout[interaction term] capturing the interaction between two variables:

$$Y_i=\beta_0+\beta_1X_i+\beta_2 D_i+\beta_3(X_i \times D_i) \quad \text{ where } D_i=\{0,1\}$$

--

- $\beta_3$ estimates the .shout[interaction term] (in this case between a dummy variable and a continuous variable)

--

- What do the different coefficients ($\beta$'s) tell us? 
  - Again, think logically by examining each group $(D_i=0$ or $D_i=1)$

---

# Interaction Effects as Two Regressions I

$$Y_i=\beta_0+\beta_1X_i+\beta_2 D_i+\beta_3 X_i * D_i$$

- .red[When `\\(D_i=0\\)` (Control group):]

$$\begin{align*}
\hat{Y_i}&=\hat{\beta_0}+\hat{\beta_1}X_i+\hat{\beta_2}(\mathbf{0})+\hat{\beta_3}X_i \times (\mathbf{0})\\
\hat{Y_i}& =\hat{\beta_0}+\hat{\beta_1}X_i\\
\end{align*}$$

--

- .blue[When `\\(D_i=1\\)` (Treatment group):]

$$\begin{align*}
\hat{Y_i}&=\hat{\beta_0}+\hat{\beta_1}X_i+\hat{\beta_2}(\mathbf{1})+\hat{\beta_3}X_i \times (\mathbf{1})\\
\hat{Y_i}&= (\hat{\beta_0}+\hat{\beta_2})+(\hat{\beta_1}+\hat{\beta_3})X_i\\
\end{align*}$$

--

- So what we really have is *two* regression lines!

---

# Interaction Effects as Two Regressions II

.pull-left[

```{r, fig.retina=3}
library(mosaic)

control=function(x){2+0.5*x}
treat=function(x){4+x}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=control, geom="line", size=2, color = "red")+
    geom_label(aes(x=5,y=control(5)), color = "red", label=expression(paste("slope=",hat(beta[1]))), size = 5)+
  stat_function(fun=treat, geom="line", size=2, color = "blue")+
    geom_label(aes(x=5,y=treat(5)), color = "blue", label=expression(paste("slope=",hat(beta[1])+hat(beta[3]))), size = 5)+
    scale_x_continuous(breaks=NULL,
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(2,4),
                     labels=c(expression(hat(beta[0])),expression(hat(beta[0])+hat(beta[2]))),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  labs(x = "",
       y = "")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)

```
]

.pull-right[

- .red[`\\(D_i=0\\)` group: `\\(Y_i=\hat{\beta_0}+\hat{\beta_1}X_i\\)`]

- .blue[`\\(D_i=1\\)` group: `\\(Y_i=(\hat{\beta_0}+\hat{\beta_2})+(\hat{\beta_1}+\hat{\beta_3})X_i\\)`]

]

---

# Interpretting Coefficients I

$$Y_i=\beta_0+\beta_1X_i+\beta_2 D_i+\beta_3 \mathbf{(X_i \times D_i)}$$

- To interpret the coefficients, compare cases after changing $X$ by $\mathbf{\Delta X}$:

--

$$Y_i+\mathbf{\Delta Y_i}=\beta_0+\beta_1(X_i\mathbf{+\Delta X_i})\beta_2D_i+\beta_3\big((X_i\mathbf{+\Delta X_i})D_i\big)$$

--

- Subtracting these two equations, the difference is:

--

$$\begin{align*}
	\Delta Y_i &= \beta_1 \Delta X_i + \beta_3 D_i \Delta X_i\\
	\mathbf{\frac{\Delta Y_i}{\Delta X_i}} &\mathbf{= \beta_1+\beta_3 D_i}\\
\end{align*}$$

--

- .onfire[The effect of `\\(X \rightarrow Y\\)` depends on the value of `\\(D_i\\)`!]

- .shout[`\\(\beta_3\\)`: *increment* to the effect of `\\(X \rightarrow Y\\)` when `\\(D_i=1\\)` (vs. `\\(D_i=0\\)`]

---

# Interpretting Coefficients II

$$Y_i=\beta_0+\beta_1X_i+\beta_2 D_i+\beta_3 \mathbf{(X_i \times D_i)}$$

- $\hat{\beta_0}$: $Y_i$ for $X_i=0$ and $D_i=0$

--

- $\beta_1$: Marginal effect of $X_i \rightarrow Y_i$ for $D_i=0$

--

- $\beta_2$: Marginal effect on $Y_i$ of difference between $D_i=0$ and $D_i=1$

--

- $\beta_3$: The **difference** of the marginal effect of $X_i \rightarrow Y_i$ between $D_i=0$ and $D_i=1$

--

- This is a bit awkward, easier to think about the two regression lines:

---

# Interpretting Coefficients III

$$Y_i=\beta_0+\beta_1X_i+\beta_2 D_i+\beta_3 \mathbf{(X_i \times D_i)}$$

- .red[For `\\(D_i=0\\)` Group:  `\\(\hat{Y_i}=\hat{\beta_0}+\hat{\beta_1}X_i\\)`]
  - Intercept $(Y_i$ for $X_i=0$): $\hat{\beta_0}$
  - Slope (Marginal effect of $X_i \rightarrow Y_i$ for $D_i=0$ group): $\hat{\beta_1}$

--

- .blue[For `\\(D_i=1\\)` Group:  `\\(\hat{Y_i}=(\hat{\beta_0}+\hat{\beta_2})+(\hat{\beta_1}+\hat{\beta_3})X_i\\)`]
  - Intercept $(Y_i$ for $X_i=0$): $\hat{\beta_0}+\hat{\beta_2}$
  - Slope (Marginal effect of $X_i \rightarrow Y_i$ for $D_i=1$ group): $\hat{\beta_1}+\hat{\beta_3}$

--

- How can we determine if the two lines have the same slope and/or intercept?
  - Same intercept? $t$-test $H_0$: $\beta_2=0$
  - Same slope? $t$-test $H_0$: $\beta_3=0$

---

# Example I

.content-box-green[
.green[**Example**]: $$\widehat{wage_i}=\hat{\beta_0}+\hat{\beta_1}exper_i+\hat{\beta_2}female_i+\hat{\beta_3}(exper_i \times female_i)$$

]

--

- .red[For males `\\((female=0)\\)`]:
$$\widehat{wage_i}=\hat{\beta_0}+\hat{\beta_1}exper$$

--

- .blue[For females `\\((female=1)\\)`]:
$$\widehat{wage_i}=\underbrace{(\hat{\beta_0}+\hat{\beta_2})}_{\text{intercept}}+\underbrace{(\hat{\beta_1}+\hat{\beta_3})}_{\text{slope}}exper$$

---

# Example II

.pull-left[

```{r interaction-plot-1, echo=T, eval=F}
interaction_plot<-ggplot(data = wages)+
  aes(x = exper,
      y = wage,
      color = as.factor(female))+ # make female factor
  geom_point()+
  #geom_smooth(method = "lm")+
  scale_y_continuous(labels=scales::dollar)+
  labs(x = "Experience (Years)",
       y = "Wage")+
  guides(color=F)+ # hide legend
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
interaction_plot
```

- Need to make sure `female` is a `factor` variable
  - Use `as.factor()` in plot

]

.pull-right[

```{r, ref.label="interaction-plot-1"}
```
]

---

# Example II

.pull-left[

```{r interaction-plot-2, echo=T, eval=F}
interaction_plot+
  geom_smooth(method="lm")
```
]

.pull-right[

```{r, ref.label="interaction-plot-2"}
```
]

---

# Example II

.pull-left[

```{r interaction-plot-3, echo=T, eval=F}
interaction_plot+
  geom_smooth(method="lm", color = "black")+
  facet_wrap(~female)
```
]

.pull-right[

```{r, ref.label="interaction-plot-3"}
```
]

---

# Example Regression in R I

- Syntax for an interaction term is easy in `R`: `var1:var2`
  - Or could just do `var1*var2` (multiply)

```{r}
interaction_reg<-lm(wage~female+exper+female:exper, data = wages)
```

---

# Example Regression in R II

.font80[
```{r, echo = T}
summary(interaction_reg)
```
]

---

# Example Regression in R III

.pull-left[

```{r huxreg1, echo=T, eval=F}
library(huxtable)
huxreg(interaction_reg,
       coefs = c("Constant" = "(Intercept)",
                 "Experience" = "exper",
                 "Female" = "female",
                 "Female * Exper" = "female:exper"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```

]

.pull-right[

```{r, ref.label="huxreg1"}
```
]

---

# Example Regression in R: Interpretting Coefficients

$$\widehat{wage_i}=6.16+0.05 \, Experience_i - 1.55 \, Female_i - 0.06 \, (Experience_i \times Female_i)$$

--

- $\hat{\beta_0}$:

---

# Example Regression in R: Interpretting Coefficients

$$\widehat{wage_i}=6.16+0.05 \, Experience_i - 1.55 \, Female_i - 0.06 \, (Experience_i \times Female_i)$$

- $\hat{\beta_0}$: Men with experience of 0 years earn $6.16

--

- $\hat{\beta_1}$:

---

# Example Regression in R: Interpretting Coefficients

$$\widehat{wage_i}=6.16+0.05 \, Experience_i - 1.55 \, Female_i - 0.06 \, (Experience_i \times Female_i)$$

- $\hat{\beta_0}$: Men with experience of 0 years earn $6.16

- $\hat{\beta_1}$: For every additional year of experience, *men* earn $0.05

--

- $\hat{\beta_2}$:

---

# Example Regression in R: Interpretting Coefficients

$$\widehat{wage_i}=6.16+0.05 \, Experience_i - 1.55 \, Female_i - 0.06 \, (Experience_i \times Female_i)$$

- $\hat{\beta_0}$: Men with experience of 0 years earn $6.16

- $\hat{\beta_1}$: For every additional year of experience, *men* earn $0.05

- $\hat{\beta_2}$: *Women* on average earn $1.55 less than men, *holding experience constant*

--

- $\hat{\beta_3}$:

---

# Example Regression in R: Interpretting Coefficients

$$\widehat{wage_i}=6.16+0.05 \, Experience_i - 1.55 \, Female_i - 0.06 \, (Experience_i \times Female_i)$$

- $\hat{\beta_0}$: Men with experience of 0 years earn $6.16

- $\hat{\beta_1}$: For every additional year of experience, *men* earn $0.05

- $\hat{\beta_2}$: *Women* on average earn $1.55 less than men, *holding experience constant*

- $\hat{\beta_3}$: *Women* earn $0.06 less than men for every additional year of experience

---

# Example Regression in R: Interpretting Coefficients as Two Regressions I

$$\widehat{wage_i}=6.16+0.05 \, Experience_i - 1.55 \, Female_i - 0.06 \, (Experience_i \times Female_i)$$

--

.red[Regression for men `\\(female=0)\\)`]
$$\widehat{wage_i}=6.16+0.05 \, Experience$$

--

- Men with no experience earn $6.16

--

- For every additional year of experience, men earn $0.05 more

---

# Example Regression in R: Interpretting Coefficients as Two Regressions II

$$\widehat{wage_i}=6.16+0.05 \, Experience_i - 1.55 \, Female_i - 0.06 \, (Experience_i \times Female_i)$$

.blue[Regression for women `\\(female=1)\\)`]
$$\begin{align*}
\widehat{wage_i}&=6.16+0.05 \, Experience - 1.55\mathbf{(1)}-0.06 \, Experience \times \mathbf{(1)}\\
&= (6.16-1.55)+(0.05-0.06) \, Experience\\
&= 4.61-0.01 \, Experience \\
\end{align*}$$

--

- Women with no experience earn $4.61

--

- For every additional year of experience, women earn $0.01 *less*

---

# Example Regression in R: Hypothesis Testing

.pull-left[
- Test the significance of interaction effects

- Think: are slopes and intercepts of the two regressions statistically significantly different?

- Are intercepts different?
  - Difference between men vs. women for no experience?
  - Is $\hat{\beta_2}$ significant?
  - Yes: $t = -3.210$, $p$-value = 0.00

- Are slopes different?
  - Difference between men vs. women for marginal effect of experience?
  - Is $\hat{\beta_3}$ significant?
  - Yes: $t = -2.48$, $p$-value = 0.01

]

.pull-right[

```{r, echo=T}
library(broom)
tidy(interaction_reg)
```
]

---

class: inverse, center, middle

# Interactions Between Two Dummy Variables

---

# Interactions Between Two Dummy Variables

$$Y_i=\beta_0+\beta_1D_{1i}+\beta_2 D_{2i}+\beta_3 \mathbf{(D_{1i} \times D_{2i})}$$

- $D_{1i}$ and $D_{2i}$ are dummy variables

--

- $\hat{\beta_1}$: effect on $Y$ of going from $D_{1i}=0$ to $D_{1i}=1$ for $D_{2i}=0$

--

- $\hat{\beta_2}$: effect on $Y$ of going from $D_{2i}=0$ to $D_{2i}=1$ for $D_{1i}=0$

--

- $\hat{\beta_3}$: effect on $Y$ of going from $D_{1i}=0$ and $D_{2i}=1$ to $D_{1i}=1$ *and* $D_{2i}=1$
  - *increment* to the effect of $D_{1i}$ going from 0 to 1 when $D_{2i}=1$ (vs. 0)
  - *increment* to the effect of $D_{2i}$ going from 0 to 1 when $D_{1i}=1$ (vs. 0)

---

- As always, best to think logically about possibilities (when each dummy $=0$ or $=1)$

---

# Interactions Between Two Dummy Variables: Interpretting Coefficients

$$Y_i=\beta_0+\beta_1D_{1i}+\beta_2 D_{2i}+\beta_3 \mathbf{(D_{1i} \times D_{2i})}$$

--

- To interpret coefficients, compare cases (for $D_{i1}$ as an example):

$$\begin{align*}
E(Y_i|D_{1i}=0, D_{2i}=d_2) = \beta_0+\beta_2 d_2\\
E(Y_i|D_{1i}=1, D_{2i}=d_2) = \beta_0+\beta_1(1)+\beta_2 d_2+\beta_3(1)d_2\\
\end{align*}$$

--

- Subtracting the two, the difference is:

$$\beta_1+\beta_3 d_2$$

- .onfire[The effect of `\\(D_{1i} \rightarrow Y_i\\)` depends on the value of `\\(D_{2i}\\)`]
  - $\hat{\beta_3}$ is the *increment* to the effect of $D_1$ on $Y$ when $D_2=1$

---

# Interactions Between Two Dummy Variables: Example

.content-box-green[
.green[**Example**]: Does the gender pay gap change if a person is married vs. single?

$$\widehat{wage_i}=\hat{\beta_0}+\hat{\beta_1}female_i+\hat{\beta_2}married_i+\hat{\beta_3}(female_i \times married_i)$$

]

--

- Logically, there are 4 possible combinations of $female_i = \{0,1\}$ and $married_i = \{0,1\}$

--

1. **Unmarried men** $(female_i=0, \, married_i=0)$
$$\widehat{wage_i}=\hat{\beta_0}$$

--

2. **Married men** $(female_i=0, \, married_i=1)$
$$\widehat{wage_i}=\hat{\beta_0}+\hat{\beta_2}$$

---

# Interactions Between Two Dummy Variables: Example

.content-box-green[
.green[**Example**]: Does the gender pay gap change if a person is married vs. single?

$$\widehat{wage_i}=\hat{\beta_0}+\hat{\beta_1}female_i+\hat{\beta_2}married_i+\hat{\beta_3}(female_i \times married_i)$$

]

- Logically, there are 4 possible combinations of $female_i = \{0,1\}$ and $married_i = \{0,1\}$

3. **Unmarried women** $(female_i=1, \, married_i=0)$
$$\widehat{wage_i}=\hat{\beta_0}+\hat{\beta_1}$$

--

4. **Married women** $(female_i=1, \, married_i=1)$
$$\widehat{wage_i}=\hat{\beta_0}+\hat{\beta_1}+\hat{\beta_2}+\hat{\beta_3}$$

---

# Looking at the Data

.font70[
.pull-left[

```{r, echo = T}
# get average wage for unmarried men
wages %>%
  filter(female == 0,
         married == 0) %>%
  summarize(mean = mean(wage))

# get average wage for married men
wages %>%
  filter(female == 0,
         married == 1) %>%
  summarize(mean = mean(wage))

```
]

.pull-right[

```{r, echo = T}
# get average wage for unmarried women
wages %>%
  filter(female == 1,
         married == 0) %>%
  summarize(mean = mean(wage))

# get average wage for married women
wages %>%
  filter(female == 1,
         married == 1) %>%
  summarize(mean = mean(wage))

```
]
]

---

# Interactions Between Two Dummy Variables: Group Means

$$\widehat{wage_i}=\hat{\beta_0}+\hat{\beta_1}female_i+\hat{\beta_2}married_i+\hat{\beta_3}(female_i \times married_i)$$

| Gender | Unmarried | Married |
|--------|-----------|---------|
| Male   | $5.17     | $7.98   |
| Female | $4.61     | $4.57   |

---

# Interactions Between Two Dummy Variables: In R I

.font70[
```{r, echo=T}
reg_dummies<-lm(wage~female+married+female:married, data = wages)
summary(reg_dummies)
```
]

---

# Interactions Between Two Dummy Variables: In R II

.pull-left[

```{r huxreg2, echo=T, eval=F}
library(huxtable)
huxreg(reg_dummies,
       coefs = c("Constant" = "(Intercept)",
                 "Female" = "female",
                 "Married" = "married",
                 "Female * Married" = "female:married"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```

]

.pull-right[

```{r, ref.label="huxreg2"}
```
]

---

# Interactions Between Two Dummy Variables: Interpretting Coefficients

$$\widehat{wage_i}=5.17-0.56 \, female_i + 2.82 \, married_i - 2.86 \, (female_i \times married_i)$$

| Gender | Unmarried | Married |
|--------|-----------|---------|
| Male   | $5.17     | $7.98   |
| Female | $4.61     | $4.57   |

--

- Wage for **unmarried men**: $\hat{\beta_0}=5.17$

--

- Wage for **married men**: $\hat{\beta_0}+\hat{\beta_2}=5.17+2.82=7.98$

--

--

- Wage for **unmarried women**: $\hat{\beta_0}+\hat{\beta_1}=5.17-0.56=4.61$

--

- Wage for **married women**: $\hat{\beta_0}+\hat{\beta_1}+\hat{\beta_2}+\hat{\beta_3}=5.17-0.56+2.82-2.86=4.57$

---

class: inverse, center, middle

# Interactions Between Two Continuous Variables

---

# Interactions Between Two Continuous Variables

$$Y_i=\beta_0+\beta_1X_{1i}+\beta_2 X_{2i}+\beta_3 \mathbf{(X_{1i} \times X_{2i})}$$

--

- To interpret coefficients, compare changes after changing $\mathbf{\Delta X_{1i}}$:

$$Y+\mathbf{\Delta Y}=\beta_0+\beta_1(X_1+\mathbf{\Delta X_1})\beta_2X_2+\beta_3((X_1+\mathbf{\Delta X_1}) \times X_2)$$

--

- Take the difference to get:

--

$$\begin{align*}
\Delta Y &= \beta_1 \Delta X_1 + \beta_3 X_2 \Delta X_1\\
\frac{\Delta Y}{\Delta X_1} &= \beta_1+\beta_3 X_2\\ 	
\end{align*}$$

--

- .onfire[The effect of `\\(X_1 \rightarrow Y_i\\)` depends on `\\(X_2\\)`]

- $\beta_3$: increment to the effect of $X_1 \rightarrow Y_i$ from a 1 unit change in $X_2$`

---

# Interactions Between Two Continuous Variables: Example

.content-box-green[
.green[**Example**]: Do education and experience interact in their determination of wages?

$$	\widehat{wage_i}=\hat{\beta_0}+\hat{\beta_1} \, educ_i+\hat{\beta_2} \, exper_i+\hat{\beta_3}(educ_i \times exper_i)$$

]

- Estimated effect of education on wages depends on the amount of experience (and vice versa)!

$$\frac{\Delta wage}{\Delta educ}=\hat{\beta_1}+\beta_3 \, exper$$

- This is a type of nonlinearity (we will examine nonlinearities next lesson)

---

# Interactions Between Two Continuous Variables: In R I

.font70[
```{r, echo=T}
reg_cont<-lm(wage~educ+exper+educ:exper, data = wages)
summary(reg_cont)
```
]

---

# Interactions Between Two Continuous Variables: In R II

.pull-left[

```{r huxreg3, echo=T, eval=F}
library(huxtable)
huxreg(reg_cont,
       coefs = c("Constant" = "(Intercept)",
                 "Education" = "educ",
                 "Experience" = "exper",
                 "Education * Experience" = "educ:exper"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 3)
```

]

.pull-right[

```{r, ref.label="huxreg3"}
```
]

---

# Interactions Between Two Continuous Variables: Interpretting Coefficients

$$\widehat{wages_i}=-2.860+0.602 \, educ_i + 0.047 \, exper_i + 0.002 \, (educ_i \times exper_i)$$

--

Changes in Education:

| Experience | $\displaystyle\frac{\Delta wage}{\Delta educ}$ |
|------------|------------------------------------------------|
| 5 years | $0.602+0.002(5)=0.612$ |
| 10 years | $0.602+0.002(10)=0.622$ |
| 15 years | $0.602+0.002(15)=0.632$ |

--

- Marginal effect of education $\rightarrow$ wages **increases** with more experience (but very insignificantly)

### 2019-11-19 content/assignment/paper-project.Rmd
Update paper deadlines and data sources

@@
 
+**Good R Packages for Getting Data in R Format**^[Some of these come from [Nick Huntington-Klein's excellent list](http://www.nickchk.com/econometrics.html).]
+
+Below are packages written by and for R users that link up with the API of key data sets for easy use in R. Each link goes to the documentation and description of each package.
+
+Don't forget to install^[`install.packages("name_of_package")`] first and then load it with `library()`.
+
+- [`wbstats`](https://cran.r-project.org/web/packages/wbstats/wbstats.pdf) provides access to all the data available on the **World Bank** API, which is basically everything on their website.
+The World Bank keeps track of *many* country-level indicators over time.
+- [`tidycensus`](https://cran.r-project.org/web/packages/tidycensus/tidycensus.pdf) gives you access to data from the **US Census** and the **American Community Survey.** These are the largest high-quality data sets you'll find of cross-sectional data on individual people in the US.
+You'll need to get a (free) API key from the website (or ask me for mine).
+- [`fredr`](https://cran.r-project.org/web/packages/fredr/vignettes/fredr.html) gets data from the **Federal Reserve's Economic Database (FRED)**.
+You'll need to get a (free) API key from the website (or ask me for mine).
+- [`tidyquant`](https://cran.r-project.org/web/packages/tidyquant/index.html) gets data from a number of financial sources (including `fredr).
+- [`icpsrdata`](https://cran.r-project.org/web/packages/icpsrdata/icpsrdata.pdf) downloads data from the **Inter-university Consortium for Political and Social Research** (you'll need an account and a keycode).
+ICPSR is a database of datasets from published social science papers for the purposes of reproducibility.
+- [`NHANES`](https://cran.r-project.org/web/packages/NHANES/NHANES.pdf) uses data from the **US National Health and Nutrition Examination Survey.**
+- [`ipumsr`](https://cran.r-project.org/web/packages/ipumsr/vignettes/ipums.html) has **census data** from *all around the world*, in addition to the US census, **American Community Survey**, and **Current Population Survey**.
+If you're doing international micro work, look at IPUMS.
+It's also the easiest way to get the **Current Population Survey (CPS)**, which is very popular for labor economics.
+Unfortunately ipumsr won't get the data from within R; you'll have to make your own data extract on the IPUMS website and download it.
+But ipumsr will read that file into R and preserve things like names and labels.
+- [`education-data-package-r`](https://github.com/UrbanInstitute/education-data-package-r)^[Note you will need to install `devtools` package first, and then install the package directly from Github with the command `devtools::install_github('UrbanInstitute/education-data-package-r')`] is the Urban Institute's data data on educational institutions in the US, including colleges (in IPEDS) and K-12 schools (in CCD).
+This package also has data on county-level poverty rates from SAIPE.
+- [`psidR`](https://cran.r-project.org/web/packages/psidR/psidR.pdf) is the **Panel Study of Income Dynamics**.
+This study doesn't just follow people over their lifetimes, it follows their children too, generationally!
+A great source for studying how things follow families through generations.
+- [`atus`](https://cran.r-project.org/web/packages/atus/atus.pdf) is th e **American Time Use Survey**, which is a large cross-sectional data set with information on how people spend their time.
+- [`Rilostat`](https://ilostat.github.io/Rilostat/index.html) uses data from the **International Labor Organization.** This contains lots of different statistics on labor, like employment, wage gaps, etc., generally aggregated to the national level and changing over time.
+- [`democracyData`](https://github.com/xmarquez/democracyData)^[Note you will need to install `devtools` package first, and then install the package directly from Github with the command `devtools::install_github('xmarquez/democracyData')`] is a great "package for accessing and manipulating existing measures of democracy."
+- [`politicaldata`](https://cran.rstudio.com/web/packages/politicaldata/index.html) provides useful functions for obtaining commonly-used data in political analysis and political science, including from sources such as the Comparative Agendas Project (which provides data on politics and policy from 20+ countries), the MIT Election and Data Science Lab, and FiveThirtyEight.
+
 Here is a list of good data sources depending on the types of topics you might be interested in writing on:
 
+- [Quality of Government Data](https://qog.pol.gu.se/data/dataextras/datasources) has an extremely wide range of data sources pertaining to measures of institutions.
+The data itself can be found [here](https://qog.pol.gu.se/data/datadownloads/qogstandarddata).
 - National and State Accounts Data: [Bureau of Economic Analysis](http://www.bea.doc.gov/)
@@
 | Abstract | 5 | October 29 | Short summary of your ideas |
-| Data Description | 10 | November 14 | Description of data sources, and some summary statistics |
-| Literature Review | 10 | December 5 | 1-3 paragraphs on 2-3 scholarly sources |
-| Presentation | 5 | December 10/12 | Short presentation of your project so far |
+| Data Description | 10 | November 24 | Description of data sources, and some summary statistics |
+| Literature Review | 10 | December 6 | 1-3 paragraphs on 2-3 scholarly sources |
+| Presentation | 5 | December 5/10 | Short presentation of your project so far |
 | Final Paper Due | 70 | December 17 | Email to me paper, data, code |

### 2019-11-19 content/class/18-class.Rmd
Update paper deadlines and data sources

@@
 
-[Problm Set 4](/assignment/04-problem-set) (on classes 3.1-3.5) is due by Tuesday November 19.
+[Problm Set 4](/assignment/04-problem-set) (on classes 3.1-3.5) is due by Thursday November 21.
+
+# Appendix: T-Test for Difference in Group Means
+
+Often we want to compare the means between two groups, and see if the difference is statistically significant.
+As an example, **is there a statistically significant difference in average hourly earnings between men and women**?
+Let:
+
+- $\mu_W$: mean hourly earnings for female college graduates
+- $\mu_M$: mean hourly earnings for male college graduates
+
+We want to run a hypothesis test for the difference $(d)$ in these two population means: 
+$$\mu_M-\mu_W=d_0$$
+
+Our null hypothesis is that there is *no* statistically significant difference. Let's also have a two-sided alternative hypothesis, simply that there *is* a difference (positive or negative).
+
+- $H_0: d=0$
+- $H_1: d \neq 0$
+
+Note a logical one-sided alternative would be $H_2: d > 0$, i.e. men earn more than women
+
+### The Sampling Distribution of $d$
+
+The *true* population means $\mu_M, \mu_W$ are unknown, we must estimate them from *samples* of men and women. Let:
+- $\bar{Y}_M$ the average earnings of a sample of $n_M$ men  
+- $\bar{Y}_W$ the average earnings of a sample of $n_W$ women
+
+We then estimate $(\mu_M-\mu_W)$ with the sample $(\bar{Y}_M-\bar{Y}_W)$.
+
+We would then run a **t-test** and calculate the **test-statistic** for the difference in means. The formula for the test statistic is:
+
+$$t = \frac{(\bar{Y_M}-\bar{Y_W})-d_0}{\sqrt{\frac{s_M^2}{n_M}+\frac{s_W^2}{n_W}}}$$
+
+We then compare $t$ against the critical value $t^*$, or calculate the $p$-value $P(T>t)$ as usual to determine if we have sufficient evidence to reject $H_0$
+
+```{r}
+library(tidyverse)
+library(wooldridge)
+# Our data comes from wage1 in the wooldridge package
+
+wages<-wooldridge::wage1
+
+# look at average wage for men
+
+wages %>%
+  filter(female==0) %>%
+  summarize(average = mean(wage),
+            sd = sd(wage))
+
+# look at average wage for women
+
+wages %>%
+  filter(female==1) %>%
+  summarize(average = mean(wage),
+            sd = sd(wage))
+
+```
+
+So our data is telling us that male and female average hourly earnings are distributed as such:
+
+$$\begin{align*}
+\bar{Y}_M &\sim N(7.10,4.16)\\
+\bar{Y}_W &\sim N(4.59,2.53)\\
+\end{align*}$$
+
+We can plot this to see visually.
+There is a lot of overlap in the two distributions, but the male average is higher than the female average, and there is also a lot more variation in males than females, noticeably the male distribution skews further to the right.
+
+```{r}
+wages$female<-as.factor(wages$female)
+
+library("ggplot2")
+ggplot(data=wages,aes(x=wage,fill=female))+
+  geom_density(alpha=0.5)+
+  scale_x_continuous(seq(0,25,5),name="Wage",labels=scales::dollar)+
+  theme_light()
+```
+
+Knowing the distributions of male and female average hourly earnings, we can estimate the **sampling distribution of the difference in group eans** between men and women as:
+
+The mean: 
+$$\begin{align*}
+\bar{d}&=\bar{Y}_M-\bar{Y}_W\\
+\bar{d}&=7.10-4.59\\
+\bar{d}&=2.51\\
+\end{align*}$$
+
+The standard error of the mean:
+$$\begin{align*}
+SE(\bar{d})&=\sqrt{\frac{s_M^2}{n_M}+\frac{s_W^2}{n_W}}\\
+&=\sqrt{\frac{4.16^2}{274}+\frac{2.33^2}{252}}\\
+& \approx 0.29\\
+\end{align*}$$
+
+So the sampling distribution of the difference in group means is distributed:
+$$\bar{d} \sim N(2.51,0.29)$$
+
+```{r}
+ggplot(data.frame(x=0:6),aes(x=x))+
+  stat_function(fun=dnorm, args=list(mean=2.51, sd=0.29), color="purple")+
+  ylab("Density")+
+  scale_x_continuous(seq(0,6,1),name="Wage Difference",labels=scales::dollar)+
+  theme_light()
+```
+
+Now we the **$t$-test** like any other:
+
+$$\begin{align*}
+t&=\frac{\text{estimate}-\text{null hypothesis}}{\text{standard error of the estimate}}\\
+&=\frac{d-0}{SE(d)}\\
+&=\frac{2.51-0}{0.29}\\
+&=8.66\\
+\end{align*}$$
+
+This is statistically significant. The $p$-value, $P(t>8.66)=$ is 0.000000000000000000410, or basically, 0.
+
+```{r}
+pt(8.66,456.33, lower.tail=FALSE)
+```
+
+### The $t$-test in `R`
+
+```{r}
+t.test(wage~female, data=wages, var.equal=FALSE)
+```
+
+```{r}
+reg<-lm(wage~female, data=wages)
+summary(reg)
+```

### 2019-11-19 content/class/19-class.Rmd
Update paper deadlines and data sources

@@
 
+Another way to visualize what we've been covering:
+
+```{r, fig.retina=3}
+library(mosaic)
+
+control=function(x){2+0.5*x}
+treat=function(x){4+x}
+
+ggplot(data.frame(x=c(0,10)), aes(x=x))+
+  stat_function(fun=control, geom="line", size=2, color = "red")+
+    geom_label(aes(x=5,y=control(5)), color = "red", label=expression(paste("slope=",hat(beta[1]))), size = 5)+
+  stat_function(fun=treat, geom="line", size=2, color = "blue")+
+    geom_label(aes(x=5,y=treat(5)), color = "blue", label=expression(paste("slope=",hat(beta[1])+hat(beta[3]))), size = 5)+
+    scale_x_continuous(breaks=NULL,
+                     limits=c(0,10),
+                     expand=expand_scale(mult=c(0,0.1)))+
+  scale_y_continuous(breaks=c(2,4),
+                     labels=c(expression(hat(beta[0])),expression(hat(beta[0])+hat(beta[2]))),
+                     limits=c(0,10),
+                     expand=expand_scale(mult=c(0,0.1)))+
+  labs(x = "",
+       y = "")+
+  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
+
+```
+
+$$\widehat{\text{wage}}_i=\hat{\beta_0}+\hat{\beta_1} \, \text{experience}_i$$
+
+$$\widehat{\text{wage}}_i=\hat{\beta_0}+\hat{\beta_1} \, \text{experience}_i+\hat{\beta_2} \, \text{gender}_i$$
+
+$$\widehat{\text{wage}}_i=\hat{\beta_0}+\hat{\beta_1} \, \text{experience}_i+\hat{\beta_2} \, \text{gender}_i+ \hat{\beta_3}(\text{experience}_i \times \text{gender}_i)$$
+
+```{r}
+library(tidyverse)
+library(wooldridge)
+# Our data comes from wage1 in the wooldridge package
+
+wages<-wooldridge::wage1
+```
+
+```{r, fig.retina=3}
+ggplot(data = wages)+
+  aes(x = exper,
+      y = wage)+
+  geom_point()+
+  geom_smooth(method="lm")+
+  scale_y_continuous(labels=scales::dollar)+
+  labs(x = "Experience (years)",
+       y = "Wage")+
+  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
+```
+
+```{r}
+library(broom)
+gender_reg<-lm(wage~exper+female, data = wages) %>%
+  tidy()
+
+gender_reg
+
+slope<-gender_reg %>%
+  filter(term=="exper") %>%
+  pull(estimate)
+slope
+
+male_intercept<-gender_reg %>%
+  filter(term=="(Intercept)")%>%
+  pull(estimate)
+
+female_intercept<-gender_reg %>%
+  filter(term=="female") %>%
+  pull(estimate)
+
+female_intercept<-female_intercept+male_intercept
+```
+
+```{r, fig.retina=3}
+colors = c("1" = "pink", "0" = "blue")
+
+ggplot(data = wages)+
+  aes(x = exper,
+      y = wage,
+      color = as.factor(female))+
+  geom_point()+
+  geom_abline(slope=slope, intercept = male_intercept, size=2, color = "blue")+
+  geom_abline(slope=slope, intercept = female_intercept, size=2, color = "pink")+
+  geom_label(x=2.5,y=6.63, color="blue", label="6.63")+
+  geom_label(x=2.5,y=4.15, color="pink", label="4.15")+
+  scale_y_continuous(labels=scales::dollar)+
+  scale_color_manual("Gender", labels=c("Men","Women"), values = colors)+
+  guides(color=F)+
+  labs(x = "Experience (years)",
+       y = "Wage",
+       title = "Parallel Slopes Model",
+       subtitle = "wage = 6.63+0.03 Experience-2.48 Female")+
+  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
+```
+]
+
 # Slides

### 2019-11-19 content/r/19-r-practice.Rmd
Fix typos on practice 19, HW 4 due Thursday not Tuesday
Update paper deadlines and data sources

@@
 
-- [<i class="fas fa-table"></i> `speeding_tickets.csv`](/data/speeding-tickets.csv)
+- [<i class="fas fa-table"></i> `speeding_tickets.csv`](/data/speeding_tickets.csv)
 
@@
 where
+
 - `df` is the name of your dataframe
-- `my_var` and `my_var2` are example variables^[As a bonus, you can try doing this with just one conditional command: `mutate_if(is.integer, factor)`.
+- `my_var` and `my_var2` are example variables^[As a bonus, you can try doing this with just one conditional command: `mutate_at(c("Black", "Hispanic", "Female", "OutTown", "OutState"),factor)`.
 See our [Data Wrangling slides](https://metricsf19.classes.ryansafner.com/slides/05-slides#95) for refreshers of all the fancy `mutate()` possibilities!]
@@
 where:
+
   - `y` is the variable we are testing
@@
 
-Make a new variable called `Male` and save it in your dataframe.^[Hint: `mutate()` and define `Male` equal to the `ifelse()` function, which has three arguments: 1.
-the condition to test, 2.
-what to define "Male" as when the condition is `TRUE`, and 3.
-what to define "Male" as when the condition is `FALSE`.]
+Make a new variable called `Male` and save it in your dataframe.^[Hint: `mutate()` and define `Male` equal to `factor(ifelse())`.
+This makes the variable a `factor` (so we don't have to later), and the `ifelse()` function has three arguments: 1.
+`test =` the condition to test, 2.
+`yes =` what to define "Male" as when the condition is `TRUE`, and 3.
+`no = ` what to define "Male" as when the condition is `FALSE`.]
 

### 2019-11-19 static/r/19-r-practice-rmd.Rmd
Fix typos on practice 19, HW 4 due Thursday not Tuesday
Update paper deadlines and data sources

@@
 - `df` is the name of your dataframe
-- `my_var` and `my_var2` are example variables^[As a bonus, you can try doing this with just one conditional command: `mutate_if(is.integer, factor)`.
+- `my_var` and `my_var2` are example variables^[As a bonus, you can try doing this with just one conditional command: `mutate_at(c("Black", "Hispanic", "Female", "OutTown", "OutState"),factor)`.
 See our [Data Wrangling slides](https://metricsf19.classes.ryansafner.com/slides/05-slides#95) for refreshers of all the fancy `mutate()` possibilities!]
@@
 ### Part A
-Make a new variable called `Male` and save it in your dataframe.^[Hint: `mutate()` and define `Male` equal to the `ifelse()` function, which has three arguments: 1.
-the condition to test, 2.
-what to define "Male" as when the condition is `TRUE`, and 3.
-what to define "Male" as when the condition is `FALSE`.]
+Make a new variable called `Male` and save it in your dataframe.^[Hint: `mutate()` and define `Male` equal to `factor(ifelse())`.
+This makes the variable a `factor` (so we don't have to later), and the `ifelse()` function has three arguments: 1.
+`test =` the condition to test, 2.
+`yes =` what to define "Male" as when the condition is `TRUE`, and 3.
+`no = ` what to define "Male" as when the condition is `FALSE`.]
 

### 2019-11-19 static/slides/20-slides.Rmd
Update paper deadlines and data sources

---
title: "3.8: Polynomial Regression"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))
```

# *Linear* Regression

.pull-left[

- OLS is commonly known as "**_linear_ regression**" as it fits a *straight line* to data points

- Often, data and relationships between variables may *not* be linear

]

--

.pull-right[

```{r, fig.retina=3}
library(gapminder)
p1<-ggplot(data = gapminder)+
  aes(x = gdpPercap,
      y = lifeExp)+
  geom_point(color="blue", alpha=0.5)+
  scale_x_continuous(labels=scales::dollar,
                     breaks=seq(0,120000,20000))+
  scale_y_continuous(breaks=seq(0,90,10),
                     limits=c(0,90))+
  labs(x = "GDP per Capita",
       y = "Life Expectancy (Years)")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
p1
```
]

---

# *Linear* Regression

.pull-left[

- OLS is commonly known as "**_linear_ regression**" as it fits a *straight line* to data points

- Often, data and relationships between variables may *not* be linear

]

.pull-right[

```{r, fig.retina=3}
p1+geom_smooth(method="lm", color="red")
```
]

---

# *Linear* Regression

.pull-left[

- OLS is commonly known as "**_linear_ regression**" as it fits a *straight line* to data points

- Often, data and relationships between variables may *not* be linear

]

.pull-right[

```{r, fig.retina=3}
p1+geom_smooth(method="lm", color="red")+
  stat_smooth(method="lm", formula=y~x+I(x^2), color = "green")
```
]

---

# *Linear* Regression

.pull-left[

- OLS is commonly known as "**_linear_ regression**" as it fits a *straight line* to data points

- Often, data and relationships between variables may *not* be linear

- Get rid of the outliers $(>60,000)$

]

.pull-right[

```{r, fig.retina=3}
gap_60<-gapminder %>%
  filter(gdpPercap<60000)

p2<-ggplot(data = gap_60)+
  aes(x = gdpPercap,
      y = lifeExp)+
  geom_point(color="blue", alpha=0.5)+
  scale_x_continuous(labels=scales::dollar,
                     breaks=seq(0,60000,20000))+
  scale_y_continuous(breaks=seq(0,90,10),
                     limits=c(0,90))+
  labs(x = "GDP per Capita",
       y = "Life Expectancy (Years)")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
p2
```
]

---

# *Linear* Regression

.pull-left[

- OLS is commonly known as "**_linear_ regression**" as it fits a *straight line* to data points

- Often, data and relationships between variables may *not* be linear

- Get rid of the outliers $(>60,000)$

]

.pull-right[

```{r, fig.retina=3}
p2+geom_smooth(method="lm", color="red")
```
]

---

# *Linear* Regression

.pull-left[

- OLS is commonly known as "**_linear_ regression**" as it fits a *straight line* to data points

- Often, data and relationships between variables may *not* be linear

- Get rid of the outliers $(>60,000)$

]

.pull-right[

```{r, fig.retina=3}
p2+geom_smooth(method="lm", color="red")+
  stat_smooth(method="lm", formula=y~x+I(x^2), color = "green")
```
]

---

# *Linear* Regression

.pull-left[

- OLS is commonly known as "**_linear_ regression**" as it fits a *straight line* to data points

- Often, data and relationships between variables may *not* be linear

- Get rid of the outliers $(>60,000)$

]

.pull-right[

```{r, fig.retina=3}
p2+geom_smooth(method="lm", color="red")+
  stat_smooth(method="lm", formula=y~x+I(x^2), color = "green")+
  stat_smooth(method = "lm", formula=y~log(x), color="orange", size=2)
```
]

---

# Nonlinear Effects in Linear Regression

- Despite being "linear regression", OLS can handle this with an easy fix

- OLS requires all *parameters* (i.e. the $\beta$'s) to be linear, the *regressors* $(X$'s) can be nonlinear:

--
$$Y_i=\beta_0+\beta_1 X_i^2 \quad  \mathbf{\checkmark}$$
--

$$Y_i=\beta_0+\beta_1^2X_i \quad  \mathbf{X}$$

--

$$Y_i=\beta_0+\beta_1 \sqrt{X_i} \quad  \mathbf{\checkmark}$$

--

$$Y_i=\beta_0+\sqrt{\beta_1} X_i \quad  \mathbf{X} $$

--

$$Y_i=\beta_0+\beta_1 (X_{1i} \times X_{2i}) \quad  \mathbf{\checkmark}$$

--

$$Y_i=\beta_0+\beta_1 ln(X_i) \quad  \mathbf{\checkmark}$$
--

- In the end, each $X$ is always just a number in the data, OLS can always estimate parameters for it

- *Plotting* the modelled points $(X_i, \hat{Y_i})$ can result in a curve!

---

# Sources of Nonlinearities I

- Effect of $X_1 \rightarrow Y$ might be nonlinear if:

--

1. $X_1 \rightarrow Y$ is different for different levels of $X_1$
  - e.g. diminishing returns: $\uparrow X_1$ increases $Y$ at a *decreasing* rate
  - e.g. increasing returns: $\uparrow X_1$ increases $Y$ at an *increasing* rate

--

2. $X_1 \rightarrow Y$ is different for different levels of $X_2$
  - e.g. interaction effects (last lesson)

---

# Sources of Nonlinearities II

.pull-left[

- **Linear**: 
  - slope $(\hat{\beta_1})$, $\frac{\Delta X}{\Delta Y}$ same for all $X$

]
.pull-right[

```{r fig.retina=3}

line=function(x){10-x}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=line, geom="line", size=2, color="blue")+
  geom_segment(aes(x=5, xend=5, y=line(5), yend=line(6)), linetype="dotted", color="black", size=1)+
  geom_segment(aes(x=5, xend=6, y=line(6), yend=line(6)), linetype="dotted", color="black", size=1)+
  scale_x_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  labs(x = "X",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```
]

---

# Sources of Nonlinearities II

.pull-left[

- **Polynomial**:
  - slope $(\hat{\beta_1})$, $\frac{\Delta X}{\Delta Y}$ *depends on the value of* $X$

]
.pull-right[

```{r fig.retina=3}

curve=function(x){10/x}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=curve, geom="line", size=2, color="blue")+
  geom_segment(aes(x=1, xend=1, y=5, yend=9), linetype="dotted", color="black", size=1)+
  geom_segment(aes(x=1, xend=2, y=5, yend=5), linetype="dotted", color="black", size=1)+
  geom_segment(aes(x=5, xend=5, y=2, yend=1), linetype="dotted", color="black", size=1)+
  geom_segment(aes(x=5, xend=9, y=1, yend=1), linetype="dotted", color="black", size=1)+
  scale_x_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  labs(x = "X",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```
]

---

# Sources of Nonlinearities III

.pull-left[

- **Interaction Effect**:
  - slope $(\hat{\beta_1})$, $\frac{\Delta X_1}{\Delta Y}$ *depends on the value of* $X_2$

- Easy example: if $X_2$ is a dummy variable:
  - .red[`\\(X_2=0\\)` (control)] vs. .green[`\\(X_2=1\\)` (treatment)]

]
.pull-right[

```{r fig.retina=3}
treat=function(x){4+x}
control=function(x){2+0.5*x}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=treat, geom="line", size=2, color="green")+
  geom_label(x=3, y=treat(3), color="green", label=expression(paste("When ",x[2]==1)))+
  geom_segment(aes(x=5, xend=6, y=treat(5), yend=treat(5)), linetype="dotted", color="black", size=1)+
  geom_segment(aes(x=6, xend=6, y=treat(5), yend=treat(6)), linetype="dotted", color="black", size=1)+
  stat_function(fun=control, geom="line", size=2, color="red")+
  geom_label(x=3, y=control(3), color="red", label=expression(paste("When ",x[2]==0)))+
  geom_segment(aes(x=5, xend=6, y=control(5), yend=control(5)), linetype="dotted", color="black", size=1)+
  geom_segment(aes(x=6, xend=6, y=control(5), yend=control(6)), linetype="dotted", color="black", size=1)+

  scale_x_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  labs(x = "X",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```
]

---

# Polynomial Functions of $X$

$$\hat{Y_i} = \hat{\beta_0} + \hat{\beta_1} X_i + \hat{\beta_2} X_i^2 + \cdots + \hat{\beta_{\mathbf{r}}} X_i^{\mathbf{r}} + \epsilon_i$$

- Where $r$ is the highest power $X_i$ is raised to
  - e.g. quadratic $r=2$, cubic $r=3$, etc

- The graph of an $r$<sup>th</sup>-degree polynomial function has $(r-1)$ bends

- Just another multivariate OLS regression model

---

# Quadratic Model

$$\hat{Y_i} = \hat{\beta_0} + \hat{\beta_1} X_i + \hat{\beta_2} X_i^2$$

- .shout[Quadratic model] has $X$ and $X^2$ variables in it (yes, need both!)

--

- How to interpret coefficients (betas)?
  - $\beta_0$ as "intercept" and $\beta_1$ as "slope" makes no sense 
  - $\beta_1$ as effect $X_i \rightarrow Y_i$ holding $X_i^2$ constant makes no sense<sup>.red[1]</sup>

.footnote[<sup>.red[1]</sup> This is *not* a multicollinearity problem! Correlation only measures *linear* relationships!]

--

- **Calculate marginal effects** by calculating predicted $\hat{Y_i}$ for different levels of $X_i$

---

# Quadratic Model: Calculating Marginal Effects

$$\hat{Y_i} = \hat{\beta_0} + \hat{\beta_1} X_i + \hat{\beta_2} X_i^2$$

- What is the .shout[marginal effect] of $\Delta X_i \rightarrow \Delta Y_i$?

--

- Take the **derivative** of $Y_i$ with respect to $X_i$:
$$\frac{d Y_i}{d X_i} = \hat{\beta_1}+2\hat{\beta_2} X_i$$

--

- .onfire[Marginal effect] of a 1 unit change in $X_i$ is a $\left(\hat{\beta_1}+2\hat{\beta_2} X_i \right)$ unit change in $Y$

---

# Quadratic Model: Example I

.content-box-green[
.green[**Example**]: $$\widehat{\text{Life Expectancy}_i} = \hat{\beta_0}+\hat{\beta_1} \, \text{GDP per capita}_i+\hat{\beta_2}\, \text{GDP per capita}^2_i$$
]

- Use `gapminder` package and data

```{r, echo=T}
library(gapminder)
```

---

# Quadratic Model: Example II

- These coefficients will be very large, so let's transform `gdpPercap` to be in $1,000's

```{r, echo=T}
gapminder<-gapminder %>%
  mutate(GDP_t = gdpPercap/1000)
gapminder %>% head() # look at it
```

---

# Quadratic Model: Example III

- Let's also add the squared term, `gdp_sq`

```{r, echo=T}
gapminder<-gapminder %>%
  mutate(GDP_sq = GDP_t^2)
gapminder %>% head() # look at it
```

---

# Quadratic Model: Example IV

- Can "manually" run a multivariate regression with `GDP_t` and `GDP_sq`

.font80[
```{r, echo=T}
reg1<-lm(lifeExp ~ GDP_t + GDP_sq, data = gapminder)
summary(reg1)
```
]

---

# Quadratic Model: Example V

- OR use `gdp_t` and add the "transform" command in regression, `I(gdp_t^2)`

.font80[
```{r, echo=T}
reg1_alt<-lm(lifeExp ~ GDP_t + I(GDP_t^2), data = gapminder)
summary(reg1_alt)
```
]

---

# Quadratic Model: Example VI

.pull-left[
```{r quadreg, eval=F, echo=T}
library(broom)
tidy(reg1)
```
]

.pull-right[
```{r, ref.label="quadreg"}
```
]

--

$$\widehat{\text{Life Expectancy}_i} = 50.52+1.55 \, \text{GDP per capita}_i - 0.02\, \text{GDP per capita}^2_i$$

--

- Marginal effect of GDP per capita on Life Expectancy:

--

$$\begin{align*}
\frac{d \, Y}{d \; X} &= \hat{\beta_1}+2\hat{\beta_2} X_i\\
\frac{d \, \text{Life Expectancy}}{d \, \text{GDP}} &= 1.55+2(-0.02) \, \text{GDP}\\
 &= \mathbf{1.55-0.04 \, \text{GDP}}\\
\end{align*}$$

---

# Quadratic Model: Example VII

$$\frac{d \, \text{Life Expectancy}}{d \, \text{GDP}} = 1.55-0.04 \, \text{GDP}$$
- Positive, with diminishing returns

- Effect on Life Expectancy of increasing GDP depends on initial value of GDP!

---

# Quadratic Model: Example VIII

$$\frac{d \, \text{Life Expectancy}}{d \, \text{GDP}} = 1.55-0.04 \, \text{GDP}$$

Marginal effect of GDP if GDP $=\$5$ (thousand):

$$\begin{align*}
\frac{d \, \text{Life Expectancy}}{d \, \text{GDP}} &= 1.55-0.04\text{GDP}\\
&= 1.55-0.04(5)\\
&= 1.55-0.20\\
&=1.35\\
\end{align*}$$

--

- i.e. for every addition $1 (thousand) in GDP per capita, average life expectancy increases by 1.35 years

---

# Quadratic Model: Example IX

$$\frac{d \, \text{Life Expectancy}}{d \, \text{GDP}} = 1.55-0.04 \, \text{GDP}$$

Marginal effect of GDP if GDP $=\$25$ (thousand):

--

$$\begin{align*}
\frac{d \, \text{Life Expectancy}}{d \, \text{GDP}} &= 1.55-0.04\text{GDP}\\
&= 1.55-0.04(25)\\
&= 1.55-1.00\\
&=0.55\\
\end{align*}$$

--

- i.e. for every addition $1 (thousand) in GDP per capita, average life expectancy increases by 0.55 years

---

# Quadratic Model: Example X

$$\frac{d \, \text{Life Expectancy}}{d \, \text{GDP}} = 1.55-0.04 \, \text{GDP}$$

Marginal effect of GDP if GDP $=\$50$ (thousand):

--

$$\begin{align*}
\frac{d \, \text{Life Expectancy}}{d \, \text{GDP}} &= 1.55-0.04\text{GDP}\\
&= 1.55-0.04(50)\\
&= 1.55-2.00\\
&=-0.45\\
\end{align*}$$

--

- i.e. for every addition $1 (thousand) in GDP per capita, average life expectancy *decreases* by 0.45 years

---

# Quadratic Model: Example XI

$$\begin{align*}\widehat{\text{Life Expectancy}_i} &= 50.52+1.55 \, \text{GDP per capita}_i - 0.02\, \text{GDP per capita}^2_i \\
\frac{d \, \text{Life Expectancy}}{d \, \text{GDP}} &= 1.55-0.04\text{GDP} \\ \end{align*}$$

| *Initial* GDP per capita | Marginal Effect<sup>.red[1]<sup> |
|----------------|-------------------:|
| $5,000 | $1.35$ years |
| $25,000 | $0.55$ years |
| $50,000 | $-0.45$ years |

.footnote[<sup>.red[1]</sup> Of +$1,000 GDP/capita on Life Expectancy.]

---

# Quadratic Model: Example XII

.pull-left[

.font90[
```{r quadplot, echo=T, eval=F}

ggplot(data = gapminder)+
  aes(x = GDP_t,
      y = lifeExp)+
  geom_point(color="blue", alpha=0.5)+
  stat_smooth(method = "lm", #<<
              formula = y ~ x + I(x^2), #<<
              color="green")+ #<<
  geom_vline(xintercept=c(5,25,50),
             linetype="dashed",
             color="red", size = 1)+
  scale_x_continuous(labels=scales::dollar,
                     breaks=seq(0,120,10))+
  scale_y_continuous(breaks=seq(0,90,10),
                     limits=c(0,90))+
  labs(x = "GDP per Capita (in Thousands)",
       y = "Life Expectancy (Years)")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]
]

.pull-right[
```{r, ref.label="quadplot", fig.retina=3}
```
]

---

# Quadratic Model: Maxima and Minima I

- For a polynomial model, we can also find the predicted **maximum** or **minimum** of $\hat{Y_i}$

--

- A quadratic model has a single global maximum or minimum (1 bend)

--

- By calculus, a minimum or maximum occurs where:

$$\begin{align*}
		\frac{ d Y_i}{d X_i} &=0\\
		\beta_1 + 2\beta_2 X_i &= 0\\
		2\beta_2 X_i&= -\beta_1\\
		X_i^*&=-\frac{1}{2}\frac{\beta_1}{\beta_2}\\
		\end{align*}$$

---

# Quadratic Model: Maxima and Minima II

.pull-left[

```{}
tidy(reg1)
```
]
--

.pull-right[

$$\begin{align*}
GDP_i^*&=-\frac{1}{2}\frac{\beta_1}{\beta_2}\\
GDP_i^*&=-\frac{1}{2}\frac{(1.55)}{(-0.015)}\\
GDP_i^*& \approx -\frac{1}{2}(-103.333)\\
GDP_i^*& \approx 51.67\\ \end{align*}$$

]

---

# Quadratic Model: Maxima and Minima III

.pull-left[

```{r quadplotmax, echo=T, eval=F}
ggplot(data = gapminder)+
  aes(x = GDP_t,
      y = lifeExp)+
  geom_point(color="blue", alpha=0.5)+
  stat_smooth(method = "lm", #<<
              formula = y ~ x + I(x^2), #<<
              color="green")+ #<<
  geom_vline(xintercept=51.67, linetype="dashed", color="red", size = 1)+
  geom_label(x=51.67, y=90, label="$51.67", color="red")+
  scale_x_continuous(labels=scales::dollar,
                     breaks=seq(0,120,10))+
  scale_y_continuous(breaks=seq(0,90,10),
                     limits=c(0,90))+
  labs(x = "GDP per Capita (in Thousands)",
       y = "Life Expectancy (Years)")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[
```{r, ref.label="quadplotmax", fig.retina=3}
```
]

---

# Determining if Higher-Order Polynomials are Necessary I

```{r, echo=F}
tidy(reg1)
```

- Is the quadratic term necessary?

--

- Determine if $\hat{\beta_2}$ (on $X_i^2)$ is statistically significant:
  - $H_0: \hat{\beta_2}=0$
  - $H_a: \hat{\beta_2} \neq 0$

--

- Statistically significant $\implies$ we should keep the quadratic model
  - If we only ran a linear model, it would be incorrect!

---

# Determining if Higher-Order Polynomials are Necessary I

.pull-left[
- Should we keep going up in polynomials?

$$\widehat{\text{Life Expectancy}_i} = \hat{\beta_0}+\hat{\beta_1} GDP_i+\hat{\beta_2}GDP^2_i+\hat{\beta_3}GDP_i^3$$

]

.pull-right[
```{r, fig.retina=3}
p2+geom_smooth(method="lm", color="red")+
  stat_smooth(method="lm", formula=y~x+I(x^2), color = "green")+
  stat_smooth(method="lm", formula=y~x+I(x^2)+I(x^3), color = "purple")
```

]

---

# Determining if Higher-Order Polynomials are Necessary II

.pull-left[

- In general, should have a compelling theoretical reason why data or relationships should "change direction" multiple times

- Or clear data patterns that have multiple "bends"

]

.pull-right[
```{r, fig.retina=3}
# make obviously cubic data
df_3<-tibble(x = seq(0,20,0.05),
           y = (500 + 0.4 * (x-10)^3)+rnorm(length(x), 10, 50)) # real function + random noise

ggplot(data = df_3)+
  aes(x = x,
      y = y)+
  geom_point(color="blue")+
  stat_smooth(method = "lm", formula = y~x+I(x^2)+I(x^3), color="red")+
  labs(x = "X",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

---

# A Second Polynomial Example I

.pull-left[

.content-box-green[
.green[**Example**]: How does a school district's average income affect Test scores?
]

$$\widehat{\text{Test Score}_i}=\hat{\beta_0}+\hat{\beta_1}\text{Income}_i$$

]

--

.pull-right[
```{r}
library(haven)
CASchool<-read_dta("../data/caschool.dta")
```

```{r, fig.retina=3}
income_plot<-ggplot(data = CASchool)+
  aes(x = avginc,
      y = testscr)+
  geom_point(color="blue")+
  scale_x_continuous(labels=scales::dollar)+
  scale_y_continuous(limits=c(600,710))+
  labs(x = "District Average Income (Thousands of $)",
       y = "Test Scores")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
income_plot
```
]

---

# A Second Polynomial Example I

.pull-left[

.content-box-green[
.green[**Example**]: How does a school district's average income affect Test scores?
]

$$\widehat{\text{Test Score}_i}=\hat{\beta_0}+\hat{\beta_1}\text{Income}_i$$

]

.pull-right[
```{r, fig.retina=3}
income_plot+geom_smooth(method="lm", color="red")
```
]

---

# A Second Polynomial Example I

.pull-left[

.content-box-green[
.green[**Example**]: How does a school district's average income affect Test scores?
]

$$\widehat{\text{Test Score}_i}=\hat{\beta_0}+\hat{\beta_1}\text{Income}_i$$

]

.pull-right[
```{r, fig.retina=3}
income_plot+geom_smooth(method="lm", color="red")+
  stat_smooth(method="lm", formula=y~x+I(x^2), color="green")
```
]

---

# A Second Polynomial Example II

.font80[
```{r}
income_quad_reg<-lm(testscr~avginc+I(avginc^2), data = CASchool)
summary(income_quad_reg)
```
]

---

# A Second Polynomial Example III

.pull-left[
- Should we keep going?

.font60[
```{r}
income_quad_reg<-lm(testscr~avginc+I(avginc^2)+I(avginc^3), data = CASchool)
summary(income_quad_reg)
```
]
]

.pull-right[
```{r, fig.retina=3}
income_plot+geom_smooth(method="lm", color="red")+
  stat_smooth(method="lm", formula=y~x+I(x^2), color="green")+
  stat_smooth(method="lm", formula=y~x+I(x^2)+I(x^3), color="orange")
```

]

---

# Strategy for Polynomial Model Specification

1. Are there good theoretical reasons for relationships changing (e.g. increasing/decreasing returns)?

--

2. Plot your data: does a straight line fit well enough?

--

3. Specify a polynomial function of a higher power (start with 2) and estimate OLS regression

--

4. Use $t$-test to determine if higher-power term is significant

--

5. Interpret effect of change in $X$ on $Y$

--

6. Repeat steps 3-5 as necessary

### 2019-11-19 content/class/20-class.Rmd
Add class 20 content

---
title: "3.8: Polynomial Regression - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

*Tuesday, November 19, 2019*

# Overview

This week we will cover non-linear models, which we anticipated last week with an interaction effect between two continuous variables: $X_1 \times X_2$, where the marginal effect of $X_1 \rightarrow Y$ is dependent on the value of $X_2$.

We cover two types of nonlinear models. *Today*, we cover **polynomial models**, where we include higher-order $X$ variables to explain a non-linear effect. *Next* class we will cover **logarithmic models**, which are more frequently used - as well as a few final tests and transformations for multivariate regression before we turn to panel data.

Today, we will spend the first half of class finishing [last class' practice problems](/practice/19-r-practice) (on dummy variables and interaction effects), then I will cover polynomial models. After we finish next class' lecture, we will do another set of R practice problems, and you will be given **HW 5** to work on this material.

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/19-slides.html)

# R Practice Problems

We will begin by finishing your work on [Last Class' Practice Problems](/r/17-r-practice). Answers are posted there tonight.

# Problem Set 4 Due Thursday Nov 21

[Problm Set 4](/assignment/04-problem-set) (on classes 3.1-3.5) is due by Thursday November 21.

### 2019-11-19 static/slides/20-slides.Rmd
Tweak slides 20

@@
 1. $X_1 \rightarrow Y$ is different for different levels of $X_1$
-  - e.g. diminishing returns: $\uparrow X_1$ increases $Y$ at a *decreasing* rate
-  - e.g. increasing returns: $\uparrow X_1$ increases $Y$ at an *increasing* rate
+  - e.g. **diminishing returns**: $\uparrow X_1$ increases $Y$ at a *decreasing* rate
+  - e.g. **increasing returns**: $\uparrow X_1$ increases $Y$ at an *increasing* rate
 
@@
 
-# Polynomial Functions of $X$
-
-$$\hat{Y_i} = \hat{\beta_0} + \hat{\beta_1} X_i + \hat{\beta_2} X_i^2 + \cdots + \hat{\beta_{\mathbf{r}}} X_i^{\mathbf{r}} + \epsilon_i$$
+# Polynomial Functions of $X$ I
+
+.pull-left[
+
+- .blue[Linear], $X$
+
+]
+
+.pull-right[
+
+```{r, fig.retina=3}
+library("mosaic")
+linear=function(x){x}
+quadratic=function(x){x^2}
+cubic=function(x){x^3}
+quartic=function(x){x^4}
+
+ggplot(data.frame(x=c(-10,10)), aes(x=x))+
+  stat_function(fun=linear, geom="line", size=2, color = "blue")+
+    scale_x_continuous(breaks=seq(-1,1,1),
+                     limits=c(-2,2),
+                     expand=expand_scale(mult=c(0,0.1)))+
+  scale_y_continuous(breaks=seq(-1,1,1),
+                     limits=c(-1,1))+
+  #                   expand=expand_scale(mult=c(0,0.1)),
+  #                   labels = function(x){paste("$", x, sep="")})+
+  coord_cartesian(clip="off")+
+  labs(x = "X",
+       y = "Y")+
+  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
+
+```
+]
+
+---
+
+# Polynomial Functions of $X$ I
+
+.pull-left[
+
+- .blue[Linear], $X$
+
+- .green[Quadratic], $X^2$
+
+]
+
+.pull-right[
+
+```{r, fig.retina=3}
+
+ggplot(data.frame(x=c(-10,10)), aes(x=x))+
+  stat_function(fun=linear, geom="line", size=2, color = "blue")+
+  stat_function(fun=quadratic, geom="line", size=2, color = "green")+
+    scale_x_continuous(breaks=seq(-1,1,1),
+                     limits=c(-2,2),
+                     expand=expand_scale(mult=c(0,0.1)))+
+  scale_y_continuous(breaks=seq(-1,1,1),
+                     limits=c(-1,1))+
+  #                   expand=expand_scale(mult=c(0,0.1)),
+  #                   labels = function(x){paste("$", x, sep="")})+
+  coord_cartesian(clip="off")+
+  labs(x = "X",
+       y = "Y")+
+  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
+
+```
+]
+
+---
+# Polynomial Functions of $X$ I
+
+.pull-left[
+
+- .blue[Linear], $X$
+
+- .green[Quadratic], $X^2$
+
+- .orange[Cubic], $X^3$
+
+]
+
+.pull-right[
+
+```{r, fig.retina=3}
+ggplot(data.frame(x=c(-10,10)), aes(x=x))+
+  stat_function(fun=linear, geom="line", size=2, color = "blue")+
+  stat_function(fun=quadratic, geom="line", size=2, color = "green")+
+  stat_function(fun=cubic, geom="line", size=2, color = "orange")+
+    scale_x_continuous(breaks=seq(-1,1,1),
+                     limits=c(-2,2),
+                     expand=expand_scale(mult=c(0,0.1)))+
+  scale_y_continuous(breaks=seq(-1,1,1),
+                     limits=c(-1,1))+
+  #                   expand=expand_scale(mult=c(0,0.1)),
+  #                   labels = function(x){paste("$", x, sep="")})+
+  coord_cartesian(clip="off")+
+  labs(x = "X",
+       y = "Y")+
+  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
+
+```
+]
+
+---
+
+# Polynomial Functions of $X$ I
+
+.pull-left[
+
+- .blue[Linear], $X$
+
+- .green[Quadratic], $X^2$
+
+- .orange[Cubic], $X^3$
+
+- .purple[Quartic], $X^4$
+]
+
+.pull-right[
+
+```{r, fig.retina=3}
+
+ggplot(data.frame(x=c(-10,10)), aes(x=x))+
+  stat_function(fun=linear, geom="line", size=2, color = "blue")+
+  stat_function(fun=quadratic, geom="line", size=2, color = "green")+
+  stat_function(fun=cubic, geom="line", size=2, color = "orange")+
+  stat_function(fun=quartic, geom="line", size=2, color = "purple")+
+    scale_x_continuous(breaks=seq(-1,1,1),
+                     limits=c(-2,2),
+                     expand=expand_scale(mult=c(0,0.1)))+
+  scale_y_continuous(breaks=seq(-1,1,1),
+                     limits=c(-1,1))+
+  #                   expand=expand_scale(mult=c(0,0.1)),
+  #                   labels = function(x){paste("$", x, sep="")})+
+  coord_cartesian(clip="off")+
+  labs(x = "X",
+       y = "Y")+
+  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
+
+```
+]
+
+---
+
+# Polynomial Functions of $X$ I
+
+$$\hat{Y_i} = \hat{\beta_0} + \hat{\beta_1} X_i + \hat{\beta_2} X_i^2 + \cdots + \hat{\beta_{\mathbf{r}}} X_i^{\mathbf{r}} + u_i$$
+--
 
 - Where $r$ is the highest power $X_i$ is raised to
-  - e.g. quadratic $r=2$, cubic $r=3$, etc
+  - quadratic $r=2$
+  - cubic $r=3$
+
+--
 
@@
 
-- Just another multivariate OLS regression model
+--
+
+- Just another multivariate OLS regression model!
+
+---
+
+class: inverse, center, middle
+
+# The Quadratic Model
 
@@
 
-Marginal effect of GDP if GDP $=\$5$ (thousand):
+Marginal effect of GDP if GDP $=5$ ($ thousand):
 
@@
 
-Marginal effect of GDP if GDP $=\$25$ (thousand):
+Marginal effect of GDP if GDP $=25$ ($ thousand):
 
@@
 
-Marginal effect of GDP if GDP $=\$50$ (thousand):
+Marginal effect of GDP if GDP $=50$ ($ thousand):
 
@@
 
+class: inverse, center, middle
+
+# The Quadratic Model: Maxima and Minima
+
+---
+
 # Quadratic Model: Maxima and Minima I
@@
 
+class: inverse, center, middle
+
+# Are Polynomials Necessary?
+
+---
+
 # Determining if Higher-Order Polynomials are Necessary I

### 2019-11-20 static/slides/20-slides.Rmd
Last tweak to slides 20
Add R practice 19 answers

@@
 - **Linear**: 
-  - slope $(\hat{\beta_1})$, $\frac{\Delta X}{\Delta Y}$ same for all $X$
+  - slope $(\hat{\beta_1})$, $\frac{\Delta Y}{\Delta X}$ same for all $X$
 
@@
 - **Polynomial**:
-  - slope $(\hat{\beta_1})$, $\frac{\Delta X}{\Delta Y}$ *depends on the value of* $X$
+  - slope $(\hat{\beta_1})$, $\frac{\Delta Y}{\Delta X}$ *depends on the value of* $X$
 
@@
 - **Interaction Effect**:
-  - slope $(\hat{\beta_1})$, $\frac{\Delta X_1}{\Delta Y}$ *depends on the value of* $X_2$
+  - slope $(\hat{\beta_1})$, $\frac{\Delta Y}{\Delta X_1}$ *depends on the value of* $X_2$
 
@@
 
-```{}
+```{r}
 tidy(reg1)
@@
 ]
+
 --

### 2019-11-21 content/assignment/04-problem-set.Rmd
Fix typos on practice 19, HW 4 due Thursday not Tuesday
Fix problem set 4 links! Also change that weird 404 gif

@@
 
-*Due by Tuesday, November 19, 2019*
+*Due by Thursday, November 21, 2019*
 
 - [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/04-problem-set-pdf.pdf)
-- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/04-problem-set.Rmd)
+- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/04-problem-set-rmd.Rmd)
 
@@
 
-# Question 7
-Consider
-
 # R Questions
@@
 
-## Question 11
+## Question 7
 

### 2019-11-21 static/homeworks/04-problem-set-pdf.Rmd
Fix problem set 4 links! Also change that weird 404 gif

--- 2019-11-21 content/assignment/04-problem-set.Rmd
@@
 title: "Problem Set 4"
-editor_options: 
-  chunk_output_type: console
+author: "ECON 480 - Fall 2019"
+date: "Due by Thursday, October 10, 2019"
+output: pdf_document
 ---
 
-```{r message=FALSE, warning=FALSE, include=FALSE}
-library(tidyverse)
-library(pander)
-source(file.path(here::here(), "lib", "graphics.R"))
-```
-
-*Due by Thursday, November 21, 2019*
-
-- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/04-problem-set-pdf.pdf)
-- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/04-problem-set-rmd.Rmd)
+<!--CLICK "KNIT" ABOVE TO RENDER TO HTML, PDF, OR WORD OUTPUT
+In fact, try knitting right away and see what this produces!
+-->
 
@@
 
-## Question 1
-In your own words, describe what omitted variable bias means. What are the two conditions for an omitted variable to cause a bias?
+1. In your own words, describe what omitted variable bias means. What are the two conditions for an omitted variable to cause a bias?
 
-## Question 2
+\vspace{3in}
+
+2.
 In your own words, describe what multicollinearity means.
@@
 
-## Question 3
-In your own words, describe what a proxy variable is. When or why would we use a proxy variable, and what effects does it have on our model?
+\clearpage
 
-## Question 4
+3. In your own words, describe what a proxy variable is. When or why would we use a proxy variable, and what effects does it have on our model?
+
+\vspace{3in}
+
+4.
 Explain how we use Directed Acyclic Graphs (DAGs) to depict a causal model: what are the two criteria that must hold for identifying a causal effect of $X$ on $Y$?
 When should we control a variable, and when should we *not* control a variable?
+
+\clearpage
 
@@
 
-## Question 5
-
-Data were collected from a random sample of 220 home sales from a community in 2017.
+5. Data were collected from a random sample of 220 home sales from a community in 2017.
 
@@
 
-### Part A
-Suppose that a homeowner converts part of an existing living space in her house to a new bathroom. What is the expected increase in the value of the house?
+a. Suppose that a homeowner converts part of an existing living space in her house to a new bathroom. What is the expected increase in the value of the house?
 
-### Part B
-Suppose a homeowner adds a new bathroom to her house, which also increases the size of the house by 100 square feet. What is the expected increase in the value of the house?
+\vspace{1.5in}
 
-### Part C
-Suppose the $R^2$ of this regression is 0.727. Calculate the adjusted $\bar{R}^2$.
+b. Suppose a homeowner adds a new bathroom to her house, which also increases the size of the house by 100 square feet. What is the expected increase in the value of the house?
 
-### Part D
-Suppose the following auxiliary regression for $BDR$ has an $R^2$ of 0.841.
+\vspace{1.5in}
+
+c. Suppose the $R^2$ of this regression is 0.727. Calculate the adjusted $\bar{R}^2$.
+
+\clearpage
+
+d. Suppose the following auxiliary regression for $BDR$ has an $R^2$ of 0.841.
 
@@
 
-Calculate the Variance Inflation Factor for $BDR$ and explain what it means.
+> Calculate the Variance Inflation Factor for $BDR$ and explain what it means.
 
-## Question 6
+\clearpage
 
-A researcher wants to investigate the effect of education on average hourly wages. Wage, education, and experience in the dataset have the following correlations:
+6. A researcher wants to investigate the effect of education on average hourly wages. Wage, education, and experience in the dataset have the following correlations:
 
@@
 
-### Part A
+a.
 If the true marginal effect of experience on wages (holding education constant) is 0.0701, calculate the omitted variable bias in the first regression caused by omitting experience.
@@
 
-### Part B
-Knowing this, what would be the *true effect* of education on wages, holding experience constant?
+\vspace{2in}
 
-### Part C
+b. Knowing this, what would be the *true effect* of education on wages, holding experience constant?
+
+\vspace{2in}
+
+c.
 The $R^2$ for the second regression is 0.0897.
@@
 Why?
+
+\clearpage
 
@@
 
-## Question 7
-
-- [<i class="fas fa-table"></i> `heightwages.csv`](http://metricsf19.classes.ryansafner.com/data/heightwages.csv)
-
+11.
 Download the `heightwages.csv` dataset.
@@
 
-### Part A
-
-Create a quick scatterplot between `height85` (as $X)$ amd `wage96` (as $Y)$.
-
-### Part B
-
-Regress wages on adult height. Write the equation of the estimated OLS regression. Interpret the coefficient on `height85`.
-
-### Part C
-
-How much would someone who is 5'10" be predicted to earn per hour, according to the model?
-
-### Part D
-
-Would adolescent height cause an omitted variable bias if it were left out? Explain using both your intuition, and some statistical evidence with `R`.
-
-### Part E
-
-Now add adolescent height to the regression, and write the new regression equation below, as before. Interpret the coefficient on `height85`.
-
-### Part F
-
-How much would someone who is 5'10" in 1985 and 4'8" in 1981 be predicted to earn, according to the model?
-
-### Part G
-
-What happened to the estimate on `height85` and its standard error?
-
-### Part H
-
-Is there multicollinearity between `height85` and `height81`? Explore with a scatterplot.^[Hint: to avoid overplotting, use `geom_jitter()` instead of `geom_point()` to get a better view of the data.]
-
-### Part I
-
-Quantify how much multicollinearity affects the variance of the OLS estimates on both heights.^[Hint: You'll need the `car` package.]
-
-### Part J
-
+a. Create a quick scatterplot between `height85` (as $X)$ amd `wage96` (as $Y)$.
+b. Regress wages on adult height. Write the equation of the estimated OLS regression. Interpret the coefficient on `height85`.
+c. How much would someone who is 5'10" be predicted to earn per hour, according to the model?
+d. Would adolescent height cause an omitted variable bias if it were left out? Explain using both your intuition, and some statistical evidence with `R`.
+e. Now add adolescent height to the regression, and write the new regression equation below, as before. Interpret the coefficient on `height85`.
+f. How much would someone who is 5'10" in 1985 and 4'8" in 1981 be predicted to earn, according to the model?
+g. What happened to the estimate on `height85` and its standard error?
+h.
+Is there multicollinearity between `height85` and `height81`?
+Explore with a scatterplot.**^[Hint: to avoid overplotting, use `geom_jitter()` instead of `geom_point()` to get a better view of the data.]
+i. Quantify how much multicollinearity affects the variance of the OLS estimates on both heights.^[Hint: You'll need the `car` package.]
+j.
 Reach the same number as in part I by running an auxiliary regression.^[Hint: There's some missing `wage96` data that may give you a different answer, so `filter()` your data here by `!is.na(wage96)` before running this regression - this will include only observations for `wage96` that are not `NA`'s.]
-
-### Part K
-
-Make a regression table from part B and D using `huxtable`.
+k. Make a regression table from part B and D using `huxtable`.

### 2019-11-21 static/homeworks/04-problem-set.Rmd
Fix problem set 4 links! Also change that weird 404 gif

--- 2019-11-21 content/assignment/04-problem-set.Rmd
@@
 title: "Problem Set 4"
-editor_options: 
-  chunk_output_type: console
+author: "YOUR NAME HERE" # put your name here! 
+date: "ECON 480 - Fall 2019"
+output: html_document # change to pdf if you'd like, this will make a webpage (you can email it and open in any browser)
 ---
 
-```{r message=FALSE, warning=FALSE, include=FALSE}
-library(tidyverse)
-library(pander)
-source(file.path(here::here(), "lib", "graphics.R"))
-```
+<!--CLICK "KNIT" ABOVE TO RENDER TO HTML, PDF, OR WORD OUTPUT
+In fact, try knitting right away and see what this produces!
+-->
 
 *Due by Thursday, November 21, 2019*
-
-- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/04-problem-set-pdf.pdf)
-- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/04-problem-set-rmd.Rmd)
 
@@
 ## Question 1
-In your own words, describe what omitted variable bias means. What are the two conditions for an omitted variable to cause a bias?
+**In your own words, describe what omitted variable bias means. What are the two conditions for an omitted variable to cause a bias?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ## Question 2
-In your own words, describe what multicollinearity means.
+**In your own words, describe what multicollinearity means.
 What is the cause, and what are the consequences of multicollinearity?
 How can we measure multicollinearity and its effects?
-What happens if multicollinearity is *perfect*?
+What happens if multicollinearity is *perfect*?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ## Question 3
-In your own words, describe what a proxy variable is. When or why would we use a proxy variable, and what effects does it have on our model?
+
+**In your own words, describe what a proxy variable is. When or why would we use a proxy variable, and what effects does it have on our model?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ## Question 4
-Explain how we use Directed Acyclic Graphs (DAGs) to depict a causal model: what are the two criteria that must hold for identifying a causal effect of $X$ on $Y$?
-When should we control a variable, and when should we *not* control a variable?
+
+**Explain how we use Directed Acyclic Graphs (DAGs) to depict a causal model: what are the two criteria that must hold for identifying a causal effect of $X$ on $Y$?
+When should we control a variable, and when should we *not* control a variable?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
@@
 ### Part A
-Suppose that a homeowner converts part of an existing living space in her house to a new bathroom. What is the expected increase in the value of the house?
+
+**Suppose that a homeowner converts part of an existing living space in her house to a new bathroom. What is the expected increase in the value of the house?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part B
-Suppose a homeowner adds a new bathroom to her house, which also increases the size of the house by 100 square feet. What is the expected increase in the value of the house?
+
+**Suppose a homeowner adds a new bathroom to her house, which also increases the size of the house by 100 square feet. What is the expected increase in the value of the house?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part C
-Suppose the $R^2$ of this regression is 0.727. Calculate the adjusted $\bar{R}^2$.
+
+**Suppose the $R^2$ of this regression is 0.727. Calculate the adjusted $\bar{R}^2$.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part D
-Suppose the following auxiliary regression for $BDR$ has an $R^2$ of 0.841.
+
+**Suppose the following auxiliary regression for $BDR$ has an $R^2$ of 0.841.**
 
@@
 
-Calculate the Variance Inflation Factor for $BDR$ and explain what it means.
+**Calculate the Variance Inflation Factor for $BDR$ and explain what it means.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
@@
 ### Part A
-If the true marginal effect of experience on wages (holding education constant) is 0.0701, calculate the omitted variable bias in the first regression caused by omitting experience.
-Does the estimate of $\hat{\beta_1}$ in the first regression overstate or understate the effect of education on wages?
+
+**If the true marginal effect of experience on wages (holding education constant) is 0.0701, calculate the omitted variable bias in the first regression caused by omitting experience.
+Does the estimate of $\hat{\beta_1}$ in the first regression overstate or understate the effect of education on wages?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part B
-Knowing this, what would be the *true effect* of education on wages, holding experience constant?
+
+**Knowing this, what would be the *true effect* of education on wages, holding experience constant?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part C
-The $R^2$ for the second regression is 0.0897.
+
+**The $R^2$ for the second regression is 0.0897.
 If she were to run a better regression including both education and experience, how much would the variance of the coefficients on education and experience increase?
-Why?
+Why?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
@@
 
-Create a quick scatterplot between `height85` (as $X)$ amd `wage96` (as $Y)$.
+**Create a quick scatterplot between `height85` (as $X)$ amd `wage96` (as $Y)$.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 
-Regress wages on adult height. Write the equation of the estimated OLS regression. Interpret the coefficient on `height85`.
+**Regress wages on adult height. Write the equation of the estimated OLS regression. Interpret the coefficient on `height85`.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 
-How much would someone who is 5'10" be predicted to earn per hour, according to the model?
+**How much would someone who is 5'10" be predicted to earn per hour, according to the model?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 
-Would adolescent height cause an omitted variable bias if it were left out? Explain using both your intuition, and some statistical evidence with `R`.
+**Would adolescent height cause an omitted variable bias if it were left out? Explain using both your intuition, and some statistical evidence with `R`.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 
-Now add adolescent height to the regression, and write the new regression equation below, as before. Interpret the coefficient on `height85`.
+**Now add adolescent height to the regression, and write the new regression equation below, as before. Interpret the coefficient on `height85`.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 
-How much would someone who is 5'10" in 1985 and 4'8" in 1981 be predicted to earn, according to the model?
+**How much would someone who is 5'10" in 1985 and 4'8" in 1981 be predicted to earn, according to the model?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 
-What happened to the estimate on `height85` and its standard error?
+**What happened to the estimate on `height85` and its standard error?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 
-Is there multicollinearity between `height85` and `height81`? Explore with a scatterplot.^[Hint: to avoid overplotting, use `geom_jitter()` instead of `geom_point()` to get a better view of the data.]
+**Is there multicollinearity between `height85` and `height81`?
+Explore with a scatterplot.**^[Hint: to avoid overplotting, use `geom_jitter()` instead of `geom_point()` to get a better view of the data.]
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 
-Quantify how much multicollinearity affects the variance of the OLS estimates on both heights.^[Hint: You'll need the `car` package.]
+**Quantify how much multicollinearity affects the variance of the OLS estimates on both heights**.^[Hint: You'll need the `car` package.]
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 
-Reach the same number as in part I by running an auxiliary regression.^[Hint: There's some missing `wage96` data that may give you a different answer, so `filter()` your data here by `!is.na(wage96)` before running this regression - this will include only observations for `wage96` that are not `NA`'s.]
+**Reach the same number as in part I by running an auxiliary regression.**^[Hint: There's some missing `wage96` data that may give you a different answer, so `filter()` your data here by `!is.na(wage96)` before running this regression - this will include only observations for `wage96` that are not `NA`'s.]
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 
-Make a regression table from part B and D using `huxtable`.
+**Make a regression table from part B and D using `huxtable`.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---

### 2019-11-21 content/r/21-r-practice.Rmd
Add class 21 content

---
title: "3.9: Logarithmic Regression - R Practice"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

# Set Up
To minimize confusion, I suggest creating a new `R Project` (e.g. `regression_practice`) and storing any data in that folder on your computer.

Alternatively, I have made a project in R Studio Cloud that you can use (and not worry about trading room computer limitations), with the data already inside (you will still need to assign it to an object).

- [<i class="fas fa-cloud"></i> View Project on R Studio Cloud](https://rstudio.cloud/project/739808)
- [<i class="fab fa-markdown"></i> `Download Markdown`](/r/21-r-practice-rmd.Rmd)

## Question 1

We are returning to the speeding tickets data that we began to explore in [R Practice 15 on Multivariate Regression](http://metricsf19.classes.ryansafner.com/practice/15-practice) and [R Practice 19 on Dummy Variables & Interaction Effects](http://metricsf19.classes.ryansafner.com/practice/19-r-practice). Download and read in (`read_csv`) the data below.

- [<i class="fas fa-table"></i> `speeding_tickets.csv`](/data/speeding_tickets.csv)

This data again comes from a paper by Makowsky and Strattman (2009) that we will examine later. Even though state law sets a formula for tickets based on how fast a person was driving, police officers in practice often deviate from that formula. This dataset includes information on all traffic stops. An amount for the fine is given only for observations in which the police officer decided to assess a fine. There are a number of variables in this dataset, but the one's we'll look at are:

| Variable | Description |
|----------|-------------|
| `Amount` | Amount of fine (in dollars) assessed for speeding |
| `Age` | Age of speeding driver (in years) |
| `MPHover` | Miles per hour over the speed limit |
| `Black` | Dummy $=1$ if driver was black, $=0$ if not |
| `Hispanic` | Dummy $=1$ if driver was Hispanic, $=0$ if not |
| `Female` | Dummy $=1$ if driver was female, $=0$ if not |
| `OutTown` | Dummy $=1$ if driver was not from local town, $=0$ if not |
| `OutState` | Dummy $=1$ if driver was not from local state, $=0$ if not |
| `StatePol` | Dummy $=1$ if driver was stopped by State Police, $=0$ if stopped by other (local) |

> We again want to explore who gets fines, and how much.

## Question 2

Run a regression of `Amount` on `Age`. Write out the estimated regression equation, and interpret the coefficient on Age.

## Question 3

Is the effect of `Age` on `Amount` nonlinear? Let's run a quadratic regression.

### Part A
Create a new variable for $Age^2$. Then run a quadratic regression.

### Part B
Try running the same regression using the alternate notation: `lm(Y~X+I(X^2))`. This method allows you to not have to create a new variable first. Do you get the same results?

### Part C
Write out the estimated regression equation.

### Part D
Is this model an improvement from the linear model?^[Check $R^2$.]

### Part E
Write an equation for the marginal effect of `Age` on `Amount`.

### Part F
Predict the marginal effect on `Amount` of being one year older when you are 18.  How about when you are 40?

### Part G
Our quadratic function is a $U$-shape. According to the model, at what age is the amount of the fine minimized?

### Part H
Create a scatterplot between `Amount` and `Age` and add a a layer of a linear regression (as always), and an additional layer of your predicted quadratic regression curve. The regression curve, just like any regression *line*, is a `geom_smooth()` layer on top of the `geom_point()` layer. We will need to customize `geom_smooth()` to `geom_smooth(method="lm", formula="y~x+I(x^2)` (copy/paste this verbatim)! This is the same as a regression line (`method="lm"`), but we are modifying the formula to a polynomial of degree 2 (quadratic): $y=a+bx+cx^2$.

### Part I

It's quite hard to see the quadratic curve with all those data points. Redo another plot and this time, only keep the quadratic `stat_smooth()` layer and leave out the `geom_point()` layer. This will only plot the regression curve.

# Question 4

Should we use a higher-order polynomial equation? Run a cubic regression, and determine whether it is necessary.

# Question 5

Run an $F$-test to check if a nonlinear model is appropriate. Your null hypothesis is $H_0: $\beta_2+\beta_3=0$ from the regression in pert (h). The command is `linearHypothesis(reg_name, c("var1", "var2"))` where `reg_name` is the name of the `lm` object you saved your regression in, and `var1` and `var2` (or more) in quotes are the names of the variables you are testing. This function requires (installing and) loading the "`car`" package (additional regression tools).

# Question 6

Now let's take a look at speed (`MPHover` the speed limit).

## Part A
Creating new variables as necessary, run a **linear-log** model of `Amount` on `MPHover`. Write down the estimated regression equation, and interpret the coefficient on `MPHover` $(\hat{\beta_1})$. Make a scatterplot with the regression line.^[Hint: The simple `geom_smooth(method="lm")` is sufficient, so long as you use the right variables on the plot!]

## Part B
Creating new variables as necessary, run a **log-linear** model of `Amount` on `MPHover`. Write down the estimated regression equation, and interpret the coefficient on `MPHover` $(\hat{\beta_1})$. Make a scatterplot with the regression line.^[Hint: The simple `geom_smooth(method="lm")` is sufficient, so long as you use the right variables on the plot!]

## Part C
Creating new variables as necessary, run a **log-log** model of `Amount` on `MPHover`. Write down the estimated regression equation, and interpret the coefficient on `MPHover` $(\hat{\beta_1})$. Make a scatterplot with the regression line.^[Hint: The simple `geom_smooth(method="lm")` is sufficient, so long as you use the right variables on the plot!]

## Part D
Which of the three log models has the best fit?^[Hint: Check $R^2$]

# Question 7
Return to the quadratic model. Run a quadratic regression of `Amount` on `Age`, `Age`$^2$, `MPHover`, and all of the race dummy variables. Test the null hypothesis: *"the race of the driver has no effect on Amount"*

# Question 8

Now let's try standardizing variables. Let's try running a regression of `Amount` on `Age` and `MPHover`, but standardizing each variable.

### Part A

Create new standardized variables for `Amount`, `Age`, and `MPHover`.^[Hint: use the `scale()` function inside of `mutate()`]

### Part B

Run a regression of standardized `Amount` on standardized `Age` and `MPHover`. Interpret $\hat{\beta_1}$ and $\hat{\beta_2}$ Which variable has a bigger effect on `Amount`?

### 2019-11-21 static/r/21-r-practice-rmd.Rmd
Add class 21 content

--- 2019-11-21 content/r/21-r-practice.Rmd
@@
 ---
-title: "3.9: Logarithmic Regression - R Practice"
-editor_options: 
-  chunk_output_type: console
+title: "3.9: Logarithmic Models - R Practice"
+author: "YOUR NAME"
+date: "`r Sys.Date()`"
+output:
+  html_document:
+    df_print: paged
+    #theme: 
+    toc: true 
+    toc_depth: 3
+    toc_float: true
+    code_folding: show
+    highlight: tango
 ---
 
-```{r message=FALSE, warning=FALSE, include=FALSE}
-library(tidyverse)
-library(pander)
-source(file.path(here::here(), "lib", "graphics.R"))
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
 ```
@@
 Run a regression of `Amount` on `Age`. Write out the estimated regression equation, and interpret the coefficient on Age.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 ### Part A
+
 Create a new variable for $Age^2$. Then run a quadratic regression.
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part B
+
 Try running the same regression using the alternate notation: `lm(Y~X+I(X^2))`. This method allows you to not have to create a new variable first. Do you get the same results?
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part C
+
 Write out the estimated regression equation.
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part D
+
 Is this model an improvement from the linear model?^[Check $R^2$.]
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part E
+
 Write an equation for the marginal effect of `Age` on `Amount`.
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part F
+
 Predict the marginal effect on `Amount` of being one year older when you are 18.  How about when you are 40?
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part G
+
 Our quadratic function is a $U$-shape. According to the model, at what age is the amount of the fine minimized?
 
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
 ### Part H
+
 Create a scatterplot between `Amount` and `Age` and add a a layer of a linear regression (as always), and an additional layer of your predicted quadratic regression curve.
@@
 This is the same as a regression line (`method="lm"`), but we are modifying the formula to a polynomial of degree 2 (quadratic): $y=a+bx+cx^2$.
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
 
@@
 
-# Question 4
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
+## Question 4
 
@@
 
-# Question 5
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
+## Question 5
 
@@
 
-# Question 6
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
+## Question 6
 
@@
 
-## Part A
+### Part A
 Creating new variables as necessary, run a **linear-log** model of `Amount` on `MPHover`.
@@
 
-## Part B
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
+### Part B
+
 Creating new variables as necessary, run a **log-linear** model of `Amount` on `MPHover`.
@@
 
-## Part C
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
+### Part C
+
 Creating new variables as necessary, run a **log-log** model of `Amount` on `MPHover`.
@@
 
-## Part D
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
+### Part D
+
 Which of the three log models has the best fit?^[Hint: Check $R^2$]
 
-# Question 7
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
+## Question 7
+
 Return to the quadratic model.
@@
 
-# Question 8
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# PUT CODE HERE
+```
+
+---
+
+## Question 8
 

### 2019-11-21 static/slides/21-slides.Rmd
Add class 21 content

---
title: "3.9: Logarithmic Regression"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))
```

# Nonlinearities

.pull-left[

- Consider the `gapminder` example again
]

.pull-right[

```{r, fig.retina=3}
library(gapminder)
gap_60<-gapminder %>%
  filter(gdpPercap<60000)

p2<-ggplot(data = gap_60)+
  aes(x = gdpPercap,
      y = lifeExp)+
  geom_point(color="blue", alpha=0.5)+
  scale_x_continuous(labels=scales::dollar,
                     breaks=seq(0,60000,20000))+
  scale_y_continuous(breaks=seq(0,90,10),
                     limits=c(0,90))+
  labs(x = "GDP per Capita",
       y = "Life Expectancy (Years)")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)

p2
```

]

---

# Nonlinearities

.pull-left[

- Consider the `gapminder` example again
  - .red[linear model]
]

.pull-right[

```{r, fig.retina=3}
p2+geom_smooth(method="lm", color="red")
```

]

---

# Nonlinearities

.pull-left[

- Consider the `gapminder` example again
  - .red[linear model]
  - .green[polynomial model (quadratic)]
]

.pull-right[

```{r, fig.retina=3}
p2+geom_smooth(method="lm", color="red")+
  stat_smooth(method="lm", formula=y~x+I(x^2), color = "green")
```

]

---

# Nonlinearities

.pull-left[

- Consider the `gapminder` example again
  - .red[linear model]
  - .green[polynomial model (quadratic)]
  - .orange[logarithmic model]
]

.pull-right[

```{r, fig.retina=3}
p2+geom_smooth(method="lm", color="red")+
  stat_smooth(method="lm", formula=y~x+I(x^2), color = "green")+
  stat_smooth(method = "lm", formula=y~log(x), color="orange", size=2)
```

]

---

class: inverse, center, middle

# Logarithmic Models

---

# Logarithmic Models

.pull-left[
```{r, fig.retina=3}
ggplot(data.frame(x=0:10),aes(x=x))+
    stat_function(fun=log, color="blue",size =2)+
  geom_label(x=9,y=log(9), label="y=log(x)", color="blue", size=5)+
  scale_x_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=c(0,0))+
  scale_y_continuous(limits=c(0,2.5),
                     expand=c(0,0))+
  labs(x = "X",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

- Another model specification for nonlinear data is the .shout[logarithmic model]<sup>.red[1]</sup>
  - We transform either $X$, $Y$, or *both* by taking the (natural) logarithm

- Logarithmic model has two additional advantages
  1. We can easily interpret coefficients as **percentage changes** or **elasticities**
  2. Useful economic shape: diminishing returns (production functions, utility functions, etc)

.footnote[<sup>.red[1]</sup> Note, this should not be confused with a **logistic model**, which is a model for dependent dummy variables.]

]

---

# The Natural Logarithm

.pull-left[
```{r, fig.retina=3}
ggplot(data.frame(x=0:10),aes(x=x))+
    stat_function(fun=log, color="blue",size =2)+
    stat_function(fun=exp, color="red", size =2)+
  stat_function(fun=function(x){x}, size=2, linetype="dashed")+
  geom_label(x=4,y=log(4), label=expression(Y==ln(X)), color="blue", size=5)+
  geom_label(x=1,y=exp(1), label=expression(Y==e^{X}), color="red", size=5)+
  scale_x_continuous(breaks=seq(0,5,1),
                     limits=c(0,5),
                     expand=c(0,0))+
  scale_y_continuous(limits=c(0,5),
                     expand=c(0,0))+
  labs(x = "X",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

- The .red[exponential function], $e^x$ or **$exp(x)$**, where base $e=2.71828...$

- .blue[Natural logarithm] is the inverse, $y=ln(x)$

]

---

# The Natural Logarithm: Review

- **Exponents** are defined as
$$b^n=\underbrace{b \times b \times ... \times b}_{n}$$
where base $b$ is multiplied by itself $n$ times

--
- .green[**Example**]: $2^3=\underbrace{2 \times 2 \times 2}_{n=3}=8$

--

- **Logarithms** the inverse, defined as the exponents in the expressions above
$$\text{If } b^n=y\text{, then }log_b(y)=n$$
$n$ is the number you must raise $b$ to in order to get $y$

--

- .green[**Example**]: $log_2(8)=3$

--

- Logarithms can have any base, but common to use the **natural logarithm $(ln)$** with base $\mathbf{e=2.71828...}$
$$\text{If } e^n=y\text{, then }ln(y)=n$$

---

# The Natural Logarithm: Properties

- Natural logs have a lot of useful properties:
  - $ln(\frac{1}{x})=-ln(x)$
  - $ln(ab)=ln(a)+ln(b)$
  - $ln(\frac{x}{a})=ln(x)-ln(a)$
  - $ln(x^a)=a \, ln(x)$
  - $\frac{d  \, ln \, x}{d \, x} = \frac{1}{x}$

---

# The Natural Logarithm: Example

- Most useful property: for small change in $x$, $\Delta x$:

$$\underbrace{ln(x+\Delta x) - ln(x)}_{\text{Difference in logs}} \approx \underbrace{\frac{\Delta x}{x}}_{\text{Relative change}}$$

--

.content-box-green[
.green[**Example**]: Let $x=100$ and $\Delta x =1$, relative change is:

$$\frac{\Delta x}{x} = \frac{(101-100)}{100} = 0.01 \text{ or }1\%$$

- The logged difference:
$$ln(101)-ln(100) = 0.00995 \approx 1\%$$
]

--

- This allows us to very easily interpret coefficients as *percent changes* or **elasticities**

---

# Elasticity

- An .shout[elasticity] between two variables, $E_{Y,X}$ describes the .whisper[responsiveness] of one variable to a change in another.

- Measured in percentages: a 1% change in $X$ will cause a $E\%$ change in $Y$

--

$$\begin{align*}
E_{y,x}&=\frac{\% \Delta y}{\% \Delta x}\\
&=\cfrac{\frac{\Delta y}{y}}{\frac{\Delta x}{x}}\\
\end{align*}$$

--

- Numerator is relative change in $Y$, Denominator is relative change in $X$

---

# Math FYI: Cobb Douglas Functions and Logs

- One of the (many) reasons why economists love Cobb-Douglas functions:
$$Y=AL^{\alpha}K^{\beta}$$

--

- Taking logs, relationship becomes linear:

--

$$ln(Y)=ln(A)+\alpha ln(L)+ \beta ln(K)$$

--

- With data on $(Y, L, K)$ and linear regression, can estimate $\alpha$ and $\beta$
  - $\alpha$: elasticity of $Y$ with respect to $L$
      - A 1% change in $L$ will lead to an $\alpha$% change in $Y$
  - $\beta$: elasticity of $Y$ with respect to $K$
      - A 1% change in $K$ will lead to a $\beta$% change in $Y$

---

# Logarithms in R I

- The `log()` function can easily take the logarithm

```{r}
gapminder<-gapminder %>%
  mutate(loggdp=log(gdpPercap))

gapminder
```

---

# Logarithms in R II

- Note, `log()` by default is the **natural logarithm $ln()$**, i.e. base `e`
  - Can change base with e.g. `log(x, base = 5)`
  - Some common built-in logs: `log10`, `log2`

```{r, echo=T}
log10(100)
log2(16)
log(19683, base=3)
```

---

# Types of Logarithmic Models

- Three types of log regression models, depending on which variables we log

--

1.  **Linear-log model:** $Y_i=\beta_0+\beta_1$ .red[`\\(ln(X_i)\\)`]

--

2. **Log-linear model:** .red[`\\(ln(Y_i)\\)`] $=\beta_0+\beta_1X_i$

--

3. **Log-log model:** .red[`\\(ln(Y_i)\\)`] $=\beta_0+\beta_1$ .red[`\\(ln(X_i)\\)`]

---

# Linear-Log Model

- .shout[Linear-log model] has an independent variable ($X$) that is logged

--

$$\begin{align*}
Y&=\beta_0+\beta_1 \mathbf{ln(X)}\\
\beta_1&=\cfrac{\Delta Y}{\big(\frac{\Delta X}{X}\big)}\\
\end{align*}$$

--

- $\hat{\beta_1}$: **a 1\% change in $X$** $\rightarrow$ **$\frac{\beta_1}{100}$ unit change in $Y$**

---

# Linear-Log Model in R

.pull-left[
```{r, echo=T}
lin_log_reg<-lm(lifeExp~loggdp, data = gapminder)
library(broom)
tidy(lin_log_reg)
```
]

.pull-right[

$$\widehat{\text{Life Expectancy}}_i=-9.10+9.41 \, \text{ln(GDP)}_i$$

]

---

# Linear-Log Model in R

.pull-left[
```{r, echo=T}
lin_log_reg<-lm(lifeExp~loggdp, data = gapminder)
library(broom)
tidy(lin_log_reg)
```
]

.pull-right[

$$\widehat{\text{Life Expectancy}}_i=-9.10+9.41 \, \text{ln(GDP)}_i$$

- $\hat{\beta_1}$: a **1% change in GDP** $\rightarrow$ a $\frac{9.41}{100}=$ **0.0941 year increase** in Life Expectancy

]

---

# Linear-Log Model in R

.pull-left[
```{r, echo=T}
lin_log_reg<-lm(lifeExp~loggdp, data = gapminder)
library(broom)
tidy(lin_log_reg)
```
]

.pull-right[

$$\widehat{\text{Life Expectancy}}_i=-9.10+9.41 \, \text{ln(GDP)}_i$$

- $\hat{\beta_1}$: a **1% change in GDP** $\rightarrow$ a $\frac{9.41}{100}=$ **0.0941 year increase** in Life Expectancy

- A **25% fall in GDP** $\rightarrow$ a $(25 \times 0.0941)=$ **2.353 year decrease** in Life Expectancy

]

---

# Linear-Log Model in R

.pull-left[
```{r, echo=T}
lin_log_reg<-lm(lifeExp~loggdp, data = gapminder)
library(broom)
tidy(lin_log_reg)
```
]

.pull-right[

$$\widehat{\text{Life Expectancy}}_i=-9.10+9.41 \, \text{ln(GDP)}_i$$

- $\hat{\beta_1}$: a **1% change in GDP** $\rightarrow$ a $\frac{9.41}{100}=$ **0.0941 year increase** in Life Expectancy

- A **25% fall in GDP** $\rightarrow$ a $(25 \times 0.0941)=$ **2.353 year decrease** in Life Expectancy

- A **100% rise in GDP** $\rightarrow$ a $(100 \times 0.0941)=$ **9.041 year increase** in Life Expectancy

]

---

# Linear-Log Model Graph I

.pull-left[
```{r lin-log-graph, echo=T, eval=F}
ggplot(data = gapminder)+
  aes(x = gdpPercap,
      y = lifeExp)+
  geom_point(color="blue", alpha=0.5)+
  geom_smooth(method="lm", #<<
              formula=y~log(x), #<<
              color="orange")+ #<<
  scale_x_continuous(labels=scales::dollar,
                     breaks=seq(0,120000,20000))+
  scale_y_continuous(breaks=seq(0,90,10),
                     limits=c(0,90))+
  labs(x = "GDP per Capita",
       y = "Life Expectancy (Years)")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[
```{r, ref.label="lin-log-graph", fig.retina=3}

```
]

---

# Linear-Log Model Graph II

.pull-left[
```{r lin-log-graph2, echo=T, eval=F}
ggplot(data = gapminder)+
  aes(x = loggdp, #<<
      y = lifeExp)+ 
  geom_point(color="blue", alpha=0.5)+
  geom_smooth(method="lm", color="orange")+ #<<
  scale_y_continuous(breaks=seq(0,90,10),
                     limits=c(0,90))+
  labs(x = "Log GDP per Capita",
       y = "Life Expectancy (Years)")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[
```{r, ref.label="lin-log-graph2", fig.retina=3}

```
]

---

# Log-Linear Model

- .shout[Log-linear model] has the dependent variable $(Y)$ logged

--

$$\begin{align*}
\mathbf{Y}&=\beta_0+\beta_1 X\\
\beta_1&=\cfrac{\big(\frac{\Delta Y}{Y}\big)}{X}\\
\end{align*}$$

--

- $\hat{\beta_1}$: **a 1 unit change in $X$** $\rightarrow$ **$\beta_1 \times 100\%$ change in $Y$**

---

# Log-Linear Model in R (Preliminaries)

- We will again have very large/small coefficients if we deal with GDP directly, again let's transform `gdpPercap` into $1,000s, call it `gdp_t`

- Then log LifeExp

--

```{r}
gapminder <- gapminder %>%
  mutate(gdp_t=gdpPercap/1000,
         loglife=log(lifeExp))
```

---

# Log-Linear Model in R

.pull-left[
```{r, echo=T}
log_lin_reg<-lm(loglife~gdp_t, data = gapminder)
tidy(log_lin_reg)
```
]

.pull-right[

$$\widehat{ln(\text{Life Expectancy})}_i=3.967+0.013 \, \text{GDP}_i$$

]

---

# Log-Linear Model in R

.pull-left[
```{r, echo=T}
log_lin_reg<-lm(loglife~gdp_t, data = gapminder)
tidy(log_lin_reg)
```
]

.pull-right[

$$\widehat{ln(\text{Life Expectancy})}_i=3.967+0.013 \, \text{GDP}_i$$

- $\hat{\beta_1}$: a **\$1 (thousand) change in (thousands of) GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3\% increase** in Life Expectancy

]

---

# Log-Linear Model in R

.pull-left[
```{r, echo=T}
log_lin_reg<-lm(loglife~gdp_t, data = gapminder)
tidy(log_lin_reg)
```
]

.pull-right[

$$\widehat{ln(\text{Life Expectancy})}_i=3.967+0.013 \, \text{GDP}_i$$

- $\hat{\beta_1}$: a **\$1 (thousand) change in (thousands of) GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3\% increase** in Life Expectancy

- A **\$25K fall in GDP** $\rightarrow$ a $(25 \times 1.3\%)=$ **32.5\% decrease** in Life Expectancy

]

---

# Log-Linear Model in R

.pull-left[
```{r, echo=T}
log_lin_reg<-lm(loglife~gdp_t, data = gapminder)
tidy(log_lin_reg)
```
]

.pull-right[

$$\widehat{ln(\text{Life Expectancy})}_i=3.967+0.013 \, \text{GDP}_i$$

- $\hat{\beta_1}$: a **\$1 (thousand) change in (thousands of) GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3\% increase** in Life Expectancy

- A **\$25K fall in GDP** $\rightarrow$ a $(25 \times 1.3\%)=$ **32.5\% decrease** in Life Expectancy

- A **\$100K rise in GDP** $\rightarrow$ a $(100 \times 1.3\%)=$ **130\% increase** in Life Expectancy

]

---

# Linear-Log Model Graph I

.pull-left[
```{r log-lin-graph, echo=T, eval=F}
ggplot(data = gapminder)+
  aes(x = gdp_t, #<<
      y = loglife)+ #<<
  geom_point(color="blue", alpha=0.5)+
  geom_smooth(method="lm", color="orange")+ #<<
  scale_x_continuous(labels=scales::dollar,
                     breaks=seq(0,120,20))+
  labs(x = "GDP per Capita ($ Thousands)",
       y = "Log Life Expectancy")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[
```{r, ref.label="log-lin-graph", fig.retina=3}

```
]

---

# Log-Log Model

- .shout[Log-log model] has both variables $(X \text{ and } Y)$ logged

--

$$\begin{align*}
\mathbf{Y}&=\beta_0+\beta_1 \mathbf{X}\\
\beta_1&=\cfrac{\big(\frac{\Delta Y}{Y}\big)}{\big(\frac{\Delta X}{X}\big)}\\
\end{align*}$$

--

- $\hat{\beta_1}$: **a 1% change in $X$** $\rightarrow$ **$\beta_1\%$ change in $Y$**

---

# Log-Log Model in R

.pull-left[
```{r, echo=T}
log_log_reg<-lm(loglife~loggdp, data = gapminder)
tidy(log_log_reg)
```
]

.pull-right[

$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.0147 \, \text{ln(GDP)}_i$$

]

---

# Log-Log Model in R

.pull-left[
```{r, echo=T}
log_log_reg<-lm(loglife~loggdp, data = gapminder)
tidy(log_log_reg)
```
]

.pull-right[

$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.0147 \, \text{ln(GDP)}_i$$

- $\hat{\beta_1}$: a **$1\% change in GDP** $\rightarrow$ a **0.147\% increase** in Life Expectancy

]

---

# Log-Log Model in R

.pull-left[
```{r, echo=T}
log_log_reg<-lm(loglife~loggdp, data = gapminder)
tidy(log_log_reg)
```
]

.pull-right[

$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.0147 \, \text{ln(GDP)}_i$$

- $\hat{\beta_1}$: a **$1\% change in GDP** $\rightarrow$ a **0.147\% increase** in Life Expectancy

- A **25\% fall in GDP** $\rightarrow$ a $(25 \times 0.147\%)=$ **3.675\% decrease** in Life Expectancy

]

---

# Log-Log Model in R

.pull-left[
```{r, echo=T}
log_log_reg<-lm(loglife~loggdp, data = gapminder)
tidy(log_log_reg)
```
]

.pull-right[

$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.0147 \, \text{ln(GDP)}_i$$

- $\hat{\beta_1}$: a **$1\% change in GDP** $\rightarrow$ a **0.147\% increase** in Life Expectancy

- A **25\% fall in GDP** $\rightarrow$ a $(25 \times 0.147\%)=$ **3.675\% decrease** in Life Expectancy

- A **100\% rise in GDP** $\rightarrow$ a $(100 \times 0.147\%)=$ **14.7\% increase** in Life Expectancy

]

---

# Log-Log Model Graph I

.pull-left[
```{r log-log-graph, echo=T, eval=F}
ggplot(data = gapminder)+
  aes(x = loggdp, #<<
      y = loglife)+ #<<
  geom_point(color="blue", alpha=0.5)+
  geom_smooth(method="lm", color="orange")+ #<<
  labs(x = "Log GDP per Capita",
       y = "Log Life Expectancy")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[
```{r, ref.label="log-log-graph", fig.retina=3}

```
]

---

# Comparing Models I

| Model | Equation | Interpretation |
|-------|----------|----------------|
| Linear-Log | $Y=\beta_0+\beta_1 \mathbf{ln(X)}$ | 1% change in $X \rightarrow \frac{\hat{\beta_1}}{100}$ **unit** change in $Y$ |
| Log-Linear | $\mathbf{ln(Y)}=\beta_0+\beta_1X$ | 1 **unit** change in $X \rightarrow \hat{\beta_1}\times 100$**%** change in $Y$ |
| Log-Log | $\mathbf{ln(Y)}=\beta_0+\beta_1\mathbf{ln(X)}$ | 1**%** change in $X \rightarrow \hat{\beta_1}$**%** change in $Y$ |

- Hint: the variable that gets logged changes in **percent** terms, the variable not logged changes in **unit** terms

---

# Comparing Models II

.pull-left[
```{r logs-table, echo=T, eval=F}
library(huxtable)
huxreg("Life Exp." = lin_log_reg,
       "Log Life Exp." = log_lin_reg,
       "Log Life Exp." = log_log_reg,
       coefs = c("Constant" = "(Intercept)",
                 "GDP ($1000s)" = "gdp_t",
                 "Log GDP" = "loggdp"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```

- Models are very different units, how to choose? 
  - Compare $R^2$'s
  - Compare graphs
  - Compare intution

]

.pull-left[
```{r, ref.label="logs-table"}
```

]

---

# When to Log?

- In practice, the following types of variables are logged (ln):
  - Variables that must always be positive (prices, sales, market values)
  - Very large numbers (population, GDP)
  - Variables expressed as percentages or percentage changes (inflation, population growth, GDP growth, labor force participation rate, unemployment rate)
  - Variables that have nonlinear scatterplots

--

- Never use logs for:
  - Variables that are less than one, decimals, 0, or negative (temperature)
  - Categorical variables
  - Time variables (year, week, day)

---

class: inverse, center, middle

# Comparing Across Units

---

# Comparing Coefficients of Different Units I

$$\hat{Y_i}=\beta_0+\beta_1 X_1+\beta_2 X_2	$$

- We often want to compare coefficients to see which variable $X_1$ or $X_2$ has a bigger effect on $Y$

- What if $X_1$ and $X_2$ are different units?

.content-box-green[
.green[**Example**]:
$$\begin{align*}
\widehat{\text{Salary}_i}&=\beta_0+\beta_1\, \text{Batting average}_i+\beta_2\, \text{Home runs}_i\\
\widehat{\text{Salary}_i}&=-\text{2,869,439.40}+\text{12,417,629.72} \, \text{Batting average}_i+\text{129,627.36}\, \text{Home runs}_i\\
\end{align*}$$
]

---

# Comparing Coefficients of Different Units II

- An easy way is to .shout[standardize] the variables (i.e. take the $Z$-score)

$$X^{std}=\frac{X-\overline{X}}{sd(X)}$$

---

# Comparing Coefficients of Different Units: Example

| Variable | Mean | Std. Dev. |
|----------|------|-----------|
| Salary | $2,024,616 | $2,764,512 |
| Batting Average | 0.267 | $0.031 |
| Home Runs | 12.11 | 10.31 |

$$\begin{align*}\scriptsize  
	\widehat{\text{Salary}_i}&=-\text{2,869,439.40}+\text{12,417,629.72} \, \text{Batting average}_i+\text{129,627.36} \, \text{Home runs}_i\\ 
	\widehat{\text{Salary}_i}^{std}&=\text{0.00}+\text{0.14} \, \text{Batting average}_i^{std}+\text{0.48} \, \text{Home runs}_i^{std}\\
	\end{align*}$$

--

- **Marginal effect** on $Y$ (in *standard deviations* of $Y$) from 1 *standard deviation* change in $X$
  - e.g. for 1 standard deviation increase in Batting Average, Salary increases by 0.14 standard deviations
  - $0.14\times 2,764,512=\$387,032$

---

# Standardizing in `R`

- Use the `scale()` command inside `mutate()` function to standardize a variable

```{r}
gapminder<-gapminder %>%
  mutate(std_life = scale(lifeExp),
         std_gdp = scale(gdpPercap))

std_reg<-lm(std_life~std_gdp, data = gapminder)
tidy(std_reg)
```

---

class: inverse, center, middle

# Joint Hypothesis Testing

---

# Joint Hypothesis Testing I

.content-box-green[
.green[**Example**]: Return again to:

$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Male+\hat{\beta_2}Northeast_i+\hat{\beta_3}Midwest_i+\hat{\beta_4}South_i$$

]

--

- Maybe region doesn't affect wages *at all*?

--

- $H_0: \beta_2=0, \, \beta_3=0, \, \beta_4=0$

--

- This is a .shout[joint hypothesis] to test

---

# Joint Hypothesis Testing II

- A .shout[joint hypothesis] tests against the null hypothesis of a value for *multiple* parameters:
$$\mathbf{H_0: \beta_1= \beta_2=0}$$
the hypotheses that multiple regressors are equal to zero (have no causal effect on the outcome)

--

- Our **alternative hypothesis** is that:
$$H_1: \text{ either } \beta_1\neq0\text{ or } \beta_2\neq0\text{ or both}$$
or simply, that $H_0$ is not true

---

# Types of Joint Hypothesis Tests

- Three main cases of joint hypothesis tests:

--

1. $H_0$: $\beta_1=\beta_2=0$
  - Testing against the claim that multiple variables don't matter
  - Useful under high multicollinearity between variables
  - $H_a$: at least one parameter $\neq$ 0

--

2. $H_0$: $\beta_1=\beta_2$
  - Testing whether two variables matter the same
  - Variables must be the same units
  - $H_a: \beta_1 (\neq, <, \text{ or }>) \beta_2$

--

3. $H_0:$ ALL $\beta$'s $=0$
  - The "**Overall F-test"**
  - Testing against claim that regression model explains *NO* variation in $Y$

---

# Joint Hypothesis Tests: F-statistic

- The .shout[F-statistic} is the test-statistic used to test joint hypotheses about regression coefficients with an .shout[F-test}

--

- This involves comparing two models:
  1. **Unrestricted model**: regression with all coefficients
  2. **Restricted model**: regression under null hypothesis (coefficients equal hypothesized values)

--

- $F$ is an .whisper[analysis of variance (ANOVA)]
  - essentially tests whether $R^2$ increases statistically significantly as we go from the restricted model$\rightarrow$unrestricted model

--

- $F$ has its own distribution, with *two* sets of degrees of freedom

---

# Joint Hypothesis F-test: Example I

.content-box-green[
.green[**Example**]: Return again to:

$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Male_i+\hat{\beta_2}Northeast_i+\hat{\beta_3}Midwest_i+\hat{\beta_4}South_i$$

]

--

- $H_0: \beta_2=\beta_3=\beta_4=0$

--

- $H_a$: $H_0$ is not true (at least one $\beta_i \neq 0$)

---

# Joint Hypothesis F-test: Example II

.content-box-green[
.green[**Example**]: Return again to:

$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Male_i+\hat{\beta_2}Northeast_i+\hat{\beta_3}Midwest_i+\hat{\beta_4}South_i$$

]

- **Unrestricted model**:

$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Male_i+\hat{\beta_2}Northeast_i+\hat{\beta_3}Midwest_i+\hat{\beta_4}South_i$$

--

- **Restricted model**:

$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Male_i$$

--

- $F$: does going from restricted to unrestricted statistically significantly improve $R^2$?

---

# Calculating the F-statistic

.pull-left[
$$F_{q,(n-k-1)}=\cfrac{\left(\displaystyle\frac{(R^2_u-R^2_r)}{q}\right)}{\left(\displaystyle\frac{(1-R^2_u)}{(n-k-1)}\right)}$$
]

.pull-right[

]

---

# Calculating the F-statistic

.pull-left[
$$F_{q,(n-k-1)}=\cfrac{\left(\displaystyle\frac{(R^2_u-R^2_r)}{q}\right)}{\left(\displaystyle\frac{(1-R^2_u)}{(n-k-1)}\right)}$$
]

.pull-right[

- $R^2_u$: the $R^2$ from the **unrestricted model** (all variables)
]

---

# Calculating the F-statistic

.pull-left[
$$F_{q,(n-k-1)}=\cfrac{\left(\displaystyle\frac{(R^2_u-R^2_r)}{q}\right)}{\left(\displaystyle\frac{(1-R^2_u)}{(n-k-1)}\right)}$$
]

.pull-right[

- $R^2_u$: the $R^2$ from the **unrestricted model** (all variables)

- $R^_r$: the $R^2$ from the **restricted model** (null hypothesis)
]

---

# Calculating the F-statistic

.pull-left[
$$F_{q,(n-k-1)}=\cfrac{\left(\displaystyle\frac{(R^2_u-R^2_r)}{q}\right)}{\left(\displaystyle\frac{(1-R^2_u)}{(n-k-1)}\right)}$$
]

.pull-right[

- $R^2_u$: the $R^2$ from the **unrestricted model** (all variables)

- $R^_r$: the $R^2$ from the **restricted model** (null hypothesis)

- $q$: number of restrictions (number of $\beta's=0$ under null hypothesis)
]

---

# Calculating the F-statistic

.pull-left[
$$F_{q,(n-k-1)}=\cfrac{\left(\displaystyle\frac{(R^2_u-R^2_r)}{q}\right)}{\left(\displaystyle\frac{(1-R^2_u)}{(n-k-1)}\right)}$$
]

.pull-right[

- $R^2_u$: the $R^2$ from the **unrestricted model** (all variables)

- $R^_r$: the $R^2$ from the **restricted model** (null hypothesis)

- $q$: number of restrictions (number of $\beta's=0$ under null hypothesis)

- $k$: number of $X$ variables in **unrestricted model** (all variables)
]

---

# Calculating the F-statistic

.pull-left[
$$F_{q,(n-k-1)}=\cfrac{\left(\displaystyle\frac{(R^2_u-R^2_r)}{q}\right)}{\left(\displaystyle\frac{(1-R^2_u)}{(n-k-1)}\right)}$$
]

.pull-right[

- $R^2_u$: the $R^2$ from the **unrestricted model** (all variables)

- $R^_r$: the $R^2$ from the **restricted model** (null hypothesis)

- $q$: number of restrictions (number of $\beta's=0$ under null hypothesis)

- $k$: number of $X$ variables in **unrestricted model** (all variables)

- $F$ has two sets of degrees of freedom, $q$ for the numerator, $n-k-1$ for the denominator
]

---

# Calculating the F-statistic II

.pull-left[
$$F_{q,(n-k-1)}=\cfrac{\left(\displaystyle\frac{(R^2_u-R^2_r)}{q}\right)}{\left(\displaystyle\frac{(1-R^2_u)}{(n-k-1)}\right)}$$
]

.pull-right[

- .whisper[Key takeaway]: The bigger the difference between $(R^2_u-R^2_r)$, the greater the improvement in fit by adding variables, the larger the $F$!

- This formula is (believe it or not) actually a simplified version (assuming homoskedasticity)
  - I give you this formula to **build your intuition of what F is measuring**

]

---

# F-test Example I

- We'll use the `wooldridge` package's `wage1` data again

```{r, echo=T}
# load in data from wooldridge package
library(wooldridge)
wages<-wooldridge::wage1

# run regressions
unrestricted_reg<-lm(wage~female+northcen+west+south, data=wages)
restricted_reg<-lm(wage~female, data=wages)
```

---

# F-test Example II

- Command is `linearHypothesis()`, comes from the `car` package

```{r, echo=T}
# load car package for additional regression tools
library("car")

# test that northcen=west=south=0
# # the first argument is the name of my regression ("unrestricted_reg")
linearHypothesis(unrestricted_reg, c("northcen", "west", "south")) 
```

---

# All F-test I

.font50[
```{r, echo=F, results="asis"}
summary(unrestricted_reg)
```
]

---

# All F-test II

```{r, echo=T}
library(broom)
glance(unrestricted_reg)
```

- "Statistic" is the All F-test, "p-value" next to it is the p value from the F test!

### 2019-11-21 static/slides/21-slides.Rmd
Fix errors on slides 21

@@
 
-- The .red[exponential function], $e^x$ or **$exp(x)$**, where base $e=2.71828...$
-
-- .blue[Natural logarithm] is the inverse, $y=ln(x)$
-
-]
-
----
-
-# The Natural Logarithm: Review
+- The .red[exponential function], $Y=e^X$ or $Y=exp(X)$, where base $e=2.71828...$
+
+- .blue[Natural logarithm] is the inverse, $Y=ln(X)$
+
+]
+
+---
+
+# The Natural Logarithm: Review I
 
 - **Exponents** are defined as
-$$b^n=\underbrace{b \times b \times ... \times b}_{n}$$
-where base $b$ is multiplied by itself $n$ times
+$$b^n=\underbrace{b \times b \times \cdots \times b}_{n \text{ times}}$$
+  - where base $b$ is multiplied by itself $n$ times
 
@@
 
-- **Logarithms** the inverse, defined as the exponents in the expressions above
+- **Logarithms** are the inverse, defined as the exponents in the expressions above
 $$\text{If } b^n=y\text{, then }log_b(y)=n$$
-$n$ is the number you must raise $b$ to in order to get $y$
+  - $n$ is the number you must raise $b$ to in order to get $y$
 
@@
 
---
+---
+
+# The Natural Logarithm: Review II
 
@@
 - Natural logs have a lot of useful properties:
-  - $ln(\frac{1}{x})=-ln(x)$
-  - $ln(ab)=ln(a)+ln(b)$
-  - $ln(\frac{x}{a})=ln(x)-ln(a)$
-  - $ln(x^a)=a \, ln(x)$
-  - $\frac{d  \, ln \, x}{d \, x} = \frac{1}{x}$
+
+1. $ln(\frac{1}{x})=-ln(x)$
+
+2. $ln(ab)=ln(a)+ln(b)$
+
+3.  $ln(\frac{x}{a})=ln(x)-ln(a)$
+
+4. $ln(x^a)=a \, ln(x)$
+
+5.  $\frac{d  \, ln \, x}{d \, x} = \frac{1}{x}$
 
@@
 
-- An .shout[elasticity] between two variables, $E_{Y,X}$ describes the .whisper[responsiveness] of one variable to a change in another.
+- An .shout[elasticity] between two variables, $E_{Y,X}$ describes the .onfire[responsiveness] of one variable to a change in another
 
@@
 
-$$\begin{align*}
-E_{y,x}&=\frac{\% \Delta y}{\% \Delta x}\\
-&=\cfrac{\frac{\Delta y}{y}}{\frac{\Delta x}{x}}\\
-\end{align*}$$
+$$E_{Y,X}=\frac{\% \Delta Y}{\% \Delta X} =\cfrac{\left(\frac{\Delta Y}{Y}\right)}{\left( \frac{\Delta X}{X}\right)}$$
 
@@
 
-```{r}
+```{r, echo=T}
 gapminder<-gapminder %>%
-  mutate(loggdp=log(gdpPercap))
-
-gapminder
+  mutate(loggdp=log(gdpPercap)) # log GDP per capita
+
+gapminder %>% head() # look at it
 ```
@@
 
-- .shout[Linear-log model] has an independent variable ($X$) that is logged
+- .shout[Linear-log model] has an independent variable $(X)$ that is logged
 
@@
 
-- $\hat{\beta_1}$: **a 1\% change in $X$** $\rightarrow$ **$\frac{\beta_1}{100}$ unit change in $Y$**
+- .onfire[**Marginal effect of** `\\(\mathbf{X \rightarrow Y}\\)`: a **1%** change in `\\(X \rightarrow\\)` a `\\(\frac{\beta_1}{100}\\)` **unit** change in `\\(Y\\)`]
 
@@
 
-- $\hat{\beta_1}$: a **1% change in GDP** $\rightarrow$ a $\frac{9.41}{100}=$ **0.0941 year increase** in Life Expectancy
+- A **1% change in GDP** $\rightarrow$ a $\frac{9.41}{100}=$ **0.0941 year increase** in Life Expectancy
 
@@
 
-- $\hat{\beta_1}$: a **1% change in GDP** $\rightarrow$ a $\frac{9.41}{100}=$ **0.0941 year increase** in Life Expectancy
-
-- A **25% fall in GDP** $\rightarrow$ a $(25 \times 0.0941)=$ **2.353 year decrease** in Life Expectancy
+- A **1% change in GDP** $\rightarrow$ a $\frac{9.41}{100}=$ **0.0941 year increase** in Life Expectancy
+
+- A **25% fall in GDP** $\rightarrow$ a $(-25 \times 0.0941)=$ **2.353 year decrease** in Life Expectancy
 
@@
 $$\begin{align*}
-\mathbf{Y}&=\beta_0+\beta_1 X\\
-\beta_1&=\cfrac{\big(\frac{\Delta Y}{Y}\big)}{X}\\
+\mathbf{ln(Y)}&=\beta_0+\beta_1 X\\
+\beta_1&=\cfrac{\big(\frac{\Delta Y}{Y}\big)}{\Delta X}\\
 \end{align*}$$
@@
 
-- $\hat{\beta_1}$: **a 1 unit change in $X$** $\rightarrow$ **$\beta_1 \times 100\%$ change in $Y$**
+- .onfire[**Marginal effect of** `\\(\mathbf{X \rightarrow Y}\\)`: a **1 unit** change in `\\(X \rightarrow\\)` a `\\(\beta_1 \times 100\\)` **%** change in `\\(Y\\)`]
 
@@
 
-```{r}
+```{r, echo=T}
 gapminder <- gapminder %>%
-  mutate(gdp_t=gdpPercap/1000,
-         loglife=log(lifeExp))
+  mutate(gdp_t=gdpPercap/1000, # first make GDP/capita in $1000s
+         loglife=log(lifeExp)) # take the log of LifeExp
+gapminder %>% head() # look at it
 ```
@@
 
-- $\hat{\beta_1}$: a **\$1 (thousand) change in (thousands of) GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3\% increase** in Life Expectancy
+- A **$1 (thousand) change in GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3% increase** in Life Expectancy
 
@@
 
-- $\hat{\beta_1}$: a **\$1 (thousand) change in (thousands of) GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3\% increase** in Life Expectancy
-
-- A **\$25K fall in GDP** $\rightarrow$ a $(25 \times 1.3\%)=$ **32.5\% decrease** in Life Expectancy
+- A **$1 (thousand) change in GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3% increase** in Life Expectancy
+
+- A **$25 (thousand) fall in GDP** $\rightarrow$ a $(-25 \times 1.3\%)=$ **32.5% decrease** in Life Expectancy
 
@@
 
-- $\hat{\beta_1}$: a **\$1 (thousand) change in (thousands of) GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3\% increase** in Life Expectancy
-
-- A **\$25K fall in GDP** $\rightarrow$ a $(25 \times 1.3\%)=$ **32.5\% decrease** in Life Expectancy
-
-- A **\$100K rise in GDP** $\rightarrow$ a $(100 \times 1.3\%)=$ **130\% increase** in Life Expectancy
+- A **$1 (thousand) change in GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3% increase** in Life Expectancy
+
+- A **$25K fall in GDP** $\rightarrow$ a $(-25 \times 1.3\%)=$ **32.5% decrease** in Life Expectancy
+
+- A **$100K rise in GDP** $\rightarrow$ a $(100 \times 1.3\%)=$ **130% increase** in Life Expectancy
 
@@
 $$\begin{align*}
-\mathbf{Y}&=\beta_0+\beta_1 \mathbf{X}\\
+\mathbf{ln(Y)}&=\beta_0+\beta_1 \mathbf{ln(X)}\\
 \beta_1&=\cfrac{\big(\frac{\Delta Y}{Y}\big)}{\big(\frac{\Delta X}{X}\big)}\\
@@
 
-- $\hat{\beta_1}$: **a 1% change in $X$** $\rightarrow$ **$\beta_1\%$ change in $Y$**
+- .onfire[**Marginal effect of** `\\(\mathbf{X \rightarrow Y}\\)`: a **1%** change in `\\(X \rightarrow\\)` a `\\(\beta_1\\)` **%** change in `\\(Y\\)`]
+
+- $\beta_1$ is the **elasticity** of $Y$ with respect to $X$!
 
@@
 
-$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.0147 \, \text{ln(GDP)}_i$$
+$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.147 \, \text{ln(GDP)}_i$$
 
@@
 
-$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.0147 \, \text{ln(GDP)}_i$$
-
-- $\hat{\beta_1}$: a **$1\% change in GDP** $\rightarrow$ a **0.147\% increase** in Life Expectancy
+$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.147 \, \text{ln(GDP)}_i$$
+
+- A **$1% change in GDP** $\rightarrow$ a **0.147% increase** in Life Expectancy
 
@@
 
-$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.0147 \, \text{ln(GDP)}_i$$
-
-- $\hat{\beta_1}$: a **$1\% change in GDP** $\rightarrow$ a **0.147\% increase** in Life Expectancy
-
-- A **25\% fall in GDP** $\rightarrow$ a $(25 \times 0.147\%)=$ **3.675\% decrease** in Life Expectancy
+$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.147 \, \text{ln(GDP)}_i$$
+
+- A **$1% change in GDP** $\rightarrow$ a **0.147% increase** in Life Expectancy
+
+- A **25% fall in GDP** $\rightarrow$ a $(-25 \times 0.147\%)=$ **3.675% decrease** in Life Expectancy
 
@@
 
-$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.0147 \, \text{ln(GDP)}_i$$
-
-- $\hat{\beta_1}$: a **$1\% change in GDP** $\rightarrow$ a **0.147\% increase** in Life Expectancy
-
-- A **25\% fall in GDP** $\rightarrow$ a $(25 \times 0.147\%)=$ **3.675\% decrease** in Life Expectancy
-
-- A **100\% rise in GDP** $\rightarrow$ a $(100 \times 0.147\%)=$ **14.7\% increase** in Life Expectancy
+$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.147 \, \text{ln(GDP)}_i$$
+
+- A **$1% change in GDP** $\rightarrow$ a **0.147% increase** in Life Expectancy
+
+- A **25% fall in GDP** $\rightarrow$ a $(-25 \times 0.147\%)=$ **3.675% decrease** in Life Expectancy
+
+- A **100% rise in GDP** $\rightarrow$ a $(100 \times 0.147\%)=$ **14.7% increase** in Life Expectancy
 
@@
 | Salary | $2,024,616 | $2,764,512 |
-| Batting Average | 0.267 | $0.031 |
+| Batting Average | 0.267 | 0.031 |
 | Home Runs | 12.11 | 10.31 |
@@
 
-- **Marginal effect** on $Y$ (in *standard deviations* of $Y$) from 1 *standard deviation* change in $X$
-  - e.g. for 1 standard deviation increase in Batting Average, Salary increases by 0.14 standard deviations
-  - $0.14\times 2,764,512=\$387,032$
+- .onfire[Marginal effect] on $Y$ (in *standard deviations* of $Y$) from 1 *standard deviation* change in $X$
+  - $\hat{\beta_1}$: a 1 standard deviation increase in Batting Average increases Salary by 0.14 standard deviations
+    - $0.14 \times 2,764,512=\$387,032$
+  - $\hat{\beta_2}$: a 1 standard deviation increase in Home Runs increases Salary by 0.48 standard deviations
+    - $0.48 \times 2,764,512=\$1,326,966$
 
@@
 
-```{r}
+```{r, echo=T}
 gapminder<-gapminder %>%
@@
 
-- The .shout[F-statistic} is the test-statistic used to test joint hypotheses about regression coefficients with an .shout[F-test}
+- The .shout[F-statistic] is the test-statistic used to test joint hypotheses about regression coefficients with an .shout[F-test]
 
@@
 
-- $F$ is an .whisper[analysis of variance (ANOVA)]
+- $F$ is an .onfire[analysis of variance (ANOVA)]
   - essentially tests whether $R^2$ increases statistically significantly as we go from the restricted model$\rightarrow$unrestricted model
@@
 
-- $R^_r$: the $R^2$ from the **restricted model** (null hypothesis)
+- $R^2_r$: the $R^2$ from the **restricted model** (null hypothesis)
 ]
@@
 
-- $R^_r$: the $R^2$ from the **restricted model** (null hypothesis)
+- $R^2_r$: the $R^2$ from the **restricted model** (null hypothesis)
 
@@
 
-- .whisper[Key takeaway]: The bigger the difference between $(R^2_u-R^2_r)$, the greater the improvement in fit by adding variables, the larger the $F$!
+- .onfire[Key takeaway]: The bigger the difference between $(R^2_u-R^2_r)$, the greater the improvement in fit by adding variables, the larger the $F$!
 
@@
 .font50[
-```{r, echo=F, results="asis"}
+```{r, echo=F}
 summary(unrestricted_reg)
@@
 
-- "Statistic" is the All F-test, "p-value" next to it is the p value from the F test!
+- "**statistic**" is the All F-test, "**p.value**" next to it is the p value from the F test!

### 2019-11-22 static/slides/21-slides.Rmd
Minor formatting tweaks to slides 21
Add comparing models table to slides 21

@@
 
-- A **$25K fall in GDP** $\rightarrow$ a $(-25 \times 1.3\%)=$ **32.5% decrease** in Life Expectancy
-
-- A **$100K rise in GDP** $\rightarrow$ a $(100 \times 1.3\%)=$ **130% increase** in Life Expectancy
+- A **$25 (thousand) fall in GDP** $\rightarrow$ a $(-25 \times 1.3\%)=$ **32.5% decrease** in Life Expectancy
+
+- A **$100 (thousand) rise in GDP** $\rightarrow$ a $(100 \times 1.3\%)=$ **130% increase** in Life Expectancy
 
@@
 
-- A **$1% change in GDP** $\rightarrow$ a **0.147% increase** in Life Expectancy
+- A **1% change in GDP** $\rightarrow$ a **0.147% increase** in Life Expectancy
 
@@
 |-------|----------|----------------|
-| Linear-Log | $Y=\beta_0+\beta_1 \mathbf{ln(X)}$ | 1% change in $X \rightarrow \frac{\hat{\beta_1}}{100}$ **unit** change in $Y$ |
-| Log-Linear | $\mathbf{ln(Y)}=\beta_0+\beta_1X$ | 1 **unit** change in $X \rightarrow \hat{\beta_1}\times 100$**%** change in $Y$ |
-| Log-Log | $\mathbf{ln(Y)}=\beta_0+\beta_1\mathbf{ln(X)}$ | 1**%** change in $X \rightarrow \hat{\beta_1}$**%** change in $Y$ |
-
-- Hint: the variable that gets logged changes in **percent** terms, the variable not logged changes in **unit** terms
+| Linear-**Log** | $Y=\beta_0+\beta_1 \mathbf{ln(X)}$ | 1% change in $X \rightarrow \frac{\hat{\beta_1}}{100}$ **unit** change in $Y$ |
+| **Log**-Linear | $\mathbf{ln(Y)}=\beta_0+\beta_1X$ | 1 **unit** change in $X \rightarrow \hat{\beta_1}\times 100$**%** change in $Y$ |
+| **Log**-**Log** | $\mathbf{ln(Y)}=\beta_0+\beta_1\mathbf{ln(X)}$ | 1**%** change in $X \rightarrow \hat{\beta_1}$**%** change in $Y$ |
+
+- Hint: the variable that gets logged changes in **percent** terms, the variable not logged changes in *unit* terms
 
@@
 
+# Comparing Models III
+
+| Linear-Log | Log-Linear | Log-Log |
+|:----------:|:----------:|:-------:|
+| ![](21-slides_files/figure-html/unnamed-chunk-14-1.png) | ![](21-slides_files/figure-html/unnamed-chunk-20-1.png) | ![](21-slides_files/figure-html/unnamed-chunk-25-1.png) |
+| $\hat{Y_i}=\hat{\beta_0}+\hat{\beta_1}\mathbf{ln(X_i)}$ | $\mathbf{ln(\hat{Y_i})}=\hat{\beta_0}+\hat{\beta_1}X_i$ | $\mathbf{ln(\hat{Y_i})}=\hat{\beta_0}+\hat{\beta_1}\mathbf{ln(X_i)}$ |
+| $R^2=0.65$ | $R^2=0.30$ | $R^2=0.61$ |
+
+---
+
 # When to Log?
 
-- In practice, the following types of variables are logged (ln):
+- In practice, the following types of variables are logged:
   - Variables that must always be positive (prices, sales, market values)
   - Very large numbers (population, GDP)
-  - Variables expressed as percentages or percentage changes (inflation, population growth, GDP growth, labor force participation rate, unemployment rate)
+  - Variables we want to talk about as percentage changes or growth rates (money supply, population, GDP)
+  - Variables that have diminishing returns (output, utility)
   - Variables that have nonlinear scatterplots
@@
 
-- Never use logs for:
-  - Variables that are less than one, decimals, 0, or negative (temperature)
-  - Categorical variables
+- Avoid logs for:
+  - Variables that are less than one, decimals, 0, or negative
+  - Categorical variables (season, gender, political party)
   - Time variables (year, week, day)

### 2019-11-22 static/slides/21-slides.Rmd
Add more F-tests to slides 21

@@
 
-- $F$ has two sets of degrees of freedom, $q$ for the numerator, $n-k-1$ for the denominator
+- $F$ has two sets of degrees of freedom:
+  - $q$ for the numerator
+  - $(n-k-1)$ for the denominator
 ]
@@
 
-- Command is `linearHypothesis()`, comes from the `car` package
+- **Unrestricted model**:
+
+$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Female_i+\hat{\beta_2}Northeast_i+\hat{\beta_3}Northcen+\hat{\beta_4}South_i$$
+
+- **Restricted model**:
+
+$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Female_i$$
+
+- $H_0: \beta_2 = \beta_3 = \beta_4 =0$
+
+- $q = 3$ restrictions (F numerator df)
+
+- $n-k-1 = 526-4-1=521$ (F denominator df)
+
+---
+
+# F-test Example III
+
+- We can use the `car` package's `linearHypothesis()` command to run an $F$-test:
+  - first argument: name of the (unrestricted) regression
+  - second argument: vector of variable names (in quotes) you are testing
 
@@
 
-# test that northcen=west=south=0
-# # the first argument is the name of my regression ("unrestricted_reg")
+# F-test
 linearHypothesis(unrestricted_reg, c("northcen", "west", "south")) 
@@
 
+# Second F-test Example: Are Two Coefficients Equal?
+
+- The second type of test is whether two coefficients equal one another
+
+.content-box-green[
+.green[**Example**]:
+
+$$\widehat{wage_i}=\beta_0+\beta_1 \text{Adolescent height}_i + \beta_2 \text{Adult height}_i + \beta_3 \text{Male}_i$$
+
+]
+
+--
+
+- Does height as an adolescent have the same effect on wages as height as an adult?
+
+$$H_0: \beta_1=\beta_2$$
+
+--
+
+- What is the restricted regression?
+
+$$\widehat{wage_i}=\beta_0+\beta_1(\text{Adolescent height}_i + \text{Adult height}_i )+ \beta_3 \text{Male}_i$$
+
+- $q=1$ restriction
+
+---
+
+# Second F-test Example: Data
+
+```{r, echo=T}
+# load in data
+heightwages<-read_csv("../data/heightwages.csv")
+
+# make a "heights" variable as the sum of adolescent (height81) and adult (height85) height
+
+heightwages <- heightwages %>%
+  mutate(heights=height81+height85)
+
+height_reg<-lm(wage96~height81+height85+male, data=heightwages)
+height_restricted_reg<-lm(wage96~heights+male, data=heightwages)
+
+```
+
+---
+
+# Second F-test Example: Data
+
+- For second argument, set two variables equal, in quotes
+
+```{r, echo=T}
+linearHypothesis(height_reg, "height81=height85") # F-test 
+```
+
+- Insufficient evidence to reject $H_0$!
+
+- The effect of adolescent and adult height on wages is the same
+
+---
+
 # All F-test I
 
-.font50[
-```{r, echo=F}
+.pull-left[
+
+```{css, echo = FALSE}
+.smallcode .remark-code { /*Change made here*/
+  font-size: 65% !important;
+}
+```
+
+.smallcode[
+```{r, echo=T}
 summary(unrestricted_reg)
@@
 ]
+]
+
+.pull-right[
+
+- Last line of regression output from `summary()` of an `lm()` object is an **All F-test**
+  - $H_0:$ all $\beta's=0$ 
+    - the regression explains no variation in $Y$
+  - Calculates an `F-statistic` that, if high enough, is significant (`p-value` $<0.05)$ enough to reject $H_0$
+]
 
@@
 # All F-test II
+
+- Alternatively, if you use `broom` instead of `summary()` of `lm` object:
+  - `glance()` command makes table of regression summary statistics
+  - `tidy()` only shows coefficients
 
@@
 
-- "**statistic**" is the All F-test, "**p.value**" next to it is the p value from the F test!
+- "**statistic**" is the All F-test, "**p.value**" next to it is the p value from the F test

### 2019-11-24 content/class/21-class.Rmd
Add tables and class page for class 21

--- 2019-11-22 static/slides/21-slides.Rmd
@@
 ---
-title: "3.9: Logarithmic Regression"
-subtitle: "ECON 480 · Econometrics · Fall 2019"
-author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
-#date:
+title: "3.9: Logarithmic Regression - Class Notes"
+citeproc: false
+bibliography: ../../static/bib/references.bib
+csl: ../../static/bib/chicago-syllabus-no-bib.csl
 output:
-  xaringan::moon_reader:
-    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
-    lib_dir: libs
-    df_print: paged
-    #seal: false
-    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
-    nature:
-      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
-      highlightStyle: github
-      highlightLines: true
-      countIncrementalSlides: false
-      ratio: "16:9"
-      slideNumberFormat: |
-        <div class="progress-bar-container">
-          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
-          </div>
-        </div>
-    includes:
-      in_header: header.html # for font awesome, used in title  
+  blogdown::html_page:
+    template: ../../pandoc/toc-title_html.template
+    toc: true
+editor_options: 
+  chunk_output_type: console
 ---
 
-```{r setup, include=FALSE}
-options(htmltools.dir.version = FALSE)
-knitr::opts_chunk$set(echo=F,
-                      message=F,
-                      warning=F)
-library(tidyverse)
-set.seed(256)
-update_geom_defaults("label", list(family = "Fira Sans Condensed"))
-```
+*Thursday, November 21, 2019*
 
-# Nonlinearities
+# Overview
 
-.pull-left[
+*Today*, we finish up our view of *nonlinear* models with **logarithmic models**, which are more frequently used.
+We also discuss a few other tests and transformations to wrap up multivariate regression before we turn to panel data: **standardizing variables** to compare effect sizes, and **joint hypothesis tests**.
 
-- Consider the `gapminder` example again
-]
-
-.pull-right[
-
-```{r, fig.retina=3}
-library(gapminder)
-gap_60<-gapminder %>%
-  filter(gdpPercap<60000)
-
-p2<-ggplot(data = gap_60)+
-  aes(x = gdpPercap,
-      y = lifeExp)+
-  geom_point(color="blue", alpha=0.5)+
-  scale_x_continuous(labels=scales::dollar,
-                     breaks=seq(0,60000,20000))+
-  scale_y_continuous(breaks=seq(0,90,10),
-                     limits=c(0,90))+
-  labs(x = "GDP per Capita",
-       y = "Life Expectancy (Years)")+
-  theme_classic(base_family = "Fira Sans Condensed",
-           base_size=20)
-
-p2
-```
-
-]
-
----
-
-# Nonlinearities
-
-.pull-left[
-
-- Consider the `gapminder` example again
-  - .red[linear model]
-]
-
-.pull-right[
-
-```{r, fig.retina=3}
-p2+geom_smooth(method="lm", color="red")
-```
-
-]
-
----
-
-# Nonlinearities
-
-.pull-left[
-
-- Consider the `gapminder` example again
-  - .red[linear model]
-  - .green[polynomial model (quadratic)]
-]
-
-.pull-right[
-
-```{r, fig.retina=3}
-p2+geom_smooth(method="lm", color="red")+
-  stat_smooth(method="lm", formula=y~x+I(x^2), color = "green")
-```
-
-]
-
----
-
-# Nonlinearities
-
-.pull-left[
-
-- Consider the `gapminder` example again
-  - .red[linear model]
-  - .green[polynomial model (quadratic)]
-  - .orange[logarithmic model]
-]
-
-.pull-right[
-
-```{r, fig.retina=3}
-p2+geom_smooth(method="lm", color="red")+
-  stat_smooth(method="lm", formula=y~x+I(x^2), color = "green")+
-  stat_smooth(method = "lm", formula=y~log(x), color="orange", size=2)
-```
-
-]
-
----
-
-class: inverse, center, middle
-
-# Logarithmic Models
-
----
-
-# Logarithmic Models
-
-.pull-left[
-```{r, fig.retina=3}
-ggplot(data.frame(x=0:10),aes(x=x))+
-    stat_function(fun=log, color="blue",size =2)+
-  geom_label(x=9,y=log(9), label="y=log(x)", color="blue", size=5)+
-  scale_x_continuous(breaks=seq(0,10,1),
-                     limits=c(0,10),
-                     expand=c(0,0))+
-  scale_y_continuous(limits=c(0,2.5),
-                     expand=c(0,0))+
-  labs(x = "X",
-       y = "Y")+
-  theme_classic(base_family = "Fira Sans Condensed",
-           base_size=20)
-```
-]
-
-.pull-right[
-
-- Another model specification for nonlinear data is the .shout[logarithmic model]<sup>.red[1]</sup>
-  - We transform either $X$, $Y$, or *both* by taking the (natural) logarithm
-
-- Logarithmic model has two additional advantages
-  1. We can easily interpret coefficients as **percentage changes** or **elasticities**
-  2. Useful economic shape: diminishing returns (production functions, utility functions, etc)
-
-.footnote[<sup>.red[1]</sup> Note, this should not be confused with a **logistic model**, which is a model for dependent dummy variables.]
-
-]
-
----
-
-# The Natural Logarithm
-
-.pull-left[
-```{r, fig.retina=3}
-ggplot(data.frame(x=0:10),aes(x=x))+
-    stat_function(fun=log, color="blue",size =2)+
-    stat_function(fun=exp, color="red", size =2)+
-  stat_function(fun=function(x){x}, size=2, linetype="dashed")+
-  geom_label(x=4,y=log(4), label=expression(Y==ln(X)), color="blue", size=5)+
-  geom_label(x=1,y=exp(1), label=expression(Y==e^{X}), color="red", size=5)+
-  scale_x_continuous(breaks=seq(0,5,1),
-                     limits=c(0,5),
-                     expand=c(0,0))+
-  scale_y_continuous(limits=c(0,5),
-                     expand=c(0,0))+
-  labs(x = "X",
-       y = "Y")+
-  theme_classic(base_family = "Fira Sans Condensed",
-           base_size=20)
-```
-]
-
-.pull-right[
-
-- The .red[exponential function], $Y=e^X$ or $Y=exp(X)$, where base $e=2.71828...$
-
-- .blue[Natural logarithm] is the inverse, $Y=ln(X)$
-
-]
-
----
-
-# The Natural Logarithm: Review I
-
-- **Exponents** are defined as
-$$b^n=\underbrace{b \times b \times \cdots \times b}_{n \text{ times}}$$
-  - where base $b$ is multiplied by itself $n$ times
-
---
-- .green[**Example**]: $2^3=\underbrace{2 \times 2 \times 2}_{n=3}=8$
-
---
-
-- **Logarithms** are the inverse, defined as the exponents in the expressions above
-$$\text{If } b^n=y\text{, then }log_b(y)=n$$
-  - $n$ is the number you must raise $b$ to in order to get $y$
-
---
-
-- .green[**Example**]: $log_2(8)=3$
-
----
-
-# The Natural Logarithm: Review II
-
-- Logarithms can have any base, but common to use the **natural logarithm $(ln)$** with base $\mathbf{e=2.71828...}$
-$$\text{If } e^n=y\text{, then }ln(y)=n$$
-
----
-
-# The Natural Logarithm: Properties
-
-- Natural logs have a lot of useful properties:
-
-1. $ln(\frac{1}{x})=-ln(x)$
-
-2. $ln(ab)=ln(a)+ln(b)$
-
-3.  $ln(\frac{x}{a})=ln(x)-ln(a)$
-
-4. $ln(x^a)=a \, ln(x)$
-
-5.  $\frac{d  \, ln \, x}{d \, x} = \frac{1}{x}$
-
----
-
-# The Natural Logarithm: Example
-
-- Most useful property: for small change in $x$, $\Delta x$:
-
-$$\underbrace{ln(x+\Delta x) - ln(x)}_{\text{Difference in logs}} \approx \underbrace{\frac{\Delta x}{x}}_{\text{Relative change}}$$
-
---
-
-.content-box-green[
-.green[**Example**]: Let $x=100$ and $\Delta x =1$, relative change is:
-
-$$\frac{\Delta x}{x} = \frac{(101-100)}{100} = 0.01 \text{ or }1\%$$
-
-- The logged difference:
-$$ln(101)-ln(100) = 0.00995 \approx 1\%$$
-]
-
---
-
-- This allows us to very easily interpret coefficients as *percent changes* or **elasticities**
-
----
-
-# Elasticity
-
-- An .shout[elasticity] between two variables, $E_{Y,X}$ describes the .onfire[responsiveness] of one variable to a change in another
-
-- Measured in percentages: a 1% change in $X$ will cause a $E\%$ change in $Y$
-
---
-
-$$E_{Y,X}=\frac{\% \Delta Y}{\% \Delta X} =\cfrac{\left(\frac{\Delta Y}{Y}\right)}{\left( \frac{\Delta X}{X}\right)}$$
-
---
-
-- Numerator is relative change in $Y$, Denominator is relative change in $X$
-
----
-
-# Math FYI: Cobb Douglas Functions and Logs
-
-- One of the (many) reasons why economists love Cobb-Douglas functions:
-$$Y=AL^{\alpha}K^{\beta}$$
-
---
-
-- Taking logs, relationship becomes linear:
-
---
-
-$$ln(Y)=ln(A)+\alpha ln(L)+ \beta ln(K)$$
-
---
-
-- With data on $(Y, L, K)$ and linear regression, can estimate $\alpha$ and $\beta$
-  - $\alpha$: elasticity of $Y$ with respect to $L$
-      - A 1% change in $L$ will lead to an $\alpha$% change in $Y$
-  - $\beta$: elasticity of $Y$ with respect to $K$
-      - A 1% change in $K$ will lead to a $\beta$% change in $Y$
-
----
-
-# Logarithms in R I
-
-- The `log()` function can easily take the logarithm
-
-```{r, echo=T}
-gapminder<-gapminder %>%
-  mutate(loggdp=log(gdpPercap)) # log GDP per capita
-
-gapminder %>% head() # look at it
-```
-
----
-
-# Logarithms in R II
-
-- Note, `log()` by default is the **natural logarithm $ln()$**, i.e. base `e`
-  - Can change base with e.g. `log(x, base = 5)`
-  - Some common built-in logs: `log10`, `log2`
-
-```{r, echo=T}
-log10(100)
-log2(16)
-log(19683, base=3)
-```
-
----
-
-# Types of Logarithmic Models
-
-- Three types of log regression models, depending on which variables we log
-
---
-
-1.  **Linear-log model:** $Y_i=\beta_0+\beta_1$ .red[`\\(ln(X_i)\\)`]
-
---
-
-2. **Log-linear model:** .red[`\\(ln(Y_i)\\)`] $=\beta_0+\beta_1X_i$
-
---
-
-3. **Log-log model:** .red[`\\(ln(Y_i)\\)`] $=\beta_0+\beta_1$ .red[`\\(ln(X_i)\\)`]
-
----
-
-# Linear-Log Model
-
-- .shout[Linear-log model] has an independent variable $(X)$ that is logged
-
---
-
-$$\begin{align*}
-Y&=\beta_0+\beta_1 \mathbf{ln(X)}\\
-\beta_1&=\cfrac{\Delta Y}{\big(\frac{\Delta X}{X}\big)}\\
-\end{align*}$$
-
---
-
-- .onfire[**Marginal effect of** `\\(\mathbf{X \rightarrow Y}\\)`: a **1%** change in `\\(X \rightarrow\\)` a `\\(\frac{\beta_1}{100}\\)` **unit** change in `\\(Y\\)`]
-
----
-
-# Linear-Log Model in R
-
-.pull-left[
-```{r, echo=T}
-lin_log_reg<-lm(lifeExp~loggdp, data = gapminder)
-library(broom)
-tidy(lin_log_reg)
-```
-]
-
-.pull-right[
-
-$$\widehat{\text{Life Expectancy}}_i=-9.10+9.41 \, \text{ln(GDP)}_i$$
-
-]
-
----
-
-# Linear-Log Model in R
-
-.pull-left[
-```{r, echo=T}
-lin_log_reg<-lm(lifeExp~loggdp, data = gapminder)
-library(broom)
-tidy(lin_log_reg)
-```
-]
-
-.pull-right[
-
-$$\widehat{\text{Life Expectancy}}_i=-9.10+9.41 \, \text{ln(GDP)}_i$$
-
-- A **1% change in GDP** $\rightarrow$ a $\frac{9.41}{100}=$ **0.0941 year increase** in Life Expectancy
-
-]
-
----
-
-# Linear-Log Model in R
-
-.pull-left[
-```{r, echo=T}
-lin_log_reg<-lm(lifeExp~loggdp, data = gapminder)
-library(broom)
-tidy(lin_log_reg)
-```
-]
-
-.pull-right[
-
-$$\widehat{\text{Life Expectancy}}_i=-9.10+9.41 \, \text{ln(GDP)}_i$$
-
-- A **1% change in GDP** $\rightarrow$ a $\frac{9.41}{100}=$ **0.0941 year increase** in Life Expectancy
-
-- A **25% fall in GDP** $\rightarrow$ a $(-25 \times 0.0941)=$ **2.353 year decrease** in Life Expectancy
-
-]
-
----
-
-# Linear-Log Model in R
-
-.pull-left[
-```{r, echo=T}
-lin_log_reg<-lm(lifeExp~loggdp, data = gapminder)
-library(broom)
-tidy(lin_log_reg)
-```
-]
-
-.pull-right[
-
-$$\widehat{\text{Life Expectancy}}_i=-9.10+9.41 \, \text{ln(GDP)}_i$$
-
-- A **1% change in GDP** $\rightarrow$ a $\frac{9.41}{100}=$ **0.0941 year increase** in Life Expectancy
-
-- A **25% fall in GDP** $\rightarrow$ a $(-25 \times 0.0941)=$ **2.353 year decrease** in Life Expectancy
-
-- A **100% rise in GDP** $\rightarrow$ a $(100 \times 0.0941)=$ **9.041 year increase** in Life Expectancy
-
-]
-
----
-
-# Linear-Log Model Graph I
-
-.pull-left[
-```{r lin-log-graph, echo=T, eval=F}
-ggplot(data = gapminder)+
-  aes(x = gdpPercap,
-      y = lifeExp)+
-  geom_point(color="blue", alpha=0.5)+
-  geom_smooth(method="lm", #<<
-              formula=y~log(x), #<<
-              color="orange")+ #<<
-  scale_x_continuous(labels=scales::dollar,
-                     breaks=seq(0,120000,20000))+
-  scale_y_continuous(breaks=seq(0,90,10),
-                     limits=c(0,90))+
-  labs(x = "GDP per Capita",
-       y = "Life Expectancy (Years)")+
-  theme_classic(base_family = "Fira Sans Condensed",
-           base_size=20)
-```
-]
-
-.pull-right[
-```{r, ref.label="lin-log-graph", fig.retina=3}
-
-```
-]
-
----
-
-# Linear-Log Model Graph II
-
-.pull-left[
-```{r lin-log-graph2, echo=T, eval=F}
-ggplot(data = gapminder)+
-  aes(x = loggdp, #<<
-      y = lifeExp)+ 
-  geom_point(color="blue", alpha=0.5)+
-  geom_smooth(method="lm", color="orange")+ #<<
-  scale_y_continuous(breaks=seq(0,90,10),
-                     limits=c(0,90))+
-  labs(x = "Log GDP per Capita",
-       y = "Life Expectancy (Years)")+
-  theme_classic(base_family = "Fira Sans Condensed",
-           base_size=20)
-```
-]
-
-.pull-right[
-```{r, ref.label="lin-log-graph2", fig.retina=3}
-
-```
-]
-
----
-
-# Log-Linear Model
-
-- .shout[Log-linear model] has the dependent variable $(Y)$ logged
-
---
-
-$$\begin{align*}
-\mathbf{ln(Y)}&=\beta_0+\beta_1 X\\
-\beta_1&=\cfrac{\big(\frac{\Delta Y}{Y}\big)}{\Delta X}\\
-\end{align*}$$
-
---
-
-- .onfire[**Marginal effect of** `\\(\mathbf{X \rightarrow Y}\\)`: a **1 unit** change in `\\(X \rightarrow\\)` a `\\(\beta_1 \times 100\\)` **%** change in `\\(Y\\)`]
-
----
-
-# Log-Linear Model in R (Preliminaries)
-
-- We will again have very large/small coefficients if we deal with GDP directly, again let's transform `gdpPercap` into $1,000s, call it `gdp_t`
-
-- Then log LifeExp
-
---
-
-```{r, echo=T}
-gapminder <- gapminder %>%
-  mutate(gdp_t=gdpPercap/1000, # first make GDP/capita in $1000s
-         loglife=log(lifeExp)) # take the log of LifeExp
-gapminder %>% head() # look at it
-```
-
----
-
-# Log-Linear Model in R
-
-.pull-left[
-```{r, echo=T}
-log_lin_reg<-lm(loglife~gdp_t, data = gapminder)
-tidy(log_lin_reg)
-```
-]
-
-.pull-right[
-
-$$\widehat{ln(\text{Life Expectancy})}_i=3.967+0.013 \, \text{GDP}_i$$
-
-]
-
----
-
-# Log-Linear Model in R
-
-.pull-left[
-```{r, echo=T}
-log_lin_reg<-lm(loglife~gdp_t, data = gapminder)
-tidy(log_lin_reg)
-```
-]
-
-.pull-right[
-
-$$\widehat{ln(\text{Life Expectancy})}_i=3.967+0.013 \, \text{GDP}_i$$
-
-- A **$1 (thousand) change in GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3% increase** in Life Expectancy
-
-]
-
----
-
-# Log-Linear Model in R
-
-.pull-left[
-```{r, echo=T}
-log_lin_reg<-lm(loglife~gdp_t, data = gapminder)
-tidy(log_lin_reg)
-```
-]
-
-.pull-right[
-
-$$\widehat{ln(\text{Life Expectancy})}_i=3.967+0.013 \, \text{GDP}_i$$
-
-- A **$1 (thousand) change in GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3% increase** in Life Expectancy
-
-- A **$25 (thousand) fall in GDP** $\rightarrow$ a $(-25 \times 1.3\%)=$ **32.5% decrease** in Life Expectancy
-
-]
-
----
-
-# Log-Linear Model in R
-
-.pull-left[
-```{r, echo=T}
-log_lin_reg<-lm(loglife~gdp_t, data = gapminder)
-tidy(log_lin_reg)
-```
-]
-
-.pull-right[
-
-$$\widehat{ln(\text{Life Expectancy})}_i=3.967+0.013 \, \text{GDP}_i$$
-
-- A **$1 (thousand) change in GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3% increase** in Life Expectancy
-
-- A **$25 (thousand) fall in GDP** $\rightarrow$ a $(-25 \times 1.3\%)=$ **32.5% decrease** in Life Expectancy
-
-- A **$100 (thousand) rise in GDP** $\rightarrow$ a $(100 \times 1.3\%)=$ **130% increase** in Life Expectancy
-
-]
-
----
-
-# Linear-Log Model Graph I
-
-.pull-left[
-```{r log-lin-graph, echo=T, eval=F}
-ggplot(data = gapminder)+
-  aes(x = gdp_t, #<<
-      y = loglife)+ #<<
-  geom_point(color="blue", alpha=0.5)+
-  geom_smooth(method="lm", color="orange")+ #<<
-  scale_x_continuous(labels=scales::dollar,
-                     breaks=seq(0,120,20))+
-  labs(x = "GDP per Capita ($ Thousands)",
-       y = "Log Life Expectancy")+
-  theme_classic(base_family = "Fira Sans Condensed",
-           base_size=20)
-```
-]
-
-.pull-right[
-```{r, ref.label="log-lin-graph", fig.retina=3}
-
-```
-]
-
----
-
-# Log-Log Model
-
-- .shout[Log-log model] has both variables $(X \text{ and } Y)$ logged
-
---
-
-$$\begin{align*}
-\mathbf{ln(Y)}&=\beta_0+\beta_1 \mathbf{ln(X)}\\
-\beta_1&=\cfrac{\big(\frac{\Delta Y}{Y}\big)}{\big(\frac{\Delta X}{X}\big)}\\
-\end{align*}$$
-
---
-
-- .onfire[**Marginal effect of** `\\(\mathbf{X \rightarrow Y}\\)`: a **1%** change in `\\(X \rightarrow\\)` a `\\(\beta_1\\)` **%** change in `\\(Y\\)`]
-
-- $\beta_1$ is the **elasticity** of $Y$ with respect to $X$!
-
----
-
-# Log-Log Model in R
-
-.pull-left[
-```{r, echo=T}
-log_log_reg<-lm(loglife~loggdp, data = gapminder)
-tidy(log_log_reg)
-```
-]
-
-.pull-right[
-
-$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.147 \, \text{ln(GDP)}_i$$
-
-]
-
----
-
-# Log-Log Model in R
-
-.pull-left[
-```{r, echo=T}
-log_log_reg<-lm(loglife~loggdp, data = gapminder)
-tidy(log_log_reg)
-```
-]
-
-.pull-right[
-
-$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.147 \, \text{ln(GDP)}_i$$
-
-- A **1% change in GDP** $\rightarrow$ a **0.147% increase** in Life Expectancy
-
-]
-
----
-
-# Log-Log Model in R
-
-.pull-left[
-```{r, echo=T}
-log_log_reg<-lm(loglife~loggdp, data = gapminder)
-tidy(log_log_reg)
-```
-]
-
-.pull-right[
-
-$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.147 \, \text{ln(GDP)}_i$$
-
-- A **1% change in GDP** $\rightarrow$ a **0.147% increase** in Life Expectancy
-
-- A **25% fall in GDP** $\rightarrow$ a $(-25 \times 0.147\%)=$ **3.675% decrease** in Life Expectancy
-
-]
-
----
-
-# Log-Log Model in R
-
-.pull-left[
-```{r, echo=T}
-log_log_reg<-lm(loglife~loggdp, data = gapminder)
-tidy(log_log_reg)
-```
-]
-
-.pull-right[
-
-$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.147 \, \text{ln(GDP)}_i$$
-
-- A **1% change in GDP** $\rightarrow$ a **0.147% increase** in Life Expectancy
-
-- A **25% fall in GDP** $\rightarrow$ a $(-25 \times 0.147\%)=$ **3.675% decrease** in Life Expectancy
-
-- A **100% rise in GDP** $\rightarrow$ a $(100 \times 0.147\%)=$ **14.7% increase** in Life Expectancy
-
-]
-
----
-
-# Log-Log Model Graph I
-
-.pull-left[
-```{r log-log-graph, echo=T, eval=F}
-ggplot(data = gapminder)+
-  aes(x = loggdp, #<<
-      y = loglife)+ #<<
-  geom_point(color="blue", alpha=0.5)+
-  geom_smooth(method="lm", color="orange")+ #<<
-  labs(x = "Log GDP per Capita",
-       y = "Log Life Expectancy")+
-  theme_classic(base_family = "Fira Sans Condensed",
-           base_size=20)
-```
-]
-
-.pull-right[
-```{r, ref.label="log-log-graph", fig.retina=3}
-
-```
-]
-
----
-
-# Comparing Models I
+Interpretting logged variables can often be difficult to remember, so here I reproduce the tables that describe the interpretations of the marginal effect of $X \rightarrow Y$, as well as some visual examples from the slides:
 
@@
 
----
-
-# Comparing Models II
-
-.pull-left[
-```{r logs-table, echo=T, eval=F}
-library(huxtable)
-huxreg("Life Exp." = lin_log_reg,
-       "Log Life Exp." = log_lin_reg,
-       "Log Life Exp." = log_log_reg,
-       coefs = c("Constant" = "(Intercept)",
-                 "GDP ($1000s)" = "gdp_t",
-                 "Log GDP" = "loggdp"),
-       statistics = c("N" = "nobs",
-                      "R-Squared" = "r.squared",
-                      "SER" = "sigma"),
-       number_format = 2)
-```
-
-- Models are very different units, how to choose? 
-  - Compare $R^2$'s
-  - Compare graphs
-  - Compare intution
-
-]
-
-.pull-left[
-```{r, ref.label="logs-table"}
-```
-
-]
-
----
-
-# Comparing Models III
-
 | Linear-Log | Log-Linear | Log-Log |
@@
 
----
+We will do another set of R practice problems, and you will be given **HW 5** to work on this material.
 
-# When to Log?
+# Slides
 
-- In practice, the following types of variables are logged:
-  - Variables that must always be positive (prices, sales, market values)
-  - Very large numbers (population, GDP)
-  - Variables we want to talk about as percentage changes or growth rates (money supply, population, GDP)
-  - Variables that have diminishing returns (output, utility)
-  - Variables that have nonlinear scatterplots
+- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/21-slides.html)
 
---
+# R Practice Problems
 
-- Avoid logs for:
-  - Variables that are less than one, decimals, 0, or negative
-  - Categorical variables (season, gender, political party)
-  - Time variables (year, week, day)
+We will do some [R Practice Problems](/r/21-r-practice) on nonlinear models, which we will continue into Tuesday November 26.
 
----
+# Problem Set 4 Due TODAY
 
-class: inverse, center, middle
-
-# Comparing Across Units
-
----
-
-# Comparing Coefficients of Different Units I
-
-$$\hat{Y_i}=\beta_0+\beta_1 X_1+\beta_2 X_2	$$
-
-- We often want to compare coefficients to see which variable $X_1$ or $X_2$ has a bigger effect on $Y$
-
-- What if $X_1$ and $X_2$ are different units?
-
-.content-box-green[
-.green[**Example**]:
-$$\begin{align*}
-\widehat{\text{Salary}_i}&=\beta_0+\beta_1\, \text{Batting average}_i+\beta_2\, \text{Home runs}_i\\
-\widehat{\text{Salary}_i}&=-\text{2,869,439.40}+\text{12,417,629.72} \, \text{Batting average}_i+\text{129,627.36}\, \text{Home runs}_i\\
-\end{align*}$$
-]
-
----
-
-# Comparing Coefficients of Different Units II
-
-- An easy way is to .shout[standardize] the variables (i.e. take the $Z$-score)
-
-$$X^{std}=\frac{X-\overline{X}}{sd(X)}$$
-
----
-
-# Comparing Coefficients of Different Units: Example
-
-| Variable | Mean | Std. Dev. |
-|----------|------|-----------|
-| Salary | $2,024,616 | $2,764,512 |
-| Batting Average | 0.267 | 0.031 |
-| Home Runs | 12.11 | 10.31 |
-
-$$\begin{align*}\scriptsize  
-	\widehat{\text{Salary}_i}&=-\text{2,869,439.40}+\text{12,417,629.72} \, \text{Batting average}_i+\text{129,627.36} \, \text{Home runs}_i\\ 
-	\widehat{\text{Salary}_i}^{std}&=\text{0.00}+\text{0.14} \, \text{Batting average}_i^{std}+\text{0.48} \, \text{Home runs}_i^{std}\\
-	\end{align*}$$
-
---
-
-- .onfire[Marginal effect] on $Y$ (in *standard deviations* of $Y$) from 1 *standard deviation* change in $X$
-  - $\hat{\beta_1}$: a 1 standard deviation increase in Batting Average increases Salary by 0.14 standard deviations
-    - $0.14 \times 2,764,512=\$387,032$
-  - $\hat{\beta_2}$: a 1 standard deviation increase in Home Runs increases Salary by 0.48 standard deviations
-    - $0.48 \times 2,764,512=\$1,326,966$
-
----
-
-# Standardizing in `R`
-
-- Use the `scale()` command inside `mutate()` function to standardize a variable
-
-```{r, echo=T}
-gapminder<-gapminder %>%
-  mutate(std_life = scale(lifeExp),
-         std_gdp = scale(gdpPercap))
-
-std_reg<-lm(std_life~std_gdp, data = gapminder)
-tidy(std_reg)
-```
-
----
-
-class: inverse, center, middle
-
-# Joint Hypothesis Testing
-
----
-
-# Joint Hypothesis Testing I
-
-.content-box-green[
-.green[**Example**]: Return again to:
-
-$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Male+\hat{\beta_2}Northeast_i+\hat{\beta_3}Midwest_i+\hat{\beta_4}South_i$$
-
-]
-
---
-
-- Maybe region doesn't affect wages *at all*?
-
---
-
-- $H_0: \beta_2=0, \, \beta_3=0, \, \beta_4=0$
-
---
-
-- This is a .shout[joint hypothesis] to test
-
----
-
-# Joint Hypothesis Testing II
-
-- A .shout[joint hypothesis] tests against the null hypothesis of a value for *multiple* parameters:
-$$\mathbf{H_0: \beta_1= \beta_2=0}$$
-the hypotheses that multiple regressors are equal to zero (have no causal effect on the outcome)
-
---
-
-- Our **alternative hypothesis** is that:
-$$H_1: \text{ either } \beta_1\neq0\text{ or } \beta_2\neq0\text{ or both}$$
-or simply, that $H_0$ is not true
-
----
-
-# Types of Joint Hypothesis Tests
-
-- Three main cases of joint hypothesis tests:
-
---
-
-1. $H_0$: $\beta_1=\beta_2=0$
-  - Testing against the claim that multiple variables don't matter
-  - Useful under high multicollinearity between variables
-  - $H_a$: at least one parameter $\neq$ 0
-
---
-
-2. $H_0$: $\beta_1=\beta_2$
-  - Testing whether two variables matter the same
-  - Variables must be the same units
-  - $H_a: \beta_1 (\neq, <, \text{ or }>) \beta_2$
-
---
-
-3. $H_0:$ ALL $\beta$'s $=0$
-  - The "**Overall F-test"**
-  - Testing against claim that regression model explains *NO* variation in $Y$
-
----
-
-# Joint Hypothesis Tests: F-statistic
-
-- The .shout[F-statistic] is the test-statistic used to test joint hypotheses about regression coefficients with an .shout[F-test]
-
---
-
-- This involves comparing two models:
-  1. **Unrestricted model**: regression with all coefficients
-  2. **Restricted model**: regression under null hypothesis (coefficients equal hypothesized values)
-
---
-
-- $F$ is an .onfire[analysis of variance (ANOVA)]
-  - essentially tests whether $R^2$ increases statistically significantly as we go from the restricted model$\rightarrow$unrestricted model
-
---
-
-- $F$ has its own distribution, with *two* sets of degrees of freedom
-
----
-
-# Joint Hypothesis F-test: Example I
-
-.content-box-green[
-.green[**Example**]: Return again to:
-
-$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Male_i+\hat{\beta_2}Northeast_i+\hat{\beta_3}Midwest_i+\hat{\beta_4}South_i$$
-
-]
-
---
-
-- $H_0: \beta_2=\beta_3=\beta_4=0$
-
---
-
-- $H_a$: $H_0$ is not true (at least one $\beta_i \neq 0$)
-
----
-
-# Joint Hypothesis F-test: Example II
-
-.content-box-green[
-.green[**Example**]: Return again to:
-
-$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Male_i+\hat{\beta_2}Northeast_i+\hat{\beta_3}Midwest_i+\hat{\beta_4}South_i$$
-
-]
-
-- **Unrestricted model**:
-
-$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Male_i+\hat{\beta_2}Northeast_i+\hat{\beta_3}Midwest_i+\hat{\beta_4}South_i$$
-
---
-
-- **Restricted model**:
-
-$$\widehat{Wage_i}=\hat{\beta_0}+\hat{\beta_1}Male_i$$
-
---
-
-- $F$: does going from restricted to unrestricted statistically significantly improve $R^2$?
-
----
-
-# Calculating the F-statistic
-
-.pull-left[
-$$F_{q,(n-k-1)}=\cfrac{\left(\displaystyle\frac{(R^2_u-R^2_r)}{q}\right)}{\left(\displaystyle\frac{(1-R^2_u)}{(n-k-1)}\right)}$$
-]
-
-.pull-right[
-
-]
-
----
-
-# Calculating the F-statistic
-
-.pull-left[
-$$F_{q,(n-k-1)}=\cfrac{\left(\displaystyle\frac{(R^2_u-R^2_r)}{q}\right)}{\left(\displaystyle\frac{(1-R^2_u)}{(n-k-1)}\right)}$$
-]
-
-.pull-right[
-
-- $R^2_u$: the $R^2$ from the **unrestricted model** (all variables)
-]
-
----
-
-# Calculating the F-statistic
-
-.pull-left[
-$$F_{q,(n-k-1)}=\cfrac{\left(\displaystyle\frac{(R^2_u-R^2_r)}{q}\right)}{\left(\displaystyle\frac{(1-R^2_u)}{(n-k-1)}\right)}$$
-]
-
-.pull-right[
-
-- $R^2_u$: the $R^2$ from the **unrestricted model** (all variables)
-
-- $R^2_r$: the $R^2$ from the **restricted model** (null hypothesis)
-]
-
----
-
-# Calculating the F-statistic
-
-.pull-left[
-$$F_{q,(n-k-1)}=\cfrac{\left(\displaystyle\frac{(R^2_u-R^2_r)}{q}\right)}{\left(\displaystyle\frac{(1-R^2_u)}{(n-k-1)}\right)}$$
-]
-
-.pull-right[
-
-- $R^2_u$: the $R^2$ from the **unrestricted model** (all variables)
-
-- $R^2_r$: the $R^2$ from the **restricted model** (null hypothesis)
-
-- $q$: number of restrictions (number of $\beta's=0$ under null hypothesis)
-]
-
----
-
-# Calculating the F-statistic
-
-.pull-left[
-$$F_{q,(n-k-1)}=\cfrac{\left(\displaystyle\frac{(R^2_u-R^2_r)}{q}\right)}{\left(\displaystyle\frac{(1-R^2_u)}{(n-k-1)}\right)}$$
-]
-
-.pull-right[
-
-- $R^2_u$: the $R^2$ from the **unrestricted model** (all variables)
-
-- $R^2_r$: the $R^2$ from the **restricted model** (null hypothesis)
-
-- $q$: number of restrictions (number of $\beta's=0$ under null hypothesis)
-
-- $k$: number of $X$ variables in **unrestricted model** (all variables)
-]
-
----
-
-# Calculating the F-statistic
-
-.pull-left[
-$$F_{q,(n-k-1)}=\cfrac{\left(\displaystyle\frac{(R^2_u-R^2_r)}{q}\right)}{\left(\displaystyle\frac{(1-R^2_u)}{(n-k-1)}\right)}$$
-]
-
-.pull-right[
-
-- $R^2_u$: the $R^2$ from the **unrestricted model** (all variables)
-
-- $R^2_r$: the $R^2$ from the **restricted model** (null hypothesis)
-
-- $q$: number of restrictions (number of $\beta's=0$ under null hypothesis)
-
-- $k$: number of $X$ variables in **unrestricted model** (all variables)
-
-- $F$ has two sets of degrees of freedom, $q$ for the numerator, $n-k-1$ for the denominator
-]
-
----
-
-# Calculating the F-statistic II
-
-.pull-left[
-$$F_{q,(n-k-1)}=\cfrac{\left(\displaystyle\frac{(R^2_u-R^2_r)}{q}\right)}{\left(\displaystyle\frac{(1-R^2_u)}{(n-k-1)}\right)}$$
-]
-
-.pull-right[
-
-- .onfire[Key takeaway]: The bigger the difference between $(R^2_u-R^2_r)$, the greater the improvement in fit by adding variables, the larger the $F$!
-
-- This formula is (believe it or not) actually a simplified version (assuming homoskedasticity)
-  - I give you this formula to **build your intuition of what F is measuring**
-
-]
-
----
-
-# F-test Example I
-
-- We'll use the `wooldridge` package's `wage1` data again
-
-```{r, echo=T}
-# load in data from wooldridge package
-library(wooldridge)
-wages<-wooldridge::wage1
-
-# run regressions
-unrestricted_reg<-lm(wage~female+northcen+west+south, data=wages)
-restricted_reg<-lm(wage~female, data=wages)
-```
-
----
-
-# F-test Example II
-
-- Command is `linearHypothesis()`, comes from the `car` package
-
-```{r, echo=T}
-# load car package for additional regression tools
-library("car")
-
-# test that northcen=west=south=0
-# # the first argument is the name of my regression ("unrestricted_reg")
-linearHypothesis(unrestricted_reg, c("northcen", "west", "south")) 
-```
-
----
-
-# All F-test I
-
-.font50[
-```{r, echo=F}
-summary(unrestricted_reg)
-```
-]
-
----
-
-# All F-test II
-
-```{r, echo=T}
-library(broom)
-glance(unrestricted_reg)
-```
-
-- "**statistic**" is the All F-test, "**p.value**" next to it is the p value from the F test!
+[Problm Set 4](/assignment/04-problem-set) (on classes 3.1-3.5) is due TODAY.

### 2019-11-24 static/homeworks/04-problem-set-answers.Rmd
Add HW 4 answers

---
title: "Problem Set 4 (Answers)"
author: "Ryan Safner" # put your name here! 
date: "ECON 480 - Fall 2019"
output: html_document # change to pdf if you'd like, this will make a webpage (you can email it and open in any browser)
---

<!--CLICK "KNIT" ABOVE TO RENDER TO HTML, PDF, OR WORD OUTPUT
In fact, try knitting right away and see what this produces!
-->

# Instructions

Answers may be longer than I would deem sufficient on an exam. Some might vary slightly based on points of interest, examples, or personal experience. These suggested answers are designed to give you both the answer and a short explanation of *why* it is the answer.

# Theory and Concepts

## Question 1
**In your own words, describe what omitted variable bias means. What are the two conditions for an omitted variable to cause a bias?**

---

<!--WRITE YOUR ANSWERS BELOW -->

All variables that might influence the dependent variable ($Y$) that we do not measure and include in our regression are a part of the error term $(\epsilon$). If we omit a variable $(Z)$, it will cause a bias if and only if it meets both of the following conditions:
    1. The variable must be a determinant of our dependent variable, $corr(Z,Y)\neq 0$, and thus would appear in the error term, $\epsilon$.
    2. The variable must be correlated with one of our independent variables of interest in our regression, $corr(Z,X) \neq 0$.

If both conditions are met, then if we did not include the omitted variable $Z$, our estimate of the causal effect of $X$ on $Y$ would be biased, because our estimate $(\hat{\beta_1}$) would pick up some of the effect of $Z$. If we include $Z$ as another independent variable, then the $\hat{\beta_1}$ on $X$ will tell us the precise effect of *only* $X\rightarrow Y$, holding $Z$ constant.

```{r, echo=F, message=F, warning=F}
library(tidyverse)
library(ggdag)
dagify(Y~Z+X,
       X~Z,
       coords=list(
         x=c(X=1,Y=2, Z=1.5),
         y=c(X=1,Y=1, Z=2)
         )) %>% 
  #tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```
---

## Question 2
**In your own words, describe what multicollinearity means. What is the cause, and what are the consequences of multicollinearity? How can we measure multicollinearity and its effects? What happens if multicollinearity is *perfect*?**

---

Multicollinearity just means that two regressors (e.g .$X_1$ and $X_2$) are correlated with each other. This fact does *not* bias the OLS estimates of these regressors (e.g. $\hat{\beta_1}$ and $\hat{\beta_2}$). In fact, the reason $X_2$ is included in the regression is because omitting it would cause omitted variable bias, since $corr(X_1,X_2)\neq 0$ and $corr(Y, X_2)\neq 0$. However, the variance of these OLS estimators is increased because it is hard to get a precise measure of $X_1\rightarrow Y$ because $X_2\rightarrow Y$ also, and $X_1$ may tend to be certain values (large or small) when $X_2$ is certain values (large or small) so we don't know counterfactuals (e.g. what if $X_1$ were the *opposite* of what it tends to be (large or small) when $X_2$ is large or small).

The strength of multicollinearity is simply given by the value of the correlation coefficient between $X_1$ and $X_2$, $r_{X_1,X_2}$. We can measure the *effect* of multicollinearity on the variance of a regressor ($X_j$)'s coefficient ($\hat{\beta_j}$) with the **Variance Inflation Factor**:

$$VIF=\frac{1}{1-R^2_j}$$

where $R^2_j$ is the $R^2$ from an auxiliary regression of $X_j$ on all of the other regressors.

Multicollinearity is *perfect* when the correlation between $X_1$ and $X_2$ is 1 or -1. This happens when one regressor (e.g. $X_1$) is an exact linear function of another regressor(s) $(e.g. X_1=\frac{X_2}{100}$. A regression cannot be run including both variables, as it creates a logical contradiction. In this example, $\hat{\beta_1}$ would be the marginal effect on $Y$ of changing $X_1$ holding $X_2$ constant -- but $X_2$ would naturally change as it is a function of $X_1$!

---

## Question 3

**In your own words, describe what a proxy variable is. When or why would we use a proxy variable, and what effects does it have on our model?**

---

<!--WRITE YOUR ANSWERS BELOW -->

If we can think of a candidate variable $Z$ that would cause omitted variable bias (meeting both conditions in question 1), we would need to include this variable in the regression to remove the bias in our OLS estimate for a regressor $X_j$. If we *do not* have data on $Z$, but we do have data on a variable strongly correlated with $Z$, call it $P$, then we can use $P$ as a **proxy** for Z.

$$\begin{align*}
corr(Y,Z)&\neq 0\\
corr(X_j,Z)&\neq 0\\
corr(Z,P)&\neq 0\\
\end{align*}$$

Including the proxy variable removes the bias that variable $Z$ would have on the estimate for $X_j$ by essentially including something strongly resembling $Z$. It is *not* important whether or not there is a strong $corr(Y,P)$ or that we have a causal explanation of how $P\rightarrow Y$. The *only* reason we included $P$ in the regression is to get rid of omitted variable bias in $X_j$ caused by $Z$!

---

## Question 4

**Explain how we use Directed Acyclic Graphs (DAGs) to depict a causal model: what are the two criteria that must hold for identifying a causal effect of $X$ on $Y$? When should we control a variable, and when should we *not* control a variable?**

---

<!--WRITE YOUR ANSWERS BELOW -->

A Directed Acyclic Graph (DAG) describes a causal model based on making our assumptions about relationships between variables explicit, and in many cases, testable.

Variables are represented as nodes, and causal effects represented as arrows from one node to another (in the direction of the causal effect). We think about the causal effect of $X \rightarrow Y$ in *counterfactual* terms: if $X$ had been different, $Y$ would have been different as a response.

When considering the causal effect of $X \rightarrow Y$, we must consider *all pathways from $X$ to $Y$* (that do not loop, or go through a variable twice), regardless of the direction of the arrows. The paths will be of two types:

- **Causal (front-door) pathways** where arrows go from $X$ into $Y$ (including through other **mediator** variables)
- **Non-causal (back-door) pathways** where an arrow leads into $X$ (implying $X$ is partially caused by that variable)

Adjusting or controlling for (in a multivariate regression, this means including the variable in the regression) a variable along a pathway closes that pathway.

Variables should be adjusted (controlled for) such that:

1. **Back-door criterion**: no backdoor pathway between $X$ and $Y$ remains open
2. **Front-door criterion**: no frontdoor pathway is closed

The one exception is a **collider** variable, where a variable along a pathway has arrows pointing into it from both directions. This *automatically blocks a path* (whether front door or back door). Controlling for a collider variable *opens* the pathway it is on.

See [R Practice on Causality and DAGs](/r/17-r-practice-answers.Rmd) for examples.

---

# Theory Problems

For the following questions, please *show all work* and explain answers as necessary. You may lose points if you only write the correct answer. You may use `R` to *verify* your answers, but you are expected to reach the answers in this section "manually."

## Question 5

Data were collected from a random sample of 220 home sales from a community in 2017.

$$\widehat{Price}=119.2+0.485 \, BDR+23.4 \, Bath+0.156 \, Hsize+0.002 \, Lsize+0.090 \, Age$$

- $Price$: selling price (in \$1,000s)
- $BDR$: number of bedrooms
- $Bath$: number of bathrooms
- $Hsize$: size of the house (in $ft^2$)
- $Lsize$: lot size (in $ft^2$)
- $Age$: age of the house (in years)

### Part A

**Suppose that a homeowner converts part of an existing living space in her house to a new bathroom. What is the expected increase in the value of the house?**

---

<!--WRITE YOUR ANSWERS BELOW -->

From $\hat{\beta_2}$, $23,400

---

### Part B

**Suppose a homeowner adds a new bathroom to her house, which also increases the size of the house by 100 square feet. What is the expected increase in the value of the house?**

---

<!--WRITE YOUR ANSWERS BELOW -->

In this case, $\Delta BDR=1$ and $\Delta Hsize=100$. The resulting expected increase in price is $23.4(1)+0.156(100)=39.0$, or $39,000.

---

### Part C

**Suppose the $R^2$ of this regression is 0.727. Calculate the adjusted $\bar{R}^2$.**

---

<!--WRITE YOUR ANSWERS BELOW -->

There are $n=220$ observations and $k=6$ variables, so:

$$\begin{align*}
\bar{R}^2&=1-\frac{n-1}{n-k-1}(1-R^2)\\
&=1-\frac{220-1}{220-6-1}(1-0.727)\\
&=1-\frac{219}{213}(0.273)\\
&=0.719\\
\end{align*}$$

---

### Part D

**Suppose the following auxiliary regression for $BDR$ has an $R^2$ of 0.841.**

$$\widehat{BDR}=\delta_0+\delta_1Bath+\delta_2Hsize+\delta_3Lsize+\delta_4Age$$

**Calculate the Variance Inflation Factor for $BDR$ and explain what it means.**

---

<!--WRITE YOUR ANSWERS BELOW -->

$$\begin{align*}
VIF&=\frac{1}{1-R^2_j}\\
&=\frac{1}{1-0.841}\\
&=\frac{1}{0.159}\\
&=6.29\\
\end{align*}$$

The variance on $\hat{\beta_2}$ (on Bath) increases by 6.29 times due to multicollinearity between Bath and other $X$-variables.

---

## Question 6

A researcher wants to investigate the effect of education on average hourly wages. Wage, education, and experience in the dataset have the following correlations:

|    | Wage | Education | Experience |
|----|------|-----------|------------|
| Wage | 1.0000 |    |    |
| Education | 0.4059 |    |    |
| Experience | 0.1129 | $-0.2995$ | 1.0000 |

She runs a simple regression first, and gets the results:

$$\widehat{\text{Wage}} = -0.9049 +  0.5414 \, Education$$

She runs another regression, and gets the results:

$$\widehat{\text{Experience}} = 35.4615 - 1.4681 \, Education$$

### Part A

**If the true marginal effect of experience on wages (holding education constant) is 0.0701, calculate the omitted variable bias in the first regression caused by omitting experience. Does the estimate of $\hat{\beta_1}$ in the first regression overstate or understate the effect of education on wages?**

---

<!--WRITE YOUR ANSWERS BELOW -->
We know that the estimate in the biased regression (first one above) is a function of:

$$\hat{\alpha_1}=\hat{\beta_1}+\hat{\beta_2}\hat{\delta_1}$$

Where:

- $\hat{\alpha_1}$: the coefficient on education in the biased regression (0.5414)
- $\hat{\beta_1}$: the true effect of education on wages (??)
- $\hat{\beta_2}$: the true effect of experience on wages (0.0701)
- $\hat{\delta_1}$: the effect of education on experience (from an auxiliary regression) (-1.4681)

$$\begin{align*}
		OMV&=\hat{\beta_2} \hat{\delta_1}\\
		&=(0.0701)(-1.4681)\\
		&=-0.1029
\end{align*}$$

Since the bias is negative, it understates the effect of education (due to experience). Note there are other factors that can bias the effect of education, but at least through experience, the bias is negative since the two are negatively related in the data (see the second regression).

---

### Part B

**Knowing this, what would be the *true effect* of education on wages, holding experience constant?**

---

<!--WRITE YOUR ANSWERS BELOW -->
We know the biased estimate for education, 0.0611. Plugging in both this and the bias, we get the "true" effect:

$$\begin{align*}
			\alpha_1&=\beta_1+\beta_2\delta_1\\
			0.5414&=\beta_1-0.1029\\
			0.6443&=\beta_1\\
\end{align*}$$

---

### Part C

**The $R^2$ for the second regression is 0.0897. If she were to run a better regression including both education and experience, how much would the variance of the coefficients on education and experience increase? Why?**

---

<!--WRITE YOUR ANSWERS BELOW -->
Here we need to calculate the Variance Inflation Factor (VIF) by using the $R^2$ from the auxiliary regression.

$$\begin{align*}
		VIF &=\frac{1}{1-R^2_j}\\
		&=\frac{1}{1-(0.0897)}\\
		&=\frac{1}{0.9103}\\
		&=1.0985
\end{align*}$$

The variance increases only by 1.0985 times due to fairly weak multicollinearity between education and experience.

---

# R Questions

Answer the following questions using `R`. When necessary, please write answers in the same document (knitted `Rmd` to `html` or `pdf`, typed `.doc(x)`, or handwritten) as your answers to the above questions. Be sure to include (email or print an `.R` file, or show in your knitted `markdown`) your code and the outputs of your code with the rest of your answers.

## Question 7

- [<i class="fas fa-table"></i> `heightwages.csv`](http://metricsf19.classes.ryansafner.com/data/heightwages.csv)

Download the `heightwages.csv` dataset. This data is a part of a larger dataset from the National Longitudinal Survey of Youth (NLSY) 1979 cohort: a nationally representative sample of 12,686 men and women aged 14-22 years old when they were first surveyed in 1979. They were subsequently interviewed every year through 1994 and then every other year afterwards. There are many included variables, but for now we will just focus on:

- `wage96`: Adult hourly wages (\$/hr) reported in 1996
- `height85`: Adult height (inches) reported in 1985
- `height81`: Adolescent height (inches) reported in 1981

> We want to figure out what is the effect of height on wages (e.g. do taller people earn more on average than shorter people?)

### Part A

**Create a quick scatterplot between `height85` (as $X)$ amd `wage96` (as $Y)$.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r, warning=F, message=F}
library(tidyverse)

# load data
heights<-read_csv("https://metricsf19.classes.ryansafner.com/Data/heightwages.csv")

# make scatterplot
ggplot(data=heights, aes(x=height85, y=wage96))+
  geom_jitter(color="blue")+
  geom_smooth(method="lm",color="red")+
  labs(x = "Adult Height in 1985 (inches)",
       y = "Hourly Wage in 1996 ($)")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

---

### Part B

**Regress wages on adult height. Write the equation of the estimated OLS regression. Interpret the coefficient on `height85`.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
reg1<-lm(wage96~height85, data=heights)
summary(reg1)
```

$\widehat{\text{Wages}_i} = -6.98+0.31\text{Height}_i$$

For every additional inch in height as an adult, we can expect someone's wages to be $0.31 higher, on average.

---

### Part C

**How much would someone who is 5'10" be predicted to earn per hour, according to the model?**

---

<!--WRITE YOUR ANSWERS BELOW -->

For a 5'10'' person (70 inches), they would earn:

$$\begin{align*}
	\widehat{Wages}&=-6.98+0.31Height\\
	&=-6.98+0.31(70)\\
	&=-6.98+21.70\\
	&=\$14.72\\	
\end{align*}$$

```{r}
# If you want to calculate it with R
library(broom)
reg1_tidy<-tidy(reg1)

beta_0<-reg1_tidy %>%
  filter(term == "(Intercept)") %>%
  pull(estimate)

beta_1<-reg1_tidy %>%
  filter(term == "height85") %>%
  pull(estimate)

beta_0+beta_1*70 # some rounding error in my calculation above
```
---

### Part D

**Would adolescent height cause an omitted variable bias if it were left out? Explain using both your intuition, and some statistical evidence with `R`.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
heights %>%
  select(wage96, height81, height85) %>%
  cor(use = "pairwise.complete.obs")
```

We see that Adolescent height `height81` is weakly correlated with `wage96` (to be fair, so is adult height), but it is strongly correlated with adult height `height85`.

---

### Part E

**Now add adolescent height to the regression, and write the new regression equation below, as before. Interpret the coefficient on `height85`.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
reg2<-lm(wage96~height85+height81, data=heights)
summary(reg2)
```

$$\widehat{\text{Wages}_i}=-9.25-0.11\text{Height85}_i+0.46\text{Height81}_i$$

For every additional inch in height as an adult, we can expect someone's wages to be \$0.11 lower, on average, holding their adolescent height constant.
---

### Part F

**How much would someone who is 5'10" in 1985 and 4'8" in 1981 be predicted to earn, according to the model?**

---

<!--WRITE YOUR ANSWERS BELOW -->

For that 5'10" person (70 inches) in 1985 and 4'10" (58") in 1981, they would earn:

$$\begin{align*}
	\widehat{Wages}&=-9.25-0.11Height85+0.46Height81\\
	&=-9.25-0.11(70)+0.46(58)\\
	&=-9.25-7.70+25.76\\
	&=\$9.73 \\	
\end{align*}$$

```{r}
# If you want to calculate it with R
library(broom)
reg2_tidy<-tidy(reg2)

multi_beta_0<-reg2_tidy %>%
  filter(term == "(Intercept)") %>%
  pull(estimate)

multi_beta_1<-reg2_tidy %>%
  filter(term == "height85") %>%
  pull(estimate)

multi_beta_2<-reg2_tidy %>%
  filter(term == "height81") %>%
  pull(estimate)

multi_beta_0+multi_beta_1*70+multi_beta_2*58 # some rounding error in my calculation above
```

---

### Part G

**What happened to the estimate on `height85` and its standard error?**

---

<!--WRITE YOUR ANSWERS BELOW -->

The effect fell by more than a half, and turned negative. The standard error also doubled (likely due to multicollinearity).

---

### Part H

**Is there multicollinearity between `height85` and `height81`? Explore with a scatterplot.**^[Hint: to avoid overplotting, use `geom_jitter()` instead of `geom_point()` to get a better view of the data.]

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
ggplot(data=heights, aes(x=height81, y=height85))+
  geom_jitter(color="blue")+
  geom_smooth(method="lm",color="red")+
  labs(x = "Adult Height in 1985 (inches)",
       y = "Adolescent Height in 1981 (inches)")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

---

### Part I

**Quantify how much multicollinearity affects the variance of the OLS estimates on both heights**.^[Hint: You'll need the `car` package.]

---

<!--WRITE YOUR ANSWERS BELOW -->

We simply run `vif()` (from the `car` package library) after the recent regression to calculate the Variance Inflation Factor (VIF).

```{r}
#install.packages("car") # install if you don't have 
library("car") # load the car library for the vif() command

vif(reg2) # run vif 
```

The variance of each OLS estimate is inflated by 8.38 times due to multicollinearity.

---

### Part J

**Reach the same number as in part I by running an auxiliary regression.**^[Hint: There's some missing `wage96` data that may give you a different answer, so `filter()` your data here by `!is.na(wage96)` before running this regression - this will include only observations for `wage96` that are not `NA`'s.]

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
aux_reg <- heights %>%
  filter(!is.na(wage96)) %>% # use only for which we have wages
  lm(data = ., height85~height81) # run regression

summary(aux_reg) # look for R-squared
```

We find that adult height is positively affected by adolescent height ($\hat{\beta_1}=0.95$, meaning for every inch that someone is when they are an adolescent, their adult height can be predicted to be about the same). The $R^2$ on the auxiliary regression is 0.88.

\begin{align*}
	VIF&=\frac{1}{1-R^2_1}\\
		&=\frac{1}{1-(0.8807)}\\
		&=\frac{1}{0.1193}\\
		&=8.38\\
\end{align*}

```{r}
# in R:

# extract r.squared using broom
aux_r_sq<-glance(aux_reg) %>%
  pull(r.squared)

# vif formula
1/(1-aux_r_sq)
```

Note if you failed to tell `R` to not drop the data with NA's you will get a different $R^2$ and hence a different VIF than before. This would use more observations that have data on `height81` and `height85` but not `wage96` which would not be used in the regression...hence, a different $n$ and $R^2$.

---

### Part K

**Make a regression table from part B and D using `huxtable`.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
library(huxtable)
huxreg("Wages (1996)"=reg1,
       "Wages (1996)"=reg2,
       coefs = c("Constant" = "(Intercept)",
                 "Adult Height (1985)" = "height85",
                 "Adolescent Height (1981)" = "height81"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```

---

### 2019-11-24 static/homeworks/04-problem-set-answers.R

@@ -1,0 +1,115 @@
+# 7 -------------------
+
+## A -------------
+library(tidyverse)
+
+# load data
+heights<-read_csv("https://metricsf19.classes.ryansafner.com/Data/heightwages.csv")
+
+# make scatterplot
+ggplot(data=heights, aes(x=height85, y=wage96))+
+  geom_jitter(color="blue")+
+  geom_smooth(method="lm",color="red")+
+  labs(x = "Adult Height in 1985 (inches)",
+       y = "Hourly Wage in 1996 ($)")+
+  theme_classic(base_family = "Fira Sans Condensed",
+                base_size=20)
+```
+
+## B -------------
+
+reg1<-lm(wage96~height85, data=heights)
+summary(reg1)
+
+
+## C -------------
+
+# If you want to calculate it with R
+library(broom)
+reg1_tidy<-tidy(reg1)
+
+beta_0<-reg1_tidy %>%
+  filter(term == "(Intercept)") %>%
+  pull(estimate)
+
+beta_1<-reg1_tidy %>%
+  filter(term == "height85") %>%
+  pull(estimate)
+
+beta_0+beta_1*70 # some rounding error in my calculation above
+
+## D -------------
+
+heights %>%
+  select(wage96, height81, height85) %>%
+  cor(use = "pairwise.complete.obs")
+
+## E -------------
+
+reg2<-lm(wage96~height85+height81, data=heights)
+summary(reg2)
+
+## F -------------
+
+# If you want to calculate it with R
+library(broom)
+reg2_tidy<-tidy(reg2)
+
+multi_beta_0<-reg2_tidy %>%
+  filter(term == "(Intercept)") %>%
+  pull(estimate)
+
+multi_beta_1<-reg2_tidy %>%
+  filter(term == "height85") %>%
+  pull(estimate)
+
+multi_beta_2<-reg2_tidy %>%
+  filter(term == "height81") %>%
+  pull(estimate)
+
+multi_beta_0+multi_beta_1*70+multi_beta_2*58 # some rounding error in my calculation above
+
+## H -------------
+
+ggplot(data=heights, aes(x=height81, y=height85))+
+  geom_jitter(color="blue")+
+  geom_smooth(method="lm",color="red")+
+  labs(x = "Adult Height in 1985 (inches)",
+       y = "Adolescent Height in 1981 (inches)")+
+  theme_classic(base_family = "Fira Sans Condensed",
+                base_size=20)
+
[…]

### 2019-11-24 static/homeworks/04-problem-set-answers.Rmd
Update format for HW 4 answers

@@
 title: "Problem Set 4 (Answers)"
-author: "Ryan Safner" # put your name here! 
+author: "Ryan Safner"
 date: "ECON 480 - Fall 2019"
-output: html_document # change to pdf if you'd like, this will make a webpage (you can email it and open in any browser)
+output:
+  html_document:
+    df_print: paged
+    #theme: simplex
+    toc: true 
+    toc_depth: 3
+    toc_float: true
+    code_folding: show
+    highlight: tango
 ---

### 2019-11-26 content/assignment/05-problem-set.Rmd
Finalize HW 5 question page, fix 21 r practice

---
title: "Problem Set 5"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

*Due by Thursday, December 5, 2019*

- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/05-problem-set-pdf.pdf)
- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/05-problem-set.Rmd)

# Instructions

For this problem set, you may submit handwritten answers on a plain sheet of paper, or download and type/handwrite on the PDF.

Alternatively, you may download the `.Rmd` file, do the homework in markdown, and email to me a single `knit`ted `html` or `pdf` file (and be sure that it shows all of your code).

You may work together (and I highly encourage that) but you must turn in your own answers. I grade homeworks 70% for completion, and for the remaining 30%, pick one question to grade for accuracy - so it is best that you try every problem, even if you are unsure how to complete it accurately.

# Theory and Concepts

## Question 1
In your own words, describe what the "dummy variable trap" means. What precisely is the problem, and what is the standard way to prevent it?

## Question 2
In your own words, describe what an interaction term is used for, and give an example. You can use any type of interaction to explain your answer.

## Question 3
In your own words, describe when and why using logged variables can be useful.

## Question 4
In your own words, describe when we would use an $F$-test, and give some example (null) hypotheses. Describe intuitively and specifically (no need for the formula) what exactly $F$ is trying to test for.

# Theory Problems

For the following questions, please *show all work* and explain answers as necessary. You may lose points if you only write the correct answer. You may use `R` to *verify* your answers, but you are expected to reach the answers in this section "manually."

## Question 5

Suppose data on many countries' legal systems (Common Law or Civil Law) and their GDP per capita gives us the following summary statistics:

| Legal System | Avg. GDP Growth Rate | Std. dev. | $n$ |
|--------------|----------------------|-----------|-----|
| Common Law | $1.84$ | $3.55$ | $19$ |
| Civil Law | $4.97$ | $4.27$ | $141$ |
| Difference | $-3.13$ | $1.02$ | $-$ |

### Part A
Using the group means, write a regression equation for a regression of GDP Growth rate on Common Law, including the standard errors in parentheses below the coefficients. Define

$$\text{Common Law}_i = \begin{cases}
    1 & \text{if country } i \text{ has common law} \\
    0 & \text{if country } i \text{ has civil law}\\
    \end{cases}$$

### Part B
How do we use the regression to find the average GDP Growth rate for common law countries? For civil law countries? For the difference?

### Part C
Looking at the coefficients, does there appear to be a statistically significant difference in average GDP Growth Rates between Civil and Common law countries?

### Part D
Is the estimate on the difference likely to be unbiased? Why or why not?

### Part E
Now using the same table above, reconstruct the regression equation if instead of Common Law, we had used:

$$\text{Civil Law}_i = \begin{cases}
    1 & \text{if country } i \text{ has civil law} \\
    0 & \text{if country } i \text{ has common law}\\
    \end{cases}$$

## Question 6

Suppose a real estate agent collects data on houses that have sold in a particular neighborhood over the past year, with the following variables:

- $Price_h$: price of house $h$ (in thousands of $)
- $Bedrooms_h:$ number of bedrooms in house $h$
- $Baths_h:$ number of bathrooms in house $h$
- $Pool_h: \begin{cases} =1 & \text{if house } h \text{ has a pool} \\ =0 & \text{if house } h \text{ does not have a pool} \end{cases}$
- $View_h: \begin{cases} =1 & \text{if house } h \text{ has a nice view} \\ =0 & \text{if house } h \text{ does not have a nice view} \end{cases}$

### Part A
Suppose he runs the following regression:
$$\widehat{\text{Price}_h}=119.20+29.76 \, \text{Bedrooms}_h+24.09 \, \text{View}_h+14.06 \, (\text{Bedrooms}_h \times \text{View}_h)$$

> What does each coefficient mean?

### Part B
Write out **two** separate regression equations, one for houses *with* a nice view, and one for homes *without* a nice view. Explain each coefficient in each regression.

### Part C
Suppose he runs the following regression:

$$\widehat{\text{Price}_h}=189.20+42.40 \, \text{Pool}_h+12.10 \, \text{View}_h+12.09 \, (\text{Pool}_h \times \text{View}_h)$$

> What does each coefficient mean?

### Part D
Find the expected price for:

- a house with no pool and no view
- a house with no pool and a view
- a house with a pool and without a view
- a house with a pool and with a view

### Part E
Suppose he runs the following regression:
$$\widehat{\text{Price}_h}=87.90+53.94 \, \text{Bedrooms}_h+15.29 \, \text{Baths}_h+16.19 \, (\text{Bedrooms}_h \times \text{Baths}_h)$$

> What is the marginal effect of adding an additional *bedroom* if the house has 1 bathroom? 2 bathrooms? 3 bathrooms?

### Part F
What is the marginal effect of adding an additional *bathroom* if the house has 1 bedroom? 2 bedrooms? 3 bedrooms?

## Question 7

Suppose we want to examine the change in average global temperature over time. We have data on the deviation in temperature from pre-industrial times (in Celcius), and the year.

### Part A
Suppose we estimate the following simple model relating deviation in  temperature to year:

$$\widehat{\text{Temperature}_i}=-10.46+0.006 \text{Year}_i$$

> Interpret the coefficient on Year (i.e. $\hat{\beta_1})$

### Part B
Predict the (deviation in) temperature for the year 1900 and for the year 2000.

### Part C
Suppose we believe temperature deviations are increasing at an increasing rate, and introduce a quadratic term and estimate the following regression model:

$$\widehat{\text{Temperature}_i}=155.68-0.116 \text{Year}_i+0.000044\text{Year}_i^2$$

What is the marginal effect on (deviation in) global temperature of one additional year elapsing?

### Part D
Predict the marginal effect on temperature of one more year elapsing starting in 1900, and in 2000.

### Part E
Our quadratic function is a $U$-shape. According to the model, at what year was temperature (deviation) at its minimum?

## Question 8

Suppose we want to examine the effect of cell phone use while driving on traffic fatalities. While we cannot measure the amount of cell phone activity while driving, we do have a good proxy variable, the number of cell phone subscriptions (in 1000s) in a state, along with traffic fatalities in that state.

### Part A
Suppose we estimate the following simple regression:

$$\widehat{\text{fatalities}_i}=123.98+0.091\text{cell plans}_i$$

Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$

### Part B
Now suppose we estimate the regression using a linear-log model:

$$\widehat{\text{fatalities}_i}=-3557.08+515.81\text{ln(cell plans}_i)$$

> Interpret the coefficient on ln(cell plans) (i.e. $\hat{\beta_1})$

### Part C
Now suppose we estimate the regression using a log-linear model:

$$\widehat{\text{ln(fatalities}_i)}=5.43+0.0001\text{cell plans}_i$$

> Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$

### Part D
Now suppose we estimate the regression using a log-log model:

$$\widehat{\text{ln(fatalities}_i)}=-0.89+0.85\text{ln(cell plans}_i)$$

> Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$

### Part E
Suppose we include several other variables into our regression and want to determine which variable(s) have the largest effects, a State's cell plans, population, or amount of miles driven. Suppose we decide to *standardize* the data to compare units, and we get:

$$\widehat{\text{fatalities}_i}=4.35+0.002\text{cell plans}^{std}-0.00007\text{population}^{std}+0.019\text{miles driven}^{std}$$

Interpret the coefficients on cell plans, population, and miles driven. Which has the largest effect on fatalities?

### Part F
Suppose we wanted to make the claim that it is *only* miles driven, and neither population nor cell phones determine traffic fatalities. Write (i) the null hypothesis for this claim and (ii) the estimated restricted regression equation.

### Part G
Suppose the $R^2$ on the original regression from (e) was 0.9221, and the $R^2$ from the restricted regression is 0.9062. With 50 observations, calculate the $F$-statistic.

# R Questions

Answer the following questions using `R`. When necessary, please write answers in the same document (knitted `Rmd` to `html` or `pdf`, typed `.doc(x)`, or handwritten) as your answers to the above questions. Be sure to include (email or print an `.R` file, or show in your knitted `markdown`) your code and the outputs of your code with the rest of your answers.

## Question 7

- [<i class="fas fa-table"></i> `LeadMortality.csv`](http://metricsf19.classes.ryansafner.com/data/LeadMortality.csv)

Lead is toxic, particularly for young children, and for this reason government regulations severely restrict the amount of lead in our environment. In the early part of the 20^th^ century, the underground water pipes in many U.S. cities contained lead, and lead from these pipes leached into drinking water. This exercise will have you investigate the effect of these lead pipes on infant mortality. Download the `LeadMortality.dta` dataset from Blackboard. This dataset contains data on:

- `infrate`: infant mortality rate (deaths per 100 in population)
- `lead`: $=1$ if city has lead water pipes, $=0$ if did not have lead pipes
- `pH`: water pH

and several demographic variables for 172 U.S. cities in 1900.

### Part A

Using `R` to examine the data, find the average infant mortality rate for cities with lead pipes and for cities without lead pipes. Calculate the difference, and run a $t$-test to determine if this difference is statistically significant.

### Part B

Run a regression of `infrate` on `lead`, and write down the estimated regression equation. Use the regression coefficients to find:

- the average infant mortality rate for cities with lead pipes
- the average infant mortality rate for cities without lead pipes
- the difference between the averages for cities with or without lead pipes

### Part C

Does the pH of the water matter? Include `ph` in your regression from part B. Write down the estimated regression equation, and interpret each coefficient (note there is no interaction effect here). What happens to the estimate on `lead`?

### Part D

The amount of lead leached from lead pipes normally depends on the chemistry of the water running through the pipes: the more acidic the water (lower pH), the more lead is leached. Create an interaction term between lead and pH, and run a regression of `infrate` on `lead`, `pH`, and your interaction term. Write down the estimated regression equation. Is this interaction significant?

### Part E

What we actually have are two different regression lines. Visualize this with a scatterplot between `infrate` $(Y)$ and `ph` $(X)$ by `lead`.

### Part F

Do the two regression lines have the same intercept? The same slope? Use the original regression in part D to test these possibilities.

### Part G

Take your regression equation from part D and rewrite it as two separate regression equations (one for no lead and one for lead). Interpret the coefficients for each.

### Part H

Double check your calculations in G are correct by running the regression in D twice, once for cities without lead pipes and once for cities with lead pipes.^[`filter()` the data first, then use the filtered data for the `data=` in each regression.]

### Part I

Use `huxtable` to make a nice output table of all of your regressions from parts B, C, and D.

## Question 8

- [<i class="fas fa-table"></i> `freedom.csv`](http://metricsf19.classes.ryansafner.com/data/freedom.csv)

Let's look at economic freedom and GDP per capita using some data I sourced from [Gapminder](https://gapminder.org/tools)^[GDP per capita (2018)], [Freedom House](https://freedomhouse.org/content/freedom-world-data-and-resources)^[Political freedom score (2018)] and [[Fraser Institute Data](https://www.fraserinstitute.org/economic-freedom/dataset?geozone=world&year=2016&page=dataset)]^[Economic Freedom score (2016)] and cleaned up for you, with the following variables:

| Variable | Description |
| `Country` | Name of country |
| `ISO` | Code of country (good for plotting) |
| `econ_freedom` | Economic Freedom Index score (2016) from 1 (least) to 10 (most free) |
| `pol_freedom` | Political freedom index score (2018) from 1 (least) top 10 (most free) |
| `gdp_pc` | GDP per capita (2018 USD) |
| `continent` | Continent of country |

### Part A
Does economic freedom affect GDP per capita? Create a scatterplot of `gdp_pc` (`Y`) against `econ_freedom` (`x`). Does the effect appear to be linear or nonlinear?

### Part B
Run a simple regression of `gdp_pc` on `econ_freedom`. Write out the estimated regression equation. What is the marginal effect of `econ_freedom` on `gdp_pc`?

### Part C
Let's try a quadratic model. Run a quadratic regression of `gdp_pc` on `econ_freedom`. Write out the estimated regression equation.

### Part D
Add the quadratic regression to your scatterplot.

### Part E
What is the marginal effect of `econ_freedom` on `gdp_pc`?

### Part F
As a quadratic model, this relationship should predict an` econ_freedom` score where `gdp_pc` is at a *minimum*. What is that minimum Economic Freedom score, and what is the associated GDP per capita?

### Part G
Run a cubic model to see if we should keep going up in polynomials. Write out the estimated regression equation. Should we add a cubic term?

### Part H
Another way we can *test* for non-linearity is to run an $F$-test on all non-linear variables - i.e. the quadratic term and the cubic term $(\hat{\beta_2}$ and $\hat{\beta_3}$) and test against the null hypothesis that:
$$H_0: \hat{\beta_2} = \hat{\beta_3} = 0$$

Run this joint hypothesis test, and what can you conclude?

### Part I
Instead of a polynomial model, try out a logarithmic model. It is hard to interpret percent changes on an index, but it is easy to understand percent changes in GDP per capita, so run a **log-linear** regression. Write out the estimated regression equation. What is the marginal effect of `econ_freedom`?

### Part G
Make a scatterplot of your log-linear model with a regression line.

### Part H
Put all of your results together in a regression output table with `huxtable` from your answers in questions B, C, G, and H.

### 2019-11-26 content/assignment/05-problem-set.Rmd
Add problem set 5 rmd

@@
 
-- $Price_h$: price of house $h$ (in thousands of $)
-- $Bedrooms_h:$ number of bedrooms in house $h$
-- $Baths_h:$ number of bathrooms in house $h$
-- $Pool_h: \begin{cases} =1 & \text{if house } h \text{ has a pool} \\ =0 & \text{if house } h \text{ does not have a pool} \end{cases}$
-- $View_h: \begin{cases} =1 & \text{if house } h \text{ has a nice view} \\ =0 & \text{if house } h \text{ does not have a nice view} \end{cases}$
+| Variable | Description |
+| `$Price_h$ | price of house $h$ (in thousands of $) |
+| $Bedrooms_h$ | number of bedrooms in house $h$ |
+| $Baths_h$ | number of bathrooms in house $h$ |
+| $Pool_h$ | $\begin{cases} =1 & \text{if house } h \text{ has a pool} \\ =0 & \text{if house } h \text{ does not have a pool} \end{cases}$ |
+| $View_h$ |  $\begin{cases} =1 & \text{if house } h \text{ has a nice view} \\ =0 & \text{if house } h \text{ does not have a nice view} \end{cases}$ |
 
 ### Part A
-Suppose he runs the following regression:
+Suppose she runs the following regression:
 $$\widehat{\text{Price}_h}=119.20+29.76 \, \text{Bedrooms}_h+24.09 \, \text{View}_h+14.06 \, (\text{Bedrooms}_h \times \text{View}_h)$$
@@
 ### Part C
-Suppose he runs the following regression:
+Suppose she runs the following regression:
 
@@
 ### Part E
-Suppose he runs the following regression:
+Suppose she runs the following regression:
 $$\widehat{\text{Price}_h}=87.90+53.94 \, \text{Bedrooms}_h+15.29 \, \text{Baths}_h+16.19 \, (\text{Bedrooms}_h \times \text{Baths}_h)$$
@@
 
-Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$
+> Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$
 
@@
 
-- `infrate`: infant mortality rate (deaths per 100 in population)
-- `lead`: $=1$ if city has lead water pipes, $=0$ if did not have lead pipes
-- `pH`: water pH
+| Variable | Description |
+| `infrate` | infant mortality rate (deaths per 100 in population) |
+| `lead` | $=1$ if city has lead water pipes, $=0$ if did not have lead pipes |
+| `pH` | water pH |
 
-and several demographic variables for 172 U.S. cities in 1900.
+> and several demographic variables for 172 U.S. cities in 1900.
 
@@
 
-Run this joint hypothesis test, and what can you conclude?
+> Run this joint hypothesis test, and what can you conclude?
 
@@
 
-### Part G
+### Part J
 Make a scatterplot of your log-linear model with a regression line.
 
-### Part H
+### Part K
 Put all of your results together in a regression output table with `huxtable` from your answers in questions B, C, G, and H.

### 2019-11-26 content/assignment/05-problem-set.Rmd
Fix PS 5 tables

@@
 | Variable | Description |
+|----------|-------------|
 | `$Price_h$ | price of house $h$ (in thousands of $) |
@@
 | Variable | Description |
+|----------}-------------}
 | `infrate` | infant mortality rate (deaths per 100 in population) |
@@
 | Variable | Description |
+|----------|-------------|
 | `Country` | Name of country |

### 2019-11-26 static/homeworks/05-problem-set.Rmd
Fix PS 5 tables

--- 2019-11-26 content/assignment/05-problem-set.Rmd
@@
 title: "Problem Set 5"
-editor_options: 
-  chunk_output_type: console
+author: "YOUR NAME HERE" # put your name here! 
+date: "ECON 480 - Fall 2019"
+output: html_document # change to pdf if you'd like, this will make a webpage (you can email it and open in any browser)
 ---
 
-```{r message=FALSE, warning=FALSE, include=FALSE}
-library(tidyverse)
-library(pander)
-source(file.path(here::here(), "lib", "graphics.R"))
-```
+<!--CLICK "KNIT" ABOVE TO RENDER TO HTML, PDF, OR WORD OUTPUT
+In fact, try knitting right away and see what this produces!
+-->
 
-*Due by Thursday, December 5, 2019*
-
-- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/05-problem-set-pdf.pdf)
-- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/05-problem-set.Rmd)
+*Due by Thursday, November 21, 2019*
 
@@
 ## Question 1
-In your own words, describe what the "dummy variable trap" means. What precisely is the problem, and what is the standard way to prevent it?
+**In your own words, describe what the "dummy variable trap" means. What precisely is the problem, and what is the standard way to prevent it?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ## Question 2
-In your own words, describe what an interaction term is used for, and give an example. You can use any type of interaction to explain your answer.
+**In your own words, describe what an interaction term is used for, and give an example. You can use any type of interaction to explain your answer.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ## Question 3
-In your own words, describe when and why using logged variables can be useful.
+
+**In your own words, describe when and why using logged variables can be useful.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ## Question 4
-In your own words, describe when we would use an $F$-test, and give some example (null) hypotheses.
-Describe intuitively and specifically (no need for the formula) what exactly $F$ is trying to test for.
+
+**In your own words, describe when we would use an $F$-test, and give some example (null) hypotheses.
+Describe intuitively and specifically (no need for the formula) what exactly $F$ is trying to test for.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
@@
 ### Part A
-Using the group means, write a regression equation for a regression of GDP Growth rate on Common Law, including the standard errors in parentheses below the coefficients. Define
+
+**Using the group means, write a regression equation for a regression of GDP Growth rate on Common Law, including the standard errors in parentheses below the coefficients. Define**
 
@@
 
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
+
 ### Part B
-How do we use the regression to find the average GDP Growth rate for common law countries? For civil law countries? For the difference?
+
+**How do we use the regression to find the average GDP Growth rate for common law countries? For civil law countries? For the difference?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part C
-Looking at the coefficients, does there appear to be a statistically significant difference in average GDP Growth Rates between Civil and Common law countries?
+
+**Looking at the coefficients, does there appear to be a statistically significant difference in average GDP Growth Rates between Civil and Common law countries?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part D
-Is the estimate on the difference likely to be unbiased? Why or why not?
+
+**Is the estimate on the difference likely to be unbiased? Why or why not?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part E
-Now using the same table above, reconstruct the regression equation if instead of Common Law, we had used:
+
+**Now using the same table above, reconstruct the regression equation if instead of Common Law, we had used:**
 
@@
     \end{cases}$$
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
@@
 ### Part A
-Suppose she runs the following regression:
+
+**Suppose she runs the following regression:**
 $$\widehat{\text{Price}_h}=119.20+29.76 \, \text{Bedrooms}_h+24.09 \, \text{View}_h+14.06 \, (\text{Bedrooms}_h \times \text{View}_h)$$
 
-> What does each coefficient mean?
+> **What does each coefficient mean?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part B
-Write out **two** separate regression equations, one for houses *with* a nice view, and one for homes *without* a nice view. Explain each coefficient in each regression.
+
+**Write out _two_ separate regression equations, one for houses _*_with_ a nice view, and one for homes _without_ a nice view. Explain each coefficient in each regression.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part C
-Suppose she runs the following regression:
+
+**Suppose she runs the following regression:**
 
@@
 
-> What does each coefficient mean?
+> **What does each coefficient mean?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part D
-Find the expected price for:
+**Find the expected price for:**
 
@@
 
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
+
 ### Part E
-Suppose she runs the following regression:
+**Suppose she runs the following regression:**
 $$\widehat{\text{Price}_h}=87.90+53.94 \, \text{Bedrooms}_h+15.29 \, \text{Baths}_h+16.19 \, (\text{Bedrooms}_h \times \text{Baths}_h)$$
 
-> What is the marginal effect of adding an additional *bedroom* if the house has 1 bathroom? 2 bathrooms? 3 bathrooms?
+> **What is the marginal effect of adding an additional _bedroom_ if the house has 1 bathroom? 2 bathrooms? 3 bathrooms?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part F
+
 What is the marginal effect of adding an additional *bathroom* if the house has 1 bedroom? 2 bedrooms? 3 bedrooms?
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
@@
 ### Part A
-Suppose we estimate the following simple model relating deviation in  temperature to year:
+
+**Suppose we estimate the following simple model relating deviation in  temperature to year:**
 
@@
 
-> Interpret the coefficient on Year (i.e. $\hat{\beta_1})$
+> **Interpret the coefficient on Year (i.e. $\hat{\beta_1})$**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part B
-Predict the (deviation in) temperature for the year 1900 and for the year 2000.
+
+**Predict the (deviation in) temperature for the year 1900 and for the year 2000.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part C
-Suppose we believe temperature deviations are increasing at an increasing rate, and introduce a quadratic term and estimate the following regression model:
+
+**Suppose we believe temperature deviations are increasing at an increasing rate, and introduce a quadratic term and estimate the following regression model:**
 
@@
 
-What is the marginal effect on (deviation in) global temperature of one additional year elapsing?
+**What is the marginal effect on (deviation in) global temperature of one additional year elapsing?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part D
-Predict the marginal effect on temperature of one more year elapsing starting in 1900, and in 2000.
+
+**Predict the marginal effect on temperature of one more year elapsing starting in 1900, and in 2000.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part E
-Our quadratic function is a $U$-shape. According to the model, at what year was temperature (deviation) at its minimum?
+
+**Our quadratic function is a $U$-shape. According to the model, at what year was temperature (deviation) at its minimum?**
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
@@
 ### Part A
-Suppose we estimate the following simple regression:
+
+**Suppose we estimate the following simple regression:**
 
@@
 
-> Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$
+> **Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part B
-Now suppose we estimate the regression using a linear-log model:
+
+**Now suppose we estimate the regression using a linear-log model:**
 
@@
 
-> Interpret the coefficient on ln(cell plans) (i.e. $\hat{\beta_1})$
+> **Interpret the coefficient on ln(cell plans) (i.e. $\hat{\beta_1})$**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part C
-Now suppose we estimate the regression using a log-linear model:
+
+**Now suppose we estimate the regression using a log-linear model:**
 
@@
 
-> Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$
+> **Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part D
-Now suppose we estimate the regression using a log-log model:
+
+**Now suppose we estimate the regression using a log-log model:**
 
@@
 
-> Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$
+> **Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part E
-Suppose we include several other variables into our regression and want to determine which variable(s) have the largest effects, a State's cell plans, population, or amount of miles driven.
-Suppose we decide to *standardize* the data to compare units, and we get:
+
+**Suppose we include several other variables into our regression and want to determine which variable(s) have the largest effects, a State's cell plans, population, or amount of miles driven.
+Suppose we decide to _standardize_ the data to compare units, and we get: **
 
@@
 
-Interpret the coefficients on cell plans, population, and miles driven. Which has the largest effect on fatalities?
+**Interpret the coefficients on cell plans, population, and miles driven. Which has the largest effect on fatalities?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part F
-Suppose we wanted to make the claim that it is *only* miles driven, and neither population nor cell phones determine traffic fatalities.
-Write (i) the null hypothesis for this claim and (ii) the estimated restricted regression equation.
+
+**Suppose we wanted to make the claim that it is _only_ miles driven, and neither population nor cell phones determine traffic fatalities.
+Write (i) the null hypothesis for this claim and (ii) the estimated restricted regression equation.***
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
 ### Part G
-Suppose the $R^2$ on the original regression from (e) was 0.9221, and the $R^2$ from the restricted regression is 0.9062. With 50 observations, calculate the $F$-statistic.
+
+**Suppose the $R^2$ on the original regression from (e) was 0.9221, and the $R^2$ from the restricted regression is 0.9062. With 50 observations, calculate the $F$-statistic.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+---
 
@@
 | Variable | Description |
-|----------}-------------}
+|----------|-------------|
 | `infrate` | infant mortality rate (deaths per 100 in population) |
@@
 
-Using `R` to examine the data, find the average infant mortality rate for cities with lead pipes and for cities without lead pipes.
-Calculate the difference, and run a $t$-test to determine if this difference is statistically significant.
+Part A
+
+**Using `R` to examine the data, find the average infant mortality rate for cities with lead pipes and for cities without lead pipes.
+Calculate the difference, and run a $t$-test to determine if this difference is statistically significant.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 
-Run a regression of `infrate` on `lead`, and write down the estimated regression equation. Use the regression coefficients to find:
+**Run a regression of `infrate` on `lead`, and write down the estimated regression equation. Use the regression coefficients to find:**
 
@@
 
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
+
 ### Part C
 
-Does the pH of the water matter?
+**Does the pH of the water matter?
 Include `ph` in your regression from part B.
 Write down the estimated regression equation, and interpret each coefficient (note there is no interaction effect here).
-What happens to the estimate on `lead`?
+What happens to the estimate on `lead`?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 
-The amount of lead leached from lead pipes normally depends on the chemistry of the water running through the pipes: the more acidic the water (lower pH), the more lead is leached.
+**The amount of lead leached from lead pipes normally depends on the chemistry of the water running through the pipes: the more acidic the water (lower pH), the more lead is leached.
 Create an interaction term between lead and pH, and run a regression of `infrate` on `lead`, `pH`, and your interaction term.
 Write down the estimated regression equation.
-Is this interaction significant?
+Is this interaction significant?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 
-What we actually have are two different regression lines. Visualize this with a scatterplot between `infrate` $(Y)$ and `ph` $(X)$ by `lead`.
+**What we actually have are two different regression lines. Visualize this with a scatterplot between `infrate` $(Y)$ and `ph` $(X)$ by `lead`.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 
-Do the two regression lines have the same intercept? The same slope? Use the original regression in part D to test these possibilities.
+**Do the two regression lines have the same intercept? The same slope? Use the original regression in part D to test these possibilities.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 
-Take your regression equation from part D and rewrite it as two separate regression equations (one for no lead and one for lead). Interpret the coefficients for each.
+**Take your regression equation from part D and rewrite it as two separate regression equations (one for no lead and one for lead). Interpret the coefficients for each.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 
-Double check your calculations in G are correct by running the regression in D twice, once for cities without lead pipes and once for cities with lead pipes.^[`filter()` the data first, then use the filtered data for the `data=` in each regression.]
+**Double check your calculations in G are correct by running the regression in D twice, once for cities without lead pipes and once for cities with lead pipes.**^[`filter()` the data first, then use the filtered data for the `data=` in each regression.]
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 
-Use `huxtable` to make a nice output table of all of your regressions from parts B, C, and D.
+**Use `huxtable` to make a nice output table of all of your regressions from parts B, C, and D.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
@@
 ### Part A
-Does economic freedom affect GDP per capita? Create a scatterplot of `gdp_pc` (`Y`) against `econ_freedom` (`x`). Does the effect appear to be linear or nonlinear?
+
+**Does economic freedom affect GDP per capita? Create a scatterplot of `gdp_pc` (`Y`) against `econ_freedom` (`x`). Does the effect appear to be linear or nonlinear?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
 ### Part B
-Run a simple regression of `gdp_pc` on `econ_freedom`. Write out the estimated regression equation. What is the marginal effect of `econ_freedom` on `gdp_pc`?
+
+**Run a simple regression of `gdp_pc` on `econ_freedom`. Write out the estimated regression equation. What is the marginal effect of `econ_freedom` on `gdp_pc`?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
 ### Part C
-Let's try a quadratic model. Run a quadratic regression of `gdp_pc` on `econ_freedom`. Write out the estimated regression equation.
+
+**Let's try a quadratic model. Run a quadratic regression of `gdp_pc` on `econ_freedom`. Write out the estimated regression equation. **
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
 ### Part D
-Add the quadratic regression to your scatterplot.
+
+**Add the quadratic regression to your scatterplot.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
 ### Part E
-What is the marginal effect of `econ_freedom` on `gdp_pc`?
+
+**What is the marginal effect of `econ_freedom` on `gdp_pc`?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
 ### Part F
-As a quadratic model, this relationship should predict an` econ_freedom` score where `gdp_pc` is at a *minimum*. What is that minimum Economic Freedom score, and what is the associated GDP per capita?
+
+**As a quadratic model, this relationship should predict an` econ_freedom` score where `gdp_pc` is at a *minimum*.
+What is that minimum Economic Freedom score, and what is the associated GDP per capita?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
 ### Part G
-Run a cubic model to see if we should keep going up in polynomials. Write out the estimated regression equation. Should we add a cubic term?
+
+**Run a cubic model to see if we should keep going up in polynomials. Write out the estimated regression equation. Should we add a cubic term?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
 ### Part H
-Another way we can *test* for non-linearity is to run an $F$-test on all non-linear variables - i.e.
-the quadratic term and the cubic term $(\hat{\beta_2}$ and $\hat{\beta_3}$) and test against the null hypothesis that:
+
+**Another way we can _test_ for non-linearity is to run an $F$-test on all non-linear variables - i.e.
+the quadratic term and the cubic term $(\hat{\beta_2}$ and $\hat{\beta_3}$) and test against the null hypothesis that:**
 $$H_0: \hat{\beta_2} = \hat{\beta_3} = 0$$
 
-> Run this joint hypothesis test, and what can you conclude?
+> **Run this joint hypothesis test, and what can you conclude?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
 ### Part I
-Instead of a polynomial model, try out a logarithmic model.
-It is hard to interpret percent changes on an index, but it is easy to understand percent changes in GDP per capita, so run a **log-linear** regression.
+
+**Instead of a polynomial model, try out a logarithmic model.
+It is hard to interpret percent changes on an index, but it is easy to understand percent changes in GDP per capita, so run a _log-linear_ regression.
 Write out the estimated regression equation.
-What is the marginal effect of `econ_freedom`?
+What is the marginal effect of `econ_freedom`?**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
 ### Part J
-Make a scatterplot of your log-linear model with a regression line.
+
+**Make a scatterplot of your log-linear model with a regression line.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```
+
+---
 
 ### Part K
-Put all of your results together in a regression output table with `huxtable` from your answers in questions B, C, G, and H.
+
+**Put all of your results together in a regression output table with `huxtable` from your answers in questions B, C, G, and H.**
+
+---
+
+<!--WRITE YOUR ANSWERS BELOW -->
+
+```{r}
+# write your code here
+```

### 2019-11-26 content/assignment/05-problem-set.Rmd
Finalize problem set 5 in all formats

@@
 |----------|-------------|
-| `$Price_h$ | price of house $h$ (in thousands of $) |
+| $Price_h$ | price of house $h$ (in thousands of $) |
 | $Bedrooms_h$ | number of bedrooms in house $h$ |
@@
 
-> What does each coefficient mean?
+What does each coefficient mean?
 
@@
 
-> What is the marginal effect of adding an additional *bedroom* if the house has 1 bathroom? 2 bathrooms? 3 bathrooms?
+What is the marginal effect of adding an additional *bedroom* if the house has 1 bathroom? 2 bathrooms? 3 bathrooms?
 
@@
 
-> Interpret the coefficient on Year (i.e. $\hat{\beta_1})$
+Interpret the coefficient on Year (i.e. $\hat{\beta_1})$
 
@@
 
-> Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$
+Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$
 
@@
 
-> Interpret the coefficient on ln(cell plans) (i.e. $\hat{\beta_1})$
+Interpret the coefficient on ln(cell plans) (i.e. $\hat{\beta_1})$
 
@@
 This exercise will have you investigate the effect of these lead pipes on infant mortality.
-Download the `LeadMortality.dta` dataset from Blackboard.
 This dataset contains data on:
@@
 
-> and several demographic variables for 172 U.S. cities in 1900.
+and several demographic variables for 172 U.S. cities in 1900.
 
@@
 
-Let's look at economic freedom and GDP per capita using some data I sourced from [Gapminder](https://gapminder.org/tools)^[GDP per capita (2018)], [Freedom House](https://freedomhouse.org/content/freedom-world-data-and-resources)^[Political freedom score (2018)] and [[Fraser Institute Data](https://www.fraserinstitute.org/economic-freedom/dataset?geozone=world&year=2016&page=dataset)]^[Economic Freedom score (2016)] and cleaned up for you, with the following variables:
+Let's look at economic freedom and GDP per capita using some data I sourced from [Gapminder](https://gapminder.org/tools)^[GDP per capita (2018)], [Freedom House](https://freedomhouse.org/content/freedom-world-data-and-resources)^[Political freedom score (2018)] and [Fraser Institute Data](https://www.fraserinstitute.org/economic-freedom/dataset?geozone=world&year=2016&page=dataset)^[Economic Freedom score (2016)] and cleaned up for you, with the following variables:
 
@@
 
-> Run this joint hypothesis test, and what can you conclude?
+Run this joint hypothesis test, and what can you conclude?
 

### 2019-11-26 static/homeworks/05-problem-set-pdf.Rmd
Finalize problem set 5 in all formats

--- 2019-11-26 content/assignment/05-problem-set.Rmd
@@
 title: "Problem Set 5"
-editor_options: 
-  chunk_output_type: console
+date: "ECON 480 - Fall 2019"
+output: pdf_document 
 ---
 
-```{r message=FALSE, warning=FALSE, include=FALSE}
-library(tidyverse)
-library(pander)
-source(file.path(here::here(), "lib", "graphics.R"))
-```
+<!--CLICK "KNIT" ABOVE TO RENDER TO HTML, PDF, OR WORD OUTPUT
+In fact, try knitting right away and see what this produces!
+-->
 
-*Due by Thursday, December 5, 2019*
-
-- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/05-problem-set-pdf.pdf)
-- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/05-problem-set.Rmd)
+*Due by Thursday, November 21, 2019*
 
@@
 
-## Question 1
-In your own words, describe what the "dummy variable trap" means. What precisely is the problem, and what is the standard way to prevent it?
+1. In your own words, describe what the "dummy variable trap" means. What precisely is the problem, and what is the standard way to prevent it?
 
-## Question 2
-In your own words, describe what an interaction term is used for, and give an example. You can use any type of interaction to explain your answer.
+---
 
-## Question 3
-In your own words, describe when and why using logged variables can be useful.
+\vspace{2in}
 
-## Question 4
+---
+
+2. In your own words, describe what an interaction term is used for, and give an example. You can use any type of interaction to explain your answer.
+
+---
+
+\clearpage
+
+3. In your own words, describe when and why using logged variables can be useful.
+
+---
+
+\vspace{3in}
+
+---
+
+4.
 In your own words, describe when we would use an $F$-test, and give some example (null) hypotheses.
 Describe intuitively and specifically (no need for the formula) what exactly $F$ is trying to test for.
+
+---
+
+\clearpage
 
@@
 
-## Question 5
-
-Suppose data on many countries' legal systems (Common Law or Civil Law) and their GDP per capita gives us the following summary statistics:
+5. Suppose data on many countries' legal systems (Common Law or Civil Law) and their GDP per capita gives us the following summary statistics:
 
@@
 
-### Part A
-Using the group means, write a regression equation for a regression of GDP Growth rate on Common Law, including the standard errors in parentheses below the coefficients. Define
-
+a. Using the group means, write a regression equation for a regression of GDP Growth rate on Common Law, including the standard errors in parentheses below the coefficients. Define
 $$\text{Common Law}_i = \begin{cases}
@@
     \end{cases}$$
-
-### Part B
-How do we use the regression to find the average GDP Growth rate for common law countries? For civil law countries? For the difference?
-
-### Part C
-Looking at the coefficients, does there appear to be a statistically significant difference in average GDP Growth Rates between Civil and Common law countries?
-
-### Part D
-Is the estimate on the difference likely to be unbiased? Why or why not?
-
-### Part E
-Now using the same table above, reconstruct the regression equation if instead of Common Law, we had used:
+b. How do we use the regression to find the average GDP Growth rate for common law countries? For civil law countries? For the difference?
+c. Looking at the coefficients, does there appear to be a statistically significant difference in average GDP Growth Rates between Civil and Common law countries?
+d. Is the estimate on the difference likely to be unbiased? Why or why not?
+e. Now using the same table above, reconstruct the regression equation if instead of Common Law, we had used:
 
@@
 
-## Question 6
+---
 
-Suppose a real estate agent collects data on houses that have sold in a particular neighborhood over the past year, with the following variables:
+\clearpage
+
+6. Suppose a real estate agent collects data on houses that have sold in a particular neighborhood over the past year, with the following variables:
 
@@
 
-### Part A
-Suppose she runs the following regression:
+a. Suppose she runs the following regression:
 $$\widehat{\text{Price}_h}=119.20+29.76 \, \text{Bedrooms}_h+24.09 \, \text{View}_h+14.06 \, (\text{Bedrooms}_h \times \text{View}_h)$$
-
 What does each coefficient mean?
-
-### Part B
-Write out **two** separate regression equations, one for houses *with* a nice view, and one for homes *without* a nice view. Explain each coefficient in each regression.
-
-### Part C
-Suppose she runs the following regression:
-
+b. Write out _two_ separate regression equations, one for houses _*_with_ a nice view, and one for homes _without_ a nice view. Explain each coefficient in each regression.
+c. Suppose she runs the following regression:
 $$\widehat{\text{Price}_h}=189.20+42.40 \, \text{Pool}_h+12.10 \, \text{View}_h+12.09 \, (\text{Pool}_h \times \text{View}_h)$$
-
 What does each coefficient mean?
-
-### Part D
-Find the expected price for:
-
+d. Find the expected price for:
     - a house with no pool and no view
@@
     - a house with a pool and with a view
+e. Suppose she runs the following regression:
+$$\widehat{\text{Price}_h}=87.90+53.94 \, \text{Bedrooms}_h+15.29 \, \text{Baths}_h+16.19 \, (\text{Bedrooms}_h \times \text{Baths}_h)$$
+What is the marginal effect of adding an additional _bedroom_ if the house has 1 bathroom? 2 bathrooms? 3 bathrooms?
+f. What is the marginal effect of adding an additional *bathroom* if the house has 1 bedroom? 2 bedrooms? 3 bedrooms?
 
-### Part E
-Suppose she runs the following regression:
-$$\widehat{\text{Price}_h}=87.90+53.94 \, \text{Bedrooms}_h+15.29 \, \text{Baths}_h+16.19 \, (\text{Bedrooms}_h \times \text{Baths}_h)$$
+---
 
-What is the marginal effect of adding an additional *bedroom* if the house has 1 bathroom? 2 bathrooms? 3 bathrooms?
+\clearpage
 
-### Part F
-What is the marginal effect of adding an additional *bathroom* if the house has 1 bedroom? 2 bedrooms? 3 bedrooms?
+7. Suppose we want to examine the change in average global temperature over time. We have data on the deviation in temperature from pre-industrial times (in Celcius), and the year.
 
-## Question 7
+a. Suppose we estimate the following simple model relating deviation in  temperature to year:
+$$\widehat{\text{Temperature}_i}=-10.46+0.006 \text{Year}_i$$
+Interpret the coefficient on Year (i.e. $\hat{\beta_1})$
+b. Predict the (deviation in) temperature for the year 1900 and for the year 2000.
+c. Suppose we believe temperature deviations are increasing at an increasing rate, and introduce a quadratic term and estimate the following regression model:
+$$\widehat{\text{Temperature}_i}=155.68-0.116 \text{Year}_i+0.000044\text{Year}_i^2$$
+What is the marginal effect on (deviation in) global temperature of one additional year elapsing?
+d. Predict the marginal effect on temperature of one more year elapsing starting in 1900, and in 2000.
+e. Our quadratic function is a $U$-shape. According to the model, at what year was temperature (deviation) at its minimum?
 
-Suppose we want to examine the change in average global temperature over time. We have data on the deviation in temperature from pre-industrial times (in Celcius), and the year.
+---
 
-### Part A
-Suppose we estimate the following simple model relating deviation in  temperature to year:
+\clearpage
 
-$$\widehat{\text{Temperature}_i}=-10.46+0.006 \text{Year}_i$$
-
-Interpret the coefficient on Year (i.e. $\hat{\beta_1})$
-
-### Part B
-Predict the (deviation in) temperature for the year 1900 and for the year 2000.
-
-### Part C
-Suppose we believe temperature deviations are increasing at an increasing rate, and introduce a quadratic term and estimate the following regression model:
-
-$$\widehat{\text{Temperature}_i}=155.68-0.116 \text{Year}_i+0.000044\text{Year}_i^2$$
-
-What is the marginal effect on (deviation in) global temperature of one additional year elapsing?
-
-### Part D
-Predict the marginal effect on temperature of one more year elapsing starting in 1900, and in 2000.
-
-### Part E
-Our quadratic function is a $U$-shape. According to the model, at what year was temperature (deviation) at its minimum?
-
-## Question 8
-
+8.
 Suppose we want to examine the effect of cell phone use while driving on traffic fatalities.
@@
 
-### Part A
-Suppose we estimate the following simple regression:
+a. Suppose we estimate the following simple regression:
+$$\widehat{\text{fatalities}_i}=123.98+0.091\text{cell plans}_i$$
+Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$
+b. Now suppose we estimate the regression using a linear-log model:
+$$\widehat{\text{fatalities}_i}=-3557.08+515.81\text{ln(cell plans}_i)$$
+Interpret the coefficient on ln(cell plans) (i.e. $\hat{\beta_1})$
+c. Now suppose we estimate the regression using a log-linear model:
+$$\widehat{\text{ln(fatalities}_i)}=5.43+0.0001\text{cell plans}_i$$
+Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$
+d. Now suppose we estimate the regression using a log-log model:
+$$\widehat{\text{ln(fatalities}_i)}=-0.89+0.85\text{ln(cell plans}_i)$$
+Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$
+e.
+Suppose we include several other variables into our regression and want to determine which variable(s) have the largest effects, a State's cell plans, population, or amount of miles driven.
+Suppose we decide to _standardize_ the data to compare units, and we get:
+$$\widehat{\text{fatalities}_i}=4.35+0.002\text{cell plans}^{std}-0.00007\text{population}^{std}+0.019\text{miles driven}^{std}$$
+Interpret the coefficients on cell plans, population, and miles driven. Which has the largest effect on fatalities?
+f.
+Suppose we wanted to make the claim that it is _only_ miles driven, and neither population nor cell phones determine traffic fatalities.
+Write (i) the null hypothesis for this claim and (ii) the estimated restricted regression equation.
+g. Suppose the $R^2$ on the original regression from (e) was 0.9221, and the $R^2$ from the restricted regression is 0.9062. With 50 observations, calculate the $F$-statistic.
 
-$$\widehat{\text{fatalities}_i}=123.98+0.091\text{cell plans}_i$$
+---
 
-Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$
-
-### Part B
-Now suppose we estimate the regression using a linear-log model:
-
-$$\widehat{\text{fatalities}_i}=-3557.08+515.81\text{ln(cell plans}_i)$$
-
-Interpret the coefficient on ln(cell plans) (i.e. $\hat{\beta_1})$
-
-### Part C
-Now suppose we estimate the regression using a log-linear model:
-
-$$\widehat{\text{ln(fatalities}_i)}=5.43+0.0001\text{cell plans}_i$$
-
-Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$
-
-### Part D
-Now suppose we estimate the regression using a log-log model:
-
-$$\widehat{\text{ln(fatalities}_i)}=-0.89+0.85\text{ln(cell plans}_i)$$
-
-Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$
-
-### Part E
-Suppose we include several other variables into our regression and want to determine which variable(s) have the largest effects, a State's cell plans, population, or amount of miles driven.
-Suppose we decide to *standardize* the data to compare units, and we get:
-
-$$\widehat{\text{fatalities}_i}=4.35+0.002\text{cell plans}^{std}-0.00007\text{population}^{std}+0.019\text{miles driven}^{std}$$
-
-Interpret the coefficients on cell plans, population, and miles driven. Which has the largest effect on fatalities?
-
-### Part F
-Suppose we wanted to make the claim that it is *only* miles driven, and neither population nor cell phones determine traffic fatalities.
-Write (i) the null hypothesis for this claim and (ii) the estimated restricted regression equation.
-
-### Part G
-Suppose the $R^2$ on the original regression from (e) was 0.9221, and the $R^2$ from the restricted regression is 0.9062. With 50 observations, calculate the $F$-statistic.
+\clearpage
 
@@
 
-## Question 9
-
-- [<i class="fas fa-table"></i> `LeadMortality.csv`](http://metricsf19.classes.ryansafner.com/data/LeadMortality.csv)
-
+9.
 Lead is toxic, particularly for young children, and for this reason government regulations severely restrict the amount of lead in our environment.
@@
 
-and several demographic variables for 172 U.S. cities in 1900.
+> and several demographic variables for 172 U.S. cities in 1900.
 
-### Part A
-
+a.
 Using `R` to examine the data, find the average infant mortality rate for cities with lead pipes and for cities without lead pipes.
 Calculate the difference, and run a $t$-test to determine if this difference is statistically significant.
-
-### Part B
-
-Run a regression of `infrate` on `lead`, and write down the estimated regression equation. Use the regression coefficients to find:
-
+b. Run a regression of `infrate` on `lead`, and write down the estimated regression equation. Use the regression coefficients to find:
     - the average infant mortality rate for cities with lead pipes
@@
     - the difference between the averages for cities with or without lead pipes
-
-### Part C
-
+c.
 Does the pH of the water matter?
@@
 What happens to the estimate on `lead`?
-
-### Part D
-
+d.
 The amount of lead leached from lead pipes normally depends on the chemistry of the water running through the pipes: the more acidic the water (lower pH), the more lead is leached.
@@
 Is this interaction significant?
+e. What we actually have are two different regression lines. Visualize this with a scatterplot between `infrate` $(Y)$ and `ph` $(X)$ by `lead`.
+f. Do the two regression lines have the same intercept? The same slope? Use the original regression in part D to test these possibilities.
+g. Take your regression equation from part D and rewrite it as two separate regression equations (one for no lead and one for lead). Interpret the coefficients for each.
+h.
+Double check your calculations in G are correct by running the regression in D twice, once for cities without lead pipes and once for cities with lead pipes.**^[`filter()` the data first, then use the filtered data for the `data=` in each regression.]
+i. Use `huxtable` to make a nice output table of all of your regressions from parts B, C, and D.
 
-### Part E
+---
 
-What we actually have are two different regression lines. Visualize this with a scatterplot between `infrate` $(Y)$ and `ph` $(X)$ by `lead`.
+\clearpage
 
-### Part F
-
-Do the two regression lines have the same intercept? The same slope? Use the original regression in part D to test these possibilities.
-
-### Part G
-
-Take your regression equation from part D and rewrite it as two separate regression equations (one for no lead and one for lead). Interpret the coefficients for each.
-
-### Part H
-
-Double check your calculations in G are correct by running the regression in D twice, once for cities without lead pipes and once for cities with lead pipes.^[`filter()` the data first, then use the filtered data for the `data=` in each regression.]
-
-### Part I
-
-Use `huxtable` to make a nice output table of all of your regressions from parts B, C, and D.
-
-## Question 10
-
-- [<i class="fas fa-table"></i> `freedom.csv`](http://metricsf19.classes.ryansafner.com/data/freedom.csv)
-
+10.
 Let's look at economic freedom and GDP per capita using some data I sourced from [Gapminder](https://gapminder.org/tools)^[GDP per capita (2018)], [Freedom House](https://freedomhouse.org/content/freedom-world-data-and-resources)^[Political freedom score (2018)] and [Fraser Institute Data](https://www.fraserinstitute.org/economic-freedom/dataset?geozone=world&year=2016&page=dataset)^[Economic Freedom score (2016)] and cleaned up for you, with the following variables:
@@
 
-### Part A
-Does economic freedom affect GDP per capita? Create a scatterplot of `gdp_pc` (`Y`) against `econ_freedom` (`x`). Does the effect appear to be linear or nonlinear?
-
-### Part B
-Run a simple regression of `gdp_pc` on `econ_freedom`. Write out the estimated regression equation. What is the marginal effect of `econ_freedom` on `gdp_pc`?
-
-### Part C
-Let's try a quadratic model. Run a quadratic regression of `gdp_pc` on `econ_freedom`. Write out the estimated regression equation.
-
-### Part D
-Add the quadratic regression to your scatterplot.
-
-### Part E
-What is the marginal effect of `econ_freedom` on `gdp_pc`?
-
-### Part F
-As a quadratic model, this relationship should predict an` econ_freedom` score where `gdp_pc` is at a *minimum*. What is that minimum Economic Freedom score, and what is the associated GDP per capita?
-
-### Part G
-Run a cubic model to see if we should keep going up in polynomials. Write out the estimated regression equation. Should we add a cubic term?
-
-### Part H
-Another way we can *test* for non-linearity is to run an $F$-test on all non-linear variables - i.e.
+a. Does economic freedom affect GDP per capita? Create a scatterplot of `gdp_pc` (`Y`) against `econ_freedom` (`x`). Does the effect appear to be linear or nonlinear?
+b. Run a simple regression of `gdp_pc` on `econ_freedom`. Write out the estimated regression equation. What is the marginal effect of `econ_freedom` on `gdp_pc`?
+c. Let's try a quadratic model. Run a quadratic regression of `gdp_pc` on `econ_freedom`. Write out the estimated regression equation.
+d. Add the quadratic regression to your scatterplot.
+e. What is the marginal effect of `econ_freedom` on `gdp_pc`?
+f.
+As a quadratic model, this relationship should predict an` econ_freedom` score where `gdp_pc` is at a *minimum*.
+What is that minimum Economic Freedom score, and what is the associated GDP per capita?
+g. Run a cubic model to see if we should keep going up in polynomials. Write out the estimated regression equation. Should we add a cubic term?
+h.
+Another way we can _test_ for non-linearity is to run an $F$-test on all non-linear variables - i.e.
 the quadratic term and the cubic term $(\hat{\beta_2}$ and $\hat{\beta_3}$) and test against the null hypothesis that:
 $$H_0: \hat{\beta_2} = \hat{\beta_3} = 0$$
-
 Run this joint hypothesis test, and what can you conclude?
-
-### Part I
+i.
 Instead of a polynomial model, try out a logarithmic model.
-It is hard to interpret percent changes on an index, but it is easy to understand percent changes in GDP per capita, so run a **log-linear** regression.
+It is hard to interpret percent changes on an index, but it is easy to understand percent changes in GDP per capita, so run a _log-linear_ regression.
 Write out the estimated regression equation.
 What is the marginal effect of `econ_freedom`?
+j. Make a scatterplot of your log-linear model with a regression line.
+k. Put all of your results together in a regression output table with `huxtable` from your answers in questions B, C, G, and H.
 
-### Part J
-Make a scatterplot of your log-linear model with a regression line.
-
-### Part K
-Put all of your results together in a regression output table with `huxtable` from your answers in questions B, C, G, and H.
+---

### 2019-11-26 static/homeworks/05-problem-set.Rmd
Finalize problem set 5 in all formats

@@
 
-> **What does each coefficient mean?**
+**What does each coefficient mean?**
 
@@
 
-> **What is the marginal effect of adding an additional _bedroom_ if the house has 1 bathroom? 2 bathrooms? 3 bathrooms?**
+**What is the marginal effect of adding an additional _bedroom_ if the house has 1 bathroom? 2 bathrooms? 3 bathrooms?**
 
@@
 
-> **Interpret the coefficient on Year (i.e. $\hat{\beta_1})$**
+**Interpret the coefficient on Year (i.e. $\hat{\beta_1})$**
 
@@
 
-> **Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$**
+**Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$**
 
@@
 
-> **Interpret the coefficient on ln(cell plans) (i.e. $\hat{\beta_1})$**
+**Interpret the coefficient on ln(cell plans) (i.e. $\hat{\beta_1})$**
 
@@
 
-> and several demographic variables for 172 U.S. cities in 1900.
+ and several demographic variables for 172 U.S. cities in 1900.
 
@@
 
-> **Run this joint hypothesis test, and what can you conclude?**
+**Run this joint hypothesis test, and what can you conclude?**
 

### 2019-11-30 content/r/21-r-practice.Rmd
Finalize HW 5 question page, fix 21 r practice
Add problem set 5 rmd
Add R Practice 21 answers

@@
 - [<i class="fab fa-markdown"></i> `Download Markdown`](/r/21-r-practice-rmd.Rmd)
+
+Answers:
+- [<i class="fas fa-external-link-square-alt"></i> `Answer Key (html)`](/r/19-r-practice-answers.html)
+- [<i class="fab fa-markdown"></i> `Answer Key (markdown)`](/r/19-r-practice-answers.Rmd)
 
@@
 Run an $F$-test to check if a nonlinear model is appropriate.
-Your null hypothesis is $H_0: $\beta_2+\beta_3=0$ from the regression in pert (h).
+Your null hypothesis is $H_0: \beta_2=\beta_3=0$ from the regression in pert (h).
 The command is `linearHypothesis(reg_name, c("var1", "var2"))` where `reg_name` is the name of the `lm` object you saved your regression in, and `var1` and `var2` (or more) in quotes are the names of the variables you are testing.
@@
 Run a regression of standardized `Amount` on standardized `Age` and `MPHover`. Interpret $\hat{\beta_1}$ and $\hat{\beta_2}$ Which variable has a bigger effect on `Amount`?
+
+# Question 9
+Make a regression output table with `huxtable` of your regressions in Questions 2, 3, 4, 6a, 6b, 6c, 7 and 8.

### 2019-12-03 content/r/21-r-practice.Rmd
Fix link to r practice 21 answers

@@
 Answers:
-- [<i class="fas fa-external-link-square-alt"></i> `Answer Key (html)`](/r/19-r-practice-answers.html)
-- [<i class="fab fa-markdown"></i> `Answer Key (markdown)`](/r/19-r-practice-answers.Rmd)
+
+- [<i class="fas fa-external-link-square-alt"></i> `Answer Key (html)`](/r/21-r-practice-answers.html)
+- [<i class="fab fa-markdown"></i> `Answer Key (markdown)`](/r/21-r-practice-answers.Rmd)
 

### 2019-12-03 static/r/21-r-practice-answers.Rmd
Changed name of R practice answers 21 to fit link

--- 2019-11-21 static/r/21-r-practice-rmd.Rmd
@@
 title: "3.9: Logarithmic Models - R Practice"
-author: "YOUR NAME"
+author: "Ryan Safner"
 date: "`r Sys.Date()`"
@@
 ```{r}
-# PUT CODE HERE
+library(tidyverse)
+
+speed<-read_csv("https://metricsf19.classes.ryansafner.com/data/speeding_tickets.csv")
 ```
@@
 ```{r}
-# PUT CODE HERE
+reg_linear<-lm(Amount ~ Age, data = speed)
+summary(reg_linear)
 ```
+
+$\widehat{\text{Amount}_i}=131.71-0.29\text{Age}_i$
+
+For every year of age, expected fines decrease by $0.29.
 
@@
 ```{r}
-# PUT CODE HERE
+# make Age_sq variable
+speed<-speed %>%
+  mutate(Age_sq=Age^2)
+
+# run quadratic regression
+reg_quad<-lm(Amount ~ Age + Age_sq, data = speed) 
+summary(reg_quad) 
 ```
@@
 ```{r}
-# PUT CODE HERE
+reg_quad_alt<-lm(Amount ~ Age + I(Age^2), data = speed) 
+summary(reg_quad_alt) 
 ```
+
+Yes, this is the same output.
 
@@
 
-```{r}
-# PUT CODE HERE
-```
+$$\widehat{\text{Amount}_i}=146.75-1.17\text{Age}_i+0.01\text{Age}_i^2$$
 
@@
 
-```{r}
-# PUT CODE HERE
-```
+Yes, a slight improvement. $R^2$ went from 0.004 on the linear model to 0.006 on the quadratic model.
 
@@
 <!--ANSWER BELOW HERE-->
+
+The marginal effect is measured by the first derivative of the regression equation with respect to Age. But you can just remember the resulting formula and plug in the parameters:
+
+$$\begin{align*}
+	\frac{d \, Y}{d \, X} &= \beta_1+2\beta_2 X\\
+	\frac{d \, Amount}{d \, Age} &= -1.17+2(0.01) \, Age\\
+	&=-1.17+0.02 \, Age\\
+\end{align*}$$
 
@@
 
+For 18 year olds:
+
+$$\begin{align*}
+	\frac{d \, Amount}{d \, Age} &=-1.17+0.02(18)\\
+ 		&=-1.17+0.36\\
+ 		&=-\$0.81\\
+\end{align*}$$
+
+For 40 year olds:
+
+$$\begin{align*}
+	\frac{d \, Amount}{d \, Age} &=-1.17+0.02(40)\\
+ 		&=-1.17+0.80\\
+ 		&=-\$0.37\\
+ 	\end{align*}$$
+
 ```{r}
-# PUT CODE HERE
+# Let's do this in R:
+
+# we need broom
+library(broom)
+tidy_reg_quad<-tidy(reg_quad)
+
+tidy_reg_quad
+
+# save beta 1 
+quad_beta_1<-tidy_reg_quad %>%
+  filter(term=="Age") %>%
+  pull(estimate)
+
+# save beta 2
+quad_beta_2<-tidy_reg_quad %>%
+  filter(term=="Age_sq") %>%
+  pull(estimate)
+
+# create function to estimate marginal effects
+marginal_effect<-function(x){
+  return(quad_beta_1+2*quad_beta_2*x)
+}
+
+# run the function on the 18-year-old and the 40-year-old
+marginal_effect(c(18,40))
+
+# close enough, we had some rounding error
 ```
@@
 
+We can set the derivative equal to 0, or you can just remember the formula and plug in the parameters:
+
+$$\begin{align*}
+	\frac{d Y}{d X}  &= \beta_1+2\beta_2 X\\
+	0 &=\beta_1+2\beta_2 X\\
+	-\beta_1&=2\beta_2 X\\
+	-\frac{1}{2} \frac{\beta_1}{\beta_2}&=Age^*\\
+	-\frac{1}{2} \frac{-1.17}{0.01} &= Age^*\\
+	-\frac{1}{2} 117 & \approx Age^*\\
+	58.5 & \approx Age ^*\\
+\end{align*}$$
+
 ```{r}
-# PUT CODE HERE
+# Let's do this in R:
+
+-0.5*(quad_beta_1/quad_beta_2)
+
+# again, some rounding error
 ```
@@
 
-```{r}
-# PUT CODE HERE
+```{r, fig.retina=3}
+ggplot(data=speed)+
+  aes(x = Age,
+      y = Amount)+
+  geom_point()+
+  geom_smooth(method="lm",
+              formula="y~poly(x,2)",
+              color="red")
 ```
@@
 
-```{r}
-# PUT CODE HERE
+```{r, fig.retina=3}
+ggplot(data=speed)+
+  aes(x = Age,
+      y = Amount)+
+  geom_smooth(method="lm",
+              formula="y~poly(x,2)",
+              color="red")
 ```
@@
 ```{r}
-# PUT CODE HERE
+reg_cube<-lm(Amount ~ Age + I(Age^2) + I(Age^3), data = speed)
+summary(reg_cube)
+```
+
+$$\widehat{\text{Amount}_i}=151.95-1.61\text{Age}_i+0.02\text{Age}_i^2-0.00008\text{Age}^3$$
+
+The $t$-statistic on Age$^3$ is small (-1.03) and the $p$-value is 0.31, so the cubic term does not have a significant impact on `Amount`. We should *not* include it.
+
+Just for fun, would the cubic model *look* any better?
+
+```{r, fig.retina=3}
+ggplot(data=speed)+
+  aes(x = Age,
+      y = Amount)+
+  geom_point()+
+  geom_smooth(method="lm", formula="y~x+I(x^2)", color="red")+
+  geom_smooth(method="lm", formula="y~x+I(x^2)+I(x^3)", color="orange")
+```
+
+```{r, fig.retina=3}
+ggplot(data=speed)+
+  aes(x = Age,
+      y = Amount)+
+  geom_smooth(method="lm", formula="y~x+I(x^2)", color="red")+
+  geom_smooth(method="lm", formula="y~x+I(x^2)+I(x^3)", color="orange")
 ```
@@
 ```{r}
-# PUT CODE HERE
+# install.packages("car") # install once if you don't have
+library("car") # load car package
+linearHypothesis(reg_cube, c("I(Age^2)", "I(Age^3)")) # F-test 
 ```
+
+We get a large $F$ of 26.43, with a very small $p$-value.
+Therefore, we can reject the null hyothesis that the model is linear $(\beta_2=0, \beta_3=0)$.
+We should in fact *not* use a linear model.
+Note it does *not* tell us if the model should be *quadratic* or *cubic* (or even *logarithmic* of some sort), *only* that it is not linear.
+Remember, this was a *joint* hypothesis of all *non-linear* terms $(beta_2$ and $\beta_3)$!
 
@@
 ```{r}
-# PUT CODE HERE
+# create log of MPHover
+speed<-speed %>%
+  mutate(log_mph=log(MPHover))
+
+# Run linear-log regression
+linear_log_reg<-lm(Amount ~ log_mph, data = speed)
+summary(linear_log_reg)
+
+# note we could have done this without creating the variable
+# just take log() inside the regression:
+linear_log_reg_alt<-lm(Amount ~ log(MPHover), data = speed)
+summary(linear_log_reg)
 ```
+
+$$\widehat{\text{Amount}_i}=-200.10+115.75\text{ln(MPHover}_i)$$
+
+A 1% increase in speed (over the speed limit) increases the fine by $\frac{115.75}{100}\approx \$1.16$
 
@@
 ```{r}
-# PUT CODE HERE
+# create log of Amount
+speed<-speed %>%
+  mutate(log_Amount=log(Amount))
+
+# Run log-linear regression
+log_linear_reg<-lm(log_Amount ~ MPHover, data = speed)
+summary(log_linear_reg)
+
+# again we could have done this without creating the variable
+# just take log() inside the regression:
+log_linear_reg_alt<-lm(log(Amount) ~ MPHover, data = speed)
+summary(log_linear_reg_alt)
 ```
+
+$$\widehat{\text{ln(Amount}_i)}=3.87+0.05\text{MPHover}_i$$
+
+For every 1 MPH in speed (over the speed limit), expected fine increases by $0.05 \times 100\%=5\%$
 
@@
 ```{r}
-# PUT CODE HERE
+# Run log-log regression
+log_log_reg<-lm(log_Amount ~ log_mph, data = speed)
+summary(log_log_reg)
+
+# again we could have done this just taking log()s inside the regression:
+log_log_reg_alt<-lm(log(Amount) ~ log(MPHover), data = speed)
+summary(log_log_reg_alt)
 ```
+
+$$\widehat{\text{ln(Amount}_i)}=2.31+0.86\text{ln(MPHover}_i)$$
+
+For every 1% increase in speed (over the speed limit), expected fine increases by 0.86%.
 
@@
 
+We can compare the $R^2$'s of the three models or compare scatterplots with the regression lines. I will make a table of the three regressions with `huxreg` for easy comparison of fit:
+
 ```{r}
-# PUT CODE HERE
+library(huxtable)
+huxreg("Linear-Log" = linear_log_reg,
+       "Log-Linear" = log_linear_reg,
+       "Log-Log" = log_log_reg,
+       coefs = c("Constant" = "(Intercept)",
+                 "MPH Over" = "MPHover",
+                 "Log(MPH Over)" = "log_mph"),
+       statistics = c("N" = "nobs",
+                      "R-Squared" = "r.squared",
+                      "SER" = "sigma"),
+       number_format = 2)
+```
+
+It appears the linear-log model has the best fit with the highest $R^2$ out of the three, but not by very much.
+
+If we wanted to compare scatterplots:
+
+```{r, fig.height=2}
+ggplot(data = speed)+
+  aes(x = log_mph,
+      y = Amount)+
+  geom_point()+
+  geom_smooth(method="lm", color="red")+
+  labs(title="Linear-Log Model")
+```
+
+```{r, fig.height=2}
+ggplot(data = speed)+
+  aes(x = MPHover,
+      y = log_Amount)+
+  geom_point()+
+  geom_smooth(method="lm", color="red")+
+  labs(title="Log-Linear Model")
+```
+
+```{r, fig.height=2}
+ggplot(data = speed)+
+  aes(x = log_mph,
+      y = log_Amount)+
+  geom_point()+
+  geom_smooth(method="lm", color="red")+
+  labs(title="Log-Log Model")
 ```
@@
 ```{r}
-# PUT CODE HERE
+full_reg<-lm(Amount~Age+Age_sq+MPHover+Black+Hispanic, data=speed)
+summary(full_reg)
+
+library(car)
+linearHypothesis(full_reg, c("Black", "Hispanic"))
 ```
@@
 ```{r}
-# PUT CODE HERE
+# make standardized variables
+speed<-speed %>%
+  mutate(std_Amount = scale(Amount),
+         std_Age = scale(Age),
+         std_MPH = scale(MPHover))
 ```
@@
 Run a regression of standardized `Amount` on standardized `Age` and `MPHover`. Interpret $\hat{\beta_1}$ and $\hat{\beta_2}$ Which variable has a bigger effect on `Amount`?
+
+---
+
+<!--ANSWER BELOW HERE-->
+
+```{r}
+# make standardized variables
+std_reg<-lm(std_Amount ~ std_Age + std_MPH, data = speed)
+summary(std_reg)
+```

### 2019-12-03 static/slides/22-slides.Rmd
Tweak 22 slides

---
title: "4.1: Panel Data and Fixed Effects"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))
```

```{r}
phones<-read_csv("../Data/cellphones.csv")
phones<-phones %>%
  select(-state_numeric) %>%
  mutate(year_num = year) %>%
  mutate_at(c("year", "state", "cell_ban", "text_ban"), as.factor) %>%
  rename(deaths = DeathsPerBillionMiles,
         cell_plans = cell_per10thous_pop)
```

# Types of Data I

.pull-left[

- .red[**Cross-sectional data**]: individual $i$

```{r}
phones %>%
  filter(year=="2012") %>%
  select(state, year, deaths, cell_plans) %>%
  head()
```
]

--

.pull-right[

- .blue[**Time-series data**]: time $t$

```{r}
phones %>%
  filter(state=="Maryland") %>%
  select(state, year, deaths, cell_plans) %>%
  head()
```
]

---

# Types of Data I

.pull-left[

- .red[**Cross-sectional data**]: individual $i$

```{r, fig.height=4, fig.retina=3}
phones %>%
  filter(year=="2012") %>%
  ggplot(data = .)+
  aes(x = cell_plans,
      y = deaths)+
  geom_point(color = "red")+
  labs(x = "Cell Phones Per 10,000 People",
       y = "Deaths Per Billion Miles Driven")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=18)
```
]

.pull-right[

- .blue[**Time-series data**]: time $t$

```{r, fig.height=4, fig.retina=3}
phones %>%
  filter(state=="Maryland") %>%
  ggplot(data = .)+
  aes(x = year_num,
      y = deaths)+
  geom_point(color="blue")+
  geom_path(color="blue", size = 1)+
  labs(x = "Year",
       y = "Deaths Per Billion Miles Driven")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=18)
```
]

--

- .purple[**Panel data**]: combines these dimensions: individual $i$ **at** time $t$

---

# Panel Data I

```{r, fig.retina=3, fig.width=14, fig.align="center"}
phones %>%
  filter(state!="District of Columbia") %>%
  #group_by(state) %>%
  ggplot(data = .)+
  aes(x = cell_plans,
      y = deaths,
      color = state)+
  geom_point()+
  scale_color_viridis_d()+
  geom_path(size = 1)+
  labs(x = "Cell Phones Per 10,000 People",
       y = "Deaths Per Billion Miles Driven")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=18)+
  theme(legend.position = "none")
```

---

# Panel Data II

.pull-left[
```{r}
phones %>%
  arrange(state) %>%
  select(state, year, deaths, cell_plans)
```

]

.pull-right[

- .purple[**Panel**] or .purple[**Longitudinal**] data contains 
  - .blue[repeated observations] $(t)$
  - on .red[multiple individuals] $(i)$
]

---

# Panel Data II

.pull-left[
```{r}
phones %>%
  arrange(state) %>%
  select(state, year, deaths, cell_plans)
```

]

.pull-right[

- .purple[**Panel**] or .purple[**Longitudinal**] data contains 
  - .blue[repeated observations] $(t)$
  - on .red[multiple individuals] $(i)$

- Thus, our regression equation looks like:

$$
\begin{align}
  \hat{Y_{\color{red}{i}\color{blue}{t}}} = \beta_0 + \beta_1 X_{\color{red}{i}\color{blue}{t}} + u_{\color{red}{i}\color{blue}{t}}
\end{align}
$$

> for .red[individual `\\(i\\)`] in .blue[time `\\(t\\)`].

]

---

# Panel Data: Our Motivating Example

.pull-left[
```{r}
phones %>%
  arrange(state) %>%
  select(state, year, deaths, cell_plans)
```

]

.pull-right[

.content-box-green[
.green[**Example**]: Do cell phones cause more traffic fatalities?

- No measure of cell phones *used* while driving
  - `cell_plans` as a **proxy** for cell phone usage

- State-level data over 6 years
]
]

---

# The Data I

```{r, echo=T}
glimpse(phones)
```

---

# The Data II

.pull-left[

```{r, echo=T}
phones %>%
  count(state)
```
]

--

.pull-right[
```{r, echo=T}
phones %>%
  count(year)
```

]

---

# The Data III

.pull-left[

```{r, echo=T}
phones %>%
  distinct(state)
```
]

--

.pull-right[
```{r, echo=T}
phones %>%
  distinct(year)
```

]

---

# The Data IV

```{r, echo=T}
phones %>%
  summarize(States = n_distinct(state),
            Years = n_distinct(year))
```

---

# The Data: With plm

.pull-left[
```{r, echo=T}
# install.packages("plm")
library(plm)

pdim(phones, index=c("state","year"))
```

]

.pull-right[

- **`plm` package** for panel data in R

- `pdim()` checks dimensions of panel dataset
  - `index=` vector of "group" & "year" variables

- Returns with a summary of:
  - `n` groups
  - `T` periods
  - `N` total observaitons
]

---

# Pooled Regression I

- What if we just ran a standard regression:

$$\hat{Y_{it}}=\beta_0+\beta_1X_{it}+u_{it}$$

--
  - $N$ number of $i$ groups (e.g. U.S. States)
  - $T$ number of $t$ periods (e.g. years)

- This is a .shout[pooled regression model]: treats all observations as independent

---

# Pooled Regression II

```{r, echo=T}
pooled<-lm(deaths~cell_plans, data=phones)
summary(pooled)
```

---

# Pooled Regression III

.pull-left[
```{r pooled-plot,echo=T, eval=F}
ggplot(data = phones)+
  aes(x = cell_plans,
      y = deaths)+
  geom_point()+
  labs(x = "Cell Phones Per 10,000 People",
       y = "Deaths Per Billion Miles Driven")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=18)
```

]

.pull-right[
```{r, ref.label="pooled-plot", fig.retina=3}

```
]

---

# Pooled Regression III

.pull-left[
```{r pooled-plot2,echo=T, eval=F}
ggplot(data = phones)+
  aes(x = cell_plans,
      y = deaths)+
  geom_point()+
  geom_smooth(method = "lm", color = "red")+ #<<
  labs(x = "Cell Phones Per 10,000 People",
       y = "Deaths Per Billion Miles Driven")+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=18)
```

]

.pull-right[
```{r, ref.label="pooled-plot2", fig.retina=3}

```
]

---

# Recap: Assumptions about Errors

.pull-left[

- Recall the .onfire[4 critical **assumptions** about `\\(u\\)`]:

1. The expected value of the residuals is 0
$$E[u]=0$$

2. The variance of the residuals over $X$ is constant, written:
$$var(u|X)=\sigma^2_{u}$$

3. Errors are not correlated across observations: 
$$cor(u_i,u_j)=0 \quad \forall i \neq j$$

4. No correlation between $X$ and the error term: 
$$cor(X, u)=0 \text{ or } E[u|X]=0$$

]

.pull-right[

.center[
![](https://www.dropbox.com/s/95rh9ow982y5htb/error.png?raw=1)
]
]

---

# Biases of Pooled Regression

$$\hat{Y_{it}}=\beta_0+\beta_1X_{it}+\epsilon_{it}$$

- Assumption 3: $cor(u_i,u_j)=0 \quad \forall i \neq j$

- Pooled regression model is **biased** because it ignores:
  - Multiple observations from same group $i$ 
  - Multiple observations from same time $t$

- Thus, errors are .shout[serially] or .shout[auto-correlated]; $cor(u_i, u_j) \neq 0$ within same $i$ and within same $t$

---

# Biases of Pooled Regression: Our Example

$$\widehat{\text{Deaths}_{it}}=\beta_0+\beta_1 \, \text{Cell Phones}_{it}+u_{it}$$

- Multiple observations from same state $i$
  - Probably similarities among $u$ for obs in same state

- Multiple observations from same year $t$
  - Probably similarities among $u$ for obs in same year

---

# Look at 5 States

.pull-left[
```{r 5state-plot, echo=T, eval=F}
phones %>%
  filter(state %in% c("District of Columbia",
                      "Maryland", "Texas",
                      "California", "Kansas")) %>%
ggplot(data = .)+
  aes(x = cell_plans,
      y = deaths)+
  geom_point(aes(color = state))+ #<<
  geom_smooth(method = "lm", color = "red")+
  labs(x = "Cell Phones Per 10,000 People",
       y = "Deaths Per Billion Miles Driven",
       color = NULL)+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=18)+
  theme(legend.position = "top")
```

]

.pull-right[
```{r, ref.label="5state-plot", fig.retina=3}

```
]

---

# Look at 5 States

.pull-left[
```{r 5state-plot1, echo=T, eval=F}
phones %>%
  filter(state %in% c("District of Columbia",
                      "Maryland", "Texas",
                      "California", "Kansas")) %>%
ggplot(data = .)+
  aes(x = cell_plans,
      y = deaths,
      color = state)+ #<<
  geom_point()+ #<<
  geom_smooth(method = "lm")+ #<<
  labs(x = "Cell Phones Per 10,000 People",
       y = "Deaths Per Billion Miles Driven",
       color = NULL)+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=18)+
  theme(legend.position = "top")
```

]

.pull-right[
```{r, ref.label="5state-plot1", fig.retina=3}

```
]

---
# Look at 5 States

.pull-left[
```{r 5state-plot2, echo=T, eval=F}
phones %>%
  filter(state %in% c("District of Columbia",
                      "Maryland", "Texas",
                      "California", "Kansas")) %>%
ggplot(data = .)+
  aes(x = cell_plans,
      y = deaths,
      color = state)+ 
  geom_point()+ 
  geom_smooth(method = "lm")+ 
  labs(x = "Cell Phones Per 10,000 People",
       y = "Deaths Per Billion Miles Driven",
       color = NULL)+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=18)+
  theme(legend.position = "none")+ #<<
  facet_wrap(~state, ncol=3) #<<
```

]

.pull-right[
```{r, ref.label="5state-plot2", fig.retina=3}

```
]

---

# Look at All States

.pull-left[
```{r all-state-plot, echo=T, eval=F}
ggplot(data = phones)+ #<<
  aes(x = cell_plans,
      y = deaths,
      color = state)+ 
  geom_point()+ 
  geom_smooth(method = "lm")+ 
  labs(x = "Cell Phones Per 10,000 People",
       y = "Deaths Per Billion Miles Driven",
       color = NULL)+
  theme_bw(base_family = "Fira Sans Condensed")+
  theme(legend.position = "none")+
  facet_wrap(~state, ncol=7) #<<
```

]

.pull-right[
```{r, ref.label="all-state-plot", fig.retina=3}

```
]

---

# The Bias in our Pooled Regression

$$\widehat{\text{Deaths}_{it}}=\beta_0+\beta_1 \, \text{Cell Phones}_{it}+\text{u}_{it}$$

- $\text{Cell Phones}_{it}$ is endogenous:

--

$$cor(\text{u}_{it}, \text{cell phones}_{it}) \neq 0 \quad \quad E[\text{u}_{it}|\text{cell phones}_{it}] \neq 0$$

--

- Things in $u_{it}$ correlated with $\text{Cell phones}_{it}$:
  - infrastructure spending, population, urban vs. rural, more/less cautious citizens, cultural attitudes towards driving, texting, etc

--

- A lot of these things vary systematically **by State**!
  - $cor(\text{u}_{it_1}, \text{u}_{it_2})\neq 0$
    - Error in State $i$ during $t_1$ correlates with error in State $i$ during $t_2$ (things in State that don't change over time)

---

class: inverse, center, middle

# Fixed Effects Model

---

# Fixed Effects: Decomposing $\text{u}_{it}$

- Much of the endogeneity in $X_{it}$ can be explained by systematic differences across $i$ (groups)

--

- Exploit the systematic variation across groups with a .shout[fixed effects model]

--

- *Decompose* the error term into two parts:

$$\text{u}_{it} = \alpha_i + \epsilon_{it}$$

---

# Fixed Effects: $\alpha_i$

$$\text{u}_{it} = \alpha_i + \epsilon_{it}$$

- $\alpha_i$ are .onfire[group-specific fixed effects]
  - group $i$ tends to have higher or lower $\hat{Y}$ than other groups given regressor(s) $X_{it}$
  - estimate a separate $\alpha_i$ for each group $i$
  - essentially, a separate constant (or intercept), $(\hat{\beta_0})$ _for each group_

- .onfire[This includes **all** factors that do not change _within_ group $i$ over time]

---

# Fixed Effects: $\epsilon_{it}$

$$\text{u}_{it} = \alpha_i + \epsilon_{it}$$

- $\epsilon_{it}$ is the remaining random error
  - As usual in OLS, assume the four assumptions about this error:
    - $E[\epsilon_{it}]=0$, $var[\epsilon_{it}]=\sigma^2_{\epsilon}$, $cor(\epsilon_{it}, \epsilon_{jt})=0$, $cor(\epsilon_{it}, X_{it})=0$

- Includes all other factors affecting $Y_{it}$ *not* contained in group effect $\alpha_i$
  - i.e. differences *within* each group that *change* over time

- $X_{it}$ can still be endogenous from non-group-specific effects!

---

# Fixed Effects: New Regression Equation

$$\widehat{Y}_{it} = \beta_0+\beta_1 X_{it} +\alpha_i + \epsilon_{it}$$

- We've pulled $\alpha_i$ out of the original error term into the regression

- Essentially we'll estimate an intercept for each group (minus one, which is $\beta_0)$

- Must have multiple observations (over time) for each group (i.e. panel data)

---

# Fixed Effects: Our Example

$$\widehat{\text{Deaths}}_{it} = \beta_0+\beta_1 \text{Cell phones}_{it} +\alpha_i + \epsilon_{it}$$

- $\alpha_i$ is the .onfire[State fixed effect]
  - Captures everything unique about each state $i$ that *does not change over time*
    - culture, institutions, history, geography, climate, etc!

- There could *still* be factors in $\epsilon_{it}$ that are correlated with $\text{Cell phones}_{it}$!
  - things that do change over time within States
  - Perhaps a State has a cell phone ban only for *some* years in our data

---

# Estimating Fixed Effects Models

$$\widehat{Y}_{it} = \beta_0+\beta_1 X_{it} +\alpha_i$$

- Two methods to estimate fixed effects models:

1. Least Squares Dummy Variable (LSDV) approach

2. De-meaned data approach

---

class: inverse, center, middle

# Least Squares Dummy Variable Approach

---

# Least Squares Dummy Variable Approach

$$\widehat{Y_{it}}=\beta_0+\beta_1X_{it}+\beta_2 D_{1i}+ \beta_3 D_{2i} + \cdots +\beta_N D_{(N-1)i}$$

- A dummy variable $D_{i} = \begin{cases} 1 \text{ if observation is from group }i\\ 0 \text{ if observation is not from group }i\\ \end{cases}$ for each possible group $=1$ if observation $it$ is from group $i$, otherwise $=0$

--

- If there are $N$ groups:
  - Include $N-1$ dummies (to avoid **dummy variable trap**) and $\beta_0$ is the reference category<sup>.red[1]</sup>

--

- So we are estimating different intercepts for each group

--

- Sounds like a lot of work, automatic in `R`

--

- This soaks up *anything* in the error term fixed within groups over time!

.footnote[<sup>.red[1]</sup> If we do not estimate `\\(\beta_0\\)`, we could include all N dummies. In either case, `\\(\beta_0\\)` takes the place of one category-dummy.]

---

# Least Squares Dummy Variable Approach: Our Example

.content-box-green[
.green[**Example**]: 
$$\widehat{\text{Deaths}_{it}}=\beta_1\text{Cell Phones}_{it}+\text{Alabama}_i+\text{Alaska}_i+ \cdots +\text{Wyoming}_i$$
]

---

# Our Example in R I

$$\widehat{\text{Deaths}_{it}}=\beta_1\text{Cell Phones}_{it}+\text{Alabama}_i+\text{Alaska}_i+ \cdots +\text{Wyoming}_i$$

- If `state` is a `factor` variable, just include it in the regression

- `R` automatically creates $N-1$ dummy variables and includes them in the regression
  - Keeps intercept and leaves out first group dummy

---

# Our Example in R II

.pull-right[
```{css, echo = FALSE}
.smallcode .remark-code { /*Change made here*/
  font-size: 65% !important;
}
```

```{r fe-reg, echo=T, eval=F}
fe_reg_1<-lm(deaths~cell_plans+state, data = phones)
summary(fe_reg_1)
```

]

--

.pull-left[
.smallcode[
```{r, ref.label="fe-reg"}
```
]
]

---

# Our Example in R II

```{r}
library(broom)
tidy(fe_reg_1)
```

---

class: inverse, center, middle

# De-meaned Approach

---

# De-meaned Approach I

- Alternatively, we can hold group fixed effects constant without directly estimating them

- We simply **de-mean** the data for each group

- For each group $i$, find the means (over time, $t)$:
$$\bar{Y}_i=\beta_0+\beta_1 \bar{X}_i+\bar{\alpha}_i+\bar{\epsilon}_i$$
- $\bar{Y}_i$: average value of $Y_{it}$ for group $i$
- $\bar{X}_i$: average value of $X_{it}$ for group $i$
- $\bar{\alpha}_i$: average value of $\alpha_{i}$ for group $i$ ($=\alpha_i$)
- $\bar{\epsilon}_{i}=0$, by assumption

---

# De-meaned Approach II

$$\begin{align*}
\widehat{Y_{it}}&=\beta_0+\beta_1X_{it}+u_{it}\\
\bar{Y}_i&=\beta_0+\beta_1 \bar{X}_i+\bar{\alpha}_i+\bar{\epsilon}_i\\
\end{align*}$$

- Subtract the means equation from the pooled equation to get:

$$\begin{align*}
Y_i-\bar{Y}_i&=\beta_1(X_{it}-\bar{X}_i)+\tilde{\epsilon}_{it}\\
\tilde{Y}_{it}&=\beta_1 \tilde{X}_{it}+\tilde{\epsilon}_{it}\\
\end{align*}$$

- Within each group $i$, the de-meaned variables $\tilde{Y}_{it}$ and $\tilde{X}_{it}$'s all have a mean of 0<sup>.red[1]</sup>

- Variables that don't change over time will drop out of analysis altogether

- Removes any source of variation **across** groups to only work with variation **within** each group

.footnote[<sup>.red[1]</sup> Recall **Rule 4** from the [class notes](class/07-class/#the-summation-operator) on the Summation Operator: $\sum(X_i-\bar{X})=0$]

---

# De-meaned Approach III

$$\tilde{Y}_{it}=\beta_1 \tilde{X}_{it}+\tilde{\epsilon}_{it}$$

- Yields identical results to dummy variable approach

- More useful when we have many groups (would be many dummies)

- Demonstrates **intuition** behind fixed effects:
  - Converts all data to deviations from the mean of each group
    - All groups are "centered" at 0
  - Fixed effects are often called the .onfire["within" estimators], they exploit variation *within* groups, not *across* groups

---

# De-meaned Approach IV

- We are basically comparing groups *to themselves* over time
  - apples to apples comparison
  - e.g. Maryland in 2000 vs. Maryland in 2005

- Ignore all differences *between* groups, only look at differences *within* groups over time

---

# De-Meaning the Data in R I

.pull-left[
```{r demeaning, echo=T, eval=F}
# get means of Y and X by state
means_state<-phones %>%
  group_by(state) %>%
  summarize(avg_deaths = mean(deaths),
            avg_phones = mean(cell_plans))

# look at it
means_state
```
]

--

.pull-right[
```{r, ref.label="demeaning"}

```
]

---

# De-Meaning the Data in R II

.pull-left[

```{r demeaning-plot, echo=T, eval=F}
ggplot(data = means_state)+
  aes(x = fct_reorder(state,avg_deaths),
      y = avg_deaths,
      color = state)+
  geom_point()+
  geom_segment(aes(y=0, yend=avg_deaths, x=state, xend=state))+
  coord_flip()+
  labs(x = "Cell Phones Per 10,000 People",
       y = "Deaths Per Billion Miles Driven",
       color = NULL)+
  theme_bw(base_family = "Fira Sans Condensed",
           base_size=10)+
  theme(legend.position = "none")
```
]

.pull-right[
```{r, ref.label="demeaning-plot", fig.retina=3}

```

]

---

# Visualizing "Within Estimates" for the 5 States

```{r}
library(gganimate)
phones2<-phones %>%
filter(state %in% c("District of Columbia",
                      "Maryland", "Texas",
                      "California", "Kansas")) %>%
  group_by(state) %>%
  mutate(mean_phones = mean(cell_plans),
         mean_deaths = mean(deaths))

before_cor <- paste("1. Raw data: cor(cell plans, deaths) = ",round(cor(phones2$cell_plans,phones2$deaths),3),sep='')
after_cor <- paste("6. Analyze what's left! cor(cell plans, deaths) = ",round(cor(phones2$cell_plans-phones2$mean_phones,phones2$deaths-phones2$mean_deaths),3),sep='')

#Add step 2 in which X is demeaned, and 3 in which both X and Y are, and 4 which just changes label
dffull <- rbind(
  #Step 1: Raw data only
  phones2 %>% mutate(mean_phones=NA,mean_deaths=NA,time=before_cor),
  #Step 2: Add x-lines
  phones2 %>% mutate(mean_deaths=NA,time='2. Figure out any between-State differences in cell plans'),
  #Step 3: X de-meaned 
  phones2 %>% mutate(cell_plans = cell_plans - mean_phones,mean_phones=0,mean_deaths=NA,time="3. Remove all between-State differences in cell plans"),
  #Step 4: Remove X lines, add Y
  phones2 %>% mutate(cell_plans = cell_plans - mean_phones,mean_phones=NA,time="4. Figure out any between-State differences in deaths"),
  #Step 5: Y de-meaned
  phones2 %>% mutate(cell_plans = cell_plans - mean_phones,deaths = deaths - mean_deaths,mean_phones=NA,mean_deaths=0,time="5. Remove all between-State differences in deaths"),
  #Step 6: Raw demeaned data only
  phones2 %>% mutate(cell_plans = cell_plans - mean_phones,deaths = deaths - mean_deaths,mean_phones=NA,mean_deaths=NA,time=after_cor))
```

```{r, fig.retina=3}
p <- ggplot(dffull,aes(y=deaths,x=cell_plans,color=as.factor(state)))+geom_point()+
  geom_vline(aes(xintercept=mean_phones,color=as.factor(state)))+
  geom_hline(aes(yintercept=mean_deaths,color=as.factor(state)))+
  labs(x = "Cell Phones Per 10,000 People",
       y = "Deaths Per Billion Miles Driven",
       title = 'The Relationship between Cell Plans and Deaths, with State Fixed Effects \n{next_state}')+
  theme_bw(base_family = "Fira Sans Condensed")+
  theme(legend.position = "none")+
  transition_states(time,transition_length=c(12,32,12,32,12,12),state_length=c(160,100,75,100,75,160),wrap=FALSE)+
  ease_aes('sine-in-out')+
  exit_fade()+enter_fade()

animate(p,nframes=200)
```

---

# Visualizing "Within Estimates" for All 51 States

```{r}
phones3<-phones %>%
  group_by(state) %>%
  mutate(mean_phones = mean(cell_plans),
         mean_deaths = mean(deaths))

before_cor <- paste("1. Raw data: cor(cell plans, deaths) = ",round(cor(phones3$cell_plans,phones3$deaths),3),sep='')
after_cor <- paste("6. Analyze what's left! cor(cell plans, deaths) = ",round(cor(phones3$cell_plans-phones3$mean_phones,phones3$deaths-phones3$mean_deaths),3),sep='')

#Add step 2 in which X is demeaned, and 3 in which both X and Y are, and 4 which just changes label
dffull2 <- rbind(
  #Step 1: Raw data only
  phones3 %>% mutate(mean_phones=NA,mean_deaths=NA,time=before_cor),
  #Step 2: Add x-lines
  phones3 %>% mutate(mean_deaths=NA,time='2. Figure out any between-State differences in cell plans'),
  #Step 3: X de-meaned 
  phones3 %>% mutate(cell_plans = cell_plans - mean_phones,mean_phones=0,mean_deaths=NA,time="3. Remove all between-State differences in cell plans"),
  #Step 4: Remove X lines, add Y
  phones3 %>% mutate(cell_plans = cell_plans - mean_phones,mean_phones=NA,time="4. Figure out any between-State differences in deaths"),
  #Step 5: Y de-meaned
  phones3 %>% mutate(cell_plans = cell_plans - mean_phones,deaths = deaths - mean_deaths,mean_phones=NA,mean_deaths=0,time="5. Remove all between-State differences in deaths"),
  #Step 6: Raw demeaned data only
  phones3 %>% mutate(cell_plans = cell_plans - mean_phones,deaths = deaths - mean_deaths,mean_phones=NA,mean_deaths=NA,time=after_cor))
```

```{r}
p2 <- ggplot(dffull2,aes(y=deaths,x=cell_plans,color=as.factor(state)))+geom_point()+
  geom_vline(aes(xintercept=mean_phones,color=as.factor(state)))+
  geom_hline(aes(yintercept=mean_deaths,color=as.factor(state)))+
  labs(x = "Cell Phones Per 10,000 People",
       y = "Deaths Per Billion Miles Driven",
       title = 'The Relationship between Cell Plans and Deaths, with State Fixed Effects \n{next_state}')+
  theme_bw(base_family = "Fira Sans Condensed")+
  theme(legend.position = "none")+
  transition_states(time,transition_length=c(12,32,12,32,12,12),state_length=c(160,100,75,100,75,160),wrap=FALSE)+
  ease_aes('sine-in-out')+
  exit_fade()+enter_fade()

animate(p2,nframes=200)
```

---

# De-meaned Approach in R I

- The `plm` package is designed for panel data

- `plm()` function is just like `lm()`, with some additional arguments:
  - `index="group_variable_name"` set equal to the name of your `factor` variable for the groups
  - `model=` set equal to `"within"` to use fixed-effects (within-estimator)

```{r plm-fe, echo=T}
#install.packages("plm")
library(plm)
fe_reg_1_alt<-plm(deaths ~ cell_plans,
                  data = phones,
                  index = "state",
                  model = "within")
```

---

# De-meaned Approach in R II

```{r, echo=T}
summary(fe_reg_1_alt)
```

---

class: inverse, center, middle

# Two-Way Fixed Effects:

---

# Two-Way Fixed Effects

- So far, we've looked at a .shout[one-way fixed effects model] that estimates a fixed effect for **groups**

--

- .shout[Two-way fixed effects model] estimates a fixed effect for *both* the **groups** *and* the **time periods**
$$\hat{Y_{it}}=\beta_0+\beta_1 X_{it}+ \alpha_{i} + \theta_{t} + \nu_{it}$$

- $\alpha_i$: group fixed effects
    - accounts for **time-invariant differences across groups**

- $\theta_t$: time fixed effects
    - acounts for **group-invariant differences over time**

---

# Two-Way Fixed Effects: Our Example

$$\widehat{\text{Deaths}}_{it} = \beta_0+\beta_1 \text{Cell phones}_{it} +\alpha_i + \theta_t +  \epsilon_{it}$$

- $\alpha_i$: State fixed effects
    - differences **across states** that are **stable over time**
    - e.g. geography, culture, (unchanging) state laws

- $\theta_t$: Year fixed effects
    - differences **over time** that are **stable across states**
    - e.g. economy-wide macroeconomic changes, *federal* laws passed

---

# Visualizing Year Effects I

```{r, echo=T}
# find averages for years
means_year<-phones %>%
  group_by(year) %>%
  summarize(avg_deaths = mean(deaths),
            avg_phones = mean(cell_plans))
means_year
```

---

# Visualizing Year Effects II

.pull-left[
```{r years-plot,echo=T, eval=F}
ggplot(data = means_year)+
  aes(x = year,
      y = avg_deaths)+
  geom_point(data = phones,
             aes(x = year,
                 y = deaths,
                 color = year))+
  geom_point(size=3)+
  geom_path()
```

]

.pull-right[
```{r, ref.label="years-plot", fig.retina=3}

```
]

---

# Estimating Two-Way Fixed Effects

$$\widehat{Y}_{it} = \beta_0+\beta_1 X_{it} +\alpha_i+\theta_t$$

- As before, several equivalent ways to estimate two-way fixed effects models:

1. **Least Squares Dummy Variable (LSDV) Approach**: add dummies for both groups and time periods (separate intercepts for groups and times)

--

2. **Fully De-meaned data**: 
$$\tilde{Y}_{it}=\beta_1\tilde{X}_{it}+\tilde{\nu}_{it}$$

where each $\tilde{variable}_{it}=variable_{it}-\overline{variable}_{t}-\overline{variable}_{i}$

--

3. **Hybrid**: de-mean for one effect (groups or times) and add dummies for the other effect (times or groups)

---

# LSDV Method

.pull-right[
```{css, echo = FALSE}
.tinycode .remark-code { /*Change made here*/
  font-size: 45% !important;
}
```

```{r fe2-reg, echo=T, eval=F}
fe2_reg_1<-lm(deaths~cell_plans+state+year, data = phones)
summary(fe2_reg_1)
```

]

--

.pull-left[
.tinycode[
```{r, ref.label="fe2-reg"}
```
]
]

---

# With plm

.pull-right[
```{css, echo = FALSE}
.tinycode .remark-code { /*Change made here*/
  font-size: 45% !important;
}
```

```{r fe2-reg2, echo=T, eval=F}
fe2_reg_2<-plm(deaths~cell_plans, index=c("state", "year"), model="within", data = phones)
summary(fe2_reg_2)
```

]

--

.pull-left[
- `plm()` command allows for multiple effects to be fit inside `index=c("group", "time")`

.tinycode[
```{r, ref.label="fe2-reg2"}
```
]
]

```{r}
fe2_reg_2alt<-plm(deaths~cell_plans, data=phones,index=c("state","year"), model="within") 
summary(fe2_reg_2alt)

```

---

# Adding Covariates I

.pull-left[

- Can still add covariates to remove endogeneity not soaked up by fixed effects
    - Factors that change within groups over time 
    - e.g. some states pass bans over the time period in data (some years before, some years after)

]

.pull-right[

$$\widehat{\text{Deaths}_{it}}=\beta_1\text{Cell Phones}_{it}+\alpha_i+\theta_t+\text{urban pct}_{it}+\text{cell ban}_{it}+\text{text ban}_{it}$$
]

---

# Adding Covariates II

.pull-left[
```{r, echo=T}
fe2_controls_reg<-plm(deaths~cell_plans+text_ban+urban_percent+cell_ban, data=phones,index=c("state","year"), model="within", effect="twoways") 
summary(fe2_controls_reg)
```

]

---

# Comparing Models

```{r huxout, echo=T, eval =F}
library(huxtable)
huxreg("Pooled" = pooled,
       "State Effects" = fe_reg_1,
       "State and Year Effects" = fe2_reg_1,
       "With Controls" = fe2_controls_reg,
       coefs = c("Intercept" = "(Intercept)",
                 "Cell phones" = "cell_plans",
                 "Cell Ban" = "cell_ban1",
                 "Texting Ban" = "text_ban1",
                 "Urbanization Rate" = "urban_percent"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 4)
```

---

# Comparing Models

```{r, ref.label="huxout"}
huxout
```

### 2019-12-05 static/papers/Presentatation_example.Rmd
Add presentation example, class 23 content

---
title: "The Effect of Writing Longer Papers"
subtitle: "ECON 480 - Econometrics - Fall 2018"
author: "Ryan Safner"
date: "Dec 10, 2018"
output: 
  beamer_presentation:
    incremental: true 
    fig_caption: yes
    toc: true 
    slide_level: 3
    includes:
      in_header: template.tex
    keep_tex: no 
    latex_engine: xelatex #required for Fira Sans font to render properly

classoption: aspectratio=169
---

```{r setup, include=FALSE}
# For making transparent ggplot2 graphs, from https://gist.github.com/cboettig/5600558

# Set plotting to bw plot default, but with transparent background elements.  
# Note transparency requires the panel.background, plot.background, and device background all be set!
library(tidyverse)
library(knitr)
library(stargazer)

theme_set(theme_bw(base_size=12))
theme_update(panel.background = element_rect(fill = "transparent", colour = NA),
             plot.background = element_rect(fill = "transparent", colour = NA))
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, dev.args=list(bg="transparent"))
```

```{r, load-data}
papers<-read.csv("paperlengthsregcsv.csv")
```

### Overview

- As a student writing an empirical research paper, does writing a longer paper attain a higher grade?
- Simple OLS regression
- Sample data collected from previous classes with paper assignments 
- Strong positive effect: for every marginal page written, grades improve by about 1-2 points 
- Robust to different models

### Theory

- Dilemma:
1. Longer papers might imply students work hard and write a good paper 
2. But students might also put in low quality filler hoping to inflate their grades

### Data

- I collected data from 7 different classes at 2 different colleges where I assigned a longer term paper

\columnsbegin

\column{0.33\textwidth}

```{r}
papers %>% 
  count(Class) %>%
  kable(.)
```

\column{0.33\textwidth}

```{r}
papers %>% 
  count(School) %>%
  kable(.)
```

\column{0.33\textwidth}

```{r}
papers %>% 
  count(Year) %>%
  kable(.)
```

\columnsend

### Variables Used

\begin{table}
  \begin{tabular}{ll}
  Variable & Description \\ \toprule
  Grade & Grade on paper assignment (0-100)\\ \midrule
  Pages & Number of pages written\\ 
  Female & $=1$ if student is female, $=0$ if male\\
  Morning & $=1$ if class met in morning, $=0$ if afternoon\\
  Hood & $=1$ if class taught at Hood, $=0$ if other school\\
  Metrics & $=1$ if class was Econometrics, $=0$ if other class\\ \bottomrule 
  \end{tabular} 
\end{table}

### Summary Statistics

```{r, results="asis"}
source("nice.summary.R")
sumstats<-nice.summary(papers,Grade,Pages)
sumstats
```

### Data: Histogram of Pages Written

```{r, fig.height=3.5}
library("ggplot2")
h<-ggplot(papers, aes(x=Pages))+
  geom_histogram(color="black", fill="indianred",breaks=seq(0,24,2))+
  scale_x_continuous(breaks=seq(0,24,2))+xlab("Pages Written")+ylab("Number of Students")+theme_light()
h
```

```{r, echo=F}
papers$Female<-ifelse(papers$Sex=="F",1,0)
papers$Morning<-ifelse(papers$Time=="Morning",1,0)
papers$Hood<-ifelse(papers$School=="Hood",1,0)
papers$Metrics<-ifelse(papers$Class=="Econometrics",1,0)
```

### Data: Histogram of Paper Grades

```{r}
g<-ggplot(papers, aes(x=Grade))+
  geom_histogram(color="black", fill="indianred",breaks=seq(0,100,10))+
  scale_x_continuous(breaks=seq(0,100,10))+xlab("Paper Grade")+ylab("Number of Students")+theme_light()
g
```

### Scatterplot of Pages Written and Grade with Outliers (0's) Dropped

```{r}

papers0<-subset(papers, Grade>0)

s0<-ggplot(papers0, aes(x=Pages, y=Grade))+
  geom_jitter(color="blue")+
  geom_smooth(method="lm", color="red")+
  scale_x_continuous(breaks=seq(0,24,2))+
  scale_y_continuous(breaks=seq(0,100,10))+xlab("Pages Written")+ylab("Paper Grades")+theme_light()
s0
```

### Scatterplot For Econometrics Classes vs Non-Econometrics Classes

```{r}
papers$Metrics<-ifelse(papers$Class=="Econometrics",1,0)

s0+facet_grid(cols=vars(Metrics))
```

### Empirical Model

$$Grade_i=Length_i+Final_i+Female_i+Morning_i+Hood_i+Metrics_i\epsilon_i$$

- $Length$ is the most important variable we care about
- $Length$ probably endogenous, correlated with other Grade-determining factors
    - Why we included $Final$, $Female$, $Morning$, $Hood$, and $Metrics$
- We probably expect $Length$ to be positive and small 
- **Endogeneity**: unmeasured attributes (e.g. paper topic, student's diligence, conscienciousness, intelligence, etc) correlated with both number of `Length` and `Grade`

### Results

```{r, regs, results="asis"}
basicreg<-lm(Grade~Pages, data=papers)
no0reg<-lm(Grade~Pages, data=papers0)
finalreg<-lm(Grade~Pages+Final, data=papers0)
sexreg<-lm(Grade~Pages+Final+Female, data=papers0)
timereg<-lm(Grade~Pages+Final+Female+Morning, data=papers0)
hoodreg<-lm(Grade~Pages+Final+Female+Morning+Hood, data=papers0)
metricsreg<-lm(Grade~Pages+Final+Female+Morning+Hood+Metrics, data=papers0)

stargazer(basicreg,finalreg,sexreg,timereg,hoodreg,metricsreg, type="latex", header=FALSE, float=FALSE, column.labels = c("All", "No 0s", "No 0s", "No 0s", "No 0s", "No 0s", "No 0s"), font.size = "tiny", column.sep.width = "-1pt", omit.stat=c("F", "rsq"), dep.var.caption = "")
```

### Results II

- For each additional page written, paper grades increase by about about 1-2 points 
- Still probably strongly endogenous 
- Poor advice to just write long papers!

### 2019-12-05 static/slides/23-slides.Rmd
Add presentation example, class 23 content

---
title: "4.2: Differences-in-Differences Models"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))
```

class: inverse, center, middle

# Difference-in-Difference Models

---

# Difference-in-Difference Models I

- Often, we want to examine the consequences of a change, such as a law or policy
    - e.g. States that implemented law $X$ saw a change in $Y$
    - **Treatment**: States that implement law $X$
    - **Control**: States that did not implement law $X$
    - If we have **panel data** with observations for all states before *and* after the change:
- Simple logic: compare difference in outcomes of treatment group (before and after treatment) with those of non-treated group (before and after same treatment period)

---

# Difference-in-Difference Models II

.center[
![](https://www.dropbox.com/s/vpk560fok07yo9t/dndhot.jpg?raw=1)
]

---

# Difference-in-Difference Models III

- The .shout[difference-in-difference model] (aka .shout["diff-in-diff"] or .shout[DND"]) identifies treatment effect by differencing the difference pre- and post-treatment between treatment and control groups

--

$$\hat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$

--

- $Treated_i= \begin{cases}1 \text{ if } i \text{ is in treatment group}\\ 0 \text{ if } i \text{ is not in treatment group}\end{cases}$

- $After_t= \begin{cases}1 \text{ if } t \text{ is after treatment period}\\ 0 \text{ if } t \text{ is before treatment period}\end{cases}$

--

|    | Control | Treatment | **Group Diff** $(\Delta Y_i)$ |
|----|---------|-----------|---------------------------|
| Before | $\beta_0$ | $\beta_0+\beta_1$ | $\beta_1$ |
| After | $\beta_0+\beta_2$ | $\beta_0+\beta_1+\beta_2+\beta_3$ | $\beta_1+\beta_3$ |
| **Time Diff** $(\Delta Y_t)$ | $\beta_2$ | $\beta_2+\beta_3$ | **Diff-in-diff** $\Delta_i \Delta_t: \beta_3$ |

---

```{r}
control<-tribble(
  ~x, ~y,
  1, 1,
  2, 3
)
treatment<-tribble(
  ~x, ~y,
  1, 4,
  2, 8
)
```

# Visualizing Diff-in-Diff

$$\hat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$

.pull-left[

```{r, fig.height=6, fig.retina=3}
ggplot(data = tibble(x=1,3))+
  aes(x = x)+
  # control
  geom_point(data = control, aes(x=x,y=y), size = 3, color = "red")+
  geom_text(x=1,y=0.5, color = "red", label=expression(C[1]))+
  geom_text(x=2,y=3.5, color = "red", label=expression(C[2]))+

  # line
  geom_segment(x=1, y=1, xend=2, yend=3, color="red", size = 2)+

  # beta 0 
  geom_label(x=0.9,y=1, color = "red", label=expression(hat(beta)[0]))+

  # line slope
  geom_segment(x=1, y=1, xend=2, yend=1, color="red", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=1, xend=2, yend=3, color="red", linetype = "dotted", size = 1)+
  # beta 2
  geom_label(x=2.1,y=2, color = "red", label=expression(hat(beta)[2]))+

  scale_x_continuous(breaks=c(1,2),
                  labels=c(expression(t[before]),expression(t[after])),
                  limits=c(0.5,2.5))+
  scale_y_continuous(breaks=NULL,
                     limits=c(0,9))+
  labs(x = "Time",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

- .red[Control group `\\((Treated = 0)\\)`]

- .red[`\\(\hat{\beta_0}\\)`]: value of $Y$ for **control** group **before** treatment

- .red[`\\(\hat{\beta_2}\\)`]: time *difference* (for **control** group)

]

---

# Visualizing Diff-in-Diff

$$\hat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$

.pull-left[

```{r, fig.height=6, fig.retina=3}
ggplot(data = tibble(x=1,3))+
  aes(x = x)+
  # control
  geom_point(data = control, aes(x=x,y=y), size = 3, color = "red")+
  geom_text(x=1,y=0.5, color = "red", label=expression(C[1]))+
  geom_text(x=2,y=3.5, color = "red", label=expression(C[2]))+

  # line
  geom_segment(x=1, y=1, xend=2, yend=3, color="red", size = 2)+

  # beta 0 
  geom_label(x=0.9,y=1, color = "red", label=expression(hat(beta)[0]))+

  # line slope
  geom_segment(x=1, y=1, xend=2, yend=1, color="red", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=1, xend=2, yend=3, color="red", linetype = "dotted", size = 1)+
  # beta 2
  geom_label(x=2.1,y=2, color = "red", label=expression(hat(beta)[2]))+

  # treatment
  geom_point(data = treatment, aes(x=x,y=y), size = 3, color = "green")+
  geom_text(x=1,y=4.5, color = "green", label=expression(T[1]))+
  geom_text(x=2,y=8.5, color = "green", label=expression(T[2]))+

    # line
  geom_segment(x=1, y=4, xend=2, yend=8, color="green", size = 2)+

  # beta 0 plus beta 1 
  geom_label(x=0.8,y=4, color = "green", label=expression(hat(beta)[0]+hat(beta)[1]))+

    # line slope
  geom_segment(x=1, y=4, xend=2, yend=4, color="green", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  # diag
  geom_segment(x=1, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  # beta 2
  geom_label(x=2.1,y=5, color = "red",label=expression(hat(beta)[2]))+

  scale_x_continuous(breaks=c(1,2),
                  labels=c(expression(t[before]),expression(t[after])),
                  limits=c(0.5,2.5))+
  scale_y_continuous(breaks=NULL,
                     limits=c(0,9))+
  labs(x = "Time",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

- .red[Control group `\\((Treated = 0)\\)`]

- .red[`\\(\hat{\beta_0}\\)`]: value of $Y$ for **control** group **before** treatment

- .red[`\\(\hat{\beta_2}\\)`]: time *difference* (for **control** group)

- .green[Treated group `\\((Treated = 1)\\)`]

]

---

# Visualizing Diff-in-Diff

$$\hat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$

.pull-left[

```{r, fig.height=6, fig.retina=3}
ggplot(data = tibble(x=1,3))+
  aes(x = x)+
  # control
  geom_point(data = control, aes(x=x,y=y), size = 3, color = "red")+
  geom_text(x=1,y=0.5, color = "red", label=expression(C[1]))+
  geom_text(x=2,y=3.5, color = "red", label=expression(C[2]))+

  # line
  geom_segment(x=1, y=1, xend=2, yend=3, color="red", size = 2)+

  # beta 0 
  geom_label(x=0.9,y=1, color = "red", label=expression(hat(beta)[0]))+

  # line slope
  geom_segment(x=1, y=1, xend=2, yend=1, color="red", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=1, xend=2, yend=3, color="red", linetype = "dotted", size = 1)+
  # beta 2
  geom_label(x=2.1,y=2, color = "red", label=expression(hat(beta)[2]))+

  # treatment
  geom_point(data = treatment, aes(x=x,y=y), size = 3, color = "green")+
  geom_text(x=1,y=4.5, color = "green", label=expression(T[1]))+
  geom_text(x=2,y=8.5, color = "green", label=expression(T[2]))+

    # line
  geom_segment(x=1, y=4, xend=2, yend=8, color="green", size = 2)+

  # beta 0 plus beta 1 
  geom_label(x=0.8,y=4, color = "green", label=expression(hat(beta)[0]+hat(beta)[1]))+

  # beta 1 dif
  geom_segment(x=1, y=1, xend=1, yend=4, color="black", linetype = "dashed", size = 1)+
  geom_label(x=0.9,y=2.5, color = "black", label=expression(hat(beta)[1]))+

    # line slope
  geom_segment(x=1, y=4, xend=2, yend=4, color="green", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  # diag
  geom_segment(x=1, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  # beta 2
  geom_label(x=2.1,y=5, color = "red",label=expression(hat(beta)[2]))+

  scale_x_continuous(breaks=c(1,2),
                  labels=c(expression(t[before]),expression(t[after])),
                  limits=c(0.5,2.5))+
  scale_y_continuous(breaks=NULL,
                     limits=c(0,9))+
  labs(x = "Time",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

- .red[Control group `\\((Treated = 0)\\)`]

- .red[`\\(\hat{\beta_0}\\)`]: value of $Y$ for **control** group **before** treatment

- .red[`\\(\hat{\beta_2}\\)`]: time *difference* (for **control** group)

- .green[Treated group `\\((Treated = 1)\\)`]

- $\hat{\beta_1}$: *difference* between groups **before** treatment

]

---

# Visualizing Diff-in-Diff

$$\hat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$

.pull-left[

```{r, fig.height=6, fig.retina=3}
ggplot(data = tibble(x=1,3))+
  aes(x = x)+
  # control
  geom_point(data = control, aes(x=x,y=y), size = 3, color = "red")+
  geom_text(x=1,y=0.5, color = "red", label=expression(C[1]))+
  geom_text(x=2,y=3.5, color = "red", label=expression(C[2]))+

  # line
  geom_segment(x=1, y=1, xend=2, yend=3, color="red", size = 2)+

  # beta 0 
  geom_label(x=0.9,y=1, color = "red", label=expression(hat(beta)[0]))+

  # line slope
  geom_segment(x=1, y=1, xend=2, yend=1, color="red", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=1, xend=2, yend=3, color="red", linetype = "dotted", size = 1)+
  # beta 2
  geom_label(x=2.1,y=2, color = "red", label=expression(hat(beta)[2]))+

  # treatment
  geom_point(data = treatment, aes(x=x,y=y), size = 3, color = "green")+
  geom_text(x=1,y=4.5, color = "green", label=expression(T[1]))+
  geom_text(x=2,y=8.5, color = "green", label=expression(T[2]))+

    # line
  geom_segment(x=1, y=4, xend=2, yend=8, color="green", size = 2)+

  # beta 0 plus beta 1 
  geom_label(x=0.8,y=4, color = "green", label=expression(hat(beta)[0]+hat(beta)[1]))+

  # beta 1 dif
  geom_segment(x=1, y=1, xend=1, yend=4, color="black", linetype = "dashed", size = 1)+
  geom_label(x=0.9,y=2.5, color = "black", label=expression(hat(beta)[1]))+

    # line slope
  geom_segment(x=1, y=4, xend=2, yend=4, color="green", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  # diag
  geom_segment(x=1, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  geom_segment(x=2, y=6, xend=2, yend=8, color="blue", linetype = "dashed", size = 1)+

  # beta 2
  geom_label(x=2.1,y=5, color = "red",label=expression(hat(beta)[2]))+

  # beta 3
  geom_label(x=2.1,y=7, color = "blue", label=expression(hat(beta)[3]))+

  scale_x_continuous(breaks=c(1,2),
                  labels=c(expression(t[before]),expression(t[after])),
                  limits=c(0.5,2.5))+
  scale_y_continuous(breaks=NULL,
                     limits=c(0,9))+
  labs(x = "Time",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

- .red[Control group `\\((Treated = 0)\\)`]

- .red[`\\(\hat{\beta_0}\\)`]: value of $Y$ for **control** group **before** treatment

- .red[`\\(\hat{\beta_2}\\)`]: time *difference* (for **control** group)

- .green[Treated group `\\((Treated = 1)\\)`]

- $\hat{\beta_1}$: *difference* between groups **before** treatment

- .blue[`\\(\hat{\beta_3}\\)`]: difference-in-difference: .blue[treatment effect]
]

---

# Visualizing Diff-in-Diff II

$$\hat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$

.pull-left[

```{r, fig.height=6, fig.retina=3}
ggplot(data = tibble(x=1,3))+
  aes(x = x)+
  # control
  geom_point(data = control, aes(x=x,y=y), size = 3, color = "red")+
  geom_text(x=1,y=0.5, color = "red", label=expression(C[1]))+
  geom_text(x=2,y=3.5, color = "red", label=expression(C[2]))+

  # line
  geom_segment(x=1, y=1, xend=2, yend=3, color="red", size = 2)+

  # beta 0 
  geom_label(x=0.9,y=1, color = "black", label=expression(hat(beta)[0]))+

  # line slope
  geom_segment(x=1, y=1, xend=2, yend=1, color="red", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=1, xend=2, yend=3, color="red", linetype = "dotted", size = 1)+

  # treatment
  geom_point(data = treatment, aes(x=x,y=y), size = 3, color = "green")+
  geom_text(x=1,y=4.5, color = "green", label=expression(T[1]))+
  geom_text(x=2,y=8.5, color = "green", label=expression(T[2]))+

    # line
  geom_segment(x=1, y=4, xend=2, yend=8, color="green", size = 2)+

  # line slope
  geom_segment(x=1, y=4, xend=2, yend=4, color="green", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  # diag
  geom_segment(x=1, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  geom_segment(x=2, y=6, xend=2, yend=8, color="blue", linetype = "dashed", size = 1)+

  scale_x_continuous(breaks=c(1,2),
                  labels=c(expression(t[before]),expression(t[after])),
                  limits=c(0.5,2.5))+
  scale_y_continuous(breaks=NULL,
                     limits=c(0,9))+
  labs(x = "Time",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

- $Y_i$ for **Control** group **before**: $\hat{\beta_0}$

]

---

# Visualizing Diff-in-Diff II

$$\hat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$

.pull-left[

```{r, fig.height=6, fig.retina=3}
ggplot(data = tibble(x=1,3))+
  aes(x = x)+
  # control
  geom_point(data = control, aes(x=x,y=y), size = 3, color = "red")+
  geom_text(x=1,y=0.5, color = "red", label=expression(C[1]))+
  geom_text(x=2,y=3.5, color = "red", label=expression(C[2]))+

  # line
  geom_segment(x=1, y=1, xend=2, yend=3, color="red", size = 2)+

  # beta 0 
  geom_label(x=0.9,y=1, color = "black", label=expression(hat(beta)[0]))+

  # line slope
  geom_segment(x=1, y=1, xend=2, yend=1, color="red", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=1, xend=2, yend=3, color="red", linetype = "dotted", size = 1)+
  # beta 1+2
  geom_label(x=2.2,y=3, color = "black", label=expression(hat(beta)[0]+hat(beta)[2]))+

  # treatment
  geom_point(data = treatment, aes(x=x,y=y), size = 3, color = "green")+
  geom_text(x=1,y=4.5, color = "green", label=expression(T[1]))+
  geom_text(x=2,y=8.5, color = "green", label=expression(T[2]))+

    # line
  geom_segment(x=1, y=4, xend=2, yend=8, color="green", size = 2)+

  # line slope
  geom_segment(x=1, y=4, xend=2, yend=4, color="green", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  # diag
  geom_segment(x=1, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  geom_segment(x=2, y=6, xend=2, yend=8, color="blue", linetype = "dashed", size = 1)+

  scale_x_continuous(breaks=c(1,2),
                  labels=c(expression(t[before]),expression(t[after])),
                  limits=c(0.5,2.5))+
  scale_y_continuous(breaks=NULL,
                     limits=c(0,9))+
  labs(x = "Time",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

- $Y_i$ for **Control** group **before**: $\hat{\beta_0}$

- $Y_i$ for **Control** group **after**: $\hat{\beta_0}+\hat{\beta_2}$

]

---

# Visualizing Diff-in-Diff II

$$\hat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$

.pull-left[

```{r, fig.height=6, fig.retina=3}
ggplot(data = tibble(x=1,3))+
  aes(x = x)+
  # control
  geom_point(data = control, aes(x=x,y=y), size = 3, color = "red")+
  geom_text(x=1,y=0.5, color = "red", label=expression(C[1]))+
  geom_text(x=2,y=3.5, color = "red", label=expression(C[2]))+

  # line
  geom_segment(x=1, y=1, xend=2, yend=3, color="red", size = 2)+

  # beta 0 
  geom_label(x=0.9,y=1, color = "black", label=expression(hat(beta)[0]))+

  # line slope
  geom_segment(x=1, y=1, xend=2, yend=1, color="red", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=1, xend=2, yend=3, color="red", linetype = "dotted", size = 1)+
  # beta 1+2
  geom_label(x=2.2,y=3, color = "black", label=expression(hat(beta)[0]+hat(beta)[2]))+

  # treatment
  geom_point(data = treatment, aes(x=x,y=y), size = 3, color = "green")+
  geom_text(x=1,y=4.5, color = "green", label=expression(T[1]))+
  geom_text(x=2,y=8.5, color = "green", label=expression(T[2]))+

    # line
  geom_segment(x=1, y=4, xend=2, yend=8, color="green", size = 2)+

  # beta 0 plus beta 1 
  geom_label(x=0.8,y=4, color = "black", label=expression(hat(beta)[0]+hat(beta)[1]))+

  # line slope
  geom_segment(x=1, y=4, xend=2, yend=4, color="green", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  # diag
  geom_segment(x=1, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  geom_segment(x=2, y=6, xend=2, yend=8, color="blue", linetype = "dashed", size = 1)+

  scale_x_continuous(breaks=c(1,2),
                  labels=c(expression(t[before]),expression(t[after])),
                  limits=c(0.5,2.5))+
  scale_y_continuous(breaks=NULL,
                     limits=c(0,9))+
  labs(x = "Time",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

- $Y_i$ for **Control** group **before**: $\hat{\beta_0}$

- $Y_i$ for **Control** group **after**: $\hat{\beta_0}+\hat{\beta_2}$

- $Y_i$ for **Treatment** group **before**: $\hat{\beta_0}+\hat{\beta_1}$

]

---

# Visualizing Diff-in-Diff II

$$\hat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$

.pull-left[

```{r, fig.height=6, fig.retina=3}
ggplot(data = tibble(x=1,3))+
  aes(x = x)+
  # control
  geom_point(data = control, aes(x=x,y=y), size = 3, color = "red")+
  geom_text(x=1,y=0.5, color = "red", label=expression(C[1]))+
  geom_text(x=2,y=3.5, color = "red", label=expression(C[2]))+

  # line
  geom_segment(x=1, y=1, xend=2, yend=3, color="red", size = 2)+

  # beta 0 
  geom_label(x=0.9,y=1, color = "black", label=expression(hat(beta)[0]))+

  # line slope
  geom_segment(x=1, y=1, xend=2, yend=1, color="red", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=1, xend=2, yend=3, color="red", linetype = "dotted", size = 1)+
  # beta 1+2
  geom_label(x=2.2,y=3, color = "black", label=expression(hat(beta)[0]+hat(beta)[2]))+

  # treatment
  geom_point(data = treatment, aes(x=x,y=y), size = 3, color = "green")+
  geom_text(x=1,y=4.5, color = "green", label=expression(T[1]))+
  geom_text(x=2,y=8.5, color = "green", label=expression(T[2]))+

    # line
  geom_segment(x=1, y=4, xend=2, yend=8, color="green", size = 2)+

  # beta 0 plus beta 1 
  geom_label(x=0.8,y=4, color = "black", label=expression(hat(beta)[0]+hat(beta)[1]))+

  # line slope
  geom_segment(x=1, y=4, xend=2, yend=4, color="green", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  # diag
  geom_segment(x=1, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  geom_segment(x=2, y=6, xend=2, yend=8, color="blue", linetype = "dashed", size = 1)+

  # beta 3
  geom_label(x=2.25,y=8, color = "black",
             label=expression(hat(beta)[0]+hat(beta)[1]+hat(beta)[2]+hat(beta)[3]))+

  scale_x_continuous(breaks=c(1,2),
                  labels=c(expression(t[before]),expression(t[after])),
                  limits=c(0.5,2.5))+
  scale_y_continuous(breaks=NULL,
                     limits=c(0,9))+
  labs(x = "Time",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

- $Y_i$ for **Control** group **before**: $\hat{\beta_0}$

- $Y_i$ for **Control** group **after**: $\hat{\beta_0}+\hat{\beta_2}$

- $Y_i$ for **Treatment** group **before**: $\hat{\beta_0}+\hat{\beta_1}$

- $Y_i$ for **Treatment** group **after**: $\hat{\beta_0}+\hat{\beta_1}+\hat{\beta_2}+\hat{\beta_3}$

]

---

# Visualizing Diff-in-Diff II

$$\hat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$

.pull-left[

```{r, fig.height=6, fig.retina=3}
ggplot(data = tibble(x=1,3))+
  aes(x = x)+
  # control
  geom_point(data = control, aes(x=x,y=y), size = 3, color = "red")+
  geom_text(x=1,y=0.5, color = "red", label=expression(C[1]))+
  geom_text(x=2,y=3.5, color = "red", label=expression(C[2]))+

  # line
  geom_segment(x=1, y=1, xend=2, yend=3, color="red", size = 2)+

  # beta 0 
  geom_label(x=0.9,y=1, color = "black", label=expression(hat(beta)[0]))+

  # line slope
  geom_segment(x=1, y=1, xend=2, yend=1, color="red", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=1, xend=2, yend=3, color="red", linetype = "dotted", size = 1)+
  # beta 1+2
  geom_label(x=2.2,y=3, color = "black", label=expression(hat(beta)[0]+hat(beta)[2]))+

  # treatment
  geom_point(data = treatment, aes(x=x,y=y), size = 3, color = "green")+
  geom_text(x=1,y=4.5, color = "green", label=expression(T[1]))+
  geom_text(x=2,y=8.5, color = "green", label=expression(T[2]))+

    # line
  geom_segment(x=1, y=4, xend=2, yend=8, color="green", size = 2)+

  # beta 0 plus beta 1 
  geom_label(x=0.8,y=4, color = "black", label=expression(hat(beta)[0]+hat(beta)[1]))+

  # line slope
  geom_segment(x=1, y=4, xend=2, yend=4, color="green", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  # diag
  geom_segment(x=1, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  geom_segment(x=2, y=6, xend=2, yend=8, color="blue", linetype = "dashed", size = 1)+

  # beta 3
  geom_label(x=2.25,y=8, color = "black",
             label=expression(hat(beta)[0]+hat(beta)[1]+hat(beta)[2]+hat(beta)[3]))+

  # beta 1 dif
  annotate("segment", x = 1, xend = 1, y = 1, yend = 4, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="both", type="closed"))+
  geom_label(x=1,y=2.5, size = 5, color = "black", label=expression(hat(beta)[1]))+

    scale_x_continuous(breaks=c(1,2),
                  labels=c(expression(t[before]),expression(t[after])),
                  limits=c(0.5,2.5))+
  scale_y_continuous(breaks=NULL,
                     limits=c(0,9))+
  labs(x = "Time",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

- $Y_i$ for **Control** group **before**: $\hat{\beta_0}$

- $Y_i$ for **Control** group **after**: $\hat{\beta_0}+\hat{\beta_2}$

- $Y_i$ for **Treatment** group **before**: $\hat{\beta_0}+\hat{\beta_1}$

- $Y_i$ for **Treatment** group **after**: $\hat{\beta_0}+\hat{\beta_1}+\hat{\beta_2}+\hat{\beta_3}$

- **Group Difference (before)**: $\hat{\beta_1}$

]

---

# Visualizing Diff-in-Diff II

$$\hat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$

.pull-left[

```{r, fig.height=6, fig.retina=3}
ggplot(data = tibble(x=1,3))+
  aes(x = x)+
  # control
  geom_point(data = control, aes(x=x,y=y), size = 3, color = "red")+
  geom_text(x=1,y=0.5, color = "red", label=expression(C[1]))+
  geom_text(x=2,y=3.5, color = "red", label=expression(C[2]))+

  # line
  geom_segment(x=1, y=1, xend=2, yend=3, color="red", size = 2)+

  # beta 0 
  geom_label(x=0.9,y=1, color = "black", label=expression(hat(beta)[0]))+

  # line slope
  geom_segment(x=1, y=1, xend=2, yend=1, color="red", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=1, xend=2, yend=3, color="red", linetype = "dotted", size = 1)+
  # beta 1+2
  geom_label(x=2.2,y=3, color = "black", label=expression(hat(beta)[0]+hat(beta)[2]))+

  # treatment
  geom_point(data = treatment, aes(x=x,y=y), size = 3, color = "green")+
  geom_text(x=1,y=4.5, color = "green", label=expression(T[1]))+
  geom_text(x=2,y=8.5, color = "green", label=expression(T[2]))+

    # line
  geom_segment(x=1, y=4, xend=2, yend=8, color="green", size = 2)+

  # beta 0 plus beta 1 
  geom_label(x=0.8,y=4, color = "black", label=expression(hat(beta)[0]+hat(beta)[1]))+

  # line slope
  geom_segment(x=1, y=4, xend=2, yend=4, color="green", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  # diag
  geom_segment(x=1, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  geom_segment(x=2, y=6, xend=2, yend=8, color="blue", linetype = "dashed", size = 1)+

  # beta 3
  geom_label(x=2.25,y=8, color = "black",
             label=expression(hat(beta)[0]+hat(beta)[1]+hat(beta)[2]+hat(beta)[3]))+

  # beta 1 dif
  annotate("segment", x = 1, xend = 1, y = 1, yend = 4, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="both", type="closed"))+
  geom_label(x=1,y=2.5, size = 5, color = "black", label=expression(hat(beta)[1]))+

  # beta 2 dif
  annotate("segment", x = 2, xend = 2, y = 1, yend = 3, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="both", type="closed"))+
  geom_label(x=2,y=2, size = 5, color = "black", label=expression(hat(beta)[2]))+

  # beta 2 dif
  annotate("segment", x = 2, xend = 2, y = 4, yend = 6, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="both", type="closed"))+
  geom_label(x=2,y=5, size = 5, color = "black", label=expression(hat(beta)[2]))+

    scale_x_continuous(breaks=c(1,2),
                  labels=c(expression(t[before]),expression(t[after])),
                  limits=c(0.5,2.5))+
  scale_y_continuous(breaks=NULL,
                     limits=c(0,9))+
  labs(x = "Time",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

- $Y_i$ for **Control** group **before**: $\hat{\beta_0}$

- $Y_i$ for **Control** group **after**: $\hat{\beta_0}+\hat{\beta_2}$

- $Y_i$ for **Treatment** group **before**: $\hat{\beta_0}+\hat{\beta_1}$

- $Y_i$ for **Treatment** group **after**: $\hat{\beta_0}+\hat{\beta_1}+\hat{\beta_2}+\hat{\beta_3}$

- **Group Difference (before)**: $\hat{\beta_1}$

- **Time Difference**: $\hat{\beta_2}$

]

---

# Visualizing Diff-in-Diff II

$$\hat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$

.pull-left[

```{r, fig.height=6, fig.retina=3}
ggplot(data = tibble(x=1,3))+
  aes(x = x)+
  # control
  geom_point(data = control, aes(x=x,y=y), size = 3, color = "red")+
  geom_text(x=1,y=0.5, color = "red", label=expression(C[1]))+
  geom_text(x=2,y=3.5, color = "red", label=expression(C[2]))+

  # line
  geom_segment(x=1, y=1, xend=2, yend=3, color="red", size = 2)+

  # beta 0 
  geom_label(x=0.9,y=1, color = "black", label=expression(hat(beta)[0]))+

  # line slope
  geom_segment(x=1, y=1, xend=2, yend=1, color="red", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=1, xend=2, yend=3, color="red", linetype = "dotted", size = 1)+
  # beta 1+2
  geom_label(x=2.2,y=3, color = "black", label=expression(hat(beta)[0]+hat(beta)[2]))+

  # treatment
  geom_point(data = treatment, aes(x=x,y=y), size = 3, color = "green")+
  geom_text(x=1,y=4.5, color = "green", label=expression(T[1]))+
  geom_text(x=2,y=8.5, color = "green", label=expression(T[2]))+

    # line
  geom_segment(x=1, y=4, xend=2, yend=8, color="green", size = 2)+

  # beta 0 plus beta 1 
  geom_label(x=0.8,y=4, color = "black", label=expression(hat(beta)[0]+hat(beta)[1]))+

  # line slope
  geom_segment(x=1, y=4, xend=2, yend=4, color="green", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  # diag
  geom_segment(x=1, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  geom_segment(x=2, y=6, xend=2, yend=8, color="blue", linetype = "dashed", size = 1)+

  # beta 3
  geom_label(x=2.25,y=8, color = "black",
             label=expression(hat(beta)[0]+hat(beta)[1]+hat(beta)[2]+hat(beta)[3]))+

  # beta 1 dif
  annotate("segment", x = 1, xend = 1, y = 1, yend = 4, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="both", type="closed"))+
  geom_label(x=1,y=2.5, size = 5, color = "black", label=expression(hat(beta)[1]))+

  # beta 2 dif
  annotate("segment", x = 2, xend = 2, y = 1, yend = 3, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="both", type="closed"))+
  geom_label(x=2,y=2, size = 5, color = "black", label=expression(hat(beta)[2]))+

  # beta 2 dif
  annotate("segment", x = 2, xend = 2, y = 4, yend = 6, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="both", type="closed"))+
  geom_label(x=2,y=5, size = 5, color = "black", label=expression(hat(beta)[2]))+

  # beta 2 dif
  annotate("segment", x = 2, xend = 2, y = 6, yend = 8, colour = "blue", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="both", type="closed"))+
  geom_label(x=2,y=7, size = 5, color = "blue", label=expression(hat(beta)[3]))+

    scale_x_continuous(breaks=c(1,2),
                  labels=c(expression(t[before]),expression(t[after])),
                  limits=c(0.5,2.5))+
  scale_y_continuous(breaks=NULL,
                     limits=c(0,9))+
  labs(x = "Time",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

- $Y_i$ for **Control** group **before**: $\hat{\beta_0}$

- $Y_i$ for **Control** group **after**: $\hat{\beta_0}+\hat{\beta_2}$

- $Y_i$ for **Treatment** group **before**: $\hat{\beta_0}+\hat{\beta_1}$

- $Y_i$ for **Treatment** group **after**: $\hat{\beta_0}+\hat{\beta_1}+\hat{\beta_2}+\hat{\beta_3}$

- **Group Difference (before)**: $\hat{\beta_1}$

- **Time Difference**: $\hat{\beta_2}$

- .blue[**Difference-in-difference**: `\\(\hat{\beta_3}\\)` (treatment effect)]
]

---

# Comparing Group Means (Again)

$$\hat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$

|    | Control | Treatment | **Group Diff** $(\Delta Y_i)$ |
|----|---------|-----------|---------------------------|
| Before | $\beta_0$ | $\beta_0+\beta_1$ | $\beta_1$ |
| After | $\beta_0+\beta_2$ | $\beta_0+\beta_1+\beta_2+\beta_3$ | $\beta_1+\beta_3$ |
| **Time Diff** $(\Delta Y_t)$ | $\beta_2$ | $\beta_2+\beta_3$ | **Diff-in-diff** $\Delta_i \Delta_t: \beta_3$ |

---

# Key Assumption: Counterfactual

$$\hat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$

.pull-left[

```{r, fig.height=6, fig.retina=3}
ggplot(data = tibble(x=1,3))+
  aes(x = x)+
  # control
  geom_point(data = control, aes(x=x,y=y), size = 3, color = "red")+
  geom_text(x=1,y=0.5, color = "red", label=expression(C[1]))+
  geom_text(x=2,y=3.5, color = "red", label=expression(C[2]))+

  # line
  geom_segment(x=1, y=1, xend=2, yend=3, color="red", size = 2)+

  # line slope
  geom_segment(x=1, y=1, xend=2, yend=1, color="red", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=1, xend=2, yend=3, color="red", linetype = "dotted", size = 1)+

  # treatment
  geom_point(data = treatment, aes(x=x,y=y), size = 3, color = "green")+
  geom_text(x=1,y=4.5, color = "green", label=expression(T[1]))+
  geom_text(x=2,y=8.5, color = "green", label=expression(T[2]))+

    # line
  geom_segment(x=1, y=4, xend=2, yend=8, color="green", size = 2)+

  # line slope
  geom_segment(x=1, y=4, xend=2, yend=4, color="green", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  # diag
  geom_segment(x=1, y=4, xend=2, yend=6, color="red", linetype = "dotted", size = 1)+

  geom_segment(x=2, y=6, xend=2, yend=8, color="blue", linetype = "dashed", size = 1)+

  scale_x_continuous(breaks=c(1,2),
                  labels=c(expression(t[before]),expression(t[after])),
                  limits=c(0.5,2.5))+
  scale_y_continuous(breaks=NULL,
                     limits=c(0,9))+
  labs(x = "Time",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

- Key assumption for DND: **time trends** (for treatment and control) are **parallel**

- Treatment and control groups assumed to be identical over time on average, **except for treatment**

- .shout[Counterfactual]: if the treatment group had not recieved treatment, it would have changed identically over time as the control group $(\hat{\beta_2})$

]

---

# Key Assumption: Counterfactual

$$\hat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$

.pull-left[

```{r, fig.height=6, fig.retina=3}
ggplot(data = tibble(x=1,3))+
  aes(x = x)+
  # control
  geom_point(data = control, aes(x=x,y=y), size = 3, color = "red")+
  geom_text(x=1,y=0.5, color = "red", label=expression(C[1]))+
  geom_text(x=2,y=3.5, color = "red", label=expression(C[2]))+

  # line
  geom_segment(x=1, y=1, xend=2, yend=3, color="red", size = 2)+

  # line slope
  geom_segment(x=1, y=1, xend=2, yend=1, color="red", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=1, xend=2, yend=3, color="red", linetype = "dotted", size = 1)+

  # treatment
  geom_point(data = treatment, aes(x=x,y=y), size = 3, color = "green")+
  geom_text(x=1,y=4.5, color = "green", label=expression(T[1]))+
  geom_text(x=2,y=8.5, color = "green", label=expression(T[2]))+

    # line
  geom_segment(x=1, y=4, xend=2, yend=8, color="green", size = 2)+

  # line slope
  geom_segment(x=1, y=4, xend=2, yend=4, color="green", linetype = "dotted", size = 1)+
  geom_segment(x=2, y=4, xend=2, yend=4.5, color="purple", linetype = "dotted", size = 1)+

  # diag
  geom_segment(x=1, y=4, xend=2, yend=4.5, color="purple", linetype = "dotted", size = 1)+

  geom_segment(x=2, y=4.5, xend=2, yend=8, color="blue", linetype = "dashed", size = 1)+

  scale_x_continuous(breaks=c(1,2),
                  labels=c(expression(t[before]),expression(t[after])),
                  limits=c(0.5,2.5))+
  scale_y_continuous(breaks=NULL,
                     limits=c(0,9))+
  labs(x = "Time",
       y = "Y")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[

- If the time-trends would have been *different*, a **biased** measure of the treatment effect $(\hat{\beta_3})$!

]

---

# Diff-in-Diff Example I

.content-box-green[
.green[**Example**]: In 1993 Georgia initiated a HOPE scholarship program to let state residents with at least a B average in high school attend public college in Georgia for free. Did it increase college enrollment?
]

- Micro-level data on 4,291 young individuals

--

- $\text{InCollege}_{it}=\begin{cases}1 \text{ if } i \text{ is in college during year }t\\ 0 \text{ if } i \text{ is not in college during year }t\\ \end{cases}$

--

- $\text{Georgia}_i=\begin{cases}1 \text{ if } i \text{ is a Georgia resident}\\ 0 \text{ if } i \text{ is not a Georgia resident}\\ \end{cases}$

--

- $\text{After}_t=\begin{cases}1 \text{ if } t \text{ is after 1992}\\ 0 \text{ if } t \text{ is after 1992}\\ \end{cases}$

.source[Dynarski, Susan (2000), "Hope for Whom? Financial Aid for the Middle Class and Its Impact on College Attendance"]

.footnote[Note: With a dummy *dependent* (Y) variable, coefficients estimate the probability `\\(Y=1\\)`, i.e. the *probability* a person is enrolled in college]

---

# Diff-in-Diff Example II

- We can use a DND model to measure the effect of HOPE scholarship on enrollments

- Georgia and nearby States, if not for HOPE, changes should be the same over time

--

- Treatment period: after 1992

--

- Treatment: Georgia

--

- Differences-in-differences: $$\Delta_i \Delta_t Enrolled = (\text{GA}_{after}-\text{GA}_{before})-(\text{neighbors}_{after}-\text{neighbors}_{before})$$

--

- Regression equation:
$$\widehat{Enrolled_{it}} = \beta_0+\beta_1 \, Georgia_{it}+\beta_2 \,  After_{it}+\beta_3 \, (Georgia_{it} \times After_{it})$$

---

# Example: Regression

```{r}
hope<-read_csv("https://metricsf19.classes.ryansafner.com/data/HOPE.csv")
hope<-hope %>%
  mutate_at(c("StateCode", "Year"),factor)
```

.pull-left[

.font70[
```{r, echo=T}
DND_reg<-lm(InCollege ~ Georgia + After + Georgia:After, data = hope)
summary(DND_reg)
```
]

--

.pull-right[
$$\widehat{Enrolled_{it}}=0.406-0.105 \, Georgia_{i}-0.004 \, After_{t}+0.089 \, (Georgia_{i} \times After_{t})$$
]

---

# Example: Interpretting the Regression

$$\widehat{Enrolled_{it}}=0.406-0.105 \, Georgia_{i}-0.004 \, After_{t}+0.089 \, (Georgia_{i} \times After_{t})$$

--

- $\beta_0$: 
--
A **non-Georgian** **before** 1992 was 40.6% likely to be a college student

--

- $\beta_1$:
--
**Georgians** **before** 1992 were 10.5% less likely to be college students than neighboring states

--

- $\beta_2$:
--
**After** 1992, **non-Georgians** are 0.4% less likely to be college students

--

- $\beta_3$:
--
**After** 1992, **Georgians** are 8.9% more likely to enroll in colleges than neighboring states

--

- .onfire[Treatment effect: HOPE increased enrollment likelihood by 8.9%]

---

# Example: Comparing Group Means

$$\widehat{Enrolled_{it}}=0.406-0.105 \, Georgia_{i}-0.004 \, After_{t}+0.089 \, (Georgia_{i} \times After_{t})$$

- A group mean for a dummy $Y$ is $E[Y=1]$, i.e. the probability a student is enrolled:

- **Non-Georgian enrollment probability pre-1992**:
--
$\beta_0=0.406$

--
- **Georgian enrollment probability pre-1992**:
--
$\beta_0+\beta_1=0.406-0.105=0.301$

--

- **Non-Georgian enrollment probability post-1992**:
--
$\beta_0+\beta_2=0.406-0.004=0.402$

--

- **Georgian enrollment probability post-1992**:
--
$\beta_0+\beta_1+\beta_2+\beta_3=0.406-0.105-0.004+0.089=0.386$

---

# Example: Comparing Group Means in R

.pull-left[
```{r, echo=T}
# group mean for non-Georgian before 1992
hope %>%
  filter(Georgia==0,
         After==0) %>%
  summarize(prob = mean(InCollege))

# group mean for Georgian before 1992
hope %>%
  filter(Georgia==1,
         After==0) %>%
  summarize(prob = mean(InCollege))
```
]

--

.pull-right[
```{r, echo=T}
# group mean for non-Georgian AFTER 1992
hope %>%
  filter(Georgia==0,
         After==1) %>%
  summarize(prob = mean(InCollege))

# group mean for Georgian AFTER 1992
hope %>%
  filter(Georgia==1,
         After==1) %>%
  summarize(prob = mean(InCollege))
```
]

---

# Example: Diff-in-Diff Summary

$$\widehat{Enrolled_{it}}=0.406-0.105 \, Georgia_{i}-0.004 \, After_{t}+0.089 \, (Georgia_{i} \times After_{t})$$

|    | Neighbors | Georgia | **Group Diff** $(\Delta Y_i)$ |
|----|---------|-----------|---------------------------|
| Before | $0.406$ | $0.301$ | $-0.105$ |
| After | $0.402$ | $0.386$ | $0.016$ |
| **Time Diff** $(\Delta Y_t)$ | $-0.004$ | $0.085$ | **Diff-in-diff** $0.089$ |

--

$$\begin{align*}
	\Delta_i \Delta_t Enrolled &= (\text{GA}_{after}-\text{GA}_{before})-(\text{neighbors}_{after}-\text{neighbors}_{before})\\
	&=(0.386-0.301)-(0.402-0.406)\\
	&=(0.085)-(-0.004)\\
	&=0.089\\ \end{align*}$$

---

# Example: Diff-in-Diff Graph

```{r, fig.retina=3, fig.align="center", fig.width=16}
hope %>%
  ggplot(data=.)+
  aes(x = After,
      y = InCollege,
      color = factor(Georgia))+
  geom_smooth(method="lm", se=FALSE)+
  #scale_y_continuous(breaks=seq(0,1,0.125), limits=c(0,0.05))+
  scale_x_continuous(breaks=seq(0,1,1), labels=c("Before", "After"))+
scale_color_manual(name = "State",
                   labels = c("Neighbors", "Georgia"), values = c("blue", "red"))+
  labs(x = "Before or After HOPE",
       y = "Probability of Being Enrolled in College")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=16)+
  theme(legend.position = "top")
```

---

class: inverse. center, middle

# Generalizing DND Models

---

# Generalizing DND Models

- DND can be **generalized** with a **two-way fixed effects** model:  	
$$\widehat{Y_{it}}=\alpha_i+\theta_t+\beta_3 (\text{Treated}_i * \text{After}_{it})+\nu_{it}$$
    - $\alpha_i$: **group fixed effects** (treatments/control groups)
    - $\theta_t$: **time fixed effects** (pre/post treatment)

--

- Allows *many* periods, and treatment(s) can occur at different times to different units (so long as some do not get treated)

--

- Can also add control variables that vary within units and over time 
$$\widehat{Y_{it}}=\alpha_i+\theta_t+\beta_3 \, (\text{Treated}_i \times \text{After}_{it})+\beta_4 X_{it}+\nu_{it}$$

---

# Our Example, Generalized I

$$\widehat{Enrolled_{it}} = \alpha_i+\theta_t+\beta_3 \, (Georgia_{it} \times After_{it})$$

- `StateCode` is a variable for the State $\implies$ create State fixed effect

- `Year` is a variable for the year $\implies$ create year fixed effect

- Using LSDV method...

```{r dnd-gen-reg, echo=T, eval=F}
DND_fe<-lm(InCollege ~ Georgia:After + factor(StateCode) + factor(Year),
           data = hope)
summary(DND_fe)
```

---

# Our Example, Generalized II

```{css, echo = FALSE}
.smallcode .remark-code { /*Change made here*/
  font-size: 55% !important;
}
```

.pull-left[
.smallcode[
```{r, ref.label="dnd-gen-reg"}

```
]
]

--

.pull-right[
$$\widehat{InCollege_{it}}=\alpha_i+\theta_t+0.091 \, (\text{Georgia}_i \times \text{After}_{it})$$

]

---

# Intuition behind DND

- Diff-in-diff models are the quintessential example of exploiting .shout[natural experiments]

- A major change at a point in time (change in law, a natural disaster, political crisis) separates groups where one is affected and another is not---identifies the effect of the change (treatment)

- One of the cleanest and clearest causal **identification strategies**

---

class: inverse, center, middle

# Example: "The" Card-Kreuger Minimum Wage Study

---

# Example: "The" Card-Kreuger Minimum Wage Study I

.content-box-green[
.green[**Example**]: *The* controversial minimum wage study, Card & Kreuger (1994) is a quintessential (and clever) diff-in-diff.
]

.source[Card, David, Krueger, Alan B, (1994), "Minimum Wages and Employment: A Case Study of the Fast-Food Industry in New Jersey and Pennsylvania," *American Economic Review* 84 (4): 772–793	]

---

# Card & Kreuger (1994): Background I

.pull-left[
- Card & Kreuger (1994)	compare employment in fast food restaurants on New Jersey and Pennsylvania sides of border between February and November 1992.

- Pennsylvania & New Jersey both had a minimum wage of $4.25 before February 1992

- In February 1992, New Jersey raised minimum wage from $4.25 to $5.05
]

.pull-right[
.center[
![](https://www.dropbox.com/s/gdoip37e26j253b/PANJ.png?raw=1)
]
]

---

# Card & Kreuger (1994): Background II

.pull-left[
- If we look only at New Jersey before & after change: 
  - **Omitted variable bias**: macroeconomic variables (there's a recession!), weather, etc.
	- Including PA as a control will control for these time-varying effects if they are national trends

- Surveyed 400 fast food restaurants on each side of the border, before & after min wage increase
  - .onfire[Key assumption]: Pennsylvania and New Jersey follow parallel trends, 
	- **Counterfactual**: if not for the minimum wage increase, NJ employment would have changed similar to PA employment
]

.pull-right[
.center[
![](https://www.dropbox.com/s/gdoip37e26j253b/PANJ.png?raw=1)
]
]

---

# Card & Kreuger (1994): Comparisons

.center[
![:scale 25%](https://www.dropbox.com/s/d0z7vb9so296eun/CardKreuger1.png?raw=1)
]

---

# Card & Kreuger (1994): Summary I

.center[
![:scale 65%](https://www.dropbox.com/s/9igiex5ly8u67vq/CardKreuger2.png?raw=1)
]

---

# Card & Kreuger (1994): Summary II

.center[
![:scale 65%](https://www.dropbox.com/s/gj4spkyz3jkop2z/CardKreuger3.png?raw=1)
]

---

# Card & Kreuger (1994): Identification Strategy and Model

$$\widehat{Employment_{i t}}=\beta_0+\beta_1 \, NJ_{i}+\beta_2 \, After_{t}+\beta_3 \, (NJ_i \times After_t)$$

- PA Before: $\beta_0$

- PA After: $\beta_0+\beta_2$

- NJ Before: $\beta_0+\beta_1$

- NJ After: $\beta_0+\beta_1+\beta_2+\beta_3$

- **Diff-in-diff**: $(NJ_{after}-NJ_{before})-(PA_{after}-PA_{before})$

--

|    | PA | NJ | **Group Diff** $(\Delta Y_i)$ |
|----|---------|-----------|---------------------------|
| Before | $\beta_0$ | $\beta_0+\beta_1$ | $\beta_1$ |
| After | $\beta_0+\beta_2$ | $\beta_0+\beta_1+\beta_2+\beta_3$ | $\beta_1+\beta_3$ |
| **Time Diff** $(\Delta Y_t)$ | $\beta_2$ | $\beta_2+\beta_3$ | **Diff-in-diff** $\Delta_i \Delta_t: \beta_3$ |

---

# Card & Kreuger (1994): Results

.center[
![:scale 65%](https://www.dropbox.com/s/ms5gpn8lfkws6a9/CardKreuger5.png?raw=1)
]

### 2019-12-05 static/slides/23-slides.Rmd
Update lesson 23 content

@@
 ---
-title: "4.2: Differences-in-Differences Models"
+title: "4.2: Difference-in-Difference Models"
 subtitle: "ECON 480 · Econometrics · Fall 2019"
@@
 
+# Clever Research Designs Identify Causality
+
+```{r, fig.retina=3, fig.align="center", fig.width=13}
+ggplot(data = tibble(x=c(0,10),
+                     y=c(0,10)))+
+  aes(x = x,
+      y = y)+
+  geom_label(x=1,y=2, label="Correlation", size =10, color = "#fde0dd")+
+  geom_label(x=5,y=2, label="Causation", size = 10, color = "#7a0177")+
+
+  geom_label(x=1,y=4, label="Differences", size =5, color = "#fde0dd")+
+  geom_label(x=1,y=3.5, label="Pre-Post", size =5, color = "#fde0dd")+
+  geom_label(x=2,y=4.5, label="Multiple Regression", size =5, color = "#fbb4b9")+
+  geom_label(x=2,y=3.5, label="Matching", size =5, color = "#fbb4b9")+
+    geom_label(x=3,y=5, label="Fixed Effects", size =5, color = "#f768a1")+
+    geom_label(x=3.5,y=4.5, label="Diff-in-Diff", size =5, color = "#c51b8a")+
+  geom_label(x=4,y=3.5, label="Natural Experiments", size =5, color = "#c51b8a")+
+  geom_label(x=3.5,y=4, label="Regression Discontinuity", size =5, color = "#c51b8a")+
+  geom_label(x=5,y=4.5, label="RCTs", size =5, color = "#7a0177")+
+  annotate("segment", x = 1, xend = 5, y = 3, yend = 3, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+
+  scale_x_continuous(breaks=seq(0,6,1),
+                     limits=c(0,6))+
+  scale_y_continuous(breaks=seq(1,5,1),
+                     limits=c(1,5))+
+  theme_void()
+
+```
+
+Again, this toolkit of research designs to identify causal effects is the economist's comparative advantage that firms and governments want!
+
+---
+
 class: inverse, center, middle
@@
 
+.pull-left[
+
 - Often, we want to examine the consequences of a change, such as a law or policy
-    - e.g. States that implemented law $X$ saw a change in $Y$
+
+]
+
+---
+
+# Difference-in-Difference Models I
+
+.pull-left[
+
+- Often, we want to examine the consequences of a change, such as a law or policy
+
+- .green[**Example**]: how do States that implement law $X$ see changes in $Y$
   - **Treatment**: States that implement law $X$
   - **Control**: States that did not implement law $X$
-    - If we have **panel data** with observations for all states before *and* after the change:
-- Simple logic: compare difference in outcomes of treatment group (before and after treatment) with those of non-treated group (before and after same treatment period)
-
----
-
-# Difference-in-Difference Models II
-
+
+]
+
+---
+
+# Difference-in-Difference Models I
+
+.pull-left[
+
+- Often, we want to examine the consequences of a change, such as a law or policy
+
+- .green[**Example**]: how do States that implement law $X$ see changes in $Y$
+  - **Treatment**: States that implement law $X$
+  - **Control**: States that did not implement law $X$
+
+- If we have **panel data** with observations for all states before *and* after the change...
+
+- Find the *difference* between treatment & control groups *in* their *differences* before and after the treatment period
+
+]
+
+--
+
+.pull-right[
 .center[
@@
 ]
-
----
-
-# Difference-in-Difference Models III
-
-- The .shout[difference-in-difference model] (aka .shout["diff-in-diff"] or .shout[DND"]) identifies treatment effect by differencing the difference pre- and post-treatment between treatment and control groups
+]
+
+---
+
+# Difference-in-Difference Models II
+
+- The .shout[difference-in-difference model] (aka .shout["diff-in-diff"] or .shout[DND"]) identifies treatment effect by differencing the difference pre- and post-treatment values of $Y$ between treatment and control groups
 
@@
 
-class: inverse. center, middle
+class: inverse, center, middle
 

### 2019-12-05 static/slides/23-slides.Rmd
Final tweak for slides 23

@@
 
-```{r, fig.retina=3, fig.align="center", fig.width=13}
+Again, **this toolkit** of research designs to **identify causal effects** is the economist's **comparative advantage** that firms and governments want!
+
+```{r, fig.retina=3, fig.align="center", fig.width=12}
 ggplot(data = tibble(x=c(0,10),
@@
 
-Again, this toolkit of research designs to identify causal effects is the economist's comparative advantage that firms and governments want!
-
 ---
@@
 - Regression equation:
-$$\widehat{Enrolled_{it}} = \beta_0+\beta_1 \, Georgia_{it}+\beta_2 \,  After_{it}+\beta_3 \, (Georgia_{it} \times After_{it})$$
+$$\widehat{Enrolled_{it}} = \beta_0+\beta_1 \, Georgia_{i}+\beta_2 \,  After_{t}+\beta_3 \, (Georgia_{i} \times After_{t})$$
 
@@
 ```
+]
 ]
@@
 
-# group mean for Georgian before 1992
-hope %>%
-  filter(Georgia==1,
-         After==0) %>%
-  summarize(prob = mean(InCollege))
 ```
@@
 
+```
+]
+
+---
+
+# Example: Comparing Group Means in R II
+
+.pull-left[
+```{r, echo=T}
+# group mean for Georgian before 1992
+hope %>%
+  filter(Georgia==1,
+         After==0) %>%
+  summarize(prob = mean(InCollege))
+```
+]
+
+--
+
+.pull-right[
+```{r, echo=T}
 # group mean for Georgian AFTER 1992
@@
 |    | Neighbors | Georgia | **Group Diff** $(\Delta Y_i)$ |
-|----|---------|-----------|---------------------------|
+|----|--------:|----------:|--------------------------:|
 | Before | $0.406$ | $0.301$ | $-0.105$ |
 | After | $0.402$ | $0.386$ | $0.016$ |
-| **Time Diff** $(\Delta Y_t)$ | $-0.004$ | $0.085$ | **Diff-in-diff** $0.089$ |
+| **Time Diff** $(\Delta Y_t)$ | $-0.004$ | $0.085$ | **Diff-in-diff**: $0.089$ |
 
@@
 - DND can be **generalized** with a **two-way fixed effects** model:  	
-$$\widehat{Y_{it}}=\alpha_i+\theta_t+\beta_3 (\text{Treated}_i * \text{After}_{it})+\nu_{it}$$
+$$\widehat{Y_{it}}=\alpha_i+\theta_t+\beta_3 (\text{Treated}_i * \text{After}_{t})+\nu_{it}$$
     - $\alpha_i$: **group fixed effects** (treatments/control groups)
@@
 - Can also add control variables that vary within units and over time 
-$$\widehat{Y_{it}}=\alpha_i+\theta_t+\beta_3 \, (\text{Treated}_i \times \text{After}_{it})+\beta_4 X_{it}+\nu_{it}$$
+$$\widehat{Y_{it}}=\alpha_i+\theta_t+\beta_3 \, (\text{Treated}_i \times \text{After}_{t})+\beta_4 X_{it}+\nu_{it}$$
 

### 2019-12-06 content/assignment/02-exam.Rmd
Add HW 5 answers, HW 6 and answers, finalize schedule page (put extra content after final), add final exam prep

--- 2019-10-10 content/assignment/01-exam.Rmd
@@
 ---
-title: "Midterm Exam"
+title: "Final Exam"
 editor_options: 
@@
 
-*Thursday, October 17*
+*Thursday, December 12, 11:30 AM -2:00 PM*
 
@@
 
-- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/oldexams/midtermconcepts.pdf)
+- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/oldexams/finalconcepts.pdf)
+- [<i class="fas fa-external-link-square-alt"></i> Review Questions Slides (TBD)](/slides/final-review.html)
 
@@
 
-- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/oldexams/01-old-exam.pdf)
+- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/oldexams/02-old-exam.pdf)
 
@@
 
-- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/oldexams/midtermformulas.pdf)
+- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/oldexams/finalformulas.pdf)
 

### 2019-12-06 content/assignment/05-problem-set.Rmd
Add R Practice 21 answers
Add HW 5 answers, HW 6 and answers, finalize schedule page (put extra content after final), add final exam prep

@@
 - [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/05-problem-set.Rmd)
+
+**ANSWERS:**
+
+- [<i class="fas fa-file-pdf"></i> `html`](/homeworks/05-problem-set-answers.html)
+- [<i class="fab fa-markdown"></i> `Markdown`](/homeworks/05-problem-set-answers.Rmd)
+- [<i class="fas fa-registered"></i> `R Script`](/homeworks/05-problem-set-answers.R)
 
@@
 ### Part A
-Using the group means, write a regression equation for a regression of GDP Growth rate on Common Law, including the standard errors in parentheses below the coefficients. Define
+Using the group means, write a regression equation for a regression of GDP Growth rate on Common Law. Define
 

### 2019-12-06 content/assignment/06-problem-set.Rmd
Add HW 5 answers, HW 6 and answers, finalize schedule page (put extra content after final), add final exam prep

---
title: "Problem Set 6"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

**UNGRADED: For Practice Only!**

- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/06-problem-set-pdf.pdf)
- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/06-problem-set.Rmd)

**ANSWERS:**

- [<i class="fas fa-file-pdf"></i> `html`](/homeworks/06-problem-set-answers.html)
- [<i class="fab fa-markdown"></i> `Markdown`](/homeworks/06-problem-set-answers.Rmd)

# Instructions

For this problem set, you may submit handwritten answers on a plain sheet of paper, or download and type/handwrite on the PDF.

Alternatively, you may download the `.Rmd` file, do the homework in markdown, and email to me a single `knit`ted `html` or `pdf` file (and be sure that it shows all of your code).

You may work together (and I highly encourage that) but you must turn in your own answers. I grade homeworks 70% for completion, and for the remaining 30%, pick one question to grade for accuracy - so it is best that you try every problem, even if you are unsure how to complete it accurately.

# Theory and Concepts

## Question 1
In your own words, describe what *fixed effects* are, when we can use them, and how they remove endogeneity.

## Question 2
In your own words, describe the logic of a *difference-in-difference* model: what is it comparing against what, and how does it estimate the effect of treatment? What assumption must be made about the treatment and control group for the model to be valid?

# R Questions

Answer the following questions using `R`. When necessary, please write answers in the same document (knitted `Rmd` to `html` or `pdf`, typed `.doc(x)`, or handwritten) as your answers to the above questions. Be sure to include (email or print an `.R` file, or show in your knitted `markdown`) your code and the outputs of your code with the rest of your answers.

## Question 3

- [<i class="fas fa-table"></i> `PeaceCorps.csv`](http://metricsf19.classes.ryansafner.com/data/PeacceCorps.csv)

How do people respond to changes in economic conditions? Are they more likely to pursue public service when private sector jobs are scarce? This dataset contains variables at the U.S. State (& D.C.) level:

| Variable | Description |
|----------|-------------|
| `state` | U.S. State |
| `year` | Year |
| `apps` | Applications to the Peace Corps (per capita) in State |
| `unemployrate` | State unemployment rate |

> Do more people apply to the Peace Corps when unemployment increases (and reduces other opportunities)?

### Part A

Before looking at the data, what does your economic intuition tell you? Explain your hypothesis.

### Part B

To get the hang of the data we're working with, `count` (separately) the number of `state`s, and the number of `year`s. Get the number of `n_distinct()` `state`s and `year`s^[Do this inside the `summarize()` command], as well as the `distinct()` values of each^[Don't use the `summarize()` command for this part].

### Part C

Continuing our pre-analysis inspection, (install, and) load the `plm` package, and check the dimensions of the data with `pdim`.^[Set `index=c("state","year")` to indicate the group and time dimensions.]

### Part D

Create a scatterplot of `appspc` (Y) on `unemployrate` (X). Which State is an outlier? How would this affect the pooled regression estimates? Create a *second* scatterplot that does not include this State.

### Part E

Run two *pooled* regressions, one with the outliers, and one without them. Write out the estimated regression equation for each. Interpret the coefficient, and comment on how it changes between the two regressions.

### Part F

Now run a regression with State fixed effects using the dummy variable method.^[Ensure that `state` is a factor variable, and insert in the regression. You can either `mutate()` it into a `factor` beforehand, or just do `as.factor(state)` in the `lm` command.] Interpret the marginal effect of `unemployrate` on `appspc`. How did it change?

### Part G

Find the coefficient for Maryland and interpret it. How many applications per capita does Maryland have?

### Part H
Now try using the `plm()` command, which de-means the data, and make sure you get the same results as Part F.^[Inside `plm()`, set `index = "state"` to indicate variable, and `model = "within"` to indicate a fixed effects model.] Do you get the same marginal effect of `unemployrate` on `appspc`?

### Part I

Now include *year* fixed effects in your regression, using the dummy variable method. Interpret the marginal effect of `unemployrate` on `appspc`. How did it change?

### Part J

What would be the predicted number of applications in Maryland in 2011 at an unemployment rate of 5%?

### Part K

Now try using the `plm()` command, which de-means the data, and make sure you get the same results as Part I.^[Inside `plm()`, set `index = c("state", "year")` to indicate both variables, and `effect = "twoways"` to indicate a 2-way fixed effects model.] Do you get the same marginal effect of `unemployrate` on `appspc`?

### Part L

Can there still be endogeneity in this model? Give some examples.

### Part M

Create a nice regression table (using `huxtable`) for comparison of the regressions in E, G, and I.^[It's better to use the `plm()`-generated regressions so as to avoid the multitude of coefficients for the state and year dummies! You certainly could use the dummy variable ones and manually list all of the variables to suppress in the table inside `omit_coefs()`...]

## Question 4

- [<i class="fas fa-table"></i> `TexasSchools.csv`](http://metricsf19.classes.ryansafner.com/data/TexasSchools.csv)

Are teachers paid more when school board members are elected "off cycle" when there are not major national political elections (e.g. odd years) than "on cycle?" The argument is that during "off" years, without attention on state or national elections, voters will pay less attention to the election, and teachers can more effectively mobilize for higher pay, versus "on" years where voters are paying more attention. This data comes from Anzia, Sarah (2012), "The Election Timing Effect: Evidence from a Policy Intervention in Texas." *Quarterly Journal of Political Science* 7(3): 277-297, and follows 1,020 Texas school board districts from 2003–2009.

From 2003–2006, all districts elected their school board members off-cycle. A change in Texas policy in 2006 led some, but not all, districts to elect their school board members on-cycle from 2007 onwards.

| Variable | Description |
|----------|-------------|
| `LnAvgSalary` | logged average salary of teachers in district |
| `Year` | Year |
| `OnCycle` | `=1` if school boards elected on-cycle (e.g. same year as national and state elections), `=0` if elected off-cycle |
| `pol_freedom` | Political freedom index score (2018) from 1 (least) top 10 (most free) |
| `CycleSwitch` | `=1` if district switched from off- to on-cycle elections |
| `AfterSwitch` | `=1` if year is after 2006 |

### Part A
Run a pooled regression model of `LnAvgSalary` on `OnCycle`. Write the estimated regression equation, and interpret the coefficient on `OnCycle.` Are there any sources of bias (consider in particular the argument in the question prompt)?

### Part B
Some schools decided to switch to an on-cycle election after 2006. Consider this, `CycleSwitch` the "treatment." Create a variable to indicate post-treatment years (i.e. years after 2006). Call it `After`.

### Part C
Now estimate a difference-in-difference model with your variables in Part B: `CycleSwitch` is the treatment variable, `After` is your post-treatment indicator, and add an *interaction* variable to capture the interaction effect between those districts that *switched*, and *after* the treatment. Write down the estimated regression equation (to four decimal places).

### Part D
Interpret what each coefficient means from Part C.

### Part E
Using your regression equation in Part C, calculate the expected logged average salary $(Y)$ for districts in Texas:

- i. *Before* the switch that did *not* switch
- ii. *After* the switch that did *not* switch
- iii. *Before* the switch that *did* switch
- iv. *After* the switch that *did* switch

### Part F
Confirm your estimates in Part E by finding the mean logged average salary for each of those four groups in the data.^[Hint, `filter()` properly then `summarize()`.]

### Part G
Write out the difference-in-difference equation, and calculate the difference-in-difference. Make sure it matches your estimate from the regression.

### Part H
Can we say anything about the types of districts that switched? Can we say anything about all salaries in the districts in the years after the switch?

### Part I
Now let’s generalize the diff-in-diff model. Instead of the treatment and post- treatment dummies, use district- and year-fixed effects and the interaction term. Interpret the coefficient on the interaction term.

### Part J
Create a nice regression table (using `huxtable`) for comparison of the regressions in (a), (c), and (i).

### 2019-12-06 content/class/21-class.Rmd
Add HW 5 answers, HW 6 and answers, finalize schedule page (put extra content after final), add final exam prep

@@
 |:----------:|:----------:|:-------:|
-| ![](21-slides_files/figure-html/unnamed-chunk-14-1.png) | ![](21-slides_files/figure-html/unnamed-chunk-20-1.png) | ![](21-slides_files/figure-html/unnamed-chunk-25-1.png) |
+| ![](https://metricsf19.classes.ryansafner.com/slides/21-slides_files/figure-html/unnamed-chunk-14-1.png) | ![](https://metricsf19.classes.ryansafner.com/slides/21-slides_files/figure-html/unnamed-chunk-20-1.png) | ![](https://metricsf19.classes.ryansafner.com/slides/21-slides_files/figure-html/unnamed-chunk-25-1.png) |
 | $\hat{Y_i}=\hat{\beta_0}+\hat{\beta_1}\mathbf{ln(X_i)}$ | $\mathbf{ln(\hat{Y_i})}=\hat{\beta_0}+\hat{\beta_1}X_i$ | $\mathbf{ln(\hat{Y_i})}=\hat{\beta_0}+\hat{\beta_1}\mathbf{ln(X_i)}$ |

### 2019-12-06 static/homeworks/05-problem-set-answers.Rmd
Add HW 5 answers, HW 6 and answers, finalize schedule page (put extra content after final), add final exam prep

---
title: "Problem Set 5 (Answers)"
author: "Ryan Safner" # put your name here! 
date: "ECON 480 - Fall 2019"
output:
  html_document:
    df_print: paged
    #theme: simplex
    toc: true 
    toc_depth: 3
    toc_float: true
    code_folding: show
    highlight: tango
---

<!--CLICK "KNIT" ABOVE TO RENDER TO HTML, PDF, OR WORD OUTPUT
In fact, try knitting right away and see what this produces!
-->

*Due by Thursday, November 21, 2019*

# Instructions

Answers may be longer than I would deem sufficient on an exam. Some might vary slightly based on points of interest, examples, or personal experience. These suggested answers are designed to give you both the answer and a short explanation of *why* it is the answer.

# Theory and Concepts

## Question 1
**In your own words, describe what the "dummy variable trap" means. What precisely is the problem, and what is the standard way to prevent it?**

---

<!--WRITE YOUR ANSWERS BELOW -->
The dummy variable trap is a specific form of perfect multicollinearity, which prevents an OLS regression from making logical sense. If two (or more) variables can be expressed as an exact linear function of one another, we cannot estimate a regression. In terms of dummy variables, we could have dummy variables for each category an observation `i` can fall into, such as the four seasons:

- `Spring` ($=1$ if `i` is in `Spring`, $=0$ if not)
- `Summer` ($=1$ if `i` is in Summer, $=0$ if not)
- `Fall` ($=1$ if `i` is in Fall, $=0$ if not)
- `Winter` ($=1$ if `i` is in Winter, $=0$ if not)

If we were to include all four dummy variables in the regression, each and every observation will have the following property:

$$Spring_i + Summer_i+ Fall_i +Winter_i=1$$

That is, for each and every observation `i`, the sum of the four dummy variable values must always equal 1 (because every observation must be either Spring, Summer, Fall, or Winter).

Our constant (formerly called the intercept), $\beta_0$, is the same value for every observation, as if we had a dummy variable $X_0$ that always equals 1 for each observation `i`. Now with the 4 season dummies, we have a perfectly collinear relationship between the constant and the sum of the dummies: 
$$1=X_{0i}=Spring_i+Summer_i+Fall_i+Winter_i$$

In any case, we need to drop one of these variables (either $X_0$ or one of the dummies). If we do this, then we break the collinearity. For example, if we drop $Summer_i$, then the sum of $Spring_i+Fall_i+Winter_i \neq 1$, since for observations in Summer, this equation$=0$.

---

## Question 2
**In your own words, describe what an interaction term is used for, and give an example. You can use any type of interaction to explain your answer.**

---

<!--WRITE YOUR ANSWERS BELOW -->
Suppose we have a model of: 
$$Y=\beta_0+\beta_1 X_1+ \beta_2 X_2$$

$\hat{\beta_1}$ tells us the marginal effect of $X_1 \rightarrow Y$ holding $X_2$ constant. However, it might be that the effect of $X_1 \rightarrow Y$ may be *different* for different values of $X_2$. We can explore this possibility with an interaction term, adding:

$$Y=\beta_0+\beta_1 X_1+ \beta_2 X_2+\beta_3 X_1 * X_2$$
Now, the marginal effect of on $Y$ of a change in $X_1$:

$$\frac{\Delta Y}{\Delta X_1}=\beta_1+\beta_3 X_2$$

So, the effect depends on what $X_2$ is. $\beta_3$ on the interaction term describes the *increment* to the effect of $X_1$ on $Y$ that depends on the value of $X_2$.

Note $X_1$ and $X_2$ could each be a continuous variable or a dummy variable, so we could have three possible combinations for an interaction effect:

- Two continuous variables (example above), e.g. `wage` and `experience`
- Two dummy variables, e.g. `male` ($=1$ if male, $=0$ if female) and `married` ($=1$ if married, $=0$ if unmarried)
- One continuous and one dummy variable (e.g. `male` and `experience`)

---

## Question 3

**In your own words, describe when and why using logged variables can be useful.**

---

<!--WRITE YOUR ANSWERS BELOW -->
Logged variables can help us out for theoretical reasons or data-fitting reasons.

Obviously, if a log function appears to fit nonlinear data better than a linear function (or even a polynomial), then using a log function is a good idea.

Sometimes we need functional forms that behave in a way that would be predicted by theory, such as diminishing returns, where the effect of something is always positive, but always growing smaller (such as the marginal product of labor or marginal utility). A log function always increases at a decreasing rate.

Additionally, logs can allow us to talk about percentage changes in variables, rather than unit changes. If we have both $X$ and $Y$ logged, we can talk about the elasticity between $X$ and $Y$, a 1% change in $X$ leads to a $\hat{\beta_1}$% change in $Y$.

---

## Question 4

**In your own words, describe when we would use an $F$-test, and give some example (null) hypotheses. Describe intuitively and specifically (no need for the formula) what exactly $F$ is trying to test for.**

---

<!--WRITE YOUR ANSWERS BELOW -->
$F$-tests are used to test \emph{joint}-hypotheses that hypothesizes values for \emph{multiple} parameters. This is as opposed to an ordinary $t$-test which only 	hypothesizes a value for one parameter (e.g. $H_0$: $\beta_1=0$).

We can run $F$-tests for a few different scenarios, but the most common is checking whether multiple variables are jointly-significant. The null hypothesis would be that several parameters are equal to 0, e.g.:
$$H_0 \text{: } \beta_1=0; \beta_2=0$$

The formula that calculates the value of the $F$-statistic essentially measures if the $R^2$ improves by a statistically significant amount when we move from the restricted regression (under the null hypothesis) to the unrestricted regression (where the null hypothesis is false).

For example, for a regression: 
$$\hat{Y_i}=\beta_0+\beta_1 X_{1i} + \beta_2 X_{2i} + \beta_3 X_{3i} + \beta_4 X_{4i}$$

the null hypothesis: 
$$H_0 \text{: } \beta_1=0; \beta_2=0;$$

would produce a restricted regression of: 
$$\hat{Y_i} = \beta_0+ \beta_3 X_{3i} + \beta_4 X_{4i}$$

and compare the $R^2$ of this regression to the $R^2$ of the original (unrestricted) regression. If the $R^2$ improves enough, the $F$-statistic is large enough, and hence we can reject the null hypothesis and conclude that $X_{1i}$ and $X_{2i}$ are jointly significant (we should not omit them from our regression).

---

# Theory Problems

For the following questions, please *show all work* and explain answers as necessary. You may lose points if you only write the correct answer. You may use `R` to *verify* your answers, but you are expected to reach the answers in this section "manually."

## Question 5

Suppose data on many countries' legal systems (Common Law or Civil Law) and their GDP per capita gives us the following summary statistics:

| Legal System | Avg. GDP Growth Rate | Std. dev. | $n$ |
|--------------|----------------------|-----------|-----|
| Common Law | $1.84$ | $3.55$ | $19$ |
| Civil Law | $4.97$ | $4.27$ | $141$ |
| Difference | $-3.13$ | $1.02$ | $-$ |

### Part A

**Using the group means, write a regression equation for a regression of GDP Growth rate on Common Law. Define**

$$\text{Common Law}_i = \begin{cases}
    1 & \text{if country } i \text{ has common law} \\
    0 & \text{if country } i \text{ has civil law}\\
    \end{cases}$$

---

<!--WRITE YOUR ANSWERS BELOW -->
$\widehat{\text{GDP Growth Rate}_i}=4.97-3.13 \, \text{Common Law}_i$

- $\hat{\beta_0}$: the average GDP growth rate for Civil Law countries
- $\hat{\beta_1}$: the difference in average GDP growth rate between Common Law and Civil Law countries

---

### Part B

**How do we use the regression to find the average GDP Growth rate for common law countries? For civil law countries? For the difference?**

---

<!--WRITE YOUR ANSWERS BELOW -->
This helps you construct the regression in part (a). The trick here is to recognize what each coefficient describes:

- $\beta_0$: average GDP Growth Rate for Civil Law countries (Common Law$=0)$
- $\beta_1$: difference in average GDP Growth Rates between Civil and Common Law countries
- $\beta_0+\beta_1$: average GDP Growth Rate for Common Law countries (Common Law$=1)$

---

### Part C

**Looking at the coefficients, does there appear to be a statistically significant difference in average GDP Growth Rates between Civil and Common law countries?**

---

<!--WRITE YOUR ANSWERS BELOW -->
For this, we want to look at the coefficient and standard error for $\beta_3$, which measures the difference in GDP Growth Rate between Civil and Common law countries. The standard error of $\hat{\beta_3}$ is given by the standard error of the difference in means (1.02). We run a $t$-test:

 $$t=\frac{-3.13}{1.02} \approx 3.07$$

This is a fairly large $t$-statistic, and so it is likely to be statistically significant.

---

### Part D

**Is the estimate on the difference likely to be unbiased? Why or why not?**

---

<!--WRITE YOUR ANSWERS BELOW -->
Whether or not a country is common law or civil law is likely to be endogenous. There are a host of other variables that are correlated both with a country's legal system and its GDP growth rate -- who it was colonized by, what its history is, geography, etc.

---

### Part E

**Now using the same table above, reconstruct the regression equation if instead of Common Law, we had used:**

$$\text{Civil Law}_i = \begin{cases}
    1 & \text{if country } i \text{ has civil law} \\
    0 & \text{if country } i \text{ has common law}\\
    \end{cases}$$

---

<!--WRITE YOUR ANSWERS BELOW -->

$$\widehat{\text{GDP Growth Rate}_i} = 1.84 + 3.13 \, \text{Civil Law}_i$$

---

## Question 6

Suppose a real estate agent collects data on houses that have sold in a particular neighborhood over the past year, with the following variables:

| Variable | Description |
|----------|-------------|
| $Price_h$ | price of house $h$ (in thousands of $) |
| $Bedrooms_h$ | number of bedrooms in house $h$ |
| $Baths_h$ | number of bathrooms in house $h$ |
| $Pool_h$ | $\begin{cases} =1 & \text{if house } h \text{ has a pool} \\ =0 & \text{if house } h \text{ does not have a pool} \end{cases}$ |
| $View_h$ |  $\begin{cases} =1 & \text{if house } h \text{ has a nice view} \\ =0 & \text{if house } h \text{ does not have a nice view} \end{cases}$ |

### Part A

**Suppose she runs the following regression:**
$$\widehat{\text{Price}_h}=119.20+29.76 \, \text{Bedrooms}_h+24.09 \, \text{View}_h+14.06 \, (\text{Bedrooms}_h \times \text{View}_h)$$

**What does each coefficient mean?**

---

<!--WRITE YOUR ANSWERS BELOW -->
- $\hat{\beta_0}=119.20$: Houses with 0 bedrooms and no view sell for $119.20 thousand
- $\hat{\beta_1}=29.76$: For every additional bedroom for houses with no view, price increases by $29.76 thousand
- $\hat{\beta_2}=24.09$: A house with 0 bedrooms and a view sells for $24.09 thousand higher than a house with 0 bedrooms and no view
- $\hat{\beta_3}=14.06$: A house with a view gains an $14.06 thousand *more* in price for every additional bedroom than does a house with no view

---

### Part B

**Write out _two_ separate regression equations, one for houses _*_with_ a nice view, and one for homes _without_ a nice view. Explain each coefficient in each regression.**

---

<!--WRITE YOUR ANSWERS BELOW -->

For houses without a nice view $(View_h=0)$:

$$\begin{align*}
	\widehat{Price_h} &= 119.20 +29.76 \, Bedrooms_h +24.09(\mathbf{0})+14.06\, Bedrooms_h(\mathbf{0})\\
		&=119.20+29.76\, Bedrooms_h\\
\end{align*}$$

Houses without a nice view sell for $119.20 thousand for no bedrooms, and gain $29.76 thousand for every additional bedroom.

For houses with a nice view $(View_h=1)$:

$$\begin{align*}
	\widehat{Price}_h &= 119.20 +29.76\, Bedrooms_h+24.09(\mathbf{1})+14.06\, Bedrooms_h(\mathbf{1})\\
		&=119.20+29.76\, Bedrooms_h+24.09+14.06\, Bedrooms_h\\
		&=(119.20+24.09)+(29.76+14.06)\, Bedrooms_h\\
		&=143.29+43.82\, Bedrooms_h\\
	\end{align*}$$

Houses with a nice view sell for $143.29 thousand for no bedrooms, and gain $43.82 thousand for every additional bedroom.

---

### Part C

**Suppose she runs the following regression:**

$$\widehat{\text{Price}_h}=189.20+42.40 \, \text{Pool}_h+12.10 \, \text{View}_h+12.09 \, (\text{Pool}_h \times \text{View}_h)$$

**What does each coefficient mean?**

---

<!--WRITE YOUR ANSWERS BELOW -->
- $\hat{\beta_0}=189.20$: Houses with no pool and no view sell for $189.20 thousand on average
- $\hat{\beta_1}=29.76$: Houses with no view gain $42.40 thousand for having a pool
- $\hat{\beta_2}=24.09$: Houses with no pool gain $12.10 thousand for having a view
- $\hat{\beta_3}=14.06$: Houses with a pool gain $12.09 thousand *more* for also having a view than a house with no pool also having a view. Note you could equivalently say houses with a view gain $12.09 thousand *more* for also having a pool than a house without a view also having a pool.

---

### Part D
**Find the expected price for:**

- a house with no pool and no view
- a house with no pool and a view
- a house with a pool and without a view
- a house with a pool and with a view

---

<!--WRITE YOUR ANSWERS BELOW -->
- No Pool & No View: $\beta_0=\$189.20$ thousand 
- No Pool & View: $\beta_0+\beta_2=189.20+12.10=\$201.30$ thousand
- Pool & No view: $\beta_0+\beta_1=189.20+42.40=\$231.60$ thousand
- Pool & View: $\beta_0+\beta_1+\beta_2+\beta_3=189.20+42.40+12.10+12.09=\$255.79$ thousand

---

### Part E

**Suppose she runs the following regression:**
$$\widehat{\text{Price}_h}=87.90+53.94 \, \text{Bedrooms}_h+15.29 \, \text{Baths}_h+16.19 \, (\text{Bedrooms}_h \times \text{Baths}_h)$$

**What is the marginal effect of adding an additional _bedroom_ if the house has 1 bathroom? 2 bathrooms? 3 bathrooms?**

---

<!--WRITE YOUR ANSWERS BELOW -->
- $\displaystyle\frac{\Delta Price}{\Delta Bed}|Bath=1=\beta_1+\beta_3(Bath)=53.94+16.19(1)=\$70.13$ thousand
- $\displaystyle\frac{\Delta Price}{\Delta Bed}|Bath=2=\beta_1+\beta_3(Bath)=53.94+16.19(2)=\$86.32$ thousand
- $\displaystyle\frac{\Delta Price}{\Delta Bed}|Bath=3=\beta_1+\beta_3(Bath)=53.94+16.19(3)=\$102.51$ thousand

---

### Part F

What is the marginal effect of adding an additional *bathroom* if the house has 1 bedroom? 2 bedrooms? 3 bedrooms?

---

<!--WRITE YOUR ANSWERS BELOW -->
- $\displaystyle\frac{\Delta Price}{\Delta Bath}|Bed=1=\beta_2+\beta_3(Bed)=15.29+16.19(1)=\$31.48$ thousand
- $\displaystyle\frac{\Delta Price}{\Delta Bath}|Bed=2=\beta_2+\beta_3(Bed)=15.29+16.19(2)=\$47.67$ thousand
- $\displaystyle\frac{\Delta Price}{\Delta Bath}|Bed=3=\beta_2+\beta_3(Bed)=15.29+16.19(3)=\$63.86$ thousand

---

## Question 7

Suppose we want to examine the change in average global temperature over time. We have data on the deviation in temperature from pre-industrial times (in Celcius), and the year.

### Part A

**Suppose we estimate the following simple model relating deviation in  temperature to year:**

$$\widehat{\text{Temperature}_i}=-10.46+0.006 \text{Year}_i$$

**Interpret the coefficient on Year (i.e. $\hat{\beta_1})$**

---

<!--WRITE YOUR ANSWERS BELOW -->
Every (additional) year, temperature increases by 0.006 degrees.

---

### Part B

**Predict the (deviation in) temperature for the year 1900 and for the year 2000.**

---

<!--WRITE YOUR ANSWERS BELOW -->
In 1900:
$$\begin{align*}
 		\widehat{\text{Temperature}_{1900}}&=-10.46+0.006 (1900)\\
 		&=-10.46+11.4\\
 		&=0.94\text{ degrees} \\
 \end{align*}$$

In 2000:
$$\begin{align*}
 		\widehat{\text{Temperature}_{2000}}&=-10.46+0.006 (2000)\\
 		&=-10.46+12\\
 		&=1.54\text{ degrees} \\
 	\end{align*}$$

---

### Part C

**Suppose we believe temperature deviations are increasing at an increasing rate, and introduce a quadratic term and estimate the following regression model:**

$$\widehat{\text{Temperature}_i}=155.68-0.116 \text{Year}_i+0.000044\text{Year}_i^2$$

**What is the marginal effect on (deviation in) global temperature of one additional year elapsing?**

---

<!--WRITE YOUR ANSWERS BELOW -->
The marginal effect is measured by the first derivative of the regression equation with respect to Year. But you can just remember the resulting formula and plug in the parameters:

$$\begin{align*}
	\frac{d \, Y}{d \, X} &= \beta_1+2\beta_2 X\\
	\frac{d \, Temperature}{d \, Year} &= -0.116+2(0.000044)\, Year\\
	&=0.116+0.000088 \, Year\\
	\end{align*}$$

---

### Part D

**Predict the marginal effect on temperature of one more year elapsing starting in 1900, and in 2000.**

---

<!--WRITE YOUR ANSWERS BELOW -->
For 1900:

$$\begin{align*}
 		\frac{d \, Temperature}{d \, Year} &=-0.116+0.000088(1900)\\
 		&=-0.116+0.1672\\
 		&=0.0512\text{ degrees} \\
\end{align*}$$

For 2000:

$$\begin{align*}
 		\frac{d \, Temperature}{d \, Year} &=-0.116+0.000088(2000)\\
 		&=-0.116+0.176\\
 		&=0.06\text{ degrees} \\
 \end{align*}$$

---

### Part E

**Our quadratic function is a $U$-shape. According to the model, at what year was temperature (deviation) at its minimum?**

---

<!--WRITE YOUR ANSWERS BELOW -->
We can set the derivative equal to 0, or you can just remember the formula and plug in the parameters:

$$\begin{align*}
	\frac{d Y}{d X}  &= \beta_1+2\beta_2 X\\
	0 &=\beta_1+2\beta_2 X\\
	-\beta_1&=2\beta_2 X\\
	-\frac{1}{2} \times \frac{\beta_1}{\beta_2}&=Year^*\\
	-\frac{1}{2} \times\frac{(-0.116)}{(0.000044)} &= Year^*\\
	-\frac{1}{2} \times -2636 & \approx Year^*\\
	1318 & \approx Year ^*\\
\end{align*}$$

---

## Question 8

Suppose we want to examine the effect of cell phone use while driving on traffic fatalities. While we cannot measure the amount of cell phone activity while driving, we do have a good proxy variable, the number of cell phone subscriptions (in 1000s) in a state, along with traffic fatalities in that state.

### Part A

**Suppose we estimate the following simple regression:**

$$\widehat{\text{fatalities}_i}=123.98+0.091\text{cell plans}_i$$

**Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$**

---

<!--WRITE YOUR ANSWERS BELOW -->
For every additional (thousand) cell phone plans, traffic fatalities increase by 0.091.

---

### Part B

**Now suppose we estimate the regression using a linear-log model:**

$$\widehat{\text{fatalities}_i}=-3557.08+515.81\text{ln(cell plans}_i)$$

**Interpret the coefficient on ln(cell plans) (i.e. $\hat{\beta_1})$**

---

<!--WRITE YOUR ANSWERS BELOW -->
A 1% increase in cell phone plans will increase traffic fatalities by $\frac{515.81}{100}=5.1581$.

---

### Part C

**Now suppose we estimate the regression using a log-linear model:**

$$\widehat{\text{ln(fatalities}_i)}=5.43+0.0001\text{cell plans}_i$$

**Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$**

---

<!--WRITE YOUR ANSWERS BELOW -->
For every additional (thousand) cell phone plans, traffic fatalities will increase by $0.001 \times 100\%=0.1\%$.

---

### Part D

**Now suppose we estimate the regression using a log-log model:**

$$\widehat{\text{ln(fatalities}_i)}=-0.89+0.85\text{ln(cell plans}_i)$$

**Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$**

---

<!--WRITE YOUR ANSWERS BELOW -->
This is an elasticity. A 1% increase in cell phone plans will cause a 0.85% increase in traffic fatalities.

---

### Part E

**Suppose we include several other variables into our regression and want to determine which variable(s) have the largest effects, a State's cell plans, population, or amount of miles driven. Suppose we decide to _standardize_ the data to compare units, and we get: **

$$\widehat{\text{fatalities}_i}=4.35+0.002\text{cell plans}^{std}-0.00007\text{population}^{std}+0.019\text{miles driven}^{std}$$

**Interpret the coefficients on cell plans, population, and miles driven. Which has the largest effect on fatalities?**

---

<!--WRITE YOUR ANSWERS BELOW -->
- A 1-standard-deviation-increase in cell plans will lead to a 0.002 standard deviation increase in fatalities. 
- A 1-standard-deviation-increase in population will lead to a 0.00007 standard deviation increase in fatalities. 
- A 1-standard-deviation-increase in miles driven will lead to a 0.019 standard deviation increase in fatalities.

It appears that if a State has more miles driven in it, it will have the strongest effect on fatalities.

---

### Part F

**Suppose we wanted to make the claim that it is _only_ miles driven, and neither population nor cell phones determine traffic fatalities. Write (i) the null hypothesis for this claim and (ii) the estimated restricted regression equation.***

---

<!--WRITE YOUR ANSWERS BELOW -->
$$H_0 \text{: } \beta_1=\beta_2=0$$

Restricted Regression: $\widehat{\text{fatalities}_i}=4.35+0.019\text{miles driven}^{std}$

---

### Part G

**Suppose the $R^2$ on the original regression from (e) was 0.9221, and the $R^2$ from the restricted regression is 0.9062. With 50 observations, calculate the $F$-statistic.**

---

<!--WRITE YOUR ANSWERS BELOW -->
Note $q=2$ as we are hypothesizing values for $2$ parameters, $k=3$ as we have 3 variables in our unrestricted regression (cell plans, population, and miles driven).

$$\begin{align*}
		F_{q,n-k-1}&=\cfrac{\displaystyle\frac{(R^2_u-R^2_r)}{q}}{\displaystyle\frac{(1-R^2_u)}{(n-k-1)}}\\
		& \\ 
		F_{2,50-3-1}&=\cfrac{\displaystyle\frac{(0.9221-0.9062)}{2}}{\displaystyle\frac{(1-0.9221)}{(50-3-1)}}\\
		& \\ 
		F_{2,46}&=\cfrac{\displaystyle\frac{(0.0159)}{2}}{\displaystyle\frac{(0.0779)}{(46)}}\\
		&=\frac{0.00795}{0.00169}\\
		&\approx 4.70\\
\end{align*}$$

The number here does not mean anything literally. The purpose of this exercise is to get you to try to understand exactly how $F$ is calculated and how to think about it. Essentially, we are seeing if the $R^2$ on the unrestricted model has *statistically significantly increased* from the $R^2$ on the restricted model. The larger the $F$-statistic is, the larger the increase. We can also check statistical significance by calculating the associated $p$-value.

```{r}
# FYI: We could estimate the p-value in R if we wanted
#     - similar to calculating p-value for a t-statistic
#     - syntax is: 
# pf(F.value, df1, df2, lower.tail = FALSE) 
#     - we plug in our calculated F-statistic (above) for F.value
#     - df1 and df2 are degrees of freedom from numerator and denominator
#     - lower.tail=FALSE implies we are looking at probability to the RIGHT  
#       of F-value on F-distribution, p(f>F) that f is higher than our F
pf(4.70, 2,46, lower.tail=FALSE)
```

---

# R Questions

Answer the following questions using `R`. When necessary, please write answers in the same document (knitted `Rmd` to `html` or `pdf`, typed `.doc(x)`, or handwritten) as your answers to the above questions. Be sure to include (email or print an `.R` file, or show in your knitted `markdown`) your code and the outputs of your code with the rest of your answers.

## Question 9

- [<i class="fas fa-table"></i> `LeadMortality.csv`](http://metricsf19.classes.ryansafner.com/data/LeadMortality.csv)

Lead is toxic, particularly for young children, and for this reason government regulations severely restrict the amount of lead in our environment. In the early part of the 20^th^ century, the underground water pipes in many U.S. cities contained lead, and lead from these pipes leached into drinking water. This exercise will have you investigate the effect of these lead pipes on infant mortality. This dataset contains data on:

| Variable | Description |
|----------|-------------|
| `infrate` | infant mortality rate (deaths per 100 in population) |
| `lead` | $=1$ if city has lead water pipes, $=0$ if did not have lead pipes |
| `pH` | water pH |

 and several demographic variables for 172 U.S. cities in 1900.

### Part A

Part A

**Using `R` to examine the data, find the average infant mortality rate for cities with lead pipes and for cities without lead pipes. Calculate the difference, and run a $t$-test to determine if this difference is statistically significant.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# load tidyverse
library("tidyverse")

# read in data
lead<-read_csv("http://metricsf19.classes.ryansafner.com/data/leadmortality.csv")
```

```{r}
# mean of infrate for cities with lead
mean_lead<-lead %>%
    filter(lead==1) %>%
    summarize(mean(infrate)) %>%
    pull() # to save as number

# mean of infrate for cities with no lead
mean_no_lead<-lead %>%
    filter(lead==0) %>%
    summarize(mean(infrate)) %>%
    pull() # to save as number

mean_lead-mean_no_lead
```

Cities with lead pipes have an infant mortality rate of 0.40, and cities without lead pipes have an infant mortality rate of 0.38. So the difference is 0.02.

```{r}
# run t-test of difference
t.test(infrate ~ lead, data = lead)
```

We get a $t$-statistic of $-0.90$ and a $p$-value of $0.3681$, so the difference is *not* statistically significant.

---

### Part B

**Run a regression of `infrate` on `lead`, and write down the estimated regression equation. Use the regression coefficients to find:**

- the average infant mortality rate for cities with lead pipes
- the average infant mortality rate for cities without lead pipes
- the difference between the averages for cities with or without lead pipes

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
lead_reg1<-lm(infrate ~ lead, data = lead)
summary(lead_reg1)
```

$$\widehat{\text{Infrate}_i} = 0.38+0.02 \, \text{Lead}_i$$

- Cities *without* lead pipes have an infant mortality rate of 0.38 $(\hat{\beta_0})$
- Cities *with* lead pipes have an infant mortality rate of $0.38+0.02=0.40$ $(\hat{\beta_0}+\hat{\beta_1})$
- The difference is $0.02$ $(\hat{\beta_3})$

---

### Part C

**Does the pH of the water matter? Include `ph` in your regression from part B. Write down the estimated regression equation, and interpret each coefficient (note there is no interaction effect here). What happens to the estimate on `lead`?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
lead_reg2<-lm(infrate ~ lead + ph, data = lead)
summary(lead_reg2)
```

$$\widehat{\text{Infrate}_i}=1.17+0.05 \, \text{Lead}_i-0.11 \, \text{pH}_i$$

- $\hat{\beta_0}:$ the infant mortality rate for cities with a pH of 0, holding the type of pipe constant is 1.17
- $\hat{\beta_1}$: the infant mortality rate for cities with lead pipes is 0.05 higher than cities without lead pipes, holding pH constant
- $\hat{\beta_2}$: the infant mortality rate falls by 0.11 for every increase of 1 in the city water's pH (less acidic), holding the type of pipe constant

The estimate on `lead` doubled and became significant at the 5% level.

---

### Part D

**The amount of lead leached from lead pipes normally depends on the chemistry of the water running through the pipes: the more acidic the water (lower pH), the more lead is leached. Create an interaction term between lead and pH, and run a regression of `infrate` on `lead`, `pH`, and your interaction term. Write down the estimated regression equation. Is this interaction significant?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
lead_reg3<-lm(infrate ~ lead + ph + lead:ph, data = lead)
summary(lead_reg3)
```

$$\widehat{\text{Infrate}_i}=0.92+0.46 \, \text{Lead}_i-0.08 \, \text{pH}_i- \, 0.06 \, (\text{Lead} \times \text{pH})$$

We see that this interaction is just barely insignificant, with a $p$-value of 0.06.

---

### Part E

**What we actually have are two different regression lines. Visualize this with a scatterplot between `infrate` $(Y)$ and `ph` $(X)$ by `lead`.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r, fig.retina=3}
lead_scatter<-ggplot(data = lead)+
    aes(x = ph,
        y = infrate)+
    geom_point(aes(color = as.factor(lead)))+ # making it a factot makes color discrete rather than continuous!
    geom_smooth(method = "lm")+
    # now I'm just making it pretty
    # changing color
    scale_color_viridis_d("Pipes",
                          labels=c("0"="Not Lead", "1"="Lead"))+ # changing labels for colors
  labs(x = "pH",
       y = "Infant Mortality Rate")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
lead_scatter
```

```{r}
lead_scatter+facet_wrap(~lead, 
                        labeller=labeller(lead=c("0"="Not Lead", "1"="Lead")))+ # change facet titles
    guides(color = F) # hide other legend
```

---

### Part F

**Do the two regression lines have the same intercept? The same slope? Use the original regression in part D to test these possibilities.**

---

<!--WRITE YOUR ANSWERS BELOW -->

$\beta_1$ (on `lead`) measures the difference in intercept between the lead & no lead regression lines. So we would want to test:

$$\begin{align*}
	H_0: & \beta_1=0\\
	H_a: & \beta_1 \neq 0\\	
\end{align*}$$

The `R` output tells us $\hat{\beta_1}$ is $0.46$ with a standard error of 0.22, so the $t$-statistic for this test is $2.09$ with a $p$-value of 0.04, so there is a statistically significant difference at the 5% level.

$\beta_3$ (on the interaction term) measures the difference in slope between the lead & no lead regression lines. So we would want to test:

$$\begin{align*}
	H_0: & \beta_3=0\\
	H_a: & \beta_3 \neq 0\\	
\end{align*}$$

The output tells us $\hat{\beta_3}$ is $-0.06$ with a standard error of 0.3, so the $t$-statistic for this test is $-1.87$ with a $p$-value of $0.06$, so there is not a statistically significant difference at the 5% level.

Therefore, they have *difference* intercepts, and the *same* slopes, statistically.

---

### Part G

**Take your regression equation from part D and rewrite it as two separate regression equations (one for no lead and one for lead). Interpret the coefficients for each.**

---

<!--WRITE YOUR ANSWERS BELOW -->

For no lead (`lead=0`):

$$\begin{align*}
	\widehat{Infrate}&=0.92+0.46 \, Lead-0.08 \, pH-0.06\, (Lead \times pH)\\
		&=0.92+0.46(\mathbf{0})-0.08pH-0.06((\mathbf{0}) \times pH)\\
		&=0.92-0.08 \, pH\\
\end{align*}$$

For lead (`lead=1`):

$$\begin{align*}
	\widehat{Infrate}&=0.92+0.46 \, Lead-0.08 \, pH-0.06\, (Lead \times pH)\\
		&=0.92+0.46(\mathbf{1})-0.08pH-0.06((\mathbf{1}) \times pH)\\
		&=(0.92+0.46)+(-0.08-0.06) \, pH\\
		&=1.30-0.14 \, pH\\
\end{align*}$$

Cities without lead pipes have an infant mortality rate of $0.92$ $(\hat{\beta_0}$ from reg 1) vs. cities with lead pipes have an infant mortality rate of $1.30$ ($\hat{\beta_0}$ from reg 2). For every additional unit of pH, the infant mortality rate of a city without lead pipes would decrease by 0.06 $(\hat{\beta_1}$ from reg 1) vs. a city with lead pipes which would see a fall of 0.14.

So again, we can see the difference in infant mortality rates between cities with lead pipes vs. those that don't, with a water pH of 0 is $1.30-0.92=\$0.38$, $(\hat{\beta_0}$ from the regression in (part d)), and the cities with lead raise their infant mortality rates by $0.14-0.08=0.06$ more from each unit of pH than cities without lead do $(\hat{\beta_3})$ from the regression in (part d)).

---

### Part H

**Double check your calculations in G are correct by running the regression in D twice, once for cities without lead pipes and once for cities with lead pipes.**^[`filter()` the data first, then use the filtered data for the `data=` in each regression.]

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# regression for no lead

lead %>%
    filter(lead==0) %>%
    lm(data = ., infrate ~ lead + ph + lead:ph) %>%
    summary()

# regression for lead

lead %>%
    filter(lead==1) %>%
    lm(data = ., infrate ~ lead + ph + lead:ph) %>%
    summary()
```

---

### Part I

**Use `huxtable` to make a nice output table of all of your regressions from parts B, C, and D.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
library(huxtable)
huxreg(lead_reg1,
       lead_reg2,
       lead_reg3,
       coefs = c("Constant" = "(Intercept)",
                 "Lead Pipes" = "lead",
                 "pH" = "ph",
                 "Lead * pH" = "lead:ph"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```

---

## Question 10

- [<i class="fas fa-table"></i> `freedom.csv`](http://metricsf19.classes.ryansafner.com/data/freedom.csv)

Let's look at economic freedom and GDP per capita using some data I sourced from [Gapminder](https://gapminder.org/tools)^[GDP per capita (2018)], [Freedom House](https://freedomhouse.org/content/freedom-world-data-and-resources)^[Political freedom score (2018)] and [Fraser Institute Data](https://www.fraserinstitute.org/economic-freedom/dataset?geozone=world&year=2016&page=dataset)^[Economic Freedom score (2016)] and cleaned up for you, with the following variables:

| Variable | Description |
|----------|-------------|
| `Country` | Name of country |
| `ISO` | Code of country (good for plotting) |
| `econ_freedom` | Economic Freedom Index score (2016) from 1 (least) to 10 (most free) |
| `pol_freedom` | Political freedom index score (2018) from 1 (least) top 10 (most free) |
| `gdp_pc` | GDP per capita (2018 USD) |
| `continent` | Continent of country |

### Part A

**Does economic freedom affect GDP per capita? Create a scatterplot of `gdp_pc` (`Y`) against `econ_freedom` (`x`). Does the effect appear to be linear or nonlinear?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# load data
freedom<-read_csv("https://metricsf19.classes.ryansafner.com/data/freedom.csv")
```

```{r, fig.retina=3}
# scatterplot

freedom_plot<-ggplot(data = freedom)+
    aes(x = econ_freedom,
        y = gdp_pc)+
    geom_point(aes(color = continent))+
    scale_y_continuous(labels=scales::dollar)+
  labs(x = "Economic Freedom Score (0-10)",
       y = "GDP per Capita")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
freedom_plot
```

The effect appears to be nonlinear.

---

### Part B

**Run a simple regression of `gdp_pc` on `econ_freedom`. Write out the estimated regression equation. What is the marginal effect of `econ_freedom` on `gdp_pc`?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
freedom_reg1<-lm(gdp_pc ~ econ_freedom, data = freedom)

summary(freedom_reg1)
```

---

### Part C

**Let's try a quadratic model. Run a quadratic regression of `gdp_pc` on `econ_freedom`. Write out the estimated regression equation. **

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
freedom_reg2<-lm(gdp_pc ~ econ_freedom + I(econ_freedom^2), data = freedom)

summary(freedom_reg2)
```

---

### Part D

**Add the quadratic regression to your scatterplot.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r,fig.retina=3}
freedom_plot+geom_smooth(method = "lm", formula = "y~x+I(x^2)", color = "green")
```

---

### Part E

**What is the marginal effect of `econ_freedom` on `gdp_pc`?**

---

<!--WRITE YOUR ANSWERS BELOW -->

The marginal effect can be found by taking the derivative of the regression with respect to `econ_freedom` (or recalling the rule):

$$\begin{align*}
	\frac{d \, Y}{d \, X} &= \beta_1+2\beta_2 X\\
	\frac{d \, GDPpc}{d \, econfreedom} &= -96618+2(8327)econfreedom\, econ\\
	&=-96618+16654 \, econfreedom\\
	\end{align*}$$

```{r}
# if you want to calculate it in R

library(broom)

freedom_reg2_tidy<-tidy(freedom_reg2)

freedom_beta_1<-freedom_reg2_tidy %>%
    filter(term == "econ_freedom") %>%
    pull(estimate)

freedom_beta_2<-freedom_reg2_tidy %>%
    filter(term == "I(econ_freedom^2)") %>%
    pull(estimate)

# freedom_beta_1+2*freedom_beta_2* # number
```

---

### Part F

**As a quadratic model, this relationship should predict an` econ_freedom` score where `gdp_pc` is at a *minimum*. What is that minimum Economic Freedom score, and what is the associated GDP per capita?**

---

<!--WRITE YOUR ANSWERS BELOW -->

We can set the derivative equal to 0, or you can just remember the formula and plug in the parameters:

$$\begin{align*}
	\frac{d Y}{d X}  &= \beta_1+2\beta_2 X\\
	0 &=\beta_1+2\beta_2 X\\
	-\beta_1&=2\beta_2 X\\
	-\frac{1}{2} \times \frac{\beta_1}{\beta_2}&=econfreedom^*\\
	-\frac{1}{2} \times\frac{(-96618)}{(8327)} &= econfreedom^*\\
	-\frac{1}{2} \times -11.603 & \approx econfreedom^*\\
	5.801 & \approx econfreedom ^*\\
\end{align*}$$

```{r}
# to calculate in R

min<--0.5*(freedom_beta_1/freedom_beta_2)
min
```

```{r,fig.retina=3}
# let's visualize on the scatterplot

freedom_plot+geom_smooth(method = "lm", formula = "y~x+I(x^2)", color = "green")+
  geom_vline(xintercept = min, linetype = "dashed", size = 1)+
  geom_label(x = min, y = 75000, label = round(min,2))

```

---

### Part G

**Run a cubic model to see if we should keep going up in polynomials. Write out the estimated regression equation. Should we add a cubic term?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
freedom_reg3<-lm(gdp_pc ~ econ_freedom + I(econ_freedom^2)+ I(econ_freedom^3), data = freedom)

summary(freedom_reg3)
```

There's no good theoretical reason why we should expect economic freedom to "change direction" twice - go down, then up, then down again - in its effect on GDP.

Statistically, we can see that $\hat{\beta_3}$ on `I(econ_freedom^3)` is not significant (`p-value` is `0.590`), so we should *not* include the cubic term.

```{r}
# let's visualize it on the scatterplot
freedom_plot+
  geom_smooth(method = "lm", formula = "y~x+I(x^2)", color = "green")+
  geom_smooth(method = "lm", formula = "y~x+I(x^2)+I(x^3)", color = "orange")
```

---

### Part H

**Another way we can _test_ for non-linearity is to run an $F$-test on all non-linear variables - i.e. the quadratic term and the cubic term $(\hat{\beta_2}$ and $\hat{\beta_3}$) and test against the null hypothesis that:**
$$H_0: \hat{\beta_2} = \hat{\beta_3} = 0$$

**Run this joint hypothesis test, and what can you conclude?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# run F test
library(car)
linearHypothesis(freedom_reg3, c("I(econ_freedom^2)", "I(econ_freedom^3)"))
```

The null hypothesis is that the polynomial terms (quadratic and cubic) jointly do not matter (and the relationship is therefore linear). We have sufficient evidence to reject that hypothesis (`p-value` is very small). Thus, the relationship is in fact *not* linear.

---

### Part I

**Instead of a polynomial model, try out a logarithmic model. It is hard to interpret percent changes on an index, but it is easy to understand percent changes in GDP per capita, so run a _log-linear_ regression. Write out the estimated regression equation. What is the marginal effect of `econ_freedom`?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# log linear model 
freedom_reg4<-lm(log(gdp_pc) ~ econ_freedom, data = freedom)

summary(freedom_reg4)
```

For every 1 point increase on the economic freedom index, a country's GDP per capita increases by $1.2889 \times 100\%=128.90\%$

---

### Part J

**Make a scatterplot of your log-linear model with a regression line.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r, fig.retina=3}
log_freedom_plot<-ggplot(data = freedom)+
    aes(x = econ_freedom,
        y = log(gdp_pc))+
    geom_point(aes(color = continent))+
    geom_smooth(method = "lm")+
  labs(x = "Economic Freedom Score (0-10)",
       y = "Log GDP per Capita")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
log_freedom_plot
```

---

### Part K

**Put all of your results together in a regression output table with `huxtable` from your answers in questions B, C, G, and H.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
huxreg("GDP per Capita" = freedom_reg1,
       "GDP per Capita" = freedom_reg2,
       "Log(GDP per Capita)" = freedom_reg3,
       coefs = c("Constant" = "(Intercept)",
                 "Economic Freedom Score (0-10)" = "econ_freedom",
                 "Economic Freedom Squared" = "I(econ_freedom^2)",
                 "Economic Freedom Cubed" = "I(econ_freedom^3)"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```

### 2019-12-06 static/homeworks/06-problem-set-answers.Rmd
Add HW 5 answers, HW 6 and answers, finalize schedule page (put extra content after final), add final exam prep

---
title: "Problem Set 6 (Answers)"
author: "Ryan Safner" # put your name here! 
date: "ECON 480 - Fall 2019"
output:
  html_document:
    df_print: paged
    #theme: simplex
    toc: true 
    toc_depth: 3
    toc_float: true
    code_folding: show
    highlight: tango
---

<!--CLICK "KNIT" ABOVE TO RENDER TO HTML, PDF, OR WORD OUTPUT
In fact, try knitting right away and see what this produces!
-->

*Due by Thursday, November 21, 2019*

# Instructions

Answers may be longer than I would deem sufficient on an exam. Some might vary slightly based on points of interest, examples, or personal experience. These suggested answers are designed to give you both the answer and a short explanation of *why* it is the answer.

# Theory and Concepts

## Question 1
**In your own words, describe what *fixed effects* are, when we can use them, and how they remove endogeneity.**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

## Question 2
**In your own words, describe the logic of a *difference-in-difference* model: what is it comparing against what, and how does it estimate the effect of treatment? What assumption must be made about the treatment and control group for the model to be valid?**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

# R Questions

Answer the following questions using `R`. When necessary, please write answers in the same document (knitted `Rmd` to `html` or `pdf`, typed `.doc(x)`, or handwritten) as your answers to the above questions. Be sure to include (email or print an `.R` file, or show in your knitted `markdown`) your code and the outputs of your code with the rest of your answers.

## Question 3

- [<i class="fas fa-table"></i> `PeaceCorps.csv`](http://metricsf19.classes.ryansafner.com/data/PeacceCorps.csv)

How do people respond to changes in economic conditions? Are they more likely to pursue public service when private sector jobs are scarce? This dataset contains variables at the U.S. State (& D.C.) level:

| Variable | Description |
|----------|-------------|
| `state` | U.S. State |
| `year` | Year |
| `appspc` | Applications to the Peace Corps (per capita) in State |
| `unemployrate` | State unemployment rate |

> Do more people apply to the Peace Corps when unemployment increases (and reduces other opportunities)?

### Part A

**Before looking at the data, what does your economic intuition tell you? Explain your hypothesis.**

---

<!--WRITE YOUR ANSWERS BELOW -->

If joining the Peace Corps is a substitute for private sector jobs, then as unemployment rises, so too should applications. Although, it could also be possible that people also are more willing to opt out of the workforce when the economy is strong, so we should examine this empirically to be sure.

---

### Part B

**To get the hang of the data we're working with, `count` (separately) the number of `state`s, and the number of `year`s. Get the number of `n_distinct()` `state`s and `year`s^[Do this inside the `summarize()` command], as well as the `distinct()` values of each^[Don't use the `summarize()` command for this part].**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
library(tidyverse)

# load in data
peacecorps<-read_csv("https://metricsf19.classes.ryansafner.com/data/PeaceCorps.csv")
```

```{r}
# count number of states
peacecorps %>%
  count(state)

# count number of years
peacecorps %>%
  count(year)

# get number of distinct states and years
peacecorps %>%
  summarize(Num_states = n_distinct(state),
            Num_years = n_distinct(year))

# get distinct values of states
peacecorps %>%
  distinct(state)

# get distinct values of years
peacecorps %>%
  distinct(year)
```

---

### Part C

**Continuing our pre-analysis inspection, (install, and) load the `plm` package, and check the dimensions of the data with `pdim`.**^[Set `index=c("state","year")` to indicate the group and time dimensions.]

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# install.packages("plm")

# load plm
library(plm)

pdim(peacecorps, index=c("state", "year"))
```

---

### Part D

**Create a scatterplot of `appspc` (Y) on `unemployrate` (X). Which State is an outlier? How would this affect the pooled regression estimates? Create a *second* scatterplot that does not include this State.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r, fig.retina=3}
ggplot(data = peacecorps)+
    aes(x = unemployrate,
        y = appspc)+
    geom_point(aes(color = as.factor(state)))+
    geom_smooth(method = "lm")+
  scale_x_continuous(breaks=seq(0,20,2),
                     labels = function(x){paste(x,"%", sep="")})+
  labs(x = "Unemployment Rate",
       y = "Peace Corps Applications (per capita)")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=18)+
  theme(legend.position = "none") # remove legend for State colors
```

We can see there are very clear outliers at the top. Let's plot `text` instead of `point`s, using the `stateshort` to see which observations are which states:

```{r, fig.retina=3}
ggplot(data = peacecorps)+
    aes(x = unemployrate,
        y = appspc)+
    geom_text(aes(color = as.factor(state), label = stateshort))+ #<<
    geom_smooth(method = "lm")+
  scale_x_continuous(breaks=seq(0,20,2),
                     labels = function(x){paste(x,"%", sep="")})+
  labs(x = "Unemployment Rate",
       y = "Peace Corps Applications (per capita)")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=18)+
  theme(legend.position = "none") # remove legend for State colors
```

Clearly `DIS`, which is District of Columbia, are the outliers. Let's make a second scatterplot without them:

```{r, fig.retina=3}
peacecorps %>%
  filter(state != "DISTRICT OF COLUMBIA") %>%
ggplot(data = .)+
    aes(x = unemployrate,
        y = appspc)+
    geom_point(aes(color = as.factor(state)))+
    geom_smooth(method = "lm")+
  scale_x_continuous(breaks=seq(0,20,2),
                     labels = function(x){paste(x,"%", sep="")})+
  labs(x = "Unemployment Rate",
       y = "Peace Corps Applications (per capita)")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=18)+
  theme(legend.position = "none") # remove legend for State colors
```

---

### Part E

**Run two *pooled* regressions, one with the outliers, and one without them. Write out the estimated regression equation for each. Interpret the coefficient, and comment on how it changes between the two regressions.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
pooled<-lm(appspc ~ unemployrate, data = peacecorps)
summary(pooled)
```

$$\widehat{\text{Apps per capita}_{it}}=46.23+0.84 \, \text{Unemployment Rate}_{it}$$

For every 1 percentage point increase in unemployment, there are 0.46 more applications to the Peace Corps per capita.

```{r}
peacecorps_no_outliers<-peacecorps %>%
  filter(state != "DISTRICT OF COLUMBIA")

pooled_no_outliers<-lm(appspc ~ unemployrate, data = peacecorps_no_outliers)
summary(pooled_no_outliers)
```

$$\widehat{\text{Apps per capita}_{it}}=48.59-0.37 \, \text{Unemployment Rate}_{it}$$

For every 1 percentage point increase in unemployment, there are 0.37 *fewer* applications to the Peace Corps per capita.

The coefficient changes signs and becomes a smaller magnitude by dropping the outliers (DC)!

---

### Part F

**Now run a regression with State fixed effects using the dummy variable method.^[Ensure that `state` is a factor variable, and insert in the regression. You can either `mutate()` it into a `factor` beforehand, or just do `as.factor(state)` in the `lm` command.] Interpret the marginal effect of `unemployrate` on `appspc`. How did it change?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r, fig.retina=3}
fe_reg<-lm(appspc ~ unemployrate + factor(state), data = peacecorps)
summary(fe_reg)
```

For every 1 percentage point increase in the unemployment rate, there are 0.66 more applications to the Peace Corps per capita. It remained positive, and shrank in size, from the original regression (with outliers).

---

### Part G

**Find the coefficient for Maryland and interpret it. How many applications per capita does Maryland have?**

---

<!--WRITE YOUR ANSWERS BELOW -->

This is just like any model with dummy variables representing different categories. The constant (12.61) represents the *reference category* not put in the regression (to avoid the dummy variable trap), i.e. `ALABAMA.`

The coefficient on Maryland is 45.19. This means that, on average, there are 45.19 *more* applications per capita in Maryland *than Alabama* over the time period in the data. To get the average for Maryland, we add $12.61+45.19=57.80$.

```{r}
# if we want to extract it using R
library(broom)

fe_reg_tidy<-tidy(fe_reg)

AL<-fe_reg_tidy %>%
  filter(term == "(Intercept)") %>%
  pull(estimate)

MD_AL_diff<-fe_reg_tidy %>%
  filter(term == "factor(state)MARYLAND") %>%
  pull(estimate)

AL+MD_AL_diff
```

---

### Part H

**Now try using the `plm()` command, which de-means the data, and make sure you get the same results as Part F.^[Inside `plm()`, set `index = "state"` to indicate variable, and `model = "within"` to indicate a fixed effects model.] Do you get the same marginal effect of `unemployrate` on `appspc`?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
fe_reg_alt<-plm(appspc ~ unemployrate, index = "state", model = "within", data = peacecorps)
summary(fe_reg_alt)
```

---

### Part I

**Now include *year* fixed effects in your regression, using the dummy variable method. Interpret the marginal effect of `unemployrate` on `appspc`. How did it change?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
fe_2way_reg<-lm(appspc ~ unemployrate + factor(state) + factor(year), data = peacecorps)
summary(fe_2way_reg)
```

For every 1 percentage point increase in the unemployment rate, there are 0.71 applications per capita to the Peace Corps.

---

### Part I

**What would be the predicted number of applications in Maryland in 2011 at an unemployment rate of 5%?**

---

<!--WRITE YOUR ANSWERS BELOW -->

$$\begin{align*}
\widehat{\text{Apps per capita}_{it}}&=14.49 +0.71 \, \text{Unemployment Rate}_{it} + 45.24 \, \text{Maryland_i} -8.89 \text{2007}_t \\
\widehat{\text{Apps per capita}_{MD, 2011}}&=14.49 +0.71 \, (5) + 45.24 \, (1) -8.89 (1) \\
\widehat{\text{Apps per capita}_{MD, 2011}}& \approx 54.09
\end{align*}$$

```{r}
# if we want to extract it using R
fe_2way_reg<-tidy(fe_2way_reg)

constant<-fe_2way_reg %>%
  filter(term == "(Intercept)") %>%
  pull(estimate)

unemploy_me<-fe_2way_reg %>%
  filter(term == "unemployrate") %>%
  pull(estimate)

MD_dif<-fe_2way_reg %>%
  filter(term == "factor(state)MARYLAND") %>%
  pull(estimate)

dif_2007<-fe_2way_reg %>%
  filter(term == "factor(year)2007") %>%
  pull(estimate)

constant+unemploy_me*5+MD_dif+dif_2007

# some rounding error
```

---

### Part K

**Now try using the `plm()` command, which de-means the data, and make sure you get the same results as Part I.^[Inside `plm()`, set `index = c("state", "year")` to indicate both variables, and `effect = "twoways"` to indicate a 2-way fixed effects model.] Do you get the same marginal effect of `unemployrate` on `appspc`?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
fe_2way_reg_alt<-plm(appspc ~ unemployrate, index = c("state", "year"), data = peacecorps, effect = "twoways")
summary(fe_2way_reg_alt)
```

---

### Part L

**Can there still be endogeneity in this model? Give some examples.**

---

<!--WRITE YOUR ANSWERS BELOW -->

Yes. There may be other time-varying variables that are omitted and correlated with unemployment, but not picked up in the time-fixed effect since they are not the same in each state. Examples could be a local market bubble some years in Texas or Nevada, but not in Maine, or bad weather in Florida one year, but not in Wyoming, etc.^[It's better to use the `plm()`-generated regressions so as to avoid the multitude of coefficients for the state and year dummies! You certainly could use the dummy variable ones and manually list all of the variables to suppress in the table inside `omit_coefs()`...]

---

### Part M
**Create a nice regression table (using `huxtable`) for comparison of the regressions in E, G, and I.**

---

<!--WRITE YOUR ANSWERS BELOW -->
```{r}
library(huxtable)
huxreg(pooled,
       pooled_no_outliers,
       fe_reg_alt,
       fe_2way_reg_alt,
         coefs = c("Constant" = "(Intercept)",
                 "Unemployment Rate" = "unemployrate"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       note = NULL, # suppress footnote for stars, to insert Fixed Effects Row below
       number_format = 3) %>%

add_rows(c("Outliers", "Yes", "No DC", "Yes", "Yes"), # add row for outliers
         after = 5) %>% # insert after 5th row
add_rows(c("Fixed Effects", "None", "None", "State", "State and Year"), # add fixed effects row
         after = 6) # insert after 6th row

```

## Question 4

- [<i class="fas fa-table"></i> `TexasSchools.csv`](http://metricsf19.classes.ryansafner.com/data/TexasSchools.csv)

Are teachers paid more when school board members are elected "off cycle" when there are not major national political elections (e.g. odd years) than "on cycle?" The argument is that during "off" years, without attention on state or national elections, voters will pay less attention to the election, and teachers can more effectively mobilize for higher pay, versus "on" years where voters are paying more attention. This data comes from Anzia, Sarah (2012), "The Election Timing Effect: Evidence from a Policy Intervention in Texas." *Quarterly Journal of Political Science* 7(3): 277-297, and follows 1,020 Texas school board districts from 2003–2009.

From 2003–2006, all districts elected their school board members off-cycle. A change in Texas policy in 2006 led some, but not all, districts to elect their school board members on-cycle from 2007 onwards.

| Variable | Description |
|----------|-------------|
| `LnAvgSalary` | logged average salary of teachers in district |
| `Year` | Year |
| `OnCycle` | `=1` if school boards elected on-cycle (e.g. same year as national and state elections), `=0` if elected off-cycle |
| `pol_freedom` | Political freedom index score (2018) from 1 (least) top 10 (most free) |
| `CycleSwitch` | `=1` if district switched from off- to on-cycle elections |
| `AfterSwitch` | `=1` if year is after 2006 |

### Part A

**Run a pooled regression model of `LnAvgSalary` on `OnCycle`. Write the estimated regression equation, and interpret the coefficient on `OnCycle.` Are there any sources of bias (consider in particular the argument in the question prompt)?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# load in data
texas<-read_csv("https://metricsf19.classes.ryansafner.com/data/TexasSchools.csv")
```

```{r, fig.retina=3}
pooled_reg<-lm(LnAvgSalary~OnCycle, data = texas)
summary(pooled_reg)
```

$$\widehat{\text{log(Average Salary})_it} = 10.67-0.03 \, \text{OnCycle}_{it}$$

Recognize this is a linear-log model and $X$ is a dummy variable. School board elections that are held `OnCycle` $(=1)$ lead to $\hat{\beta_1 \times 100\% = −0.03 \times 100\% = −3\%$ lower teacher salaries. This is highly statistically significant.

This is almost certainly biased. There are probably things correlated at the district level between both whether or not districts vote (or switch to voting) on cycle/off cycle and with teacher salaries. Perhaps some districts have strong/weak teachers unions and that determines whether they choose to vote on or off cycle (off cycle would be preferable, and perhaps districts with strong teachers unions kept the elections off cycle).

Also, there could be changes in the entire State over time-perhaps all Texas teacher salaries were increasing or decreasing over the years.

---

### Part B

**Some schools decided to switch to an on-cycle election after 2006. Consider this, `CycleSwitch` the "treatment." Create a variable to indicate post-treatment years (i.e. years after 2006). Call it `After`. Create a second, *interaction* variable to capture the interaction effect between those districts that *switched*, and *after* the treatment.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# create a variable, that is a factor, using ifelse() function
# # if the year is greater than 2006, mark "After" as 1,
# # If the year is NOT greater than 2006, market "After" a 0
texas<-texas %>%
  mutate(After = factor(ifelse(test = Year>2006,
                        yes = 1,
                        no = 0)))
head(texas)
```

---

### Part C

**Now estimate a difference-in-difference model with your variables in Part B: `CycleSwitch` is the treatment variable, `After` is your post-treatment indicator, and add an _interaction_ variable to capture the interaction effect between those districts that _switched_, and _after_ the treatment. Write down the estimated regression equation (to four decimal places). **

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
dnd<-lm(LnAvgSalary ~ CycleSwitch + After + CycleSwitch:After, data = texas)
summary(dnd)
```

$$\widehat{\text{Log(Average Salary)}_{it}}=10.6711-0.0240 \, \text{Cycle Switch}_{i} + 0.0093 \, \text{After}_t - 0.0086 \, (\text{Cycle Switch}_i \times \text{After}_t)$$

---

### Part D

**Interpret what each coefficient means from Part C.**

---

<!--WRITE YOUR ANSWERS BELOW -->

- $\hat{\beta_0}$: districts before 2007 that did not switch had `ln(AvgSalary)` of 10.6711
- $\hat{\beta_1}$: districts before 2007 that did switch had `ln(AvgSalary)` 0.0240 lower
than districts that did not switch
- $\hat{\beta_2}$: all districts after the change had `ln(AvgSalary)` 0.0093 higher than before the change
- $\hat{\beta_3}$: districts that made a switch had a 0.0086 *lower* change in `ln(AvgSalary)` over time than districts that did not switch

---

### Part E

**Using your regression equation in Part C, calculate the expected logged average salary $(Y)$ for districts in Texas:**

- i. *Before* the switch that did *not* switch
- ii. *After* the switch that did *not* switch
- iii. *Before* the switch that *did* switch
- iv. *After* the switch that *did* switch

---

<!--WRITE YOUR ANSWERS BELOW -->

- i. Before change and did not switch: $\hat{\beta_0} = 10.6711$
- ii. After change and did not switch: $\hat{\beta_0}+\hat{\beta_2} = 10.6711 + 0.0093 = 10.6804$
- iii. Before change and did switch: $\hat{\beta_0}+\hat{\beta_1} = 10.6711 − 0.0240 = 10.6471$
- iv. After change and did switch: $\hat{\beta_0}+\hat{\beta_1}+\hat{\beta_2}+\hat{\beta_3} = 10.6711−0.0240+ 0.0093 − 0.0086 = 10.6478$

---

### Part F

**Confirm your estimates in Part E by finding the mean logged average salary for each of those four groups in the data.^[Hint, `filter()` properly then `summarize()`.]**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# avg log salary for districts BEFORE that did NOT switch
texas %>%
  filter(CycleSwitch == 0,
         After == 0) %>%
  summarize(average = mean(LnAvgSalary, na.rm=T)) # need to remove NA's

# avg log salary for districts AFTER that did NOT switch
texas %>%
  filter(CycleSwitch == 0,
         After == 1) %>%
  summarize(average = mean(LnAvgSalary, na.rm=T)) # need to remove NA's

# avg log salary for districts BEFORE that DID switch
texas %>%
  filter(CycleSwitch == 1,
         After == 0) %>%
  summarize(average = mean(LnAvgSalary, na.rm=T)) # need to remove NA's

# avg log salary for districts AFTER that DID switch
texas %>%
  filter(CycleSwitch == 1,
         After == 1) %>%
  summarize(average = mean(LnAvgSalary, na.rm=T)) # need to remove NA's
```

---

### Part G

**Write out the difference-in-difference equation, and calculate the difference-in-difference. Make sure it matches your estimate from the regression.**

---

<!--WRITE YOUR ANSWERS BELOW -->

$$\begin{align*}
		\Delta \Delta ln(AvgSalary)&=(Switched_{after}-Switched_{before})-(Didn't_{after}-Didn't_{before})\\
		&=(10.6478-10.6471)-(10.6804-10.6711)\\
		&=(0.0007)-(0.0093)\\
		&=-0.0086\\ \end{align*}$$

This is $\beta_3$.

---

### Part H

**Can we say anything about the types of districts that switched? Can we say anything about all salaries in the districts in the years after the switch?**

---

<!--WRITE YOUR ANSWERS BELOW -->

We can see that the districts that switched paid 2.3% less to their teachers to begin with. $\beta_2$ shows the difference, before treatment, between the treated (districts that switched) and control groups (districts that did not switch).

We can see that all districts saw a 0.93% increase in teacher salaries after the years after the switch. $\beta_3$ shows the difference, for both the treatment and control, of salaries over time (before and after the switch).

---

### Part I

**Now let’s generalize the diff-in-diff model. Instead of the treatment and post-treatment dummies, use district-and year-fixed effects and the interaction term. Interpret the coefficient on the interaction term**^[This is doable with the dummy variable method, but there will be a *lot* of dummies! I suggest using `plm()`.]

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# make sure DistNumber and Year are factors for plm
texas<-texas %>%
  mutate_at(c("DistNumber", "Year"), factor)
head(texas) # make sure it worked

dnd_fe_reg<-plm(LnAvgSalary ~ CycleSwitch*After, index=c("DistNumber", "Year"), effect = "twoways", data = texas)
summary(dnd_fe_reg)
```

Districts that switched to On Cycle saw teacher salaries decrease by 8.67%. No change from our original model!

---

### Part J

**Create a nice regression table (using `huxtable`) for comparison of the regressions in (a), (c), and (i).**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
huxreg(pooled_reg,
       dnd,
       dnd_fe_reg,
         coefs = c("Constant" = "(Intercept)",
                 "On Cycle" = "OnCycle",
                 "Switched" = "CycleSwitch",
                 "After Switch" = "After1",
                 "Interaction" = "CycleSwitch:After1"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       note = NULL, # suppress footnote for stars, to insert Fixed Effects Row below
       number_format = 3) %>%
add_rows(c("Fixed Effects", "None", "None", "District and Year"), # add fixed effects row
         after = 11) # insert after 11th row
```

### 2019-12-06 static/homeworks/05-problem-set-answers.R

@@ -1,0 +1,208 @@
+# 9 -------------------
+
+## A -------------
+# load tidyverse
+library("tidyverse")
+
+# read in data
+lead<-read_csv("http://metricsf19.classes.ryansafner.com/data/leadmortality.csv")
+
+# mean of infrate for cities with lead
+mean_lead<-lead %>%
+  filter(lead==1) %>%
+  summarize(mean(infrate)) %>%
+  pull() # to save as number
+
+# mean of infrate for cities with no lead
+mean_no_lead<-lead %>%
+  filter(lead==0) %>%
+  summarize(mean(infrate)) %>%
+  pull() # to save as number
+
+mean_lead-mean_no_lead
+
+# run t-test of difference
+t.test(infrate ~ lead, data = lead)
+
+## B -------------
+
+lead_reg1<-lm(infrate ~ lead, data = lead)
+summary(lead_reg1)
+
+
+## C -------------
+
+lead_reg2<-lm(infrate ~ lead + ph, data = lead)
+summary(lead_reg2)
+
+## D -------------
+
+lead_reg3<-lm(infrate ~ lead + ph + lead:ph, data = lead)
+summary(lead_reg3)
+
+## E -------------
+
+lead_scatter<-ggplot(data = lead)+
+  aes(x = ph,
+      y = infrate)+
+  geom_point(aes(color = as.factor(lead)))+ # making it a factot makes color discrete rather than continuous!
+  
+  # now I'm just making it pretty
+  # changing color
+  scale_color_viridis_d("Pipes",
+                        labels=c("0"="Not Lead", "1"="Lead"))+ # changing labels for colors
+  labs(x = "pH",
+       y = "Infant Mortality Rate")+
+  theme_classic(base_family = "Fira Sans Condensed",
+                base_size=20)
+lead_scatter
+
+lead_scatter+facet_wrap(~lead, 
+                        labeller=labeller(lead=c("0"="Not Lead", "1"="Lead")))+ # change facet titles
+  guides(color = F) # hide other legend
+
+## H -------------
+
+# regression for no lead
+
+lead %>%
+  filter(lead==0) %>%
+  lm(data = ., infrate ~ lead + ph + lead:ph) %>%
+  summary()
+
+# regression for lead
+
+lead %>%
+  filter(lead==1) %>%
+  lm(data = ., infrate ~ lead + ph + lead:ph) %>%
+  summary()
+
+## I -------------
+
[…]

### 2019-12-06 static/homeworks/06-problem-set-answers.Rmd
Add HW 6 empty markdown file, answers to #1-2 in HW 6 answers

@@
 <!--WRITE YOUR ANSWERS BELOW -->
+Fixed effects can be used for panel data, where each observation$_{it}$ belongs to a group $i$ and a time period $t$.
+Running a simple OLS model as usual, known as a **pooled model** would be biased due to systematic relationships within and between groups and time periods that cause $X_{it}$ to be endogenous:
+
+$$\widehat{Y_{it}}=\beta_0+\beta_1 X_{it}+u_{it}$$
+
+Group-fixed effects $(\alpha_i$) pull out of the error term all of the factors that explain $Y_{it}$ that are stable over time within each individual group ($i$), but vary across groups.
+For example, if groups are U.S.
+states, state-fixed effects pull out all of the error term all of the idiosyncrasies of each state that do not change over time, such as cultural differences, geographical differences, historical differences, population differences, etc.
+Group fixed effects *do not* pull out factors that change over time, such as federal laws passed, recessions affecting all states, etc.
+
+Time-fixed effects $(\theta_i)$ pull out of the error term all of the factors that explain $Y_{it}$ that change over time but do not vary across groups.
+For example, if groups are U.S.
+states, and time is in years, year-fixed effects pull out all of the error term all of the changes over the years that affect all U.S.
+states, such as recessions, inflation, national population growth, national immigration trends, or federal laws passed that affect all states.
+Time-fixed effects *do not* pull out factors that do not change over time.
+
+$$\widehat{Y_{it}}=\beta_0+\beta_1 X_{it}+ \alpha_{i} + \theta_{t} + \epsilon_{it}$$
+
+Mechanically, OLS estimates a separate constant for each group (and/or each time period), giving the expected value of $Y_{it}$ for that group or time-period.
+This can be done by either de-meaning the data and calculating OLS coefficients by exploiting the variation *within* each group and/or time period (which is why fixed effects estimators are called "within" estimators), or by creating a dummy variable for each group and/or time period (and omitting one of each, to avoid the dummy variable trap).
 
@@
 <!--WRITE YOUR ANSWERS BELOW -->
+$$\widehat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$
+
+A difference-in-difference model compares the difference before and after a treatment between a group that receives the treatment and a group that does not.
+By doing so, we can isolate the effect of the treatment.
+It is easiest to see in the following equation:
+
+$$\Delta \Delta Y = (\text{Treated}_{after}-\text{Treated}_{before})-(\text{Control}_{after}-\text{Control}_{before})$$
+
+In OLS regression, we can simply make two dummy variables for observations$_it$ depending on the group and the time period each observation is:
+
+$$\text{Treated}_i = \begin{cases}
+        1 & \text{if } i \text{ receives treatment} \\
+   			0 & \text{if } i \text{ does not receive treatment }\\
+  		\end{cases}$$
+
+$$\text{After}_t = \begin{cases}
+    		1 & \text{if } t \text{is after treatment period} \\
+   			0 & \text{if } t \text{is before treatment period}\\
+  		\end{cases}$$
+
+Lastly, we make an interaction term $\text{Treated}_i * \text{After}_t$ which isolates the treatment effect, captured by the coefficient on this term, $\beta_3$.
+
+Diff-in-diff models assume a counterfactual that if the treated group did not receive treatment, it would have experience the same change as the control group over the pre- to post-treatment period.
+
+A classic example is if a state(s) passes a law and others do not.
+We want to compare the difference in the difference before and after the law was passed between states that passed the law and states that did not:
+
+$$\Delta_i \Delta_t Outcome = (\text{Passed}_{after}-\text{Passed}_{before})-(\text{Didn't}_{after}-\text{Didn't}_{before})$$
+
+"Treatment$_i$" and time-fixed effects for "After$_t$" to estimate the treatment effect (still $\beta_3$): 
+$$\widehat{Y_{it}}=\alpha_i +\theta_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+\epsilon_{it}$$
 

### 2019-12-06 static/homeworks/06-problem-set.Rmd
Add HW 6 empty markdown file, answers to #1-2 in HW 6 answers

--- 2019-12-06 static/homeworks/06-problem-set-answers.Rmd
@@
 
-If joining the Peace Corps is a substitute for private sector jobs, then as unemployment rises, so too should applications.
-Although, it could also be possible that people also are more willing to opt out of the workforce when the economy is strong, so we should examine this empirically to be sure.
+```{r}
+# Write your code here
+```
 
@@
 ```{r}
-library(tidyverse)
-
-# load in data
-peacecorps<-read_csv("https://metricsf19.classes.ryansafner.com/data/PeaceCorps.csv")
-```
-
-```{r}
-# count number of states
-peacecorps %>%
-  count(state)
-
-# count number of years
-peacecorps %>%
-  count(year)
-
-# get number of distinct states and years
-peacecorps %>%
-  summarize(Num_states = n_distinct(state),
-            Num_years = n_distinct(year))
-
-# get distinct values of states
-peacecorps %>%
-  distinct(state)
-
-# get distinct values of years
-peacecorps %>%
-  distinct(year)
+# Write your code here
 ```
@@
 ```{r}
-# install.packages("plm")
-
-# load plm
-library(plm)
-
-pdim(peacecorps, index=c("state", "year"))
+# Write your code here
 ```
@@
 
-```{r, fig.retina=3}
-ggplot(data = peacecorps)+
-    aes(x = unemployrate,
-        y = appspc)+
-    geom_point(aes(color = as.factor(state)))+
-    geom_smooth(method = "lm")+
-  scale_x_continuous(breaks=seq(0,20,2),
-                     labels = function(x){paste(x,"%", sep="")})+
-  labs(x = "Unemployment Rate",
-       y = "Peace Corps Applications (per capita)")+
-  theme_classic(base_family = "Fira Sans Condensed",
-           base_size=18)+
-  theme(legend.position = "none") # remove legend for State colors
-```
-
-We can see there are very clear outliers at the top. Let's plot `text` instead of `point`s, using the `stateshort` to see which observations are which states:
-
-```{r, fig.retina=3}
-ggplot(data = peacecorps)+
-    aes(x = unemployrate,
-        y = appspc)+
-    geom_text(aes(color = as.factor(state), label = stateshort))+ #<<
-    geom_smooth(method = "lm")+
-  scale_x_continuous(breaks=seq(0,20,2),
-                     labels = function(x){paste(x,"%", sep="")})+
-  labs(x = "Unemployment Rate",
-       y = "Peace Corps Applications (per capita)")+
-  theme_classic(base_family = "Fira Sans Condensed",
-           base_size=18)+
-  theme(legend.position = "none") # remove legend for State colors
-```
-
-Clearly `DIS`, which is District of Columbia, are the outliers. Let's make a second scatterplot without them:
-
-```{r, fig.retina=3}
-peacecorps %>%
-  filter(state != "DISTRICT OF COLUMBIA") %>%
-ggplot(data = .)+
-    aes(x = unemployrate,
-        y = appspc)+
-    geom_point(aes(color = as.factor(state)))+
-    geom_smooth(method = "lm")+
-  scale_x_continuous(breaks=seq(0,20,2),
-                     labels = function(x){paste(x,"%", sep="")})+
-  labs(x = "Unemployment Rate",
-       y = "Peace Corps Applications (per capita)")+
-  theme_classic(base_family = "Fira Sans Condensed",
-           base_size=18)+
-  theme(legend.position = "none") # remove legend for State colors
+```{r}
+# Write your code here
 ```
@@
 ```{r}
-pooled<-lm(appspc ~ unemployrate, data = peacecorps)
-summary(pooled)
+# Write your code here
 ```
-
-$$\widehat{\text{Apps per capita}_{it}}=46.23+0.84 \, \text{Unemployment Rate}_{it}$$
-
-For every 1 percentage point increase in unemployment, there are 0.46 more applications to the Peace Corps per capita.
-
-```{r}
-peacecorps_no_outliers<-peacecorps %>%
-  filter(state != "DISTRICT OF COLUMBIA")
-
-pooled_no_outliers<-lm(appspc ~ unemployrate, data = peacecorps_no_outliers)
-summary(pooled_no_outliers)
-```
-
-$$\widehat{\text{Apps per capita}_{it}}=48.59-0.37 \, \text{Unemployment Rate}_{it}$$
-
-For every 1 percentage point increase in unemployment, there are 0.37 *fewer* applications to the Peace Corps per capita.
-
-The coefficient changes signs and becomes a smaller magnitude by dropping the outliers (DC)!
 
@@
 
-```{r, fig.retina=3}
-fe_reg<-lm(appspc ~ unemployrate + factor(state), data = peacecorps)
-summary(fe_reg)
+```{r}
+# Write your code here
 ```
-
-For every 1 percentage point increase in the unemployment rate, there are 0.66 more applications to the Peace Corps per capita.
-It remained positive, and shrank in size, from the original regression (with outliers).
 
@@
 
-This is just like any model with dummy variables representing different categories.
-The constant (12.61) represents the *reference category* not put in the regression (to avoid the dummy variable trap), i.e.
-`ALABAMA.`
-
-The coefficient on Maryland is 45.19.
-This means that, on average, there are 45.19 *more* applications per capita in Maryland *than Alabama* over the time period in the data.
-To get the average for Maryland, we add $12.61+45.19=57.80$.
-
 ```{r}
-# if we want to extract it using R
-library(broom)
-
-fe_reg_tidy<-tidy(fe_reg)
-
-AL<-fe_reg_tidy %>%
-  filter(term == "(Intercept)") %>%
-  pull(estimate)
-
-MD_AL_diff<-fe_reg_tidy %>%
-  filter(term == "factor(state)MARYLAND") %>%
-  pull(estimate)
-
-AL+MD_AL_diff
+# Write your code here
 ```
@@
 ```{r}
-fe_reg_alt<-plm(appspc ~ unemployrate, index = "state", model = "within", data = peacecorps)
-summary(fe_reg_alt)
+# Write your code here
 ```
@@
 ```{r}
-fe_2way_reg<-lm(appspc ~ unemployrate + factor(state) + factor(year), data = peacecorps)
-summary(fe_2way_reg)
+# Write your code here
 ```
-
-For every 1 percentage point increase in the unemployment rate, there are 0.71 applications per capita to the Peace Corps.
 
@@
 
-$$\begin{align*}
-\widehat{\text{Apps per capita}_{it}}&=14.49 +0.71 \, \text{Unemployment Rate}_{it} + 45.24 \, \text{Maryland_i} -8.89 \text{2007}_t \\
-\widehat{\text{Apps per capita}_{MD, 2011}}&=14.49 +0.71 \, (5) + 45.24 \, (1) -8.89 (1) \\
-\widehat{\text{Apps per capita}_{MD, 2011}}& \approx 54.09
-\end{align*}$$
-
 ```{r}
-# if we want to extract it using R
-fe_2way_reg<-tidy(fe_2way_reg)
-
-constant<-fe_2way_reg %>%
-  filter(term == "(Intercept)") %>%
-  pull(estimate)
-
-unemploy_me<-fe_2way_reg %>%
-  filter(term == "unemployrate") %>%
-  pull(estimate)
-
-MD_dif<-fe_2way_reg %>%
-  filter(term == "factor(state)MARYLAND") %>%
-  pull(estimate)
-
-dif_2007<-fe_2way_reg %>%
-  filter(term == "factor(year)2007") %>%
-  pull(estimate)
-
-constant+unemploy_me*5+MD_dif+dif_2007
-
-# some rounding error
+# Write your code here
 ```
@@
 ```{r}
-fe_2way_reg_alt<-plm(appspc ~ unemployrate, index = c("state", "year"), data = peacecorps, effect = "twoways")
-summary(fe_2way_reg_alt)
+# Write your code here
 ```
@@
 
-Yes.
-There may be other time-varying variables that are omitted and correlated with unemployment, but not picked up in the time-fixed effect since they are not the same in each state.
-Examples could be a local market bubble some years in Texas or Nevada, but not in Maine, or bad weather in Florida one year, but not in Wyoming, etc.^[It's better to use the `plm()`-generated regressions so as to avoid the multitude of coefficients for the state and year dummies!
-You certainly could use the dummy variable ones and manually list all of the variables to suppress in the table inside `omit_coefs()`...]
+```{r}
+# Write your code here
+```
 
@@
 <!--WRITE YOUR ANSWERS BELOW -->
+
 ```{r}
-library(huxtable)
-huxreg(pooled,
-       pooled_no_outliers,
-       fe_reg_alt,
-       fe_2way_reg_alt,
-         coefs = c("Constant" = "(Intercept)",
-                 "Unemployment Rate" = "unemployrate"),
-       statistics = c("N" = "nobs",
-                      "R-Squared" = "r.squared",
-                      "SER" = "sigma"),
-       note = NULL, # suppress footnote for stars, to insert Fixed Effects Row below
-       number_format = 3) %>%
-
-add_rows(c("Outliers", "Yes", "No DC", "Yes", "Yes"), # add row for outliers
-         after = 5) %>% # insert after 5th row
-add_rows(c("Fixed Effects", "None", "None", "State", "State and Year"), # add fixed effects row
-         after = 6) # insert after 6th row
-
+# Write your code here
 ```
@@
 ```{r}
-# load in data
-texas<-read_csv("https://metricsf19.classes.ryansafner.com/data/TexasSchools.csv")
+# Write your code here
 ```
-
-```{r, fig.retina=3}
-pooled_reg<-lm(LnAvgSalary~OnCycle, data = texas)
-summary(pooled_reg)
-```
-
-$$\widehat{\text{log(Average Salary})_it} = 10.67-0.03 \, \text{OnCycle}_{it}$$
-
-Recognize this is a linear-log model and $X$ is a dummy variable.
-School board elections that are held `OnCycle` $(=1)$ lead to $\hat{\beta_1 \times 100\% = −0.03 \times 100\% = −3\%$ lower teacher salaries.
-This is highly statistically significant.
-
-This is almost certainly biased.
-There are probably things correlated at the district level between both whether or not districts vote (or switch to voting) on cycle/off cycle and with teacher salaries.
-Perhaps some districts have strong/weak teachers unions and that determines whether they choose to vote on or off cycle (off cycle would be preferable, and perhaps districts with strong teachers unions kept the elections off cycle).
-
-Also, there could be changes in the entire State over time-perhaps all Texas teacher salaries were increasing or decreasing over the years.
 
@@
 ```{r}
-# create a variable, that is a factor, using ifelse() function
-# # if the year is greater than 2006, mark "After" as 1,
-# # If the year is NOT greater than 2006, market "After" a 0
-texas<-texas %>%
-  mutate(After = factor(ifelse(test = Year>2006,
-                        yes = 1,
-                        no = 0)))
-head(texas)
+# Write your code here
 ```
@@
 ```{r}
-dnd<-lm(LnAvgSalary ~ CycleSwitch + After + CycleSwitch:After, data = texas)
-summary(dnd)
+# Write your code here
 ```
-
-$$\widehat{\text{Log(Average Salary)}_{it}}=10.6711-0.0240 \, \text{Cycle Switch}_{i} + 0.0093 \, \text{After}_t - 0.0086 \, (\text{Cycle Switch}_i \times \text{After}_t)$$
 
@@
 
-- $\hat{\beta_0}$: districts before 2007 that did not switch had `ln(AvgSalary)` of 10.6711
-- $\hat{\beta_1}$: districts before 2007 that did switch had `ln(AvgSalary)` 0.0240 lower
-than districts that did not switch
-- $\hat{\beta_2}$: all districts after the change had `ln(AvgSalary)` 0.0093 higher than before the change
-- $\hat{\beta_3}$: districts that made a switch had a 0.0086 *lower* change in `ln(AvgSalary)` over time than districts that did not switch
+```{r}
+# Write your code here
+```
 
@@
 
-- i. Before change and did not switch: $\hat{\beta_0} = 10.6711$
-- ii. After change and did not switch: $\hat{\beta_0}+\hat{\beta_2} = 10.6711 + 0.0093 = 10.6804$
-- iii. Before change and did switch: $\hat{\beta_0}+\hat{\beta_1} = 10.6711 − 0.0240 = 10.6471$
-- iv. After change and did switch: $\hat{\beta_0}+\hat{\beta_1}+\hat{\beta_2}+\hat{\beta_3} = 10.6711−0.0240+ 0.0093 − 0.0086 = 10.6478$
+```{r}
+# Write your code here
+```
 
@@
 ```{r}
-# avg log salary for districts BEFORE that did NOT switch
-texas %>%
-  filter(CycleSwitch == 0,
-         After == 0) %>%
-  summarize(average = mean(LnAvgSalary, na.rm=T)) # need to remove NA's
-
-# avg log salary for districts AFTER that did NOT switch
-texas %>%
-  filter(CycleSwitch == 0,
-         After == 1) %>%
-  summarize(average = mean(LnAvgSalary, na.rm=T)) # need to remove NA's
-
-# avg log salary for districts BEFORE that DID switch
-texas %>%
-  filter(CycleSwitch == 1,
-         After == 0) %>%
-  summarize(average = mean(LnAvgSalary, na.rm=T)) # need to remove NA's
-
-# avg log salary for districts AFTER that DID switch
-texas %>%
-  filter(CycleSwitch == 1,
-         After == 1) %>%
-  summarize(average = mean(LnAvgSalary, na.rm=T)) # need to remove NA's
+# Write your code here
 ```
@@
 
-$$\begin{align*}
-		\Delta \Delta ln(AvgSalary)&=(Switched_{after}-Switched_{before})-(Didn't_{after}-Didn't_{before})\\
-		&=(10.6478-10.6471)-(10.6804-10.6711)\\
-		&=(0.0007)-(0.0093)\\
-		&=-0.0086\\ \end{align*}$$
-
-This is $\beta_3$.
+```{r}
+# Write your code here
+```
 
@@
 
-We can see that the districts that switched paid 2.3% less to their teachers to begin with.
-$\beta_2$ shows the difference, before treatment, between the treated (districts that switched) and control groups (districts that did not switch).
-
-We can see that all districts saw a 0.93% increase in teacher salaries after the years after the switch.
-$\beta_3$ shows the difference, for both the treatment and control, of salaries over time (before and after the switch).
+```{r}
+# Write your code here
+```
 
@@
 ```{r}
-# make sure DistNumber and Year are factors for plm
-texas<-texas %>%
-  mutate_at(c("DistNumber", "Year"), factor)
-head(texas) # make sure it worked
-
-dnd_fe_reg<-plm(LnAvgSalary ~ CycleSwitch*After, index=c("DistNumber", "Year"), effect = "twoways", data = texas)
-summary(dnd_fe_reg)
+# Write your code here
 ```
-
-Districts that switched to On Cycle saw teacher salaries decrease by 8.67%. No change from our original model!
 
@@
 ```{r}
-huxreg(pooled_reg,
-       dnd,
-       dnd_fe_reg,
-         coefs = c("Constant" = "(Intercept)",
-                 "On Cycle" = "OnCycle",
-                 "Switched" = "CycleSwitch",
-                 "After Switch" = "After1",
-                 "Interaction" = "CycleSwitch:After1"),
-       statistics = c("N" = "nobs",
-                      "R-Squared" = "r.squared",
-                      "SER" = "sigma"),
-       note = NULL, # suppress footnote for stars, to insert Fixed Effects Row below
-       number_format = 3) %>%
-add_rows(c("Fixed Effects", "None", "None", "District and Year"), # add fixed effects row
-         after = 11) # insert after 11th row
+# Write your code here
 ```

### 2019-12-10 static/slides/final-review.Rmd
Add final review

---
title: "Final Review Questions"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F)
library(tidyverse)
set.seed(256)
update_geom_defaults("label", list(family = "Fira Sans Condensed"))
```

# Major Models and Extensions

- Multivariate OLS
    - Omitted Variable Bias
    - Variance/Multicollinearity
    - Causal inference/DAGs, controls, proxies
- Categorical data
    - Using categorical variables as dummies
    - dummy variable trap
    - linear probability model $(Y$ is a dummy)
    - interaction terms $X_1 \times X_2$
- Nonlinear Models
    - quadratic model, polynomial models, logarithmic models
- Panel Data
    - fixed effects (two-way group effects and time effects)
    - difference-in-difference models

---

# Question 1

What are the two conditions for a variable $Z$ to cause .shout[omitted variable bias] if it is left out of the regression?

---

# Question 2

$$\widehat{Wages}_i=\beta_0+\beta_1 \, Education_i + \beta_2 \, Age_i + \beta_3 \, Experience_i + \epsilon_i$$

$Education_i$ and $Age_i$ are highly correlated

--

- Does this bias $\hat{\beta_1}$ and $\hat{\beta_2}$?

--

- What will happen to the variance of $\hat{\beta_2}$ and $\hat{\beta_2}$? How can we measure this?

---

# Question 3

$$\widehat{Cholesterol}=\beta_0+\beta_1Treated+u_i$$

- $Treated_i$ is a dummy variable $=1$ if person received treatment, $=0$ if did not

--
- What is $\hat{\beta_0}$?

--

- What is $\hat{\beta_1}$?

--

- What is the average cholesterol level for someone who recieved treatment?

---

# Question 4

$$\widehat{Y}_i=\beta_0+\beta_1 \, Red_i+\beta_2 \, Orange_i+\beta_3 \, Yellow_i+\beta_4 \, Green_i+\beta_5 \, Blue_i$$

Suppose observation $i$ can be either $\{$Red, Orange, Yellow, Green, Blue, Purple $\}$

--

- What is $\hat{\beta_0}$?

--

- What is $\hat{\beta_1}$?

--

- What is the average value of $Y_i$ for $Green$ shapes?

--

- Why can't we add $\beta_6 \, Purple_i$?

---

# Question 5

$$\widehat{Utility}_i=\beta_0+\beta_1 \, Eggs_i+\beta_2 \, Breakfast_i+\beta_3 (Eggs_i \times Breakfast_i)$$

$Breakfast_i$ is a dummy variable $=1$ if meal is breakfast, $=0$ if not

--

- What is $\hat{\beta_1}$?

--

- What is $\hat{\beta_2}$?

--

- What is $\hat{\beta_3}$?

--

- We have two regressions (one for Breakfast; one for Not Breakfast)
    - how can we determine if the intercepts are different?
    - how can we determine if the slopes are different?

---

# Question 6

$$\widehat{Utility}=2+4\text{ Ice Cream Cones}_i-1\text{ Ice Cream Cones}_i^2$$

--

- What is the marginal effect of eating 1 more Ice Cream Cone?

--

- What if we *start* with 1 Ice Cream Cone?

--

- What if we *start* with 4 Ice Cream Cones?
--

- What amount of ice cream cones will *maximize* utility?

--

- How would we know if we should add $\text{Ice Cream Cones}_i^3$?

---

# Question 7

$$ln(GDP_i)=10+2\text{ population (in billions)}_i$$

- Interpret $\hat{\beta_1}$ in context.

--

$$ln(GDP_i)=10+0.1 \, ln(\text{population}_i)$$

- Interpret $\hat{\beta_1}$ in context.

---

# Question 8

- Explain what an $F$-test is used for

--

- Explain *how* an $F$-statistic is generated

---

# Question 9

$$\widehat{\text{Divorce Rate}_{it}}=\beta_0+\beta_1 \text{Divorce Law}_{it}+\alpha_i+\theta_t+\epsilon_{it}$$

--

- Why do we need $\alpha_i$ and $\theta_t$?

--

- What sorts of things are in $\alpha_i$?

--

- What sorts of things are in $\theta_t$?

---

# Question 10

$$\widehat{\text{Crime Rate}_{it}}=\beta_0+\beta_1 \, \text{Maryland}_{i}+\beta_2 \, \text{After}_t+\beta_3  \, (\text{Maryland}_i \times \text{After}_t)$$

- Suppose Maryland passes a law (and other States do not) that affects crime rates

--

- What must we assume about Maryland over time?

--

- What is the average crime rate for other states before the law?

--

- What is the average crime rate for Maryland after the law?

--

- What is the *causal effect* of passing the law?

### 2019-12-14 content/assignment/paper-project.Rmd
Add presentation example, class 23 content
Add link to example empirical paper

@@
 - [<i class="fas fa-external-link-square-alt"></i> Writing and Reading Empirical Papers Slides](/slides/paper-slides.html)
+- [<i class="fas fa-github"></i> Example Paper Project](https://github.com/ryansafner/example_empirical_paper)
 
@@
 Using some slides to show tables, models, and plots is highly recommended.
+  - [<i class="fas fa-file-pdf"></i> Example](/papers/Presentation_example.pdf)
 

### 2019-12-14 content/assignment/paper-project.Rmd
Change github logo link on empirical example

--- 2019-11-19 content/assignment/paper-project.Rmd
@@
 - [<i class="fas fa-external-link-square-alt"></i> Writing and Reading Empirical Papers Slides](/slides/paper-slides.html)
+- [<i class="fab fa-github"></i> Example Paper Project](https://github.com/ryansafner/example_empirical_paper)
 
